### 2023-10-24 paper/lensbans.tex
Adjoints folds

\newsavebox{\lbansbox}
\newcommand{\lbansmacro}{%
\begin{tikzpicture}[baseline=0.45ex,xscale=0.006em,yscale=0.012ex]%
% curvey bananas
%\draw[solid,join=round,fill=yellow] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9) to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;%
% fitted lenses
\draw[solid,join=round] (1.8,6) -- (1.5,5.9) to[out=-120, in=90] (0.7,3) to[out=-90, in=120] (1.5,0.1) -- (1.8,0) -- cycle;%
\end{tikzpicture}}
\savebox{\lbansbox}{\lbansmacro}
\newcommand{\lbans}{\mathopen{\usebox{\lbansbox}\mspace{1mu}}}

\newsavebox{\rbansbox}
\newcommand{\rbansmacro}{%
\begin{tikzpicture}[baseline=0.45ex,xscale=-0.006em,yscale=0.012ex]%
% curvey bananas
%\draw[solid,join=round,fill=yellow] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9) to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;%
% fitted lenses
\draw[solid,join=round] (1.8,6) -- (1.5,5.9) to[out=-120, in=90] (0.7,3) to[out=-90, in=120] (1.5,0.1) -- (1.8,0) -- cycle;%
\end{tikzpicture}}
\savebox{\rbansbox}{\rbansmacro}
\newcommand{\rbans}{\mathclose{\mspace{1mu}\usebox{\rbansbox}}}

\newsavebox{\llensbox}
\newcommand{\llensmacro}{%
\begin{tikzpicture}[baseline=0.45ex,xscale=0.006em,yscale=0.012ex]%
\draw[solid,join=round] (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9) to[out=-120, in=90] (0.7,3) to[out=-90, in=120] (1.5,0.1) -- cycle;%
\end{tikzpicture}}
\savebox{\llensbox}{\llensmacro}
\newcommand{\llens}{\mathopen{\usebox{\llensbox}\mspace{1mu}}}

\newsavebox{\rlensbox}
\newcommand{\rlensmacro}{%
\begin{tikzpicture}[baseline=0.45ex,xscale=-0.006em,yscale=0.012ex]%
\draw[solid,join=round] (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9) to[out=-120, in=90] (0.7,3) to[out=-90, in=120] (1.5,0.1) -- cycle;%
\end{tikzpicture}}
\savebox{\rlensbox}{\rlensmacro}
\newcommand{\rlens}{\mathclose{\mspace{1mu}\usebox{\rlensbox}}}

### 2023-10-24 paper/paper.tex
Adjoints folds

\documentclass{article}

\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
\usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
\addbibresource{references.bib}
\usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{mathpartir}
\usepackage{bussproofs}
\usepackage{xparse}
\usepackage[usenames, dvipsnames]{xcolor}
\usepackage{lipsum}
\usepackage{xargs}
\usepackage{hyperref}
\usepackage{tikz-cd}
\usepackage{todonotes}
\usepackage{url}
\usepackage{xspace}
\usepackage{rotating}
\usepackage{quiver}
\usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
\addbibresource{references.bib}
\usepackage{array}   % for \newcolumntype macro
\newcolumntype{C}{>{$}c<{$}} % math-mode version of "l" column type
\newcolumntype{L}{>{$}l<{$}} % math-mode version of "l" column type

% \usepackage{draftwatermark}
% \SetWatermarkText{Confidential}
% \SetWatermarkScale{4}
% \SetWatermarkColor[gray]{0.9}

\hypersetup{
  linktocpage,
  colorlinks,
  citecolor=BlueViolet,
  filecolor=red,
  linkcolor=Blue,
  urlcolor=BrickRed
}
\newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
\newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
\newenvironment{proofof}[1] {\begin{proof}[Proof of {#1}]}{\end{proof}}
\newcommand{\eqdef}{\stackrel{\mathrm{\Delta}}{=}}
\newcommand{\bnfeq}{\mathrel{::=}}
\newcommand{\defeq}{\triangleq}
\newcommand{\rul}[3]{\frac{#2}{#3}\;  {\textrulelabel{#1}}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\jud}[3]{#1 \vdash #2 : #3}
\newcommand{\bden}[1]{\llparenthesis#1 \rrparenthesis}
\newcommand{\bigslant}[2]{{\raisebox{.2em}{$#1$}\left/\raisebox{-.2em}{$#2$}\right.}}
\newcommand{\quotient}[2]{\bigslant{#1}{#2}}

\newcommand{\curry}{\Lambda}
\newcommand{\uncurry}{\begin{sideways}\begin{sideways}$\Lambda$\end{sideways}\end{sideways}}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Sets}{\mathbf{Set}}
\newcommand{\blacklater}{\blacktriangleright}
\newcommand{\tot}{\mathcal{S}}
\newcommand{\PSh}{\ensuremath{\textbf{PSh}(\omega)}}

\input{lensbans}

\newcommand{\anamor}[1]{{\llens\, #1\, \rlens}}
\newcommand{\catamor}[1]{\lbans\, #1\, \rbans}
\newcommand{\cata}[1]{\lbans #1 \rbans}
\newcommand{\ana}[1]{\llens #1 \rlens}
\newcommand{\fold}[1]{\catamor{#1}}
\newcommand{\unfold}[1]{\anamor{#1}}

\newcommand{\Alg}{\text{-Alg}}
\newcommand{\Free}{\text{Free\xspace}}

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}

\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}

\title{Hylomorphisms in Coq}

\author{}

\begin{document}

\maketitle

\section{Adjoint Folds}

\subsection{Canonical Adjoint Folds}
\[\begin{tikzcd}
	{\CatC} & \CatD
	\arrow[""{name=0, anchor=center, inner sep=0}, "L"', shift right=2, from=1-2, to=1-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "R"', shift right=2, from=1-1, to=1-2]
	\arrow["\dashv"{anchor=center, rotate=-90}, draw=none, from=0, to=1]
    \arrow["D"', loop, distance=2em, in=35, out=325, from=1-2, to=1-2]
    \arrow["LDR"', loop, distance=2em, in=215, out=145, from=1-1, to=1-1]
\end{tikzcd}
\]

\[\begin{tikzcd}
	{LD\mu D} & LDRB && {D\mu D} & DRB \\
	{L\mu D} & B & {} & {\mu D} & RB
	\arrow["{\cata{alg}}"', from=2-4, to=2-5]
	\arrow[""{name=0, anchor=center, inner sep=0}, "{\iso}"', from=1-4, to=2-4]
	\arrow["{D(\cata{alg})}", from=1-4, to=1-5]
	\arrow["{\floor{alg}}", from=1-5, to=2-5]
	\arrow["{\ceiling{\cata{alg}}}"', from=2-1, to=2-2]
	\arrow["{L(\iso)}"', from=1-1, to=2-1]
	\arrow["{LD(\cata{alg})}"', from=1-1, to=1-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "alg", from=1-2, to=2-2]
	\arrow[shorten <=28pt, shorten >=28pt, Rightarrow, 2tail reversed, from=1, to=0]
\end{tikzcd}\]

\subsection{Example}
\[\begin{tikzcd}
	{\CatC \times \CatC} & \CatC
	\arrow[""{name=0, anchor=center, inner sep=0}, "\Delta"', shift right=2, from=1-2, to=1-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\times"', shift right=2, from=1-1, to=1-2]
	\arrow["\dashv"{anchor=center, rotate=-90}, draw=none, from=0, to=1]
    \arrow["D"', loop, distance=2em, in=35, out=325, from=1-2, to=1-2]
    \arrow["\Delta\circ D\circ \times "', loop, distance=2em, in=215, out=145, from=1-1, to=1-1]
\end{tikzcd}
\]

\[\begin{tikzcd}
	{\Delta\circ D\mu D} & (\Delta \circ D \circ \times)B && {D\mu D} & (D\circ \times) B \\
	{\Delta\mu D} & B & {} & {\mu D} & RB
	\arrow["{\cata{alg}}"', from=2-4, to=2-5]
	\arrow[""{name=0, anchor=center, inner sep=0}, "{\iso}"', from=1-4, to=2-4]
	\arrow["{D(\cata{alg})}", from=1-4, to=1-5]
	\arrow["{\floor{alg}}", from=1-5, to=2-5]
	\arrow["{\ceiling{\cata{alg}}}"', from=2-1, to=2-2]
	\arrow["{D(\iso)}"', from=1-1, to=2-1]
	\arrow["{\Delta D(\cata{alg})}"', from=1-1, to=1-2]
	\arrow[""{name=1, anchor=center, inner sep=0}, "alg", from=1-2, to=2-2]
	\arrow[shorten <=28pt, shorten >=28pt, Rightarrow, 2tail reversed, from=1, to=0]
\end{tikzcd}\]

\end{document}

### 2023-10-25 paper/paper.tex
Adjoint Folds with example.

@@
 
+\newcommand{\Bool}{\mathbb{B}}
 \newcommand{\N}{\mathbb{N}}
+\newcommand{\Nat}{\N}
 \newcommand{\Sets}{\mathbf{Set}}
@@
 
+\newcommand{\operator}[1]{\textsf{#1}}
 \newcommand{\Alg}{\text{-Alg}}
@@
 \newcommand{\floor}[1]{\lfloor #1 \rfloor}
+
+\newcommand{\pair}[2]{\langle #1, #2 \rangle}
+\newcommand{\InIso}{\text{In}}
 
@@
 
-\[\begin{tikzcd}
+The following is a natural isomorphism arising from the adjunction $L \dashv R$.
+\begin{equation}
+  \label{eq:adjoint-iso}
+  \floor{\cdot} : \CatC(LA, B) \iso \CatD (A, RB) : \ceiling{\cdot}
+\end{equation}
+Some properties of adjunctions are worth recalling.  The unit and counit of the
+adjunction can be defined from the above isomorphism as follows:
+\[
+  \eta_{A} = \floor{id_{LA}} \qquad \epsilon_{B}  = \ceiling{id_{RB}}
+\]
+Conversely, the isomorphism can be defined in terms of the two functors $L$ and
+$R$ and the units of the adjunction:
+\begin{align*}
+  & \floor{f : LA \to B} = A \xrightarrow{\eta_{A}} RLA \xrightarrow{R(f)} RB\\
+  & \ceiling{g : A \to RB} = LA \xrightarrow{L(g)} LRB \xrightarrow{\epsilon_{B}} B
+\end{align*}
+These definitions are going to be useful later to understand the definitions of
+the recursion schemes we are going to prove correct.
+
+As a result, the following two diagrams (recursion schemes) are equivalent. The
+on the right hand side is a \emph{catamorphism} taking a $D$ algebra on $RB$ and
+construting an inductive function over the data type $\mu D$. The one on the
+left hand side is the derived recursion scheme where the data type has been
+wrapped up in $L$. The transformation is done by just applying the isomorphism
+(\ref{eq:adjoint-iso}) given by the adjunction.
+
+\begin{equation}
+  \label{eq:canonical-adjoint-fold}
+  \begin{tikzcd}
 	{LD\mu D} & LDRB && {D\mu D} & DRB \\
 	{L\mu D} & B & {} & {\mu D} & RB
-	\arrow["{\cata{alg}}"', from=2-4, to=2-5]
-	\arrow[""{name=0, anchor=center, inner sep=0}, "{\iso}"', from=1-4, to=2-4]
-	\arrow["{D(\cata{alg})}", from=1-4, to=1-5]
-	\arrow["{\floor{alg}}", from=1-5, to=2-5]
-	\arrow["{\ceiling{\cata{alg}}}"', from=2-1, to=2-2]
-	\arrow["{L(\iso)}"', from=1-1, to=2-1]
-	\arrow["{LD(\cata{alg})}"', from=1-1, to=1-2]
-	\arrow[""{name=1, anchor=center, inner sep=0}, "alg", from=1-2, to=2-2]
+	\arrow["{\cata{a}}"', from=2-4, to=2-5]
+	\arrow[""{name=0, anchor=center, inner sep=0}, "{\InIso}"', from=1-4, to=2-4]
+	\arrow["{D(\cata{a})}", from=1-4, to=1-5]
+	\arrow["{\floor{a}}", from=1-5, to=2-5]
+	\arrow["{\ceiling{\cata{a}}}"', from=2-1, to=2-2]
+	\arrow["{L(\InIso)}"', from=1-1, to=2-1]
+	\arrow["{LD(\cata{a})}", from=1-1, to=1-2]
+	\arrow[""{name=1, anchor=center, inner sep=0}, "a", from=1-2, to=2-2]
 	\arrow[shorten <=28pt, shorten >=28pt, Rightarrow, 2tail reversed, from=1, to=0]
-\end{tikzcd}\]
+  \end{tikzcd}
+\end{equation}
+
+In the following section we instantiate the adjoint fold with some instances
+taken from the literature.
 
 \subsection{Example}
+Say that we want to prove that the two functions below are well-defined
+
+\begin{tabular}{LL}
+  \operator{even} : \Nat \to \Bool                  &  \operator{odd} : \Nat \to \Bool\\
+  \operator{even} (0) = \top                        &  \operator{odd}(0) = \top\\
+  \operator{even} (n+1) = \neg \operator{odd}(n)    &  \operator{odd}(n+1) = \neg (\operator{even}(n))
+\end{tabular}
+
+One option is to prove that the following function (given by a catamorphism) is
+equivalent to the above
+\begin{align*}
+  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
+  & \operator{even-odd} (0) = (\top, \bot)\\
+  & \operator{even-odd} (n+1) = \operator{let } (x,y) \leftarrow \operator{even-odd}(n) \operator{ in } (\neg x, \neg y)
+\end{align*}
+This function is clearly a catarmorphism:
+\begin{align*}
+  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
+  & \operator{even-odd} = \cata{a} \operator{ where }\\
+  & \qquad \qquad a : 1 + (\Bool \times \Bool)\\
+  & \qquad \qquad a (\operator{nil}) = (\top, \bot)\\
+  & \qquad \qquad a (\operator{x,y}) = (\neg x, \neg y)
+\end{align*}
+where $\cata{\cdot}$ takes a $(1+-)$ algebra on $\Bool \times \Bool$ and returns
+the associated recursive function.
+
+In order to transform $\operator{even-odd}$ into two functions $\operator{even}$
+and $\operator{odd}$ we are going to use the adjunction $\Delta \vdash \times$
+where $\Delta : \CatC \to \CatC \times \CatC$ is the diagonal functor
+$\Delta X = (X,X)$ and $\times : \CatC \times \CatC$ is the product seen as a
+bifunctor
+
 \[\begin{tikzcd}
@@
 \]
+To get the natural isomorphism from this adjunction we can instantiate
+(\ref{eq:adjoint-iso}) as follows
+\[
+  \floor{\cdot} : \CatC \times \CatC ((A, A) , (B_{1}, B_{2}) \iso \CatC (A , B_{1} \times B_{2}): \ceiling{\cdot}
+\]
+We can instantiate further the iso by setting $A = \Nat$, $B_{1} = \Bool$ and
+$B_{2} = \Bool$.
+\[
+  \floor{\cdot}: \CatC \times \CatC ((\Nat, \Nat) , (\Bool, \Bool) \iso \CatC (\Nat , \Bool \times \Bool) : \ceiling{\cdot}
+\]
 
-\[\begin{tikzcd}
-	{\Delta\circ D\mu D} & (\Delta \circ D \circ \times)B && {D\mu D} & (D\circ \times) B \\
-	{\Delta\mu D} & B & {} & {\mu D} & RB
-	\arrow["{\cata{alg}}"', from=2-4, to=2-5]
-	\arrow[""{name=0, anchor=center, inner sep=0}, "{\iso}"', from=1-4, to=2-4]
-	\arrow["{D(\cata{alg})}", from=1-4, to=1-5]
-	\arrow["{\floor{alg}}", from=1-5, to=2-5]
-	\arrow["{\ceiling{\cata{alg}}}"', from=2-1, to=2-2]
-	\arrow["{D(\iso)}"', from=1-1, to=2-1]
-	\arrow["{\Delta D(\cata{alg})}"', from=1-1, to=1-2]
-	\arrow[""{name=1, anchor=center, inner sep=0}, "alg", from=1-2, to=2-2]
-	\arrow[shorten <=28pt, shorten >=28pt, Rightarrow, 2tail reversed, from=1, to=0]
-\end{tikzcd}\]
+To define this isomorphism we can just define the units of the adjunctions:
+\begin{align*}
+  & \eta_{A} : A \to A \times A\\
+  & \eta_{A}= \pair{id_{A}}{id_{a}}\\
+  & \epsilon_{(X,Y)} : ((X \times Y), (X \times Y) \to (X,Y)\\
+  & \epsilon_{(X,Y)} = (\pi_{1}, \pi_{2})
+\end{align*}
+Notice that, in particular, the counit is given by the projections out of the
+product.  Now, the right transponse $\ceiling{\cdot}$ of the adjunction can be
+defined using the projections provided by the counit.  And in fact, the ceiling
+can be defined as
+\[
+  \ceiling{g} = (\pi_{1},\pi_{2}) \cdot \Delta(g) =  (\pi_{1},\pi_{2}) \cdot (g,g) = (\pi_{1} \cdot g,\pi_{2}\cdot g)
+\]
+and the floor can be defined as
+\[
+  \floor{f}(a) = \pair{f(a)}{f(a)}
+\]
+At this point we are pretty much done as we can instantiate the diagrams
+(\ref{eq:canonical-adjoint-fold}) where $D X = 1 + X$
+\[
+  \begin{tikzcd}
+	{(1 + \Nat, 1 + \Nat)} && (1 + \Bool \times \Bool, 1 + \Bool \times \Bool) & { 1 + \Nat} & 1 + (\Bool \times \Bool) \\
+	{(\Nat, \Nat)} && (\Bool, \Bool) & {\Nat} & \Bool \times \Bool
+	\arrow["{\cata{a}}"', from=2-4, to=2-5]
+	\arrow[""{name=0, anchor=center, inner sep=0}, "{\InIso}"', from=1-4, to=2-4]
+	\arrow["{1 + \cata{a}}", from=1-4, to=1-5]
+	\arrow["a", from=1-5, to=2-5]
+	\arrow["{\ceiling{\cata{a}}}"', from=2-1, to=2-3]
+	\arrow["{(1 +  \InIso,1 +  \InIso)}"', from=1-1, to=2-1]
+	\arrow["{(1 + \cata{a}, 1 + \cata{a})}", from=1-1, to=1-3]
+	\arrow[""{name=1, anchor=center, inner sep=0}, "\ceiling{a}", from=1-3, to=2-3]
+	\arrow[shorten <=35pt, shorten >=35pt, Rightarrow, 2tail reversed, from=1, to=0]
+  \end{tikzcd}
+\]
+Now we define $a$ to be the following $D$ algebra:
+\begin{align*}
+  & a : 1 + \Bool \times \Bool \to \Bool\\
+  & a (\operator{nil}) = (\top, \bot)\\
+  & a (b_{1}, b_{2}) = (\neg b_{1}, \neg b_{2})
+\end{align*}
 
+It is easy to see that the catamorphism over the algebra $a$ equals
+$\operator{even-odd}$ and by the adjoint fold property we know that its right
+transpose satisfies the right hand side diagram. We now have to show that the
+pair of functions $(\operator{even}, \operator{odd})$ equals the right transpose
+of $\operator{even-odd}$, but this is easy to see by definition of the ceiling:
+\[
+  \ceiling{\operator{even-odd}} = (\pi_{1}\cdot \operator{even-odd}, \pi_{2}\cdot \operator{even-odd})
+\]
+and then it is an easy calculation to see that $\pi_{1}\cdot \operator{even-odd}$ equals $\operator{even}$ and
+that $\pi_{2}\cdot \operator{even-odd}$ equals $\operator{odd}$.
 \end{document}

### 2023-10-26 paper/paper.tex
Reshuffling the explanation and expand.

@@
 \maketitle
+\newcommand{\comp}{\cdot}
+\section{Introduction}
+Say that we want to prove that the two functions below are well-defined
 
+\begin{tabular}{LL}
+  \operator{even} : \Nat \to \Bool                  &  \operator{odd} : \Nat \to \Bool\\
+  \operator{even} (0) = \top                        &  \operator{odd}(0) = \bot\\
+  \operator{even} (n+1) = \neg \operator{odd}(n)    &  \operator{odd}(n+1) = \neg (\operator{even}(n))
+\end{tabular}
+
+One option is to prove that the following function (given by a catamorphism) is
+equivalent to the above
+\begin{align*}
+  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
+  & \operator{even-odd} (0) = (\top, \bot)\\
+  & \operator{even-odd} (n+1) = \operator{let } (x,y) \leftarrow \operator{even-odd}(n) \operator{ in } (\neg x, \neg y)
+\end{align*}
+This function is clearly a catarmorphism:
+\begin{align*}
+  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
+  & \operator{even-odd} = \cata{a} \operator{ where }\\
+  & \qquad \qquad a : 1 + (\Bool \times \Bool)\\
+  & \qquad \qquad a (\operator{nil}) = (\top, \bot)\\
+  & \qquad \qquad a (\operator{x,y}) = (\neg x, \neg y)
+\end{align*}
+where $\cata{\cdot}$ takes a $(1+-)$ algebra on $\Bool \times \Bool$ and returns
+the associated recursive function.
+
+Now we recover the original two functions by setting
+$\operator{even}' \eqdef \pi_{1} \comp \operator{even-odd}$
+$\operator{odd}' \eqdef \pi_{2} \comp \operator{even-odd}$. The proof that these two pairs of functions are the same is an easy induction over the natural numbers.
+
+This is an instance of a more general phenonema whose argumet uses category theory.
 \section{Adjoint Folds}
+In order to transform $\operator{even-odd}$ into two functions $\operator{even}$ and
+$\operator{odd}$ we are going to use something called an ``adjunction''.
+In particular, we have two functors. The diagonal functor $\Delta : \CatC \to \CatC \times \CatC$ defined as
+$\Delta X = (X,X)$ and the product functor $\times : \CatC \times \CatC$. These two functors form an adjunction:
+\[\begin{tikzcd}
+	{\CatC \times \CatC} & \CatC
+	\arrow[""{name=0, anchor=center, inner sep=0}, "\Delta"', shift right=2, from=1-2, to=1-1]
+	\arrow[""{name=1, anchor=center, inner sep=0}, "\times"', shift right=2, from=1-1, to=1-2]
+	\arrow["\dashv"{anchor=center, rotate=-90}, draw=none, from=0, to=1]
+    \arrow["D"', loop, distance=2em, in=35, out=325, from=1-2, to=1-2]
+    \arrow["\Delta\circ D\circ \times "', loop, distance=2em, in=215, out=145, from=1-1, to=1-1]
+\end{tikzcd}
+\]
+More formally, this means there is a one-to-one correspondence of maps, i.e.
+every map $(A,A) \to (B_{1}, B_{2})$ corresponds univocally to a map of type $A \to B_{1} \times B_{2}$.
+In the isomorphism below we use $\ceiling{\cdot}$ and $\floor{\cdot}$ to denote the maps witnessing the isomorphism called respectively the right and left transpose of the adjunction:
+\[
+  \floor{\cdot} : \CatC \times \CatC ((A, A) , (B_{1}, B_{2}) \iso \CatC (A , B_{1} \times B_{2}): \ceiling{\cdot}
+\]
+the right transponse $\ceiling{\cdot}$ of the adjunction can be
+defined using the projections:
+\[
+  \ceiling{g} = (\pi_{1} \cdot g,\pi_{2}\cdot g)
+\]
+and the floor can be defined as using the pair into the product:
+\[
+  \floor{f}(a) = \pair{f(a)}{f(a)}
+\]
 
-\subsection{Canonical Adjoint Folds}
+We can instantiate further the iso by setting $A = \Nat$, $B_{1} = \Bool$ and
+$B_{2} = \Bool$.
+\[
+  \floor{\cdot}: \CatC \times \CatC ((\Nat, \Nat) , (\Bool, \Bool) \iso \CatC (\Nat , \Bool \times \Bool) : \ceiling{\cdot}
+\]
+% At this point we are pretty much done as we can instantiate the diagrams
+% (\ref{eq:canonical-adjoint-fold}) where $D X = 1 + X$
+% \[
+%   \begin{tikzcd}
+% 	{(1 + \Nat, 1 + \Nat)} && (1 + \Bool \times \Bool, 1 + \Bool \times \Bool) & { 1 + \Nat} & 1 + (\Bool \times \Bool) \\
+% 	{(\Nat, \Nat)} && (\Bool, \Bool) & {\Nat} & \Bool \times \Bool
+% 	\arrow["{\cata{a}}"', from=2-4, to=2-5]
+% 	\arrow[""{name=0, anchor=center, inner sep=0}, "{\InIso}"', from=1-4, to=2-4]
+% 	\arrow["{1 + \cata{a}}", from=1-4, to=1-5]
+% 	\arrow["a", from=1-5, to=2-5]
+% 	\arrow["{\ceiling{\cata{a}}}"', from=2-1, to=2-3]
+% 	\arrow["{(1 +  \InIso,1 +  \InIso)}"', from=1-1, to=2-1]
+% 	\arrow["{(1 + \cata{a}, 1 + \cata{a})}", from=1-1, to=1-3]
+% 	\arrow[""{name=1, anchor=center, inner sep=0}, "\ceiling{a}", from=1-3, to=2-3]
+% 	\arrow[shorten <=35pt, shorten >=35pt, Rightarrow, 2tail reversed, from=1, to=0]
+%   \end{tikzcd}
+% \]
+% Now we define $a$ to be the following $D$ algebra:
+% \begin{align*}
+%   & a : 1 + \Bool \times \Bool \to \Bool\\
+%   & a (\operator{nil}) = (\top, \bot)\\
+%   & a (b_{1}, b_{2}) = (\neg b_{1}, \neg b_{2})
+% \end{align*}
+
+We already seen that the catamorphism over the algebra $a$ equals
+$\operator{even-odd}$. By using the right transpose we can derive the pair of original functions $(\operator{even}, \operator{odd})$:
+\[
+  \ceiling{\operator{even-odd}} = (\pi_{1}\cdot \operator{even-odd}, \pi_{2}\cdot \operator{even-odd})
+\]
+To show this pair of functions are the same as the original ones we appeal to
+uniqueness of catamorphisms. Recall that catamorphisms for a generic $F$ algebra $a$ are the unique maps such that
+\[
+  \cata{a} = a \comp F(\cata{a})\comp \InIso^{\circ}
+\]
+This is represented on the right hand side of the following diagrams:
+\[
+  \begin{tikzcd}
+	{(1 + \Nat, 1 + \Nat)} && (1 + \Bool^{2}, 1 + \Bool^{2}) & {1 + \Nat} & 1 + \Bool^{2} \\
+	{(\Nat, \Nat)} && (\Bool, \Bool) & {\Nat} & \Bool^{2}
+	\arrow["{\operator{eo}}"', dotted,  from=2-4, to=2-5]
+	\arrow[""{name=0, anchor=center, inner sep=0}, "{\InIso}"', from=1-4, to=2-4]
+	\arrow["{1 + \operator{eo}}", dotted, from=1-4, to=1-5]
+	\arrow["a", from=1-5, to=2-5]
+	\arrow["{\ceiling{\operator{eo}}}"', from=2-1, to=2-3]
+	\arrow["{(1 +  \InIso,1 +  \InIso)}"', from=1-1, to=2-1]
+	\arrow["{(1 + \operator{eo}, 1 + \operator{eo})}", from=1-1, to=1-3]
+	\arrow[""{name=1, anchor=center, inner sep=0}, "\ceiling{a}", from=1-3, to=2-3]
+	\arrow[shorten <=20pt, shorten >=20pt, Rightarrow, 2tail reversed, from=1, to=0]
+  \end{tikzcd}
+\]
+The double implication in the middle of these two diagrams states that the catamorphism on the right is in one-to-one correspondences with the right transpose of $\operator{eo}$ satisfying the right-hand side diagram.
+Notice the right-hand side is exactly the definition of the pair of functions $\operator{even}$ and $\operator{odd}$ and the diagram states these correspond to $\operator{even-odd}$ given by a catamorphism on the left.
+Since this latter is the unique catamorphism the one on the right is also unique such that it makes the right-hand side diagram commute.
+\subsection{The General Case}
+
 \[\begin{tikzcd}
@@
 taken from the literature.
-
-\subsection{Example}
-Say that we want to prove that the two functions below are well-defined
-
-\begin{tabular}{LL}
-  \operator{even} : \Nat \to \Bool                  &  \operator{odd} : \Nat \to \Bool\\
-  \operator{even} (0) = \top                        &  \operator{odd}(0) = \top\\
-  \operator{even} (n+1) = \neg \operator{odd}(n)    &  \operator{odd}(n+1) = \neg (\operator{even}(n))
-\end{tabular}
-
-One option is to prove that the following function (given by a catamorphism) is
-equivalent to the above
-\begin{align*}
-  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
-  & \operator{even-odd} (0) = (\top, \bot)\\
-  & \operator{even-odd} (n+1) = \operator{let } (x,y) \leftarrow \operator{even-odd}(n) \operator{ in } (\neg x, \neg y)
-\end{align*}
-This function is clearly a catarmorphism:
-\begin{align*}
-  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
-  & \operator{even-odd} = \cata{a} \operator{ where }\\
-  & \qquad \qquad a : 1 + (\Bool \times \Bool)\\
-  & \qquad \qquad a (\operator{nil}) = (\top, \bot)\\
-  & \qquad \qquad a (\operator{x,y}) = (\neg x, \neg y)
-\end{align*}
-where $\cata{\cdot}$ takes a $(1+-)$ algebra on $\Bool \times \Bool$ and returns
-the associated recursive function.
-
-In order to transform $\operator{even-odd}$ into two functions $\operator{even}$
-and $\operator{odd}$ we are going to use the adjunction $\Delta \vdash \times$
-where $\Delta : \CatC \to \CatC \times \CatC$ is the diagonal functor
-$\Delta X = (X,X)$ and $\times : \CatC \times \CatC$ is the product seen as a
-bifunctor
-
-\[\begin{tikzcd}
-	{\CatC \times \CatC} & \CatC
-	\arrow[""{name=0, anchor=center, inner sep=0}, "\Delta"', shift right=2, from=1-2, to=1-1]
-	\arrow[""{name=1, anchor=center, inner sep=0}, "\times"', shift right=2, from=1-1, to=1-2]
-	\arrow["\dashv"{anchor=center, rotate=-90}, draw=none, from=0, to=1]
-    \arrow["D"', loop, distance=2em, in=35, out=325, from=1-2, to=1-2]
-    \arrow["\Delta\circ D\circ \times "', loop, distance=2em, in=215, out=145, from=1-1, to=1-1]
-\end{tikzcd}
-\]
-To get the natural isomorphism from this adjunction we can instantiate
-(\ref{eq:adjoint-iso}) as follows
-\[
-  \floor{\cdot} : \CatC \times \CatC ((A, A) , (B_{1}, B_{2}) \iso \CatC (A , B_{1} \times B_{2}): \ceiling{\cdot}
-\]
-We can instantiate further the iso by setting $A = \Nat$, $B_{1} = \Bool$ and
-$B_{2} = \Bool$.
-\[
-  \floor{\cdot}: \CatC \times \CatC ((\Nat, \Nat) , (\Bool, \Bool) \iso \CatC (\Nat , \Bool \times \Bool) : \ceiling{\cdot}
-\]
-
-To define this isomorphism we can just define the units of the adjunctions:
-\begin{align*}
-  & \eta_{A} : A \to A \times A\\
-  & \eta_{A}= \pair{id_{A}}{id_{a}}\\
-  & \epsilon_{(X,Y)} : ((X \times Y), (X \times Y) \to (X,Y)\\
-  & \epsilon_{(X,Y)} = (\pi_{1}, \pi_{2})
-\end{align*}
-Notice that, in particular, the counit is given by the projections out of the
-product.  Now, the right transponse $\ceiling{\cdot}$ of the adjunction can be
-defined using the projections provided by the counit.  And in fact, the ceiling
-can be defined as
-\[
-  \ceiling{g} = (\pi_{1},\pi_{2}) \cdot \Delta(g) =  (\pi_{1},\pi_{2}) \cdot (g,g) = (\pi_{1} \cdot g,\pi_{2}\cdot g)
-\]
-and the floor can be defined as
-\[
-  \floor{f}(a) = \pair{f(a)}{f(a)}
-\]
-At this point we are pretty much done as we can instantiate the diagrams
-(\ref{eq:canonical-adjoint-fold}) where $D X = 1 + X$
-\[
-  \begin{tikzcd}
-	{(1 + \Nat, 1 + \Nat)} && (1 + \Bool \times \Bool, 1 + \Bool \times \Bool) & { 1 + \Nat} & 1 + (\Bool \times \Bool) \\
-	{(\Nat, \Nat)} && (\Bool, \Bool) & {\Nat} & \Bool \times \Bool
-	\arrow["{\cata{a}}"', from=2-4, to=2-5]
-	\arrow[""{name=0, anchor=center, inner sep=0}, "{\InIso}"', from=1-4, to=2-4]
-	\arrow["{1 + \cata{a}}", from=1-4, to=1-5]
-	\arrow["a", from=1-5, to=2-5]
-	\arrow["{\ceiling{\cata{a}}}"', from=2-1, to=2-3]
-	\arrow["{(1 +  \InIso,1 +  \InIso)}"', from=1-1, to=2-1]
-	\arrow["{(1 + \cata{a}, 1 + \cata{a})}", from=1-1, to=1-3]
-	\arrow[""{name=1, anchor=center, inner sep=0}, "\ceiling{a}", from=1-3, to=2-3]
-	\arrow[shorten <=35pt, shorten >=35pt, Rightarrow, 2tail reversed, from=1, to=0]
-  \end{tikzcd}
-\]
-Now we define $a$ to be the following $D$ algebra:
-\begin{align*}
-  & a : 1 + \Bool \times \Bool \to \Bool\\
-  & a (\operator{nil}) = (\top, \bot)\\
-  & a (b_{1}, b_{2}) = (\neg b_{1}, \neg b_{2})
-\end{align*}
-
-It is easy to see that the catamorphism over the algebra $a$ equals
-$\operator{even-odd}$ and by the adjoint fold property we know that its right
-transpose satisfies the right hand side diagram. We now have to show that the
-pair of functions $(\operator{even}, \operator{odd})$ equals the right transpose
-of $\operator{even-odd}$, but this is easy to see by definition of the ceiling:
-\[
-  \ceiling{\operator{even-odd}} = (\pi_{1}\cdot \operator{even-odd}, \pi_{2}\cdot \operator{even-odd})
-\]
-and then it is an easy calculation to see that $\pi_{1}\cdot \operator{even-odd}$ equals $\operator{even}$ and
-that $\pi_{2}\cdot \operator{even-odd}$ equals $\operator{odd}$.
 \end{document}

### 2023-10-27 paper/paper.tex
don't know what I've done
goals and motivations

@@
 
+\newcommand{\comp}{\cdot}
+
 \title{Hylomorphisms in Coq}
@@
 \maketitle
-\newcommand{\comp}{\cdot}
-\section{Introduction}
+\section{Goals and Motivations}
+\begin{itemize}
+  \item We provide a framework for \emph{generic programming} with recursion schemes in Coq
+  \item The hylomorphism and its connected theory with adjunctions provides the most general framework we can work with and provides some built-in optimisations as well
+        \item We formalise algorithms for
+        \begin{itemize}
+          \item  divide and conquer
+          \item dynamic programming
+          \item mutual recursion
+        \end{itemize}
+  \item We show these can be optimised in Coq and extracted to OCaml.
+\end{itemize}
+
+\section{Background on Adjoint Folds}
 Say that we want to prove that the two functions below are well-defined
@@
 $\operator{even}' \eqdef \pi_{1} \comp \operator{even-odd}$
-$\operator{odd}' \eqdef \pi_{2} \comp \operator{even-odd}$. The proof that these two pairs of functions are the same is an easy induction over the natural numbers.
+$\operator{odd}' \eqdef \pi_{2} \comp \operator{even-odd}$. The proof that these
+two pairs of functions are the same is an easy induction over the natural
+numbers.
 
-This is an instance of a more general phenonema whose argumet uses category theory.
+This is an instance of a more general phenonema whose argumet uses category
+theory.
 \section{Adjoint Folds}
-In order to transform $\operator{even-odd}$ into two functions $\operator{even}$ and
-$\operator{odd}$ we are going to use something called an ``adjunction''.
-In particular, we have two functors. The diagonal functor $\Delta : \CatC \to \CatC \times \CatC$ defined as
-$\Delta X = (X,X)$ and the product functor $\times : \CatC \times \CatC$. These two functors form an adjunction:
+In order to transform $\operator{even-odd}$ into two functions $\operator{even}$
+and $\operator{odd}$ we are going to use something called an ``adjunction''.  In
+particular, we have two functors. The diagonal functor
+$\Delta : \CatC \to \CatC \times \CatC$ defined as $\Delta X = (X,X)$ and the
+product functor $\times : \CatC \times \CatC$. These two functors form an
+adjunction:
 \[\begin{tikzcd}
@@
 More formally, this means there is a one-to-one correspondence of maps, i.e.
-every map $(A,A) \to (B_{1}, B_{2})$ corresponds univocally to a map of type $A \to B_{1} \times B_{2}$.
-In the isomorphism below we use $\ceiling{\cdot}$ and $\floor{\cdot}$ to denote the maps witnessing the isomorphism called respectively the right and left transpose of the adjunction:
+every map $(A,A) \to (B_{1}, B_{2})$ corresponds univocally to a map of type
+$A \to B_{1} \times B_{2}$. In the isomorphism below we use $\ceiling{\cdot}$
+and $\floor{\cdot}$ to denote the maps witnessing the isomorphism called
+respectively the right and left transpose of the adjunction:
 \[
@@
 We already seen that the catamorphism over the algebra $a$ equals
-$\operator{even-odd}$. By using the right transpose we can derive the pair of original functions $(\operator{even}, \operator{odd})$:
+$\operator{even-odd}$. By using the right transpose we can derive the pair of
+original functions $(\operator{even}, \operator{odd})$:
 \[
@@
 To show this pair of functions are the same as the original ones we appeal to
-uniqueness of catamorphisms. Recall that catamorphisms for a generic $F$ algebra $a$ are the unique maps such that
+uniqueness of catamorphisms. Recall that catamorphisms for a generic $F$ algebra
+$a$ are the unique maps such that
 \[
@@
 \]
-The double implication in the middle of these two diagrams states that the catamorphism on the right is in one-to-one correspondences with the right transpose of $\operator{eo}$ satisfying the right-hand side diagram.
-Notice the right-hand side is exactly the definition of the pair of functions $\operator{even}$ and $\operator{odd}$ and the diagram states these correspond to $\operator{even-odd}$ given by a catamorphism on the left.
-Since this latter is the unique catamorphism the one on the right is also unique such that it makes the right-hand side diagram commute.
+The double implication in the middle of these two diagrams states that the
+catamorphism on the right is in one-to-one correspondences with the right
+transpose of $\operator{eo}$ satisfying the right-hand side diagram. Notice the
+right-hand side is exactly the definition of the pair of functions
+$\operator{even}$ and $\operator{odd}$ and the diagram states these correspond
+to $\operator{even-odd}$ given by a catamorphism on the left. Since this latter
+is the unique catamorphism the one on the right is also unique such that it
+makes the right-hand side diagram commute.
+
 \subsection{The General Case}

### 2023-12-31 paper/paper.tex
A draft of the motivations and intro

--- 2023-10-26 paper/paper.tex
@@
 \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
+\usepackage{minted}
+\usepackage{newunicodechar}
+
 \addbibresource{references.bib}
@@
 \newcommand{\unfold}[1]{\anamor{#1}}
-
+\newcommand{\comp}{\cdot}
 \newcommand{\operator}[1]{\textsf{#1}}
@@
 
-\title{Hylomorphisms in Coq}
+\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
+
+\title{A Mechanised Library for Generic (Co)Programming}
 
@@
 \maketitle
-\newcommand{\comp}{\cdot}
 \section{Introduction}
+Recursive definitions cannot be proven well-defined automatically. Modern proof
+assistants like Coq or Agda provide a sound, but incomplete algorithm which
+automatically checks for termination or productivity. For recursion, this is
+done by trying to (automatically) infer which argument in the recursive call
+gets smaller with respects to the original input argument. For productivity, the
+algorithm checks that the (co)recursive call appears directly under a
+constructor to make sure that this function always produce at least one element
+after each recursive step.
+
+As mentioned, these algorithms are incomplete meaning that some functions,
+though well-defined, cannot be recognized so by the proof assistant.
+
+One example of this fact is the quicksort algorithm given by the
+\haskell{partition} and \haskell{quicksort} function below (written in Haskell
+code):
+\begin{minted}{haskell}
+partition :: [a] -> ([a], a, [a])
+partition (x:xs) = ((filter (x <=) xs), x, (filter (x >) xs))
+
+quicksort :: [a] -> [a]
+quicksort [] = []
+quicksort ls = let (xs, x, ys) = partition ls in
+                        (quicksort xs) ++ [x] ++ (quicksort ys)
+\end{minted}
+While both \haskell{partition} and the \haskell{quicksort} program using it are
+well-defined functions a proof assistant such as Agda or Coq cannot accept such
+definitions.
+
+The argument can be given on pen-and-paper by showing the \haskell{partition} is
+a function that always returns two sublists which are smaller than the one we
+started with which means the repeated application of \haskell{partition} will
+eventually lead to termination.
+
+The quicksort algorithm is a \emph{divide-and-conquer} algorithm. Such
+algorithms are structured in three parts: first, the input is broken down in
+smaller sub-problems by means of a \emph{well-founded coalgebra}.Secondly, these
+sub-problems are computed recursively and, finally, the results of the recursive
+call are put together by a an algebra.
+
+Practically, every program can be turned into an hylomorphism by means of the
+conjugate rule~\cite{HinzeWG15}. This can be paraphrased by saying
+\begin{center}
+  ``\emph{every recursion scheme can be formalised as an hylomorphism}''
+\end{center}
+
+This has been a useful for structuring functional programs in a way that it
+guarantees termination (or productivity) even in programming languages with
+general recursion~\cite{DanielssonHJG06}.
+
+With this work we intend to begin a study of recursion scheme beyond the world
+of functional programming to the one of dependently typed languages such as a
+the ones employed in proof assistants like Agda or Coq.
+
+One of the challenges here is to be able to keep the theory of recursions
+schemes as general as possible while maintaining its practical usefulness. In
+order to do this, we have to make some design choices which differ a bit from
+the general theory.  For instance, formalising the conjugate rule would require
+us to mechanise a non-trivial fragment of category theory inside the prover's
+logic.  Secondly, the general conjugate rule is a statement of correspondence of
+arrows rather than a proof that these programs actually exists. In functional
+programming languages like for instance Haskell this is not a problem, since
+every functor is strong and has a fixed-point.  On the other hand, in a type
+theory we have to restrict to those so-called polynomial functors. In this work
+we do not make use of these assumptions. By restricting ourselves to polynomial
+functors given in terms of containers~\cite{AbbottAG05} we are able to prove the
+recursion schemes within the proof assistant and at the same time to maintain a
+certain level of data abstraction.
+
+First we prove that the hylo equation has a unique solution
+\[
+  x = a \comp F(x) \comp c
+\]
+for every recursive coalgebra $c$ for a polynomial $F$. Then we prove each
+recursion scheme by reducing it to a hylo. The following recursion schemes can
+all be proven by reducing them to an hylo: folds, mutual recursive functions,
+recursion schemes from comonads and their respective coinductive dual.
+
+The contributions we make in this paper are as follows:
+\begin{itemize}
+  \item We provide a framework for \emph{generic programming} with recursion schemes in Coq
+  \item We formalise algorithms for divide and conquer, dynamic programming and mutual recursion
+  \item We validated optimisations in Coq and extract the resulting code to OCaml.
+\end{itemize}
+
+\section{Background on Adjoint Folds}
 Say that we want to prove that the two functions below are well-defined
@@
 $\operator{even}' \eqdef \pi_{1} \comp \operator{even-odd}$
-$\operator{odd}' \eqdef \pi_{2} \comp \operator{even-odd}$. The proof that these two pairs of functions are the same is an easy induction over the natural numbers.
+$\operator{odd}' \eqdef \pi_{2} \comp \operator{even-odd}$. The proof that these
+two pairs of functions are the same is an easy induction over the natural
+numbers.
 
-This is an instance of a more general phenonema whose argumet uses category theory.
+This is an instance of a more general phenonema whose argumet uses category
+theory.
+
 \section{Adjoint Folds}
-In order to transform $\operator{even-odd}$ into two functions $\operator{even}$ and
-$\operator{odd}$ we are going to use something called an ``adjunction''.
-In particular, we have two functors. The diagonal functor $\Delta : \CatC \to \CatC \times \CatC$ defined as
-$\Delta X = (X,X)$ and the product functor $\times : \CatC \times \CatC$. These two functors form an adjunction:
+In order to transform $\operator{even-odd}$ into two functions $\operator{even}$
+and $\operator{odd}$ we are going to use something called an ``adjunction''.  In
+particular, we have two functors. The diagonal functor
+$\Delta : \CatC \to \CatC \times \CatC$ defined as $\Delta X = (X,X)$ and the
+product functor $\times : \CatC \times \CatC$. These two functors form an
+adjunction:
 \[\begin{tikzcd}
@@
 taken from the literature.
+
+\printbibliography[heading=bibintoc]
+
 \end{document}
+
+%%% Local Variables:
+%%% TeX-command-extra-options: "-shell-escape"
+%%% End:

### 2023-12-31 paper/paper.tex
Draft introduction.

@@
 \end{minted}
-While both \haskell{partition} and the \haskell{quicksort} program using it are
-well-defined functions a proof assistant such as Agda or Coq cannot accept such
-definitions.
+While \haskell{quicksort} is a well-defined mathematical function, a proof
+assistant such as Agda or Coq is not able to accept such definitions. The reason
+is that the function \emph{destructuring} the input, that is the
+\haskell{partition} function is not simply destructuring the list returning the
+head and the tail of it, but it returns two sublists with the head as a pivot.
 
-The argument can be given on pen-and-paper by showing the \haskell{partition} is
-a function that always returns two sublists which are smaller than the one we
-started with which means the repeated application of \haskell{partition} will
-eventually lead to termination.
+While the syntactic checker of a proof assistant cannot understand that these
+two sublists are eventually going to be smaller than the original input we can
+can give a pen-and-paper by showing the \haskell{partition} is a
+\emph{well-founded coalgebra} on the functor $(- \times A \times -)$ for a fixed
+type $A$. Here the term ``well-founded'' refers to the fact that the infinite
+iteration of this coalgebra will produce a \emph{finite} binary tree with nodes
+in $A$.
 
-The quicksort algorithm is a \emph{divide-and-conquer} algorithm. Such
-algorithms are structured in three parts: first, the input is broken down in
-smaller sub-problems by means of a \emph{well-founded coalgebra}.Secondly, these
-sub-problems are computed recursively and, finally, the results of the recursive
-call are put together by a an algebra.
+More generally, this is an instance of a \emph{divide-and-conquer} algorithm.
+Such algorithms are structured in three parts: first, the input is broken down
+in smaller sub-problems by means of a well-founded coalgebra. Secondly,
+these sub-problems are computed recursively and, finally, the results of the
+recursive call are put together by a an algebra. This particular way of
+doing recursion has been named in the functional programming literature as a hylomorphism~\cite{MeijerFP91, HuIT96}.
 
-Practically, every program can be turned into an hylomorphism by means of the
-conjugate rule~\cite{HinzeWG15}. This can be paraphrased by saying
+Many years later Hinze et al.~\cite{HinzeWG15} showed that every recursive
+program can be turned into an hylomorphism by means of the conjugate rule. The
+essence of this result is that every complex recursion scheme can be explained by its associated \emph{adjunction} which establishes a way of reducing the proof obligation
+for the complex recursion scheme into the proof obligation for a basic hylomorphism.
+This can be paraphrased by saying
 \begin{center}
@@
 
-This has been a useful for structuring functional programs in a way that it
-guarantees termination (or productivity) even in programming languages with
-general recursion~\cite{DanielssonHJG06}.
+Since then this result has had huge impact in the functional programming
+community for structuring functional programs in a way that it guarantees
+termination (or productivity) in programming languages with general
+recursion~\cite{DanielssonHJG06}.
+
+There is a tension:
+\begin{itemize}
+\item On one hand we can allow general recursion and the employ recursion schemes to help the programmer structuring programs so that they terminate
+  \item on the other hand we could forbid non-termination, but at this point there is no algorithm to tell termination and non-termination apart.
+\end{itemize}
+
+The first option, helps the programmer but does not guarantee termination, since the programmer can still feed to the recursion scheme a functor that does not have a fixed-point
+or that has an algebra that does not terminate.
+
+The second option forbids non-termination, but does not allow all terminating programs.
 
 With this work we intend to begin a study of recursion scheme beyond the world
-of functional programming to the one of dependently typed languages such as a
-the ones employed in proof assistants like Agda or Coq.
+of functional programming to the one of dependent type theories such as a the
+ones employed in proof assistants like Agda or Coq where non-termination is not
+included in the language. For type theories it has been a long-standing
+challenge to come up with a sound algorithm which rejects all programs that do
+not terminate (soundness) and accepts as many well-defined programs as possible
+(completeness). Our aim is to provide an additional tool to the every-day
+programmer to be able to extend the range of recursive programs we can write.
 

### 2023-12-31 paper/paper.tex
some edits to the intro, related work with more citations.

@@
 \section{Introduction}
-Recursive definitions cannot be proven well-defined automatically. Modern proof
-assistants like Coq or Agda provide a sound, but incomplete algorithm which
-automatically checks for termination or productivity. For recursion, this is
-done by trying to (automatically) infer which argument in the recursive call
-gets smaller with respects to the original input argument. For productivity, the
-algorithm checks that the (co)recursive call appears directly under a
-constructor to make sure that this function always produce at least one element
-after each recursive step.
+Recursive definitions cannot be proven well-defined automatically due to the
+halting problem. Modern proof assistants like Coq or Agda provide a sound, but
+incomplete algorithm which syntactically checks for termination or productivity.
+For recursion, this is done by trying to (automatically) infer which argument in
+the recursive call gets smaller with respects to the original input argument.
+For productivity, the algorithm checks that the (co)recursive call appears
+directly under a constructor to make sure that this function always produce at
+least one element after each recursive step.
 
-As mentioned, these algorithms are incomplete meaning that some functions,
-though well-defined, cannot be recognized so by the proof assistant.
-
-One example of this fact is the quicksort algorithm given by the
-\haskell{partition} and \haskell{quicksort} function below (written in Haskell
-code):
+As mentioned, some functions, though well-defined, cannot be accepted by the
+proof assistant. One example of this fact is the quicksort algorithm given by
+the \haskell{partition} and \haskell{quicksort} function below (written in
+Haskell code):
 \begin{minted}{haskell}
 partition :: [a] -> ([a], a, [a])
-partition (x:xs) = ((filter (x <=) xs), x, (filter (x >) xs))
+partition (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
 
@@
 \end{minted}
-While \haskell{quicksort} is a well-defined mathematical function, a proof
-assistant such as Agda or Coq is not able to accept such definitions. The reason
-is that the function \emph{destructuring} the input, that is the
-\haskell{partition} function is not simply destructuring the list returning the
-head and the tail of it, but it returns two sublists with the head as a pivot.
+While \haskell{quicksort} is a well-defined mathematical function it cannot be
+accepted by a proof assistant.  The reason is that the \haskell{partition}
+function \emph{destructuring} the input ``dives'' deeper in the input returning
+two sublists with the head as a pivot.
 
-While the syntactic checker of a proof assistant cannot understand that these
-two sublists are eventually going to be smaller than the original input we can
-can give a pen-and-paper by showing the \haskell{partition} is a
+A pen-and-paper technique would be to show that the \haskell{partition} is a
 \emph{well-founded coalgebra} on the functor $(- \times A \times -)$ for a fixed
@@
 Such algorithms are structured in three parts: first, the input is broken down
-in smaller sub-problems by means of a well-founded coalgebra. Secondly,
-these sub-problems are computed recursively and, finally, the results of the
-recursive call are put together by a an algebra. This particular way of
-doing recursion has been named in the functional programming literature as a hylomorphism~\cite{MeijerFP91, HuIT96}.
+in smaller sub-problems by means of a well-founded coalgebra. Secondly, these
+sub-problems are computed recursively and, finally, the results of the recursive
+call are put together by a an algebra. This particular way of doing recursion
+has been named in the functional programming literature as a
+hylomorphism~\cite{MeijerFP91, HuIT96}.
 
@@
 program can be turned into an hylomorphism by means of the conjugate rule. The
-essence of this result is that every complex recursion scheme can be explained by its associated \emph{adjunction} which establishes a way of reducing the proof obligation
-for the complex recursion scheme into the proof obligation for a basic hylomorphism.
-This can be paraphrased by saying
+essence of this result is that every complex recursion scheme can be explained
+by its associated \emph{adjunction} which establishes a way of reducing the
+proof obligation for the complex recursion scheme into the proof obligation for
+a basic hylomorphism. This can be paraphrased by saying
 \begin{center}
@@
 \begin{itemize}
-\item On one hand we can allow general recursion and the employ recursion schemes to help the programmer structuring programs so that they terminate
-  \item on the other hand we could forbid non-termination, but at this point there is no algorithm to tell termination and non-termination apart.
+  \item On one hand we can allow general recursion and the employ recursion
+schemes to help the programmer structuring programs so that they terminate
+  \item on the other hand we could forbid non-termination, but at this point
+there is no algorithm to tell termination and non-termination apart.
 \end{itemize}
 
-The first option, helps the programmer but does not guarantee termination, since the programmer can still feed to the recursion scheme a functor that does not have a fixed-point
-or that has an algebra that does not terminate.
+The first option, helps the programmer but does not guarantee termination, since
+the programmer can still feed to the recursion scheme a functor that does not
+have a fixed-point or that has an algebra that does not terminate.
 
-The second option forbids non-termination, but does not allow all terminating programs.
+The second option forbids non-termination, but does not allow all terminating
+programs.
 
-With this work we intend to begin a study of recursion scheme beyond the world
-of functional programming to the one of dependent type theories such as a the
-ones employed in proof assistants like Agda or Coq where non-termination is not
-included in the language. For type theories it has been a long-standing
-challenge to come up with a sound algorithm which rejects all programs that do
-not terminate (soundness) and accepts as many well-defined programs as possible
+With this work we intend to begin a study of the applications of recursion
+schemes that goes beyond functional programming, that is to dependent type
+theories such as a the ones employed in proof assistants like Agda or Coq where
+non-termination is outright rejected.  This has been a long-standing challenge
+to come up with a sound algorithm which rejects all programs that do not
+terminate (soundness) and accepts as many well-defined programs as possible
 (completeness). Our aim is to provide an additional tool to the every-day
@@
 \begin{itemize}
-  \item We provide a framework for \emph{generic programming} with recursion schemes in Coq
-  \item We formalise algorithms for divide and conquer, dynamic programming and mutual recursion
-  \item We validated optimisations in Coq and extract the resulting code to OCaml.
+  \item We provide a framework for \emph{generic programming} with recursion
+schemes in Coq
+  \item We formalise algorithms for divide and conquer, dynamic programming and
+mutual recursion
+  \item We validated optimisations in Coq and extract the resulting code to
+OCaml.
 \end{itemize}
@@
 
+\section{Related Work}
+
+\begin{itemize}
+  \item McBride devised a generalisation of Bove and Capretta's general
+recursive function monad~\cite{McBride15, BoveC01}
+  \item Guarded recursion for productive definitions and general recursion
+~\cite{AtkeyM13, PaviottiMB15}
+  \item PACO library~\cite{HurNDV13}
+  \item Sized Types \cite{AbelPTS13, AbelP16}
+\end{itemize}
+
 \printbibliography[heading=bibintoc]

### 2024-01-18 paper/paper.tex
Notes to introduction + Related work

@@
 \newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
+\newcommand{\dcas}[1]{\textcolor{red}{\textsc{David}: #1}}
 \newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
@@
 \section{Introduction}
+
 Recursive definitions cannot be proven well-defined automatically due to the
@@
 
+\dcas{My take would be different here from now on. See below:}
+
+Recursion schemes enjoy generic algebraic laws for program manipulation. Many
+common optimisations and transformations are instances of such laws. Examples
+of this include shortcut deforestation~\cite{TakanoM95}, and
+parallelisations~\cite{Gibbons96:Third}. \dcas{TODO: add more examples?}
+
+With this work, we study the formalisation of hylomorphism laws in proof
+assistants like Coq to derive efficient and correct recursive algorithms by
+program calculation techniques. Furthermore, our encoding of hylomorphisms
+allows Coq's code extraction mechanism to produce OCaml code that resembles
+what a programmer would have writen manually.
+
+\dcas{END (TODO: merge with Marco's. I would omit (or take parts of the) below)}
+
 There is a tension:
@@
 
+\dcas{Note: the below is useful to keep in the intro}
 One of the challenges here is to be able to keep the theory of recursions
@@
   \item Sized Types \cite{AbelPTS13, AbelP16}
+  \item IMPORTANT!!!!! Coq extraction from recursion
+    schemes~\cite{larchey2022braga} \dcas{Check bibtex}
+  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
+  \item Recursion schemes in Coq (but not hylos)~\cite{MurataE19}
+  \item EVEN MORE IMPORTANT!!! 2023 POPL paper on divide and conquer code in
+    Coq~\cite{AbreuDHJMS23}.  Some differences to point out: they require
+    functional-extensionality axiom. We are axiom-free. Little discussion about
+    extraction. They focus mostly on erasing termination proofs, but they do
+    not discuss "what does their extracted code look like". Their
+    representation of Functors leads me to think that they will produce loads
+    of "Obj.magic", but we'd need to check this. 
 \end{itemize}

### 2024-01-21 paper/paper.tex
small edits

--- 2023-12-31 paper/paper.tex
@@
 As mentioned, some functions, though well-defined, cannot be accepted by the
-proof assistant. One example of this fact is the quicksort algorithm given by
-the \haskell{partition} and \haskell{quicksort} function below (written in
+proof assistant. One example is the quicksort algorithm. This is given by the \haskell{partition} and \haskell{quicksort} function below (written in
 Haskell code):
@@
 
-Since then this result has had huge impact in the functional programming
-community for structuring functional programs in a way that it guarantees
-termination (or productivity) in programming languages with general
-recursion~\cite{DanielssonHJG06}.
+This result has given a much needed unifying view of recursion scheme which
+helped understanding their nature and also in providing a unified way of
+reasoning about recursive programs by means of fusion laws.
 
-There is a tension:
-\begin{itemize}
-  \item On one hand we can allow general recursion and the employ recursion
-schemes to help the programmer structuring programs so that they terminate
-  \item on the other hand we could forbid non-termination, but at this point
-there is no algorithm to tell termination and non-termination apart.
-\end{itemize}
+All in all, the functional programming way of dealing with recursion is to just
+allow all types of recursion and then restricting ourselves to only the
+recursion schemes provided by the given libraries. This gives a way of ``fast
+and loose reasoning''~\cite{DanielssonHJG06} which works provided that the user
+is somewhat aware of the perils of feeding this libraries with potentially
+non-terminating programs.
 
-The first option, helps the programmer but does not guarantee termination, since
-the programmer can still feed to the recursion scheme a functor that does not
-have a fixed-point or that has an algebra that does not terminate.
-
-The second option forbids non-termination, but does not allow all terminating
-programs.
+The type theoretical approach instead is to reject all definitions which cannot
+be proven well-defined. It is safer, but incomplete.
 
@@
 theories such as a the ones employed in proof assistants like Agda or Coq where
-non-termination is outright rejected.  This has been a long-standing challenge
-to come up with a sound algorithm which rejects all programs that do not
-terminate (soundness) and accepts as many well-defined programs as possible
-(completeness). Our aim is to provide an additional tool to the every-day
-programmer to be able to extend the range of recursive programs we can write.
+non-termination is outright rejected.
+
+This has been a long-standing challenge to come up with a sound algorithm which
+rejects all programs that do not terminate (soundness) and accepts as many
+well-defined programs as possible (completeness). Our aim is to provide an
+additional tool to the every-day programmer to be able to extend the range of
+recursive programs we can write.
 

### 2024-01-21 paper/paper.tex
small edits

@@
 As mentioned, some functions, though well-defined, cannot be accepted by the
-proof assistant. One example is the quicksort algorithm. This is given by the \haskell{partition} and \haskell{quicksort} function below (written in
+pThis specific module poses a considerable challenge for the current student cohort.
+However, upon evaluating their performance in assignments, I can affirm that the majority are managing well.
+Recognizing the need for change in the teaching methodology to foster increased engagement, I am considering shifting from traditional slides to utilising my iPad/whiteboard during lectures.
+Additionally, incorporating in-class group exercises aims to slow down the lecture pace and encourage active participation
+
+The classic whiteboard teaching approach is known to enhance student involvement by compelling the lecturer to moderate their speed, allowing students to absorb the material more gradually.
+Supplementary materials will be provided to support their learning outside the classroom.
+I intend to meticulously gauge the impact of these modifications through assignment evaluations and ongoing observation of student participation.
+Any deviations from the expected outcomes, indicated by a lack of in-class responses, will be promptly addressed.
+Encouragingly, I have already shared these proposed changes with my peers in the PGCHE, receiving positive feedback and support for the initiative.
+
+I've already shared some of my plans with the PGCHE instructors, and the response has been quite positive.
+I will continue to stay in touch with them for further feedback on future changes I intend to make.
+roof assistant. One example is the quicksort algorithm. This is given by the \haskell{partition} and \haskell{quicksort} function below (written in
 Haskell code):

### 2024-01-21 paper/paper.tex

--- 2024-01-18 paper/paper.tex
@@
 As mentioned, some functions, though well-defined, cannot be accepted by the
-proof assistant. One example of this fact is the quicksort algorithm given by
-the \haskell{partition} and \haskell{quicksort} function below (written in
+pThis specific module poses a considerable challenge for the current student cohort.
+However, upon evaluating their performance in assignments, I can affirm that the majority are managing well.
+Recognizing the need for change in the teaching methodology to foster increased engagement, I am considering shifting from traditional slides to utilising my iPad/whiteboard during lectures.
+Additionally, incorporating in-class group exercises aims to slow down the lecture pace and encourage active participation
+
+The classic whiteboard teaching approach is known to enhance student involvement by compelling the lecturer to moderate their speed, allowing students to absorb the material more gradually.
+Supplementary materials will be provided to support their learning outside the classroom.
+I intend to meticulously gauge the impact of these modifications through assignment evaluations and ongoing observation of student participation.
+Any deviations from the expected outcomes, indicated by a lack of in-class responses, will be promptly addressed.
+Encouragingly, I have already shared these proposed changes with my peers in the PGCHE, receiving positive feedback and support for the initiative.
+
+I've already shared some of my plans with the PGCHE instructors, and the response has been quite positive.
+I will continue to stay in touch with them for further feedback on future changes I intend to make.
+roof assistant. One example is the quicksort algorithm. This is given by the \haskell{partition} and \haskell{quicksort} function below (written in
 Haskell code):
@@
 
-Since then this result has had huge impact in the functional programming
-community for structuring functional programs in a way that it guarantees
-termination (or productivity) in programming languages with general
-recursion~\cite{DanielssonHJG06}.
+This result has given a much needed unifying view of recursion scheme which
+helped understanding their nature and also in providing a unified way of
+reasoning about recursive programs by means of fusion laws.
 
@@
 
-The first option, helps the programmer but does not guarantee termination, since
-the programmer can still feed to the recursion scheme a functor that does not
-have a fixed-point or that has an algebra that does not terminate.
-
-The second option forbids non-termination, but does not allow all terminating
-programs.
+The type theoretical approach instead is to reject all definitions which cannot
+be proven well-defined. It is safer, but incomplete.
 

### 2024-01-21 paper/paper.tex
minor

@@
 As mentioned, some functions, though well-defined, cannot be accepted by the
-pThis specific module poses a considerable challenge for the current student cohort.
-However, upon evaluating their performance in assignments, I can affirm that the majority are managing well.
-Recognizing the need for change in the teaching methodology to foster increased engagement, I am considering shifting from traditional slides to utilising my iPad/whiteboard during lectures.
-Additionally, incorporating in-class group exercises aims to slow down the lecture pace and encourage active participation
-
-The classic whiteboard teaching approach is known to enhance student involvement by compelling the lecturer to moderate their speed, allowing students to absorb the material more gradually.
-Supplementary materials will be provided to support their learning outside the classroom.
-I intend to meticulously gauge the impact of these modifications through assignment evaluations and ongoing observation of student participation.
-Any deviations from the expected outcomes, indicated by a lack of in-class responses, will be promptly addressed.
-Encouragingly, I have already shared these proposed changes with my peers in the PGCHE, receiving positive feedback and support for the initiative.
-
-I've already shared some of my plans with the PGCHE instructors, and the response has been quite positive.
-I will continue to stay in touch with them for further feedback on future changes I intend to make.
-roof assistant. One example is the quicksort algorithm. This is given by the \haskell{partition} and \haskell{quicksort} function below (written in
+proof assistant. One example is the quicksort algorithm. This is given by the \haskell{partition} and \haskell{quicksort} function below (written in
 Haskell code):

### 2024-01-23 paper/paper.tex
cata, ana, hylo, mutu and accumulators.

@@
 \newcommand{\ana}[1]{\llens #1 \rlens}
+\newcommand{\hylo}[2]{\text{hylo}\;#1\;#2}
 \newcommand{\fold}[1]{\catamor{#1}}
@@
 \newcommand{\Free}{\text{Free\xspace}}
-
+\newcommand{\fmap}[1]{\text{fmap}\;#1}
+\newcommand{\Inop}{\operator{In}^{\circ}}
+\newcommand{\In}{\operator{In}}
 \newcommand{\CatC}{\mathcal{C}}
@@
 \end{itemize}
+
+\section{Recursive Equations}
+\subsection{Catamorphisms and Anamorphism}
+A catamorphism is the unique program defined as
+\[
+  x = a \comp \fmap{x} \comp \Inop
+\]
+Can be implemented:
+\[
+  \cata{a} = \hylo{a}{\Inop}
+\]
+
+An anamorphism is the unique program satisfying the equation
+\[
+  x = \In \comp \fmap{x} \comp{c}
+\]
+and can be implemented by hylomorphisms
+\[
+  \ana{c} = \hylo{\In}{c}
+\]
+
+\subsection{Accumulators}
+For an algebra $a : D(A^{P}) \to A$:
+\begin{align*}
+  & x : (\mu F, A) \to B\\
+  & x (d, p) = a (\fmap{(\curry f)} \comp \Inop(d), p)
+\end{align*}
+
+\subsection{Mutu-Hylo}
+For an algebra $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and an algebra
+$a_{2} : D(A_{1} \times A_{2}) \to A_{2}$:
+\begin{align*}
+  & (f,g) : (\mu F, \mu F) \to (B_{1}, B_{2})\\
+  & (f,g) = (a_{1}, a_{2}) \comp (\fmap{\pair{f}{g}},\fmap{\pair{f}{g}}) \comp (\Inop, \Inop)
+\end{align*}
 

### 2024-01-25 paper/lensbans.tex
Recovering mutu, course-of-value and accu from hylomorphisms.

@@
 \newcommand{\rlens}{\mathclose{\mspace{1mu}\usebox{\rlensbox}}}
+
+\newcommand{\lbanana}{%
+  \mathopen{\tikz[baseline=0.25em,xscale=0.006em,yscale=0.006em]
+  \draw[join=round] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
+              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;\mspace{1mu}}}
+\newcommand{\rbanana}{%
+  \mathclose{\mspace{1mu}\tikz[baseline=0.25em,xscale=-0.006em,yscale=0.006em]
+  \draw[join=round] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
+              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;}}
+% filled versions
+\newcommand{\Lbanana}{%
+  \mathopen{\mspace{1mu}\tikz[baseline=0.25em,xscale=0.006em,yscale=0.006em]
+  \fill (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
+              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;\mspace{1mu}}}
+\newcommand{\Rbanana}{%
+  \mathclose{\mspace{1mu}\tikz[baseline=0.25em,xscale=-0.006em,yscale=0.006em]
+  \fill (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
+              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;}}
+% % lens brackets (using TikZ)
+% \newcommand{\llens}{%
+%   \mathopen{\tikz[baseline=0.25em,xscale=0.006em,yscale=0.006em]
+%   \draw[join=round] (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9)
+%         to[out=-120,in=90] (0.7,3) to[out=-90,in=120] (1.5,0.1) -- cycle;\mspace{1mu}}}
+% \newcommand{\rlens}{%
+%   \mathclose{\mspace{1mu}\tikz[baseline=0.25em,xscale=-0.006em,yscale=0.006em]
+%   \draw[join=round] (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9)
+%         to[out=-120,in=90] (0.7,3) to[out=-90,in=120] (1.5,0.1) -- cycle;}}
+% filled versions
+\newcommand{\Llens}{%
+  \mathopen{\tikz[baseline=0.25em,xscale=0.006em,yscale=0.006em]
+  \fill (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9)
+        to[out=-120,in=90] (0.7,3) to[out=-90,in=120] (1.5,0.1) -- cycle;\mspace{1mu}}}
+\newcommand{\Rlens}{%
+  \mathclose{\mspace{1mu}\tikz[baseline=0.25em,xscale=-0.006em,yscale=0.006em]
+  \fill (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9)
+        to[out=-120,in=90] (0.7,3) to[out=-90,in=120] (1.5,0.1) -- cycle;}}

### 2024-01-25 paper/paper.tex
Recovering mutu, course-of-value and accu from hylomorphisms.

--- 2024-01-21 paper/paper.tex
@@
 \newcommand{\quotient}[2]{\bigslant{#1}{#2}}
-
 \newcommand{\curry}{\Lambda}
 \newcommand{\uncurry}{\begin{sideways}\begin{sideways}$\Lambda$\end{sideways}\end{sideways}}
-
 \newcommand{\Bool}{\mathbb{B}}
@@
 \newcommand{\ana}[1]{\llens #1 \rlens}
+\newcommand{\catafree}[1]{\Lbanana #1 \Rbanana}
+\newcommand{\anacofree}[1]{\Llens #1 \Rlens}
+\newcommand{\hylo}[2]{\cata{#1 \to #2}}
+\newcommand{\cohylo}[2]{\ana{#1 \to #2}}
 \newcommand{\fold}[1]{\catamor{#1}}
@@
 \newcommand{\operator}[1]{\textsf{#1}}
+\newcommand{\head}{\operator{head}}
+\newcommand{\tail}{\operator{tail}}
 \newcommand{\Alg}{\text{-Alg}}
 \newcommand{\Free}{\text{Free\xspace}}
+\newcommand{\fmap}[1]{\text{fmap}\;#1}
+
+\newcommand{\InOp}{\operator{in}^{\circ}}
+\newcommand{\InIso}{\operator{in}}
+\newcommand{\OutOp}{\operator{out}^{\circ}}
+\newcommand{\OutIso}{\operator{out}}
+\newcommand{\call}{\operator{call}}
 
@@
 \newcommand{\CatI}{\mathcal{I}}
-
 \newcommand{\Set}{\mathbf{Set}}
@@
 \newcommand{\floor}[1]{\lfloor #1 \rfloor}
-
 \newcommand{\pair}[2]{\langle #1, #2 \rangle}
-\newcommand{\InIso}{\text{In}}
-
 \newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
@@
 \section{Introduction}
-
 Recursive definitions cannot be proven well-defined automatically due to the
@@
 As mentioned, some functions, though well-defined, cannot be accepted by the
-proof assistant. One example is the quicksort algorithm. This is given by the \haskell{partition} and \haskell{quicksort} function below (written in
-Haskell code):
+proof assistant. One example is the quicksort algorithm. This is given by the
+\haskell{partition} and \haskell{quicksort} function below (written in Haskell
+code):
 \begin{minted}{haskell}
@@
 Recursion schemes enjoy generic algebraic laws for program manipulation. Many
-common optimisations and transformations are instances of such laws. Examples
-of this include shortcut deforestation~\cite{TakanoM95}, and
+common optimisations and transformations are instances of such laws. Examples of
+this include shortcut deforestation~\cite{TakanoM95}, and
 parallelisations~\cite{Gibbons96:Third}. \dcas{TODO: add more examples?}
@@
 program calculation techniques. Furthermore, our encoding of hylomorphisms
-allows Coq's code extraction mechanism to produce OCaml code that resembles
-what a programmer would have writen manually.
+allows Coq's code extraction mechanism to produce OCaml code that resembles what
+a programmer would have writen manually.
 
@@
 
-\section{Background on Adjoint Folds}
+\section{Recursive Equations}
+
+\subsection{Catamorphisms and Anamorphism}
+A catamorphism is the unique program defined as
+\[
+  x = a \comp \fmap{x} \comp \InOp
+\]
+Can be implemented:
+\[
+  \cata{a} = \hylo{\InOp}{a}
+\]
+
+An anamorphism is the unique program satisfying the equation
+\[
+  x = \OutOp \comp \fmap{x} \comp{c}
+\]
+The anamorphism is indicated by $\ana{c}$. Recursion schemes given by a
+corecursive algebra such as $\OutIso$ are also called hylomorphism but they are
+indicated with the symbol for the anamorphism, $\cohylo{c}{\OutOp}$.
+
+\paragraph{Free Monads}
+Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
+solution to the following domain equation
+\[
+  D^{*}A = A + DD^{*}A
+\]
+where $\eta_{A} : A \to D^{*}A$ is the left injection and
+$\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
+$a : DA \to A$ there is a unique solution to the following recursive equation
+\[
+  x \comp \call_{A} = a \comp D(x)
+\]
+that is $\catafree{a} = \cata{[id_{A}; a]}$.
+
+\paragraph{Cofree Comonads}
+Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
+greatest solution to the following domain equation
+\[
+  D_{\infty}A = A + DD_{\infty}A
+\]
+where $\epsilon_{A} : D_{\infty}A \to A$ is the first projection and
+$\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.
+
+Given an algebra $c : A \to DA$ there is a unique solution to the following
+recursive equation
+\[
+  \tail_{A} \comp x = D(x) \comp c
+\]
+that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$.
+
+\subsection{Accumulators}
+For an algebra $a : D(A^{P})\times P \to A$:
+\begin{align*}
+  & g : \mu D \times P \to B\\
+  & g (d, p) = a (\fmap{(\curry g)} \comp \InOp(d), p)
+\end{align*}
+
+\paragraph{Hylo Implementation}
+Given an algebra $a : D(A^{P}) \to A^{P}$ the accumorphism corresponds to the unique
+solution to the following equation
+\begin{align*}
+  & f : \mu D \to A^{P}\\
+  & f = a \comp \fmap (f) \comp \InOp
+\end{align*}
+thus $f = \hylo{\InOp}{a}$.
+
+The complex recursion scheme is then
+$g = \uncurry\hylo{\InOp}{a} : \mu D \times P \to B$.
+
+\subsection{Mutu-Hylo}
+For an algebra $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and an algebra
+$a_{2} : D(A_{1} \times A_{2}) \to A_{2}$:
+\begin{align*}
+  & (g_{1}, g_{2}) : (\mu D, \mu D) \to (A_{1}, A_{2})\\
+  & (g_{1}, g_{2}) = (a_{1}, a_{2}) \comp (\fmap{\pair{g_{1}}{g_{2}}},\fmap{\pair{g_{1}}{g_{2}}}) \comp (\InOp, \InOp)
+\end{align*}
+
+\paragraph{Hylo Implementation}
+Given an algebra $a : D(A_{1} \times A_{2}) \to A_{1} \times A_{2}$
+\begin{align*}
+  & f : \mu D \to A_{1} \times A_{2}\\
+  & f = a \comp \fmap{f} \comp \InOp
+\end{align*}
+for which the solution is an hylo $f = \hylo{\InOp}{a}$.
+
+The mutumorphism can be implemented as
+$(g_{1}, g_{2}) = (\pi_{1} \comp f, \pi_{2} \comp f)$.
+
+\subsubsection{Example}
 Say that we want to prove that the two functions below are well-defined
@@
 theory.
+
+\subsection{Course-of-Value Recursion}
+First we have to assume a distributive law $\lambda : DG \to GD$.
+For an algebra $a : D(D_{\infty} A) \to A$,
+\begin{align*}
+  & g : \mu F \to A\\
+  & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
+\end{align*}
+where $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX  $ is a $D$-algebra on $GX$. The catamorphism of this algebra has the following type
+\[
+  \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
+\]
+which is the type of a $G$-algebra used to initialise the memory table.
+
+\paragraph{Hylo Implementation}
+Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
+what is the unique solution of the following equation
+\begin{align*}
+  & f : \mu F \to G_{\infty} A\\
+  & f = a \comp \fmap{f} \comp \InOp
+\end{align*}
+The solution to this equation can be found again using an hylo
+$f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.
+The course-of-value recursion can be recovered from this by setting
+$g = \head \comp f$
+
+\subsection{Course-of-Value Recursion with Finite Memory Table}
+It can be proven that if $c$ is a recursive coalgebra, then the hylomorphism for
+any algebra $a$ generates a finite output.  In other words, as far as the
+hylomorphism is concerned $G_{\infty}$ can be regarded as its finite version
+\[
+  G_{*} A = \mu X. A \times G X
+\]
+We now rewrite course-of-value recursion using the finite memory table
+For an algebra $a : D(D_{*} A) \to A$,
+\begin{align*}
+  & g : \mu F \to A
+  & g = a \comp D(G_{*}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
+\end{align*}
 

### 2024-01-26 paper/paper.tex
Coq implementation of the Mutu

@@
 \newcommand{\Nat}{\N}
+\newcommand{\R}{\mathbb{R}}
 \newcommand{\Sets}{\mathbf{Set}}
@@
 
+\paragraph{Coq Implementation}
+Given the two algebras  $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and
+$a_{2} : D(A_{1} \times A_{2}) \to A_{2}$.
+We define the two recursive functions  $f_{1}$ and $f_{2}$ as
+
+\[
+  f_{1} = \pi_{1} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}} \qquad f_{2} = \pi_{2} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}}
+\]
+
 \subsubsection{Example}
@@
 \begin{align*}
-  & g : \mu F \to A\\
+  & g : \mu D \to A\\
   & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
@@
 The solution to this equation can be found again using an hylo
-$f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.
-The course-of-value recursion can be recovered from this by setting
-$g = \head \comp f$
+$f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.  The course-of-value recursion
+can be recovered from this by setting $g = \head \comp f$
 
@@
   & g = a \comp D(G_{*}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
+\end{align*}
+
+\subsubsection{Histomorphism}
+The knapsack algorithm
+\begin{align*}
+  & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
+  & \operator{knapsack}\; c = \operator{maximum}_{0} [v + \operator{knapsack}\;wvs\; (c - w) \mid (w,v) \leftarrow wvs, w \le c]
+\end{align*}
+
+Optimisation of the knapsack with Dynamic Programming.
+\begin{align*}
+  & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
+  & \operator{knapsack}\;wvs\;c= table !! c \\
+  & \qquad table  = [ks\;i \mid i \leftarrow [0..c]]\\
+  & \qquad ks\; i = maximum_{0} [v + table!! (i-w) \mid (w,v) \leftarrow wvs, w \le i]
+\end{align*}
+
+\paragraph{Histomorphism}
+
+\begin{align*}
+  & \operator{dyna} : [(\Nat, \R)] \to \Nat \to \R\\
+  & \operator{knapsack}\;wvs\;c= table !! c\\
+  & \qquad table  = [ks\;i \mid i \leftarrow [0..c]]\\
+  & \qquad ks\; i = maximum_{0} [v + table!! (i-w) \mid (w,v) \leftarrow wvs, w \le i]
 \end{align*}

### 2024-01-31 paper/paper.tex
Recursion schemes from comonads

--- 2024-01-25 paper/paper.tex
@@
 \newcommand{\head}{\operator{head}}
+\newcommand{\inl}{\operator{inl}}
+\newcommand{\inr}{\operator{inr}}
 \newcommand{\tail}{\operator{tail}}
@@
 \newcommand{\fmap}[1]{\text{fmap}\;#1}
+
+\newcommand{\zero}{\operator{zero}}
+\newcommand{\Nil}{\operator{nil}}
+\newcommand{\Succ}{\operator{succ}}
 
@@
 \]
-that is $\catafree{a} = \cata{[id_{A}; a]}$.
+that is $\catafree{a} = \cata{[id_{A}; a]}$. With this operation we can extend
+the free algebra $\call$ to the multiplication $\mu : D^{*}D^{*}A \to D^{*}A$
+defined as $\mu = \catafree{\call}$.
 
@@
 \]
-where $\epsilon_{A} : D_{\infty}A \to A$ is the first projection and
+where $\head : D_{\infty}A \to A$ is the first projection and
 $\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.
-
 Given an algebra $c : A \to DA$ there is a unique solution to the following
@@
 \]
-that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$.
+that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$. With the latter operation we
+can extend $\tail$ to the comultiplication for the comonad
+$\delta : D_{\infty}A \to D_{\infty} D_{\infty}A$ defined by
+$\delta = \anacofree{\tail}$.
 
@@
   & \qquad \qquad a : 1 + (\Bool \times \Bool)\\
-  & \qquad \qquad a (\operator{nil}) = (\top, \bot)\\
+  & \qquad \qquad a (\Nil) = (\top, \bot)\\
   & \qquad \qquad a (\operator{x,y}) = (\neg x, \neg y)
@@
 \subsection{Course-of-Value Recursion}
-First we have to assume a distributive law $\lambda : DG \to GD$.
-For an algebra $a : D(D_{\infty} A) \to A$,
+The knapsack algorithm is defined as following in pseudo-Haskell code:
 \begin{align*}
-  & g : \mu F \to A\\
-  & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
+  & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
+  & \operator{knapsack}\;wvs\; c = \operator{maximum}_{0}
+    [v + \operator{knapsack}\;wvs\; (c - w) \mid (w,v) \leftarrow wvs, w \le c]
 \end{align*}
-where $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX  $ is a $D$-algebra on $GX$. The catamorphism of this algebra has the following type
-\[
-  \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
-\]
-which is the type of a $G$-algebra used to initialise the memory table.
+%
+with dynamic programming knapsack can be rephrased like this
+\begin{align*}
+  & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
+  & \operator{knapsack}\;wvs\;c= table !! c \\
+  & \qquad table  = [ks\;i \mid i \leftarrow [0..c]]\\
+  & \qquad ks\; i = maximum_{0} [v + table!! (i - w) \mid (w,v) \leftarrow wvs, w \le i]
+\end{align*}
 
-\paragraph{Hylo Implementation}
-Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
-what is the unique solution of the following equation
+The recursion scheme is justified by solving a recursive equation.  First we
+define an operation to create a memory table representing the list of capacities
+$[0..c]$. We first define a functor $D X = 1 + X$ representing the base functor
+for constructing the naturals $\Nat = \mu D$ the we define the generating
+function by induction on the naturals
 \begin{align*}
-  & f : \mu F \to G_{\infty} A\\
-  & f = a \comp \fmap{f} \comp \InOp
+  & \operator{fan} : \Nat \to D_{\infty}\Nat\\
+  & \operator{fan} (0)    = \pair{0}{\inl}\\
+  & \operator{fan} (n+1)  = \pair{n+1}{\inr (\operator{fan}(n))}
 \end{align*}
-The solution to this equation can be found again using an hylo
-$f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.
-The course-of-value recursion can be recovered from this by setting
-$g = \head \comp f$
+The recursion scheme from a comonad is then defined as follows
+\begin{align*}
+  & \operator{rsfc} : \Nat \to A\\
+  & \operator{rsfc} = alg \comp DD_{\infty}(\operator{rsfc}) \comp D(\operator{fan}) \comp \InOp
+\end{align*}
+At his point we can given the definition of the algebra $alg$ which effectively implements knapsack.
+Firs we define the $\operator{lookup}$ function which returns (if present) the element indexed in the memory table
+\begin{align*}
+  & \operator{lookup} : \Nat \to D_{\infty} \R \to \R_{\bot}\\
+  & \operator{lookup} (0,   \pair{r}{\textunderscore})  = r\\
+  & \operator{lookup} (n+1, \pair{r}{\inl})         = \bot\\
+  & \operator{lookup} (n+1, \pair{r}{\inr(table)})  = \operator{lookup} (n, table)
+\end{align*}
+and now we can define the algebra over the reals
+\begin{align*}
+  & \operator{alg} : DD_{\infty} \R \to \R\\
+  & \operator{alg}\; (\inl)  = 0\\
+  & \operator{alg}\; (\inr(table))  = \operator{maximun}_{0} [ v + \operator{lookup}\; table\; (i - w) \mid (w,v) \leftarrow wvs, w \le i]
+\end{align*}
+
+It remains to show how an $\operator{rsfc}$ is implemented in terms of hylomorhisms.
+
+% A recursion scheme from a comonad $D_{\infty}$ for a data functor $F$
+
+% For example, for the functor $(1 + -)$ an algebra over this functor is a
+% function $1 + X \to X$ is a pair of functions $\Nil : 1 \to X$ and
+% $\Succ : X \to X$. The cofree comonad over $D$ is defined as follows
+% \[
+%   D_{\infty} A \iso A \times (1 + D_{\infty})
+% \]
+% which is the the type of non-empty possibly infinite lists with elements in $A$
+
+% First we have to assume a distributive law
+% $\lambda : DD_{\infty} \to D_{\infty}D$.  For an algebra
+% $a : D(D_{\infty} A) \to A$,
+% \begin{align*}
+%   & g : \mu D \to A\\
+%   & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
+% \end{align*}
+% where
+% $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX$
+% is a $D$-algebra on $GX$. The catamorphism of this algebra has the following
+% type
+% \[
+%   \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
+% \]
+% which is the type of a $G$-algebra used to initialise the memory table.
+
+% Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
+% what is the unique solution of the following equation
+% \begin{align*}
+%   & f : \mu F \to G_{\infty} A\\
+%   & f = a \comp \fmap{f} \comp \InOp
+% \end{align*}
+% The solution to this equation can be found again using an hylo
+% $f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.  The course-of-value recursion
+% can be recovered from this by setting $g = \head \comp f$
+
+% \subsubsection{Histomorphism}
+
+% The definition of $\lambda$ is:
+% \begin{align*}
+%   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
+%   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
+% \end{align*}
+% For example, for $GX = 1 + X$ we have to define
+% \begin{align*}
+%   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
+%   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
+%   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
+% \end{align*}
+
+% % \begin{align*}
+% %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
+% %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
+% %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
+% %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
+% % \end{align*}
+
+% % Defining the algebra
+% % \begin{align*}
+% %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
+% %   & \operator{alg}\; Zero           = 0\\
+% %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
+% % \end{align*}
 

### 2024-01-31 paper/paper.tex
Some structure, plus some Coq code

@@
 
+\section{Mechanising Extractable Containers}
+
+\dcas{Subsections are tentative topics to remember what I need to fill in here,
+but these are not final}
+
+Our goals are: (1) that the mechanisation of hylomorphisms and their laws are
+axiom-free; and (2) Coq hylomorphisms can be extracted to recursive OCaml
+functions that use no instances of \mintinline{OCaml}{Obj.magic}. Note that
+although the use of \mintinline{OCaml}{Obj.magic} is safe in the code extracted
+by Coq, its presence may prevent some optimisations by the OCaml
+compiler\footnote{\dcas{I read this somewhere, search \& add reference/link}}.
+
+\subsection{Functors and Containers}
+
+\dcas{A first approach (problems with extraction, lots of Obj.magic, etc)}
+\begin{minted}{Coq}
+  Record Container (Shape : Set) (Pos : Shape -> Set) (X : Set) :=
+    { shape : Shape; contents : Pos shape -> X }.
+\end{minted}
+
+\subsection{A Type for Morphisms}
+
+\section{Extracting Certified Hylomorphisms}
+
+\dcas{TODO: Hylo mechanisation, Hylo laws}
+
+\subsection{Algebras}
+
+\subsection{(Recursive) Coalgebras}
+
+\section{Examples}
+
+\paragraph{Quicksort \& Mergesort}
+
+\dcas{What else?}
+
 \section{Related Work}

### 2024-02-01 paper/paper.tex
Implementation of rsfc into hylo

@@
 
-It remains to show how an $\operator{rsfc}$ is implemented in terms of hylomorhisms.
+It remains to show how an $\operator{rsfc}$ is implemented in terms of
+hylomorhisms.
+
+\paragraph{Implementation into hylomorphisms}
+In order to implement rsfcs as hylos we need a
+parametric distributive law
+\[
+  \lambda_{X} : GG_{\infty}X \to G_{\infty}GX
+\]
+This can be defined as $\lambda = G_{\infty}G(\head) \comp \anacofree{G(\tail)}$.
+The distributive law is instrumental to be able to lift coalgebras in the
+following sense. Give a $G$-coalgebra on $X$, a lifting $H$ is a functor which
+lifts this coalgebra to a $G$-coalgebra on $HX$.  In particular, the lifting
+$G_{\lambda}$ takes $G_{\infty}$-coalgebras on $X$ and transforms them into
+$G_{\infty}$-coalgebras on $GX$
+\begin{align*}
+  & G_{\lambda}(c) : GA \to G_{\infty} GA\\
+  & G_{\lambda}(c) = \lambda \comp G (c)
+\end{align*}
+where $c : C \to GC$.
+
+Now, given an algebra $a : GG_{\infty} A \to A$ on the carrier $A$ we can
+convert it into an algebra for the carrier $G_{\infty} A$ by
+\begin{align*}
+  & \floor{a} : GG_{\infty} A \to G_{\infty} A\\
+  & \floor{a} = G_{\infty}(a) \comp \anacofree{G_{\lambda}(\delta)}
+\end{align*}
+where $\eta_{A} : A \to G_{\infty} A$ is the unit of the comonad creating an
+infinite constant list.
+
+At this point the hylo is simply defined by
+$\hylo{\InOp}{a} : \mu F \to G_{\infty}A$  and the rsfc is derived as follows
+\begin{align*}
+  & \operator{rsfc} : \mu G \to A \\
+  & \operator{rsfc} = \head \comp \hylo{\InOp}{a}
+\end{align*}
 

### 2024-02-02 paper/paper.tex
minor
Paper

@@
 \paragraph{Implementation into hylomorphisms}
-In order to implement rsfcs as hylos we need a
-parametric distributive law
+In order to implement rsfcs as hylos we need a parametric distributive law
 \[
@@
 \]
-This can be defined as $\lambda = G_{\infty}G(\head) \comp \anacofree{G(\tail)}$.
-The distributive law is instrumental to be able to lift coalgebras in the
-following sense. Give a $G$-coalgebra on $X$, a lifting $H$ is a functor which
-lifts this coalgebra to a $G$-coalgebra on $HX$.  In particular, the lifting
-$G_{\lambda}$ takes $G_{\infty}$-coalgebras on $X$ and transforms them into
-$G_{\infty}$-coalgebras on $GX$
+This can be defined as
+$\lambda = G_{\infty}G(\head) \comp \anacofree{G(\tail)}$.  The distributive law
+is instrumental to be able to lift coalgebras in the following sense. Give a
+$G$-coalgebra on $X$, a lifting $H$ is a functor which lifts this coalgebra to a
+$G$-coalgebra on $HX$.  In particular, the lifting $G_{\lambda}$ takes
+$G_{\infty}$-coalgebras on $X$ and transforms them into $G_{\infty}$-coalgebras
+on $GX$
 \begin{align*}
@@
 
-Our goals are: (1) that the mechanisation of hylomorphisms and their laws are
-axiom-free; and (2) Coq hylomorphisms can be extracted to recursive OCaml
+Our goals are: (1) the mechanisation of hylomorphisms and their laws should be
+axiom-free; and (2) Coq hylomorphisms should be extracted to recursive OCaml
 functions that use no instances of \mintinline{OCaml}{Obj.magic}. Note that
 although the use of \mintinline{OCaml}{Obj.magic} is safe in the code extracted
-by Coq, its presence may prevent some optimisations by the OCaml
-compiler\footnote{\dcas{I read this somewhere, search \& add reference/link}}.
+by Coq, it increases the Trusted Code Base, makes code less legible for
+programmers, and its presence will likely inhibit optimisations by the OCaml
+compiler\footnote{\dcas{I read this somewhere, search \& add reference/link?}}.
 
@@
 
+Mechanising functors in Coq is straightforward. But taking least/greatest
+fixpoints of the said functors may fail, due to not being strictly positive. A
+common approach to construct strictly positive functors in Coq is to use
+\emph{containers}. In essence, a container 
 \dcas{A first approach (problems with extraction, lots of Obj.magic, etc)}
@@
 
-\subsection{A Type for Morphisms}
+\dcas{A better approach for extraction. Unfortunately, to remain axiom-free and
+heterogeneous equality we need to have a decidable ``Valid''}
+\begin{minted}{Coq}
+  Record Container (Shape : Set) (Pos : Set) 
+                   (Valid : Shape -> Pos -> Prop) (X : Set) :=
+    { shape : Shape; contents : {p | Valid shape pos} -> X }.
+\end{minted}
+
+\dcas{Solution to the previous problem}
+\begin{minted}{Coq}
+  Record Container (Shape : Set) (Pos : Set) 
+                   (valid : Shape -> Pos -> bool) (X : Set) :=
+    { shape : Shape; contents : {p | valid shape pos = true} -> X }.
+\end{minted}
+
+\dcas{A shape, a position, and a decidable predicate on the validity of a
+position in a shape define a functor}
+
+\begin{minted}{Coq}
+  Coercion is_true (b : bool) : Prop := b = true.
+  Class functor (Shape Pos : Set) := { valid : Shape -> Pos -> bool }.
+  Record Container `{functor Shape Pos} (X : Set) :=
+    { shape : Shape; contents : {p | valid shape pos} -> X }.
+\end{minted}
+
+\subsection{Types and Morphisms}
+
+Reasoning about the equality of containers requires reasoning about the
+equality of the contained elements. But the contained elements are given by a
+function. This means that container equality needs to be extensional, if we
+want to avoid the functional extensionality axiom. Furthermore, the equality of
+functions also needs to be extensional. We capture this fact by associating
+each type with an equivalence relation:
+
+\begin{minted}{Coq}
+  Class Equiv (T : Type) := 
+  { equiv : T -> T -> Prop;
+    e_refl : forall x, equiv x x;
+    e_sym : forall x y, equiv x y -> equiv y x;
+    e_trans : forall x y z, equiv x y -> equiv y z -> equiv x z;
+  }.
+\end{minted}
+
+\dcas{Morphisms must respect equivalences}:
+
+\begin{minted}{Coq}
+  Record Morph `{Equiv A} `{Equiv B} := 
+  { app : A -> B;
+    app_ext : forall x y, equiv x y -> equiv (app x) (app y);
+  }.
+
+\dcas{We provide common morphisms for polynomials: sums, products, const,
+  exponential}
+
+\end{minted}
 

### 2024-02-05 paper/paper.tex
add canonical reference for sized types
a bit of writing about sized types, and more references

@@
   \item PACO library~\cite{HurNDV13}
-  \item Sized Types \cite{AbelPTS13, AbelP16}
+  \item Sized Types \cite{HughesPS96, abs-1202-3496}
+  \item Copatterns \cite{AbelPTS13, AbelP16}
   \item IMPORTANT!!!!! Coq extraction from recursion
@@
 
+\subsection{Sized Types}
+
+Another approach to certifying termination of nonstructurally recursive
+functions is the use of \emph{sized types}. Sized types were introduced in
+\textcite{HughesPS96} as a way to track/verify various properties of recursive
+programs, including productivity termination. The core idea of sized types is
+that types express bounds on the sizes of recursive data structures. This
+approach has been used to express nonstructurally recursive algorithms in Agda:
+\textcite{CopelloTB14} used sized types in a straightforward formulation of
+mergesort. Such an approach has also been used in MiniAgda~\cite{abs-1202-3496}.
+
 \printbibliography[heading=bibintoc]

### 2024-02-05 paper/paper.tex
missing word
add note about obj.magic

@@
 \newcommand{\dcas}[1]{\textcolor{red}{\textsc{David}: #1}}
+\newcommand{\mvol}[1]{\textcolor{red}{\textsc{Michael}: #1}}
 \newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
@@
     of "Obj.magic", but we'd need to check this.
+\mvol{We have a story for this now: obj.magic harms interaction with other ocaml code, the resulting ocaml code is not natural to read, and in general more complex/strange extracted code is likely to be less efficient and harder to optimise.}
 \end{itemize}
@@
 \textcite{HughesPS96} as a way to track/verify various properties of recursive
-programs, including productivity termination. The core idea of sized types is
-that types express bounds on the sizes of recursive data structures. This
+programs, including productivity and termination. The core idea of sized types
+is that types express bounds on the sizes of recursive data structures. This
 approach has been used to express nonstructurally recursive algorithms in Agda:

### 2024-02-06 paper/paper.tex
Updates on containers

@@
 
-\section{Mechanising Extractable Containers}
+\section{Mechanising Extractable Container Functors}
 
@@
 
-Our goals are: (1) the mechanisation of hylomorphisms and their laws should be
-axiom-free; and (2) Coq hylomorphisms should be extracted to recursive OCaml
-functions that use no instances of \mintinline{OCaml}{Obj.magic}. Note that
-although the use of \mintinline{OCaml}{Obj.magic} is safe in the code extracted
-by Coq, it increases the Trusted Code Base, makes code less legible for
-programmers, and its presence will likely inhibit optimisations by the OCaml
-compiler\footnote{\dcas{I read this somewhere, search \& add reference/link?}}.
+Our mechanisation works in the category of Coq types, and our goals are: (1)
+the mechanisation of hylomorphisms and their laws should be axiom-free; and (2)
+Coq hylomorphisms should be extracted to recursive OCaml functions that use no
+instances of \mintinline{OCaml}{Obj.magic}.  Although the use of
+\mintinline{OCaml}{Obj.magic} is safe in the code extracted by Coq, it has been
+proven that, for higher-order programs, simple interoperations can lead to
+incorrect behaviour or even segfaults~\cite{forster:hal-04329663}.  Therefore,
+the avoidance of axioms and  \mintinline{OCaml}{Obj.magic} means that the
+extracted code will have stronger safety guarantees, and will allow the safe
+interoperation with other OCaml code.
 
@@
 
-Mechanising functors in Coq is straightforward. But taking least/greatest
-fixpoints of the said functors may fail, due to not being strictly positive. A
-common approach to construct strictly positive functors in Coq is to use
-\emph{containers}. In essence, a container 
-\dcas{A first approach (problems with extraction, lots of Obj.magic, etc)}
+In Coq, to construct least/greatest fixpoints of functors, these need to build
+\emph{strictly positive types}. A common approach to construct such strictly
+positive endofunctors is to use \emph{container functors}~\cite{Whatever}.
+However, reasoning about container and container functor equality requires to
+consider: (a) functional extensionality; (b) heterogeneous equality. This is
+problematic, due to our goals to remain axiom-free. Furthermore, standard
+representations of containers in Coq will lead to OCaml code that contains
+\mintinline{OCaml}{Obj.magic}.
+
+A containers is defined by a type of \emph{shapes} (\mintinline{Coq}{Shape }),
+and a \emph{family of position types} indexed by shapes
+(\mintinline{Coq}{Pos}).  An \emph{extension} of this container is a functor,
+whose action on objects is given by the following definition:
 \begin{minted}{Coq}
@@
 \end{minted}
+Note that, in our mechanisation, we call container extensions
+\mintinline{Coq}{Container}. Generally, any functor that is isomorphic to a
+container extension is called a \emph{container functor}. However, in our work
+we will call \emph{container functor} to the functors defined by a specific
+container extension.
+
+The first problem in extracting Coq code from such container extensions is
+the use of families of types. Since OCaml's
 

### 2024-02-06 paper/paper.tex
Containers

@@
 representations of containers in Coq will lead to OCaml code that contains
-\mintinline{OCaml}{Obj.magic}.
+\mintinline{OCaml}{Obj.magic}. We will explain why, and our mechanisation of
+containers in the remainder of this section.
 
-A containers is defined by a type of \emph{shapes} (\mintinline{Coq}{Shape }),
+A container is defined by a type of \emph{shapes} (\mintinline{Coq}{Shape }),
 and a \emph{family of position types} indexed by shapes
@@
 \end{minted}
-Note that, in our mechanisation, we call container extensions
-\mintinline{Coq}{Container}. Generally, any functor that is isomorphic to a
-container extension is called a \emph{container functor}. However, in our work
-we will call \emph{container functor} to the functors defined by a specific
-container extension.
+In our mechanisation, we use \mintinline{Coq}{Container} to represent container
+extensions.  Generally, any functor that is isomorphic to a container extension
+is called a \emph{container functor}. However, in our work we will call
+\emph{container functor} to the functors defined by a specific container
+extension.
 
-The first problem in extracting Coq code from such container extensions is
-the use of families of types. Since OCaml's
-
-\dcas{A better approach for extraction. Unfortunately, to remain axiom-free and
-heterogeneous equality we need to have a decidable ``Valid''}
+The first problem in extracting Coq code from such container extensions is the
+use of families of types. Since OCaml's type system cannot accurately capture
+such families of types, Coq's code extraction mechanism will insert unsafe
+casts. To avoid unsafe casts, we can consider representing the positions of a
+container as a type that captures \emph{all} the potential positions in all
+possible shapes, together with a predicate \mintinline{Coq}{Valid} that
+determines whether a position is valid in a shape. The family of positions is
+then equivalent to the type \mintinline{Coq}{fun shape => {p | Valid shape p}}.
 \begin{minted}{Coq}
@@
                    (Valid : Shape -> Pos -> Prop) (X : Set) :=
-    { shape : Shape; contents : {p | Valid shape pos} -> X }.
+    { shape : Shape; contents : {p | Valid shape p} -> X }.
 \end{minted}
+This approach solves the the problem of the unsafe casts, since the code
+extracted for \mintinine{Coq}{{p | Valid shape pos}} will be an OCaml
+singleton type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.
+
+However, there still remains the issue of reasoning about the equality of
+container extensions. The contents of the container extensions need to be
+accessed via a function \mintinline{Coq}{contents}. The domain of this function
+depends on the shape of the container. This means that, for a given type of shapes
+and positions.,
 

### 2024-02-07 paper/paper.tex
Up to recursive hylos

@@
 \begin{minted}{Coq}
-  Record Container (Shape : Set) (Pos : Shape -> Set) (X : Set) :=
+  Record App (Shape : Set) (Pos : Shape -> Set) (X : Set) :=
     { shape : Shape; contents : Pos shape -> X }.
 \end{minted}
-In our mechanisation, we use \mintinline{Coq}{Container} to represent container
+In our mechanisation, we use \mintinline{Coq}{App} to represent container
 extensions.  Generally, any functor that is isomorphic to a container extension
@@
 \begin{minted}{Coq}
-  Record Container (Shape : Set) (Pos : Set) 
+  Record App (Shape : Set) (Pos : Set) 
                    (Valid : Shape -> Pos -> Prop) (X : Set) :=
@@
 This approach solves the the problem of the unsafe casts, since the code
-extracted for \mintinine{Coq}{{p | Valid shape pos}} will be an OCaml
+extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml
 singleton type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.
@@
 accessed via a function \mintinline{Coq}{contents}. The domain of this function
-depends on the shape of the container. This means that, for a given type of shapes
-and positions.,
-
-\dcas{Solution to the previous problem}
+depends on the shape of the container. This means that, for a given type of
+shapes and positions, the equality of two objects of a container extension will
+need to deal with heterogeneous equality. Since we want to remain axiom-free,
+one way to deal with this is to rely on a \emph{decidable} validity predicate.
+This way, we reach our final definition for container extensions. First,
+containers are defined as pairs of shape and position types that have a
+\emph{decidable} validity predicate. We represent decidable predicates as
+functions that return a boolean, and lift booleans to propositions similarly to
+SSReflect.
 \begin{minted}{Coq}
-  Record Container (Shape : Set) (Pos : Set) 
-                   (valid : Shape -> Pos -> bool) (X : Set) :=
-    { shape : Shape; contents : {p | valid shape pos = true} -> X }.
+Class Container (Shape : Set) (Pos : Set) := {
+   valid : Shape -> Pos -> bool
+}.
+Coercion is_true (b : bool) : Prop := b = true.
+\end{minted}
+Then, we define \emph{container extensions} as
+\begin{minted}{Coq}
+  Record App `{container Shape Pos} (X : Set) :=
+    { shape : Shape; contents : {p | valid shape pos} -> X }.
 \end{minted}
 
-\dcas{A shape, a position, and a decidable predicate on the validity of a
-position in a shape define a functor}
-
-\begin{minted}{Coq}
-  Coercion is_true (b : bool) : Prop := b = true.
-  Class functor (Shape Pos : Set) := { valid : Shape -> Pos -> bool }.
-  Record Container `{functor Shape Pos} (X : Set) :=
-    { shape : Shape; contents : {p | valid shape pos} -> X }.
-\end{minted}
+\dcas{Examples of polynomials as container extensions?}
 
@@
 Reasoning about the equality of containers requires reasoning about the
-equality of the contained elements. But the contained elements are given by a
-function. This means that container equality needs to be extensional, if we
-want to avoid the functional extensionality axiom. Furthermore, the equality of
-functions also needs to be extensional. We capture this fact by associating
-each type with an equivalence relation:
+equality of the contained elements. But the contained elements are given by
+function \mintinline{Coq}{contents}. This means that container equality needs
+to be extensional, if we want to avoid the functional extensionality axiom.
+Furthermore, the equality of functions also needs to be extensional. To deal
+with the fact that different types will use different notions of equality, we
+restrict to types with an associated equivalence relation, and \emph{proper
+morphisms}, i.e. morphisms that respect the necessary equivalence relations.
 
 \begin{minted}{Coq}
-  Class Equiv (T : Type) := 
-  { equiv : T -> T -> Prop;
-    e_refl : forall x, equiv x x;
-    e_sym : forall x y, equiv x y -> equiv y x;
-    e_trans : forall x y z, equiv x y -> equiv y z -> equiv x z;
+Reserved Notation "f =e g" (at level 70, no associativity).
+Class equiv A : Type :=
+  MkEquiv
+    { eqRel : A -> A -> Prop;
+      e_refl : forall x, eqRel x x;
+      e_sym : forall x y, eqRel x y -> eqRel y x;
+      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
     }.
+Notation "f =e g" := (eqRel f g).
+
+Reserved Notation "x ~> y" (at level 95, right associativity, y at level 200).
+Context `{eA : equiv A} `{eB : equiv B}.
+
+ Structure morph :=
+   MkMorph
+     { app :> A -> B;
+       app_eq : forall x y, x =e y -> app x =e app y
+     }.
+Arguments morph A {eA} B {eB}.
+Notation "x ~> y" := (morph x y).
 \end{minted}
+\dcas{Simplify the above}
 
-\dcas{Morphisms must respect equivalences}:
+\dcas{We provide common morphisms for polynomials: sums, products, const,
+exponential (Add examples?)}
+
+By default, we associate every type with Coq's propositional equality, but allow
+the use of other forms of equality. In most cases, we will use either propositional equality, extensional
+equality, or a mixture of both, e.g. for container extensions, we use an arbitrary equality for shapes,
+extensional equality for the contained elements, but Coq's propositional equality for positions:
 
 \begin{minted}{Coq}
-  Record Morph `{Equiv A} `{Equiv B} := 
-  { app : A -> B;
-    app_ext : forall x y, equiv x y -> equiv (app x) (app y);
-  }.
+  Inductive AppR `{F : Container Sh P} (X : Type) {e : equiv X}
+           (x y : App F X) : Prop :=
+  | AppR_ext
+      (Es : shape x =e shape y)
+      (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
+\end{minted}
+The reason for this is to avoid a circularity in the definitions: arbitrary
+equivalences for positions would force us to define the contents as morphisms
+from positions to the contained elements, but that would as well force us to
+redefine container extensions so that they are only applicable to types with
+arbitrary equalities and morphisms.  We will see in the following section why
+this is problematic for defining initial algebras of container extensions.
+\dcas{I am actually not sure of what I wrote above, and it may be something I
+simply overlooked. I will look into the Coq code and try to generalise it if I
+have the time}
 
-\dcas{We provide common morphisms for polynomials: sums, products, const,
-  exponential}
-
+We associate some types with their standard equivalences:
+\begin{minted}{Coq}
+Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
+Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
+Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
+Instance prop_eq : equiv Prop.
+Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
 \end{minted}
@@
 
-\dcas{TODO: Hylo mechanisation, Hylo laws}
-
 \subsection{Algebras}
 
+Algebras on container extensions are defined as expected. The carrier of the
+algebra is \mintinline{Coq}{A}, and the operations of the algebra are
+morphisms of the following type:
+\begin{minted}{Coq}
+Notation Alg F A := (App F A ~> A).
+\end{minted}
+
+There are \emph{two} initial algebras that are, as expected, unique up to
+isomorphism. We show one such initial algebra here, and the second in the
+following subsection, plus a proof that they are isomorphic:
+
+\begin{minted}{Coq}
+Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
+\end{minted}
+This is why we could not define equality of container extensions generically:
+the contained types may not have (yet) an associated equivalence. We could have
+overcame this circularity by defining simultaneously LFix and its associated
+equivalence relation. But that would complicate definitions unnecessarily.
+\dcas{I know ... I need to rewrite/expand on this}
+
+We provide a custom induction schema for initial algebras:
+
+\begin{minted}{Coq}
+Lemma LFix_ind [P : LFix -> Prop]
+  : (forall x : App F LFix,
+        (forall e : Pos (shape x), P (cont x e)) ->
+        P (LFix_in x))
+    -> forall l : LFix, P l.
+\end{minted}
+
+Equality of initial algebras of container extensions needs to be defined
+recursively and extensionally:
+\begin{minted}{Coq}
+Fixpoint LFixR (x y : LFix) : Prop :=
+  let f_x := LFix_out x in
+  let f_y := LFix_out y in
+  shape f_x =e shape f_y /\
+    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
+\end{minted}
+
+The unique homomorphism from an initial algebra to other algebras, i.e. the catamorphism, is defined as a 
+regular recursive function:
+\begin{minted}{Coq}
+Definition cata_f `{eA : equiv A} (g : Alg F A) : LFix -> A
+:= fix f (x : LFix) :=
+  match x with
+  | LFix_in ax =>
+      let (sx, kx) := ax in
+      g (MkCont sx (fun e => f (kx e)))
+  end.
+\end{minted}
+
+It is easy to show that this function is a proper morphism, and that it is
+unique. We can use the fact that catamorphisms are proper morphisms to encode
+them as a higher-order morphism.
+\begin{minted}{Coq}
+Lemma cata_arr1 `{eA : equiv A} (g : Alg F A)
+  : forall x y, x =e y -> cata_f g x =e cata_f g y.
+Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
+
+Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
+    : f =e cata g <-> f =e g \o fmap f \o l_out.
+\end{minted}
+
 \subsection{(Recursive) Coalgebras}
+
+Terminal coalgebras are defined analogously to initial algebras:
+\begin{minted}{Coq}
+CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
+\end{minted}
+
+Their equivalence relation, and anamorphisms, and universal properties are
+easy to encode: 
+\begin{minted}{Coq}
+  Definition ana `{eA : equiv A} : Coalg F A ~> A ~> GFix...
+
+  Lemma ana_univ `{eA : equiv A} (h : Coalg F A) (f : A ~> GFix)
+    : f =e ana h <-> f =e g_in \o fmap f \o h.
+\end{minted}
+
+With terminal coalgebras, we can encode another form of initial algebras. In
+particular, the initial algebras based on terminal coalgebras are the carrier
+types of terminal coalgebras, plus a proof that they are finite:
+
+\begin{minted}{Coq}
+Inductive FinF : GFix F -> Prop :=
+| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
+\end{minted}
+
+Using this predicate, we can define the notion of \emph{recursive coalgebras}
+as those coalgebras that build finite trees. Recursive coalgebras are
+implemented in Coq recursively on the proof that they produce finite trees.
+Such proofs will be erased when extracting code to OCaml. This is standard
+practice when building general recursive functions in Coq\footnote{\dcas{This
+technique had a name in the community, but I cannot remember. Something
+predicates}}.
+\begin{minted}{Coq}
+Notation rana_f__ h :=
+  (fix f x H :=
+    let hx := h x in
+    LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e)))).
+
+Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...
+
+Lemma rana_univ A {eA : equiv A} (h : RCoalg A) (f : A ~> LFix F)
+    : f =e rana h <-> f =e l_in \o fmap f \o h.
+\end{minted}
+
+\subsection{Recursive Hylos}
+
+Recursive hylomorphisms are built from an algebra and a recursive coalgebra:
+\begin{minted}{Coq}
+Definition hylo_def (a : Alg F B) (c : Coalg F A)
+: forall (x : A), RecF c x -> B
+:= fix f x H :=
+  match c x as h0
+        return
+        (forall e : Pos (shape h0), RecF c (cont h0 e)) ->
+        B
+  with
+  | MkCont s_x c_x => fun H => a (MkCont s_x (fun e => f (c_x e) (H e)))
+  end (RecF_inv H).
+\end{minted}
+
+\begin{minted}{Coq}
+Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
+    : f =e hylo g h <-> f =e g \o fmap f \o h.
+\end{minted}
+
+Recursive hylo universal property allows us to prove common optimisations:
+
+\begin{minted}{Coq}
+    Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
+    (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
+    : f2 \o hylo g1 h1 =e hylo g2 h1.
+  Proof.
+    rewrite hylo_univ.
+    rewrite fmap_comp.
+    rewrite compA.
+    rewrite <- E2.
+    rewrite <- compA.
+    rewrite <- compA.
+    rewrite (compA g1).
+    rewrite <- hylo_unr.
+    reflexivity.
+  Qed.
+
+  Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
+    (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
+    : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
+
+  Corollary cata_ana_hylo `(F : Container Sh P) `{equiv A} `{equiv B}
+    (g : Alg F B) (h : RCoalg F A)
+  : cata g \o rana h =e hylo g h.
+\end{minted}
 

### 2024-02-10 paper/paper.tex
LICPICS

@@
-\documentclass{article}
-
+\documentclass[a4paper, UKenglish, cleveref, autoref, thm-restate]{lipics-v2021}
 \usepackage[utf8]{inputenc} % Required for inputting international characters
 \usepackage[T1]{fontenc} % Output font encoding for international characters
-\usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
-\addbibresource{references.bib}
+% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
+% \usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
+% \addbibresource{references.bib}
 \usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
@@
 \usepackage{quiver}
-\usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
 \usepackage{minted}
@@
 
-\addbibresource{references.bib}
+% \addbibresource{references.bib}
 \usepackage{array}   % for \newcolumntype macro
@@
 
-\author{}
+\bibliographystyle{plainurl}% the mandatory bibstyle
+%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line
+
+\author{Jane {Open Access}}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+Please provide at least the name of the affiliation and the country.
+The full address is optional.
+Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.
+
+\author{Joan R.
+Public\footnote{Optional footnote, e.g.
+to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
+
+\authorrunning{J. Open Access and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+
+\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+
+\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+
+\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
+
+\category{} %optional, e.g. invited paper
+
+\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
+%\relatedversiondetails[linktext={opt.
+text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g.
+Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional
+
+%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
+%\supplementdetails[linktext={opt.
+text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
+Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
+
+%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
+Please enter author specific funding statements as fifth argument of the \author macro.
+
+\acknowledgements{I want to thank \dots}%optional
+
+%\nolinenumbers %uncomment to disable line numbering
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\EventEditors{John Q. Open and Joan R. Access}
+\EventNoEds{2}
+\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+\EventShortTitle{CVIT 2016}
+\EventAcronym{CVIT}
+\EventYear{2016}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \maketitle
+
+%TODO mandatory: add short abstract of the document
+\begin{abstract}
+
+\end{abstract}
+
 \section{Introduction}
@@
 By default, we associate every type with Coq's propositional equality, but allow
-the use of other forms of equality. In most cases, we will use either propositional equality, extensional
-equality, or a mixture of both, e.g. for container extensions, we use an arbitrary equality for shapes,
-extensional equality for the contained elements, but Coq's propositional equality for positions:
+the use of other forms of equality. In most cases, we will use either
+propositional equality, extensional equality, or a mixture of both, e.g. for
+container extensions, we use an arbitrary equality for shapes, extensional
+equality for the contained elements, but Coq's propositional equality for
+positions:
 
@@
 functions is the use of \emph{sized types}. Sized types were introduced in
-\textcite{HughesPS96} as a way to track/verify various properties of recursive
+\cite{HughesPS96} as a way to track/verify various properties of recursive
 programs, including productivity and termination. The core idea of sized types
@@
 approach has been used to express nonstructurally recursive algorithms in Agda:
-\textcite{CopelloTB14} used sized types in a straightforward formulation of
+\cite{CopelloTB14} used sized types in a straightforward formulation of
 mergesort. Such an approach has also been used in MiniAgda~\cite{abs-1202-3496}.
 
-\printbibliography[heading=bibintoc]
-
+\bibliography{references.bib}
 \end{document}

### 2024-02-12 paper/paper.tex
Some cleanup about containers and container extensions

@@
 
-A container is defined by a type of \emph{shapes} (\mintinline{Coq}{Shape }),
-and a \emph{family of position types} indexed by shapes
-(\mintinline{Coq}{Pos}).  An \emph{extension} of this container is a functor,
-whose action on objects is given by the following definition:
-\begin{minted}{Coq}
-  Record App (Shape : Set) (Pos : Shape -> Set) (X : Set) :=
-    { shape : Shape; contents : Pos shape -> X }.
+A container is defined by a type of \emph{shapes}, and a \emph{family of
+position types} indexed by shapes. 
+\begin{minted}{Coq}
+Context (Shape : Set) (Pos : Shape -> Set).
+\end{minted}
+An \emph{extension} of this container is a functor,
+whose action on objects is given by:
+\begin{minted}{Coq}
+Record App (X : Set) := { shape : Shape; contents : Pos shape -> X }.
 \end{minted}
@@
 possible shapes, together with a predicate \mintinline{Coq}{Valid} that
-determines whether a position is valid in a shape. The family of positions is
-then equivalent to the type \mintinline{Coq}{fun shape => {p | Valid shape p}}.
-\begin{minted}{Coq}
-  Record App (Shape : Set) (Pos : Set)
-                   (Valid : Shape -> Pos -> Prop) (X : Set) :=
-    { shape : Shape; contents : {p | Valid shape p} -> X }.
+determines whether a position is valid in a shape. 
+\begin{minted}{Coq}
+Context (Shape : Set) (Pos : Set) (Valid : Shape -> Pos -> Prop).
+Record App (X : Set) := { shape    : Shape; 
+                          contents : {p | Valid shape p} -> X }.
 \end{minted}
@@
 \begin{minted}{Coq}
-Class Container (Shape : Set) (Pos : Set) := {
-   valid : Shape -> Pos -> bool
-}.
+Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
 Coercion is_true (b : bool) : Prop := b = true.
@@
 \begin{minted}{Coq}
-  Record App `{container Shape Pos} (X : Set) :=
+Record App `(C : Cont Shape Pos) (X : Set) :=
   { shape : Shape; contents : {p | valid shape pos} -> X }.
 \end{minted}
+
+The functorial action for container extensions on morphisms boils down to
+function composition, which trivially preserves identity and composition.
+\begin{minted}{Coq}
+Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
+  := {| shape := shape x; contents := fun e => f (contents x e) |}
+\end{minted}
+
+\subsubsection{Polynomial Functors as Container Extensions}
+
+\subsubsection{Equality of Container Extensions}
 

### 2024-02-13 paper/paper.tex
Containers and polynomial functors

@@
 \end{minted}
-
 The functorial action for container extensions on morphisms boils down to
@@
 
+Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
+omit here the proofs that these container extensions are isomorphic to
+polynomials, but this proof is straightforward. To illustrate how the identity,
+constant, sum and product container extensions work, consider the
+tree-generating functor (i.e. the functor whose fixpoint is a binary tree):
+$F\; X = \mathbb{N} + X \times X$.
+
+This functor is built as the sum of two functors: $X \mapsto \mathbb{N}$ and $X \mapsto
+X \times X$. Consider first $X \mapsto \mathbb{N}$. This is a constant functor
+that maps $X$ to the type $\mathbb{N}$. More generally, a functor that, given a
+type $A$, maps everything to it, $X \mapsto A$, only needs one shape ($A$, L9
+of Fig~\ref{fig:poly}), and zero positions (In Coq, \mintinline{Coq}{False},
+L10 of Fig~\ref{fig:poly}). The reason why there are zero valid positions in
+a constant container extension is because, given something of type 
+\mintinline{Coq}{App (Const (Ks A) Kp) X}, it should be impossible to get
+something of type \mintinline{Coq}{X}. But from the container shape, we will
+always be able to get the element of type $A$\footnote{%
+  Our decision to use \mintinline{Coq}{False} is arbitrary. In general, we
+  could have chosen any time, with a validity predicate that always returns
+  \mintinline{Coq}{false}. It would make no difference for code extraction.
+}. This functor is, therefore \mintinline{Coq}{Const nat}.
+
+Consider now $X \mapsto X \times X$. This is the product of two
+identity functors $X \mapsto X$. The identity functor only has one shape, and
+we can use the unit type to represent such shapes. From this shape we should be
+able to get an element of type $X$, and we can capture this by using the unit
+type for positions. To build an element of the container extension type, we
+will need to provide a function of type \mintinline{Coq}{unit -> X}, which is
+isomorphic to \mintinline{Coq}{X}.
+
+The shape of the product of two container extensions is the product of their
+shapes. The product of such container extensions should allow us to get either
+the left, or the right element in the product. Therefore, the positions of this
+container extension is the sum of the positions of the two container
+extensions. What is a valid position in a given shape? If the position is
+\mintinline{Coq}{inl p} (or \mintinline{Coq}{inr p}), then it is only valid if
+\mintinline{Coq}{p} is valid in the first (resp. second) component of the
+shape. The product of our two container extensions is
+\mintinline{Coq}{Prod Id Id}.
+
+Finally, the shape the sum of two container extensions is the sum of their
+shapes. The positions are, like for products, the sum of their shapes. But to
+check the validity of shapes in positions, their constructor must match, e.g.
+shape \mintinline{Coq}{inl s} and shape \mintinline{Coq}{inl p}, with
+\mintinline{Coq}{valid s p}.
+
+Our final container extension is the following: 
+\mintinline{Coq}{Sum (Const nat) (Prod Id Id)}.
+The shape of this functor is: 
+\mintinline{Coq}{nat + unit * unit}. The positions of this functor
+are \mintinline[breaklines]{Coq}{False + (unit + unit)}, which is isomorphic
+to \mintinline{Coq}{unit + unit}, which will only be valid in the RHS.
+
+The Coq value \mintinline{Coq}{inl n} of type
+\mintinline{Coq}{nat + bool * bool} would look as follows as a container
+extension 
+\mintinline{Coq}{App (Sum (Const nat) (Prod Id Id)) bool}:
+\begin{minted}{Coq}
+{| shape := inl n; cont := fun e => uninhabited e|}
+\end{minted}
+We omit the definition of \mintinline{Coq}{uninhabited}, which exploits the
+fact that \mintinline{Coq}{e} for \mintinline{Coq}{inl n} implies
+\mintinline{Coq}{False} to build an element of type \mintinline{Coq}{bool}.
+
+Similarly, the Coq value \mintinline{Coq}{inr (true, false)} as a container
+extension is as follows:
+\begin{minted}{Coq}
+  {| shape := inr (tt,tt);
+     cont := fun e : {p | valid (inr (tt, tt)) p} =>
+        match proj1_sig e with
+        | inl f => match f with end
+        | inr (inl tt) => fun _ => true
+        | inr (inr tt) => fun _ => false
+        end
+  |}
+\end{minted}
+Here, the position \mintinline{Coq}{inl f} is impossible, as
+\mintinline{Coq}{f} would need to be a proof of \mintinline{Coq}{False}.
+Of the two remaining positions, the left one will return 
+\mintinline{Coq}{true}, and the right position \mintinline{Coq}{false}.
+\begin{figure}
+\begin{minipage}{.52\textwidth}
+  \begin{minted}[linenos, fontsize=\small]{Coq}
+Context `(C1 : Cont S1 P1) 
+        `(C2 : Cont S2 P2).
+
+Definition Is := unit.
+Definition Ip := unit.
+Instance Id : Cont Is Ip := {
+  valid := fun _ _ => true }.
+
+Definition Ks (A : Set) := A.
+Definition Kp := False.
+Instance Const A : Cont (Ks A) Kp := {
+  valid := fun _ p =>
+    match p with end }.
+
+Definition Ps : Type := S1 * S2.
+Definition Pp : Type := P1 + P2.
+Instance Prod : Cont Ps Pp := {
+  valid := fun s p => 
+    match p with
+    | inl p' => valid (fst s) p'
+    | inr p' => valid (snd s) p'
+    end }.
+\end{minted}
+\end{minipage}
+\begin{minipage}{.47\textwidth}
+  \begin{minted}[linenos, firstnumber=23, fontsize=\small]{Coq}
+Definition Ss : Type := S1 + S2.
+Definition Sp : Type := P1 + P2.
+Instance Sum : Cont Ss Sp := {
+  valid := fun s p => 
+    match s, p with
+    | inl s', inl p' => valid s' p'
+    | inr s', inr p' => valid s' p'
+    | _, _ => false
+    end }.
+
+Definition Cs : Type := App F S2.
+Definition Cp : Type := P1 * P2.
+Instance Comp : Cont Cs Cp := 
+{ valid :=  
+    fun (s : App F S2) (p : P1 * P2) =>
+        match valid (shape s) (fst p) 
+        with
+        | true  => fun H =>
+            valid (cont s {fst p | H}) 
+                  (snd p)
+        | false => fun _ => false
+        end eq_refl }.
+  \end{minted}
+\end{minipage}
+  \caption{Polynomials as Container Extensions. In order: identity (L4-7),
+  constant (L9-13), product (L15-22), sum (L23-31), composition (L33-44).}
+  \label{fig:poly}
+\end{figure}
+
 \subsubsection{Equality of Container Extensions}
-
-\dcas{Examples of polynomials as container extensions?}
 
@@
 \begin{minted}{Coq}
-  Inductive AppR `{F : Container Sh P} (X : Type) {e : equiv X}
+  Inductive AppR `{F : Cont Sh P} (X : Type) {e : equiv X}
            (x y : App F X) : Prop :=
@@
 
-  Corollary cata_ana_hylo `(F : Container Sh P) `{equiv A} `{equiv B}
+  Corollary cata_ana_hylo `(F : Cont Sh P) `{equiv A} `{equiv B}
     (g : Alg F B) (h : RCoalg F A)

### 2024-02-15 paper/paper.tex
Equality of containers

@@
 fact that \mintinline{Coq}{e} for \mintinline{Coq}{inl n} implies
-\mintinline{Coq}{False} to build an element of type \mintinline{Coq}{bool}.
+\mintinline{Coq}{False} to build an element of type \mintinline{Coq}{bool}%
+\footnote{Note that, since \mintinline{Coq}{e} is uninhabited we could return
+any arbitrary value instead of \mintinline{Coq}{uninhabited e}.}.
 
@@
 \subsubsection{Equality of Container Extensions}
+
+The equality of container extensions needs to be extensional, since the
+contents are obtained by applying a function. We want to equate container
+extensions that have the same shape, and that, for equal shape and position,
+they return the same element. 
+\begin{minted}{Coq}
+  { shape : Shape; contents : {p | valid shape pos} -> X }.
+\end{minted}
 

### 2024-02-16 paper/paper.tex
Paper

@@
 functions that return a boolean, and lift booleans to propositions similarly to
-SSReflect.
+SSReflect. For simplicity, we define the dependent pair of valid positions as a
+record, which will be extracted to a singleton type with the validity proof
+erased by Coq's code extraction mechanism.
 \begin{minted}{Coq}
@@
 Coercion is_true (b : bool) : Prop := b = true.
+Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
+  ValidPos { val : Pos; IsValid : valid shape pos }.
 \end{minted}
@@
 Record App `(C : Cont Shape Pos) (X : Set) :=
-  { shape : Shape; contents : {p | valid shape pos} -> X }.
+  { shape : Shape; contents : PosOf C shape -> X }.
 \end{minted}
@@
 constant, sum and product container extensions work, consider the
-tree-generating functor (i.e. the functor whose fixpoint is a binary tree):
+tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
 $F\; X = \mathbb{N} + X \times X$.
@@
 \end{minipage}
-  \caption{Polynomials as Container Extensions. In order: identity (L4-7),
-  constant (L9-13), product (L15-22), sum (L23-31), composition (L33-44).}
+  \caption{Polynomials as Container Extensions (simplified). In order: identity
+  (L4-7), constant (L9-13), product (L15-22), sum (L23-31), composition
+  (L33-44).}
   \label{fig:poly}
@@
 extensions that have the same shape, and that, for equal shape and position,
-they return the same element. 
-\begin{minted}{Coq}
-  { shape : Shape; contents : {p | valid shape pos} -> X }.
-\end{minted}
+they return the same element. To avoid the use of functional extensionality
+axioms, we capture this relation with the following inductive proposition in
+Coq:
+\begin{minted}{Coq}
+Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
+| AppR_ext (Es : shape x = shape y)
+           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
+\end{minted}
+Note that  we do not care about the validity proof of the positions, only their
+value. This is to simplify (slightly) our proofs. This relation is
+trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.
+
+However, the use of a different equality for container extensions now forces us
+to deal with the fact that some types have different definitions of equality.
+In particular, we want to reason about the equality of functions of types such
+as \mintinline{Coq}{App F A -> B} (or \mintinline{Coq}{B -> App F A}). 
+We want the following kind of properties to hold: 
+\begin{minted}{Coq}
+forall (x y : App C A) (f : App C A -> B), AppR x y -> f x = f y
+\end{minted}
+Moreover, we want this to be true for any \mintinline{Coq}{B}, and equality.
+We address this in the following section.
 
@@
 
-Reasoning about the equality of containers requires reasoning about the
-equality of the contained elements. But the contained elements are given by
-function \mintinline{Coq}{contents}. This means that container equality needs
-to be extensional, if we want to avoid the functional extensionality axiom.
-Furthermore, the equality of functions also needs to be extensional. To deal
-with the fact that different types will use different notions of equality, we
-restrict to types with an associated equivalence relation, and \emph{proper
-morphisms}, i.e. morphisms that respect the necessary equivalence relations.
+The need to reason about different forms of equality forces us to couple every
+type with an associated equivalence relation. Furthermore, we will only allow
+to reason about the equality of \emph{proper morphisms}, i.e. morphisms that
+respect the necessary equivalence relations.
 
@@
 \end{minted}
-\dcas{Simplify the above}
 
@@
 positions:
-
-\begin{minted}{Coq}
-  Inductive AppR `{F : Cont Sh P} (X : Type) {e : equiv X}
-           (x y : App F X) : Prop :=
-  | AppR_ext
-      (Es : shape x =e shape y)
+\begin{minted}{Coq}
+  Inductive AppR `{Cont Sh P} `{equiv X} (x y : App F X) : Prop :=
+  | AppR_ext (Es : shape x =e shape y)
              (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
 \end{minted}
-The reason for this is to avoid a circularity in the definitions: arbitrary
-equivalences for positions would force us to define the contents as morphisms
-from positions to the contained elements, but that would as well force us to
-redefine container extensions so that they are only applicable to types with
-arbitrary equalities and morphisms.  We will see in the following section why
-this is problematic for defining initial algebras of container extensions.
+Notice the use of \mintinline{Coq}{=e} for both shapes, and the continuations
+of container extensions. 
+The reason to use Coq's propositional equality for positions is to avoid a
+circularity in the definitions: arbitrary equivalences for positions would
+force us to define the contents as morphisms from positions to the contained
+elements, but that would as well force us to redefine container extensions so
+that they are only applicable to types with arbitrary equalities and morphisms.
+We will see in the following section why this is problematic for defining
+initial algebras of container extensions.
 \dcas{I am actually not sure of what I wrote above, and it may be something I
@@
 
+\subsubsection{Polynomial Types, Equivalences, and Morphisms}
 We associate some types with their standard equivalences:
@@
 \end{minted}
+
+\subsubsection{Isomorphisms with Container Extensions}
 

### 2024-02-22 paper/paper.tex
Moved adjoint folds to the apendix to save space
Types and morphisms

@@
 
+\section{Mechanising Extractable Container Functors}
+
+\dcas{Subsections are tentative topics to remember what I need to fill in here,
+but these are not final}
+
+Our mechanisation works in the category of Coq types, and our goals are: (1)
+the mechanisation of hylomorphisms and their laws should be axiom-free; and (2)
+Coq hylomorphisms should be extracted to recursive OCaml functions that use no
+instances of \mintinline{OCaml}{Obj.magic}.  Although the use of
+\mintinline{OCaml}{Obj.magic} is safe in the code extracted by Coq, it has been
+proven that, for higher-order programs, simple interoperations can lead to
+incorrect behaviour or even segfaults~\cite{forster:hal-04329663}.  Therefore,
+the avoidance of axioms and  \mintinline{OCaml}{Obj.magic} means that the
+extracted code will have stronger safety guarantees, and will allow the safe
+interoperation with other OCaml code.
+
+\subsection{Functors and Containers}
+
+In Coq, to construct least/greatest fixpoints of functors, these need to build
+\emph{strictly positive types}. A common approach to construct such strictly
+positive endofunctors is to use \emph{container functors}~\cite{Whatever}.
+However, reasoning about container and container functor equality requires to
+consider: (a) functional extensionality; (b) heterogeneous equality. This is
+problematic, due to our goals to remain axiom-free. Furthermore, standard
+representations of containers in Coq will lead to OCaml code that contains
+\mintinline{OCaml}{Obj.magic}. We will explain why, and our mechanisation of
+containers in the remainder of this section.
+
+A container is defined by a type of \emph{shapes}, and a \emph{family of
+position types} indexed by shapes. 
+\begin{minted}{Coq}
+Context (Shape : Set) (Pos : Shape -> Set).
+\end{minted}
+An \emph{extension} of this container is a functor,
+whose action on objects is given by:
+\begin{minted}{Coq}
+Record App (X : Set) := { shape : Shape; contents : Pos shape -> X }.
+\end{minted}
+In our mechanisation, we use \mintinline{Coq}{App} to represent container
+extensions.  Generally, any functor that is isomorphic to a container extension
+is called a \emph{container functor}. However, in our work we will call
+\emph{container functor} to the functors defined by a specific container
+extension.
+
+The first problem in extracting Coq code from such container extensions is the
+use of families of types. Since OCaml's type system cannot accurately capture
+such families of types, Coq's code extraction mechanism will insert unsafe
+casts. To avoid unsafe casts, we can consider representing the positions of a
+container as a type that captures \emph{all} the potential positions in all
+possible shapes, together with a predicate \mintinline{Coq}{Valid} that
+determines whether a position is valid in a shape. 
+\begin{minted}{Coq}
+Context (Shape : Set) (Pos : Set) (Valid : Shape -> Pos -> Prop).
+Record App (X : Set) := { shape    : Shape; 
+                          contents : {p | Valid shape p} -> X }.
+\end{minted}
+This approach solves the the problem of the unsafe casts, since the code
+extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml
+singleton type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.
+
+However, there still remains the issue of reasoning about the equality of
+container extensions. The contents of the container extensions need to be
+accessed via a function \mintinline{Coq}{contents}. The domain of this function
+depends on the shape of the container. This means that, for a given type of
+shapes and positions, the equality of two objects of a container extension will
+need to deal with heterogeneous equality. Since we want to remain axiom-free,
+one way to deal with this is to rely on a \emph{decidable} validity predicate.
+This way, we reach our final definition for container extensions. First,
+containers are defined as pairs of shape and position types that have a
+\emph{decidable} validity predicate. We represent decidable predicates as
+functions that return a boolean, and lift booleans to propositions similarly to
+SSReflect. For simplicity, we define the dependent pair of valid positions as a
+record, which will be extracted to a singleton type with the validity proof
+erased by Coq's code extraction mechanism.
+\begin{minted}{Coq}
+Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
+Coercion is_true (b : bool) : Prop := b = true.
+Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
+  ValidPos { val : Pos; IsValid : valid shape pos }.
+\end{minted}
+Then, we define \emph{container extensions} as
+\begin{minted}{Coq}
+Record App `(C : Cont Shape Pos) (X : Set) :=
+  { shape : Shape; contents : PosOf C shape -> X }.
+\end{minted}
+The functorial action for container extensions on morphisms boils down to
+function composition, which trivially preserves identity and composition.
+\begin{minted}{Coq}
+Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
+  := {| shape := shape x; contents := fun e => f (contents x e) |}
+\end{minted}
+
+\subsubsection{Polynomial Functors as Container Extensions}
+
+Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
+omit here the proofs that these container extensions are isomorphic to
+polynomials, but this proof is straightforward. To illustrate how the identity,
+constant, sum and product container extensions work, consider the
+tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
+$F\; X = \mathbb{N} + X \times X$.
+
+This functor is built as the sum of two functors: $X \mapsto \mathbb{N}$ and $X \mapsto
+X \times X$. Consider first $X \mapsto \mathbb{N}$. This is a constant functor
+that maps $X$ to the type $\mathbb{N}$. More generally, a functor that, given a
+type $A$, maps everything to it, $X \mapsto A$, only needs one shape ($A$, L9
+of Fig~\ref{fig:poly}), and zero positions (In Coq, \mintinline{Coq}{False},
+L10 of Fig~\ref{fig:poly}). The reason why there are zero valid positions in
+a constant container extension is because, given something of type 
+\mintinline{Coq}{App (Const (Ks A) Kp) X}, it should be impossible to get
+something of type \mintinline{Coq}{X}. But from the container shape, we will
+always be able to get the element of type $A$\footnote{%
+  Our decision to use \mintinline{Coq}{False} is arbitrary. In general, we
+  could have chosen any time, with a validity predicate that always returns
+  \mintinline{Coq}{false}. It would make no difference for code extraction.
+}. This functor is, therefore \mintinline{Coq}{Const nat}.
+
+Consider now $X \mapsto X \times X$. This is the product of two
+identity functors $X \mapsto X$. The identity functor only has one shape, and
+we can use the unit type to represent such shapes. From this shape we should be
+able to get an element of type $X$, and we can capture this by using the unit
+type for positions. To build an element of the container extension type, we
+will need to provide a function of type \mintinline{Coq}{unit -> X}, which is
+isomorphic to \mintinline{Coq}{X}.
+
+The shape of the product of two container extensions is the product of their
+shapes. The product of such container extensions should allow us to get either
+the left, or the right element in the product. Therefore, the positions of this
+container extension is the sum of the positions of the two container
+extensions. What is a valid position in a given shape? If the position is
+\mintinline{Coq}{inl p} (or \mintinline{Coq}{inr p}), then it is only valid if
+\mintinline{Coq}{p} is valid in the first (resp. second) component of the
+shape. The product of our two container extensions is
+\mintinline{Coq}{Prod Id Id}.
+
+Finally, the shape the sum of two container extensions is the sum of their
+shapes. The positions are, like for products, the sum of their shapes. But to
+check the validity of shapes in positions, their constructor must match, e.g.
+shape \mintinline{Coq}{inl s} and shape \mintinline{Coq}{inl p}, with
+\mintinline{Coq}{valid s p}.
+
+Our final container extension is the following: 
+\mintinline{Coq}{Sum (Const nat) (Prod Id Id)}.
+The shape of this functor is: 
+\mintinline{Coq}{nat + unit * unit}. The positions of this functor
+are \mintinline[breaklines]{Coq}{False + (unit + unit)}, which is isomorphic
+to \mintinline{Coq}{unit + unit}, which will only be valid in the RHS.
+
+The Coq value \mintinline{Coq}{inl n} of type
+\mintinline{Coq}{nat + bool * bool} would look as follows as a container
+extension 
+\mintinline{Coq}{App (Sum (Const nat) (Prod Id Id)) bool}:
+\begin{minted}{Coq}
+{| shape := inl n; cont := fun e => uninhabited e|}
+\end{minted}
+We omit the definition of \mintinline{Coq}{uninhabited}, which exploits the
+fact that \mintinline{Coq}{e} for \mintinline{Coq}{inl n} implies
+\mintinline{Coq}{False} to build an element of type \mintinline{Coq}{bool}%
+\footnote{Note that, since \mintinline{Coq}{e} is uninhabited we could return
+any arbitrary value instead of \mintinline{Coq}{uninhabited e}.}.
+
+Similarly, the Coq value \mintinline{Coq}{inr (true, false)} as a container
+extension is as follows:
+\begin{minted}{Coq}
+  {| shape := inr (tt,tt);
+     cont := fun e : {p | valid (inr (tt, tt)) p} =>
+        match proj1_sig e with
+        | inl f => match f with end
+        | inr (inl tt) => fun _ => true
+        | inr (inr tt) => fun _ => false
+        end
+  |}
+\end{minted}
+Here, the position \mintinline{Coq}{inl f} is impossible, as
+\mintinline{Coq}{f} would need to be a proof of \mintinline{Coq}{False}.
+Of the two remaining positions, the left one will return 
+\mintinline{Coq}{true}, and the right position \mintinline{Coq}{false}.
+\begin{figure}
+\begin{minipage}{.52\textwidth}
+  \begin{minted}[linenos, fontsize=\small]{Coq}
+Context `(C1 : Cont S1 P1) 
+        `(C2 : Cont S2 P2).
+
+Definition Is := unit.
+Definition Ip := unit.
+Instance Id : Cont Is Ip := {
+  valid := fun _ _ => true }.
+
+Definition Ks (A : Set) := A.
+Definition Kp := False.
+Instance Const A : Cont (Ks A) Kp := {
+  valid := fun _ p =>
+    match p with end }.
+
+Definition Ps : Type := S1 * S2.
+Definition Pp : Type := P1 + P2.
+Instance Prod : Cont Ps Pp := {
+  valid := fun s p => 
+    match p with
+    | inl p' => valid (fst s) p'
+    | inr p' => valid (snd s) p'
+    end }.
+\end{minted}
+\end{minipage}
+\begin{minipage}{.47\textwidth}
+  \begin{minted}[linenos, firstnumber=23, fontsize=\small]{Coq}
+Definition Ss : Type := S1 + S2.
+Definition Sp : Type := P1 + P2.
+Instance Sum : Cont Ss Sp := {
+  valid := fun s p => 
+    match s, p with
+    | inl s', inl p' => valid s' p'
+    | inr s', inr p' => valid s' p'
+    | _, _ => false
+    end }.
+
+Definition Cs : Type := App F S2.
+Definition Cp : Type := P1 * P2.
+Instance Comp : Cont Cs Cp := 
+{ valid :=  
+    fun (s : App F S2) (p : P1 * P2) =>
+        match valid (shape s) (fst p) 
+        with
+        | true  => fun H =>
+            valid (cont s {fst p | H}) 
+                  (snd p)
+        | false => fun _ => false
+        end eq_refl }.
+  \end{minted}
+\end{minipage}
+  \caption{Polynomials as Container Extensions (simplified). In order: identity
+  (L4-7), constant (L9-13), product (L15-22), sum (L23-31), composition
+  (L33-44).}
+  \label{fig:poly}
+\end{figure}
+
+\subsubsection{Equality of Container Extensions}
+
+The equality of container extensions needs to be extensional, since the
+contents are obtained by applying a function. We want to equate container
+extensions that have the same shape, and that, for equal shape and position,
+they return the same element. To avoid the use of functional extensionality
+axioms, we capture this relation with the following inductive proposition in
+Coq:
+\begin{minted}{Coq}
+Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
+| AppR_ext (Es : shape x = shape y)
+           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
+\end{minted}
+Note that  we do not care about the validity proof of the positions, only their
+value. This is to simplify (slightly) our proofs. This relation is
+trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.
+
+However, the use of a different equality for container extensions now forces us
+to deal with the fact that some types have different definitions of equality.
+In particular, we want to reason about the equality of functions of types such
+as \mintinline{Coq}{App F A -> B} (or \mintinline{Coq}{B -> App F A}). Since 
+these types now come with their own equivalence, any function that manipulates
+them needs to be \emph{respectful}. I.e.\ given 
+\mintinline{Coq}{R : X -> X -> Prop} and \mintinline{Coq}{R' : Y -> Y -> Prop},
+we want functions (morphisms) that satisfy the following property:
+\begin{minted}{Coq}
+forall (x y : X), R x y -> R' (f x) (f y)
+\end{minted}
+
+\subsection{Types and Morphisms}
+
+We address the different forms of equality by coupling every type with an
+associated equivalence relation, and every morphism with a proof that it is
+\emph{respectful} (i.e. \emph{proper} morphisms with respect to the function
+\emph{respectfulness} relation). We use \mintinline{Coq}{=e} to denote the
+equivalence relation associated with a type. Every Coq type is associated by
+default with the standard Coq propositional equality, unless a different
+equivalence is specified. We use the type-class mechanism to help Coq's code
+extraction mechanism to remove any occurence of custom equivalence relations
+in the extracted OCaml code.
+\begin{minted}{Coq}
+Reserved Notation "f =e g" (at level 70, no associativity).
+Class equiv A : Type :=
+  MkEquiv
+    { eqRel : A -> A -> Prop;
+      e_refl : forall x, eqRel x x;
+      e_sym : forall x y, eqRel x y -> eqRel y x;
+      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
+    }.
+Notation "f =e g" := (eqRel f g).
+\end{minted}
+Given types \mintinline{Coq}{A} and  \mintinline{Coq}{B}, with their respective
+equivalence relations \mintinline{Coq}{eA : A -> A -> Prop} and
+\mintinline{Coq}{eB : B -> B -> Prop}, the we define the type
+\mintinline{Coq}{A ~> B} to represents \emph{proper} morphisms of the
+respectfulness relation of \mintinline{Coq}{R} to \mintinline{Coq}{R'}.
+\begin{minted}{Coq}
+Structure morph A {eA : equiv A} B {eB : equiv B} :=
+  MkMorph
+    { app :> A -> B;
+      app_eq : forall x y, x =e y -> app x =e app y
+    }.
+Notation "A ~> B" := (@morph A _ B _).
+\end{minted}
+Note that we leverage on Coq's type class mechanism to fill in the necessary
+equivalence relations. Since Coq's code extraction mechanism will erase any
+occurrence of \mintinline{Coq}{Prop} in the code, so objects of type
+\mintinline{Coq}{A ~> B} will be extracted to the OCaml equivalent to
+\mintinline{Coq}{A -> B}.
+
+\subsubsection{Polynomial Types}
+
+We define a number of equivalences for polynomial types. These equivalences are
+be used when building proper morphisms that manipulate their respective types.
+\begin{minted}{Coq}
+Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
+Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
+Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
+Instance prop_eq : equiv Prop.
+Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
+\end{minted}
+
+Most of the definitions that involve functions and polynomial types are
+straightworward.  Identity and composition are defined as 
+\mintinline{Coq}{fun x => x} and \mintinline{Coq}{fun f g x => f (g x)} 
+respectively, and the proofs that they are proper morphisms is straightforward,
+and automatically discharged by Coq. 
+Products are built using function \mintinline{Coq}{fun f g x => (f x, g x)},
+with the projections being the standard Coq  \mintinline{Coq}{fst} and
+\mintinline{Coq}{snd} functions. Similarly, sum injections are encoded using
+Coq's \mintinline{Coq}{inl} and \mintinline{Coq}{inr} constructors, and
+pattern matching on them uses the function 
+\mintinline{Coq}{fun f g x => match x with | inl y => f y | inr y => g y end}.
+The proofs that these morphisms are proper are straightforward.  Finally, we
+also provide functions for currying/uncurrying, and partially applying a proper
+morphism. We avoid using definitions, or unfold them, to help Coq's code
+extraction mechanism to inline as many of these small definitions as possible.
+
+These definitions are isomorphic to the equivalent container extension.
+
+We illustrate how these definitions
+\paragraph*{Isomorphisms with Container Extensions} Our definitions involving
+polynomial types are isomorphic to polynomial functors as container extensions.
+
+\subsubsection{Generalised Rewriting for Morphisms}
+
+\section{Extracting Certified Hylomorphisms}
+
+\subsection{Algebras}
+
+Algebras on container extensions are defined as expected. The carrier of the
+algebra is \mintinline{Coq}{A}, and the operations of the algebra are
+morphisms of the following type:
+\begin{minted}{Coq}
+Notation Alg F A := (App F A ~> A).
+\end{minted}
+
+There are \emph{two} initial algebras that are, as expected, unique up to
+isomorphism. We show one such initial algebra here, and the second in the
+following subsection, plus a proof that they are isomorphic:
+
+\begin{minted}{Coq}
+Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
+\end{minted}
+This is why we could not define equality of container extensions generically:
+the contained types may not have (yet) an associated equivalence. We could have
+overcame this circularity by defining simultaneously LFix and its associated
+equivalence relation. But that would complicate definitions unnecessarily.
+\dcas{I know ... I need to rewrite/expand on this}
+
+We provide a custom induction schema for initial algebras:
+
+\begin{minted}{Coq}
+Lemma LFix_ind [P : LFix -> Prop]
+  : (forall x : App F LFix,
+        (forall e : Pos (shape x), P (cont x e)) ->
+        P (LFix_in x))
+    -> forall l : LFix, P l.
+\end{minted}
+
+Equality of initial algebras of container extensions needs to be defined
+recursively and extensionally:
+\begin{minted}{Coq}
+Fixpoint LFixR (x y : LFix) : Prop :=
+  let f_x := LFix_out x in
+  let f_y := LFix_out y in
+  shape f_x =e shape f_y /\
+    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
+\end{minted}
+
+The unique homomorphism from an initial algebra to other algebras, i.e. the catamorphism, is defined as a
+regular recursive function:
+\begin{minted}{Coq}
+Definition cata_f `{eA : equiv A} (g : Alg F A) : LFix -> A
+:= fix f (x : LFix) :=
+  match x with
+  | LFix_in ax =>
+      let (sx, kx) := ax in
+      g (MkCont sx (fun e => f (kx e)))
+  end.
+\end{minted}
+
+It is easy to show that this function is a proper morphism, and that it is
+unique. We can use the fact that catamorphisms are proper morphisms to encode
+them as a higher-order morphism.
+\begin{minted}{Coq}
+Lemma cata_arr1 `{eA : equiv A} (g : Alg F A)
+  : forall x y, x =e y -> cata_f g x =e cata_f g y.
+Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
+
+Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
+    : f =e cata g <-> f =e g \o fmap f \o l_out.
+\end{minted}
+
+\subsection{(Recursive) Coalgebras}
+
+Terminal coalgebras are defined analogously to initial algebras:
+\begin{minted}{Coq}
+CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
+\end{minted}
+
+Their equivalence relation, and anamorphisms, and universal properties are
+easy to encode:
+\begin{minted}{Coq}
+  Definition ana `{eA : equiv A} : Coalg F A ~> A ~> GFix...
+
+  Lemma ana_univ `{eA : equiv A} (h : Coalg F A) (f : A ~> GFix)
+    : f =e ana h <-> f =e g_in \o fmap f \o h.
+\end{minted}
+
+With terminal coalgebras, we can encode another form of initial algebras. In
+particular, the initial algebras based on terminal coalgebras are the carrier
+types of terminal coalgebras, plus a proof that they are finite:
+
+\begin{minted}{Coq}
+Inductive FinF : GFix F -> Prop :=
+| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
+\end{minted}
+
+Using this predicate, we can define the notion of \emph{recursive coalgebras}
+as those coalgebras that build finite trees. Recursive coalgebras are
+implemented in Coq recursively on the proof that they produce finite trees.
+Such proofs will be erased when extracting code to OCaml. This is standard
+practice when building general recursive functions in Coq\footnote{\dcas{This
+technique had a name in the community, but I cannot remember. Something
+predicates}}.
+\begin{minted}{Coq}
+Notation rana_f__ h :=
+  (fix f x H :=
+    let hx := h x in
+    LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e)))).
+
+Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...
+
+Lemma rana_univ A {eA : equiv A} (h : RCoalg A) (f : A ~> LFix F)
+    : f =e rana h <-> f =e l_in \o fmap f \o h.
+\end{minted}
+
+\subsection{Recursive Hylos}
+
+Recursive hylomorphisms are built from an algebra and a recursive coalgebra:
+\begin{minted}{Coq}
+Definition hylo_def (a : Alg F B) (c : Coalg F A)
+: forall (x : A), RecF c x -> B
+:= fix f x H :=
+  match c x as h0
+        return
+        (forall e : Pos (shape h0), RecF c (cont h0 e)) ->
+        B
+  with
+  | MkCont s_x c_x => fun H => a (MkCont s_x (fun e => f (c_x e) (H e)))
+  end (RecF_inv H).
+\end{minted}
+
+\begin{minted}{Coq}
+Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
+    : f =e hylo g h <-> f =e g \o fmap f \o h.
+\end{minted}
+
+Recursive hylo universal property allows us to prove common optimisations:
+
+\begin{minted}{Coq}
+    Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
+    (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
+    : f2 \o hylo g1 h1 =e hylo g2 h1.
+  Proof.
+    rewrite hylo_univ.
+    rewrite fmap_comp.
+    rewrite compA.
+    rewrite <- E2.
+    rewrite <- compA.
+    rewrite <- compA.
+    rewrite (compA g1).
+    rewrite <- hylo_unr.
+    reflexivity.
+  Qed.
+
+  Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
+    (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
+    : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
+
+  Corollary cata_ana_hylo `(F : Cont Sh P) `{equiv A} `{equiv B}
+    (g : Alg F B) (h : RCoalg F A)
+  : cata g \o rana h =e hylo g h.
+\end{minted}
+
+\section{Examples}
+
+\paragraph{Quicksort \& Mergesort}
+
+\dcas{What else?}
+
+\section{Related Work}
+
+\begin{itemize}
+  \item McBride devised a generalisation of Bove and Capretta's general
+recursive function monad~\cite{McBride15, BoveC01}
+  \item Guarded recursion for productive definitions and general recursion
+~\cite{AtkeyM13, PaviottiMB15}
+  \item PACO library~\cite{HurNDV13}
+  \item Sized Types \cite{HughesPS96, abs-1202-3496}
+  \item Copatterns \cite{AbelPTS13, AbelP16}
+  \item IMPORTANT!!!!! Coq extraction from recursion
+    schemes~\cite{larchey2022braga} \dcas{Check bibtex}
+  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
+  \item Recursion schemes in Coq (but not hylos)~\cite{MurataE19}
+  \item EVEN MORE IMPORTANT!!! 2023 POPL paper on divide and conquer code in
+    Coq~\cite{AbreuDHJMS23}.  Some differences to point out: they require
+    functional-extensionality axiom. We are axiom-free. Little discussion about
+    extraction. They focus mostly on erasing termination proofs, but they do
+    not discuss "what does their extracted code look like". Their
+    representation of Functors leads me to think that they will produce loads
+    of "Obj.magic", but we'd need to check this.
+\mvol{We have a story for this now: obj.magic harms interaction with other ocaml code, the resulting ocaml code is not natural to read, and in general more complex/strange extracted code is likely to be less efficient and harder to optimise.}
+\end{itemize}
+
+\subsection{Sized Types}
+
+Another approach to certifying termination of nonstructurally recursive
+functions is the use of \emph{sized types}. Sized types were introduced in
+\cite{HughesPS96} as a way to track/verify various properties of recursive
+programs, including productivity and termination. The core idea of sized types
+is that types express bounds on the sizes of recursive data structures. This
+approach has been used to express nonstructurally recursive algorithms in Agda:
+\cite{CopelloTB14} used sized types in a straightforward formulation of
+mergesort. Such an approach has also been used in MiniAgda~\cite{abs-1202-3496}.
+
+\bibliography{references.bib}
+\appendix
+
 \section{Adjoint Folds}
@@
 
-\section{Mechanising Extractable Container Functors}
-
-\dcas{Subsections are tentative topics to remember what I need to fill in here,
-but these are not final}
-
-Our mechanisation works in the category of Coq types, and our goals are: (1)
-the mechanisation of hylomorphisms and their laws should be axiom-free; and (2)
-Coq hylomorphisms should be extracted to recursive OCaml functions that use no
-instances of \mintinline{OCaml}{Obj.magic}.  Although the use of
-\mintinline{OCaml}{Obj.magic} is safe in the code extracted by Coq, it has been
-proven that, for higher-order programs, simple interoperations can lead to
-incorrect behaviour or even segfaults~\cite{forster:hal-04329663}.  Therefore,
-the avoidance of axioms and  \mintinline{OCaml}{Obj.magic} means that the
-extracted code will have stronger safety guarantees, and will allow the safe
-interoperation with other OCaml code.
-
-\subsection{Functors and Containers}
-
-In Coq, to construct least/greatest fixpoints of functors, these need to build
-\emph{strictly positive types}. A common approach to construct such strictly
-positive endofunctors is to use \emph{container functors}~\cite{Whatever}.
-However, reasoning about container and container functor equality requires to
-consider: (a) functional extensionality; (b) heterogeneous equality. This is
-problematic, due to our goals to remain axiom-free. Furthermore, standard
-representations of containers in Coq will lead to OCaml code that contains
-\mintinline{OCaml}{Obj.magic}. We will explain why, and our mechanisation of
-containers in the remainder of this section.
-
-A container is defined by a type of \emph{shapes}, and a \emph{family of
-position types} indexed by shapes. 
-\begin{minted}{Coq}
-Context (Shape : Set) (Pos : Shape -> Set).
-\end{minted}
-An \emph{extension} of this container is a functor,
-whose action on objects is given by:
-\begin{minted}{Coq}
-Record App (X : Set) := { shape : Shape; contents : Pos shape -> X }.
-\end{minted}
-In our mechanisation, we use \mintinline{Coq}{App} to represent container
-extensions.  Generally, any functor that is isomorphic to a container extension
-is called a \emph{container functor}. However, in our work we will call
-\emph{container functor} to the functors defined by a specific container
-extension.
-
-The first problem in extracting Coq code from such container extensions is the
-use of families of types. Since OCaml's type system cannot accurately capture
-such families of types, Coq's code extraction mechanism will insert unsafe
-casts. To avoid unsafe casts, we can consider representing the positions of a
-container as a type that captures \emph{all} the potential positions in all
-possible shapes, together with a predicate \mintinline{Coq}{Valid} that
-determines whether a position is valid in a shape. 
-\begin{minted}{Coq}
-Context (Shape : Set) (Pos : Set) (Valid : Shape -> Pos -> Prop).
-Record App (X : Set) := { shape    : Shape; 
-                          contents : {p | Valid shape p} -> X }.
-\end{minted}
-This approach solves the the problem of the unsafe casts, since the code
-extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml
-singleton type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.
-
-However, there still remains the issue of reasoning about the equality of
-container extensions. The contents of the container extensions need to be
-accessed via a function \mintinline{Coq}{contents}. The domain of this function
-depends on the shape of the container. This means that, for a given type of
-shapes and positions, the equality of two objects of a container extension will
-need to deal with heterogeneous equality. Since we want to remain axiom-free,
-one way to deal with this is to rely on a \emph{decidable} validity predicate.
-This way, we reach our final definition for container extensions. First,
-containers are defined as pairs of shape and position types that have a
-\emph{decidable} validity predicate. We represent decidable predicates as
-functions that return a boolean, and lift booleans to propositions similarly to
-SSReflect. For simplicity, we define the dependent pair of valid positions as a
-record, which will be extracted to a singleton type with the validity proof
-erased by Coq's code extraction mechanism.
-\begin{minted}{Coq}
-Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
-Coercion is_true (b : bool) : Prop := b = true.
-Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
-  ValidPos { val : Pos; IsValid : valid shape pos }.
-\end{minted}
-Then, we define \emph{container extensions} as
-\begin{minted}{Coq}
-Record App `(C : Cont Shape Pos) (X : Set) :=
-  { shape : Shape; contents : PosOf C shape -> X }.
-\end{minted}
-The functorial action for container extensions on morphisms boils down to
-function composition, which trivially preserves identity and composition.
-\begin{minted}{Coq}
-Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
-  := {| shape := shape x; contents := fun e => f (contents x e) |}
-\end{minted}
-
-\subsubsection{Polynomial Functors as Container Extensions}
-
-Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
-omit here the proofs that these container extensions are isomorphic to
-polynomials, but this proof is straightforward. To illustrate how the identity,
-constant, sum and product container extensions work, consider the
-tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
-$F\; X = \mathbb{N} + X \times X$.
-
-This functor is built as the sum of two functors: $X \mapsto \mathbb{N}$ and $X \mapsto
-X \times X$. Consider first $X \mapsto \mathbb{N}$. This is a constant functor
-that maps $X$ to the type $\mathbb{N}$. More generally, a functor that, given a
-type $A$, maps everything to it, $X \mapsto A$, only needs one shape ($A$, L9
-of Fig~\ref{fig:poly}), and zero positions (In Coq, \mintinline{Coq}{False},
-L10 of Fig~\ref{fig:poly}). The reason why there are zero valid positions in
-a constant container extension is because, given something of type 
-\mintinline{Coq}{App (Const (Ks A) Kp) X}, it should be impossible to get
-something of type \mintinline{Coq}{X}. But from the container shape, we will
-always be able to get the element of type $A$\footnote{%
-  Our decision to use \mintinline{Coq}{False} is arbitrary. In general, we
-  could have chosen any time, with a validity predicate that always returns
-  \mintinline{Coq}{false}. It would make no difference for code extraction.
-}. This functor is, therefore \mintinline{Coq}{Const nat}.
-
-Consider now $X \mapsto X \times X$. This is the product of two
-identity functors $X \mapsto X$. The identity functor only has one shape, and
-we can use the unit type to represent such shapes. From this shape we should be
-able to get an element of type $X$, and we can capture this by using the unit
-type for positions. To build an element of the container extension type, we
-will need to provide a function of type \mintinline{Coq}{unit -> X}, which is
-isomorphic to \mintinline{Coq}{X}.
-
-The shape of the product of two container extensions is the product of their
-shapes. The product of such container extensions should allow us to get either
-the left, or the right element in the product. Therefore, the positions of this
-container extension is the sum of the positions of the two container
-extensions. What is a valid position in a given shape? If the position is
-\mintinline{Coq}{inl p} (or \mintinline{Coq}{inr p}), then it is only valid if
-\mintinline{Coq}{p} is valid in the first (resp. second) component of the
-shape. The product of our two container extensions is
-\mintinline{Coq}{Prod Id Id}.
-
-Finally, the shape the sum of two container extensions is the sum of their
-shapes. The positions are, like for products, the sum of their shapes. But to
-check the validity of shapes in positions, their constructor must match, e.g.
-shape \mintinline{Coq}{inl s} and shape \mintinline{Coq}{inl p}, with
-\mintinline{Coq}{valid s p}.
-
-Our final container extension is the following: 
-\mintinline{Coq}{Sum (Const nat) (Prod Id Id)}.
-The shape of this functor is: 
-\mintinline{Coq}{nat + unit * unit}. The positions of this functor
-are \mintinline[breaklines]{Coq}{False + (unit + unit)}, which is isomorphic
-to \mintinline{Coq}{unit + unit}, which will only be valid in the RHS.
-
-The Coq value \mintinline{Coq}{inl n} of type
-\mintinline{Coq}{nat + bool * bool} would look as follows as a container
-extension 
-\mintinline{Coq}{App (Sum (Const nat) (Prod Id Id)) bool}:
-\begin{minted}{Coq}
-{| shape := inl n; cont := fun e => uninhabited e|}
-\end{minted}
-We omit the definition of \mintinline{Coq}{uninhabited}, which exploits the
-fact that \mintinline{Coq}{e} for \mintinline{Coq}{inl n} implies
-\mintinline{Coq}{False} to build an element of type \mintinline{Coq}{bool}%
-\footnote{Note that, since \mintinline{Coq}{e} is uninhabited we could return
-any arbitrary value instead of \mintinline{Coq}{uninhabited e}.}.
-
-Similarly, the Coq value \mintinline{Coq}{inr (true, false)} as a container
-extension is as follows:
-\begin{minted}{Coq}
-  {| shape := inr (tt,tt);
-     cont := fun e : {p | valid (inr (tt, tt)) p} =>
-        match proj1_sig e with
-        | inl f => match f with end
-        | inr (inl tt) => fun _ => true
-        | inr (inr tt) => fun _ => false
-        end
-  |}
-\end{minted}
-Here, the position \mintinline{Coq}{inl f} is impossible, as
-\mintinline{Coq}{f} would need to be a proof of \mintinline{Coq}{False}.
-Of the two remaining positions, the left one will return 
-\mintinline{Coq}{true}, and the right position \mintinline{Coq}{false}.
-\begin{figure}
-\begin{minipage}{.52\textwidth}
-  \begin{minted}[linenos, fontsize=\small]{Coq}
-Context `(C1 : Cont S1 P1) 
-        `(C2 : Cont S2 P2).
-
-Definition Is := unit.
-Definition Ip := unit.
-Instance Id : Cont Is Ip := {
-  valid := fun _ _ => true }.
-
-Definition Ks (A : Set) := A.
-Definition Kp := False.
-Instance Const A : Cont (Ks A) Kp := {
-  valid := fun _ p =>
-    match p with end }.
-
-Definition Ps : Type := S1 * S2.
-Definition Pp : Type := P1 + P2.
-Instance Prod : Cont Ps Pp := {
-  valid := fun s p => 
-    match p with
-    | inl p' => valid (fst s) p'
-    | inr p' => valid (snd s) p'
-    end }.
-\end{minted}
-\end{minipage}
-\begin{minipage}{.47\textwidth}
-  \begin{minted}[linenos, firstnumber=23, fontsize=\small]{Coq}
-Definition Ss : Type := S1 + S2.
-Definition Sp : Type := P1 + P2.
-Instance Sum : Cont Ss Sp := {
-  valid := fun s p => 
-    match s, p with
-    | inl s', inl p' => valid s' p'
-    | inr s', inr p' => valid s' p'
-    | _, _ => false
-    end }.
-
-Definition Cs : Type := App F S2.
-Definition Cp : Type := P1 * P2.
-Instance Comp : Cont Cs Cp := 
-{ valid :=  
-    fun (s : App F S2) (p : P1 * P2) =>
-        match valid (shape s) (fst p) 
-        with
-        | true  => fun H =>
-            valid (cont s {fst p | H}) 
-                  (snd p)
-        | false => fun _ => false
-        end eq_refl }.
-  \end{minted}
-\end{minipage}
-  \caption{Polynomials as Container Extensions (simplified). In order: identity
-  (L4-7), constant (L9-13), product (L15-22), sum (L23-31), composition
-  (L33-44).}
-  \label{fig:poly}
-\end{figure}
-
-\subsubsection{Equality of Container Extensions}
-
-The equality of container extensions needs to be extensional, since the
-contents are obtained by applying a function. We want to equate container
-extensions that have the same shape, and that, for equal shape and position,
-they return the same element. To avoid the use of functional extensionality
-axioms, we capture this relation with the following inductive proposition in
-Coq:
-\begin{minted}{Coq}
-Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
-| AppR_ext (Es : shape x = shape y)
-           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
-\end{minted}
-Note that  we do not care about the validity proof of the positions, only their
-value. This is to simplify (slightly) our proofs. This relation is
-trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.
-
-However, the use of a different equality for container extensions now forces us
-to deal with the fact that some types have different definitions of equality.
-In particular, we want to reason about the equality of functions of types such
-as \mintinline{Coq}{App F A -> B} (or \mintinline{Coq}{B -> App F A}). 
-We want the following kind of properties to hold: 
-\begin{minted}{Coq}
-forall (x y : App C A) (f : App C A -> B), AppR x y -> f x = f y
-\end{minted}
-Moreover, we want this to be true for any \mintinline{Coq}{B}, and equality.
-We address this in the following section.
-
-\subsection{Types and Morphisms}
-
-The need to reason about different forms of equality forces us to couple every
-type with an associated equivalence relation. Furthermore, we will only allow
-to reason about the equality of \emph{proper morphisms}, i.e. morphisms that
-respect the necessary equivalence relations.
-
-\begin{minted}{Coq}
-Reserved Notation "f =e g" (at level 70, no associativity).
-Class equiv A : Type :=
-  MkEquiv
-    { eqRel : A -> A -> Prop;
-      e_refl : forall x, eqRel x x;
-      e_sym : forall x y, eqRel x y -> eqRel y x;
-      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
-    }.
-Notation "f =e g" := (eqRel f g).
-
-Reserved Notation "x ~> y" (at level 95, right associativity, y at level 200).
-Context `{eA : equiv A} `{eB : equiv B}.
-
- Structure morph :=
-   MkMorph
-     { app :> A -> B;
-       app_eq : forall x y, x =e y -> app x =e app y
-     }.
-Arguments morph A {eA} B {eB}.
-Notation "x ~> y" := (morph x y).
-\end{minted}
-
-\dcas{We provide common morphisms for polynomials: sums, products, const,
-exponential (Add examples?)}
-
-By default, we associate every type with Coq's propositional equality, but allow
-the use of other forms of equality. In most cases, we will use either
-propositional equality, extensional equality, or a mixture of both, e.g. for
-container extensions, we use an arbitrary equality for shapes, extensional
-equality for the contained elements, but Coq's propositional equality for
-positions:
-\begin{minted}{Coq}
-  Inductive AppR `{Cont Sh P} `{equiv X} (x y : App F X) : Prop :=
-  | AppR_ext (Es : shape x =e shape y)
-             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
-\end{minted}
-Notice the use of \mintinline{Coq}{=e} for both shapes, and the continuations
-of container extensions. 
-The reason to use Coq's propositional equality for positions is to avoid a
-circularity in the definitions: arbitrary equivalences for positions would
-force us to define the contents as morphisms from positions to the contained
-elements, but that would as well force us to redefine container extensions so
-that they are only applicable to types with arbitrary equalities and morphisms.
-We will see in the following section why this is problematic for defining
-initial algebras of container extensions.
-\dcas{I am actually not sure of what I wrote above, and it may be something I
-simply overlooked. I will look into the Coq code and try to generalise it if I
-have the time}
-
-\subsubsection{Polynomial Types, Equivalences, and Morphisms}
-We associate some types with their standard equivalences:
-\begin{minted}{Coq}
-Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
-Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
-Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
-Instance prop_eq : equiv Prop.
-Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
-\end{minted}
-
-\subsubsection{Isomorphisms with Container Extensions}
-
-\section{Extracting Certified Hylomorphisms}
-
-\subsection{Algebras}
-
-Algebras on container extensions are defined as expected. The carrier of the
-algebra is \mintinline{Coq}{A}, and the operations of the algebra are
-morphisms of the following type:
-\begin{minted}{Coq}
-Notation Alg F A := (App F A ~> A).
-\end{minted}
-
-There are \emph{two} initial algebras that are, as expected, unique up to
-isomorphism. We show one such initial algebra here, and the second in the
-following subsection, plus a proof that they are isomorphic:
-
-\begin{minted}{Coq}
-Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
-\end{minted}
-This is why we could not define equality of container extensions generically:
-the contained types may not have (yet) an associated equivalence. We could have
-overcame this circularity by defining simultaneously LFix and its associated
-equivalence relation. But that would complicate definitions unnecessarily.
-\dcas{I know ... I need to rewrite/expand on this}
-
-We provide a custom induction schema for initial algebras:
-
-\begin{minted}{Coq}
-Lemma LFix_ind [P : LFix -> Prop]
-  : (forall x : App F LFix,
-        (forall e : Pos (shape x), P (cont x e)) ->
-        P (LFix_in x))
-    -> forall l : LFix, P l.
-\end{minted}
-
-Equality of initial algebras of container extensions needs to be defined
-recursively and extensionally:
-\begin{minted}{Coq}
-Fixpoint LFixR (x y : LFix) : Prop :=
-  let f_x := LFix_out x in
-  let f_y := LFix_out y in
-  shape f_x =e shape f_y /\
-    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
-\end{minted}
-
-The unique homomorphism from an initial algebra to other algebras, i.e. the catamorphism, is defined as a
-regular recursive function:
-\begin{minted}{Coq}
-Definition cata_f `{eA : equiv A} (g : Alg F A) : LFix -> A
-:= fix f (x : LFix) :=
-  match x with
-  | LFix_in ax =>
-      let (sx, kx) := ax in
-      g (MkCont sx (fun e => f (kx e)))
-  end.
-\end{minted}
-
-It is easy to show that this function is a proper morphism, and that it is
-unique. We can use the fact that catamorphisms are proper morphisms to encode
-them as a higher-order morphism.
-\begin{minted}{Coq}
-Lemma cata_arr1 `{eA : equiv A} (g : Alg F A)
-  : forall x y, x =e y -> cata_f g x =e cata_f g y.
-Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
-
-Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
-    : f =e cata g <-> f =e g \o fmap f \o l_out.
-\end{minted}
-
-\subsection{(Recursive) Coalgebras}
-
-Terminal coalgebras are defined analogously to initial algebras:
-\begin{minted}{Coq}
-CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
-\end{minted}
-
-Their equivalence relation, and anamorphisms, and universal properties are
-easy to encode:
-\begin{minted}{Coq}
-  Definition ana `{eA : equiv A} : Coalg F A ~> A ~> GFix...
-
-  Lemma ana_univ `{eA : equiv A} (h : Coalg F A) (f : A ~> GFix)
-    : f =e ana h <-> f =e g_in \o fmap f \o h.
-\end{minted}
-
-With terminal coalgebras, we can encode another form of initial algebras. In
-particular, the initial algebras based on terminal coalgebras are the carrier
-types of terminal coalgebras, plus a proof that they are finite:
-
-\begin{minted}{Coq}
-Inductive FinF : GFix F -> Prop :=
-| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
-\end{minted}
-
-Using this predicate, we can define the notion of \emph{recursive coalgebras}
-as those coalgebras that build finite trees. Recursive coalgebras are
-implemented in Coq recursively on the proof that they produce finite trees.
-Such proofs will be erased when extracting code to OCaml. This is standard
-practice when building general recursive functions in Coq\footnote{\dcas{This
-technique had a name in the community, but I cannot remember. Something
-predicates}}.
-\begin{minted}{Coq}
-Notation rana_f__ h :=
-  (fix f x H :=
-    let hx := h x in
-    LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e)))).
-
-Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...
-
-Lemma rana_univ A {eA : equiv A} (h : RCoalg A) (f : A ~> LFix F)
-    : f =e rana h <-> f =e l_in \o fmap f \o h.
-\end{minted}
-
-\subsection{Recursive Hylos}
-
-Recursive hylomorphisms are built from an algebra and a recursive coalgebra:
-\begin{minted}{Coq}
-Definition hylo_def (a : Alg F B) (c : Coalg F A)
-: forall (x : A), RecF c x -> B
-:= fix f x H :=
-  match c x as h0
-        return
-        (forall e : Pos (shape h0), RecF c (cont h0 e)) ->
-        B
-  with
-  | MkCont s_x c_x => fun H => a (MkCont s_x (fun e => f (c_x e) (H e)))
-  end (RecF_inv H).
-\end{minted}
-
-\begin{minted}{Coq}
-Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
-    : f =e hylo g h <-> f =e g \o fmap f \o h.
-\end{minted}
-
-Recursive hylo universal property allows us to prove common optimisations:
-
-\begin{minted}{Coq}
-    Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
-    (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
-    : f2 \o hylo g1 h1 =e hylo g2 h1.
-  Proof.
-    rewrite hylo_univ.
-    rewrite fmap_comp.
-    rewrite compA.
-    rewrite <- E2.
-    rewrite <- compA.
-    rewrite <- compA.
-    rewrite (compA g1).
-    rewrite <- hylo_unr.
-    reflexivity.
-  Qed.
-
-  Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
-    (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
-    : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
-
-  Corollary cata_ana_hylo `(F : Cont Sh P) `{equiv A} `{equiv B}
-    (g : Alg F B) (h : RCoalg F A)
-  : cata g \o rana h =e hylo g h.
-\end{minted}
-
-\section{Examples}
-
-\paragraph{Quicksort \& Mergesort}
-
-\dcas{What else?}
-
-\section{Related Work}
-
-\begin{itemize}
-  \item McBride devised a generalisation of Bove and Capretta's general
-recursive function monad~\cite{McBride15, BoveC01}
-  \item Guarded recursion for productive definitions and general recursion
-~\cite{AtkeyM13, PaviottiMB15}
-  \item PACO library~\cite{HurNDV13}
-  \item Sized Types \cite{HughesPS96, abs-1202-3496}
-  \item Copatterns \cite{AbelPTS13, AbelP16}
-  \item IMPORTANT!!!!! Coq extraction from recursion
-    schemes~\cite{larchey2022braga} \dcas{Check bibtex}
-  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
-  \item Recursion schemes in Coq (but not hylos)~\cite{MurataE19}
-  \item EVEN MORE IMPORTANT!!! 2023 POPL paper on divide and conquer code in
-    Coq~\cite{AbreuDHJMS23}.  Some differences to point out: they require
-    functional-extensionality axiom. We are axiom-free. Little discussion about
-    extraction. They focus mostly on erasing termination proofs, but they do
-    not discuss "what does their extracted code look like". Their
-    representation of Functors leads me to think that they will produce loads
-    of "Obj.magic", but we'd need to check this.
-\mvol{We have a story for this now: obj.magic harms interaction with other ocaml code, the resulting ocaml code is not natural to read, and in general more complex/strange extracted code is likely to be less efficient and harder to optimise.}
-\end{itemize}
-
-\subsection{Sized Types}
-
-Another approach to certifying termination of nonstructurally recursive
-functions is the use of \emph{sized types}. Sized types were introduced in
-\cite{HughesPS96} as a way to track/verify various properties of recursive
-programs, including productivity and termination. The core idea of sized types
-is that types express bounds on the sizes of recursive data structures. This
-approach has been used to express nonstructurally recursive algorithms in Agda:
-\cite{CopelloTB14} used sized types in a straightforward formulation of
-mergesort. Such an approach has also been used in MiniAgda~\cite{abs-1202-3496}.
-
-\bibliography{references.bib}
 \end{document}

### 2024-02-22 paper/paper.tex
Working on background section 2.

@@
 \newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
-\newcommand{\dcas}[1]{\textcolor{red}{\textsc{David}: #1}}
-\newcommand{\mvol}[1]{\textcolor{red}{\textsc{Michael}: #1}}
+\newcommand{\dcas}[1]{\textcolor{green}{\textsc{David}: #1}}
+\newcommand{\mvol}[1]{\textcolor{blue}{\textsc{Michael}: #1}}
 \newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
@@
 \newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
-
+\newcommand{\coq}[1]{\mintinline{coq}{#1}}
 \title{A Mechanised Library for Generic (Co)Programming}
@@
 code):
-\begin{minted}{haskell}
-partition :: [a] -> ([a], a, [a])
-partition (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
-
-quicksort :: [a] -> [a]
-quicksort [] = []
-quicksort ls = let (xs, x, ys) = partition ls in
-                        (quicksort xs) ++ [x] ++ (quicksort ys)
-\end{minted}
 While \haskell{quicksort} is a well-defined mathematical function it cannot be
@@
 
-\section{Recursive Equations}
-
-\subsection{Catamorphisms and Anamorphism}
-A catamorphism is the unique program defined as
-\[
-  x = a \comp \fmap{x} \comp \InOp
-\]
-Can be implemented:
-\[
-  \cata{a} = \hylo{\InOp}{a}
-\]
-
-An anamorphism is the unique program satisfying the equation
-\[
-  x = \OutOp \comp \fmap{x} \comp{c}
-\]
-The anamorphism is indicated by $\ana{c}$. Recursion schemes given by a
-corecursive algebra such as $\OutIso$ are also called hylomorphism but they are
-indicated with the symbol for the anamorphism, $\cohylo{c}{\OutOp}$.
-
-\paragraph{Free Monads}
-Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
-solution to the following domain equation
-\[
-  D^{*}A = A + DD^{*}A
-\]
-where $\eta_{A} : A \to D^{*}A$ is the left injection and
-$\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
-$a : DA \to A$ there is a unique solution to the following recursive equation
-\[
-  x \comp \call_{A} = a \comp D(x)
-\]
-that is $\catafree{a} = \cata{[id_{A}; a]}$. With this operation we can extend
-the free algebra $\call$ to the multiplication $\mu : D^{*}D^{*}A \to D^{*}A$
-defined as $\mu = \catafree{\call}$.
-
-\paragraph{Cofree Comonads}
-Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
-greatest solution to the following domain equation
-\[
-  D_{\infty}A = A + DD_{\infty}A
-\]
-where $\head : D_{\infty}A \to A$ is the first projection and
-$\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.
-Given an algebra $c : A \to DA$ there is a unique solution to the following
-recursive equation
-\[
-  \tail_{A} \comp x = D(x) \comp c
-\]
-that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$. With the latter operation we
-can extend $\tail$ to the comultiplication for the comonad
-$\delta : D_{\infty}A \to D_{\infty} D_{\infty}A$ defined by
-$\delta = \anacofree{\tail}$.
-
-\subsection{Accumulators}
-For an algebra $a : D(A^{P})\times P \to A$:
-\begin{align*}
-  & g : \mu D \times P \to B\\
-  & g (d, p) = a (\fmap{(\curry g)} \comp \InOp(d), p)
-\end{align*}
-
-\paragraph{Hylo Implementation}
-Given an algebra $a : D(A^{P}) \to A^{P}$ the accumorphism corresponds to the unique
-solution to the following equation
-\begin{align*}
-  & f : \mu D \to A^{P}\\
-  & f = a \comp \fmap (f) \comp \InOp
-\end{align*}
-thus $f = \hylo{\InOp}{a}$.
-
-The complex recursion scheme is then
-$g = \uncurry\hylo{\InOp}{a} : \mu D \times P \to B$.
-
-\subsection{Mutu-Hylo}
-For an algebra $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and an algebra
-$a_{2} : D(A_{1} \times A_{2}) \to A_{2}$:
-\begin{align*}
-  & (g_{1}, g_{2}) : (\mu D, \mu D) \to (A_{1}, A_{2})\\
-  & (g_{1}, g_{2}) = (a_{1}, a_{2}) \comp (\fmap{\pair{g_{1}}{g_{2}}},\fmap{\pair{g_{1}}{g_{2}}}) \comp (\InOp, \InOp)
-\end{align*}
-
-\paragraph{Hylo Implementation}
-Given an algebra $a : D(A_{1} \times A_{2}) \to A_{1} \times A_{2}$
-\begin{align*}
-  & f : \mu D \to A_{1} \times A_{2}\\
-  & f = a \comp \fmap{f} \comp \InOp
-\end{align*}
-for which the solution is an hylo $f = \hylo{\InOp}{a}$.
-
-The mutumorphism can be implemented as
-$(g_{1}, g_{2}) = (\pi_{1} \comp f, \pi_{2} \comp f)$.
-
-\paragraph{Coq Implementation}
-Given the two algebras  $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and
-$a_{2} : D(A_{1} \times A_{2}) \to A_{2}$.
-We define the two recursive functions  $f_{1}$ and $f_{2}$ as
-
-\[
-  f_{1} = \pi_{1} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}} \qquad f_{2} = \pi_{2} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}}
-\]
-
-\subsubsection{Example}
+\section{Background on Recursion Schemes}
+Recursion schemes provide a systematic way to work with data types, such as
+trees or lists, by abstracting away the recursion into higher-order functions.
+Canonical recursion schemes include \emph{folds} (catamorphisms) which consume
+data and \emph{unfolds} (anamorphisms) which produce data. These are the unique
+solutions (when they exist) to the following two equations respectively
+\[
+  h \comp \InIso = a \comp F h \qquad \OutIso \comp k = F k \comp c
+\]
+where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
+can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
+$F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
+$F$-coalgebra), $\InIso$ is the isomoprhism on the least fixed-point of $F$,
+namely $\mu F \xrightarrow{\iso} F\mu F$, and $\OutIso$ is the isomorphism on
+the greatest fixed-point of $F$, namely $\nu F \xrightarrow{\iso} F\nu F$. When
+the unique solutions to the above equations exist we denote them as $\cata{a}$
+and $\ana{c}$ respectively.
+
+However, catamorphisms (dually anamorphisms) do not capture every recursion
+scheme we can think of.  As an example, consider the quicksort
+algorithm:\footnote[1]{This program is written in Coq notation to be consistent,
+but it is not a program that Coq can accept.}
+
+\mpav{I need this to be written in Coq notation}
+\begin{minted}{haskell}
+partition :: [a] -> ([a], a, [a])
+partition (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
+
+quicksort :: [a] -> [a]
+quicksort [] = []
+quicksort ls = let (xs, x, ys) = partition ls in
+                        (quicksort xs) ++ [x] ++ (quicksort ys)
+\end{minted}
+This program is formed of essentially two parts. The first one is the
+$\operator{partition}$ program destructuring the input list in three parts, the
+pivot and the other two sublists, and the second part recursing over the
+sublists and composing the results back together.
+
+Notice that if we define the functor \coq{TreeF} $X = X \times A \times X$ with
+the obvious functorial action then \coq{partition} can be viewed as a
+\coq{TreeF}-coalgebra on lists over the type $A$. While we can define an \coq{TreeF}-algebra on lists over $A$
+as well, that is a map \coq{alg} $:$ \coq{TreeF} $[A] \to [A]$ defined
+as \coq{alg (left, x, right) = left ++ [x] ++ right}.
+
+More specifically, in the context of recursive data types, such as in Haskell,
+where induction and coinduction is conflated, an hylomorphism can be implemented
+in terms of a catarmorphism (folding) and an anamorphism (unfolding)
+\[
+  \hylo{a}{c} = \cata{a} \comp \ana{c}
+\]
+They first generate a data structure using an anamorphism and then immediately
+fold or reduce it using a catamorphism.
+
+The hylomorphism pattern is particularly useful for defining recursive
+algorithms in a functional programming context, where recursion is a fundamental
+tool for solving problems. By separating the generation of data structures from
+their consumption, hylomorphisms enable clearer and more modular code, making it
+easier to reason about and maintain recursive algorithms.
+
+However, in a type theory inductive and coinductive data types are separate.
+
+In this section we will introduce the notion of hylomorphism. By way of examples
+we will also give high level intution as to way all recursion schemes can be
+defined in terms of a basic
+
 Say that we want to prove that the two functions below are well-defined
@@
 
-One option is to prove that the following function (given by a catamorphism) is
+One option is to prove that the following function is
 equivalent to the above
@@
 
+\subsection{Catamorphisms and Anamorphism}
+A catamorphism is the unique program defined as
+\[
+  x = a \comp \fmap{x} \comp \InOp
+\]
+Can be implemented:
+\[
+  \cata{a} = \hylo{\InOp}{a}
+\]
+
+An anamorphism is the unique program satisfying the equation
+\[
+  x = \OutOp \comp \fmap{x} \comp{c}
+\]
+The anamorphism is indicated by $\ana{c}$. Recursion schemes given by a
+corecursive algebra such as $\OutIso$ are also called hylomorphism but they are
+indicated with the symbol for the anamorphism, $\cohylo{c}{\OutOp}$.
+
+\paragraph{Free Monads}
+Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
+solution to the following domain equation
+\[
+  D^{*}A = A + DD^{*}A
+\]
+where $\eta_{A} : A \to D^{*}A$ is the left injection and
+$\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
+$a : DA \to A$ there is a unique solution to the following recursive equation
+\[
+  x \comp \call_{A} = a \comp D(x)
+\]
+that is $\catafree{a} = \cata{[id_{A}; a]}$. With this operation we can extend
+the free algebra $\call$ to the multiplication $\mu : D^{*}D^{*}A \to D^{*}A$
+defined as $\mu = \catafree{\call}$.
+
+\paragraph{Cofree Comonads}
+Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
+greatest solution to the following domain equation
+\[
+  D_{\infty}A = A + DD_{\infty}A
+\]
+where $\head : D_{\infty}A \to A$ is the first projection and
+$\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.
+Given an algebra $c : A \to DA$ there is a unique solution to the following
+recursive equation
+\[
+  \tail_{A} \comp x = D(x) \comp c
+\]
+that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$. With the latter operation we
+can extend $\tail$ to the comultiplication for the comonad
+$\delta : D_{\infty}A \to D_{\infty} D_{\infty}A$ defined by
+$\delta = \anacofree{\tail}$.
+
+\subsection{Accumulators}
+For an algebra $a : D(A^{P})\times P \to A$:
+\begin{align*}
+  & g : \mu D \times P \to B\\
+  & g (d, p) = a (\fmap{(\curry g)} \comp \InOp(d), p)
+\end{align*}
+
+\paragraph{Hylo Implementation}
+Given an algebra $a : D(A^{P}) \to A^{P}$ the accumorphism corresponds to the unique
+solution to the following equation
+\begin{align*}
+  & f : \mu D \to A^{P}\\
+  & f = a \comp \fmap (f) \comp \InOp
+\end{align*}
+thus $f = \hylo{\InOp}{a}$.
+
+The complex recursion scheme is then
+$g = \uncurry\hylo{\InOp}{a} : \mu D \times P \to B$.
+
+\subsection{Mutu-Hylo}
+For an algebra $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and an algebra
+$a_{2} : D(A_{1} \times A_{2}) \to A_{2}$:
+\begin{align*}
+  & (g_{1}, g_{2}) : (\mu D, \mu D) \to (A_{1}, A_{2})\\
+  & (g_{1}, g_{2}) = (a_{1}, a_{2}) \comp (\fmap{\pair{g_{1}}{g_{2}}},\fmap{\pair{g_{1}}{g_{2}}}) \comp (\InOp, \InOp)
+\end{align*}
+
+\paragraph{Hylo Implementation}
+Given an algebra $a : D(A_{1} \times A_{2}) \to A_{1} \times A_{2}$
+\begin{align*}
+  & f : \mu D \to A_{1} \times A_{2}\\
+  & f = a \comp \fmap{f} \comp \InOp
+\end{align*}
+for which the solution is an hylo $f = \hylo{\InOp}{a}$.
+
+The mutumorphism can be implemented as
+$(g_{1}, g_{2}) = (\pi_{1} \comp f, \pi_{2} \comp f)$.
+
+\paragraph{Coq Implementation}
+Given the two algebras  $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and
+$a_{2} : D(A_{1} \times A_{2}) \to A_{2}$.
+We define the two recursive functions  $f_{1}$ and $f_{2}$ as
+
+\[
+  f_{1} = \pi_{1} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}} \qquad f_{2} = \pi_{2} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}}
+\]
 % A recursion scheme from a comonad $D_{\infty}$ for a data functor $F$
@@
 
-\subsection{Course-of-Value Recursion with Finite Memory Table}
-It can be proven that if $c$ is a recursive coalgebra, then the hylomorphism for
-any algebra $a$ generates a finite output.  In other words, as far as the
-hylomorphism is concerned $G_{\infty}$ can be regarded as its finite version
-\[
-  G_{*} A = \mu X. A \times G X
-\]
-We now rewrite course-of-value recursion using the finite memory table
-For an algebra $a : D(D_{*} A) \to A$,
-\begin{align*}
-  & g : \mu F \to A
-  & g = a \comp D(G_{*}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
-\end{align*}
-
 \section{Mechanising Extractable Container Functors}
@@
 \emph{strictly positive types}. A common approach to construct such strictly
-positive endofunctors is to use \emph{container functors}~\cite{Whatever}.
+positive endofunctors is to use \emph{container functors}~\cite{AbbottAG05}.
 However, reasoning about container and container functor equality requires to
@@
 \appendix
+\subsection{Course-of-Value Recursion with Finite Memory Table}
+It can be proven that if $c$ is a recursive coalgebra, then the hylomorphism for
+any algebra $a$ generates a finite output.  In other words, as far as the
+hylomorphism is concerned $G_{\infty}$ can be regarded as its finite version
+\[
+  G_{*} A = \mu X. A \times G X
+\]
+We now rewrite course-of-value recursion using the finite memory table
+For an algebra $a : D(D_{*} A) \to A$,
+\begin{align*}
+  & g : \mu F \to A
+  & g = a \comp D(G_{*}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
+\end{align*}
 

### 2024-02-22 paper/paper.tex
QSort & MSort with nice extraction

@@
 Algebras on container extensions are defined as expected. The carrier of the
-algebra is \mintinline{Coq}{A}, and the operations of the algebra are
-morphisms of the following type:
+algebra is \mintinline{Coq}{A}, and the operations of the algebra are morphisms
+of the following type:
 \begin{minted}{Coq}
@@
 \end{minted}
-
-There are \emph{two} initial algebras that are, as expected, unique up to
+We define \emph{two} initial algebras that are, as expected, unique up to
 isomorphism. We show one such initial algebra here, and the second in the
 following subsection, plus a proof that they are isomorphic:
-
 \begin{minted}{Coq}
@@
 \end{minted}
-This is why we could not define equality of container extensions generically:
-the contained types may not have (yet) an associated equivalence. We could have
+This definition highlights why we cannot define equality of the positions and
+the contained elements generically:
+the contained types may not have (yet) an associated equivalence. In other
+words, \mintinline{Coq}{LFix} may  We could have
 overcame this circularity by defining simultaneously LFix and its associated

### 2024-02-22 paper/paper.tex
Initial algebras

--- 2024-02-22 paper/paper.tex
@@
 \end{minted}
-
-There are \emph{two} initial algebras that are, as expected, unique up to
-isomorphism. We show one such initial algebra here, and the second in the
-following subsection, plus a proof that they are isomorphic:
-
+Given a container extension \mintinline{Coq}{F}, we define \emph{two}
+(isomorphic) initial algebras. We will show the need for the second initial
+algebra in the following section. Our fist initial algebra is defined as the
+straightforward least fixpoint of the container extension \mintinline{Coq}{F}:
 \begin{minted}{Coq}
@@
 \end{minted}
-This is why we could not define equality of container extensions generically:
-the contained types may not have (yet) an associated equivalence. We could have
-overcame this circularity by defining simultaneously LFix and its associated
-equivalence relation. But that would complicate definitions unnecessarily.
-\dcas{I know ... I need to rewrite/expand on this}
-
-We provide a custom induction schema for initial algebras:
-
+This definition highlights why we cannot define equality of the positions and
+the contained elements generically (recall Section~\ref{sec:TODO}): the
+contained types may not have (yet) an associated equivalence. In other words,
+we would need to provide an equivalence for \mintinline{Coq}{LFix}, but the
+only way to do this is to define such equivalence simultaneously to the
+definition for  \mintinline{Coq}{LFix}.  We could have overcame this
+circularity by defining  \mintinline{Coq}{LFix} independently of container
+extensions, and define later the necessary
+\mintinline{Coq}{in}/\mintinline{Coq}{out} functions to convert between
+\mintinline{Coq}{LFix} and \mintinline{Coq}{App}. However, this extra layer
+interferes with code extraction, in that it requires destructing/constructing
+subsequent \mintinline{Coq}{LFix} and  \mintinline{Coq}{App}, which Coq is not
+always able to inline.
+
+We provide a custom induction scheme for initial algebras:
 \begin{minted}{Coq}
@@
 \end{minted}
-
-Equality of initial algebras of container extensions needs to be defined
-recursively and extensionally:
+We define the equality of initial algebras of container extensions recursively,
+by traversing the structure and equating the shapes and positions. This
+definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
+Section~\ref{sec:TODO}.
 \begin{minted}{Coq}
@@
 \end{minted}
-
-The unique homomorphism from an initial algebra to other algebras, i.e. the catamorphism, is defined as a
-regular recursive function:
-\begin{minted}{Coq}
-Definition cata_f `{eA : equiv A} (g : Alg F A) : LFix -> A
+From such initial algebras, we define \emph{catamorphisms} by recursively
+traversing their structure. As usual, catamorphisms take an
+\mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
+function from \mintinline{Coq}{LFix F} to \mintinline{Coq}{A}.
+\begin{minted}{Coq}
+Definition cata_f `{equiv A} (g : Alg F A) : LFix -> A
 := fix f (x : LFix) :=

### 2024-02-22 paper/paper.tex
Background section.

--- 2024-02-22 paper/paper.tex
@@
 \usepackage{newunicodechar}
-
+\usepackage{microtype}
 % \addbibresource{references.bib}
@@
 \newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
-\newcommand{\dcas}[1]{\textcolor{green}{\textsc{David}: #1}}
+\newcommand{\dcas}[1]{\textcolor{ForestGreen}{\textsc{David}: #1}}
 \newcommand{\mvol}[1]{\textcolor{blue}{\textsc{Michael}: #1}}
@@
 \newcommand{\coq}[1]{\mintinline{coq}{#1}}
+
 \title{A Mechanised Library for Generic (Co)Programming}
@@
 \[
-  h \comp \InIso = a \comp F h \qquad \OutIso \comp k = F k \comp c
+  h  = a \comp F h \comp \InOp \qquad k = \OutOp \comp F k \comp c
 \]
@@
 $F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
-$F$-coalgebra), $\InIso$ is the isomoprhism on the least fixed-point of $F$,
-namely $\mu F \xrightarrow{\iso} F\mu F$, and $\OutIso$ is the isomorphism on
-the greatest fixed-point of $F$, namely $\nu F \xrightarrow{\iso} F\nu F$. When
+$F$-coalgebra), $\InOp$ is the isomoprhism on the least fixed-point of $F$,
+namely $\InOp : \mu F \to F\mu F$, and $\OutOp$ is the isomorphism on
+the greatest fixed-point of $F$, namely $F \nu F \to \nu F$. When
 the unique solutions to the above equations exist we denote them as $\cata{a}$
@@
 
-However, catamorphisms (dually anamorphisms) do not capture every recursion
-scheme we can think of.  As an example, consider the quicksort
-algorithm:\footnote[1]{This program is written in Coq notation to be consistent,
-but it is not a program that Coq can accept.}
+\subsection{Hylomorphisms}
+Catamorphisms (dually anamorphisms) do not capture every recursion
+scheme we can think of.  As an example, consider an example of a classic
+\emph{divide-and-conquer} program such as quicksort algorithm:\footnote[1]{This
+program is written in Coq notation to be consistent, but it is not a program
+that Coq can accept.}
 
 \mpav{I need this to be written in Coq notation}
-\begin{minted}{haskell}
-partition :: [a] -> ([a], a, [a])
-partition (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
-
-quicksort :: [a] -> [a]
-quicksort [] = []
-quicksort ls = let (xs, x, ys) = partition ls in
-                        (quicksort xs) ++ [x] ++ (quicksort ys)
-\end{minted}
-This program is formed of essentially two parts. The first one is the
-$\operator{partition}$ program destructuring the input list in three parts, the
-pivot and the other two sublists, and the second part recursing over the
-sublists and composing the results back together.
+\begin{minted}{coq}
+qs :: [a] -> [a]
+qs [] = []
+qs (x:xs) = qs (filter (x <=) xs) ++ [x] ++ (qs (filter (x >) xs))
+\end{minted}
+This program is formed of essentially two parts. The first one is what is
+commonly thought as the \emph{partition} function destructuring the input list
+in three parts, the pivot and the other two sublists, and the second part
+recursing over the sublists and composing the results back together.
 
@@
 the obvious functorial action then \coq{partition} can be viewed as a
-\coq{TreeF}-coalgebra on lists over the type $A$. While we can define an \coq{TreeF}-algebra on lists over $A$
-as well, that is a map \coq{alg} $:$ \coq{TreeF} $[A] \to [A]$ defined
-as \coq{alg (left, x, right) = left ++ [x] ++ right}.
-
-More specifically, in the context of recursive data types, such as in Haskell,
-where induction and coinduction is conflated, an hylomorphism can be implemented
-in terms of a catarmorphism (folding) and an anamorphism (unfolding)
-\[
-  \hylo{a}{c} = \cata{a} \comp \ana{c}
-\]
-They first generate a data structure using an anamorphism and then immediately
-fold or reduce it using a catamorphism.
-
-The hylomorphism pattern is particularly useful for defining recursive
-algorithms in a functional programming context, where recursion is a fundamental
-tool for solving problems. By separating the generation of data structures from
-their consumption, hylomorphisms enable clearer and more modular code, making it
-easier to reason about and maintain recursive algorithms.
-
-However, in a type theory inductive and coinductive data types are separate.
-
-In this section we will introduce the notion of hylomorphism. By way of examples
-we will also give high level intution as to way all recursion schemes can be
-defined in terms of a basic
-
-Say that we want to prove that the two functions below are well-defined
+\coq{TreeF}-coalgebra on lists over the type $A$
+\begin{minted}{coq}
+c : [A] -> TreeF [A]
+c (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
+\end{minted}
+
+While we can define an \coq{TreeF}-algebra on lists
+over $A$ as well, that is a map
+\begin{minted}{coq}
+a : TreeF [A] -> [A]
+a (left, x, right) = left ++ [x] ++ right
+\end{minted}
+
+It should be easy to see that the program \coq{qs} can now be defined as the
+unique program that runs the coalgebra $c$ recurses over the sublists produced
+and then composes the results together. The structure exemplified by this
+program is a typical divide-and-conquer algorithm which can be implemented using
+an \emph{hylomorphism}. An hylomorphism is the unique solution $x : C \to A$ to
+the hylo equation
+\begin{equation}
+  \label{eq:hylo}
+  x =  a \comp Fx \comp c
+\end{equation}
+for a $F$-algebra on $A$, an $F$-coalgebra on $C$ and a functor $F$.  If for
+every $F$-algebra there exists a unique solution to (\ref{eq:hylo}) then we say
+the $F$-coalgebra $c$ is \emph{recursive}. Dually if for every $F$-coalgebra $c$
+there exists a unique solution to (\ref{eq:hylo}) we say the $F$-algebra $a$ is
+\emph{corecursive}. We denote the unique solution to the hylo equation
+(\ref{eq:hylo}) as $\hylo{a}{c}$.
+
+Moreover, it shoud be easy to see that catamorphisms and anamorphism arise as a
+special case of hylomorphisms
+\begin{equation}
+  \label{eq:cata}
+  \cata{a} = \hylo{\InOp}{a} \qquad \ana{c} = \hylo{c}{\OutOp}
+\end{equation}
+since $\InOp$ is indeed a recursive coalgebra and $\OutOp$ is a corecursive
+algebra.
+
+\subsection{Implementing Recursion Schemes}
+Hylomorphisms can implement all structured recursion schemes~\cite{HinzeWG15,
+HinzeW16}.  Here we show the result by means of a very basic example: mutual
+recursion. We refer the reader to Hinze et al.~\cite{HinzeWG15} for the
+theoretical result and to Yang and Wu~\cite{YangW22} for more examples of
+recursion schemes.
+
+% More specifically, in the context of recursive data types, such as in Haskell,
+% where induction and coinduction is conflated, an hylomorphism can be implemented
+% in terms of a catarmorphism (folding) and an anamorphism (unfolding)
+% \[
+%   \hylo{a}{c} = \cata{a} \comp \ana{c}
+% \]
+% They first generate a data structure using an anamorphism and then immediately
+% fold or reduce it using a catamorphism.
+
+% The hylomorphism pattern is particularly useful for defining recursive
+% algorithms in a functional programming context, where recursion is a fundamental
+% tool for solving problems. By separating the generation of data structures from
+% their consumption, hylomorphisms enable clearer and more modular code, making it
+% easier to reason about and maintain recursive algorithms.
+
+% However, in a type theory inductive and coinductive data types are separate.
+
+% In this section we will introduce the notion of hylomorphism. By way of examples
+% we will also give high level intution as to way all recursion schemes can be
+% defined in terms of a basic
+
+Say that we want to prove that the two mutually recursive functions below are
+well-defined
 
@@
 \end{align*}
-This function is clearly a catarmorphism:
-\begin{align*}
-  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
-  & \operator{even-odd} = \cata{a} \operator{ where }\\
-  & \qquad \qquad a : 1 + (\Bool \times \Bool)\\
-  & \qquad \qquad a (\Nil) = (\top, \bot)\\
-  & \qquad \qquad a (\operator{x,y}) = (\neg x, \neg y)
-\end{align*}
-where $\cata{\cdot}$ takes a $(1+-)$ algebra on $\Bool \times \Bool$ and returns
-the associated recursive function.
-
-Now we recover the original two functions by setting
-$\operator{even}' \eqdef \pi_{1} \comp \operator{even-odd}$
-$\operator{odd}' \eqdef \pi_{2} \comp \operator{even-odd}$. The proof that these
-two pairs of functions are the same is an easy induction over the natural
-numbers.
-
-This is an instance of a more general phenonema whose argumet uses category
-theory.
-
-\subsection{Course-of-Value Recursion}
-The knapsack algorithm is defined as following in pseudo-Haskell code:
-\begin{align*}
-  & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
-  & \operator{knapsack}\;wvs\; c = \operator{maximum}_{0}
-    [v + \operator{knapsack}\;wvs\; (c - w) \mid (w,v) \leftarrow wvs, w \le c]
-\end{align*}
-%
-with dynamic programming knapsack can be rephrased like this
-\begin{align*}
-  & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
-  & \operator{knapsack}\;wvs\;c= table !! c \\
-  & \qquad table  = [ks\;i \mid i \leftarrow [0..c]]\\
-  & \qquad ks\; i = maximum_{0} [v + table!! (i - w) \mid (w,v) \leftarrow wvs, w \le i]
-\end{align*}
-
-The recursion scheme is justified by solving a recursive equation.  First we
-define an operation to create a memory table representing the list of capacities
-$[0..c]$. We first define a functor $D X = 1 + X$ representing the base functor
-for constructing the naturals $\Nat = \mu D$ the we define the generating
-function by induction on the naturals
-\begin{align*}
-  & \operator{fan} : \Nat \to D_{\infty}\Nat\\
-  & \operator{fan} (0)    = \pair{0}{\inl}\\
-  & \operator{fan} (n+1)  = \pair{n+1}{\inr (\operator{fan}(n))}
-\end{align*}
-The recursion scheme from a comonad is then defined as follows
-\begin{align*}
-  & \operator{rsfc} : \Nat \to A\\
-  & \operator{rsfc} = alg \comp DD_{\infty}(\operator{rsfc}) \comp D(\operator{fan}) \comp \InOp
-\end{align*}
-At his point we can given the definition of the algebra $alg$ which effectively implements knapsack.
-Firs we define the $\operator{lookup}$ function which returns (if present) the element indexed in the memory table
-\begin{align*}
-  & \operator{lookup} : \Nat \to D_{\infty} \R \to \R_{\bot}\\
-  & \operator{lookup} (0,   \pair{r}{\textunderscore})  = r\\
-  & \operator{lookup} (n+1, \pair{r}{\inl})         = \bot\\
-  & \operator{lookup} (n+1, \pair{r}{\inr(table)})  = \operator{lookup} (n, table)
-\end{align*}
-and now we can define the algebra over the reals
-\begin{align*}
-  & \operator{alg} : DD_{\infty} \R \to \R\\
-  & \operator{alg}\; (\inl)  = 0\\
-  & \operator{alg}\; (\inr(table))  = \operator{maximun}_{0} [ v + \operator{lookup}\; table\; (i - w) \mid (w,v) \leftarrow wvs, w \le i]
-\end{align*}
-
-It remains to show how an $\operator{rsfc}$ is implemented in terms of
-hylomorhisms.
-
-\paragraph{Implementation into hylomorphisms}
-In order to implement rsfcs as hylos we need a parametric distributive law
-\[
-  \lambda_{X} : GG_{\infty}X \to G_{\infty}GX
-\]
-This can be defined as
-$\lambda = G_{\infty}G(\head) \comp \anacofree{G(\tail)}$.  The distributive law
-is instrumental to be able to lift coalgebras in the following sense. Give a
-$G$-coalgebra on $X$, a lifting $H$ is a functor which lifts this coalgebra to a
-$G$-coalgebra on $HX$.  In particular, the lifting $G_{\lambda}$ takes
-$G_{\infty}$-coalgebras on $X$ and transforms them into $G_{\infty}$-coalgebras
-on $GX$
-\begin{align*}
-  & G_{\lambda}(c) : GA \to G_{\infty} GA\\
-  & G_{\lambda}(c) = \lambda \comp G (c)
-\end{align*}
-where $c : C \to GC$.
-
-Now, given an algebra $a : GG_{\infty} A \to A$ on the carrier $A$ we can
-convert it into an algebra for the carrier $G_{\infty} A$ by
-\begin{align*}
-  & \floor{a} : GG_{\infty} A \to G_{\infty} A\\
-  & \floor{a} = G_{\infty}(a) \comp \anacofree{G_{\lambda}(\delta)}
-\end{align*}
-where $\eta_{A} : A \to G_{\infty} A$ is the unit of the comonad creating an
-infinite constant list.
-
-At this point the hylo is simply defined by
-$\hylo{\InOp}{a} : \mu F \to G_{\infty}A$  and the rsfc is derived as follows
-\begin{align*}
-  & \operator{rsfc} : \mu G \to A \\
-  & \operator{rsfc} = \head \comp \hylo{\InOp}{a}
-\end{align*}
-
-\subsection{Catamorphisms and Anamorphism}
-A catamorphism is the unique program defined as
-\[
-  x = a \comp \fmap{x} \comp \InOp
-\]
-Can be implemented:
-\[
-  \cata{a} = \hylo{\InOp}{a}
-\]
-
-An anamorphism is the unique program satisfying the equation
-\[
-  x = \OutOp \comp \fmap{x} \comp{c}
-\]
-The anamorphism is indicated by $\ana{c}$. Recursion schemes given by a
-corecursive algebra such as $\OutIso$ are also called hylomorphism but they are
-indicated with the symbol for the anamorphism, $\cohylo{c}{\OutOp}$.
-
-\paragraph{Free Monads}
-Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
-solution to the following domain equation
-\[
-  D^{*}A = A + DD^{*}A
-\]
-where $\eta_{A} : A \to D^{*}A$ is the left injection and
-$\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
-$a : DA \to A$ there is a unique solution to the following recursive equation
-\[
-  x \comp \call_{A} = a \comp D(x)
-\]
-that is $\catafree{a} = \cata{[id_{A}; a]}$. With this operation we can extend
-the free algebra $\call$ to the multiplication $\mu : D^{*}D^{*}A \to D^{*}A$
-defined as $\mu = \catafree{\call}$.
-
-\paragraph{Cofree Comonads}
-Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
-greatest solution to the following domain equation
-\[
-  D_{\infty}A = A + DD_{\infty}A
-\]
-where $\head : D_{\infty}A \to A$ is the first projection and
-$\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.
-Given an algebra $c : A \to DA$ there is a unique solution to the following
-recursive equation
-\[
-  \tail_{A} \comp x = D(x) \comp c
-\]
-that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$. With the latter operation we
-can extend $\tail$ to the comultiplication for the comonad
-$\delta : D_{\infty}A \to D_{\infty} D_{\infty}A$ defined by
-$\delta = \anacofree{\tail}$.
-
-\subsection{Accumulators}
-For an algebra $a : D(A^{P})\times P \to A$:
-\begin{align*}
-  & g : \mu D \times P \to B\\
-  & g (d, p) = a (\fmap{(\curry g)} \comp \InOp(d), p)
-\end{align*}
-
-\paragraph{Hylo Implementation}
-Given an algebra $a : D(A^{P}) \to A^{P}$ the accumorphism corresponds to the unique
-solution to the following equation
-\begin{align*}
-  & f : \mu D \to A^{P}\\
-  & f = a \comp \fmap (f) \comp \InOp
-\end{align*}
-thus $f = \hylo{\InOp}{a}$.
-
-The complex recursion scheme is then
-$g = \uncurry\hylo{\InOp}{a} : \mu D \times P \to B$.
-
-\subsection{Mutu-Hylo}
-For an algebra $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and an algebra
-$a_{2} : D(A_{1} \times A_{2}) \to A_{2}$:
-\begin{align*}
-  & (g_{1}, g_{2}) : (\mu D, \mu D) \to (A_{1}, A_{2})\\
-  & (g_{1}, g_{2}) = (a_{1}, a_{2}) \comp (\fmap{\pair{g_{1}}{g_{2}}},\fmap{\pair{g_{1}}{g_{2}}}) \comp (\InOp, \InOp)
-\end{align*}
-
-\paragraph{Hylo Implementation}
-Given an algebra $a : D(A_{1} \times A_{2}) \to A_{1} \times A_{2}$
-\begin{align*}
-  & f : \mu D \to A_{1} \times A_{2}\\
-  & f = a \comp \fmap{f} \comp \InOp
-\end{align*}
-for which the solution is an hylo $f = \hylo{\InOp}{a}$.
-
-The mutumorphism can be implemented as
-$(g_{1}, g_{2}) = (\pi_{1} \comp f, \pi_{2} \comp f)$.
-
-\paragraph{Coq Implementation}
-Given the two algebras  $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and
-$a_{2} : D(A_{1} \times A_{2}) \to A_{2}$.
-We define the two recursive functions  $f_{1}$ and $f_{2}$ as
-
-\[
-  f_{1} = \pi_{1} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}} \qquad f_{2} = \pi_{2} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}}
-\]
-% A recursion scheme from a comonad $D_{\infty}$ for a data functor $F$
-
-% For example, for the functor $(1 + -)$ an algebra over this functor is a
-% function $1 + X \to X$ is a pair of functions $\Nil : 1 \to X$ and
-% $\Succ : X \to X$. The cofree comonad over $D$ is defined as follows
+which is clearly a catamorphism $\cata{a}$ with $a (\Nil) = (\top, \bot)$ and
+$a(\operator{x,y}) = (\neg x, \neg y)$ and hence an hylomorphism as defined in (\ref{eq:cata}).  Now we are left to show they are
+equivalent. Well, it is easy to do so since
+$\operator{even} = \pi_{1} \comp \operator{even-odd}$
+$\operator{odd} = \pi_{2} \comp \operator{even-odd}$. The proof that these two
+pairs of functions are the same is an easy induction over the natural numbers.
+
+This is an instance of a more general phenonema whose argument uses the concept
+of an adjunction~\cite{MacLane71}.  For this particular example, the argument is
+as follows.  There is an isomorphism between pairs of maps
+$(A \to B_{1}, A \to B_{2})$ and maps $(A \to B_{1} \times B_{2})$.  This is
+given by $\Phi (f,g) = \pair{f}{g}$ and
+$\Psi(f) = (\pi_{1} \comp f, \pi_{2} \comp f)$ where the projections $\pi$ are
+part of the counit of the adjunction between the diagonal functor
+$\Delta X = (X,X)$ and the product $\times$.  Once this correspondence is
+established we can implement the map $A \to B_{1} \times B_{2}$ as an
+hylomorphism and derive the original and more complex recursion scheme.
+
+% In general an adjunction is a natural correspondence of maps
 % \[
-%   D_{\infty} A \iso A \times (1 + D_{\infty})
+%   LA \to B \iff A \to RB
 % \]
-% which is the the type of non-empty possibly infinite lists with elements in $A$
-
-% First we have to assume a distributive law
-% $\lambda : DD_{\infty} \to D_{\infty}D$.  For an algebra
-% $a : D(D_{\infty} A) \to A$,
+% where $L$ and $R$ are called adjoints.
+
+However, formalising adjunctions in Coq would require to formalise a
+non-negligible amount of category theory which goes beyond the scope of the
+present work. In this paper, we will instead derive the recursion schemes as needed and
+implement them as hylos.
+% \subsubsection{Course-of-Value Recursion}
+% The knapsack algorithm is defined as following in pseudo-Haskell code:
 % \begin{align*}
-%   & g : \mu D \to A\\
-%   & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
+%   & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
+%   & \operator{knapsack}\;wvs\; c = \operator{maximum}_{0}
+%     [v + \operator{knapsack}\;wvs\; (c - w) \mid (w,v) \leftarrow wvs, w \le c]
 % \end{align*}
-% where
-% $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX$
-% is a $D$-algebra on $GX$. The catamorphism of this algebra has the following
-% type
+% %
+% with dynamic programming knapsack can be rephrased like this
+% \begin{align*}
+%   & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
+%   & \operator{knapsack}\;wvs\;c= table !! c \\
+%   & \qquad table  = [ks\;i \mid i \leftarrow [0..c]]\\
+%   & \qquad ks\; i = maximum_{0} [v + table!! (i - w) \mid (w,v) \leftarrow wvs, w \le i]
+% \end{align*}
+
+% The recursion scheme is justified by solving a recursive equation.  First we
+% define an operation to create a memory table representing the list of capacities
+% $[0..c]$. We first define a functor $D X = 1 + X$ representing the base functor
+% for constructing the naturals $\Nat = \mu D$ the we define the generating
+% function by induction on the naturals
+% \begin{align*}
+%   & \operator{fan} : \Nat \to D_{\infty}\Nat\\
+%   & \operator{fan} (0)    = \pair{0}{\inl}\\
+%   & \operator{fan} (n+1)  = \pair{n+1}{\inr (\operator{fan}(n))}
+% \end{align*}
+% The recursion scheme from a comonad is then defined as follows
+% \begin{align*}
+%   & \operator{rsfc} : \Nat \to A\\
+%   & \operator{rsfc} = alg \comp DD_{\infty}(\operator{rsfc}) \comp D(\operator{fan}) \comp \InOp
+% \end{align*}
+% At his point we can given the definition of the algebra $alg$ which effectively implements knapsack.
+% Firs we define the $\operator{lookup}$ function which returns (if present) the element indexed in the memory table
+% \begin{align*}
+%   & \operator{lookup} : \Nat \to D_{\infty} \R \to \R_{\bot}\\
+%   & \operator{lookup} (0,   \pair{r}{\textunderscore})  = r\\
+%   & \operator{lookup} (n+1, \pair{r}{\inl})         = \bot\\
+%   & \operator{lookup} (n+1, \pair{r}{\inr(table)})  = \operator{lookup} (n, table)
+% \end{align*}
+% and now we can define the algebra over the reals
+% \begin{align*}
+%   & \operator{alg} : DD_{\infty} \R \to \R\\
+%   & \operator{alg}\; (\inl)  = 0\\
+%   & \operator{alg}\; (\inr(table))  = \operator{maximun}_{0} [ v + \operator{lookup}\; table\; (i - w) \mid (w,v) \leftarrow wvs, w \le i]
+% \end{align*}
+
+% It remains to show how an $\operator{rsfc}$ is implemented in terms of
+% hylomorhisms.
+
+% \paragraph{Implementation into hylomorphisms}
+% In order to implement rsfcs as hylos we need a parametric distributive law
 % \[
-%   \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
+%   \lambda_{X} : GG_{\infty}X \to G_{\infty}GX
 % \]
-% which is the type of a $G$-algebra used to initialise the memory table.
-
-% Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
-% what is the unique solution of the following equation
+% This can be defined as
+% $\lambda = G_{\infty}G(\head) \comp \anacofree{G(\tail)}$.  The distributive law
+% is instrumental to be able to lift coalgebras in the following sense. Give a
+% $G$-coalgebra on $X$, a lifting $H$ is a functor which lifts this coalgebra to a
+% $G$-coalgebra on $HX$.  In particular, the lifting $G_{\lambda}$ takes
+% $G_{\infty}$-coalgebras on $X$ and transforms them into $G_{\infty}$-coalgebras
+% on $GX$
 % \begin{align*}
-%   & f : \mu F \to G_{\infty} A\\
+%   & G_{\lambda}(c) : GA \to G_{\infty} GA\\
+%   & G_{\lambda}(c) = \lambda \comp G (c)
+% \end{align*}
+% where $c : C \to GC$.
+
+% Now, given an algebra $a : GG_{\infty} A \to A$ on the carrier $A$ we can
+% convert it into an algebra for the carrier $G_{\infty} A$ by
+% \begin{align*}
+%   & \floor{a} : GG_{\infty} A \to G_{\infty} A\\
+%   & \floor{a} = G_{\infty}(a) \comp \anacofree{G_{\lambda}(\delta)}
+% \end{align*}
+% where $\eta_{A} : A \to G_{\infty} A$ is the unit of the comonad creating an
+% infinite constant list.
+
+% At this point the hylo is simply defined by
+% $\hylo{\InOp}{a} : \mu F \to G_{\infty}A$  and the rsfc is derived as follows
+% \begin{align*}
+%   & \operator{rsfc} : \mu G \to A \\
+%   & \operator{rsfc} = \head \comp \hylo{\InOp}{a}
+% \end{align*}
+
+% % \subsection{Catamorphisms and Anamorphism}
+% % A catamorphism is the unique program defined as
+% % \[
+% %   x = a \comp \fmap{x} \comp \InOp
+% % \]
+% % Can be implemented:
+% % \[
+% %   \cata{a} = \hylo{\InOp}{a}
+% % \]
+
+% % An anamorphism is the unique program satisfying the equation
+% % \[
+% %   x = \OutOp \comp \fmap{x} \comp{c}
+% % \]
+% % The anamorphism is indicated by $\ana{c}$. Recursion schemes given by a
+% % corecursive algebra such as $\OutIso$ are also called hylomorphism but they are
+% % indicated with the symbol for the anamorphism, $\cohylo{c}{\OutOp}$.
+
+% \paragraph{Free Monads}
+% Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
+% solution to the following domain equation
+% \[
+%   D^{*}A = A + DD^{*}A
+% \]
+% where $\eta_{A} : A \to D^{*}A$ is the left injection and
+% $\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
+% $a : DA \to A$ there is a unique solution to the following recursive equation
+% \[
+%   x \comp \call_{A} = a \comp D(x)
+% \]
+% that is $\catafree{a} = \cata{[id_{A}; a]}$. With this operation we can extend
+% the free algebra $\call$ to the multiplication $\mu : D^{*}D^{*}A \to D^{*}A$
+% defined as $\mu = \catafree{\call}$.
+
+% \paragraph{Cofree Comonads}
+% Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
+% greatest solution to the following domain equation
+% \[
+%   D_{\infty}A = A + DD_{\infty}A
+% \]
+% where $\head : D_{\infty}A \to A$ is the first projection and
+% $\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.
+% Given an algebra $c : A \to DA$ there is a unique solution to the following
+% recursive equation
+% \[
+%   \tail_{A} \comp x = D(x) \comp c
+% \]
+% that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$. With the latter operation we
+% can extend $\tail$ to the comultiplication for the comonad
+% $\delta : D_{\infty}A \to D_{\infty} D_{\infty}A$ defined by
+% $\delta = \anacofree{\tail}$.
+
+% \subsection{Accumulators}
+% For an algebra $a : D(A^{P})\times P \to A$:
+% \begin{align*}
+%   & g : \mu D \times P \to B\\
+%   & g (d, p) = a (\fmap{(\curry g)} \comp \InOp(d), p)
+% \end{align*}
+
+% \paragraph{Hylo Implementation}
+% Given an algebra $a : D(A^{P}) \to A^{P}$ the accumorphism corresponds to the unique
+% solution to the following equation
+% \begin{align*}
+%   & f : \mu D \to A^{P}\\
+%   & f = a \comp \fmap (f) \comp \InOp
+% \end{align*}
+% thus $f = \hylo{\InOp}{a}$.
+
+% The complex recursion scheme is then
+% $g = \uncurry\hylo{\InOp}{a} : \mu D \times P \to B$.
+
+% \subsection{Mutu-Hylo}
+% For an algebra $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and an algebra
+% $a_{2} : D(A_{1} \times A_{2}) \to A_{2}$:
+% \begin{align*}
+%   & (g_{1}, g_{2}) : (\mu D, \mu D) \to (A_{1}, A_{2})\\
+%   & (g_{1}, g_{2}) = (a_{1}, a_{2}) \comp (\fmap{\pair{g_{1}}{g_{2}}},\fmap{\pair{g_{1}}{g_{2}}}) \comp (\InOp, \InOp)
+% \end{align*}
+
+% \paragraph{Hylo Implementation}
+% Given an algebra $a : D(A_{1} \times A_{2}) \to A_{1} \times A_{2}$
+% \begin{align*}
+%   & f : \mu D \to A_{1} \times A_{2}\\
 %   & f = a \comp \fmap{f} \comp \InOp
 % \end{align*}
-% The solution to this equation can be found again using an hylo
-% $f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.  The course-of-value recursion
-% can be recovered from this by setting $g = \head \comp f$
-
-% \subsubsection{Histomorphism}
-
-% The definition of $\lambda$ is:
-% \begin{align*}
-%   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
-%   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
-% \end{align*}
-% For example, for $GX = 1 + X$ we have to define
-% \begin{align*}
-%   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
-%   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
-%   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
-% \end{align*}
-
+% for which the solution is an hylo $f = \hylo{\InOp}{a}$.
+
+% The mutumorphism can be implemented as
+% $(g_{1}, g_{2}) = (\pi_{1} \comp f, \pi_{2} \comp f)$.
+
+% \paragraph{Coq Implementation}
+% Given the two algebras  $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and
+% $a_{2} : D(A_{1} \times A_{2}) \to A_{2}$.
+% We define the two recursive functions  $f_{1}$ and $f_{2}$ as
+
+% \[
+%   f_{1} = \pi_{1} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}} \qquad f_{2} = \pi_{2} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}}
+% \]
+% % A recursion scheme from a comonad $D_{\infty}$ for a data functor $F$
+
+% % For example, for the functor $(1 + -)$ an algebra over this functor is a
+% % function $1 + X \to X$ is a pair of functions $\Nil : 1 \to X$ and
+% % $\Succ : X \to X$. The cofree comonad over $D$ is defined as follows
+% % \[
+% %   D_{\infty} A \iso A \times (1 + D_{\infty})
+% % \]
+% % which is the the type of non-empty possibly infinite lists with elements in $A$
+
+% % First we have to assume a distributive law
+% % $\lambda : DD_{\infty} \to D_{\infty}D$.  For an algebra
+% % $a : D(D_{\infty} A) \to A$,
 % % \begin{align*}
-% %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
-% %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
-% %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
-% %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
+% %   & g : \mu D \to A\\
+% %   & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
 % % \end{align*}
-
-% % Defining the algebra
+% % where
+% % $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX$
+% % is a $D$-algebra on $GX$. The catamorphism of this algebra has the following
+% % type
+% % \[
+% %   \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
+% % \]
+% % which is the type of a $G$-algebra used to initialise the memory table.
+
+% % Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
+% % what is the unique solution of the following equation
 % % \begin{align*}
-% %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
-% %   & \operator{alg}\; Zero           = 0\\
-% %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
+% %   & f : \mu F \to G_{\infty} A\\
+% %   & f = a \comp \fmap{f} \comp \InOp
 % % \end{align*}
+% % The solution to this equation can be found again using an hylo
+% % $f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.  The course-of-value recursion
+% % can be recovered from this by setting $g = \head \comp f$
+
+% % \subsubsection{Histomorphism}
+
+% % The definition of $\lambda$ is:
+% % \begin{align*}
+% %   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
+% %   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
+% % \end{align*}
+% % For example, for $GX = 1 + X$ we have to define
+% % \begin{align*}
+% %   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
+% %   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
+% %   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
+% % \end{align*}
+
+% % % \begin{align*}
+% % %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
+% % %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
+% % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
+% % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
+% % % \end{align*}
+
+% % % Defining the algebra
+% % % \begin{align*}
+% % %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
+% % %   & \operator{alg}\; Zero           = 0\\
+% % %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
+% % % \end{align*}
 
@@
 \end{minted}
-
-\section{Examples}
-
-\paragraph{Quicksort \& Mergesort}
-
-\dcas{What else?}
 
@@
 \mvol{We have a story for this now: obj.magic harms interaction with other ocaml code, the resulting ocaml code is not natural to read, and in general more complex/strange extracted code is likely to be less efficient and harder to optimise.}
+        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
 \end{itemize}

### 2024-02-22 paper/paper.tex
Updates to paper

@@
 definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
-Section~\ref{sec:TODO}.
+Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
+positions, the contents must be related by \mintinline{Coq}{LFixR}.
 \begin{minted}{Coq}
@@
 \end{minted}
-
-It is easy to show that this function is a proper morphism, and that it is
-unique. We can use the fact that catamorphisms are proper morphisms to encode
-them as a higher-order morphism.
-\begin{minted}{Coq}
-Lemma cata_arr1 `{eA : equiv A} (g : Alg F A)
-  : forall x y, x =e y -> cata_f g x =e cata_f g y.
+It is easy to show that this function is a proper morphism, and we use this
+fact to encode them as a higher-order morphism, with the following type:
+\begin{minted}{Coq}
 Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
-
+\end{minted}
+Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
+they are the unique solution to the equation \mintinline{Coq}{f =e g \o fmap f
+\o l_out}). The proof is by induction on the input to \mintinline{Coq}{f} and
+\mintinline{Coq}{cata g}.
+\begin{minted}{Coq}
 Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
@@
 \end{minted}
-
 Their equivalence relation, and anamorphisms, and universal properties are
@@
 \end{minted}
-
-With terminal coalgebras, we can encode another form of initial algebras. In
-particular, the initial algebras based on terminal coalgebras are the carrier
-types of terminal coalgebras, plus a proof that they are finite:
-
+This definition, however, prevents us from composing catamorphisms and
+anamorphisms. To address this issue, we consider another alternative definition
+for initial $F$-algebras. In particular, we show that the terminal coalgebras
+for $F$, coupled with a proof that they are finite are also initial
+$F$-algebras:
 \begin{minted}{Coq}
@@
 \end{minted}
-
 Using this predicate, we can define the notion of \emph{recursive coalgebras}
-as those coalgebras that build finite trees. Recursive coalgebras are
-implemented in Coq recursively on the proof that they produce finite trees.
-Such proofs will be erased when extracting code to OCaml. This is standard
-practice when building general recursive functions in Coq\footnote{\dcas{This
-technique had a name in the community, but I cannot remember. Something
-predicates}}.
+as those that build finite trees. Recursive coalgebras are implemented in Coq
+recursively on the proof that they produce finite trees.  Such proofs will be
+erased when extracting code to OCaml. This is standard practice when building
+general recursive functions in Coq\footnote{\dcas{This technique had a name in
+the community, but I cannot remember. Something predicates}}.
 \begin{minted}{Coq}

### 2024-02-23 paper/paper.tex
minor
Fixed problem in paper
Minimal, just started working on Section 3 and correct David's obnoxious uncompilable code.

@@
 recursion schemes from comonads and their respective coinductive dual.
+\mpav{Moved this from Section 3}
+goals are: (1) the
+mechanisation of hylomorphisms and their laws should be axiom-free; and (2) Coq
+hylomorphisms should be extracted to recursive OCaml functions that use no
+instances of \mintinline{OCaml}{Obj.magic}.  Although the use of
+\mintinline{OCaml}{Obj.magic} is safe in the code extracted by Coq, it has been
+proven that, for higher-order programs, simple interoperations can lead to
+incorrect behaviour or even segfaults~\cite{forster:hal-04329663}.  Therefore,
+the avoidance of axioms and  \mintinline{OCaml}{Obj.magic} means that the
+extracted code will have stronger safety guarantees, and will allow the safe
+interoperation with other OCaml code.
 
@@
 which is clearly a catamorphism $\cata{a}$ with $a (\Nil) = (\top, \bot)$ and
-$a(\operator{x,y}) = (\neg x, \neg y)$ and hence an hylomorphism as defined in (\ref{eq:cata}).  Now we are left to show they are
-equivalent. Well, it is easy to do so since
-$\operator{even} = \pi_{1} \comp \operator{even-odd}$
+$a(\operator{x,y}) = (\neg x, \neg y)$ and hence an hylomorphism as defined in
+(\ref{eq:cata}).  Now we are left to show they are equivalent. Well, it is easy
+to do so since $\operator{even} = \pi_{1} \comp \operator{even-odd}$
 $\operator{odd} = \pi_{2} \comp \operator{even-odd}$. The proof that these two
@@
 non-negligible amount of category theory which goes beyond the scope of the
-present work. In this paper, we will instead derive the recursion schemes as needed and
-implement them as hylos.
+present work. In this paper, we will instead derive the recursion schemes as
+needed and implement them as hylos.
 % \subsubsection{Course-of-Value Recursion}
@@
 \section{Mechanising Extractable Container Functors}
-
 \dcas{Subsections are tentative topics to remember what I need to fill in here,
@@
 
-Our mechanisation works in the category of Coq types, and our goals are: (1)
-the mechanisation of hylomorphisms and their laws should be axiom-free; and (2)
-Coq hylomorphisms should be extracted to recursive OCaml functions that use no
-instances of \mintinline{OCaml}{Obj.magic}.  Although the use of
-\mintinline{OCaml}{Obj.magic} is safe in the code extracted by Coq, it has been
-proven that, for higher-order programs, simple interoperations can lead to
-incorrect behaviour or even segfaults~\cite{forster:hal-04329663}.  Therefore,
-the avoidance of axioms and  \mintinline{OCaml}{Obj.magic} means that the
-extracted code will have stronger safety guarantees, and will allow the safe
-interoperation with other OCaml code.
+In order to abstract recursion patters we need to be able to talk about functors
+which have the suitable fixed-point properties. Instances of these functors are
+the polynomial ones which are in one-to-one correspondence with the so called
+\emph{containers}~\cite{AbbottAG05}.
+In this section we introduce containers
 
 \subsection{Functors and Containers}
-
-In Coq, to construct least/greatest fixpoints of functors, these need to build
+\label{sec:containers}
+In order to To construct fixpoints of functors, these need to build
 \emph{strictly positive types}. A common approach to construct such strictly
@@
 However, reasoning about container and container functor equality requires to
-consider: (a) functional extensionality; (b) heterogeneous equality. This is
-problematic, due to our goals to remain axiom-free. Furthermore, standard
-representations of containers in Coq will lead to OCaml code that contains
-\mintinline{OCaml}{Obj.magic}. We will explain why, and our mechanisation of
-containers in the remainder of this section.
+consider both functional extensionality and heterogeneous equality. This is
+problematic since, as stated in Section~\ref{sec:intro}, our goals to remain
+axiom-free.  Furthermore, standard representations of containers in Coq will
+lead to OCaml code that contains \mintinline{OCaml}{Obj.magic}. We will explain
+why, and our mechanisation of containers in the remainder of this section.
 
@@
 \end{minted}
-An \emph{extension} of this container is a functor,
-whose action on objects is given by:
+An \emph{extension} of this container is a functor, whose action on objects is
+given by:
 \begin{minted}{Coq}
@@
 This approach solves the the problem of the unsafe casts, since the code
-extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml
-singleton type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.
+extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml singleton
+type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.
 
@@
 \subsubsection{Polynomial Functors as Container Extensions}
-
 Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
@@
 \subsubsection{Equality of Container Extensions}
-
 The equality of container extensions needs to be extensional, since the
@@
 \subsection{Types and Morphisms}
-
+\label{sec:types-morphisms}
 We address the different forms of equality by coupling every type with an
@@
 \subsubsection{Polynomial Types}
-
 We define a number of equivalences for polynomial types. These equivalences are
@@
 Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
-they are the unique solution to the equation \mintinline{Coq}{f =e g \o fmap f
-\o l_out}). The proof is by induction on the input to \mintinline{Coq}{f} and
+they are the unique solution to the equation \mintinline{Coq}{f =e g \o fmap f \o l_out}).
+The proof is by induction on the input to \mintinline{Coq}{f} and
 \mintinline{Coq}{cata g}.

### 2024-02-23 paper/paper.tex
Writing for section 5

@@
 
-\subsection{Sized Types}
+\subsection{Termination of Nonstructural Recursion}
+
+Termination and productivity are non-trivial properties of programs, and various
+methods have been proposed for checking that these properties hold. A common
+approach is guardedness checking~\cite{10.5555/646535.695850}, which is a
+\emph{syntactic} check that definitions avoid the introduction of
+non-normalizable terms. However, some desirable definitions will not be accepted
+by such checks.
+
+In particular, the problem of \emph{nonstructural recursion} is
+well-studied~\cite{BOVE_KRAUSS_SOZEAU_2016}.
+% A distinction between standard and
+% nonstandard recursion lies in the fact that standard recursion adheres strictly
+% to recursing on pattern variables, avoiding reliance on results from other
+% function calls. Nonstandard recursion may deviate from this convention, allowing
+% for more flexibility.
+Certain functions that are not structurally recursive can be reformulated using
+a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
+Take, for instance, division by iterated subtraction, which is inherently
+non-structurally recursive since it involves recursion based on the result of a
+subtraction. There is a nonstandard implementation of divivion found on Coq's
+standard library, which involves a four-argument function that effectively
+combines subtraction and division. Similarly, the mergesort in Coq's standard
+library uses an ``explicit stack of pending merges'' in order to avoid issues
+with nonstructural definitions. There is a major downside, however; as noted by
+Abreu et al., the result is ``barely recognizable as a form of
+mergesort''~\cite{AbreuDHJMS23}.
+
+\subsubsection{Sized Types}
 
 Another approach to certifying termination of nonstructurally recursive
-functions is the use of \emph{sized types}. Sized types were introduced in
-\cite{HughesPS96} as a way to track/verify various properties of recursive
-programs, including productivity and termination. The core idea of sized types
-is that types express bounds on the sizes of recursive data structures. This
-approach has been used to express nonstructurally recursive algorithms in Agda:
-\cite{CopelloTB14} used sized types in a straightforward formulation of
-mergesort. Such an approach has also been used in MiniAgda~\cite{abs-1202-3496}.
+functions is the use of \emph{sized types}. Sized types were introduced by
+Hughes et al.\ as a way to track/verify various properties of recursive
+programs, including productivity and termination~\cite{HughesPS96}. The core
+idea of sized types is that types express bounds on the sizes of recursive data
+structures. With this approach, algorithms are implemented in the standard
+functional way, following the divide-and-conquer pattern of splitting, recurring,
+and merging, with the addition of the data types being indexed by a static
+approximation of the relative \emph{size}. An advantage of this approach is
+that it allows for code to be written in a natural way.
+
+This approach has been used to express nonstructurally recursive
+algorithms in Agda: Copello et al.\ used sized types in a straightforward
+formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
+in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
+Coq as well~\cite{CHAN_2023}.
+
+\subsubsection{Divide and Conquer Recursion}
+
+Abreu et al.\ propose\ldots~\cite{AbreuDHJMS23}. Their approach has similarities to what is presented here.
+
+They differ in requiring the assumption of the functional extensionality axiom.
+Additionally, they are concerned with the erasure of termination proofs but have
+little discussion of what the resulting extracted code \emph{looks like} (that
+is, whether the extracted OCaml code resembles the natural formulation of the
+recursive function).
+\mvol{We should find out for sure if they use \texttt{obj.magic}.}
 

### 2024-02-23 paper/paper.tex
Recursive coalgebras

@@
 From such initial algebras, we define \emph{catamorphisms} by recursively
-traversing their structure. As usual, catamorphisms take an
+traversing their structure. As usual, a catamorphism takes an
 \mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
@@
 \end{minted}
-Using this predicate, we can define the notion of \emph{recursive coalgebras}
-as those that build finite trees. Recursive coalgebras are implemented in Coq
-recursively on the proof that they produce finite trees.  Such proofs will be
-erased when extracting code to OCaml. This is standard practice when building
-general recursive functions in Coq\footnote{\dcas{This technique had a name in
-the community, but I cannot remember. Something predicates}}.
-\begin{minted}{Coq}
-Notation rana_f__ h :=
-  (fix f x H :=
-    let hx := h x in
-    LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e)))).
-
+Catamorphisms from this initial algebra are denoted by \mintinline{Coq}{ccata},
+and they can be used to define the isomorphism between 
+\mintinline{Coq}{{x : GFix F | FinF x}} and
+\mintinline{Coq}{LFix F}. In particular: 
+\begin{minted}{Coq}
+Lemma cata_ccata `{equiv A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
+Lemma ccata_cata `{equiv A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
+\end{minted}
+Using this new initial algebra, we can compose catamorphisms and anamorphisms,
+\emph{as long as the anamorphisms produce finite trees}. However, if
+anamorphisms produce finite trees, then we can also encode in Coq the notion of
+\emph{recursive anamorphisms} and \emph{recursive coalgebras}~\cite{TODO}.
+Recursive coalgebras are coalgebras whose anamorphism produces finite trees.
+The following predicate captures this fact. Note how all the predicate states
+is that \mintinline{Coq}{h} can only be applied finitely many times:
+\begin{minted}{Coq}
+Inductive RecF (h : Coalg F A) : A -> Prop :=
+| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
+\end{minted}
+We show how this predicate is equivalent to the anamorphism of such coalgebras
+producing finite trees:
+\begin{minted}{Coq}
+Lemma ana_rec_term `{equiv A} (h : Coalg F A)
+  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
+\end{minted}
+Recursive anamorphisms are implemented in Coq recursively on the proof that
+they can only be applied finitely many times (\mintinline{Coq}{RecF}).  Note
+that such proofs will be erased when extracting code to OCaml. This is standard
+practice when building general recursive functions in Coq\footnote{\dcas{This
+technique had a name in the community, but I cannot remember. Something
+predicates}}. The core definitions are the following:
+\begin{minted}{Coq}
+Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
+
+Definition rana_f__ `{equiv A} (c : Coalg F A)
+  : forall x : A, RecF c x -> LFix F
+  := fix f (x : A) (H : RecF c x) :=
+       let hx := c x in
+       LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
+\end{minted}
+Here, \mintinline{Coq}{RecF_inv} provides the structurally smaller element for
+the Coq recursive definition. We then couple coalgebras with a proof that they
+are recursive, and add a coercion from recursive coalgebras to coalgebras.
+\begin{minted}{Coq}
+Definition RCoalg F A = {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
+\end{minted}
+Similarly to the rest of the definitions, we prove that recursive anamorphisms
+are respectful, and use this fact to build them as higher order proper
+morphisms, and prove its universal law, i.e.\ that recursive anamorphisms are
+the unique solution to the equation 
+\mintinline{Coq}{f =e l_in \o fmap f \o h},
+where \mintinline{Coq}{h} is a recursive coalgebra.
+\begin{minted}{Coq}
 Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...

### 2024-02-23 paper/paper.tex
more writing for section 5

@@
 
+\subsection{Program Calculation}
+
+Within the domain of program optimization, program calculation serves as a
+well-established programming technique aimed at deriving efficient programs from
+their naive counterparts through systematic program transformation. This area
+has been extensively explored over the years. Tesson et al.\ demonstrated the
+efficacy of leveraging Coq to establish an approach for implementing a robust
+system dedicated to verifying the correctness of program transformations for
+functions that manipulate lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and
+Emoto went further and\ldots~\cite{10.1007/978-3-030-34175-6_11}
+
 \subsection{Termination of Nonstructural Recursion}
@@
 approximation of the relative \emph{size}. An advantage of this approach is
-that it allows for code to be written in a natural way.
+that it allows for code to be written in a natural way. There are costs, however.
+Programs must be written in a way that accounts for the handling of size indexes,
+and support for the size types must be added to the language.
 
@@
 in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
-Coq as well~\cite{CHAN_2023}.
+Coq as well, though it has not yet been adopted~\cite{CHAN_2023}.
 

### 2024-02-23 paper/paper.tex
Up to recursive hylos

@@
 Record App `(C : Cont Shape Pos) (X : Set) :=
-  { shape : Shape; contents : PosOf C shape -> X }.
+  MkApp { shape : Shape; contents : PosOf C shape -> X }.
 \end{minted}
@@
       let (sx, kx) := ax in
-      g (MkCont sx (fun e => f (kx e)))
+      g (MkApp sx (fun e => f (kx e)))
   end.
@@
        let hx := c x in
-       LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
+       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
 \end{minted}
@@
 
-\subsection{Recursive Hylos}
-
-Recursive hylomorphisms are built from an algebra and a recursive coalgebra:
+\subsection{Recursive Hylomorphisms}
+
+Given an $F$-algebra and $F$-coalgebra, a solution to the equation $f = a \circ
+F\; f \circ c$ is known as a hylomorphism. Unfortunately, this definition
+cannot be accepted by Coq, since there is no structurally smaller argument to
+do recursion, and there is no corecursive definition that is guaranteed to be
+guarded, due to the use of an arbitrary algebra $a$. Furthermore, solutions to
+the hylomorphism equation are \emph{not} guaranteed to be unique, or to even
+exist. To address these issues, we mechanise \textbf{recursive hylomorphisms}
+which, as we will see, are guaranteed to have a unique solution.
+
+Hylomorphisms \emph{are} the unique solution to $f = a \circ F\; f \circ c$, if
+$c$ is a recursive coalgebra. We show this by first providing the definition
+for hylomorphisms in Coq, and then proving that any function satisfying this
+equation is a hylomorphism. We call these hylomorphisms with a recursive
+coalgebra, \textbf{recursive hylomorphisms}.
 \begin{minted}{Coq}
 Definition hylo_def (a : Alg F B) (c : Coalg F A) 
-: forall (x : A), RecF c x -> B
-:= fix f x H :=
-  match c x as h0
-        return
-        (forall e : Pos (shape h0), RecF c (cont h0 e)) ->
-        B
+  : forall (x : A), RecF c x -> B := fix f x H 
+  := match c x as h0 return
+       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
      with
-  | MkCont s_x c_x => fun H => a (MkCont s_x (fun e => f (c_x e) (H e)))
+     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
      end (RecF_inv H).
 \end{minted}
-
+The definition of recursive hylomorphisms is recursive on the proof that the
+anamorphism can only be applied finitely many times on any input. We again use
+\mintinline{Coq}{RecF_inv} to obtain the structurally smaller proof to use in
+the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
+we prove that \mintinline{Coq}{hylo_def} is respectful, and use this proof to
+built the corresponding higher-order proper morphism:
+\begin{minted}{Coq}
+Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
+\end{minted}
+Finally, we show that recursive hylomorphisms are the unique solution to
+the hylomorphism equation.
 \begin{minted}{Coq}
@@
 \end{minted}
-
-Recursive hylo universal property allows us to prove common optimisations:
-
-\begin{minted}{Coq}
-    Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
-    (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
+This result can be used to prove hylomorphism laws in a way that is analogous
+to pen-and-paper program calculation techniques. We show the rule for 
+hylomorphism fusion as an example:
+\begin{minted}{Coq}
+Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) 
+  (g2 : Alg F C) (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
   : f2 \o hylo g1 h1 =e hylo g2 h1.
 Proof.
+  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
   rewrite hylo_univ.
+  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
   rewrite fmap_comp.
+  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
   rewrite compA.
+  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
   rewrite <- E2.
+  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
   rewrite <- compA.
+  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
   rewrite <- compA.
+  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
   rewrite (compA g1).
+  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
   rewrite <- hylo_unr.
+  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
   reflexivity.
 Qed.
-
+\end{minted}
+Using the hylo fusion law, we can prove the well-known \emph{deforestation
+optimisation}, where two consecutive recursive computations, one that builds a
+data structure, and another one that consumes it, can be fused together into a
+single recursive definition. This, in turn, allows us to prove that a recursive
+hylomorphism is the composition of a catamorphism and a recursive anamorphism.
+\begin{minted}{Coq}
 Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
@@
 : cata g \o rana h =e hylo g h.
+\end{minted}
+This result will allow us to \emph{calculate} programs by providing small, easy
+to verify implementations, and combining them into efficient recursive
+functions by \emph{using Coq's proofs of hylomorphism laws as optimisation
+passes}.
+
+\section{Extraction}
+
+\dcas{Talk About this}
+\begin{minted}{OCaml}
+  let rec msort x = match x with
+| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
+     let (l0, r) = splitL x [] [] in
+     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
+     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
 \end{minted}

### 2024-02-23 paper/paper.tex
minor work on sections 1 and 5

--- 2024-02-23 paper/paper.tex
@@
 recursive programs we can write.
+\mvol{What is the ``every-day programmer'' here? Are they working in Coq/Agda?}
 
@@
 \mpav{Moved this from Section 3}
-goals are: (1) the
-mechanisation of hylomorphisms and their laws should be axiom-free; and (2) Coq
+
+The primary goals of this work are:
+\begin{enumerate}
+  \item that the
+  mechanisation of hylomorphisms and their laws should be axiom-free; and
+  \item that Coq
 hylomorphisms should be extracted to recursive OCaml functions that use no
-instances of \mintinline{OCaml}{Obj.magic}.  Although the use of
-\mintinline{OCaml}{Obj.magic} is safe in the code extracted by Coq, it has been
-proven that, for higher-order programs, simple interoperations can lead to
-incorrect behaviour or even segfaults~\cite{forster:hal-04329663}.  Therefore,
-the avoidance of axioms and  \mintinline{OCaml}{Obj.magic} means that the
-extracted code will have stronger safety guarantees, and will allow the safe
-interoperation with other OCaml code.
+instances of \mintinline{OCaml}{Obj.magic}.
+\end{enumerate}
+In OCaml, \mintinline{OCaml}{Obj.magic} allows for unsafe casts between types.
+Although the use of \mintinline{OCaml}{Obj.magic} should be safe in the code
+extracted by Coq, it has been proven that, for higher-order programs, simple
+interoperations can lead to incorrect behaviour or even
+segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
+\mintinline{OCaml}{Obj.magic} means that the extracted code will have stronger
+safety guarantees, and will allow the safe interoperation with other OCaml code.
+\mvol{This is good!}
 
@@
 functions that manipulate lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and
-Emoto went further and\ldots~\cite{10.1007/978-3-030-34175-6_11}
-
-\subsection{Termination of Nonstructural Recursion}
+Emoto went further and\ldots~\cite{MurataE19}
+
+\mvol{Need more general citations for program calculation.}
+
+\subsection{Termination Checking}
 
@@
 
-Another approach to certifying termination of nonstructurally recursive
-functions is the use of \emph{sized types}. Sized types were introduced by
-Hughes et al.\ as a way to track/verify various properties of recursive
-programs, including productivity and termination~\cite{HughesPS96}. The core
-idea of sized types is that types express bounds on the sizes of recursive data
-structures. With this approach, algorithms are implemented in the standard
-functional way, following the divide-and-conquer pattern of splitting, recurring,
-and merging, with the addition of the data types being indexed by a static
-approximation of the relative \emph{size}. An advantage of this approach is
-that it allows for code to be written in a natural way. There are costs, however.
-Programs must be written in a way that accounts for the handling of size indexes,
-and support for the size types must be added to the language.
+Another approach to certifying termination of recursive functions is the use of
+\emph{sized types}. Sized types were introduced by Hughes et al.\ as a way to
+track/verify various properties of recursive programs, including productivity
+and termination~\cite{HughesPS96}. The core idea of sized types is that types
+express bounds on the sizes of recursive data structures. With this approach,
+algorithms are implemented in the standard functional way, following the
+divide-and-conquer pattern of splitting, recurring, and merging, with the
+addition of the data types being indexed by a static approximation of the
+relative \emph{size}. An advantage of this approach is that it allows for code
+to be written in a natural way. There are costs, however. Programs must be
+written in a way that accounts for the handling of size indexes, and support for
+the size types must be added to the language.
 

### 2024-02-23 paper/paper.tex
a bit more explanation of structural vs nonstructural recursion in section 5

@@
 \emph{syntactic} check that definitions avoid the introduction of
-non-normalizable terms. However, some desirable definitions will not be accepted
-by such checks.
-
-In particular, the problem of \emph{nonstructural recursion} is
-well-studied~\cite{BOVE_KRAUSS_SOZEAU_2016}.
+non-normalizable terms. This sort of check generally looks for a
+\emph{structural decrease} of arguments in the recursive calls of a function.
+Coq uses such a check, and it works for many classic functional programming
+patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
+definitions will not be accepted by such checks.
+
+In particular, the problem of \emph{nonstructural recursion} (including
+divide-and-conquer algorithms) is well-studied~\cite{BOVE_KRAUSS_SOZEAU_2016}.
 % A distinction between standard and

### 2024-02-26 paper/paper.tex
add comment to section 1 about establishing the main goals and contributions of the paper
Stuff

@@
 
+\mvol{Need some transiiton here to build up to listing the goals and contributions of the paper.
+Also, ``axiom-free'' and ``obj.magic'' come out of nowhere here, so both need to be established earlier in the introduction.}
 The primary goals of this work are:
@@
 \begin{minted}{Coq}
-Context (Shape : Set) (Pos : Shape -> Set).
+Context (Shape : Type) (Pos : Shape -> Type).
 \end{minted}
@@
 \begin{minted}{Coq}
-Record App (X : Set) := { shape : Shape; contents : Pos shape -> X }.
+Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
 \end{minted}
@@
 \begin{minted}{Coq}
-Context (Shape : Set) (Pos : Set) (Valid : Shape -> Pos -> Prop).
-Record App (X : Set) := { shape    : Shape; 
+Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
+Record App (X : Type) := { shape    : Shape; 
                           contents : {p | Valid shape p} -> X }.
@@
 \begin{minted}{Coq}
-Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
+Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
 Coercion is_true (b : bool) : Prop := b = true.
@@
 \begin{minted}{Coq}
-Record App `(C : Cont Shape Pos) (X : Set) :=
+Record App `(C : Cont Shape Pos) (X : Type) :=
   MkApp { shape : Shape; contents : PosOf C shape -> X }.
@@
 
-Definition Ks (A : Set) := A.
+Definition Ks (A : Type) := A.
 Definition Kp := False.
@@
 
-These definitions are isomorphic to the equivalent container extension.
-
-We illustrate how these definitions
-\paragraph*{Isomorphisms with Container Extensions} Our definitions involving
-polynomial types are isomorphic to polynomial functors as container extensions.
+\subsubsection{Isomorphisms of Coq Types and Container Extensions}
+
+Using custom equivalences and proper morphisms, we redefine the definitions of
+container extensions and container equality. In particular, container
+extensions require a proper morphism to check the validity of positions in
+shapes, and container equality now uses equivalences of shapes and contained
+elements: 
+\begin{minted}{Coq}
+Class Cont (Shape : Type) (Pos : Type) := { valid : Shape ~> Pos ~> bool }.
+Inductive AppR `{e : equiv X} (x y : App F X) : Prop :=
+  | AppR_ext (Es : shape x =e shape y)
+             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
+\end{minted}
+Note the use of \mintinline{Coq}{=e} instead of Coq's standard equality.  For
+positions, however, we chose to use Coq's propositional equality, since these
+would leads to simpler code. We explain why in Section~\ref{sec:cata}, and the
+mechanisation of initial algebras of container extensions. The definitions in
+Figure~\ref{fig:poly} are redefined to account for these changes in a
+straightforward way.
+
+We prove the isomorphisms of polynomial types and the equivalent (redefined)
+container extensions. As an example, we discuss (informally) the isomorphisms
+of pairs with their equivalent container extensions. Suppose that we
+know that \mintinline{Coq}{App F X} is isomorphic to \mintinline{Coq}{A}, and
+\mintinline{Coq}{App G X} is isomorphic to \mintinline{Coq}{B}. Then we can
+show that \mintinline{Coq}{App (Prod F G) X} is isomorphic to
+\mintinline{Coq}{A * B}.  If we have an element of type 
+\mintinline{Coq}{App (Prod F G) X},
+using the \mintinline{Coq}{inl} position, we can obtain
+\mintinline{Coq}{App F X}.  Similarly, using \mintinline{Coq}{inr}, we can
+obtain \mintinline{Coq}{App G X}. Since these are the only two valid positions
+in the shape of pairs, we have finished. It is now sufficient to use the
+isomorphisms of \mintinline{Coq}{App F X} and \mintinline{Coq}{App G X} to
+obtain \mintinline{Coq}{A * B}. Similarly if we have \mintinline{Coq}{A * B},
+we can first use the isomorphisms of 
+\mintinline{Coq}{A} and
+\mintinline{Coq}{B} to obtain
+\mintinline{Coq}{App F X * App G X}, and then construct the necessary container
+extension. Given \mintinline{Coq}{p_inl}/\mintinline{Coq}{p_inr} that act as
+\mintinline{Coq}{inl/inr} but on valid product positions, the relevant
+functions that define the isomorphism are:
+\begin{minted}{Coq}
+Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
+  ({| shape := shape (fst x); cont := fun e => cont x (e_inl e) |}, 
+   {| shape := shape (snd x); cont := fun e => cont x (e_inr e) |}).
+Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
+{| shape := (shape (fst x), shape (snd x));
+   cont := fun e => match val e with 
+                    | inl p => fun H => cont (fst x) (ValidPos p H)
+                    | inr p => fun H => cont (snd x) (ValidPos p H)
+                    end (IsValid e)
+|}.
+\end{minted}
+Proving that these functions indeed form an isomorphism is straightforward,
+using the fact that Product containers only have two valid positions.
 
 \subsubsection{Generalised Rewriting for Morphisms}
+
+Since our framework only allows types with equivalences, and proper morphisms,
+we can leverage Coq's \emph{setoid rewrite}
+
+\dcas{Blabla: examples, no need to explain setoid rewrite in detail here}
 

### 2024-02-26 paper/paper.tex
I've made a bunch of changes which I don't have any idea right now what hey are.

--- 2024-02-23 paper/paper.tex
@@
 \section{Introduction}
+\label{sec:intro}
 Recursive definitions cannot be proven well-defined automatically due to the
@@
 recursive programs we can write.
-\mvol{What is the ``every-day programmer'' here? Are they working in Coq/Agda?}
 
@@
 \mpav{Moved this from Section 3}
-
-The primary goals of this work are:
-\begin{enumerate}
-  \item that the
-  mechanisation of hylomorphisms and their laws should be axiom-free; and
-  \item that Coq
+goals are: (1) the
+mechanisation of hylomorphisms and their laws should be axiom-free; and (2) Coq
 hylomorphisms should be extracted to recursive OCaml functions that use no
-instances of \mintinline{OCaml}{Obj.magic}.
-\end{enumerate}
-In OCaml, \mintinline{OCaml}{Obj.magic} allows for unsafe casts between types.
-Although the use of \mintinline{OCaml}{Obj.magic} should be safe in the code
-extracted by Coq, it has been proven that, for higher-order programs, simple
-interoperations can lead to incorrect behaviour or even
-segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
-\mintinline{OCaml}{Obj.magic} means that the extracted code will have stronger
-safety guarantees, and will allow the safe interoperation with other OCaml code.
-\mvol{This is good!}
+instances of \mintinline{OCaml}{Obj.magic}.  Although the use of
+\mintinline{OCaml}{Obj.magic} is safe in the code extracted by Coq, it has been
+proven that, for higher-order programs, simple interoperations can lead to
+incorrect behaviour or even segfaults~\cite{forster:hal-04329663}.  Therefore,
+the avoidance of axioms and  \mintinline{OCaml}{Obj.magic} means that the
+extracted code will have stronger safety guarantees, and will allow the safe
+interoperation with other OCaml code.
 
@@
 In order to abstract recursion patters we need to be able to talk about functors
-which have the suitable fixed-point properties. Instances of these functors are
-the polynomial ones which are in one-to-one correspondence with the so called
-\emph{containers}~\cite{AbbottAG05}.
-In this section we introduce containers
+which have the suitable fixed-point properties, i.e. those who have an initial
+algebra. Instances of these functors are the polynomial ones.  A common approach
+to construct such functors is to use \emph{container}~\cite{AbbottAG05}.  In
+this section we introduce the basics of polynomial functors as containers
+(Section~\ref{sec:containers}).  However, reasoning about containers equality
+will require us to consider both functional extensionality and heterogeneous
+equality. As stated in previous sections~\ref{sec:intro}, this is rather
+problematic since a stric requirement for this work is that we should remain
+axiom-free. For this purpose in Section~\ref{sec:types-morphisms} we introduce
+a custom equivalence relation on types in order to avoid using the functional
+extensionality axiom.
 
@@
 \label{sec:containers}
-In order to To construct fixpoints of functors, these need to build
-\emph{strictly positive types}. A common approach to construct such strictly
-positive endofunctors is to use \emph{container functors}~\cite{AbbottAG05}.
-However, reasoning about container and container functor equality requires to
-consider both functional extensionality and heterogeneous equality. This is
-problematic since, as stated in Section~\ref{sec:intro}, our goals to remain
-axiom-free.  Furthermore, standard representations of containers in Coq will
-lead to OCaml code that contains \mintinline{OCaml}{Obj.magic}. We will explain
-why, and our mechanisation of containers in the remainder of this section.
-
-A container is defined by a type of \emph{shapes}, and a \emph{family of
-position types} indexed by shapes.
+A container is defined by a type of \emph{shapes} and a \emph{family of position
+types} indexed by shapes.
 \begin{minted}{Coq}
@@
 \end{minted}
-An \emph{extension} of this container is a functor, whose action on objects is
-given by:
+An \emph{extension} of this container is a functor defined as follows:
 \begin{minted}{Coq}
@@
 \end{minted}
-In our mechanisation, we use \mintinline{Coq}{App} to represent container
-extensions.  Generally, any functor that is isomorphic to a container extension
-is called a \emph{container functor}. However, in our work we will call
-\emph{container functor} to the functors defined by a specific container
-extension.
+with the obvious action on morphisms. Generally, any functor that is isomorphic
+to a container extension is called a \emph{container functor}. However, in our
+work we will call \emph{container functor} to the functors defined by a specific
+container extension.
 
@@
 Record App `(C : Cont Shape Pos) (X : Set) :=
-  MkApp { shape : Shape; contents : PosOf C shape -> X }.
+  { shape : Shape; contents : PosOf C shape -> X }.
 \end{minted}
@@
 From such initial algebras, we define \emph{catamorphisms} by recursively
-traversing their structure. As usual, a catamorphism takes an
+traversing their structure. As usual, catamorphisms take an
 \mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
@@
       let (sx, kx) := ax in
-      g (MkApp sx (fun e => f (kx e)))
+      g (MkCont sx (fun e => f (kx e)))
   end.
@@
 \end{minted}
-Catamorphisms from this initial algebra are denoted by \mintinline{Coq}{ccata},
-and they can be used to define the isomorphism between 
-\mintinline{Coq}{{x : GFix F | FinF x}} and
-\mintinline{Coq}{LFix F}. In particular: 
-\begin{minted}{Coq}
-Lemma cata_ccata `{equiv A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
-Lemma ccata_cata `{equiv A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
-\end{minted}
-Using this new initial algebra, we can compose catamorphisms and anamorphisms,
-\emph{as long as the anamorphisms produce finite trees}. However, if
-anamorphisms produce finite trees, then we can also encode in Coq the notion of
-\emph{recursive anamorphisms} and \emph{recursive coalgebras}~\cite{TODO}.
-Recursive coalgebras are coalgebras whose anamorphism produces finite trees.
-The following predicate captures this fact. Note how all the predicate states
-is that \mintinline{Coq}{h} can only be applied finitely many times:
-\begin{minted}{Coq}
-Inductive RecF (h : Coalg F A) : A -> Prop :=
-| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
-\end{minted}
-We show how this predicate is equivalent to the anamorphism of such coalgebras
-producing finite trees:
-\begin{minted}{Coq}
-Lemma ana_rec_term `{equiv A} (h : Coalg F A)
-  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
-\end{minted}
-Recursive anamorphisms are implemented in Coq recursively on the proof that
-they can only be applied finitely many times (\mintinline{Coq}{RecF}).  Note
-that such proofs will be erased when extracting code to OCaml. This is standard
-practice when building general recursive functions in Coq\footnote{\dcas{This
-technique had a name in the community, but I cannot remember. Something
-predicates}}. The core definitions are the following:
-\begin{minted}{Coq}
-Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
-
-Definition rana_f__ `{equiv A} (c : Coalg F A)
-  : forall x : A, RecF c x -> LFix F
-  := fix f (x : A) (H : RecF c x) :=
-       let hx := c x in
-       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
-\end{minted}
-Here, \mintinline{Coq}{RecF_inv} provides the structurally smaller element for
-the Coq recursive definition. We then couple coalgebras with a proof that they
-are recursive, and add a coercion from recursive coalgebras to coalgebras.
-\begin{minted}{Coq}
-Definition RCoalg F A = {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
-\end{minted}
-Similarly to the rest of the definitions, we prove that recursive anamorphisms
-are respectful, and use this fact to build them as higher order proper
-morphisms, and prove its universal law, i.e.\ that recursive anamorphisms are
-the unique solution to the equation 
-\mintinline{Coq}{f =e l_in \o fmap f \o h},
-where \mintinline{Coq}{h} is a recursive coalgebra.
-\begin{minted}{Coq}
+Using this predicate, we can define the notion of \emph{recursive coalgebras}
+as those that build finite trees. Recursive coalgebras are implemented in Coq
+recursively on the proof that they produce finite trees.  Such proofs will be
+erased when extracting code to OCaml. This is standard practice when building
+general recursive functions in Coq\footnote{\dcas{This technique had a name in
+the community, but I cannot remember. Something predicates}}.
+\begin{minted}{Coq}
+Notation rana_f__ h :=
+  (fix f x H :=
+    let hx := h x in
+    LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e)))).
+
 Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...
@@
 
-\subsection{Recursive Hylomorphisms}
-
-Given an $F$-algebra and $F$-coalgebra, a solution to the equation $f = a \circ
-F\; f \circ c$ is known as a hylomorphism. Unfortunately, this definition
-cannot be accepted by Coq, since there is no structurally smaller argument to
-do recursion, and there is no corecursive definition that is guaranteed to be
-guarded, due to the use of an arbitrary algebra $a$. Furthermore, solutions to
-the hylomorphism equation are \emph{not} guaranteed to be unique, or to even
-exist. To address these issues, we mechanise \textbf{recursive hylomorphisms}
-which, as we will see, are guaranteed to have a unique solution.
-
-Hylomorphisms \emph{are} the unique solution to $f = a \circ F\; f \circ c$, if
-$c$ is a recursive coalgebra. We show this by first providing the definition
-for hylomorphisms in Coq, and then proving that any function satisfying this
-equation is a hylomorphism. We call these hylomorphisms with a recursive
-coalgebra, \textbf{recursive hylomorphisms}.
+\subsection{Recursive Hylos}
+
+Recursive hylomorphisms are built from an algebra and a recursive coalgebra:
 \begin{minted}{Coq}
 Definition hylo_def (a : Alg F B) (c : Coalg F A)
-  : forall (x : A), RecF c x -> B := fix f x H 
-  := match c x as h0 return
-       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
+: forall (x : A), RecF c x -> B
+:= fix f x H :=
+  match c x as h0
+        return
+        (forall e : Pos (shape h0), RecF c (cont h0 e)) ->
+        B
   with
-     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
+  | MkCont s_x c_x => fun H => a (MkCont s_x (fun e => f (c_x e) (H e)))
   end (RecF_inv H).
 \end{minted}
-The definition of recursive hylomorphisms is recursive on the proof that the
-anamorphism can only be applied finitely many times on any input. We again use
-\mintinline{Coq}{RecF_inv} to obtain the structurally smaller proof to use in
-the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
-we prove that \mintinline{Coq}{hylo_def} is respectful, and use this proof to
-built the corresponding higher-order proper morphism:
-\begin{minted}{Coq}
-Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
-\end{minted}
-Finally, we show that recursive hylomorphisms are the unique solution to
-the hylomorphism equation.
+
 \begin{minted}{Coq}
@@
 \end{minted}
-This result can be used to prove hylomorphism laws in a way that is analogous
-to pen-and-paper program calculation techniques. We show the rule for 
-hylomorphism fusion as an example:
-\begin{minted}{Coq}
-Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) 
-  (g2 : Alg F C) (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
+
+Recursive hylo universal property allows us to prove common optimisations:
+
+\begin{minted}{Coq}
+    Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
+    (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
     : f2 \o hylo g1 h1 =e hylo g2 h1.
   Proof.
-  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
     rewrite hylo_univ.
-  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
     rewrite fmap_comp.
-  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
     rewrite compA.
-  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
     rewrite <- E2.
-  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
     rewrite <- compA.
-  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
     rewrite <- compA.
-  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
     rewrite (compA g1).
-  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
     rewrite <- hylo_unr.
-  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
     reflexivity.
   Qed.
-\end{minted}
-Using the hylo fusion law, we can prove the well-known \emph{deforestation
-optimisation}, where two consecutive recursive computations, one that builds a
-data structure, and another one that consumes it, can be fused together into a
-single recursive definition. This, in turn, allows us to prove that a recursive
-hylomorphism is the composition of a catamorphism and a recursive anamorphism.
-\begin{minted}{Coq}
+
   Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
@@
   : cata g \o rana h =e hylo g h.
-\end{minted}
-This result will allow us to \emph{calculate} programs by providing small, easy
-to verify implementations, and combining them into efficient recursive
-functions by \emph{using Coq's proofs of hylomorphism laws as optimisation
-passes}.
-
-\section{Extraction}
-
-\dcas{Talk About this}
-\begin{minted}{OCaml}
-  let rec msort x = match x with
-| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
-     let (l0, r) = splitL x [] [] in
-     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
-     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
 \end{minted}
@@
 
-\subsection{Program Calculation}
-
-Within the domain of program optimization, program calculation serves as a
-well-established programming technique aimed at deriving efficient programs from
-their naive counterparts through systematic program transformation. This area
-has been extensively explored over the years. Tesson et al.\ demonstrated the
-efficacy of leveraging Coq to establish an approach for implementing a robust
-system dedicated to verifying the correctness of program transformations for
-functions that manipulate lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and
-Emoto went further and\ldots~\cite{MurataE19}
-
-\mvol{Need more general citations for program calculation.}
-
-\subsection{Termination Checking}
-
-Termination and productivity are non-trivial properties of programs, and various
-methods have been proposed for checking that these properties hold. A common
-approach is guardedness checking~\cite{10.5555/646535.695850}, which is a
-\emph{syntactic} check that definitions avoid the introduction of
-non-normalizable terms. However, some desirable definitions will not be accepted
-by such checks.
-
-In particular, the problem of \emph{nonstructural recursion} is
-well-studied~\cite{BOVE_KRAUSS_SOZEAU_2016}.
-% A distinction between standard and
-% nonstandard recursion lies in the fact that standard recursion adheres strictly
-% to recursing on pattern variables, avoiding reliance on results from other
-% function calls. Nonstandard recursion may deviate from this convention, allowing
-% for more flexibility.
-Certain functions that are not structurally recursive can be reformulated using
-a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
-Take, for instance, division by iterated subtraction, which is inherently
-non-structurally recursive since it involves recursion based on the result of a
-subtraction. There is a nonstandard implementation of divivion found on Coq's
-standard library, which involves a four-argument function that effectively
-combines subtraction and division. Similarly, the mergesort in Coq's standard
-library uses an ``explicit stack of pending merges'' in order to avoid issues
-with nonstructural definitions. There is a major downside, however; as noted by
-Abreu et al., the result is ``barely recognizable as a form of
-mergesort''~\cite{AbreuDHJMS23}.
-
-\subsubsection{Sized Types}
-
-Another approach to certifying termination of recursive functions is the use of
-\emph{sized types}. Sized types were introduced by Hughes et al.\ as a way to
-track/verify various properties of recursive programs, including productivity
-and termination~\cite{HughesPS96}. The core idea of sized types is that types
-express bounds on the sizes of recursive data structures. With this approach,
-algorithms are implemented in the standard functional way, following the
-divide-and-conquer pattern of splitting, recurring, and merging, with the
-addition of the data types being indexed by a static approximation of the
-relative \emph{size}. An advantage of this approach is that it allows for code
-to be written in a natural way. There are costs, however. Programs must be
-written in a way that accounts for the handling of size indexes, and support for
-the size types must be added to the language.
-
-This approach has been used to express nonstructurally recursive
-algorithms in Agda: Copello et al.\ used sized types in a straightforward
-formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
-in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
-Coq as well, though it has not yet been adopted~\cite{CHAN_2023}.
-
-\subsubsection{Divide and Conquer Recursion}
-
-Abreu et al.\ propose\ldots~\cite{AbreuDHJMS23}. Their approach has similarities to what is presented here.
-
-They differ in requiring the assumption of the functional extensionality axiom.
-Additionally, they are concerned with the erasure of termination proofs but have
-little discussion of what the resulting extracted code \emph{looks like} (that
-is, whether the extracted OCaml code resembles the natural formulation of the
-recursive function).
-\mvol{We should find out for sure if they use \texttt{obj.magic}.}
+\subsection{Sized Types}
+
+Another approach to certifying termination of nonstructurally recursive
+functions is the use of \emph{sized types}. Sized types were introduced in
+\cite{HughesPS96} as a way to track/verify various properties of recursive
+programs, including productivity and termination. The core idea of sized types
+is that types express bounds on the sizes of recursive data structures. This
+approach has been used to express nonstructurally recursive algorithms in Agda:
+\cite{CopelloTB14} used sized types in a straightforward formulation of
+mergesort. Such an approach has also been used in MiniAgda~\cite{abs-1202-3496}.
 

### 2024-02-27 paper/paper.tex
Stuff

@@
-\documentclass[a4paper, UKenglish, cleveref, autoref, thm-restate]{lipics-v2021}
+\documentclass[anonymous, a4paper, UKenglish, cleveref, autoref, thm-restate]{lipics-v2021}
 \usepackage[utf8]{inputenc} % Required for inputting international characters
@@
 
-\author{Jane {Open Access}}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{David Castro-Perez}{School of Computing, University of Kent}{d-castro-perez@kent.ac.uk}{}{}
+\author{Marco Paviotti}{School of Computing, University of Kent}{m.paviotti@kent.ac.uk}{}{}
+\author{Michael Vollmer}{School of Computing, University of Kent}{m.vollmer@kent.ac.uk}{}{}
+
+%\author{Jane {Open Access}}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.
@@
 
-\author{Joan R.
+%\author{Joan R.
 Public\footnote{Optional footnote, e.g.
@@
 
-\authorrunning{J. Open Access and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\authorrunning{D. Castro-Perez, M. Paviotti, and M. Vollmer} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+
+\Copyright{D. Castro-Perez, M. Paviotti, and M. Vollmer} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 
-\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
+\keywords{hylomorphisms, Coq, code extraction, verification} %TODO mandatory; please add comma-separated list of keywords
 
@@
 
-\acknowledgements{I want to thank \dots}%optional
+%\acknowledgements{I want to thank \dots}%optional
 
@@
 container as a type that captures \emph{all} the potential positions in all
-possible shapes, together with a predicate \mintinline{Coq}{Valid} that
+possible shapes, together with a predicate \coq{Valid} that
 determines whether a position is valid in a shape.
@@
 This approach solves the the problem of the unsafe casts, since the code
-extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml singleton
-type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.
+extracted for \coq{{p | Valid shape pos}} will be an OCaml singleton
+type defined as the OCaml equivalent to \coq{Pos}.
 
@@
 container extensions. The contents of the container extensions need to be
-accessed via a function \mintinline{Coq}{contents}. The domain of this function
+accessed via a function \coq{contents}. The domain of this function
 depends on the shape of the container. This means that, for a given type of
@@
 type $A$, maps everything to it, $X \mapsto A$, only needs one shape ($A$, L9
-of Fig~\ref{fig:poly}), and zero positions (In Coq, \mintinline{Coq}{False},
+of Fig~\ref{fig:poly}), and zero positions (In Coq, \coq{False},
 L10 of Fig~\ref{fig:poly}). The reason why there are zero valid positions in
 a constant container extension is because, given something of type 
-\mintinline{Coq}{App (Const (Ks A) Kp) X}, it should be impossible to get
-something of type \mintinline{Coq}{X}. But from the container shape, we will
+\coq{App (Const (Ks A) Kp) X}, it should be impossible to get
+something of type \coq{X}. But from the container shape, we will
 always be able to get the element of type $A$\footnote{%
-  Our decision to use \mintinline{Coq}{False} is arbitrary. In general, we
+  Our decision to use \coq{False} is arbitrary. In general, we
   could have chosen any time, with a validity predicate that always returns
-  \mintinline{Coq}{false}. It would make no difference for code extraction.
-}. This functor is, therefore \mintinline{Coq}{Const nat}.
+  \coq{false}. It would make no difference for code extraction.
+}. This functor is, therefore \coq{Const nat}.
 
@@
 type for positions. To build an element of the container extension type, we
-will need to provide a function of type \mintinline{Coq}{unit -> X}, which is
-isomorphic to \mintinline{Coq}{X}.
+will need to provide a function of type \coq{unit -> X}, which is
+isomorphic to \coq{X}.
 
@@
 extensions. What is a valid position in a given shape? If the position is
-\mintinline{Coq}{inl p} (or \mintinline{Coq}{inr p}), then it is only valid if
-\mintinline{Coq}{p} is valid in the first (resp. second) component of the
+\coq{inl p} (or \coq{inr p}), then it is only valid if
+\coq{p} is valid in the first (resp. second) component of the
 shape. The product of our two container extensions is
-\mintinline{Coq}{Prod Id Id}.
+\coq{Prod Id Id}.
 
@@
 check the validity of shapes in positions, their constructor must match, e.g.
-shape \mintinline{Coq}{inl s} and shape \mintinline{Coq}{inl p}, with
-\mintinline{Coq}{valid s p}.
+shape \coq{inl s} and shape \coq{inl p}, with
+\coq{valid s p}.
 
 Our final container extension is the following: 
-\mintinline{Coq}{Sum (Const nat) (Prod Id Id)}.
+\coq{Sum (Const nat) (Prod Id Id)}.
 The shape of this functor is: 
-\mintinline{Coq}{nat + unit * unit}. The positions of this functor
+\coq{nat + unit * unit}. The positions of this functor
 are \mintinline[breaklines]{Coq}{False + (unit + unit)}, which is isomorphic
-to \mintinline{Coq}{unit + unit}, which will only be valid in the RHS.
-
-The Coq value \mintinline{Coq}{inl n} of type
-\mintinline{Coq}{nat + bool * bool} would look as follows as a container
+to \coq{unit + unit}, which will only be valid in the RHS.
+
+The Coq value \coq{inl n} of type
+\coq{nat + bool * bool} would look as follows as a container
 extension 
-\mintinline{Coq}{App (Sum (Const nat) (Prod Id Id)) bool}:
+\coq{App (Sum (Const nat) (Prod Id Id)) bool}:
 \begin{minted}{Coq}
@@
 \end{minted}
-We omit the definition of \mintinline{Coq}{uninhabited}, which exploits the
-fact that \mintinline{Coq}{e} for \mintinline{Coq}{inl n} implies
-\mintinline{Coq}{False} to build an element of type \mintinline{Coq}{bool}%
-\footnote{Note that, since \mintinline{Coq}{e} is uninhabited we could return
-any arbitrary value instead of \mintinline{Coq}{uninhabited e}.}.
-
-Similarly, the Coq value \mintinline{Coq}{inr (true, false)} as a container
+We omit the definition of \coq{uninhabited}, which exploits the
+fact that \coq{e} for \coq{inl n} implies
+\coq{False} to build an element of type \coq{bool}%
+\footnote{Note that, since \coq{e} is uninhabited we could return
+any arbitrary value instead of \coq{uninhabited e}.}.
+
+Similarly, the Coq value \coq{inr (true, false)} as a container
 extension is as follows:
@@
 \end{minted}
-Here, the position \mintinline{Coq}{inl f} is impossible, as
-\mintinline{Coq}{f} would need to be a proof of \mintinline{Coq}{False}.
+Here, the position \coq{inl f} is impossible, as
+\coq{f} would need to be a proof of \coq{False}.
 Of the two remaining positions, the left one will return 
-\mintinline{Coq}{true}, and the right position \mintinline{Coq}{false}.
+\coq{true}, and the right position \coq{false}.
 \begin{figure}
@@
 In particular, we want to reason about the equality of functions of types such
-as \mintinline{Coq}{App F A -> B} (or \mintinline{Coq}{B -> App F A}). Since 
+as \coq{App F A -> B} (or \coq{B -> App F A}). Since 
 these types now come with their own equivalence, any function that manipulates
 them needs to be \emph{respectful}. I.e.\ given 
-\mintinline{Coq}{R : X -> X -> Prop} and \mintinline{Coq}{R' : Y -> Y -> Prop},
+\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
 we want functions (morphisms) that satisfy the following property:
@@
 \label{sec:types-morphisms}
-We address the different forms of equality by coupling every type with an
-associated equivalence relation, and every morphism with a proof that it is
-\emph{respectful} (i.e. \emph{proper} morphisms with respect to the function
-\emph{respectfulness} relation). We use \mintinline{Coq}{=e} to denote the
-equivalence relation associated with a type. Every Coq type is associated by
-default with the standard Coq propositional equality, unless a different
-equivalence is specified. We use the type-class mechanism to help Coq's code
-extraction mechanism to remove any occurence of custom equivalence relations
-in the extracted OCaml code.
+We address the different forms of equality by defining a class of setoids,
+i.e.\ types with na associated equivalence relation, and considering only
+functions that respect the associated equivalences, i.e. \emph{proper}
+morphisms with respect to the function \emph{respectfulness} relation.  We use
+the type-class mechanism, instead of setoids in Coq's standard library, to help
+Coq's code extraction mechanism remove any occurence of custom equivalence
+relations in the extracted OCaml code.  We use \coq{=e} to denote
+the equivalence relation of a setoid. By default, we associate every Coq type
+with the standard propositional equality, unless a different equivalence is
+specified (i.e.\ we allow overlapping instances, and Coq's propositional
+equality takes the lowest priority). 
 \begin{minted}{Coq}
 Reserved Notation "f =e g" (at level 70, no associativity).
-Class equiv A : Type :=
+Class setoid A : Type :=
   MkEquiv
@@
 \end{minted}
-Given types \mintinline{Coq}{A} and  \mintinline{Coq}{B}, with their respective
-equivalence relations \mintinline{Coq}{eA : A -> A -> Prop} and
-\mintinline{Coq}{eB : B -> B -> Prop}, the we define the type
-\mintinline{Coq}{A ~> B} to represents \emph{proper} morphisms of the
-respectfulness relation of \mintinline{Coq}{R} to \mintinline{Coq}{R'}.
-\begin{minted}{Coq}
-Structure morph A {eA : equiv A} B {eB : equiv B} :=
+Given types \coq{A} and  \coq{B}, with their respective
+equivalence relations \coq{eA : A -> A -> Prop} and
+\coq{eB : B -> B -> Prop}, the we define the type
+\coq{A ~> B} to represents \emph{proper} morphisms of the
+respectfulness relation of \coq{R} to \coq{R'}.
+\begin{minted}{Coq}
+Structure morph A {eA : setoid A} B {eB : setoid B} :=
   MkMorph
@@
 equivalence relations. Since Coq's code extraction mechanism will erase any
-occurrence of \mintinline{Coq}{Prop} in the code, so objects of type
-\mintinline{Coq}{A ~> B} will be extracted to the OCaml equivalent to
-\mintinline{Coq}{A -> B}.
+occurrence of \coq{Prop} in the code, so objects of type
+\coq{A ~> B} will be extracted to the OCaml equivalent to
+\coq{A -> B}.
 
@@
 \begin{minted}{Coq}
-Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
-Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
-Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
-Instance prop_eq : equiv Prop.
-Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
+Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
+Instance pair_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A * B).
+Instance sum_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A + B).
+Instance prop_eq : setoid Prop.
+Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
 \end{minted}
@@
 straightworward.  Identity and composition are defined as 
-\mintinline{Coq}{fun x => x} and \mintinline{Coq}{fun f g x => f (g x)} 
+\coq{fun x => x} and \coq{fun f g x => f (g x)} 
 respectively, and the proofs that they are proper morphisms is straightforward,
 and automatically discharged by Coq. 
-Products are built using function \mintinline{Coq}{fun f g x => (f x, g x)},
-with the projections being the standard Coq  \mintinline{Coq}{fst} and
-\mintinline{Coq}{snd} functions. Similarly, sum injections are encoded using
-Coq's \mintinline{Coq}{inl} and \mintinline{Coq}{inr} constructors, and
+Products are built using function \coq{fun f g x => (f x, g x)},
+with the projections being the standard Coq  \coq{fst} and
+\coq{snd} functions. Similarly, sum injections are encoded using
+Coq's \coq{inl} and \coq{inr} constructors, and
 pattern matching on them uses the function 
-\mintinline{Coq}{fun f g x => match x with | inl y => f y | inr y => g y end}.
+\coq{fun f g x => match x with | inl y => f y | inr y => g y end}.
 The proofs that these morphisms are proper are straightforward.  Finally, we
@@
 Class Cont (Shape : Type) (Pos : Type) := { valid : Shape ~> Pos ~> bool }.
-Inductive AppR `{e : equiv X} (x y : App F X) : Prop :=
+Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
   | AppR_ext (Es : shape x =e shape y)
@@
 \end{minted}
-Note the use of \mintinline{Coq}{=e} instead of Coq's standard equality.  For
+Note the use of \coq{=e} instead of Coq's standard equality.  For
 positions, however, we chose to use Coq's propositional equality, since these
@@
 of pairs with their equivalent container extensions. Suppose that we
-know that \mintinline{Coq}{App F X} is isomorphic to \mintinline{Coq}{A}, and
-\mintinline{Coq}{App G X} is isomorphic to \mintinline{Coq}{B}. Then we can
-show that \mintinline{Coq}{App (Prod F G) X} is isomorphic to
-\mintinline{Coq}{A * B}.  If we have an element of type 
-\mintinline{Coq}{App (Prod F G) X},
-using the \mintinline{Coq}{inl} position, we can obtain
-\mintinline{Coq}{App F X}.  Similarly, using \mintinline{Coq}{inr}, we can
-obtain \mintinline{Coq}{App G X}. Since these are the only two valid positions
+know that \coq{App F X} is isomorphic to \coq{A}, and
+\coq{App G X} is isomorphic to \coq{B}. Then we can
+show that \coq{App (Prod F G) X} is isomorphic to
+\coq{A * B}.  If we have an element of type 
+\coq{App (Prod F G) X},
+using the \coq{inl} position, we can obtain
+\coq{App F X}.  Similarly, using \coq{inr}, we can
+obtain \coq{App G X}. Since these are the only two valid positions
 in the shape of pairs, we have finished. It is now sufficient to use the
-isomorphisms of \mintinline{Coq}{App F X} and \mintinline{Coq}{App G X} to
-obtain \mintinline{Coq}{A * B}. Similarly if we have \mintinline{Coq}{A * B},
+isomorphisms of \coq{App F X} and \coq{App G X} to
+obtain \coq{A * B}. Similarly if we have \coq{A * B},
 we can first use the isomorphisms of 
-\mintinline{Coq}{A} and
-\mintinline{Coq}{B} to obtain
-\mintinline{Coq}{App F X * App G X}, and then construct the necessary container
-extension. Given \mintinline{Coq}{p_inl}/\mintinline{Coq}{p_inr} that act as
-\mintinline{Coq}{inl/inr} but on valid product positions, the relevant
+\coq{A} and
+\coq{B} to obtain
+\coq{App F X * App G X}, and then construct the necessary container
+extension. Given \coq{p_inl}/\coq{p_inr} that act as
+\coq{inl/inr} but on valid product positions, the relevant
 functions that define the isomorphism are:
@@
 Algebras on container extensions are defined as expected. The carrier of the
-algebra is \mintinline{Coq}{A}, and the operations of the algebra are morphisms
+algebra is \coq{A}, and the operations of the algebra are morphisms
 of the following type:
@@
 \end{minted}
-Given a container extension \mintinline{Coq}{F}, we define \emph{two}
+Given a container extension \coq{F}, we define \emph{two}
 (isomorphic) initial algebras. We will show the need for the second initial
 algebra in the following section. Our fist initial algebra is defined as the
-straightforward least fixpoint of the container extension \mintinline{Coq}{F}:
+straightforward least fixpoint of the container extension \coq{F}:
 \begin{minted}{Coq}
@@
 contained types may not have (yet) an associated equivalence. In other words,
-we would need to provide an equivalence for \mintinline{Coq}{LFix}, but the
+we would need to provide an equivalence for \coq{LFix}, but the
 only way to do this is to define such equivalence simultaneously to the
-definition for  \mintinline{Coq}{LFix}.  We could have overcame this
-circularity by defining  \mintinline{Coq}{LFix} independently of container
+definition for  \coq{LFix}.  We could have overcame this
+circularity by defining  \coq{LFix} independently of container
 extensions, and define later the necessary
-\mintinline{Coq}{in}/\mintinline{Coq}{out} functions to convert between
-\mintinline{Coq}{LFix} and \mintinline{Coq}{App}. However, this extra layer
+\coq{in}/\coq{out} functions to convert between
+\coq{LFix} and \coq{App}. However, this extra layer
 interferes with code extraction, in that it requires destructing/constructing
-subsequent \mintinline{Coq}{LFix} and  \mintinline{Coq}{App}, which Coq is not
+subsequent \coq{LFix} and  \coq{App}, which Coq is not
 always able to inline.
@@
 by traversing the structure and equating the shapes and positions. This
-definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
+definition is equivalent to taking the fixpoint of \coq{AppR} in
 Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
-positions, the contents must be related by \mintinline{Coq}{LFixR}.
+positions, the contents must be related by \coq{LFixR}.
 \begin{minted}{Coq}
@@
 traversing their structure. As usual, a catamorphism takes an
-\mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
-function from \mintinline{Coq}{LFix F} to \mintinline{Coq}{A}.
-\begin{minted}{Coq}
-Definition cata_f `{equiv A} (g : Alg F A) : LFix -> A
+\coq{F}-algebra with carrier \coq{A}, and produces a
+function from \coq{LFix F} to \coq{A}.
+\begin{minted}{Coq}
+Definition cata_f `{setoid A} (g : Alg F A) : LFix -> A
 := fix f (x : LFix) :=
@@
 \begin{minted}{Coq}
-Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
+Definition cata `{eA : setoid A} : Alg F A ~> LFix ~> A := ...
 \end{minted}
 Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
-they are the unique solution to the equation \mintinline{Coq}{f =e g \o fmap f \o l_out}).
-The proof is by induction on the input to \mintinline{Coq}{f} and
-\mintinline{Coq}{cata g}.
-\begin{minted}{Coq}
-Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
+they are the unique solution to the equation \coq{f =e g \o fmap f \o l_out}).
+The proof is by induction on the input to \coq{f} and
+\coq{cata g}.
+\begin{minted}{Coq}
+Lemma cata_univ `{eA : setoid A} (g : Alg F A) (f : LFix ~> A)
     : f =e cata g <-> f =e g \o fmap f \o l_out.
@@
 \begin{minted}{Coq}
-  Definition ana `{eA : equiv A} : Coalg F A ~> A ~> GFix...
-
-  Lemma ana_univ `{eA : equiv A} (h : Coalg F A) (f : A ~> GFix)
+  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...
+
+  Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
     : f =e ana h <-> f =e g_in \o fmap f \o h.
@@
 \end{minted}
-Catamorphisms from this initial algebra are denoted by \mintinline{Coq}{ccata},
+Catamorphisms from this initial algebra are denoted by \coq{ccata},
 and they can be used to define the isomorphism between 
-\mintinline{Coq}{{x : GFix F | FinF x}} and
-\mintinline{Coq}{LFix F}. In particular: 
-\begin{minted}{Coq}
-Lemma cata_ccata `{equiv A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
-Lemma ccata_cata `{equiv A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
+\coq{{x : GFix F | FinF x}} and
+\coq{LFix F}. In particular: 
+\begin{minted}{Coq}
+Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
+Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
 \end{minted}
 Using this new initial algebra, we can compose catamorphisms and anamorphisms,
-\emph{as long as the anamorphisms produce finite trees}. However, if
-anamorphisms produce finite trees, then we can also encode in Coq the notion of
-\emph{recursive anamorphisms} and \emph{recursive coalgebras}~\cite{TODO}.
-Recursive coalgebras are coalgebras whose anamorphism produces finite trees.
-The following predicate captures this fact. Note how all the predicate states
-is that \mintinline{Coq}{h} can only be applied finitely many times:
+\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
+the supplementary material and Coq code to see how we encode this. In this
+paper, however, we focus on the fact that, anamorphisms produce finite trees,
+then their return type can also be the \emph{least} fixed point of the
+container extension. We encode in Coq the notion of \emph{recursive
+anamorphisms} and \emph{recursive coalgebras}~\cite{TODO}.  Recursive
+coalgebras are coalgebras whose anamorphism produces finite trees.  The
+following predicate captures this fact. The following predicate states that
+\coq{h} can only be applied finitely many times:
 \begin{minted}{Coq}
@@
 \begin{minted}{Coq}
-Lemma ana_rec_term `{equiv A} (h : Coalg F A)
+Lemma ana_rec_term `{setoid A} (h : Coalg F A)
   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
 \end{minted}
-Recursive anamorphisms are implemented in Coq recursively on the proof that
-they can only be applied finitely many times (\mintinline{Coq}{RecF}).  Note
-that such proofs will be erased when extracting code to OCaml. This is standard
-practice when building general recursive functions in Coq\footnote{\dcas{This
-technique had a name in the community, but I cannot remember. Something
-predicates}}. The core definitions are the following:
+Recursive anamorphisms are implemented in Coq recursively on the structure of
+the proof that they can only be applied finitely many times
+(\coq{RecF}).  Note that such proofs will be erased when extracting
+code to OCaml. This is standard practice when building general recursive
+functions in Coq\footnote{\dcas{This technique had a name in the community, but
+I cannot remember. Something predicates}}. The core definitions are the
+following:
 \begin{minted}{Coq}
@@
 
-Definition rana_f__ `{equiv A} (c : Coalg F A)
+Definition rana_f__ `{setoid A} (c : Coalg F A)
   : forall x : A, RecF c x -> LFix F
@@
 \end{minted}
-Here, \mintinline{Coq}{RecF_inv} provides the structurally smaller element for
+Here, \coq{RecF_inv} provides the structurally smaller element for
 the Coq recursive definition. We then couple coalgebras with a proof that they
@@
 \begin{minted}{Coq}
-Definition RCoalg F A = {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
-\end{minted}
+Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
+\end{minted}
+In some cases, proving that coalgebras can only be applied finitely many
+times is straightforward (e.g.\ they are recursive on some inductive type in
+Coq). However, in many cases, proving that a coalgebra is recursive directly
+can be tedious and cumbersome. To ease development, we prove that if a 
+given coalgebra respects a \emph{well-founded relation}, then it is 
+necessarily recursive.
+\begin{minted}{Coq}
+Definition respects_relation `{setoid A} (c : Coalg F A)
+      {B} (m : A -> B) (R : B -> B -> Prop)
+  := forall x (p : Pos (shape (c x))), R (m (cont (c x) p)) (m x).
+
+Lemma wf_coalg_rec `{setoid A} {B}
+  (m : A -> B) (R : B -> B -> Prop) (WF : well_founded R)
+  (c : Coalg F A) (RR : respects_relation c m R) : RecP c.
+\end{minted}
+In the above definition, \coq{m} is used as a form of ``metric''
+for termination, and the well-founded relation is on \coq{B}. This
+works equivalently to transporting a well-founded relation 
+\coq{R}
+on
+\coq{B} to \coq{A} as 
+\coq{R' := fun x y => R (m x) (m y)}, which is also trivially
+well-founded. Using these definitions, we can construct recursive coalgebras by
+proving that they respect a well-founded relation, which generally leads to
+simpler proofs.
+
 Similarly to the rest of the definitions, we prove that recursive anamorphisms
 are respectful, and use this fact to build them as higher order proper
-morphisms, and prove its universal law, i.e.\ that recursive anamorphisms are
+morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
 the unique solution to the equation 
-\mintinline{Coq}{f =e l_in \o fmap f \o h},
-where \mintinline{Coq}{h} is a recursive coalgebra.
-\begin{minted}{Coq}
-Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...
-
-Lemma rana_univ A {eA : equiv A} (h : RCoalg A) (f : A ~> LFix F)
+\coq{f =e l_in \o fmap f \o h},
+where \coq{h} is a recursive coalgebra.
+\begin{minted}{Coq}
+Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F :=...
+
+Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
     : f =e rana h <-> f =e l_in \o fmap f \o h.
@@
 exist. To address these issues, we mechanise \textbf{recursive hylomorphisms}
-which, as we will see, are guaranteed to have a unique solution.
-
-Hylomorphisms \emph{are} the unique solution to $f = a \circ F\; f \circ c$, if
-$c$ is a recursive coalgebra. We show this by first providing the definition
+which are guaranteed to have a unique solution to the hylomorphism equation.
+
+If $c$ is a recursive coalgebra, then the equation $f = a \circ F\; f \circ c$ 
+does have a unique solution.
+We show this by first providing the definition
 for hylomorphisms in Coq, and then proving that any function satisfying this
@@
 anamorphism can only be applied finitely many times on any input. We again use
-\mintinline{Coq}{RecF_inv} to obtain the structurally smaller proof to use in
+\coq{RecF_inv} to obtain the structurally smaller proof to use in
 the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
-we prove that \mintinline{Coq}{hylo_def} is respectful, and use this proof to
+we prove that \coq{hylo_def} is respectful, and use this proof to
 built the corresponding higher-order proper morphism:
@@
 
-Corollary cata_ana_hylo `(F : Cont Sh P) `{equiv A} `{equiv B}
+Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
   (g : Alg F B) (h : RCoalg F A)

### 2024-02-27 paper/paper.tex
Some style changes

@@
 \newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
-\newcommand{\coq}[1]{\mintinline{coq}{#1}}
+
+%\newcommand{\coq}[1]{\mintinline{coq}{#1}}
+\newmintinline[coq]{coq}{fontsize=\small}
+\newminted{coq}{fontsize=\small}
 
@@
 \mpav{I need this to be written in Coq notation}
-\begin{minted}{coq}
+\begin{coqcode}
 qs :: [a] -> [a]
@@
 qs (x:xs) = qs (filter (x <=) xs) ++ [x] ++ (qs (filter (x >) xs))
-\end{minted}
+\end{coqcode}
 This program is formed of essentially two parts. The first one is what is
@@
 \coq{TreeF}-coalgebra on lists over the type $A$
-\begin{minted}{coq}
+\begin{coqcode}
 c : [A] -> TreeF [A]
 c (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
-\end{minted}
+\end{coqcode}
 
@@
 over $A$ as well, that is a map
-\begin{minted}{coq}
+\begin{coqcode}
 a : TreeF [A] -> [A]
 a (left, x, right) = left ++ [x] ++ right
-\end{minted}
+\end{coqcode}
 
@@
 types} indexed by shapes.
-\begin{minted}{Coq}
+\begin{coqcode}
 Context (Shape : Type) (Pos : Shape -> Type).
-\end{minted}
+\end{coqcode}
 An \emph{extension} of this container is a functor defined as follows:
-\begin{minted}{Coq}
+\begin{coqcode}
 Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
-\end{minted}
+\end{coqcode}
 with the obvious action on morphisms. Generally, any functor that is isomorphic
@@
 determines whether a position is valid in a shape.
-\begin{minted}{Coq}
+\begin{coqcode}
 Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
@@
                           contents : {p | Valid shape p} -> X }.
-\end{minted}
+\end{coqcode}
 This approach solves the the problem of the unsafe casts, since the code
@@
 erased by Coq's code extraction mechanism.
-\begin{minted}{Coq}
+\begin{coqcode}
 Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
@@
   ValidPos { val : Pos; IsValid : valid shape pos }.
-\end{minted}
+\end{coqcode}
 Then, we define \emph{container extensions} as
-\begin{minted}{Coq}
+\begin{coqcode}
 Record App `(C : Cont Shape Pos) (X : Type) :=
   MkApp { shape : Shape; contents : PosOf C shape -> X }.
-\end{minted}
+\end{coqcode}
 The functorial action for container extensions on morphisms boils down to
 function composition, which trivially preserves identity and composition.
-\begin{minted}{Coq}
+\begin{coqcode}
 Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
   := {| shape := shape x; contents := fun e => f (contents x e) |}
-\end{minted}
+\end{coqcode}
 
@@
 \coq{App (Sum (Const nat) (Prod Id Id)) bool}:
-\begin{minted}{Coq}
+\begin{coqcode}
 {| shape := inl n; cont := fun e => uninhabited e|}
-\end{minted}
+\end{coqcode}
 We omit the definition of \coq{uninhabited}, which exploits the
@@
 extension is as follows:
-\begin{minted}{Coq}
+\begin{coqcode}
   {| shape := inr (tt,tt);
@@
   |}
-\end{minted}
+\end{coqcode}
 Here, the position \coq{inl f} is impossible, as
@@
 Coq:
-\begin{minted}{Coq}
+\begin{coqcode}
 Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
@@
            (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
-\end{minted}
+\end{coqcode}
 Note that  we do not care about the validity proof of the positions, only their
@@
 we want functions (morphisms) that satisfy the following property:
-\begin{minted}{Coq}
+\begin{coqcode}
 forall (x y : X), R x y -> R' (f x) (f y)
-\end{minted}
+\end{coqcode}
 
@@
 equality takes the lowest priority). 
-\begin{minted}{Coq}
+\begin{coqcode}
 Reserved Notation "f =e g" (at level 70, no associativity).
@@
 Notation "f =e g" := (eqRel f g).
-\end{minted}
+\end{coqcode}
 Given types \coq{A} and  \coq{B}, with their respective
@@
 respectfulness relation of \coq{R} to \coq{R'}.
-\begin{minted}{Coq}
+\begin{coqcode}
 Structure morph A {eA : setoid A} B {eB : setoid B} :=
@@
 Notation "A ~> B" := (@morph A _ B _).
-\end{minted}
+\end{coqcode}
 Note that we leverage on Coq's type class mechanism to fill in the necessary
@@
 be used when building proper morphisms that manipulate their respective types.
-\begin{minted}{Coq}
+\begin{coqcode}
 Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
@@
 Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
-\end{minted}
+\end{coqcode}
 
@@
 elements: 
-\begin{minted}{Coq}
+\begin{coqcode}
 Class Cont (Shape : Type) (Pos : Type) := { valid : Shape ~> Pos ~> bool }.
@@
              (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
-\end{minted}
+\end{coqcode}
 Note the use of \coq{=e} instead of Coq's standard equality.  For
@@
 functions that define the isomorphism are:
-\begin{minted}{Coq}
+\begin{coqcode}
 Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
@@
 |}.
-\end{minted}
+\end{coqcode}
 Proving that these functions indeed form an isomorphism is straightforward,
-using the fact that Product containers only have two valid positions.
+using the fact that \coq{Prod} containers only have two valid positions.
 
@@
 of the following type:
-\begin{minted}{Coq}
+\begin{coqcode}
 Notation Alg F A := (App F A ~> A).
-\end{minted}
+\end{coqcode}
 Given a container extension \coq{F}, we define \emph{two}
@@
 straightforward least fixpoint of the container extension \coq{F}:
-\begin{minted}{Coq}
+\begin{coqcode}
 Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
-\end{minted}
+\end{coqcode}
 This definition highlights why we cannot define equality of the positions and
@@
 We provide a custom induction scheme for initial algebras:
-\begin{minted}{Coq}
+\begin{coqcode}
 Lemma LFix_ind [P : LFix -> Prop]
@@
     -> forall l : LFix, P l.
-\end{minted}
+\end{coqcode}
 We define the equality of initial algebras of container extensions recursively,
@@
 positions, the contents must be related by \coq{LFixR}.
-\begin{minted}{Coq}
+\begin{coqcode}
 Fixpoint LFixR (x y : LFix) : Prop :=
@@
     (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
-\end{minted}
+\end{coqcode}
 From such initial algebras, we define \emph{catamorphisms} by recursively
@@
 function from \coq{LFix F} to \coq{A}.
-\begin{minted}{Coq}
+\begin{coqcode}
 Definition cata_f `{setoid A} (g : Alg F A) : LFix -> A
@@
   end.
-\end{minted}
+\end{coqcode}
 It is easy to show that this function is a proper morphism, and we use this
 fact to encode them as a higher-order morphism, with the following type:
-\begin{minted}{Coq}
+\begin{coqcode}
 Definition cata `{eA : setoid A} : Alg F A ~> LFix ~> A := ...
-\end{minted}
+\end{coqcode}
 Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
@@
 \coq{cata g}.
-\begin{minted}{Coq}
+\begin{coqcode}
 Lemma cata_univ `{eA : setoid A} (g : Alg F A) (f : LFix ~> A)
     : f =e cata g <-> f =e g \o fmap f \o l_out.
-\end{minted}
+\end{coqcode}
 
@@
 Terminal coalgebras are defined analogously to initial algebras:
-\begin{minted}{Coq}
+\begin{coqcode}
 CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
-\end{minted}
+\end{coqcode}
 Their equivalence relation, and anamorphisms, and universal properties are
 easy to encode:
-\begin{minted}{Coq}
+\begin{coqcode}
   Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...
@@
     : f =e ana h <-> f =e g_in \o fmap f \o h.
-\end{minted}
+\end{coqcode}
 This definition, however, prevents us from composing catamorphisms and
@@
 $F$-algebras:
-\begin{minted}{Coq}
+\begin{coqcode}
 Inductive FinF : GFix F -> Prop :=
 | FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
-\end{minted}
+\end{coqcode}
 Catamorphisms from this initial algebra are denoted by \coq{ccata},
@@
 \coq{LFix F}. In particular: 
-\begin{minted}{Coq}
+\begin{coqcode}
 Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
 Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
-\end{minted}
+\end{coqcode}
 Using this new initial algebra, we can compose catamorphisms and anamorphisms,
@@
 \coq{h} can only be applied finitely many times:
-\begin{minted}{Coq}
+\begin{coqcode}
 Inductive RecF (h : Coalg F A) : A -> Prop :=
 | RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
-\end{minted}
+\end{coqcode}
 We show how this predicate is equivalent to the anamorphism of such coalgebras
 producing finite trees:
-\begin{minted}{Coq}
+\begin{coqcode}
 Lemma ana_rec_term `{setoid A} (h : Coalg F A)
   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
-\end{minted}
+\end{coqcode}
 Recursive anamorphisms are implemented in Coq recursively on the structure of
@@
 following:
-\begin{minted}{Coq}
+\begin{coqcode}
 Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
@@
        LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
-\end{minted}
+\end{coqcode}
 Here, \coq{RecF_inv} provides the structurally smaller element for
@@
 are recursive, and add a coercion from recursive coalgebras to coalgebras.
-\begin{minted}{Coq}
+\begin{coqcode}
 Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
-\end{minted}
+\end{coqcode}
 In some cases, proving that coalgebras can only be applied finitely many
@@
 necessarily recursive.
-\begin{minted}{Coq}
+\begin{coqcode}
 Definition respects_relation `{setoid A} (c : Coalg F A)
@@
   (c : Coalg F A) (RR : respects_relation c m R) : RecP c.
-\end{minted}
+\end{coqcode}
 In the above definition, \coq{m} is used as a form of ``metric''
@@
 where \coq{h} is a recursive coalgebra.
-\begin{minted}{Coq}
+\begin{coqcode}
 Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F :=...
@@
     : f =e rana h <-> f =e l_in \o fmap f \o h.
-\end{minted}
+\end{coqcode}
 
@@
 coalgebra, \textbf{recursive hylomorphisms}.
-\begin{minted}{Coq}
+\begin{coqcode}
 Definition hylo_def (a : Alg F B) (c : Coalg F A) 
@@
      end (RecF_inv H).
-\end{minted}
+\end{coqcode}
 The definition of recursive hylomorphisms is recursive on the proof that the
@@
 built the corresponding higher-order proper morphism:
-\begin{minted}{Coq}
+\begin{coqcode}
 Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
-\end{minted}
+\end{coqcode}
 Finally, we show that recursive hylomorphisms are the unique solution to
 the hylomorphism equation.
-\begin{minted}{Coq}
+\begin{coqcode}
 Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
     : f =e hylo g h <-> f =e g \o fmap f \o h.
-\end{minted}
+\end{coqcode}
 This result can be used to prove hylomorphism laws in a way that is analogous
@@
 hylomorphism fusion as an example:
-\begin{minted}{Coq}
+\begin{coqcode}
 Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) 
@@
 Qed.
-\end{minted}
+\end{coqcode}
 Using the hylo fusion law, we can prove the well-known \emph{deforestation
@@
 hylomorphism is the composition of a catamorphism and a recursive anamorphism.
-\begin{minted}{Coq}
+\begin{coqcode}
 Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
@@
 : cata g \o rana h =e hylo g h.
-\end{minted}
+\end{coqcode}
 This result will allow us to \emph{calculate} programs by providing small, easy

### 2024-02-27 paper/paper.tex
Stuff

@@
 \coq{A ~> B} will be extracted to the OCaml equivalent to
-\coq{A -> B}.
+\coq{A -> B}. Furthermore, we add a coercion from \coq{A ~> B} to \coq{A -> B}.
+On top of this, we define basic function composition and identity functions:
+\begin{coqcode}
+Notation "f \o g" = (comp f g).
+Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
+Definition id : A ~> A := ...
+\end{coqcode}
+
+Using custom equivalences and proper morphisms, we redefine the definitions of
+container extensions and container equality. In particular, container
+extensions require a proper morphism to check the validity of positions in
+shapes, and container equality now uses equivalences of shapes and contained
+elements: 
+\begin{coqcode}
+Class Cont (Shape : Type) (Pos : Type) := { valid : Shape * Pos ~> bool }.
+Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
+  | AppR_ext (Es : shape x =e shape y)
+             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
+\end{coqcode}
+Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
+however, we chose to use Coq's propositional equality, since these would leads
+to simpler code. We explain why in Section~\ref{sec:cata}, and the
+mechanisation of initial algebras of container extensions. The definitions in
+Figure~\ref{fig:poly} are redefined to account for these changes in a
+straightforward way, i.e. we prove that validity predicates are indeed proper
+morphisms.  \dcas{Should we show, instead, the actual full definition? It
+becomes a bit more complicated ...}
+
+This definition leads to the well-known ``setoid hell'', which we mitigate by
+providing tactics and notations to automatically discharge proofs of
+\coq{app_eq} for morphisms, whenever the types use the standard propositional
+equality, or a combination of propositional and extensional equality. However,
+our compositional approach allows to build morphisms by plugging in other
+morphisms to our combinator, In our framework, our goal is that the
+user-provided functions remain small, with relatively straightforward proofs of
+\coq{app_eq}.
+
+However, by using this mechanisation, we gain simplified proofs via the use of
+Coq's \emph{Generalised Rewriting}.  Since every morphism \coq{f : A ~> B}
+satisfies that if \coq{x =e y}, then \coq{f x =e f y}, we can add every
+morphism as a proper element of Coq's respectfulness relation. In practice,
+this means that we can use the \coq{rewrite} tactic on proofs of type 
+\coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
+arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},
+and \coq{transitivity} tactics. For example:
+\begin{coqcode}
+Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D), 
+  h \o g =e id -> h \o (g \o f) =e f.
+Proof.
+  intros H.  rewrite compA H.  reflexivity.
+Qed.
+\end{coqcode}
 
@@
 extraction mechanism to inline as many of these small definitions as possible.
-
-\subsubsection{Isomorphisms of Coq Types and Container Extensions}
-
-Using custom equivalences and proper morphisms, we redefine the definitions of
-container extensions and container equality. In particular, container
-extensions require a proper morphism to check the validity of positions in
-shapes, and container equality now uses equivalences of shapes and contained
-elements: 
-\begin{coqcode}
-Class Cont (Shape : Type) (Pos : Type) := { valid : Shape ~> Pos ~> bool }.
-Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
-  | AppR_ext (Es : shape x =e shape y)
-             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
-\end{coqcode}
-Note the use of \coq{=e} instead of Coq's standard equality.  For
-positions, however, we chose to use Coq's propositional equality, since these
-would leads to simpler code. We explain why in Section~\ref{sec:cata}, and the
-mechanisation of initial algebras of container extensions. The definitions in
-Figure~\ref{fig:poly} are redefined to account for these changes in a
-straightforward way.
 
@@
 extension. Given \coq{p_inl}/\coq{p_inr} that act as
-\coq{inl/inr} but on valid product positions, the relevant
+\coq{inl/inr} on valid product positions, the relevant
 functions that define the isomorphism are:
@@
 using the fact that \coq{Prod} containers only have two valid positions.
-
-\subsubsection{Generalised Rewriting for Morphisms}
-
-Since our framework only allows types with equivalences, and proper morphisms,
-we can leverage Coq's \emph{setoid rewrite}
-
-\dcas{Blabla: examples, no need to explain setoid rewrite in detail here}
 

### 2024-02-27 paper/paper.tex
I did some minor rewriting

--- 2024-02-27 paper/paper.tex
@@
 to a container extension is called a \emph{container functor}. However, in our
-work we will call \emph{container functor} to the functors defined by a specific
+work we will call \emph{container functor} the functors defined by a specific
 container extension.
@@
 \end{coqcode}
-This approach solves the the problem of the unsafe casts, since the code
+This approach solves the problem of the unsafe casts, since the code
 extracted for \coq{{p | Valid shape pos}} will be an OCaml singleton
@@
 
-However, there still remains the issue of reasoning about the equality of
+Additionaly we would like to reason about equality of
 container extensions. The contents of the container extensions need to be
-accessed via a function \coq{contents}. The domain of this function
-depends on the shape of the container. This means that, for a given type of
-shapes and positions, the equality of two objects of a container extension will
-need to deal with heterogeneous equality. Since we want to remain axiom-free,
-one way to deal with this is to rely on a \emph{decidable} validity predicate.
-This way, we reach our final definition for container extensions. First,
-containers are defined as pairs of shape and position types that have a
+accessed via a function \coq{contents} whose domain depends on the shape of the container.
+This means that for a given type of shapes and positions, the equality of two
+objects of a container extension will need to deal with heterogeneous equality.
+Since we want to remain axiom-free, one way to deal with this is to rely on a
+\emph{decidable} validity predicate.
+
+Finally we can define the type of containers we need for performing Coq
+extraction correctly. First, containers are defined as pairs of shape and
+position types that have a
 \emph{decidable} validity predicate. We represent decidable predicates as
-functions that return a boolean, and lift booleans to propositions similarly to
-SSReflect. For simplicity, we define the dependent pair of valid positions as a
-record, which will be extracted to a singleton type with the validity proof
-erased by Coq's code extraction mechanism.
+functions that return a boolean and lift booleans to propositions similarly to the implementation provided by
+SSReflect~\cite{GonthierL09}.
+For simplicity, we define the dependent pair of valid positions as a record,
+which will be extracted to a singleton type with the validity proof erased by
+Coq's code extraction mechanism.
 \begin{coqcode}
@@
 \end{coqcode}
-Then, we define \emph{container extensions} as
+Then, we define \emph{container extensions} as follows:
 \begin{coqcode}

### 2024-02-27 paper/paper.tex
Hylo stuff

@@
 \begin{coqcode}
-Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
+Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
 \end{coqcode}
@@
 only way to do this is to define such equivalence simultaneously to the
-definition for  \coq{LFix}.  We could have overcame this
-circularity by defining  \coq{LFix} independently of container
-extensions, and define later the necessary
-\coq{in}/\coq{out} functions to convert between
-\coq{LFix} and \coq{App}. However, this extra layer
-interferes with code extraction, in that it requires destructing/constructing
-subsequent \coq{LFix} and  \coq{App}, which Coq is not
-always able to inline.
-
-We provide a custom induction scheme for initial algebras:
-\begin{coqcode}
-Lemma LFix_ind [P : LFix -> Prop]
-  : (forall x : App F LFix,
-        (forall e : Pos (shape x), P (cont x e)) ->
-        P (LFix_in x))
-    -> forall l : LFix, P l.
-\end{coqcode}
+definition for  \coq{LFix}. Such definition would be rejected by Coq.
+There are other approaches that can circumvent this problem, by separating
+the definition of least fixed points from the proofs that the functions that
+get the contained elements are proper morphisms. But this leads to longer,
+unnecessarily complicated proofs.  Furthermore, this extra layer interferes
+with code extraction, in that it requires destructing/constructing subsequent
+\coq{LFix} and  \coq{App}, and Coq may not always able to inline them.
+
+%%\dcas{Boring stuff}
+%The definition of \coq{LFix} in terms of \coq{App} requires us to define our
+%own custom induction scheme for initial algebras:
+%\begin{coqcode}
+%Lemma LFix_ind [P : LFix -> Prop]
+%  : (forall x : App F LFix,
+%        (forall e : Pos (shape x), P (cont x e)) ->
+%        P (LFix_in x))
+%    -> forall l : LFix, P l.
+%\end{coqcode}
 We define the equality of initial algebras of container extensions recursively,
@@
 \end{coqcode}
-From such initial algebras, we define \emph{catamorphisms} by recursively
-traversing their structure. As usual, a catamorphism takes an
-\coq{F}-algebra with carrier \coq{A}, and produces a
-function from \coq{LFix F} to \coq{A}.
-\begin{coqcode}
-Definition cata_f `{setoid A} (g : Alg F A) : LFix -> A
+
+Least fixed points of container extensions come with two functions that are
+defined as wrappers around constructor \coq{LFix_in} and projection
+\coq{LFix_out}.
+\begin{coqcode}
+l_in : App F (LFix F) ~> LFix F
+l_out : LFix F ~> App F (LFix F)
+\end{coqcode}
+
+We define \emph{catamorphisms} by recursively traversing their structure of the
+input \coq{LFix}. As usual, a catamorphism takes an \coq{F}-algebra with
+carrier \coq{A}, and produces a function from \coq{LFix F} to \coq{A}.
+\begin{coqcode}
+Definition cata_f (g : Alg F A) : LFix -> A
 := fix f (x : LFix) :=
@@
 \begin{coqcode}
-Definition cata `{eA : setoid A} : Alg F A ~> LFix ~> A := ...
+Definition cata : Alg F A ~> LFix ~> A := ...
 \end{coqcode}
@@
 
-Terminal coalgebras are defined analogously to initial algebras:
-\begin{coqcode}
+Terminal coalgebras are defined analogously to initial algebras (see Figure~\ref{fig:term_coalg}).
+\begin{figure}
+\begin{coqcode}
+Context `(F : Cont Sh Po).
 CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
-\end{coqcode}
+Definition GFixR : GFix -> GFix -> Prop :=...
+Instance GFix_setoid : setoid GFix := ...GFixR...
+Definition g_in : App F GFix ~> GFix := ...
+Definition g_out : GFix ~> App F GFix := ...
+\end{coqcode}
+\caption{Terminal Coalgebras}\label{fig:term_coalg}
+\end{figure}
 Their equivalence relation, and anamorphisms, and universal properties are
-easy to encode:
+easy to encode, in an analogous way to catamorphisms and initial algebras:
 \begin{coqcode}
@@
 This definition, however, prevents us from composing catamorphisms and
-anamorphisms. To address this issue, we consider another alternative definition
+anamorphisms: we cannot mix \coq{GFix} and \coq{LFix}.
+To address this issue, we consider another alternative definition
 for initial $F$-algebras. In particular, we show that the terminal coalgebras
@@
 the supplementary material and Coq code to see how we encode this. In this
-paper, however, we focus on the fact that, anamorphisms produce finite trees,
-then their return type can also be the \emph{least} fixed point of the
+paper, however, we focus on the fact that, if anamorphisms produce finite
+trees, then their return type can also be the \emph{least} fixed point of the
 container extension. We encode in Coq the notion of \emph{recursive
@@
 coalgebras are coalgebras whose anamorphism produces finite trees.  The
-following predicate captures this fact. The following predicate states that
+following predicate captures this fact, \coq{h} is recursive if
 \coq{h} can only be applied finitely many times:
@@
 \end{coqcode}
-We show how this predicate is equivalent to the anamorphism of such coalgebras
-producing finite trees:
+This predicate is equivalent to the proposition that the anamorphism of such
+coalgebras produce only finite trees:
 \begin{coqcode}
@@
 Recursive anamorphisms are implemented in Coq recursively on the structure of
-the proof that they can only be applied finitely many times
-(\coq{RecF}).  Note that such proofs will be erased when extracting
-code to OCaml. This is standard practice when building general recursive
-functions in Coq\footnote{\dcas{This technique had a name in the community, but
-I cannot remember. Something predicates}}. The core definitions are the
-following:
+the proof that they can only be applied finitely many times (\coq{RecF}).  Note
+that such proofs will be erased when extracting code to OCaml, and this is
+standard practice when building general recursive functions in
+Coq\footnote{\dcas{This technique had a name in the community, but I cannot
+remember. Something predicates}}. The core definitions are the following:
 \begin{coqcode}
@@
 do recursion, and there is no corecursive definition that is guaranteed to be
-guarded, due to the use of an arbitrary algebra $a$. Furthermore, solutions to
+syntactically guarded (as required by Coq), due to the use of an arbitrary
+algebra $a$. Furthermore, solutions to
 the hylomorphism equation are \emph{not} guaranteed to be unique, or to even
@@
 If $c$ is a recursive coalgebra, then the equation $f = a \circ F\; f \circ c$
-does have a unique solution.
-We show this by first providing the definition
-for hylomorphisms in Coq, and then proving that any function satisfying this
-equation is a hylomorphism. We call these hylomorphisms with a recursive
-coalgebra, \textbf{recursive hylomorphisms}.
+does have a unique solution.  We show this by first providing the definition
+for recursive hylomorphisms in Coq, and then proving that any function
+satisfying this equation is a hylomorphism. We call these hylomorphisms with a
+recursive coalgebra, \textbf{recursive hylomorphisms}.
 \begin{coqcode}
@@
 Proof.
-  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
-  rewrite hylo_univ.
-  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
-  rewrite fmap_comp.
-  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
-  rewrite compA.
-  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
-  rewrite <- E2.
-  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
-  rewrite <- compA.
-  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
-  rewrite <- compA.
-  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
-  rewrite (compA g1).
-  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
-  rewrite <- hylo_unr.
-  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
+  rewrite hylo_univ, fmap_comp compA <- E2 <- compA <- compA (compA g1) <- hylo_unr
   reflexivity.
 Qed.
+  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
+  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
+  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
+  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
+  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
+  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
+  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
+  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
+  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
 \end{coqcode}

### 2024-02-28 paper/paper.tex
Stuff
Abreu et al POP2023 extraction

@@
 \begin{coqcode}
-Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) 
-  (g2 : Alg F C) (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
-  : f2 \o hylo g1 h1 =e hylo g2 h1.
-Proof.
-  rewrite hylo_univ, fmap_comp compA <- E2 <- compA <- compA (compA g1) <- hylo_unr
-  reflexivity.
+Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C) 
+  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
+
+  rewrite hylo_univ, fmap_comp, compA, <- E2, 
+          <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
 Qed.
@@
 
+This section covers how to use our approach to calculate recursive programs
+from simple, non-recursive specifications. The steps are the following:
+\begin{enumerate}
+  \item Write a specification as a morphism \coq{f : A ~> B}.
+  \item Use Coq's generalised rewriting and the laws of morphisms to 
+    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
+  \item Project the function \coq{h : A -> B} from \coq{g}
+    (i.e. \coq{h = app g}).
+  \item Extract \coq{h} to OCaml.
+\end{enumerate}
+Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
+\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
+proofs from the code, and inline our definitions. To ease the application of
+these steps, we provide the following helper type:
+\begin{coqcode}
+  Record Ext (f : A ~> B) := 
+    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
+\end{coqcode}
+Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
+since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
+will be the function \coq{target} that is extensionally equal to \coq{f}.
+
+\subsection{Sorting Algorithms}
 \dcas{Talk About this}
@@
 \end{minted}
+
+\subsection{Knapsack}
+
+\subsection{Harper's Regular-Expression Matcher}
 

### 2024-02-28 paper/paper.tex
working on Section 3.

--- 2024-02-27 paper/paper.tex
@@
 
-In order to abstract recursion patters we need to be able to talk about functors
-which have the suitable fixed-point properties, i.e. those who have an initial
-algebra. Instances of these functors are the polynomial ones.  A common approach
-to construct such functors is to use \emph{container}~\cite{AbbottAG05}.  In
-this section we introduce the basics of polynomial functors as containers
-(Section~\ref{sec:containers}).  However, reasoning about containers equality
-will require us to consider both functional extensionality and heterogeneous
-equality. As stated in previous sections~\ref{sec:intro}, this is rather
-problematic since a stric requirement for this work is that we should remain
-axiom-free. For this purpose in Section~\ref{sec:types-morphisms} we introduce
-a custom equivalence relation on types in order to avoid using the functional
-extensionality axiom.
+In order to abstract recursion patters we need to be able to abstract away from
+the particular shape of the data.  This is achieved by introducing the concept
+of \emph{functors} which have the suitable fixed-point properties, i.e. those
+who have a initial algebras. A common approach to construct such functors is to
+use \emph{container}~\cite{AbbottAG05} which is what we introduce in this
+Section~\ref{sec:containers}.  However, reasoning about containers equality will
+require us to consider both functional extensionality and heterogeneous
+equality. To avoid this in Section~\ref{sec:types-morphisms} we introduce a
+custom equivalence relation on types.
 
@@
 \label{sec:containers}
-A container is defined by a type of \emph{shapes} and a \emph{family of position
-types} indexed by shapes.
+To abstract away the shape of a particular data structure we use functors.
+These are functions \coq{F : Type -> Type} which additionally have a
+map \coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
+functor represents a container for abstract data that can be manipulated without
+changing the outer structure. For example, the type \coq{List : Type -> Type} is
+a functor and \coq{List A} is the type of lists over elements of type \coq{A}
+with the obvious \coq{fmap} function recursive over a list and applying the
+function \coq{A -> B} for each element of the list.
+Additionally the type of lists arises as the least fixed-point for the
+functor \coq{F X = unit + (A * X) }.
+
+In general not every functor has a fixed-point, so we are going to restrict
+ourselves to polynomial functors as represented by containers.  A container is
+defined by a type of \emph{shapes} and a \emph{family of position types} indexed
+by shapes.
 \begin{coqcode}
@@
 \end{coqcode}
-An \emph{extension} of this container is a functor defined as follows:
+In our previuos example, we would define \coq{Shape} as \coq{unit + A}
+indicating there are two constructors in the data type and one of these contains
+a piece of data of type \coq{A}. Moreover, we would define the positions
+as \coq{Pos (inl tt) = Empty_set} and \coq{Pos (inr a) = unit} indicating that
+the first constructor does not have any recursive type variables and the second
+constructor has one type variable.
+
+Therefore an \emph{extension} of this container is a functor defined as follows:
 \begin{coqcode}
@@
 \end{coqcode}
-with the obvious action on morphisms. Generally, any functor that is isomorphic
+with the obvious action on morphisms. Note that for \coq{Pos (inl tt)} we have a
+type of functions \coq{Empty_set -> X} which is only inhabited by the trivial
+map from the empty type and for \coq{Pos (inr a)} we have the type of
+functions \coq{unit -> X} which has cardinality \coq{unit}.
+
+Generally, any functor that is isomorphic
 to a container extension is called a \emph{container functor}. However, in our
@@
 
-Finally we can define the type of containers we need for performing Coq
-extraction correctly. First, containers are defined as pairs of shape and
-position types that have a
-\emph{decidable} validity predicate. We represent decidable predicates as
-functions that return a boolean and lift booleans to propositions similarly to the implementation provided by
-SSReflect~\cite{GonthierL09}.
-For simplicity, we define the dependent pair of valid positions as a record,
-which will be extracted to a singleton type with the validity proof erased by
-Coq's code extraction mechanism.
+Finally we can define the type of containers as follows:
 \begin{coqcode}
@@
 \end{coqcode}
+where \coq{valid} is the First, containers are defined as pairs of shape and
+position types that have a \emph{decidable} validity predicate. We represent
+decidable predicates as functions that return a boolean and lift booleans to
+propositions similarly to the implementation provided by
+SSReflect~\cite{GonthierL09}.  For simplicity, we define the dependent pair of
+valid positions as a record, which will be extracted to a singleton type with
+the validity proof erased by Coq's code extraction mechanism.
+
 Then, we define \emph{container extensions} as follows:

### 2024-02-28 paper/paper.tex
Examples

@@
 In general not every functor has a fixed-point, so we are going to restrict
-ourselves to polynomial functors as represented by containers.  A container is
-defined by a type of \emph{shapes} and a \emph{family of position types} indexed
-by shapes.
+ourselves to \emph{polynomial functors} as represented by containers.  A
+container is defined by a type of \emph{shapes} and a \emph{family of position
+types} indexed by shapes.
 \begin{coqcode}
@@
 \end{coqcode}
-In our previuos example, we would define \coq{Shape} as \coq{unit + A}
-indicating there are two constructors in the data type and one of these contains
-a piece of data of type \coq{A}. Moreover, we would define the positions
-as \coq{Pos (inl tt) = Empty_set} and \coq{Pos (inr a) = unit} indicating that
-the first constructor does not have any recursive type variables and the second
-constructor has one type variable.
-
-Therefore an \emph{extension} of this container is a functor defined as follows:
+For the functor \coq{F} we introduced in the previous paragraph we would
+define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
+the data type and one of these contains a piece of data of type \coq{A}.
+Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
+and \coq{Pos (inr a) = unit} indicating that the first constructor does not have
+any recursive type variables and the second constructor has one type variable.
+
+At this point an \emph{extension} of this container is a functor defined as
+follows:
 \begin{coqcode}
@@
 \end{coqcode}
-with the obvious action on morphisms. Note that for \coq{Pos (inl tt)} we have a
-type of functions \coq{Empty_set -> X} which is only inhabited by the trivial
-map from the empty type and for \coq{Pos (inr a)} we have the type of
-functions \coq{unit -> X} which has cardinality \coq{unit}.
-
-Generally, any functor that is isomorphic
-to a container extension is called a \emph{container functor}. However, in our
-work we will call \emph{container functor} the functors defined by a specific
-container extension.
-
-The first problem in extracting Coq code from such container extensions is the
-use of families of types. Since OCaml's type system cannot accurately capture
-such families of types, Coq's code extraction mechanism will insert unsafe
-casts. To avoid unsafe casts, we can consider representing the positions of a
-container as a type that captures \emph{all} the potential positions in all
-possible shapes, together with a predicate \coq{Valid} that
-determines whether a position is valid in a shape.
+with the obvious action on morphisms given by post-composition
+with \coq{contents}.  In our running example, the type \coq{App X}, for
+some \coq{X} has two inhabitants.  The first is a pair composed
+by \coq{inl tt : Shape} and a function \coq{Pos (inl tt) -> X}. This latter is
+in fact a function of type we \coq{Empty_set -> X}  by definitions which in turn
+is isomorphic to the type \coq{unit}, the type with only one inhabitant.  The
+second inhabitant is the pair composed by \coq{inr a : Shape}, for
+some \coq{a : A}, and a function \coq{Pos (inr a) -> X}.  This latter type is
+equal to \coq{unit -> X} which represents the elements of \coq{X} and therefore
+is isomorphic to \coq{X}.  This particular pair of shapes and positions is
+isomorphic to the type \coq{unit + A * X}.
+
+The correspondence between containers and polynomial functors has been
+formalised in this work and has been moved to the
+Appendix~\ref{sec:pol-containers} for space reasons. Generally, any functor that
+is isomorphic to a container extension is called a \emph{container functor}.
+Here we will call \emph{container functor} the functors defined
+by a specific container extension.
+
+\subsection{Extractable Containers}
+In the previous section we defined containers using dependent types.  However,
+Ocaml's type system is not equipped to handle these and to solve this problem
+Coq's code extraction mechanism will insert unsafe casts.
+
+Here consider instead representing the positions of a
+container using a validity predicate which assigns shapes to positions.
+A container extension is then defined as follows:
 \begin{coqcode}
@@
 \end{coqcode}
-This approach solves the problem of the unsafe casts, since the code
-extracted for \coq{{p | Valid shape pos}} will be an OCaml singleton
-type defined as the OCaml equivalent to \coq{Pos}.
-
-Additionaly we would like to reason about equality of
-container extensions. The contents of the container extensions need to be
-accessed via a function \coq{contents} whose domain depends on the shape of the container.
+This approach solves the problem of the unsafe casts, since the code extracted
+for \coq{{p | Valid shape pos}} will be an OCaml singleton type defined as the
+OCaml equivalent to \coq{Pos}.
+
+Additionaly we would like to reason about equality of container extensions. The
+contents of the container extensions need to be accessed via a
+function \coq{contents} whose domain depends on the shape of the container.
 This means that for a given type of shapes and positions, the equality of two
@@
 \end{coqcode}
-
-\subsubsection{Polynomial Functors as Container Extensions}
-Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
-omit here the proofs that these container extensions are isomorphic to
-polynomials, but this proof is straightforward. To illustrate how the identity,
-constant, sum and product container extensions work, consider the
-tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
-$F\; X = \mathbb{N} + X \times X$.
-
-This functor is built as the sum of two functors: $X \mapsto \mathbb{N}$ and $X \mapsto
-X \times X$. Consider first $X \mapsto \mathbb{N}$. This is a constant functor
-that maps $X$ to the type $\mathbb{N}$. More generally, a functor that, given a
-type $A$, maps everything to it, $X \mapsto A$, only needs one shape ($A$, L9
-of Fig~\ref{fig:poly}), and zero positions (In Coq, \coq{False},
-L10 of Fig~\ref{fig:poly}). The reason why there are zero valid positions in
-a constant container extension is because, given something of type 
-\coq{App (Const (Ks A) Kp) X}, it should be impossible to get
-something of type \coq{X}. But from the container shape, we will
-always be able to get the element of type $A$\footnote{%
-  Our decision to use \coq{False} is arbitrary. In general, we
-  could have chosen any time, with a validity predicate that always returns
-  \coq{false}. It would make no difference for code extraction.
-}. This functor is, therefore \coq{Const nat}.
-
-Consider now $X \mapsto X \times X$. This is the product of two
-identity functors $X \mapsto X$. The identity functor only has one shape, and
-we can use the unit type to represent such shapes. From this shape we should be
-able to get an element of type $X$, and we can capture this by using the unit
-type for positions. To build an element of the container extension type, we
-will need to provide a function of type \coq{unit -> X}, which is
-isomorphic to \coq{X}.
-
-The shape of the product of two container extensions is the product of their
-shapes. The product of such container extensions should allow us to get either
-the left, or the right element in the product. Therefore, the positions of this
-container extension is the sum of the positions of the two container
-extensions. What is a valid position in a given shape? If the position is
-\coq{inl p} (or \coq{inr p}), then it is only valid if
-\coq{p} is valid in the first (resp. second) component of the
-shape. The product of our two container extensions is
-\coq{Prod Id Id}.
-
-Finally, the shape the sum of two container extensions is the sum of their
-shapes. The positions are, like for products, the sum of their shapes. But to
-check the validity of shapes in positions, their constructor must match, e.g.
-shape \coq{inl s} and shape \coq{inl p}, with
-\coq{valid s p}.
-
-Our final container extension is the following: 
-\coq{Sum (Const nat) (Prod Id Id)}.
-The shape of this functor is: 
-\coq{nat + unit * unit}. The positions of this functor
-are \mintinline[breaklines]{Coq}{False + (unit + unit)}, which is isomorphic
-to \coq{unit + unit}, which will only be valid in the RHS.
-
-The Coq value \coq{inl n} of type
-\coq{nat + bool * bool} would look as follows as a container
-extension 
-\coq{App (Sum (Const nat) (Prod Id Id)) bool}:
-\begin{coqcode}
-{| shape := inl n; cont := fun e => uninhabited e|}
-\end{coqcode}
-We omit the definition of \coq{uninhabited}, which exploits the
-fact that \coq{e} for \coq{inl n} implies
-\coq{False} to build an element of type \coq{bool}%
-\footnote{Note that, since \coq{e} is uninhabited we could return
-any arbitrary value instead of \coq{uninhabited e}.}.
-
-Similarly, the Coq value \coq{inr (true, false)} as a container
-extension is as follows:
-\begin{coqcode}
-  {| shape := inr (tt,tt);
-     cont := fun e : {p | valid (inr (tt, tt)) p} =>
-        match proj1_sig e with
-        | inl f => match f with end
-        | inr (inl tt) => fun _ => true
-        | inr (inr tt) => fun _ => false
-        end
-  |}
-\end{coqcode}
-Here, the position \coq{inl f} is impossible, as
-\coq{f} would need to be a proof of \coq{False}.
-Of the two remaining positions, the left one will return 
-\coq{true}, and the right position \coq{false}.
-\begin{figure}
-\begin{minipage}{.52\textwidth}
-  \begin{minted}[linenos, fontsize=\small]{Coq}
-Context `(C1 : Cont S1 P1) 
-        `(C2 : Cont S2 P2).
-
-Definition Is := unit.
-Definition Ip := unit.
-Instance Id : Cont Is Ip := {
-  valid := fun _ _ => true }.
-
-Definition Ks (A : Type) := A.
-Definition Kp := False.
-Instance Const A : Cont (Ks A) Kp := {
-  valid := fun _ p =>
-    match p with end }.
-
-Definition Ps : Type := S1 * S2.
-Definition Pp : Type := P1 + P2.
-Instance Prod : Cont Ps Pp := {
-  valid := fun s p => 
-    match p with
-    | inl p' => valid (fst s) p'
-    | inr p' => valid (snd s) p'
-    end }.
-\end{minted}
-\end{minipage}
-\begin{minipage}{.47\textwidth}
-  \begin{minted}[linenos, firstnumber=23, fontsize=\small]{Coq}
-Definition Ss : Type := S1 + S2.
-Definition Sp : Type := P1 + P2.
-Instance Sum : Cont Ss Sp := {
-  valid := fun s p => 
-    match s, p with
-    | inl s', inl p' => valid s' p'
-    | inr s', inr p' => valid s' p'
-    | _, _ => false
-    end }.
-
-Definition Cs : Type := App F S2.
-Definition Cp : Type := P1 * P2.
-Instance Comp : Cont Cs Cp := 
-{ valid :=  
-    fun (s : App F S2) (p : P1 * P2) =>
-        match valid (shape s) (fst p) 
-        with
-        | true  => fun H =>
-            valid (cont s {fst p | H}) 
-                  (snd p)
-        | false => fun _ => false
-        end eq_refl }.
-  \end{minted}
-\end{minipage}
-  \caption{Polynomials as Container Extensions (simplified). In order: identity
-  (L4-7), constant (L9-13), product (L15-22), sum (L23-31), composition
-  (L33-44).}
-  \label{fig:poly}
-\end{figure}
 
@@
 
-\section{Extracting Certified Hylomorphisms}
+\section{Certified Hylomorphisms}
 
@@
 \subsection{Harper's Regular-Expression Matcher}
+
+\subsection{Shortcut Deforestation}
+
+\dcas{foldr/build examples from original paper (or hylo papers)}
 
@@
 \appendix
-\subsection{Course-of-Value Recursion with Finite Memory Table}
+\section{Polynomial Functors as Container Extensions}
+\label{sec:pol-containers}
+Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
+omit here the proofs that these container extensions are isomorphic to
+polynomials, but this proof is straightforward. To illustrate how the identity,
+constant, sum and product container extensions work, consider the
+tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
+$F\; X = \mathbb{N} + X \times X$.
+
+This functor is built as the sum of two functors: $X \mapsto \mathbb{N}$ and $X \mapsto
+X \times X$. Consider first $X \mapsto \mathbb{N}$. This is a constant functor
+that maps $X$ to the type $\mathbb{N}$. More generally, a functor that, given a
+type $A$, maps everything to it, $X \mapsto A$, only needs one shape ($A$, L9
+of Fig~\ref{fig:poly}), and zero positions (In Coq, \coq{False},
+L10 of Fig~\ref{fig:poly}). The reason why there are zero valid positions in
+a constant container extension is because, given something of type
+\coq{App (Const (Ks A) Kp) X}, it should be impossible to get
+something of type \coq{X}. But from the container shape, we will
+always be able to get the element of type $A$\footnote{%
+  Our decision to use \coq{False} is arbitrary. In general, we
+  could have chosen any time, with a validity predicate that always returns
+  \coq{false}. It would make no difference for code extraction.
+}. This functor is, therefore \coq{Const nat}.
+
+Consider now $X \mapsto X \times X$. This is the product of two
+identity functors $X \mapsto X$. The identity functor only has one shape, and
+we can use the unit type to represent such shapes. From this shape we should be
+able to get an element of type $X$, and we can capture this by using the unit
+type for positions. To build an element of the container extension type, we
+will need to provide a function of type \coq{unit -> X}, which is
+isomorphic to \coq{X}.
+
+The shape of the product of two container extensions is the product of their
+shapes. The product of such container extensions should allow us to get either
+the left, or the right element in the product. Therefore, the positions of this
+container extension is the sum of the positions of the two container
+extensions. What is a valid position in a given shape? If the position is
+\coq{inl p} (or \coq{inr p}), then it is only valid if
+\coq{p} is valid in the first (resp. second) component of the
+shape. The product of our two container extensions is
+\coq{Prod Id Id}.
+
+Finally, the shape the sum of two container extensions is the sum of their
+shapes. The positions are, like for products, the sum of their shapes. But to
+check the validity of shapes in positions, their constructor must match, e.g.
+shape \coq{inl s} and shape \coq{inl p}, with
+\coq{valid s p}.
+
+Our final container extension is the following:
+\coq{Sum (Const nat) (Prod Id Id)}.
+The shape of this functor is:
+\coq{nat + unit * unit}. The positions of this functor
+are \mintinline[breaklines]{Coq}{False + (unit + unit)}, which is isomorphic
+to \coq{unit + unit}, which will only be valid in the RHS.
+
+The Coq value \coq{inl n} of type
+\coq{nat + bool * bool} would look as follows as a container
+extension
+\coq{App (Sum (Const nat) (Prod Id Id)) bool}:
+\begin{coqcode}
+{| shape := inl n; cont := fun e => uninhabited e|}
+\end{coqcode}
+We omit the definition of \coq{uninhabited}, which exploits the
+fact that \coq{e} for \coq{inl n} implies
+\coq{False} to build an element of type \coq{bool}%
+\footnote{Note that, since \coq{e} is uninhabited we could return
+any arbitrary value instead of \coq{uninhabited e}.}.
+
+Similarly, the Coq value \coq{inr (true, false)} as a container
+extension is as follows:
+\begin{coqcode}
+  {| shape := inr (tt,tt);
+     cont := fun e : {p | valid (inr (tt, tt)) p} =>
+        match proj1_sig e with
+        | inl f => match f with end
+        | inr (inl tt) => fun _ => true
+        | inr (inr tt) => fun _ => false
+        end
+  |}
+\end{coqcode}
+Here, the position \coq{inl f} is impossible, as
+\coq{f} would need to be a proof of \coq{False}.
+Of the two remaining positions, the left one will return
+\coq{true}, and the right position \coq{false}.
+\begin{figure}
+\begin{minipage}{.52\textwidth}
+  \begin{minted}[linenos, fontsize=\small]{Coq}
+Context `(C1 : Cont S1 P1)
+        `(C2 : Cont S2 P2).
+
+Definition Is := unit.
+Definition Ip := unit.
+Instance Id : Cont Is Ip := {
+  valid := fun _ _ => true }.
+
+Definition Ks (A : Type) := A.
+Definition Kp := False.
+Instance Const A : Cont (Ks A) Kp := {
+  valid := fun _ p =>
+    match p with end }.
+
+Definition Ps : Type := S1 * S2.
+Definition Pp : Type := P1 + P2.
+Instance Prod : Cont Ps Pp := {
+  valid := fun s p =>
+    match p with
+    | inl p' => valid (fst s) p'
+    | inr p' => valid (snd s) p'
+    end }.
+\end{minted}
+\end{minipage}
+\begin{minipage}{.47\textwidth}
+  \begin{minted}[linenos, firstnumber=23, fontsize=\small]{Coq}
+Definition Ss : Type := S1 + S2.
+Definition Sp : Type := P1 + P2.
+Instance Sum : Cont Ss Sp := {
+  valid := fun s p =>
+    match s, p with
+    | inl s', inl p' => valid s' p'
+    | inr s', inr p' => valid s' p'
+    | _, _ => false
+    end }.
+
+Definition Cs : Type := App F S2.
+Definition Cp : Type := P1 * P2.
+Instance Comp : Cont Cs Cp :=
+{ valid :=
+    fun (s : App F S2) (p : P1 * P2) =>
+        match valid (shape s) (fst p)
+        with
+        | true  => fun H =>
+            valid (cont s {fst p | H})
+                  (snd p)
+        | false => fun _ => false
+        end eq_refl }.
+  \end{minted}
+\end{minipage}
+  \caption{Polynomials as Container Extensions (simplified). In order: identity
+  (L4-7), constant (L9-13), product (L15-22), sum (L23-31), composition
+  (L33-44).}
+  \label{fig:poly}
+\end{figure}
+
+\section{Course-of-Value Recursion with Finite Memory Table}
 It can be proven that if $c$ is a recursive coalgebra, then the hylomorphism for

### 2024-02-28 paper/paper.tex
Some more writing in Section 3.

@@
 is isomorphic to \coq{X}.  This particular pair of shapes and positions is
-isomorphic to the type \coq{unit + A * X}.
-
-The correspondence between containers and polynomial functors has been
-formalised in this work and has been moved to the
-Appendix~\ref{sec:pol-containers} for space reasons. Generally, any functor that
-is isomorphic to a container extension is called a \emph{container functor}.
-Here we will call \emph{container functor} the functors defined
-by a specific container extension.
-
-\subsection{Extractable Containers}
-In the previous section we defined containers using dependent types.  However,
-Ocaml's type system is not equipped to handle these and to solve this problem
-Coq's code extraction mechanism will insert unsafe casts.
-
-Here consider instead representing the positions of a
-container using a validity predicate which assigns shapes to positions.
-A container extension is then defined as follows:
+isomorphic to the type \coq{unit + A * X}. The correspondence between containers
+and polynomial functors has been formalised in this work and has been moved to
+the Appendix~\ref{sec:pol-containers} for space reasons.
+
+Generally, any functor that is isomorphic to a container extension is called a
+\emph{container functor}. However, in our work we will call \emph{container
+functor} the functors defined by a specific container extension.
+
+The first problem in extracting Coq code from such container extensions is the
+use of families of types. Since OCaml's type system cannot accurately capture
+such families of types, Coq's code extraction mechanism will insert unsafe
+casts. To avoid unsafe casts, we can consider representing the positions of a
+container as a type that captures \emph{all} the potential positions in all
+possible shapes, together with a predicate \coq{Valid} that determines whether a
+position is valid in a shape.
 \begin{coqcode}
@@
 \subsection{Harper's Regular-Expression Matcher}
-
-\subsection{Shortcut Deforestation}
-
-\dcas{foldr/build examples from original paper (or hylo papers)}
 
@@
 \appendix
+\mpav{Dumped}
 \section{Polynomial Functors as Container Extensions}
-\label{sec:pol-containers}
 Polynomial functors can be defined as container extensions~\ref{fig:poly}. We

### 2024-02-28 paper/paper.tex
Working on Extractable Container Extensions
Intro

--- 2024-02-28 paper/paper.tex
@@
 \newmintinline[coq]{coq}{fontsize=\small}
+\newmintinline[ocaml]{OCaml}{fontsize=\small}
 \newminted{coq}{fontsize=\small}
@@
 halting problem. Modern proof assistants like Coq or Agda provide a sound, but
-incomplete algorithm which syntactically checks for termination or productivity.
-For recursion, this is done by trying to (automatically) infer which argument in
-the recursive call gets smaller with respects to the original input argument.
-For productivity, the algorithm checks that the (co)recursive call appears
-directly under a constructor to make sure that this function always produce at
-least one element after each recursive step.
+incomplete algorithm which syntactically checks for termination or
+productivity.  For recursion, this is done by trying to (automatically) infer
+which argument in the recursive call gets smaller with respects to the original
+input argument.  For productivity, the algorithm checks that the (co)recursive
+call appears directly under a constructor to make sure that this function
+always produce at least one element after each recursive step.
 
 As mentioned, some functions, though well-defined, cannot be accepted by the
-proof assistant. One example is the quicksort algorithm. This is given by the
-\haskell{partition} and \haskell{quicksort} function below (written in Haskell
-code):
-While \haskell{quicksort} is a well-defined mathematical function it cannot be
-accepted by a proof assistant.  The reason is that the \haskell{partition}
+proof assistant. One example is the quicksort algorith, written in OCaml below:
+\begin{minted}{OCaml}
+  let divide = function
+  |  [] -> None
+  |  x::xs -> Some (x, List.partition (fun y -> y < x) xs);;
+
+  let rec quicksort xs = match divide xs with
+    | None -> []
+    | Some (x, (smaller, larger)) -> quicksort smaller @ (x::quicksort larger)
+\end{minted}
+While \ocaml{quicksort} is a well-defined mathematical function it cannot be
+accepted by a proof assistant.  The reason is that the \ocaml{divide}
 function \emph{destructuring} the input ``dives'' deeper in the input returning
@@
 
-A pen-and-paper technique would be to show that the \haskell{partition} is a
-\emph{well-founded coalgebra} on the functor $(- \times A \times -)$ for a fixed
-type $A$. Here the term ``well-founded'' refers to the fact that the infinite
-iteration of this coalgebra will produce a \emph{finite} binary tree with nodes
-in $A$.
+The main approach for implementing nonstructural recursion in Coq is to use
+\emph{well-founded recursion}, where recursive definitions are coupled with
+\emph{termination proofs}. Using well-founded recursion, the recursive calls
+will happen on \emph{structurally smaller termination proofs}. A benefit of
+this approach is that, extracting verified nonstructural recursive functions to
+OCaml will erase the termination proofs, and so produce code that will be
+closer to what a programmer may have written directly in OCaml. However,
+reasoning about program equivalences requires dealing with such termination
+proofs, and it is common practice to use custom reduction lemmas that will be
+used extensively when proving properties  about the nonstructural recursive
+functions.
+
+A better approach for reasoning about program equivalences comes from
+\emph{structured recursion schemes}, and their associated algebraic laws.  In
+particular, \ocaml{quicksort} is an instance of a \emph{divide-and-conquer}
+algorithm which, in the functional programming literature is known as a
+hylomorphism~\cite{MeijerFP91, HuIT96}.  Hinze et al.~\cite{HinzeWG15} showed
+that every recursive program can be turned into an hylomorphism, which implies
+that, ``\emph{every recursion scheme can be formalised as an hylomorphism}''.
+Furthermore, the laws that are satisfied by hylomorphisms ease reasoning about
+program equivalences, since they can capture a number of useful equivalences,
+ranging from common optimisations such as \emph{short-cut
+fusion}~\cite{TakanoM95}, to (semi-)automatic
+parallelisations~\cite{Gibbons96:Third}.
+
+Despite the benefits of hylomorphisms for reasoning about program equivalences,
+not many approaches mechanise structured recursion schemes in proof assistants.
+A recent approach by Abreu et al.\~\cite{AbreuDHJMS23} encode an algebraic
+approach to divide-and-conquer computations, in which termination is entirely
+enforced by the typing discipline of the recursion scheme. However, although
+their approach solves the problem of termination proofs, as well as the
+performance of the code that is run \emph{within Coq}, it no longer allows the
+extraction of clean, naturally-looking code.  By naturally-looking and clean,
+we mean code that (1) preserves the recursive structure intended by the
+programmer; and (2) does not lead to unsafe casts \ocaml{Obj.magic} in the
+generated code.
+
+In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
+Although the use of \ocaml{Obj.magic} should be safe in the code
+extracted by Coq, it has been proven that, for higher-order programs, simple
+interoperations can lead to incorrect behaviour or even
+segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
+\ocaml{Obj.magic} means that the extracted code will have stronger
+safety guarantees, and will allow the safe interoperation with other OCaml
+code.
+
+This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
+well as their algebraic laws that: (1) is \textbf{fully axiom-free}; and (2)
+allows the extraction of naturally-looking, clean OCaml code. While programmers
+still need to reason about the termination of their programs, our mechanisation
+will allow the use of the algebraic laws of hylomorphisms for program
+reasoning, as well as extracting to clean, naturally-looking code. As an
+example our mechanisation of \ocaml{quicksort} as a hylomorphism will produce
+the following code:
+\begin{coqcode}
+let rec qsort = function
+  | [] -> []
+  | h :: t ->
+    let (l, r) = partition (fun x0 -> (<=) x0 h) t in
+    let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
+    app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
+\end{coqcode}
+
+The contributions we make in this paper are as follows:
+\begin{itemize}
+  \item We provide a framework for \emph{generic programming} with recursion
+    schemes in Coq, with a strong focus on proving their algebraic laws and 
+    program equivalences, and clean code extraction.
+  \item We formalise algorithms for divide and conquer, dynamic programming and
+    mutual recursion.
+  \item We verified the well-known short-cut fusion optimisation in Coq and
+    extract the resulting code to OCaml.
+\end{itemize}
+
+\dcas{Rewritten up to here}
+In
+\ocaml{quicksort} above, a pen-and-paper technique to prove termination would
+be to show that \ocaml{divide} is a \emph{well-founded coalgebra} on the
+functor $(1 + A \times - \times -)$ for a fixed type $A$. Here the term
+``well-founded'' refers to the fact that the infinite iteration of this
+coalgebra will produce a \emph{finite} binary tree with nodes in $A$.
 
@@
 hylomorphisms should be extracted to recursive OCaml functions that use no
-instances of \mintinline{OCaml}{Obj.magic}.
+instances of \ocaml{Obj.magic}.
 \end{enumerate}
-In OCaml, \mintinline{OCaml}{Obj.magic} allows for unsafe casts between types.
-Although the use of \mintinline{OCaml}{Obj.magic} should be safe in the code
+In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
+Although the use of \ocaml{Obj.magic} should be safe in the code
 extracted by Coq, it has been proven that, for higher-order programs, simple
@@
 segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
-\mintinline{OCaml}{Obj.magic} means that the extracted code will have stronger
+\ocaml{Obj.magic} means that the extracted code will have stronger
 safety guarantees, and will allow the safe interoperation with other OCaml code.

### 2024-02-28 paper/paper.tex
Extractable Containers

--- 2024-02-28 paper/paper.tex
@@
 is isomorphic to a container extension is called a \emph{container functor}.
-Here we will call \emph{container functor} the functors defined
-by a specific container extension.
+Here we will call \emph{container functor} the functors defined by a specific
+container extension.
 
@@
 
-Here consider instead representing the positions of a
-container using a validity predicate which assigns shapes to positions.
-A container extension is then defined as follows:
-\begin{coqcode}
-Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
-Record App (X : Type) := { shape    : Shape; 
-                          contents : {p | Valid shape p} -> X }.
-\end{coqcode}
-This approach solves the problem of the unsafe casts, since the code extracted
-for \coq{{p | Valid shape pos}} will be an OCaml singleton type defined as the
-OCaml equivalent to \coq{Pos}.
-
-Additionaly we would like to reason about equality of container extensions. The
-contents of the container extensions need to be accessed via a
-function \coq{contents} whose domain depends on the shape of the container.
-This means that for a given type of shapes and positions, the equality of two
-objects of a container extension will need to deal with heterogeneous equality.
-Since we want to remain axiom-free, one way to deal with this is to rely on a
-\emph{decidable} validity predicate.
-
-Finally we can define the type of containers as follows:
+Here consider instead representing the positions of a container using a validity
+predicate \coq{Valid : Shape -> Pos -> bool} which assigns shapes to positions.
+Moreover, it is crucial for our purposes of remaining axiom-free that this
+predicate is \emph{decidable} in order to be able to reason about equality of
+container extensions and, as a result, the contents of the container extensions
+need to be accessed via a function \coq{contents} whose domain depends on the
+shape of the container.  This means that for a given type of shapes and
+positions, the equality of two objects of a container extension will need to
+deal with heterogeneous equality.
+\mpav{This part needs more explanation.}
+
 \begin{coqcode}
@@
 \end{coqcode}
-where \coq{valid} is the First, containers are defined as pairs of shape and
-position types that have a \emph{decidable} validity predicate. We represent
-decidable predicates as functions that return a boolean and lift booleans to
-propositions similarly to the implementation provided by
-SSReflect~\cite{GonthierL09}.  For simplicity, we define the dependent pair of
-valid positions as a record, which will be extracted to a singleton type with
-the validity proof erased by Coq's code extraction mechanism.
-
-Then, we define \emph{container extensions} as follows:
+where \coq{valid} is the \emph{decidable} validity predicate. Note that booleans
+can be lifted to propositions similarly to the implementation provided by
+SSReflect~\cite{GonthierL09}. This allows us to define the dependent pair of
+valid positions as a record which will be extracted to a singleton type since
+the validity proof is erased by Coq's code extraction mechanism.
+
+We now define \emph{container extensions} as follows:
 \begin{coqcode}
@@
 \end{coqcode}
+
+% A container extension is then defined as follows:
+% \begin{coqcode}
+% Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
+% Record App (X : Type) := { shape    : Shape;
+%                           contents : {p | Valid shape p} -> X }.
+% \end{coqcode}
+% This approach solves the problem of the unsafe casts, since the code extracted
+% for \coq{{p | Valid shape pos}} will be an OCaml singleton type defined as the
+% OCaml equivalent to \coq{Pos}.
+
+Finally we can define the type of containers as follows:
 

### 2024-02-28 paper/paper.tex
Intro

--- 2024-02-28 paper/paper.tex
@@
     | x::xs -> Some (x, List.partition (fun y -> y < x) xs);;
-
   let rec quicksort xs = match divide xs with
@@
 not many approaches mechanise structured recursion schemes in proof assistants.
-A recent approach by Abreu et al.\~\cite{AbreuDHJMS23} encode an algebraic
+A recent approach by Abreu et al.~\cite{AbreuDHJMS23} encode an algebraic
 approach to divide-and-conquer computations, in which termination is entirely

### 2024-02-28 paper/paper.tex
Shameless self-reference

@@
     | x::xs -> Some (x, List.partition (fun y -> y < x) xs);;
-  let rec quicksort xs = match divide xs with
+  let rec qsort xs = match divide xs with
     | None -> []
-    | Some (x, (smaller, larger)) -> quicksort smaller @ (x::quicksort larger)
+    | Some (x, (smaller, larger)) -> qsort smaller @ (x::qsort larger)
 \end{minted}
-While \ocaml{quicksort} is a well-defined mathematical function it cannot be
+While \ocaml{qsort} is a well-defined mathematical function it cannot be
 accepted by a proof assistant.  The reason is that the \ocaml{divide}
@@
 \emph{structured recursion schemes}, and their associated algebraic laws.  In
-particular, \ocaml{quicksort} is an instance of a \emph{divide-and-conquer}
+particular, \ocaml{qsort} is an instance of a \emph{divide-and-conquer}
 algorithm which, in the functional programming literature is known as a
@@
 fusion}~\cite{TakanoM95}, to (semi-)automatic
-parallelisations~\cite{Gibbons96:Third}.
+parallelisations~\cite{Gibbons96:Third, farmsCastro}.
 
@@
 reasoning, as well as extracting to clean, naturally-looking code. As an
-example our mechanisation of \ocaml{quicksort} as a hylomorphism will produce
+example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
 the following code:

### 2024-02-29 paper/paper.tex
Moved away background on recursion schemes. Going to integrate that in to section on hylomorphisms.

@@
 
-\section{Background on Recursion Schemes}
-Recursion schemes provide a systematic way to work with data types, such as
-trees or lists, by abstracting away the recursion into higher-order functions.
-Canonical recursion schemes include \emph{folds} (catamorphisms) which consume
-data and \emph{unfolds} (anamorphisms) which produce data. These are the unique
-solutions (when they exist) to the following two equations respectively
-\[
-  h  = a \comp F h \comp \InOp \qquad k = \OutOp \comp F k \comp c
-\]
-where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
-can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
-$F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
-$F$-coalgebra), $\InOp$ is the isomoprhism on the least fixed-point of $F$,
-namely $\InOp : \mu F \to F\mu F$, and $\OutOp$ is the isomorphism on
-the greatest fixed-point of $F$, namely $F \nu F \to \nu F$. When
-the unique solutions to the above equations exist we denote them as $\cata{a}$
-and $\ana{c}$ respectively.
-
-\subsection{Hylomorphisms}
-Catamorphisms (dually anamorphisms) do not capture every recursion
-scheme we can think of.  As an example, consider an example of a classic
-\emph{divide-and-conquer} program such as quicksort algorithm:\footnote[1]{This
-program is written in Coq notation to be consistent, but it is not a program
-that Coq can accept.}
-
-\mpav{I need this to be written in Coq notation}
-\begin{coqcode}
-qs :: [a] -> [a]
-qs [] = []
-qs (x:xs) = qs (filter (x <=) xs) ++ [x] ++ (qs (filter (x >) xs))
-\end{coqcode}
-This program is formed of essentially two parts. The first one is what is
-commonly thought as the \emph{partition} function destructuring the input list
-in three parts, the pivot and the other two sublists, and the second part
-recursing over the sublists and composing the results back together.
-
-Notice that if we define the functor \coq{TreeF} $X = X \times A \times X$ with
-the obvious functorial action then \coq{partition} can be viewed as a
-\coq{TreeF}-coalgebra on lists over the type $A$
-\begin{coqcode}
-c : [A] -> TreeF [A]
-c (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
-\end{coqcode}
-
-While we can define an \coq{TreeF}-algebra on lists
-over $A$ as well, that is a map
-\begin{coqcode}
-a : TreeF [A] -> [A]
-a (left, x, right) = left ++ [x] ++ right
-\end{coqcode}
-
-It should be easy to see that the program \coq{qs} can now be defined as the
-unique program that runs the coalgebra $c$ recurses over the sublists produced
-and then composes the results together. The structure exemplified by this
-program is a typical divide-and-conquer algorithm which can be implemented using
-an \emph{hylomorphism}. An hylomorphism is the unique solution $x : C \to A$ to
-the hylo equation
-\begin{equation}
-  \label{eq:hylo}
-  x =  a \comp Fx \comp c
-\end{equation}
-for a $F$-algebra on $A$, an $F$-coalgebra on $C$ and a functor $F$.  If for
-every $F$-algebra there exists a unique solution to (\ref{eq:hylo}) then we say
-the $F$-coalgebra $c$ is \emph{recursive}. Dually if for every $F$-coalgebra $c$
-there exists a unique solution to (\ref{eq:hylo}) we say the $F$-algebra $a$ is
-\emph{corecursive}. We denote the unique solution to the hylo equation
-(\ref{eq:hylo}) as $\hylo{a}{c}$.
-
-Moreover, it shoud be easy to see that catamorphisms and anamorphism arise as a
-special case of hylomorphisms
-\begin{equation}
-  \label{eq:cata}
-  \cata{a} = \hylo{\InOp}{a} \qquad \ana{c} = \hylo{c}{\OutOp}
-\end{equation}
-since $\InOp$ is indeed a recursive coalgebra and $\OutOp$ is a corecursive
-algebra.
-
-\subsection{Implementing Recursion Schemes}
-Hylomorphisms can implement all structured recursion schemes~\cite{HinzeWG15,
-HinzeW16}.  Here we show the result by means of a very basic example: mutual
-recursion. We refer the reader to Hinze et al.~\cite{HinzeWG15} for the
-theoretical result and to Yang and Wu~\cite{YangW22} for more examples of
-recursion schemes.
-
-% More specifically, in the context of recursive data types, such as in Haskell,
-% where induction and coinduction is conflated, an hylomorphism can be implemented
-% in terms of a catarmorphism (folding) and an anamorphism (unfolding)
+% \section{Background on Recursion Schemes}
+% Recursion schemes provide a systematic way to work with data types, such as
+% trees or lists, by abstracting away the recursion into higher-order functions.
+% Canonical recursion schemes include \emph{folds} (catamorphisms) which consume
+% data and \emph{unfolds} (anamorphisms) which produce data. These are the unique
+% solutions (when they exist) to the following two equations respectively
 % \[
-%   \hylo{a}{c} = \cata{a} \comp \ana{c}
+%   h  = a \comp F h \comp \InOp \qquad k = \OutOp \comp F k \comp c
 % \]
-% They first generate a data structure using an anamorphism and then immediately
-% fold or reduce it using a catamorphism.
-
-% The hylomorphism pattern is particularly useful for defining recursive
-% algorithms in a functional programming context, where recursion is a fundamental
-% tool for solving problems. By separating the generation of data structures from
-% their consumption, hylomorphisms enable clearer and more modular code, making it
-% easier to reason about and maintain recursive algorithms.
-
-% However, in a type theory inductive and coinductive data types are separate.
-
-% In this section we will introduce the notion of hylomorphism. By way of examples
-% we will also give high level intution as to way all recursion schemes can be
-% defined in terms of a basic
-
-Say that we want to prove that the two mutually recursive functions below are
-well-defined
-
-\begin{tabular}{LL}
-  \operator{even} : \Nat \to \Bool                  &  \operator{odd} : \Nat \to \Bool\\
-  \operator{even} (0) = \top                        &  \operator{odd}(0) = \bot\\
-  \operator{even} (n+1) = \neg \operator{odd}(n)    &  \operator{odd}(n+1) = \neg (\operator{even}(n))
-\end{tabular}
-
-One option is to prove that the following function is
-equivalent to the above
-\begin{align*}
-  & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
-  & \operator{even-odd} (0) = (\top, \bot)\\
-  & \operator{even-odd} (n+1) = \operator{let } (x,y) \leftarrow \operator{even-odd}(n) \operator{ in } (\neg x, \neg y)
-\end{align*}
-which is clearly a catamorphism $\cata{a}$ with $a (\Nil) = (\top, \bot)$ and
-$a(\operator{x,y}) = (\neg x, \neg y)$ and hence an hylomorphism as defined in
-(\ref{eq:cata}).  Now we are left to show they are equivalent. Well, it is easy
-to do so since $\operator{even} = \pi_{1} \comp \operator{even-odd}$
-$\operator{odd} = \pi_{2} \comp \operator{even-odd}$. The proof that these two
-pairs of functions are the same is an easy induction over the natural numbers.
-
-This is an instance of a more general phenonema whose argument uses the concept
-of an adjunction~\cite{MacLane71}.  For this particular example, the argument is
-as follows.  There is an isomorphism between pairs of maps
-$(A \to B_{1}, A \to B_{2})$ and maps $(A \to B_{1} \times B_{2})$.  This is
-given by $\Phi (f,g) = \pair{f}{g}$ and
-$\Psi(f) = (\pi_{1} \comp f, \pi_{2} \comp f)$ where the projections $\pi$ are
-part of the counit of the adjunction between the diagonal functor
-$\Delta X = (X,X)$ and the product $\times$.  Once this correspondence is
-established we can implement the map $A \to B_{1} \times B_{2}$ as an
-hylomorphism and derive the original and more complex recursion scheme.
-
-% In general an adjunction is a natural correspondence of maps
-% \[
-%   LA \to B \iff A \to RB
-% \]
-% where $L$ and $R$ are called adjoints.
-
-However, formalising adjunctions in Coq would require to formalise a
-non-negligible amount of category theory which goes beyond the scope of the
-present work. In this paper, we will instead derive the recursion schemes as
-needed and implement them as hylos.
-% \subsubsection{Course-of-Value Recursion}
-% The knapsack algorithm is defined as following in pseudo-Haskell code:
+% where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
+% can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
+% $F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
+% $F$-coalgebra), $\InOp$ is the isomoprhism on the least fixed-point of $F$,
+% namely $\InOp : \mu F \to F\mu F$, and $\OutOp$ is the isomorphism on
+% the greatest fixed-point of $F$, namely $F \nu F \to \nu F$. When
+% the unique solutions to the above equations exist we denote them as $\cata{a}$
+% and $\ana{c}$ respectively.
+
+% \subsection{Hylomorphisms}
+% Catamorphisms (dually anamorphisms) do not capture every recursion
+% scheme we can think of.  As an example, consider an example of a classic
+% \emph{divide-and-conquer} program such as quicksort algorithm:\footnote[1]{This
+% program is written in Coq notation to be consistent, but it is not a program
+% that Coq can accept.}
+
+% \mpav{I need this to be written in Coq notation}
+% \begin{coqcode}
+% qs :: [a] -> [a]
+% qs [] = []
+% qs (x:xs) = qs (filter (x <=) xs) ++ [x] ++ (qs (filter (x >) xs))
+% \end{coqcode}
+% This program is formed of essentially two parts. The first one is what is
+% commonly thought as the \emph{partition} function destructuring the input list
+% in three parts, the pivot and the other two sublists, and the second part
+% recursing over the sublists and composing the results back together.
+
+% Notice that if we define the functor \coq{TreeF} $X = X \times A \times X$ with
+% the obvious functorial action then \coq{partition} can be viewed as a
+% \coq{TreeF}-coalgebra on lists over the type $A$
+% \begin{coqcode}
+% c : [A] -> TreeF [A]
+% c (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
+% \end{coqcode}
+
+% While we can define an \coq{TreeF}-algebra on lists
+% over $A$ as well, that is a map
+% \begin{coqcode}
+% a : TreeF [A] -> [A]
+% a (left, x, right) = left ++ [x] ++ right
+% \end{coqcode}
+
+% It should be easy to see that the program \coq{qs} can now be defined as the
+% unique program that runs the coalgebra $c$ recurses over the sublists produced
+% and then composes the results together. The structure exemplified by this
+% program is a typical divide-and-conquer algorithm which can be implemented using
+% an \emph{hylomorphism}. An hylomorphism is the unique solution $x : C \to A$ to
+% the hylo equation
+% \begin{equation}
+%   \label{eq:hylo}
+%   x =  a \comp Fx \comp c
+% \end{equation}
+% for a $F$-algebra on $A$, an $F$-coalgebra on $C$ and a functor $F$.  If for
+% every $F$-algebra there exists a unique solution to (\ref{eq:hylo}) then we say
+% the $F$-coalgebra $c$ is \emph{recursive}. Dually if for every $F$-coalgebra $c$
+% there exists a unique solution to (\ref{eq:hylo}) we say the $F$-algebra $a$ is
+% \emph{corecursive}. We denote the unique solution to the hylo equation
+% (\ref{eq:hylo}) as $\hylo{a}{c}$.
+
+% Moreover, it shoud be easy to see that catamorphisms and anamorphism arise as a
+% special case of hylomorphisms
+% \begin{equation}
+%   \label{eq:cata}
+%   \cata{a} = \hylo{\InOp}{a} \qquad \ana{c} = \hylo{c}{\OutOp}
+% \end{equation}
+% since $\InOp$ is indeed a recursive coalgebra and $\OutOp$ is a corecursive
+% algebra.
+
+% \subsection{Implementing Recursion Schemes}
+% Hylomorphisms can implement all structured recursion schemes~\cite{HinzeWG15,
+% HinzeW16}.  Here we show the result by means of a very basic example: mutual
+% recursion. We refer the reader to Hinze et al.~\cite{HinzeWG15} for the
+% theoretical result and to Yang and Wu~\cite{YangW22} for more examples of
+% recursion schemes.
+
+% % More specifically, in the context of recursive data types, such as in Haskell,
+% % where induction and coinduction is conflated, an hylomorphism can be implemented
+% % in terms of a catarmorphism (folding) and an anamorphism (unfolding)
+% % \[
+% %   \hylo{a}{c} = \cata{a} \comp \ana{c}
+% % \]
+% % They first generate a data structure using an anamorphism and then immediately
+% % fold or reduce it using a catamorphism.
+
+% % The hylomorphism pattern is particularly useful for defining recursive
+% % algorithms in a functional programming context, where recursion is a fundamental
+% % tool for solving problems. By separating the generation of data structures from
+% % their consumption, hylomorphisms enable clearer and more modular code, making it
+% % easier to reason about and maintain recursive algorithms.
+
+% % However, in a type theory inductive and coinductive data types are separate.
+
+% % In this section we will introduce the notion of hylomorphism. By way of examples
+% % we will also give high level intution as to way all recursion schemes can be
+% % defined in terms of a basic
+
+% Say that we want to prove that the two mutually recursive functions below are
+% well-defined
+
+% \begin{tabular}{LL}
+%   \operator{even} : \Nat \to \Bool                  &  \operator{odd} : \Nat \to \Bool\\
+%   \operator{even} (0) = \top                        &  \operator{odd}(0) = \bot\\
+%   \operator{even} (n+1) = \neg \operator{odd}(n)    &  \operator{odd}(n+1) = \neg (\operator{even}(n))
+% \end{tabular}
+
+% One option is to prove that the following function is
+% equivalent to the above
 % \begin{align*}
-%   & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
-%   & \operator{knapsack}\;wvs\; c = \operator{maximum}_{0}
-%     [v + \operator{knapsack}\;wvs\; (c - w) \mid (w,v) \leftarrow wvs, w \le c]
+%   & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
+%   & \operator{even-odd} (0) = (\top, \bot)\\
+%   & \operator{even-odd} (n+1) = \operator{let } (x,y) \leftarrow \operator{even-odd}(n) \operator{ in } (\neg x, \neg y)
 % \end{align*}
-% %
-% with dynamic programming knapsack can be rephrased like this
-% \begin{align*}
-%   & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
-%   & \operator{knapsack}\;wvs\;c= table !! c \\
-%   & \qquad table  = [ks\;i \mid i \leftarrow [0..c]]\\
-%   & \qquad ks\; i = maximum_{0} [v + table!! (i - w) \mid (w,v) \leftarrow wvs, w \le i]
-% \end{align*}
-
-% The recursion scheme is justified by solving a recursive equation.  First we
-% define an operation to create a memory table representing the list of capacities
-% $[0..c]$. We first define a functor $D X = 1 + X$ representing the base functor
-% for constructing the naturals $\Nat = \mu D$ the we define the generating
-% function by induction on the naturals
-% \begin{align*}
-%   & \operator{fan} : \Nat \to D_{\infty}\Nat\\
-%   & \operator{fan} (0)    = \pair{0}{\inl}\\
-%   & \operator{fan} (n+1)  = \pair{n+1}{\inr (\operator{fan}(n))}
-% \end{align*}
-% The recursion scheme from a comonad is then defined as follows
-% \begin{align*}
-%   & \operator{rsfc} : \Nat \to A\\
-%   & \operator{rsfc} = alg \comp DD_{\infty}(\operator{rsfc}) \comp D(\operator{fan}) \comp \InOp
-% \end{align*}
-% At his point we can given the definition of the algebra $alg$ which effectively implements knapsack.
-% Firs we define the $\operator{lookup}$ function which returns (if present) the element indexed in the memory table
-% \begin{align*}
-%   & \operator{lookup} : \Nat \to D_{\infty} \R \to \R_{\bot}\\
-%   & \operator{lookup} (0,   \pair{r}{\textunderscore})  = r\\
-%   & \operator{lookup} (n+1, \pair{r}{\inl})         = \bot\\
-%   & \operator{lookup} (n+1, \pair{r}{\inr(table)})  = \operator{lookup} (n, table)
-% \end{align*}
-% and now we can define the algebra over the reals
-% \begin{align*}
-%   & \operator{alg} : DD_{\infty} \R \to \R\\
-%   & \operator{alg}\; (\inl)  = 0\\
-%   & \operator{alg}\; (\inr(table))  = \operator{maximun}_{0} [ v + \operator{lookup}\; table\; (i - w) \mid (w,v) \leftarrow wvs, w \le i]
-% \end{align*}
-
-% It remains to show how an $\operator{rsfc}$ is implemented in terms of
-% hylomorhisms.
-
-% \paragraph{Implementation into hylomorphisms}
-% In order to implement rsfcs as hylos we need a parametric distributive law
-% \[
-%   \lambda_{X} : GG_{\infty}X \to G_{\infty}GX
-% \]
-% This can be defined as
-% $\lambda = G_{\infty}G(\head) \comp \anacofree{G(\tail)}$.  The distributive law
-% is instrumental to be able to lift coalgebras in the following sense. Give a
-% $G$-coalgebra on $X$, a lifting $H$ is a functor which lifts this coalgebra to a
-% $G$-coalgebra on $HX$.  In particular, the lifting $G_{\lambda}$ takes
-% $G_{\infty}$-coalgebras on $X$ and transforms them into $G_{\infty}$-coalgebras
-% on $GX$
-% \begin{align*}
-%   & G_{\lambda}(c) : GA \to G_{\infty} GA\\
-%   & G_{\lambda}(c) = \lambda \comp G (c)
-% \end{align*}
-% where $c : C \to GC$.
-
-% Now, given an algebra $a : GG_{\infty} A \to A$ on the carrier $A$ we can
-% convert it into an algebra for the carrier $G_{\infty} A$ by
-% \begin{align*}
-%   & \floor{a} : GG_{\infty} A \to G_{\infty} A\\
-%   & \floor{a} = G_{\infty}(a) \comp \anacofree{G_{\lambda}(\delta)}
-% \end{align*}
-% where $\eta_{A} : A \to G_{\infty} A$ is the unit of the comonad creating an
-% infinite constant list.
-
-% At this point the hylo is simply defined by
-% $\hylo{\InOp}{a} : \mu F \to G_{\infty}A$  and the rsfc is derived as follows
-% \begin{align*}
-%   & \operator{rsfc} : \mu G \to A \\
-%   & \operator{rsfc} = \head \comp \hylo{\InOp}{a}
-% \end{align*}
-
-% % \subsection{Catamorphisms and Anamorphism}
-% % A catamorphism is the unique program defined as
+% which is clearly a catamorphism $\cata{a}$ with $a (\Nil) = (\top, \bot)$ and
+% $a(\operator{x,y}) = (\neg x, \neg y)$ and hence an hylomorphism as defined in
+% (\ref{eq:cata}).  Now we are left to show they are equivalent. Well, it is easy
+% to do so since $\operator{even} = \pi_{1} \comp \operator{even-odd}$
+% $\operator{odd} = \pi_{2} \comp \operator{even-odd}$. The proof that these two
+% pairs of functions are the same is an easy induction over the natural numbers.
+
+% This is an instance of a more general phenonema whose argument uses the concept
+% of an adjunction~\cite{MacLane71}.  For this particular example, the argument is
+% as follows.  There is an isomorphism between pairs of maps
+% $(A \to B_{1}, A \to B_{2})$ and maps $(A \to B_{1} \times B_{2})$.  This is
+% given by $\Phi (f,g) = \pair{f}{g}$ and
+% $\Psi(f) = (\pi_{1} \comp f, \pi_{2} \comp f)$ where the projections $\pi$ are
+% part of the counit of the adjunction between the diagonal functor
+% $\Delta X = (X,X)$ and the product $\times$.  Once this correspondence is
+% established we can implement the map $A \to B_{1} \times B_{2}$ as an
+% hylomorphism and derive the original and more complex recursion scheme.
+
+% % In general an adjunction is a natural correspondence of maps
 % % \[
-% %   x = a \comp \fmap{x} \comp \InOp
+% %   LA \to B \iff A \to RB
 % % \]
-% % Can be implemented:
+% % where $L$ and $R$ are called adjoints.
+
+% However, formalising adjunctions in Coq would require to formalise a
+% non-negligible amount of category theory which goes beyond the scope of the
+% present work. In this paper, we will instead derive the recursion schemes as
+% needed and implement them as hylos.
+% % \subsubsection{Course-of-Value Recursion}
+% % The knapsack algorithm is defined as following in pseudo-Haskell code:
+% % \begin{align*}
+% %   & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
+% %   & \operator{knapsack}\;wvs\; c = \operator{maximum}_{0}
+% %     [v + \operator{knapsack}\;wvs\; (c - w) \mid (w,v) \leftarrow wvs, w \le c]
+% % \end{align*}
+% % %
+% % with dynamic programming knapsack can be rephrased like this
+% % \begin{align*}
+% %   & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
+% %   & \operator{knapsack}\;wvs\;c= table !! c \\
+% %   & \qquad table  = [ks\;i \mid i \leftarrow [0..c]]\\
+% %   & \qquad ks\; i = maximum_{0} [v + table!! (i - w) \mid (w,v) \leftarrow wvs, w \le i]
+% % \end{align*}
+
+% % The recursion scheme is justified by solving a recursive equation.  First we
+% % define an operation to create a memory table representing the list of capacities
+% % $[0..c]$. We first define a functor $D X = 1 + X$ representing the base functor
+% % for constructing the naturals $\Nat = \mu D$ the we define the generating
+% % function by induction on the naturals
+% % \begin{align*}
+% %   & \operator{fan} : \Nat \to D_{\infty}\Nat\\
+% %   & \operator{fan} (0)    = \pair{0}{\inl}\\
+% %   & \operator{fan} (n+1)  = \pair{n+1}{\inr (\operator{fan}(n))}
+% % \end{align*}
+% % The recursion scheme from a comonad is then defined as follows
+% % \begin{align*}
+% %   & \operator{rsfc} : \Nat \to A\\
+% %   & \operator{rsfc} = alg \comp DD_{\infty}(\operator{rsfc}) \comp D(\operator{fan}) \comp \InOp
+% % \end{align*}
+% % At his point we can given the definition of the algebra $alg$ which effectively implements knapsack.
+% % Firs we define the $\operator{lookup}$ function which returns (if present) the element indexed in the memory table
+% % \begin{align*}
+% %   & \operator{lookup} : \Nat \to D_{\infty} \R \to \R_{\bot}\\
+% %   & \operator{lookup} (0,   \pair{r}{\textunderscore})  = r\\
+% %   & \operator{lookup} (n+1, \pair{r}{\inl})         = \bot\\
+% %   & \operator{lookup} (n+1, \pair{r}{\inr(table)})  = \operator{lookup} (n, table)
+% % \end{align*}
+% % and now we can define the algebra over the reals
+% % \begin{align*}
+% %   & \operator{alg} : DD_{\infty} \R \to \R\\
+% %   & \operator{alg}\; (\inl)  = 0\\
+% %   & \operator{alg}\; (\inr(table))  = \operator{maximun}_{0} [ v + \operator{lookup}\; table\; (i - w) \mid (w,v) \leftarrow wvs, w \le i]
+% % \end{align*}
+
+% % It remains to show how an $\operator{rsfc}$ is implemented in terms of
+% % hylomorhisms.
+
+% % \paragraph{Implementation into hylomorphisms}
+% % In order to implement rsfcs as hylos we need a parametric distributive law
 % % \[
-% %   \cata{a} = \hylo{\InOp}{a}
+% %   \lambda_{X} : GG_{\infty}X \to G_{\infty}GX
 % % \]
-
-% % An anamorphism is the unique program satisfying the equation
+% % This can be defined as
+% % $\lambda = G_{\infty}G(\head) \comp \anacofree{G(\tail)}$.  The distributive law
+% % is instrumental to be able to lift coalgebras in the following sense. Give a
+% % $G$-coalgebra on $X$, a lifting $H$ is a functor which lifts this coalgebra to a
+% % $G$-coalgebra on $HX$.  In particular, the lifting $G_{\lambda}$ takes
+% % $G_{\infty}$-coalgebras on $X$ and transforms them into $G_{\infty}$-coalgebras
+% % on $GX$
+% % \begin{align*}
+% %   & G_{\lambda}(c) : GA \to G_{\infty} GA\\
+% %   & G_{\lambda}(c) = \lambda \comp G (c)
+% % \end{align*}
+% % where $c : C \to GC$.
+
+% % Now, given an algebra $a : GG_{\infty} A \to A$ on the carrier $A$ we can
+% % convert it into an algebra for the carrier $G_{\infty} A$ by
+% % \begin{align*}
+% %   & \floor{a} : GG_{\infty} A \to G_{\infty} A\\
+% %   & \floor{a} = G_{\infty}(a) \comp \anacofree{G_{\lambda}(\delta)}
+% % \end{align*}
+% % where $\eta_{A} : A \to G_{\infty} A$ is the unit of the comonad creating an
+% % infinite constant list.
+
+% % At this point the hylo is simply defined by
+% % $\hylo{\InOp}{a} : \mu F \to G_{\infty}A$  and the rsfc is derived as follows
+% % \begin{align*}
+% %   & \operator{rsfc} : \mu G \to A \\
+% %   & \operator{rsfc} = \head \comp \hylo{\InOp}{a}
+% % \end{align*}
+
+% % % \subsection{Catamorphisms and Anamorphism}
+% % % A catamorphism is the unique program defined as
+% % % \[
+% % %   x = a \comp \fmap{x} \comp \InOp
+% % % \]
+% % % Can be implemented:
+% % % \[
+% % %   \cata{a} = \hylo{\InOp}{a}
+% % % \]
+
+% % % An anamorphism is the unique program satisfying the equation
+% % % \[
+% % %   x = \OutOp \comp \fmap{x} \comp{c}
+% % % \]
+% % % The anamorphism is indicated by $\ana{c}$. Recursion schemes given by a
+% % % corecursive algebra such as $\OutIso$ are also called hylomorphism but they are
+% % % indicated with the symbol for the anamorphism, $\cohylo{c}{\OutOp}$.
+
+% % \paragraph{Free Monads}
+% % Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
+% % solution to the following domain equation
 % % \[
-% %   x = \OutOp \comp \fmap{x} \comp{c}
+% %   D^{*}A = A + DD^{*}A
 % % \]
-% % The anamorphism is indicated by $\ana{c}$. Recursion schemes given by a
-% % corecursive algebra such as $\OutIso$ are also called hylomorphism but they are
-% % indicated with the symbol for the anamorphism, $\cohylo{c}{\OutOp}$.
-
-% \paragraph{Free Monads}
-% Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
-% solution to the following domain equation
-% \[
-%   D^{*}A = A + DD^{*}A
-% \]
-% where $\eta_{A} : A \to D^{*}A$ is the left injection and
-% $\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
-% $a : DA \to A$ there is a unique solution to the following recursive equation
-% \[
-%   x \comp \call_{A} = a \comp D(x)
-% \]
-% that is $\catafree{a} = \cata{[id_{A}; a]}$. With this operation we can extend
-% the free algebra $\call$ to the multiplication $\mu : D^{*}D^{*}A \to D^{*}A$
-% defined as $\mu = \catafree{\call}$.
-
-% \paragraph{Cofree Comonads}
-% Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
-% greatest solution to the following domain equation
-% \[
-%   D_{\infty}A = A + DD_{\infty}A
-% \]
-% where $\head : D_{\infty}A \to A$ is the first projection and
-% $\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.
-% Given an algebra $c : A \to DA$ there is a unique solution to the following
-% recursive equation
-% \[
-%   \tail_{A} \comp x = D(x) \comp c
-% \]
-% that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$. With the latter operation we
-% can extend $\tail$ to the comultiplication for the comonad
-% $\delta : D_{\infty}A \to D_{\infty} D_{\infty}A$ defined by
-% $\delta = \anacofree{\tail}$.
-
-% \subsection{Accumulators}
-% For an algebra $a : D(A^{P})\times P \to A$:
-% \begin{align*}
-%   & g : \mu D \times P \to B\\
-%   & g (d, p) = a (\fmap{(\curry g)} \comp \InOp(d), p)
-% \end{align*}
-
-% \paragraph{Hylo Implementation}
-% Given an algebra $a : D(A^{P}) \to A^{P}$ the accumorphism corresponds to the unique
-% solution to the following equation
-% \begin{align*}
-%   & f : \mu D \to A^{P}\\
-%   & f = a \comp \fmap (f) \comp \InOp
-% \end{align*}
-% thus $f = \hylo{\InOp}{a}$.
-
-% The complex recursion scheme is then
-% $g = \uncurry\hylo{\InOp}{a} : \mu D \times P \to B$.
-
-% \subsection{Mutu-Hylo}
-% For an algebra $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and an algebra
-% $a_{2} : D(A_{1} \times A_{2}) \to A_{2}$:
-% \begin{align*}
-%   & (g_{1}, g_{2}) : (\mu D, \mu D) \to (A_{1}, A_{2})\\
-%   & (g_{1}, g_{2}) = (a_{1}, a_{2}) \comp (\fmap{\pair{g_{1}}{g_{2}}},\fmap{\pair{g_{1}}{g_{2}}}) \comp (\InOp, \InOp)
-% \end{align*}
-
-% \paragraph{Hylo Implementation}
-% Given an algebra $a : D(A_{1} \times A_{2}) \to A_{1} \times A_{2}$
-% \begin{align*}
-%   & f : \mu D \to A_{1} \times A_{2}\\
-%   & f = a \comp \fmap{f} \comp \InOp
-% \end{align*}
-% for which the solution is an hylo $f = \hylo{\InOp}{a}$.
-
-% The mutumorphism can be implemented as
-% $(g_{1}, g_{2}) = (\pi_{1} \comp f, \pi_{2} \comp f)$.
-
-% \paragraph{Coq Implementation}
-% Given the two algebras  $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and
-% $a_{2} : D(A_{1} \times A_{2}) \to A_{2}$.
-% We define the two recursive functions  $f_{1}$ and $f_{2}$ as
-
-% \[
-%   f_{1} = \pi_{1} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}} \qquad f_{2} = \pi_{2} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}}
-% \]
-% % A recursion scheme from a comonad $D_{\infty}$ for a data functor $F$
-
-% % For example, for the functor $(1 + -)$ an algebra over this functor is a
-% % function $1 + X \to X$ is a pair of functions $\Nil : 1 \to X$ and
-% % $\Succ : X \to X$. The cofree comonad over $D$ is defined as follows
+% % where $\eta_{A} : A \to D^{*}A$ is the left injection and
+% % $\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
+% % $a : DA \to A$ there is a unique solution to the following recursive equation
 % % \[
-% %   D_{\infty} A \iso A \times (1 + D_{\infty})
+% %   x \comp \call_{A} = a \comp D(x)
 % % \]
-% % which is the the type of non-empty possibly infinite lists with elements in $A$
-
-% % First we have to assume a distributive law
-% % $\lambda : DD_{\infty} \to D_{\infty}D$.  For an algebra
-% % $a : D(D_{\infty} A) \to A$,
+% % that is $\catafree{a} = \cata{[id_{A}; a]}$. With this operation we can extend
+% % the free algebra $\call$ to the multiplication $\mu : D^{*}D^{*}A \to D^{*}A$
+% % defined as $\mu = \catafree{\call}$.
+
+% % \paragraph{Cofree Comonads}
+% % Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
+% % greatest solution to the following domain equation
+% % \[
+% %   D_{\infty}A = A + DD_{\infty}A
+% % \]
+% % where $\head : D_{\infty}A \to A$ is the first projection and
+% % $\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.
+% % Given an algebra $c : A \to DA$ there is a unique solution to the following
+% % recursive equation
+% % \[
+% %   \tail_{A} \comp x = D(x) \comp c
+% % \]
+% % that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$. With the latter operation we
+% % can extend $\tail$ to the comultiplication for the comonad
+% % $\delta : D_{\infty}A \to D_{\infty} D_{\infty}A$ defined by
+% % $\delta = \anacofree{\tail}$.
+
+% % \subsection{Accumulators}
+% % For an algebra $a : D(A^{P})\times P \to A$:
 % % \begin{align*}
-% %   & g : \mu D \to A\\
-% %   & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
+% %   & g : \mu D \times P \to B\\
+% %   & g (d, p) = a (\fmap{(\curry g)} \comp \InOp(d), p)
 % % \end{align*}
-% % where
-% % $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX$
-% % is a $D$-algebra on $GX$. The catamorphism of this algebra has the following
-% % type
-% % \[
-% %   \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
-% % \]
-% % which is the type of a $G$-algebra used to initialise the memory table.
-
-% % Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
-% % what is the unique solution of the following equation
+
+% % \paragraph{Hylo Implementation}
+% % Given an algebra $a : D(A^{P}) \to A^{P}$ the accumorphism corresponds to the unique
+% % solution to the following equation
 % % \begin{align*}
-% %   & f : \mu F \to G_{\infty} A\\
+% %   & f : \mu D \to A^{P}\\
+% %   & f = a \comp \fmap (f) \comp \InOp
+% % \end{align*}
+% % thus $f = \hylo{\InOp}{a}$.
+
+% % The complex recursion scheme is then
+% % $g = \uncurry\hylo{\InOp}{a} : \mu D \times P \to B$.
+
+% % \subsection{Mutu-Hylo}
+% % For an algebra $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and an algebra
+% % $a_{2} : D(A_{1} \times A_{2}) \to A_{2}$:
+% % \begin{align*}
+% %   & (g_{1}, g_{2}) : (\mu D, \mu D) \to (A_{1}, A_{2})\\
+% %   & (g_{1}, g_{2}) = (a_{1}, a_{2}) \comp (\fmap{\pair{g_{1}}{g_{2}}},\fmap{\pair{g_{1}}{g_{2}}}) \comp (\InOp, \InOp)
+% % \end{align*}
+
+% % \paragraph{Hylo Implementation}
+% % Given an algebra $a : D(A_{1} \times A_{2}) \to A_{1} \times A_{2}$
+% % \begin{align*}
+% %   & f : \mu D \to A_{1} \times A_{2}\\
 % %   & f = a \comp \fmap{f} \comp \InOp
 % % \end{align*}
-% % The solution to this equation can be found again using an hylo
-% % $f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.  The course-of-value recursion
-% % can be recovered from this by setting $g = \head \comp f$
-
-% % \subsubsection{Histomorphism}
-
-% % The definition of $\lambda$ is:
-% % \begin{align*}
-% %   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
-% %   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
-% % \end{align*}
-% % For example, for $GX = 1 + X$ we have to define
-% % \begin{align*}
-% %   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
-% %   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
-% %   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
-% % \end{align*}
-
+% % for which the solution is an hylo $f = \hylo{\InOp}{a}$.
+
+% % The mutumorphism can be implemented as
+% % $(g_{1}, g_{2}) = (\pi_{1} \comp f, \pi_{2} \comp f)$.
+
+% % \paragraph{Coq Implementation}
+% % Given the two algebras  $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and
+% % $a_{2} : D(A_{1} \times A_{2}) \to A_{2}$.
+% % We define the two recursive functions  $f_{1}$ and $f_{2}$ as
+
+% % \[
+% %   f_{1} = \pi_{1} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}} \qquad f_{2} = \pi_{2} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}}
+% % \]
+% % % A recursion scheme from a comonad $D_{\infty}$ for a data functor $F$
+
+% % % For example, for the functor $(1 + -)$ an algebra over this functor is a
+% % % function $1 + X \to X$ is a pair of functions $\Nil : 1 \to X$ and
+% % % $\Succ : X \to X$. The cofree comonad over $D$ is defined as follows
+% % % \[
+% % %   D_{\infty} A \iso A \times (1 + D_{\infty})
+% % % \]
+% % % which is the the type of non-empty possibly infinite lists with elements in $A$
+
+% % % First we have to assume a distributive law
+% % % $\lambda : DD_{\infty} \to D_{\infty}D$.  For an algebra
+% % % $a : D(D_{\infty} A) \to A$,
 % % % \begin{align*}
-% % %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
-% % %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
-% % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
-% % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
+% % %   & g : \mu D \to A\\
+% % %   & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
 % % % \end{align*}
-
-% % % Defining the algebra
+% % % where
+% % % $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX$
+% % % is a $D$-algebra on $GX$. The catamorphism of this algebra has the following
+% % % type
+% % % \[
+% % %   \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
+% % % \]
+% % % which is the type of a $G$-algebra used to initialise the memory table.
+
+% % % Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
+% % % what is the unique solution of the following equation
 % % % \begin{align*}
-% % %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
-% % %   & \operator{alg}\; Zero           = 0\\
-% % %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
+% % %   & f : \mu F \to G_{\infty} A\\
+% % %   & f = a \comp \fmap{f} \comp \InOp
 % % % \end{align*}
+% % % The solution to this equation can be found again using an hylo
+% % % $f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.  The course-of-value recursion
+% % % can be recovered from this by setting $g = \head \comp f$
+
+% % % \subsubsection{Histomorphism}
+
+% % % The definition of $\lambda$ is:
+% % % \begin{align*}
+% % %   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
+% % %   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
+% % % \end{align*}
+% % % For example, for $GX = 1 + X$ we have to define
+% % % \begin{align*}
+% % %   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
+% % %   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
+% % %   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
+% % % \end{align*}
+
+% % % % \begin{align*}
+% % % %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
+% % % %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
+% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
+% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
+% % % % \end{align*}
+
+% % % % Defining the algebra
+% % % % \begin{align*}
+% % % %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
+% % % %   & \operator{alg}\; Zero           = 0\\
+% % % %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
+% % % % \end{align*}
 
@@
 
-\section{Certified Hylomorphisms}
-
+\section{Algebras, Coalgebras and Hylomorphisms}
+In the previous sections we have introduced the concept of a
 \subsection{Algebras}

### 2024-02-29 gen_makefile.sh
Simplifying proofs

@@ -3,3 +3,5 @@
 coq_makefile -f _CoqProject -o Makefile
 
 [ ! -d extraction ] && mkdir extraction
+
+exit 0

### 2024-02-29 paper/paper.tex
minor
On basics about hylomorhisms.

@@
 \section{Algebras, Coalgebras and Hylomorphisms}
-In the previous sections we have introduced the concept of a
+Recursion schemes provide an abstract way to consume and generate data.  The
+most simple and well-known recursion scheme is the \emph{fold} or
+\emph{catamorphism} which consumes data. Its dual is the \emph{unfold} or
+\emph{anamorphisms} which produces data.
+
+For example, assuming the type of lists over a type \coq{A},
+namely \coq{List A}, and given an
+algebra \coq{alg : unit + A * List A -> List A} a
+catamorphism \coq{cata alg : List A -> B} destructs a list into its head and
+tail using a destructor \coq{InOp : List A -> unit + A * List A}. This latter is
+actually an isomorphim with its inverse \coq{In : unit + A * List A -> List A}
+defined by the maps \coq{Empty : unit -> List A}
+and \coq{Cons : A * List A -> List A}.
+
+After destructing the input the catamorphism is applying itsefl recursively.
+This is done using functoriality of the shape of the data, which is a map of
+type \coq{unit + A * List A -> unit + A * List B}. Once the recursion as
+returned it result it applies the algebra \coq{alg} which composes together the
+results of the recursion yielding the result of type \coq{B}.
+
+More abstractly, assuming a base functor \coq{F} a catamorphism is the unique solution to the equation
+\coq{cata alg = alg . fmap (cata alg) . InOp}.
+
+% where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
+% can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
+% $F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
+% $F$-coalgebra), $\InOp$ is the isomoprhism on the least fixed-point of $F$,
+% namely $\InOp : \mu F \to F\mu F$, and $\OutOp$ is the isomorphism on the
+% greatest fixed-point of $F$, namely $F \nu F \to \nu F$. When the unique
+% solutions to the above equations exist we denote them as $\cata{a}$ and
+% $\ana{c}$ respectively.
+
+% C
+Hylomorphism are more general recursion schemes capturing the
+\emph{divide-and-conquer} idea. In short, the input is destructured by its not
+structurally. In the example, we gave in the introduction the destructuring
+program is given by \coq{divide} which is not an isomorphism.  The question the
+hylomorphism satisfies is more general than the catamorphism in that the
+destructing map \coq{InOp} is generalised to a \emph{recursive coalgebra}, that
+is intuitively a coalgebra that can be applied only a finite amount of times.
+In fact, in the case of \coq{divide} this is a recursive coalgebra because the
+size of the list is getting smaller.
+
+% This program is formed of essentially two parts. The first one is what is
+% commonly thought as the \emph{partition} function destructuring the input list
+% in three parts, the pivot and the other two sublists, and the second part
+% recursing over the sublists and composing the results back together.
+
+With this in mind we know proceed onto describing how to formalise hylomorphisms
+in Coq.
+
 \subsection{Algebras}
-
 Algebras on container extensions are defined as expected. The carrier of the
-algebra is \coq{A}, and the operations of the algebra are morphisms
-of the following type:
+algebra is \coq{A}, and the operations of the algebra are morphisms of the
+following type:
 \begin{coqcode}

### 2024-02-29 paper/paper.tex
Centered a bunch of stuff and inserted cit for recursive coalgebras.

@@
 tail using a destructor \coq{InOp : List A -> unit + A * List A}. This latter is
-actually an isomorphim with its inverse \coq{In : unit + A * List A -> List A}
+actually an isomorphim with its inverse
+\begin{minted}{coq}
+In : unit + A * List A -> List A
+\end{minted}
+
 defined by the maps \coq{Empty : unit -> List A}
@@
 This is done using functoriality of the shape of the data, which is a map of
-type \coq{unit + A * List A -> unit + A * List B}. Once the recursion as
-returned it result it applies the algebra \coq{alg} which composes together the
-results of the recursion yielding the result of type \coq{B}.
-
-More abstractly, assuming a base functor \coq{F} a catamorphism is the unique solution to the equation
-\coq{cata alg = alg . fmap (cata alg) . InOp}.
+type
+\begin{minted}{coq}
+unit + A * List A -> unit + A * List B
+\end{minted}
+Once the recursion as returned it result it applies the algebra \coq{alg} which
+composes together the results of the recursion yielding the result of
+type \coq{B}.
+
+More abstractly, assuming a base functor \coq{F} a catamorphism is the unique
+solution to the equation
+\begin{minted}{coq}
+cata alg = alg . fmap (cata alg) . InOp
+\end{minted}
 
@@
 hylomorphism satisfies is more general than the catamorphism in that the
-destructing map \coq{InOp} is generalised to a \emph{recursive coalgebra}, that
-is intuitively a coalgebra that can be applied only a finite amount of times.
-In fact, in the case of \coq{divide} this is a recursive coalgebra because the
-size of the list is getting smaller.
+destructing map \coq{InOp} is generalised to a \emph{recursive
+coalgebra}~\cite{AdamekMM19,CaprettaUV04}, that is intuitively a coalgebra that
+can be applied only a finite amount of times.  In fact, in the case
+of \coq{divide} this is a recursive coalgebra because the size of the list is
+getting smaller.
 
@@
 container extension. We encode in Coq the notion of \emph{recursive
-anamorphisms} and \emph{recursive coalgebras}~\cite{TODO}.  Recursive
+anamorphisms} and \emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  Recursive
 coalgebras are coalgebras whose anamorphism produces finite trees.  The

### 2024-03-01 paper/paper.tex
Algebras and Initial Algebras

@@
 \section{Algebras, Coalgebras and Hylomorphisms}
-Recursion schemes provide an abstract way to consume and generate data.  The
-most simple and well-known recursion scheme is the \emph{fold} or
+Recursion schemes provide an abstract way to consume and generate data. We know
+proceed onto describing how to formalise hylomorphisms in Coq. We first
+formalise algebras for container extensions (Section~\ref{sec:algebras}), then
+we formalise recursive coalgebras (Section~\ref{sec:rec-coalgebras}) and then we
+put together these notions to formalise hylomorphisms (Section~\ref{sec:hylo}).
+
+\subsection{Algebras}
+\label{sec:algebras}
+The most simple and well-known recursion scheme is the \emph{fold} or
 \emph{catamorphism} which consumes data. Its dual is the \emph{unfold} or
@@
 
-For example, assuming the type of lists over a type \coq{A},
-namely \coq{List A}, and given an
-algebra \coq{alg : unit + A * List A -> List A} a
-catamorphism \coq{cata alg : List A -> B} destructs a list into its head and
-tail using a destructor \coq{InOp : List A -> unit + A * List A}. This latter is
-actually an isomorphim with its inverse
-\begin{minted}{coq}
-In : unit + A * List A -> List A
-\end{minted}
-
-defined by the maps \coq{Empty : unit -> List A}
-and \coq{Cons : A * List A -> List A}.
-
-After destructing the input the catamorphism is applying itsefl recursively.
-This is done using functoriality of the shape of the data, which is a map of
-type
-\begin{minted}{coq}
-unit + A * List A -> unit + A * List B
-\end{minted}
-Once the recursion as returned it result it applies the algebra \coq{alg} which
-composes together the results of the recursion yielding the result of
-type \coq{B}.
-
-More abstractly, assuming a base functor \coq{F} a catamorphism is the unique
+A fold consumes elements of an inductive data type. These can be thought as
+least ``sets'' generated by some constructors. For example, the type of lists
+over a type \coq{A}, namely \coq{List A} is the ``least set'' generated by the
+constructors \coq{Empty : unit -> List A} and \coq{Cons : A * List A -> List A}.
+Together, \coq{Empty} and \coq{Cons} form an \emph{algebra} of
+type \coq{In : unit + A * List A -> List A}. In particular, this algebra is an
+isomorphism with the obvious inverse \coq{InOp : List A -> unit + A * List A}
+returning \coq{tt} is the list is empty and the head and tail of the list
+otherwise. This type of algebra is called \emph{initial algebra}.  This is
+because for for any other algebra \coq{alg : unit + A * List B -> B} there
+exists a unique map which we call \coq{cata alg : List A -> B} which destructs a
+list into its head and tail using a
+destructor \coq{InOp : List A -> unit + A * List A}, applies itself recursively,
+composes together the results of the recursion using the algebra \coq{alg}.
+
+To generalise recursion schemes we abstract the shape of the data using functors.
+Assuming a base functor \coq{F} a catamorphism is then the unique
 solution to the equation
@@
 \end{minted}
+which does precisely what we described in the previous paragraph.
 
@@
 
-% C
-Hylomorphism are more general recursion schemes capturing the
-\emph{divide-and-conquer} idea. In short, the input is destructured by its not
-structurally. In the example, we gave in the introduction the destructuring
-program is given by \coq{divide} which is not an isomorphism.  The question the
-hylomorphism satisfies is more general than the catamorphism in that the
-destructing map \coq{InOp} is generalised to a \emph{recursive
-coalgebra}~\cite{AdamekMM19,CaprettaUV04}, that is intuitively a coalgebra that
-can be applied only a finite amount of times.  In fact, in the case
-of \coq{divide} this is a recursive coalgebra because the size of the list is
-getting smaller.
-
-% This program is formed of essentially two parts. The first one is what is
-% commonly thought as the \emph{partition} function destructuring the input list
-% in three parts, the pivot and the other two sublists, and the second part
-% recursing over the sublists and composing the results back together.
-
-With this in mind we know proceed onto describing how to formalise hylomorphisms
-in Coq.
-
-\subsection{Algebras}
-Algebras on container extensions are defined as expected. The carrier of the
-algebra is \coq{A}, and the operations of the algebra are morphisms of the
-following type:
+Given a type `\coq{A}` and a container \coq{F} an algebra for \coq{F} over the
+carrier `\coq{A}` is a map of the following type:
 \begin{coqcode}
@@
 \end{coqcode}
-Given a container extension \coq{F}, we define \emph{two}
-(isomorphic) initial algebras. We will show the need for the second initial
-algebra in the following section. Our fist initial algebra is defined as the
-straightforward least fixpoint of the container extension \coq{F}:
+When \coq{A} is the least fixed-point of \coq{F} this map is an isomorphism and
+it is called the \emph{initial algebra} and denotes the inductive data type shaped by the functor represented by \coq{F}:
 \begin{coqcode}
@@
 the contained elements generically (recall Section~\ref{sec:TODO}): the
-contained types may not have (yet) an associated equivalence. In other words,
-we would need to provide an equivalence for \coq{LFix}, but the
-only way to do this is to define such equivalence simultaneously to the
-definition for  \coq{LFix}. Such definition would be rejected by Coq.
-There are other approaches that can circumvent this problem, by separating
-the definition of least fixed points from the proofs that the functions that
-get the contained elements are proper morphisms. But this leads to longer,
-unnecessarily complicated proofs.  Furthermore, this extra layer interferes
-with code extraction, in that it requires destructing/constructing subsequent
-\coq{LFix} and  \coq{App}, and Coq may not always able to inline them.
+contained types may not have (yet) an associated equivalence. In other words, we
+would need to provide an equivalence for \coq{LFix}, but the only way to do this
+is to define such equivalence simultaneously to the definition for  \coq{LFix}.
+Such definition would be rejected by Coq. There are other approaches that can
+circumvent this problem, by separating the definition of least fixed points from
+the proofs that the functions that get the contained elements are proper
+morphisms. But this leads to longer, unnecessarily complicated proofs.
+Furthermore, this extra layer interferes with code extraction, in that it
+requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
+may not always able to inline them.
 
@@
 \subsection{(Recursive) Coalgebras}
+\label{sec:rec-coalgebras}
+
+Hylomorphism are more general recursion schemes capturing the
+\emph{divide-and-conquer} idea. In short, the input is destructured by its not
+structurally. In the example, we gave in the introduction the destructuring
+program is given by \coq{divide} which is not an isomorphism.  The question the
+hylomorphism satisfies is more general than the catamorphism in that the
+destructing map \coq{InOp} is generalised to a \emph{recursive
+coalgebra}~\cite{AdamekMM19,CaprettaUV04}, that is intuitively a coalgebra that
+can be applied only a finite amount of times.  In fact, in the case
+of \coq{divide} this is a recursive coalgebra because the size of the list is
+getting smaller.
+
+% This program is formed of essentially two parts. The first one is what is
+% commonly thought as the \emph{partition} function destructuring the input list
+% in three parts, the pivot and the other two sublists, and the second part
+% recursing over the sublists and composing the results back together.
 
@@
 
-\subsection{Recursive Hylomorphisms}
-
-Given an $F$-algebra and $F$-coalgebra, a solution to the equation $f = a \circ
-F\; f \circ c$ is known as a hylomorphism. Unfortunately, this definition
-cannot be accepted by Coq, since there is no structurally smaller argument to
-do recursion, and there is no corecursive definition that is guaranteed to be
-syntactically guarded (as required by Coq), due to the use of an arbitrary
-algebra $a$. Furthermore, solutions to
-the hylomorphism equation are \emph{not} guaranteed to be unique, or to even
-exist. To address these issues, we mechanise \textbf{recursive hylomorphisms}
-which are guaranteed to have a unique solution to the hylomorphism equation.
+\subsection{Hylomorphisms}
+\label{sec:hylo}
+
+Given an $F$-algebra and $F$-coalgebra, a solution to the equation
+$f = a \circ F\; f \circ c$ is known as a hylomorphism. Unfortunately, this
+definition cannot be accepted by Coq, since there is no structurally smaller
+argument to do recursion, and there is no corecursive definition that is
+guaranteed to be syntactically guarded (as required by Coq), due to the use of
+an arbitrary algebra $a$. Furthermore, solutions to the hylomorphism equation
+are \emph{not} guaranteed to be unique, or to even exist. To address these
+issues, we mechanise \textbf{recursive hylomorphisms} which are guaranteed to
+have a unique solution to the hylomorphism equation.
 
 If $c$ is a recursive coalgebra, then the equation $f = a \circ F\; f \circ c$
-does have a unique solution.  We show this by first providing the definition
-for recursive hylomorphisms in Coq, and then proving that any function
-satisfying this equation is a hylomorphism. We call these hylomorphisms with a
-recursive coalgebra, \textbf{recursive hylomorphisms}.
+does have a unique solution.  We show this by first providing the definition for
+recursive hylomorphisms in Coq, and then proving that any function satisfying
+this equation is a hylomorphism. We call these hylomorphisms with a recursive
+coalgebra, \textbf{recursive hylomorphisms}.
 \begin{coqcode}

### 2024-03-05 paper/paper.tex
Algebras and Coalgebras

@@
 \label{sec:algebras}
+An algebra is a set of abstract operations on a type \coq{X}. For example,
+give such a type \coq{X} we can define the list operations on \coq{X} as
+\coq{Empty : unit -> X} and \coq{Cons : A * X -> X}.
+
+We can generalise this using containers. Given a type `\coq{A}` and a
+container \coq{F} an algebra for \coq{F} over the carrier `\coq{A}` is a map of
+the following type:
+\begin{coqcode}
+Notation Alg F A := (App F A ~> A).
+\end{coqcode}
+
+The type of lists over a type \coq{A}, namely \coq{List A} is the ``least set''
+generated by the constructors \coq{Empty : unit -> List A}
+and \coq{Cons : A * List A -> List A}.  Together, \coq{Empty} and \coq{Cons}
+form an \emph{algebra} of type \coq{In : unit + A * List A -> List A}.
+
+When \coq{A} is the least fixed-point of \coq{F} this map is an isomorphism and
+it is called the \emph{initial algebra} and denotes the inductive data type
+shaped by the functor represented by \coq{F}:
+\begin{coqcode}
+Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
+\end{coqcode}
+In particular, this algebra is an isomorphism with the obvious
+inverse \coq{InOp : List A -> unit + A * List A} returning \coq{tt} is the list
+is empty and the head and tail of the list otherwise. This type of algebra is
+called \emph{initial algebra}.  This is because for for any other
+algebra \coq{alg : unit + A * List B -> B} there exists a unique map which we
+call \coq{cata alg : List A -> B} which destructs a list into its head and tail
+using a destructor \coq{InOp : List A -> unit + A * List A}, applies itself
+recursively, composes together the results of the recursion using the
+algebra \coq{alg}.
+
 The most simple and well-known recursion scheme is the \emph{fold} or
@@
 A fold consumes elements of an inductive data type. These can be thought as
-least ``sets'' generated by some constructors. For example, the type of lists
-over a type \coq{A}, namely \coq{List A} is the ``least set'' generated by the
-constructors \coq{Empty : unit -> List A} and \coq{Cons : A * List A -> List A}.
-Together, \coq{Empty} and \coq{Cons} form an \emph{algebra} of
-type \coq{In : unit + A * List A -> List A}. In particular, this algebra is an
-isomorphism with the obvious inverse \coq{InOp : List A -> unit + A * List A}
-returning \coq{tt} is the list is empty and the head and tail of the list
-otherwise. This type of algebra is called \emph{initial algebra}.  This is
-because for for any other algebra \coq{alg : unit + A * List B -> B} there
-exists a unique map which we call \coq{cata alg : List A -> B} which destructs a
-list into its head and tail using a
-destructor \coq{InOp : List A -> unit + A * List A}, applies itself recursively,
-composes together the results of the recursion using the algebra \coq{alg}.
-
-To generalise recursion schemes we abstract the shape of the data using functors.
-Assuming a base functor \coq{F} a catamorphism is then the unique
-solution to the equation
-\begin{minted}{coq}
-cata alg = alg . fmap (cata alg) . InOp
-\end{minted}
-which does precisely what we described in the previous paragraph.
+least ``sets'' generated by some constructors.
+% To generalise recursion schemes we abstract the shape of the data using functors.
+% Assuming a base functor \coq{F} a catamorphism is then the unique
+% solution to the equation
+% \begin{minted}{coq}
+% cata alg = alg . fmap (cata alg) . InOp
+% \end{minted}
+% which does precisely what we described in the previous paragraph.
 
@@
 % $\ana{c}$ respectively.
-
-Given a type `\coq{A}` and a container \coq{F} an algebra for \coq{F} over the
-carrier `\coq{A}` is a map of the following type:
-\begin{coqcode}
-Notation Alg F A := (App F A ~> A).
-\end{coqcode}
-When \coq{A} is the least fixed-point of \coq{F} this map is an isomorphism and
-it is called the \emph{initial algebra} and denotes the inductive data type shaped by the functor represented by \coq{F}:
-\begin{coqcode}
-Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
-\end{coqcode}
-This definition highlights why we cannot define equality of the positions and
-the contained elements generically (recall Section~\ref{sec:TODO}): the
-contained types may not have (yet) an associated equivalence. In other words, we
-would need to provide an equivalence for \coq{LFix}, but the only way to do this
-is to define such equivalence simultaneously to the definition for  \coq{LFix}.
-Such definition would be rejected by Coq. There are other approaches that can
-circumvent this problem, by separating the definition of least fixed points from
-the proofs that the functions that get the contained elements are proper
-morphisms. But this leads to longer, unnecessarily complicated proofs.
-Furthermore, this extra layer interferes with code extraction, in that it
-requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
-may not always able to inline them.
 
@@
 \end{coqcode}
+\subsection{Coalgebras}
+Terminal coalgebras are defined analogously to initial algebras (see Figure~\ref{fig:term_coalg}).
+\begin{figure}
+\begin{coqcode}
+Context `(F : Cont Sh Po).
+CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
+Definition GFixR : GFix -> GFix -> Prop :=...
+Instance GFix_setoid : setoid GFix := ...GFixR...
+Definition g_in : App F GFix ~> GFix := ...
+Definition g_out : GFix ~> App F GFix := ...
+\end{coqcode}
+\caption{Terminal Coalgebras}\label{fig:term_coalg}
+\end{figure}
+Their equivalence relation, and anamorphisms, and universal properties are
+easy to encode, in an analogous way to catamorphisms and initial algebras:
+\begin{coqcode}
+  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...
+
+  Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
+    : f =e ana h <-> f =e g_in \o fmap f \o h.
+\end{coqcode}
+This definition, however, prevents us from composing catamorphisms and
+anamorphisms: we cannot mix \coq{GFix} and \coq{LFix}.
+To address this issue, we consider another alternative definition
+for initial $F$-algebras.
+
+\subsubsection{On Equality}
+This definition highlights why we cannot define equality of the positions and
+the contained elements generically (recall Section~\ref{sec:TODO}): the
+contained types may not have (yet) an associated equivalence. In other words, we
+would need to provide an equivalence for \coq{LFix}, but the only way to do this
+is to define such equivalence simultaneously to the definition for  \coq{LFix}.
+Such definition would be rejected by Coq. There are other approaches that can
+circumvent this problem, by separating the definition of least fixed points from
+the proofs that the functions that get the contained elements are proper
+morphisms. But this leads to longer, unnecessarily complicated proofs.
+Furthermore, this extra layer interferes with code extraction, in that it
+requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
+may not always able to inline them.
 
@@
 
-Terminal coalgebras are defined analogously to initial algebras (see Figure~\ref{fig:term_coalg}).
-\begin{figure}
-\begin{coqcode}
-Context `(F : Cont Sh Po).
-CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
-Definition GFixR : GFix -> GFix -> Prop :=...
-Instance GFix_setoid : setoid GFix := ...GFixR...
-Definition g_in : App F GFix ~> GFix := ...
-Definition g_out : GFix ~> App F GFix := ...
-\end{coqcode}
-\caption{Terminal Coalgebras}\label{fig:term_coalg}
-\end{figure}
-Their equivalence relation, and anamorphisms, and universal properties are
-easy to encode, in an analogous way to catamorphisms and initial algebras:
-\begin{coqcode}
-  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...
-
-  Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
-    : f =e ana h <-> f =e g_in \o fmap f \o h.
-\end{coqcode}
-This definition, however, prevents us from composing catamorphisms and
-anamorphisms: we cannot mix \coq{GFix} and \coq{LFix}.
-To address this issue, we consider another alternative definition
-for initial $F$-algebras. In particular, we show that the terminal coalgebras
+In particular, we show that the terminal coalgebras
 for $F$, coupled with a proof that they are finite are also initial

### 2024-03-05 paper/paper.tex
Algebra.

@@
 
-\subsection{Algebras}
-\label{sec:algebras}
-An algebra is a set of abstract operations on a type \coq{X}. For example,
-give such a type \coq{X} we can define the list operations on \coq{X} as
-\coq{Empty : unit -> X} and \coq{Cons : A * X -> X}.
-
-We can generalise this using containers. Given a type `\coq{A}` and a
-container \coq{F} an algebra for \coq{F} over the carrier `\coq{A}` is a map of
-the following type:
+\subsection{Algebras and Catamorphisms}
+\label{sec:algebras} An algebra is a set \coq{X} together with some operations
+on it\footnote[1]{Sometimes and algebra comes equipped with an equational
+theory, e.g. the theory of monoids, but here we focus on algerbraically free
+algebras}.  For example, give such a type \coq{X} we can define the monoid
+operations as \coq{Empty : unit -> X} and \coq{Concat : X * X -> X}. Notice
+that \coq{unit + X * X} is a functor on \coq{X} which means we can equivalently
+describe this algebra as a map \coq{unit + X * X -> X} (together with the
+monoidal laws).  In this work we will be using algebras to describe the
+operations of a data type and so we will not be needing additional equations on
+these operations.
+
+Since in this work we will be considering only polynomial functors we need to
+formalise algebras in terms of containers and containers extensions. Given a
+type `\coq{A}` and a container \coq{F} an algebra for \coq{F} over the carrier
+`\coq{A}` is a map of the following type:
 \begin{coqcode}
@@
 
-The type of lists over a type \coq{A}, namely \coq{List A} is the ``least set''
-generated by the constructors \coq{Empty : unit -> List A}
-and \coq{Cons : A * List A -> List A}.  Together, \coq{Empty} and \coq{Cons}
-form an \emph{algebra} of type \coq{In : unit + A * List A -> List A}.
-
-When \coq{A} is the least fixed-point of \coq{F} this map is an isomorphism and
-it is called the \emph{initial algebra} and denotes the inductive data type
-shaped by the functor represented by \coq{F}:
+Now, the \emph{easiest} way to turn any type \coq{A} into a monoid is to define
+it as the least fixed-point of the functor \coq{unit + A + X * X} on \coq{X}.
+This is isomorphic to the type of lists over \coq{A} hence it is the least
+fixed-point for the functor \coq{unit + A * X} as well. This is the least
+``set'' generated by the constructors \coq{Empty : unit -> List A}
+and \coq{Cons : A * List A -> List A}. These constructors give an
+algebra \coq{LFix_in : unit + A * List A -> List A} which has and inverse which
+we  call \coq{LFix_out}.
+
+We now formalise this idea using container extensions.  For a container
+extension a data type is formalised as an inductive data type parametrised by
+the functor represented by \coq{F}:
 \begin{coqcode}
@@
 \end{coqcode}
-In particular, this algebra is an isomorphism with the obvious
-inverse \coq{InOp : List A -> unit + A * List A} returning \coq{tt} is the list
-is empty and the head and tail of the list otherwise. This type of algebra is
-called \emph{initial algebra}.  This is because for for any other
-algebra \coq{alg : unit + A * List B -> B} there exists a unique map which we
-call \coq{cata alg : List A -> B} which destructs a list into its head and tail
-using a destructor \coq{InOp : List A -> unit + A * List A}, applies itself
-recursively, composes together the results of the recursion using the
-algebra \coq{alg}.
-
-The most simple and well-known recursion scheme is the \emph{fold} or
-\emph{catamorphism} which consumes data. Its dual is the \emph{unfold} or
-\emph{anamorphisms} which produces data.
-
-A fold consumes elements of an inductive data type. These can be thought as
-least ``sets'' generated by some constructors.
 % To generalise recursion schemes we abstract the shape of the data using functors.
@@
 %\end{coqcode}
-We define the equality of initial algebras of container extensions recursively,
-by traversing the structure and equating the shapes and positions. This
-definition is equivalent to taking the fixpoint of \coq{AppR} in
-Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
-positions, the contents must be related by \coq{LFixR}.
-\begin{coqcode}
-Fixpoint LFixR (x y : LFix) : Prop :=
-  let f_x := LFix_out x in
-  let f_y := LFix_out y in
-  shape f_x =e shape f_y /\
-    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
-\end{coqcode}
-
-Least fixed points of container extensions come with two functions that are
-defined as wrappers around constructor \coq{LFix_in} and projection
+We define smart constructors for the isomorphism of fixed points
 \coq{LFix_out}.
@@
 
-We define \emph{catamorphisms} by recursively traversing their structure of the
-input \coq{LFix}. As usual, a catamorphism takes an \coq{F}-algebra with
-carrier \coq{A}, and produces a function from \coq{LFix F} to \coq{A}.
-\begin{coqcode}
-Definition cata_f (g : Alg F A) : LFix -> A
+The most simple and well-known recursion scheme is the \emph{fold} or
+\emph{catamorphism} which consumes data. First a fold \emph{destructs} a data
+type using \coq{LFix_out}, then calls itself recursively and then folds the
+results of the recursive call using an \coq{F}-algebra.
+\begin{coqcode}
+Definition cata_f (alg : Alg F A) : LFix -> A
 := fix f (x : LFix) :=
@@
       let (sx, kx) := ax in
-      g (MkApp sx (fun e => f (kx e)))
+      alg (MkApp sx (fun e => f (kx e)))
   end.
 \end{coqcode}
-It is easy to show that this function is a proper morphism, and we use this
-fact to encode them as a higher-order morphism, with the following type:
-\begin{coqcode}
-Definition cata : Alg F A ~> LFix ~> A := ...
-\end{coqcode}
-Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
-they are the unique solution to the equation \coq{f =e g \o fmap f \o l_out}).
-The proof is by induction on the input to \coq{f} and
-\coq{cata g}.
-\begin{coqcode}
-Lemma cata_univ `{eA : setoid A} (g : Alg F A) (f : LFix ~> A)
-    : f =e cata g <-> f =e g \o fmap f \o l_out.
-\end{coqcode}
-\subsection{Coalgebras}
-Terminal coalgebras are defined analogously to initial algebras (see Figure~\ref{fig:term_coalg}).
-\begin{figure}
+It is easy to show that this function is a morphism of \coq{F}-algebras. In fact,
+it is possible to define it as a map of the following type:
+\begin{coqcode}
+Alg F A ~> LFix ~> A
+\end{coqcode}
+
+We prove that catamorphisms satisfy the universal law, that is
+that there is a unique morphism of \coq{F}-algebras from the least fixed-point of \coq{F}:
+\begin{coqcode}
+Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
+    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
+\end{coqcode}
+
+\subsection{Coalgebras and Anamorphisms}
+Terminal coalgebras are defined analogously to initial algebras (see
+Figure~\ref{fig:term_coalg}).
+% \begin{figure}
 \begin{coqcode}
@@
 \end{coqcode}
-\caption{Terminal Coalgebras}\label{fig:term_coalg}
-\end{figure}
+% \caption{Terminal Coalgebras}\label{fig:term_coalg}
+% \end{figure}
 Their equivalence relation, and anamorphisms, and universal properties are
@@
 
-\subsection{(Recursive) Coalgebras}
+\mpav{Dumped here}
+We define the equality of initial algebras of container extensions recursively,
+by traversing the structure and equating the shapes and positions. This
+definition is equivalent to taking the fixpoint of \coq{AppR} in
+Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
+positions, the contents must be related by \coq{LFixR}.
+\begin{coqcode}
+Fixpoint LFixR (x y : LFix) : Prop :=
+  let f_x := LFix_out x in
+  let f_y := LFix_out y in
+  shape f_x =e shape f_y /\
+    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
+\end{coqcode}
+
+\subsection{Recursive Coalgebras and Hylomorphism}
 \label{sec:rec-coalgebras}
@@
 
-\subsection{Hylomorphisms}
-\label{sec:hylo}
+\subsubsection{Hylomorphisms}
 

### 2024-03-05 paper/paper.tex
On Coalgebras

@@
 \newcommand{\eqdef}{\stackrel{\mathrm{\Delta}}{=}}
+\newcommand{\eqiff}{\stackrel{\triangle}{\iff}}
 \newcommand{\bnfeq}{\mathrel{::=}}
@@
 \newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
-
 %\newcommand{\coq}[1]{\mintinline{coq}{#1}}
@@
 
-\section{Algebras, Coalgebras and Hylomorphisms}
+\section{Formalising Recursion Schemes}
 Recursion schemes provide an abstract way to consume and generate data. We know
@@
 algebras}.  For example, give such a type \coq{X} we can define the monoid
-operations as \coq{Empty : unit -> X} and \coq{Concat : X * X -> X}. Notice
-that \coq{unit + X * X} is a functor on \coq{X} which means we can equivalently
-describe this algebra as a map \coq{unit + X * X -> X} (together with the
-monoidal laws).  In this work we will be using algebras to describe the
-operations of a data type and so we will not be needing additional equations on
-these operations.
+operations as \coq{u : unit -> X} for unit and \coq{m : X * X -> X} for
+multiplication. Notice that \coq{FX = unit + X * X} is a functor on \coq{X}
+which means we can equivalently describe these two maps as a single one of
+type \coq{F X -> X} which we call an algebra for the functor \coq{F}
+satisfying the unit and associative laws of the monoid.  In this work we will be
+using algebras to describe the operations of a data type and so we will not be
+needing additional equations on these operations.
 
@@
 
-Now, the \emph{easiest} way to turn any type \coq{A} into a monoid is to define
-it as the least fixed-point of the functor \coq{unit + A + X * X} on \coq{X}.
-This is isomorphic to the type of lists over \coq{A} hence it is the least
-fixed-point for the functor \coq{unit + A * X} as well. This is the least
-``set'' generated by the constructors \coq{Empty : unit -> List A}
-and \coq{Cons : A * List A -> List A}. These constructors give an
-algebra \coq{LFix_in : unit + A * List A -> List A} which has and inverse which
-we  call \coq{LFix_out}.
+Now, the \emph{easiest} and most effective way to turn any type $A$ into a
+monoid is to define it as the least solution to the equation
+\begin{coqcode}
+Free A = A + unit + Free A * Free A
+\end{coqcode}
+with \coq{Free : Type -> Type}. In Coq, this is the least fixed-point of the
+functor \coq{unit + A + X * X} on \coq{X}.  This is isomorphic to the type of
+lists over \coq{A} hence it is the least fixed-point for the
+functor \coq{unit + A * X} as well. This is the least ``set'' generated by the
+constructors \coq{Empty : unit -> List A} and \coq{Cons : A * List A -> List A}.
+These constructors give an algebra \coq{LFix_in : unit + A * List A -> List A}
+which has and inverse which we call \coq{LFix_out}.
 
@@
 %\end{coqcode}
-We define smart constructors for the isomorphism of fixed points
+We define smart constructors for the isomorphism of least fixed points
 \coq{LFix_out}.
@@
 
-We prove that catamorphisms satisfy the universal law, that is
-that there is a unique morphism of \coq{F}-algebras from the least fixed-point of \coq{F}:
+We prove that catamorphisms satisfy the universal law, that is that there is a
+unique morphism of \coq{F}-algebras from the least fixed-point of \coq{F}:
 \begin{coqcode}
@@
 \subsection{Coalgebras and Anamorphisms}
+While folds \emph{destruct} data, their dual, the unfolds, \emph{construct}
+data.  In particular, given a seed value and an \emph{observation}, the unfold
+corecursively  ``runs'' the observations starting from the initial seed.
+
+% Consider the transition system below composed by three states \coq{x1}, \coq{x2}
+% and \coq{x3} with the transitions labelled by the set of labels \coq{\{a, b\}} and define the map \coq{c} such that it implements the transition system below
+% \[\begin{tikzcd}
+%     {x_1} & {x_2} & {x_3} \arrow["b"', loop, distance=2em, in=35, out=325]
+% 	\arrow["a", from=1-1, to=1-2]
+% 	\arrow["b", from=1-2, to=1-3]
+% \end{tikzcd}\]
+
+For example, for a type of states \coq{X} and an alphabet type \coq{L} we can
+define a (labelled) transition system (LTS) on \coq{X} as a
+function \coq{c : X -> L * X} implementing the \emph{transition} map. In
+particular, for a state \coq{x1 : X}, \coq{c x1} returns a
+pair \coq{(l, x2) : L * X} where \coq{l : L} is the observable action
+and \coq{x2 : X} is the next state. We could then add the following notation
+\mpav{is this notation duable in Cock?}
+\[
+  x_{1} \xrightarrow{l} x_{2} \eqiff c x_{1} = (l, x_{2})
+\]
+\begin{coqcode}
+Notation x1 --l> x2 := c(x1) = x2.
+\end{coqcode}
+
 Terminal coalgebras are defined analogously to initial algebras (see
 Figure~\ref{fig:term_coalg}).
-% \begin{figure}
-\begin{coqcode}
-Context `(F : Cont Sh Po).
-CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
-Definition GFixR : GFix -> GFix -> Prop :=...
-Instance GFix_setoid : setoid GFix := ...GFixR...
-Definition g_in : App F GFix ~> GFix := ...
-Definition g_out : GFix ~> App F GFix := ...
+
+\begin{coqcode}
+CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
+\end{coqcode}
+
+We define smart constructors for the isomorphism of greatest fixed points
+\begin{coqcode}
+g_in : App F GFix ~> GFix
+g_out : GFix ~> App F GFix
 \end{coqcode}
@@
 % \end{figure}
+
 Their equivalence relation, and anamorphisms, and universal properties are
@@
     (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
+\end{coqcode}
+
+\begin{coqcode}
+Definition GFixR : GFix -> GFix -> Prop :=...
+Instance GFix_setoid : setoid GFix := ...GFixR...
 \end{coqcode}

### 2024-03-05 paper/paper.tex
Coalgebra completed.

@@
 
-For example, for a type of states \coq{X} and an alphabet type \coq{L} we can
-define a (labelled) transition system (LTS) on \coq{X} as a
-function \coq{c : X -> L * X} implementing the \emph{transition} map. In
-particular, for a state \coq{x1 : X}, \coq{c x1} returns a
-pair \coq{(l, x2) : L * X} where \coq{l : L} is the observable action
-and \coq{x2 : X} is the next state. We could then add the following notation
-\mpav{is this notation duable in Cock?}
-\[
-  x_{1} \xrightarrow{l} x_{2} \eqiff c x_{1} = (l, x_{2})
-\]
-\begin{coqcode}
-Notation x1 --l> x2 := c(x1) = x2.
-\end{coqcode}
-
-Terminal coalgebras are defined analogously to initial algebras (see
-Figure~\ref{fig:term_coalg}).
-
+The seed function can be modelled by a coalgebra. For example, for a type of
+states \coq{X} and an alphabet type \coq{L} we can define a (labelled)
+transition system (LTS) on \coq{X} as a function \coq{c : X -> L * X}
+implementing the \emph{transition} map. In particular, for a
+state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
+where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
+Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
+carrier \coq{X} is a map \coq{X -> FX}. In the previous example \coq{c} is
+an \coq{F}-coalgebra on the carrier \coq{X} with \coq{FX = L * X}.
+
+Now an anamorphism takes such a seed function and runs it while collecting all
+the observable behaviours. In order to do so we define the type of infinite
+streams of labels which is the greatest fixed-point for the functor \coq{A * X}.
+We define this in Coq as follows:
 \begin{coqcode}
@@
 \end{coqcode}
-
-We define smart constructors for the isomorphism of greatest fixed points
+where \coq{GFix_in}  and \coq{GFix_out} are the two inverses witnessing the
+isomorphism. Additionally, we define smart constructors for the isomorphism of greatest fixed points
 \begin{coqcode}
@@
 \end{coqcode}
-% \caption{Terminal Coalgebras}\label{fig:term_coalg}
-% \end{figure}
-
-Their equivalence relation, and anamorphisms, and universal properties are
-easy to encode, in an analogous way to catamorphisms and initial algebras:
-\begin{coqcode}
-  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...
-
+
+What witnesses the fact that this is the greatest fixed-point is the universal
+property of the anamorphism. For a given \coq{F}-coalgebra this is the unique
+map that runs the seed function and composes the information from the first step
+the result
+
+\begin{coqcode}
 Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
@@
 \end{coqcode}
-This definition, however, prevents us from composing catamorphisms and
-anamorphisms: we cannot mix \coq{GFix} and \coq{LFix}.
-To address this issue, we consider another alternative definition
-for initial $F$-algebras.
+Equivalently, we could have define an anamorphism as the
+unique \coq{F}-coalgebra homomorphism, that is a map of the following type
+\begin{coqcode}
+Coalg F A ~> A ~> GFix
+\end{coqcode}
 

### 2024-03-06 paper/paper.tex
working on recursive coalgebras

@@
 
-\subsubsection{On Equality}
-This definition highlights why we cannot define equality of the positions and
-the contained elements generically (recall Section~\ref{sec:TODO}): the
-contained types may not have (yet) an associated equivalence. In other words, we
-would need to provide an equivalence for \coq{LFix}, but the only way to do this
-is to define such equivalence simultaneously to the definition for  \coq{LFix}.
-Such definition would be rejected by Coq. There are other approaches that can
-circumvent this problem, by separating the definition of least fixed points from
-the proofs that the functions that get the contained elements are proper
-morphisms. But this leads to longer, unnecessarily complicated proofs.
-Furthermore, this extra layer interferes with code extraction, in that it
-requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
-may not always able to inline them.
-
-\mpav{Dumped here}
-We define the equality of initial algebras of container extensions recursively,
-by traversing the structure and equating the shapes and positions. This
-definition is equivalent to taking the fixpoint of \coq{AppR} in
-Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
-positions, the contents must be related by \coq{LFixR}.
-\begin{coqcode}
-Fixpoint LFixR (x y : LFix) : Prop :=
-  let f_x := LFix_out x in
-  let f_y := LFix_out y in
-  shape f_x =e shape f_y /\
-    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
-\end{coqcode}
-
-\begin{coqcode}
-Definition GFixR : GFix -> GFix -> Prop :=...
-Instance GFix_setoid : setoid GFix := ...GFixR...
-\end{coqcode}
-
 \subsection{Recursive Coalgebras and Hylomorphism}
 \label{sec:rec-coalgebras}
-
-Hylomorphism are more general recursion schemes capturing the
-\emph{divide-and-conquer} idea. In short, the input is destructured by its not
-structurally. In the example, we gave in the introduction the destructuring
-program is given by \coq{divide} which is not an isomorphism.  The question the
-hylomorphism satisfies is more general than the catamorphism in that the
-destructing map \coq{InOp} is generalised to a \emph{recursive
-coalgebra}~\cite{AdamekMM19,CaprettaUV04}, that is intuitively a coalgebra that
-can be applied only a finite amount of times.  In fact, in the case
-of \coq{divide} this is a recursive coalgebra because the size of the list is
-getting smaller.
-
-% This program is formed of essentially two parts. The first one is what is
-% commonly thought as the \emph{partition} function destructuring the input list
-% in three parts, the pivot and the other two sublists, and the second part
-% recursing over the sublists and composing the results back together.
-
-In particular, we show that the terminal coalgebras
-for $F$, coupled with a proof that they are finite are also initial
-$F$-algebras:
-\begin{coqcode}
-Inductive FinF : GFix F -> Prop :=
-| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
-\end{coqcode}
-Catamorphisms from this initial algebra are denoted by \coq{ccata},
-and they can be used to define the isomorphism between
-\coq{{x : GFix F | FinF x}} and
-\coq{LFix F}. In particular:
-\begin{coqcode}
-Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
-Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
-\end{coqcode}
-Using this new initial algebra, we can compose catamorphisms and anamorphisms,
-\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
-the supplementary material and Coq code to see how we encode this. In this
-paper, however, we focus on the fact that, if anamorphisms produce finite
-trees, then their return type can also be the \emph{least} fixed point of the
-container extension. We encode in Coq the notion of \emph{recursive
-anamorphisms} and \emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  Recursive
-coalgebras are coalgebras whose anamorphism produces finite trees.  The
-following predicate captures this fact, \coq{h} is recursive if
-\coq{h} can only be applied finitely many times:
-\begin{coqcode}
-Inductive RecF (h : Coalg F A) : A -> Prop :=
-| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
-\end{coqcode}
-This predicate is equivalent to the proposition that the anamorphism of such
-coalgebras produce only finite trees:
-\begin{coqcode}
-Lemma ana_rec_term `{setoid A} (h : Coalg F A)
-  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
-\end{coqcode}
-Recursive anamorphisms are implemented in Coq recursively on the structure of
-the proof that they can only be applied finitely many times (\coq{RecF}).  Note
-that such proofs will be erased when extracting code to OCaml, and this is
-standard practice when building general recursive functions in
-Coq\footnote{\dcas{This technique had a name in the community, but I cannot
-remember. Something predicates}}. The core definitions are the following:
-\begin{coqcode}
-Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
-
-Definition rana_f__ `{setoid A} (c : Coalg F A)
-  : forall x : A, RecF c x -> LFix F
-  := fix f (x : A) (H : RecF c x) :=
-       let hx := c x in
-       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
-\end{coqcode}
-Here, \coq{RecF_inv} provides the structurally smaller element for
-the Coq recursive definition. We then couple coalgebras with a proof that they
-are recursive, and add a coercion from recursive coalgebras to coalgebras.
-\begin{coqcode}
-Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
-\end{coqcode}
-In some cases, proving that coalgebras can only be applied finitely many
-times is straightforward (e.g.\ they are recursive on some inductive type in
-Coq). However, in many cases, proving that a coalgebra is recursive directly
-can be tedious and cumbersome. To ease development, we prove that if a
-given coalgebra respects a \emph{well-founded relation}, then it is
-necessarily recursive.
-\begin{coqcode}
-Definition respects_relation `{setoid A} (c : Coalg F A)
-      {B} (m : A -> B) (R : B -> B -> Prop)
-  := forall x (p : Pos (shape (c x))), R (m (cont (c x) p)) (m x).
-
-Lemma wf_coalg_rec `{setoid A} {B}
-  (m : A -> B) (R : B -> B -> Prop) (WF : well_founded R)
-  (c : Coalg F A) (RR : respects_relation c m R) : RecP c.
-\end{coqcode}
-In the above definition, \coq{m} is used as a form of ``metric''
-for termination, and the well-founded relation is on \coq{B}. This
-works equivalently to transporting a well-founded relation
-\coq{R}
-on
-\coq{B} to \coq{A} as
-\coq{R' := fun x y => R (m x) (m y)}, which is also trivially
-well-founded. Using these definitions, we can construct recursive coalgebras by
-proving that they respect a well-founded relation, which generally leads to
-simpler proofs.
-
-Similarly to the rest of the definitions, we prove that recursive anamorphisms
-are respectful, and use this fact to build them as higher order proper
-morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
-the unique solution to the equation
-\coq{f =e l_in \o fmap f \o h},
-where \coq{h} is a recursive coalgebra.
-\begin{coqcode}
-Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F :=...
-
-Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
-    : f =e rana h <-> f =e l_in \o fmap f \o h.
-\end{coqcode}
-
-\subsubsection{Hylomorphisms}
-
-Given an $F$-algebra and $F$-coalgebra, a solution to the equation
-$f = a \circ F\; f \circ c$ is known as a hylomorphism. Unfortunately, this
-definition cannot be accepted by Coq, since there is no structurally smaller
-argument to do recursion, and there is no corecursive definition that is
-guaranteed to be syntactically guarded (as required by Coq), due to the use of
-an arbitrary algebra $a$. Furthermore, solutions to the hylomorphism equation
-are \emph{not} guaranteed to be unique, or to even exist. To address these
-issues, we mechanise \textbf{recursive hylomorphisms} which are guaranteed to
-have a unique solution to the hylomorphism equation.
-
-If $c$ is a recursive coalgebra, then the equation $f = a \circ F\; f \circ c$
-does have a unique solution.  We show this by first providing the definition for
-recursive hylomorphisms in Coq, and then proving that any function satisfying
-this equation is a hylomorphism. We call these hylomorphisms with a recursive
-coalgebra, \textbf{recursive hylomorphisms}.
+Hylomorphism capture the concept of \emph{divide-and-conquer} algorithms.  First
+the input is destructured (\emph{divide}) in smaller parts. This operation may or may not be
+structural. Next the results from the subparts are computed recursively and
+finally the results are composed together (\emph{conquer}) by means of an algebra.
+
+A catamorphism is a special kind of hylomorphism where the coalgebra
+map \coq{InOp} only destructures the input structurally, for example,
+decomposing a list into its head and tail. A \emph{recursive
+coalgebra}~\cite{AdamekMM19,CaprettaUV04} is a coalgebra which does not
+necessarily decompose the input structurally, still maintaining the size of the
+input smaller. This is what, for example, happens in quicksort where the
+partition function, given a list, returns two smaller sublists and a pivot
+element.
+
+Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the unique solution to the equation
+\begin{equation}
+  \label{eq:hylo}
+  f = a \circ F\; f \circ c
+\end{equation}
+However this solution does not necessarily exists for an arbitrary pair of
+algebras and coalgebras.  In fact, this definition cannot be accepted by Coq,
+since there is no structurally smaller argument to do recursion on and there is
+no corecursive definition that is guaranteed to be syntactically guarded -- as
+required by Coq -- due to the use of an arbitrary algebra $a$. Furthermore,
+solutions to the hylomorphism equation are \emph{not} guaranteed to be unique,
+or to even exist. To address these issues, we mechanise \textbf{recursive
+hylomorphisms} which are guaranteed to have a unique solution to the
+hylomorphism equation.
+
+If $c$ is a recursive coalgebra, then the equation (\ref{eq:hylo}) has a unique
+solution. We show this by first providing the definition for recursive
+hylomorphisms in Coq, and then proving that any function satisfying this
+equation is a hylomorphism.
 \begin{coqcode}
@@
 passes}.
+
+\mpav{Moved here}
+In particular, we show that the terminal coalgebras
+for $F$, coupled with a proof that they are finite are also initial
+$F$-algebras:
+\begin{coqcode}
+Inductive FinF : GFix F -> Prop :=
+| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
+\end{coqcode}
+Catamorphisms from this initial algebra are denoted by \coq{ccata},
+and they can be used to define the isomorphism between
+\coq{{x : GFix F | FinF x}} and
+\coq{LFix F}. In particular:
+\begin{coqcode}
+Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
+Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
+\end{coqcode}
+Using this new initial algebra, we can compose catamorphisms and anamorphisms,
+\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
+the supplementary material and Coq code to see how we encode this. In this
+paper, however, we focus on the fact that, if anamorphisms produce finite trees,
+then their return type can also be the \emph{least} fixed point of the container
+extension. We encode in Coq the notion of \emph{recursive anamorphisms} and
+\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}. Recursive
+coalgebras are coalgebras whose anamorphism produces finite trees. The
+following predicate captures this fact, \coq{h} is recursive if \coq{h} can only
+be applied finitely many times:
+\begin{coqcode}
+Inductive RecF (h : Coalg F A) : A -> Prop :=
+| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
+\end{coqcode}
+This predicate is equivalent to the proposition that the anamorphism of such
+coalgebras produce only finite trees:
+\begin{coqcode}
+Lemma ana_rec_term `{setoid A} (h : Coalg F A)
+  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
+\end{coqcode}
+Recursive anamorphisms are implemented in Coq recursively on the structure of
+the proof that they can only be applied finitely many times (\coq{RecF}).  Note
+that such proofs will be erased when extracting code to OCaml, and this is
+standard practice when building general recursive functions in
+Coq\footnote{\dcas{This technique had a name in the community, but I cannot
+remember. Something predicates}}. The core definitions are the following:
+\begin{coqcode}
+Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
+
+Definition rana_f__ `{setoid A} (c : Coalg F A)
+  : forall x : A, RecF c x -> LFix F
+  := fix f (x : A) (H : RecF c x) :=
+       let hx := c x in
+       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
+\end{coqcode}
+Here, \coq{RecF_inv} provides the structurally smaller element for
+the Coq recursive definition. We then couple coalgebras with a proof that they
+are recursive, and add a coercion from recursive coalgebras to coalgebras.
+\begin{coqcode}
+Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
+\end{coqcode}
+In some cases, proving that coalgebras can only be applied finitely many
+times is straightforward (e.g.\ they are recursive on some inductive type in
+Coq). However, in many cases, proving that a coalgebra is recursive directly
+can be tedious and cumbersome. To ease development, we prove that if a
+given coalgebra respects a \emph{well-founded relation}, then it is
+necessarily recursive.
+\begin{coqcode}
+Definition respects_relation `{setoid A} (c : Coalg F A)
+      {B} (m : A -> B) (R : B -> B -> Prop)
+  := forall x (p : Pos (shape (c x))), R (m (cont (c x) p)) (m x).
+
+Lemma wf_coalg_rec `{setoid A} {B}
+  (m : A -> B) (R : B -> B -> Prop) (WF : well_founded R)
+  (c : Coalg F A) (RR : respects_relation c m R) : RecP c.
+\end{coqcode}
+In the above definition, \coq{m} is used as a form of ``metric''
+for termination, and the well-founded relation is on \coq{B}. This
+works equivalently to transporting a well-founded relation
+\coq{R}
+on
+\coq{B} to \coq{A} as
+\coq{R' := fun x y => R (m x) (m y)}, which is also trivially
+well-founded. Using these definitions, we can construct recursive coalgebras by
+proving that they respect a well-founded relation, which generally leads to
+simpler proofs.
+
+Similarly to the rest of the definitions, we prove that recursive anamorphisms
+are respectful, and use this fact to build them as higher order proper
+morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
+the unique solution to the equation
+\coq{f =e l_in \o fmap f \o h},
+where \coq{h} is a recursive coalgebra.
+\begin{coqcode}
+Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F :=...
+
+Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
+    : f =e rana h <-> f =e l_in \o fmap f \o h.
+\end{coqcode}
 
@@
 \end{figure}
+
+\section{On Equality}
+This definition highlights why we cannot define equality of the positions and
+the contained elements generically (recall Section~\ref{sec:TODO}): the
+contained types may not have (yet) an associated equivalence. In other words, we
+would need to provide an equivalence for \coq{LFix}, but the only way to do this
+is to define such equivalence simultaneously to the definition for  \coq{LFix}.
+Such definition would be rejected by Coq. There are other approaches that can
+circumvent this problem, by separating the definition of least fixed points from
+the proofs that the functions that get the contained elements are proper
+morphisms. But this leads to longer, unnecessarily complicated proofs.
+Furthermore, this extra layer interferes with code extraction, in that it
+requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
+may not always able to inline them.
+
+\mpav{Dumped here}
+We define the equality of initial algebras of container extensions recursively,
+by traversing the structure and equating the shapes and positions. This
+definition is equivalent to taking the fixpoint of \coq{AppR} in
+Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
+positions, the contents must be related by \coq{LFixR}.
+\begin{coqcode}
+Fixpoint LFixR (x y : LFix) : Prop :=
+  let f_x := LFix_out x in
+  let f_y := LFix_out y in
+  shape f_x =e shape f_y /\
+    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
+\end{coqcode}
+
+\begin{coqcode}
+Definition GFixR : GFix -> GFix -> Prop :=...
+Instance GFix_setoid : setoid GFix := ...GFixR...
+\end{coqcode}
 

### 2024-03-06 paper/paper.tex
Recursive anamorphisms are hylomorphisms on the initial algebra.

@@
 an \coq{F}-coalgebra on the carrier \coq{X} with \coq{FX = L * X}.
+In our development we use the following notation for coalgebras:
+\begin{coqcode}
+Notation Coalg F A := (A ~> App F A) (only parsing).
+\end{coqcode}
 
@@
 Hylomorphism capture the concept of \emph{divide-and-conquer} algorithms.  First
-the input is destructured (\emph{divide}) in smaller parts. This operation may or may not be
-structural. Next the results from the subparts are computed recursively and
-finally the results are composed together (\emph{conquer}) by means of an algebra.
+the input is destructured (\emph{divide}) in smaller parts. This operation may
+or may not be structural. Next the results from the subparts are computed
+recursively and finally the results are composed together (\emph{conquer}) by
+means of an algebra.
 
@@
 
-Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the unique solution to the equation
+Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
+unique solution to the equation
 \begin{equation}
@@
 \end{equation}
+
 However this solution does not necessarily exists for an arbitrary pair of
@@
 solutions to the hylomorphism equation are \emph{not} guaranteed to be unique,
-or to even exist. To address these issues, we mechanise \textbf{recursive
+or to even exist. To address these issues, we mechanise \emph{recursive
 hylomorphisms} which are guaranteed to have a unique solution to the
-hylomorphism equation.
-
-If $c$ is a recursive coalgebra, then the equation (\ref{eq:hylo}) has a unique
-solution. We show this by first providing the definition for recursive
-hylomorphisms in Coq, and then proving that any function satisfying this
-equation is a hylomorphism.
+hylomorphism equation. These are hylomoprhisms where the coalgebra is
+\emph{recursive}.
+
+Recursive coalgebras are coalgebras whose anamorphism produces finite data which
+what the following predicate captures:
+\begin{coqcode}
+Inductive RecF (h : Coalg F A) : A -> Prop :=
+| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
+\end{coqcode}
+Recursive anamorphisms are implemented in Coq recursively on the structure of
+the proof for the type (\coq{RecF}) as follows:
 \begin{coqcode}
@@
 \end{coqcode}
-The definition of recursive hylomorphisms is recursive on the proof that the
-anamorphism can only be applied finitely many times on any input. We again use
-\coq{RecF_inv} to obtain the structurally smaller proof to use in
-the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
-we prove that \coq{hylo_def} is respectful, and use this proof to
-built the corresponding higher-order proper morphism:
-\begin{coqcode}
-Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
-\end{coqcode}
-Finally, we show that recursive hylomorphisms are the unique solution to
-the hylomorphism equation.
+We again use \coq{RecF_inv} to obtain the structurally smaller proof to use in
+the recursive calls. As we did with catamorphisms and recursive anamorphisms, we
+prove that \coq{hylo_def} is respectful, and use this proof to built the
+corresponding higher-order proper morphism:
+\begin{coqcode}
+Alg F B ~> RCoalg F A ~> A ~> B
+\end{coqcode}
+Finally, we show that recursive hylomorphisms are the unique solution to the
+hylomorphism equation.
 \begin{coqcode}
@@
 \end{coqcode}
-This result can be used to prove hylomorphism laws in a way that is analogous
-to pen-and-paper program calculation techniques. We show the rule for
-hylomorphism fusion as an example:
+Hylomorphisms fusion falls out from the universal property:
 \begin{coqcode}
@@
   (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
-
-  rewrite hylo_univ, fmap_comp, compA, <- E2,
-          <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
-Qed.
-  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
-  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
-  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
-  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
-  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
-  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
-  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
-  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
-  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
-\end{coqcode}
+\end{coqcode}
+% Qed.
+%   rewrite hylo_univ, fmap_comp, compA, <- E2,
+%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
+
+  % rewrite hylo_univ, fmap_comp, compA, <- E2,
+  %         <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
+
+  % (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
+  % (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
+  % (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
+  % (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
+  % (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
+  % (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
+  % (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
+  % (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
+  % (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
 Using the hylo fusion law, we can prove the well-known \emph{deforestation
-optimisation}, where two consecutive recursive computations, one that builds a
-data structure, and another one that consumes it, can be fused together into a
-single recursive definition. This, in turn, allows us to prove that a recursive
-hylomorphism is the composition of a catamorphism and a recursive anamorphism.
+optimisation}. This is when two consecutive recursive computations, one that
+builds a data structure, and another one that consumes it, can be fused together
+into a single recursive definition. This, in turn, allows us to prove that a
+recursive hylomorphism is the composition of a catamorphism and a recursive
+anamorphism.
 \begin{coqcode}
@@
   : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
-
-Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
-  (g : Alg F B) (h : RCoalg F A)
-: cata g \o rana h =e hylo g h.
-\end{coqcode}
-This result will allow us to \emph{calculate} programs by providing small, easy
-to verify implementations, and combining them into efficient recursive
-functions by \emph{using Coq's proofs of hylomorphism laws as optimisation
-passes}.
-
-\mpav{Moved here}
-In particular, we show that the terminal coalgebras
-for $F$, coupled with a proof that they are finite are also initial
-$F$-algebras:
+\end{coqcode}
+
+\subsection{Recursive Anamorphisms}
+Anamorphisms specialise to an hylomorphism into an inductive data type.
+Intuitively, this is because a recursive coalgebra can be applied only finitely
+many times, so when this is applied to an anamorphism the only possible traces
+it can produce are the finite ones.  We show then that an anamorphisms applied
+to a recursive coalgebra is in one-to-one correspondence with an hylomorphism
+applied to the same recursive coalgebra and the \coq{In} map which is a map into
+the inductive data type defined by \coq{In}.
+
+To do this we first define a predicate which holds if and only if the element of
+a coinductive data type is finite:
 \begin{coqcode}
@@
 \end{coqcode}
-Catamorphisms from this initial algebra are denoted by \coq{ccata},
-and they can be used to define the isomorphism between
-\coq{{x : GFix F | FinF x}} and
-\coq{LFix F}. In particular:
+Catamorphisms from this initial algebra are denoted by \coq{ccata}, and they can
+be used to define the isomorphism between \coq{{x : GFix F | FinF x}}
+and \coq{LFix F}. In particular:
 \begin{coqcode}
@@
 extension. We encode in Coq the notion of \emph{recursive anamorphisms} and
-\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}. Recursive
-coalgebras are coalgebras whose anamorphism produces finite trees. The
-following predicate captures this fact, \coq{h} is recursive if \coq{h} can only
-be applied finitely many times:
-\begin{coqcode}
-Inductive RecF (h : Coalg F A) : A -> Prop :=
-| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
-\end{coqcode}
+\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.
+
+\begin{coqcode}
+Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
+  (g : Alg F B) (h : RCoalg F A)
+: cata g \o rana h =e hylo g h.
+\end{coqcode}
+This result will allow us to \emph{calculate} programs by providing small, easy
+to verify implementations, and combining them into efficient recursive functions
+by \emph{using Coq's proofs of hylomorphism laws as optimisation passes}.
+
 This predicate is equivalent to the proposition that the anamorphism of such
@@
 \end{coqcode}
-Recursive anamorphisms are implemented in Coq recursively on the structure of
-the proof that they can only be applied finitely many times (\coq{RecF}).  Note
-that such proofs will be erased when extracting code to OCaml, and this is
+
+\mpav{Moved here}
+
+Note that such proofs will be erased when extracting code to OCaml, and this is
 standard practice when building general recursive functions in

### 2024-03-07 paper/paper.tex
Some progress on Recursive anamorphisms

@@
 many times, so when this is applied to an anamorphism the only possible traces
-it can produce are the finite ones.  We show then that an anamorphisms applied
-to a recursive coalgebra is in one-to-one correspondence with an hylomorphism
-applied to the same recursive coalgebra and the \coq{In} map which is a map into
-the inductive data type defined by \coq{In}.
-
-To do this we first define a predicate which holds if and only if the element of
-a coinductive data type is finite:
+it can produce are the finite ones.  In particular, we can show that
+anamorphisms of type \coq{X -> GFix F} applied to a recursive \coq{F}-coalgebra
+on \coq{X} are in one-to-one correspondence with hylomorphisms of
+type \coq{X -> LFix F} on the same recursive \coq{F}-coalgebra and the
+algebra \coq{In} for the inductive data type \coq{LFix F}.
+
+To do this we first define a predicate on elements of coinductive data
+structures which holds if and only if the element is finite:
 \begin{coqcode}
@@
 \end{coqcode}
-Catamorphisms from this initial algebra are denoted by \coq{ccata}, and they can
-be used to define the isomorphism between \coq{{x : GFix F | FinF x}}
-and \coq{LFix F}. In particular:
+We can now define catamorphisms denoted by \coq{ccata_f_} from the
+subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
+any \coq{F}-algebra.
+\begin{coqcode}
+Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
+  : forall x : GFix F, FinF x -> A := fix f x H :=
+    let hx := g_out x in
+      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
+\end{coqcode}
+In particular, we can take a catamorphism from the finitary elements
+of \coq{GFix F} to the inductive data type \coq{LFix F} using
+the \coq{F}-algebra \coq{l_in} and viceversa using catamorphism on the
+restriction of \coq{g_in} to the finite elements of \coq{GFix}, which we denote
+by \coq{lg_in}.
 \begin{coqcode}
@@
 \end{coqcode}
-Using this new initial algebra, we can compose catamorphisms and anamorphisms,
-\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
-the supplementary material and Coq code to see how we encode this. In this
-paper, however, we focus on the fact that, if anamorphisms produce finite trees,
-then their return type can also be the \emph{least} fixed point of the container
-extension. We encode in Coq the notion of \emph{recursive anamorphisms} and
-\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.
+Moreover, using this new initial algebra, we can compose catamorphisms and
+recursive anamorphisms.
 
@@
 
-\mpav{Moved here}
-
-Note that such proofs will be erased when extracting code to OCaml, and this is
-standard practice when building general recursive functions in
-Coq\footnote{\dcas{This technique had a name in the community, but I cannot
-remember. Something predicates}}. The core definitions are the following:
-\begin{coqcode}
-Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
-
+\begin{coqcode}
+Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F :=...
+
+Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
+    : f =e rana h <-> f =e l_in \o fmap f \o h.
+\end{coqcode}
+
+\begin{coqcode}
 Definition rana_f__ `{setoid A} (c : Coalg F A)
@@
 \end{coqcode}
-Here, \coq{RecF_inv} provides the structurally smaller element for
-the Coq recursive definition. We then couple coalgebras with a proof that they
-are recursive, and add a coercion from recursive coalgebras to coalgebras.
-\begin{coqcode}
-Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
-\end{coqcode}
-In some cases, proving that coalgebras can only be applied finitely many
-times is straightforward (e.g.\ they are recursive on some inductive type in
-Coq). However, in many cases, proving that a coalgebra is recursive directly
-can be tedious and cumbersome. To ease development, we prove that if a
-given coalgebra respects a \emph{well-founded relation}, then it is
-necessarily recursive.
-\begin{coqcode}
-Definition respects_relation `{setoid A} (c : Coalg F A)
-      {B} (m : A -> B) (R : B -> B -> Prop)
-  := forall x (p : Pos (shape (c x))), R (m (cont (c x) p)) (m x).
-
-Lemma wf_coalg_rec `{setoid A} {B}
-  (m : A -> B) (R : B -> B -> Prop) (WF : well_founded R)
-  (c : Coalg F A) (RR : respects_relation c m R) : RecP c.
-\end{coqcode}
-In the above definition, \coq{m} is used as a form of ``metric''
-for termination, and the well-founded relation is on \coq{B}. This
-works equivalently to transporting a well-founded relation
-\coq{R}
-on
-\coq{B} to \coq{A} as
-\coq{R' := fun x y => R (m x) (m y)}, which is also trivially
-well-founded. Using these definitions, we can construct recursive coalgebras by
-proving that they respect a well-founded relation, which generally leads to
-simpler proofs.
-
-Similarly to the rest of the definitions, we prove that recursive anamorphisms
-are respectful, and use this fact to build them as higher order proper
-morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
-the unique solution to the equation
-\coq{f =e l_in \o fmap f \o h},
-where \coq{h} is a recursive coalgebra.
-\begin{coqcode}
-Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F :=...
-
-Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
-    : f =e rana h <-> f =e l_in \o fmap f \o h.
-\end{coqcode}
-
+
+% \mpav{Moved here}
+
+% Note that such proofs will be erased when extracting code to OCaml, and this is
+% standard practice when building general recursive functions in
+% Coq\footnote{\dcas{This technique had a name in the community, but I cannot
+% remember. Something predicates}}. The core definitions are the following:
+% \begin{coqcode}
+% Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
+
+% \end{coqcode}
+% Here, \coq{RecF_inv} provides the structurally smaller element for
+% the Coq recursive definition. We then couple coalgebras with a proof that they
+% are recursive, and add a coercion from recursive coalgebras to coalgebras.
+% \begin{coqcode}
+% Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
+% \end{coqcode}
+% In some cases, proving that coalgebras can only be applied finitely many
+% times is straightforward (e.g.\ they are recursive on some inductive type in
+% Coq). However, in many cases, proving that a coalgebra is recursive directly
+% can be tedious and cumbersome. To ease development, we prove that if a
+% given coalgebra respects a \emph{well-founded relation}, then it is
+% necessarily recursive.
+% \begin{coqcode}
+% Definition respects_relation `{setoid A} (c : Coalg F A)
+%       {B} (m : A -> B) (R : B -> B -> Prop)
+%   := forall x (p : Pos (shape (c x))), R (m (cont (c x) p)) (m x).
+
+% Lemma wf_coalg_rec `{setoid A} {B}
+%   (m : A -> B) (R : B -> B -> Prop) (WF : well_founded R)
+%   (c : Coalg F A) (RR : respects_relation c m R) : RecP c.
+% \end{coqcode}
+% In the above definition, \coq{m} is used as a form of ``metric''
+% for termination, and the well-founded relation is on \coq{B}. This
+% works equivalently to transporting a well-founded relation
+% \coq{R}
+% on
+% \coq{B} to \coq{A} as
+% \coq{R' := fun x y => R (m x) (m y)}, which is also trivially
+% well-founded. Using these definitions, we can construct recursive coalgebras by
+% proving that they respect a well-founded relation, which generally leads to
+% simpler proofs.
+
+% Similarly to the rest of the definitions, we prove that recursive anamorphisms
+% are respectful, and use this fact to build them as higher order proper
+% morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
+% the unique solution to the equation
+% \coq{f =e l_in \o fmap f \o h},
+% where \coq{h} is a recursive coalgebra.
 \section{Extraction}
-
 This section covers how to use our approach to calculate recursive programs

### 2024-03-07 paper/paper.tex
Recursive Anamorphisms

@@
 \subsection{Recursive Anamorphisms}
-Anamorphisms specialise to an hylomorphism into an inductive data type.
-Intuitively, this is because a recursive coalgebra can be applied only finitely
-many times, so when this is applied to an anamorphism the only possible traces
-it can produce are the finite ones.  In particular, we can show that
-anamorphisms of type \coq{X -> GFix F} applied to a recursive \coq{F}-coalgebra
-on \coq{X} are in one-to-one correspondence with hylomorphisms of
-type \coq{X -> LFix F} on the same recursive \coq{F}-coalgebra and the
-algebra \coq{In} for the inductive data type \coq{LFix F}.
-
-To do this we first define a predicate on elements of coinductive data
-structures which holds if and only if the element is finite:
-\begin{coqcode}
-Inductive FinF : GFix F -> Prop :=
-| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
-\end{coqcode}
-We can now define catamorphisms denoted by \coq{ccata_f_} from the
-subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
-any \coq{F}-algebra.
-\begin{coqcode}
-Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
-  : forall x : GFix F, FinF x -> A := fix f x H :=
-    let hx := g_out x in
-      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
-\end{coqcode}
-In particular, we can take a catamorphism from the finitary elements
-of \coq{GFix F} to the inductive data type \coq{LFix F} using
-the \coq{F}-algebra \coq{l_in} and viceversa using catamorphism on the
-restriction of \coq{g_in} to the finite elements of \coq{GFix}, which we denote
-by \coq{lg_in}.
-\begin{coqcode}
-Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
-Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
-\end{coqcode}
-Moreover, using this new initial algebra, we can compose catamorphisms and
-recursive anamorphisms.
-
-\begin{coqcode}
-Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
-  (g : Alg F B) (h : RCoalg F A)
-: cata g \o rana h =e hylo g h.
-\end{coqcode}
-This result will allow us to \emph{calculate} programs by providing small, easy
-to verify implementations, and combining them into efficient recursive functions
-by \emph{using Coq's proofs of hylomorphism laws as optimisation passes}.
-
-This predicate is equivalent to the proposition that the anamorphism of such
-coalgebras produce only finite trees:
-\begin{coqcode}
-Lemma ana_rec_term `{setoid A} (h : Coalg F A)
-  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
-\end{coqcode}
-
-\begin{coqcode}
-Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F :=...
-
-Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
-    : f =e rana h <-> f =e l_in \o fmap f \o h.
-\end{coqcode}
-
+Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
+inductive data type in the following way.  Intuitively, a recursive coalgebra
+can be applied only finitely many times, therefore when this is applied to an
+anamorphism the only possible traces it can produce from the seed function are
+the finite ones. We denote this special kind of recursion scheme \emph{recursive
+anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
+can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
+exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
+algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
+from the universality property of the hylomorphism and the fact that recursive
+anamorphisms satisfy the same equation.
+
+We now show how this is proven in the Coq development. First we define a
+recursive anamorphism as a function into \coq{LFix F}:
 \begin{coqcode}
@@
        LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
-\end{coqcode}
+
+Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F := ...
+\end{coqcode}
+We omit the boilerplate details to obtain the setoid morphism.
+The recursive anamorphism has the following universality property:
+\begin{coqcode}
+Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
+    : f =e rana h <-> f =e l_in \o fmap f \o h.
+\end{coqcode}
+We can now show that every hylomorphism is composed by a recursive anamorphism
+followed by a catamorphism:
+\begin{coqcode}
+Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
+  (g : Alg F B) (h : RCoalg F A)
+: cata g \o rana h =e hylo g h.
+\end{coqcode}
+This result will allow us to \emph{calculate} programs by providing small, easy
+to verify implementations, and combining them into efficient recursive functions
+by using Coq's proofs of hylomorphism laws as optimisation passes.
+
+\subsubsection{On the subtype of finite elements}
+In this development we have defined recursive anamorphisms on inductive data
+types.  We might have as well defined them on the subtype of finite elements of
+coinductive data types using a predicate which states when an element of a
+coinductive data type is finite:
+\begin{coqcode}
+Inductive FinF : GFix F -> Prop :=
+| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
+\end{coqcode}
+Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
+is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
+easy to see. We first define a catamorphism \coq{ccata_f_} from the
+subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
+any \coq{F}-algebra.
+\begin{coqcode}
+Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
+  : forall x : GFix F, FinF x -> A := fix f x H :=
+    let hx := g_out x in
+      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
+\end{coqcode}
+We now prove this is isomorphic to the least fixed-point of the functor \coq{F}
+We take the catamorphism from the finite elements of \coq{GFix F} to the
+inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}.
+Its inverse is the catamorphism on the restriction of \coq{g_in} to the finite elements of \coq{GFix}, which we denote by \coq{lg_in}.
+The following lemmas prove the isomorphism:
+\mpav{David, check this code works}
+\begin{coqcode}
+Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
+Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata l_in \o cata lg_in =e id.
+\end{coqcode}
+% This predicate is equivalent to the proposition that
+% the anamorphism of such coalgebras produce only finite trees:
+% \begin{coqcode}
+% Lemma ana_rec_term `{setoid A} (h : Coalg F A)
+%   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
+% \end{coqcode}
 

### 2024-03-07 paper/paper.tex
draft of abstract

--- 2024-03-07 paper/paper.tex
@@
 \begin{abstract}
-
+  Generic programming and recursion schemes provide powerful abstractions for
+reasoning about program equivalences and deriving optimisations. However,
+formalizing such abstractions in Coq often requires compromises, such as
+imposing performance penalties, requiring the assumption of additional axioms,
+or introducing unsafe casts into extracted code (for example, \ocaml{Obj.magic}
+in OCaml). This paper presents a fully axiom-free formalization of hylomorphisms
+(a recursion scheme capturing divide-and-conquer algorithms) in Coq, along with
+their algebraic laws. The formalization allows for the extraction of clean,
+natural-looking OCaml code for a variety of recursive programs.
+% while enabling
+% formal reasoning about program equivalences using the laws of hylomorphisms.
+
+  The key contribution of this paper is an axiom-free framework for generic
+programming with recursion schemes in Coq, focusing on proving algebraic laws,
+reasoning about program equivalences, and extracting safe, idiomatic code. With
+this framework, we formalize a series of divide-and conquer, dynamic
+programming, and mutual recursion programs as hylomorphisms. We demonstrate that
+the extracted OCaml code for the programs formalized in our framework is
+efficient, resembles code that a human programmer might write, and contains no
+uses of \ocaml{Obj.magic}. We also present a machine-checked proof of the
+well-known short-cut fusion optimsation.
+
+  By avoiding undafe coersions in the extracted code, this work provides a solid
+foundation for developing efficient, verified recursive programs with strong
+guarantees about the safety and interoperability of the generated OCaml code.
 \end{abstract}

### 2024-03-12 paper/paper.tex
typo
Examples

@@
 \newminted{coq}{fontsize=\small}
+\newminted{ocaml}{fontsize=\small}
 
@@
 uses of \ocaml{Obj.magic}. We also present a machine-checked proof of the
-well-known short-cut fusion optimsation.
+well-known short-cut fusion optimisation.
 
@@
 since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
-will be the function \coq{target} that is extensionally equal to \coq{f}.
+will be the function \coq{target} that is extensionally equal to \coq{f}.  We
+use a shortcut tactic, \coq{calculate}, that is a synonym to 
+\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
+\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
+complete the definition with \coq{reflexivity}, and this will instantiate the
+metavariable \coq{?Goal0} to our desired function. We will use this extensively
+in our examples.
+
+\subsection{Shortcut Deforestation}
+
+\subsubsection{Example 1}
+Shortcut deforestation can be expressed succintly in terms of hylomorphisms and
+their laws~\cite{TakanoM95}. In particular, given a function:
+\begin{coqcode}
+  s  : forall A. (App F A -> A) -> (App F A -> A)
+\end{coqcode}
+We can conclude, by parametricity, that
+\begin{coqcode}
+  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
+\end{coqcode}
+This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
+\textbf{not} provable in Coq if we want to remain axiom-free, since we would
+need to add the necessary parametricity
+axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
+
+However, we prove a specific version of this theorem for the list generating
+container (i.e.\ the container whose least fixed point is a list), and use it
+to encode the example from Takano and Meijer~\cite{TakanoM95}:
+\begin{coqcode}
+Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
+\end{coqcode}
+Here, we are defining function \coq{sf1} as the composition of \coq{length},
+\coq{Lmap f} and \coq{append ys}. Functions \coq{length} and \coq{Lmap} are
+catamorphisms. Function \coq{append ys} is also a catamorphism that
+appends \coq{ys} to an input list. It is defined by applying an algebra to
+every cons node of a list, and applying a catamorphism with the input algebra
+to \coq{ys} in the nil case:
+\begin{coqcode}
+Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
+  fun x => match x with | MkCont sx kx => match sx with
+  | s_nil => fun _ => (hylo a ilist_coalg) l
+  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
+  end kx end.
+Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
+\end{coqcode}
+Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
+\coq{ListF} container. We apply the hylo fusion law repeatedly, unfold
+definitions, and simplify in our specification for \coq{sf1}: 
+\begin{coqcode}
+Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
+  rewrite hylo_map_fusion, <- acid_rain.
+  simpl; reflexivity.
+Defined.
+\end{coqcode}
+From this, we extract the following OCaml code:
+\begin{ocamlcode}
+let rec sf1 f ys = function | [] -> let rec f0 = function
+                                      | [] -> (Uint63.of_int (0))
+                                      | _ :: t -> add (Uint63.of_int (1)) (f0 t)
+                                    in f0 ys
+                            | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
+\end{ocamlcode}
+
+\subsubsection{Example 2} We then prove that the length function fuses with
+the naive quadratic \emph{reverse} function:
+\begin{coqcode}
+Definition sf2 : Ext (length \o reverse). 
+  calculate.  unfold length, reverse. rewrite hylo_fusion_l. 
+  2:{ (* Rewrite into the fused version *) }
+  simpl; reflexivity.
+Defined.
+\end{coqcode}
+This code extracts to the optimised length function on the input list:
+\begin{ocamlcode}
+let rec ex2 = function | [] -> (Uint63.of_int (0))
+                       | _ :: t -> add (Uint63.of_int (1)) (ex2 t)
+\end{ocamlcode}
 
 \subsection{Sorting Algorithms}
-\dcas{Talk About this}
+
+So far, our examples could be modelled exclusively using catamorphisms. Here,
+we show examples that require recursive hylomorphisms and termination proofs.
+
+\subsubsection{Mergesort}
+
 \begin{minted}{OCaml}
@@
 
+\subsubsection{Quicksort}
+
+\subsubsection{Fusing a divide-and-conquer computation}
+
 \subsection{Knapsack}
-
-\subsection{Harper's Regular-Expression Matcher}
-
-\subsection{Shortcut Deforestation}
-
-\dcas{foldr/build examples from original paper (or hylo papers)}
 

### 2024-03-12 paper/paper.tex
Examples

@@
 we show examples that require recursive hylomorphisms and termination proofs.
+We complete the sorting algorithm examples by applying fusion optimisation.
+
+\subsubsection{Binary Trees}
+
+Both Mergesort and Quicksort are hylomorphisms whose recursive structure is
+described by a binary tree. In particular, the container used in their
+definition is isomorphic to \coq{TreeF (list nat) unit} and 
+\coq{TreeF unit nat} respectively, where \coq{TreeF} is defined as follows:
+\begin{coqcode}
+Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
+\end{coqcode}
+The definition as a container uses the following shapes and positions:
+\begin{coqcode}
+Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
+Inductive Tpos := | Lbranch | Rbranch. 
+\end{coqcode}
+We define the tree container, \coq{TreeF}, in a straightforward way, by making
+the positions of type \coq{Tpos} only valid in \coq{Node}.
+We define a series of definitions for tree container constructors and 
+destructors:
+\begin{coqcode}
+Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.
+(* Converting TreeF to ITreeF *)
+Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
+Notation a_node x l r := (MkCont (Node _ x)
+  (fun p => match val p with | Lbranch => l | Rbranch => r end)).
+\end{coqcode}
 
@@
 
-\begin{minted}{OCaml}
+Mergesort is a hylomorphism with the recursive structure defined by the
+container \coq{(TreeF (list int) unit)}. 
+The computationally expensive part of the Mergesort implementation is the
+algebra, since this is where most of the comparisons happen. The recursive
+coalgebra for Mergesort is defined as the following Coq functions:
+\begin{coqcode}
+Fixpoint splitL (x : list int) (accL accR : list int) :=
+  match x with
+  | nil => (accL, accR)
+  | cons x xs => splitL xs accR (cons x accL)
+  end.
+
+Definition c_split : Coalg (TreeF (list int) unit) (list int) :=
+|{ x ~> match x with
+        | nil | cons _ nil => a_leaf x
+        | _ => let (l, r) := splitL x nil nil in
+               a_node tt l r
+        end
+ }|.
+Defined.
+\end{coqcode}
+Here, the notation \coq{|{ x ~> E }|} defines the morphism \coq{fun x => E},
+and uses a custom tactic to prove that it is respectful. The recursive
+coalgebra \coq{c_split} takes a list, and it returns a leaf with the input
+list, if its size is less or equal than 1. Otherwise, it applies the function
+\coq{splitL}, and divides the list in two halves.
+
+It is easy to prove that this coalgebra is recursive, by showing that
+it respects the relation \coq{<} on the length of the lists:
+\begin{coqcode}
+Lemma split_fin : respects_relation c_split (@length int) lt.
+  Proof. (* *) Qed.
+
+Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
+  := mk_wf_coalg wf_lt split_fin.
+\end{coqcode}
+
+The algebra simply applies the \coq{mergeL} function in order to the sublists
+in every node:
+\begin{coqcode}
+Fixpoint mergeL l1 l2 {struct l1}  : list int := (* *)
+
+Definition merge : App (TreeF (list int) unit) (list int) ~> list int.
+|{ x ~> match a_out x with | node _ l r => mergeL l r | leaf e => e end }|.
+Defined.
+\end{coqcode}
+
+Putting everything together, Mergesort is defined as the following
+hylomorphism:
+\begin{coqcode}
+Definition msort := hylo merge tsplit.
+\end{coqcode}
+Note, however, that we could have composed the catamorphism and anamorphism,
+and used the deforestation result to optimise the intermediate binary tree
+away:
+\begin{coqcode}
+Definition msort : Ext (cata merge \o rana tsplit).
+calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
+\end{coqcode}
+In both cases, the result is the same, and it extracts to the usual divide and
+conquer definition of Mergesort in OCaml:
+\begin{ocamlcode}
 let rec msort x = match x with
@@
      mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
-\end{minted}
+\end{ocamlcode}
 
@@
 
+The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with
+the following algebra and coalgebra.
+\begin{coqcode}
+  Definition merge : App (TreeF unit int) (list int) ~> list int.
+|{ x : (App (TreeF unit int) (list int)) ~> (
+           match x with
+           | MkCont sx kx =>
+               match sx return (Container.Pos sx -> _) -> _ with
+               | Leaf _ _ => fun _ => nil
+               | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
+               end kx
+           end
+  )}|.
+Defined.
+
+Definition c_split : Coalg (TreeF unit int) (list int).
+|{ x ~> match x with
+        | nil => a_leaf tt
+        | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in
+                      a_node h l r
+        end}|.
+Defined.
+\end{coqcode}
+As with the Mergesort case, we prove that the coalgebra is recursive by showing
+that it respects the ``less-than'' relation on the length of the lists.
+We then define, and extract, the hylomorphism \coq{hylo merge r_split}, which
+extracts to the following OCaml code:
+\begin{ocamlcode}
+let rec qsort = function
+| [] -> [] | h :: t ->
+  let (l, r) = partition (fun x0 -> leb x0 h) t in
+  let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in 
+  app (x0 Lbranch) (h :: (x0 Rbranch))
+\end{ocamlcode}
+Note that Coq's code extraction is unable to inline \coq{x0}, but the resulting
+code is similar to a hand-written \coq{qsort}.
+
 \subsubsection{Fusing a divide-and-conquer computation}
 
+Suppose that we map a function to the result of sorting the list. We can
+use our framework to fuse both computations. In particular, consider
+the following definition:
+\begin{coqcode}
+Definition qsort_times_two : Ext (Lmap times_two \o hylo merge tsplit).
+  calculate.  apply hylo_fusion_l. 
+  (* calculate the fused stage *)
+Defined.
+\end{coqcode}
+Here, \coq{Lmap times_two} is a list map function defined as a hylomorphism,
+and \coq{times_two} multiplies every element of the list by two.
+After applying hylo fusion and the necessary rewrites, the hylomorphism that we
+extract is 
+\coq{hylo (merge \o natural times_two) tsplit}. Here, \coq{natural} defines
+a natural transformation between container extensions by applying a function
+to the shapes that preserves the validity of the positions. In this case,
+the function is \coq{times_two} that multiplies every pivot in the Quicksort
+tree by two. The extracted OCaml code is a single recursive traversal:
+\begin{ocamlcode}
+let rec qsort_times_two = function
+| [] -> []
+| h :: t ->
+  let (l, r) = partition (fun x0 -> leb x0 h) t in
+  let x0 = fun p -> qsort_times_two (match p with
+                                     | Lbranch -> l
+                                     | Rbranch -> r) in
+  app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
+\end{ocamlcode}
+
 \subsection{Knapsack}
+
+To close this section, we show how our framework can represent more advanced
+recursion schemes, such as \emph{dynamorphisms} for dynamic programming, by
+using their encoding as a hylomorphism. We use the knapsack example
+from~\cite{HinzeWG15}. First, we define memoisation tables in terms
+of containers. 
+\begin{coqcode}
+Definition MemoShape : Type := A * Sg.
+Definition MemoPos := Pg.
+Instance Memo : Cont MemoShape MemoPos := { valid := valid \o pair (snd \o fst) snd }.
+
+Definition Table := LFix Memo.
+\end{coqcode}
+Memoisation tables are the least fixed point of the container defined by shapes
+\coq{A * Sg} and positions \coq{Pg}, given a a container \coq{G : Cont Sg Pg}.
+We define a function to insert elements into the memoisation tables:
+\begin{coqcode}
+  Definition Cons : A * App G Table ~> App Memo Table := (* *)
+\end{coqcode}
+And two functions to inspect the head of a memoisation table, and remove an
+element from it:
+\begin{coqcode}
+Definition headT : Table ~> A := (* *)
+Definition tailT : Table ~> App G Table := (* *)
+\end{coqcode}
+Using these definitions, a dynamorphism is defined as follows:
+\begin{coqcode}
+Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
+:= headT \o hylo (l_in \o Cons \o pair a id) c.
+\end{coqcode}
+Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
+memoisation table. The definition of the algebra can use this table to lookup
+elements, instead of triggering a further recursive call. Elements are inserted
+into the memoisation table by the use of \coq{Cons} to the result of applying
+the algebra. The algebra for the dynamorphism looks up the previously computed
+elements to produce the result, thus saving the corresponding recursive calls:
+\begin{coqcode}
+Fixpoint memo_knap table wvs :=
+  match wvs with | nil => nil | h :: t =>
+      match lookupT (Nat.pred (fst h)) table with
+      | Some u => (u + snd h)%sint63 :: memo_knapsack table t
+      | None => memo_knapsack table t
+      end
+  end.
+
+Definition knapsack_alg (wvs : list (nat * int))
+  (x : App NatF (Table NatF int)) : int :=
+  match x with | MkCont sx kx => match sx with
+  | inl tt => fun kx => 0%sint63
+  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
+  end kx end.
+Definition knapsackA wvs : App NatF (Table NatF int) ~> int := 
+  (* [knapsack_alg wvs] as a respecful morphism *)
+\end{coqcode}
+The hylomorphism for knapsack is as follows, where \coq{out_nat} is the
+recursive coalgebra for \coq{nat}.
+\begin{coqcode}
+Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
+\end{coqcode}
+Coq's code extraction mechanism is unable to inline several definitions in
+this case, but the extracted code can be trivially inlined to produce the
+following:
+\begin{ocamlcode}
+let knapsack wvs x =
+  let (y, _) =
+    (let rec f x0 =
+      if x0=0 then Uint63.of_int (0)
+      else let fn := f (x0-1) in { lFix_out = { 
+           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
+           cont = fun _ -> fn } }
+     in f x).lFix_out.shape
+  in y
+\end{ocamlcode}
+Note how the recursive calls of \ocaml{f} build the memoisation table, and how
+this memoisation table is used to compute the intermediate results in
+\ocaml{memo_knap}, which is finally discarded to produce the final result.
 

### 2024-03-14 paper/paper.tex
Introduction

@@
 \begin{abstract}
-  Generic programming and recursion schemes provide powerful abstractions for
-reasoning about program equivalences and deriving optimisations. However,
-formalizing such abstractions in Coq often requires compromises, such as
-imposing performance penalties, requiring the assumption of additional axioms,
-or introducing unsafe casts into extracted code (for example, \ocaml{Obj.magic}
-in OCaml). This paper presents a fully axiom-free formalization of hylomorphisms
-(a recursion scheme capturing divide-and-conquer algorithms) in Coq, along with
-their algebraic laws. The formalization allows for the extraction of clean,
-natural-looking OCaml code for a variety of recursive programs.
-% while enabling
-% formal reasoning about program equivalences using the laws of hylomorphisms.
-
-The key contribution of this paper is an axiom-free framework for generic
-programming with recursion schemes in Coq, focusing on proving algebraic laws,
-reasoning about program equivalences, and extracting safe, idiomatic code. With
-this framework, we formalize a series of divide-and conquer, dynamic
-programming, and mutual recursion programs as hylomorphisms. We demonstrate that
-the extracted OCaml code for the programs formalized in our framework is
-efficient, resembles code that a human programmer might write, and contains no
-uses of \ocaml{Obj.magic}. We also present a machine-checked proof of the
-well-known short-cut fusion optimisation.
-
-By avoiding undafe coersions in the extracted code, this work provides a solid
-foundation for developing efficient, verified recursive programs with strong
-guarantees about the safety and interoperability of the generated OCaml code.
+  Generic programming with recursion schemes provides a powerful abstraction
+  for reasoning about program equivalences and deriving optimisations which has
+  been successfully applied to functional programming. However, formalising
+  this technique in a type theory, (e.g. Coq) often requires compromises, such
+  as imposing performance penalties, requiring the assumption of additional
+  axioms, or introducing unsafe casts into extracted code (for example,
+  \ocaml{Obj.magic} in OCaml).
+
+  This paper presents a formalisation of hylomorphisms (a recursion scheme
+  capturing divide-and-conquer algorithms) in Coq, along with their algebraic
+  laws. The key contribution of this paper is that this formalisation is fully
+  axiom-free, and it allows for the extraction of safe, clean, idiomatic OCaml
+  code for a variety of recursive programs. We formalise a series of divide-and
+  conquer, dynamic programming, and mutual recursion programs and demonstrate
+  that the extracted OCaml code for the programs formalised in our framework is
+  efficient, resembles code that a human programmer would write, and contains
+  no occurrences of \ocaml{Obj.magic}. We also present a machine-checked proof
+  of the well-known short-cut fusion optimisation.
 \end{abstract}
@@
 incomplete algorithm which syntactically checks for termination or
-productivity.  For recursion, this is done by trying to (automatically) infer
+productivity.  For recursion, this is done by automatically inferring
 which argument in the recursive call gets smaller with respects to the original
-input argument.  For productivity, the algorithm checks that the (co)recursive
+input argument.  For productivity, the algorithm checks that the corecursive
 call appears directly under a constructor to make sure that this function
-always produce at least one element after each recursive step.
-
-As mentioned, some functions, though well-defined, cannot be accepted by the
-proof assistant. One example is the quicksort algorith, written in OCaml below:
+always produces at least one element after each recursive step.
+This implies that some functions, though well-defined, cannot be accepted by
+the proof assistant. Examples of this include common sorting algorithms, such
+as Quicksort and Mergesort.
+%One example is Quicksort, written in OCaml
+%below:
 \begin{minted}{OCaml}
-  let divide = function
-    | [] -> None
-    | x::xs -> Some (x, List.partition (fun y -> y < x) xs);;
-  let rec qsort xs = match divide xs with
-    | None -> []
-    | Some (x, (smaller, larger)) -> qsort smaller @ (x::qsort larger)
+let rec qsort xs = match divide xs with | None -> []
+ | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
 \end{minted}
@@
 \emph{termination proofs}.  Using well-founded recursion, the recursive calls
-will happen on \emph{structurally smaller termination proofs}. A benefit of
-this approach is that, extracting verified nonstructural recursive functions to
-OCaml will erase the termination proofs, and so produce code that will be
-closer to what a programmer may have written directly in OCaml. However,
-reasoning about program equivalences requires dealing with such termination
-proofs, and it is common practice to use custom reduction lemmas that will be
-used extensively when proving properties  about the nonstructural recursive
-functions.
-
-A better approach for reasoning about program equivalences comes from
-\emph{structured recursion schemes}, and their associated algebraic laws.  In
-particular, \ocaml{qsort} is an instance of a \emph{divide-and-conquer}
+will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
+one approach is using the \coq{Fix} combinator, in which recursion is done on
+an \emph{accesibility} predicate on the input, which is a proof that there are
+no infinitely decreasing chains.}. A benefit of this approach is that,
+extracting verified nonstructural recursive functions to OCaml will erase the
+termination proofs, and so produce code that will be closer to what a
+programmer may have written directly in OCaml. However, reasoning about program
+equivalences requires dealing with such termination proofs, and it is common
+practice to use custom reduction lemmas that will be used extensively when
+proving properties about nonstructural recursive functions.
+
+\emph{Structured recursion schemes} and their associated algebraic laws provide
+an easier framework for reasoning about program equivalences.  In particular,
+\ocaml{qsort} is an instance of a \emph{divide-and-conquer}
 algorithm which, in the functional programming literature is known as a
 hylomorphism~\cite{MeijerFP91, HuIT96}.  Hinze et al.~\cite{HinzeWG15} showed
-that every recursive program can be turned into an hylomorphism, which implies
-that, ``\emph{every recursion scheme can be formalised as an hylomorphism}''.
-Furthermore, the laws that are satisfied by hylomorphisms ease reasoning about
-program equivalences, since they can capture a number of useful equivalences,
+that every recursive program (recursion scheme) can be formalised as a hylomorphism.
+Furthermore, hylomorphism laws can capture a number of useful equivalences,
 ranging from common optimisations such as \emph{short-cut
-fusion}~\cite{TakanoM95}, to (semi-)automatic
+fusion}~\cite{TakanoM95}, to semi-automatic
 parallelisations~\cite{Gibbons96:Third, farmsCastro}.
 
-Despite the benefits of hylomorphisms for reasoning about program equivalences,
-not many approaches mechanise structured recursion schemes in proof assistants.
-A recent approach by Abreu et al.~\cite{AbreuDHJMS23} encode an algebraic
-approach to divide-and-conquer computations, in which termination is entirely
-enforced by the typing discipline of the recursion scheme. However, although
-their approach solves the problem of termination proofs, as well as the
-performance of the code that is run \emph{within Coq}, it no longer allows the
-extraction of clean, naturally-looking code.  By naturally-looking and clean,
-we mean code that (1) preserves the recursive structure intended by the
-programmer; and (2) does not lead to unsafe casts \ocaml{Obj.magic} in the
-generated code.
-
+Not many approaches mechanise structured recursion
+schemes in proof assistants.  Recently, Abreu et al.~\cite{AbreuDHJMS23} encode
+an algebraic approach to divide-and-conquer computations, in which termination
+is entirely enforced by the typing discipline of the recursion scheme. However,
+although their approach solves the problem of termination proofs, as well as
+the performance of the code that is run \emph{within Coq}, it no longer allows
+the extraction of clean, naturally-looking code.  By this, we refer to code
+that (1) does not preserve the recursive structure of common implementations;
+and (2) lead to unsafe casts \ocaml{Obj.magic} in the generated code. 
 In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
@@
 \begin{coqcode}
-let rec qsort = function
-  | [] -> []
-  | h :: t ->
-    let (l, r) = partition (fun x0 -> (<=) x0 h) t in
+let rec qsort = function | [] -> []
+  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
@@
 \end{coqcode}
-
 The contributions we make in this paper are as follows:
@@
   \item We provide a framework for \emph{generic programming} with recursion
-    schemes in Coq, with a strong focus on proving their algebraic laws and 
-    program equivalences, and clean code extraction.
-  \item We formalise algorithms for divide and conquer, dynamic programming and
-    mutual recursion.
-  \item We verified the well-known short-cut fusion optimisation in Coq and
-    extract the resulting code to OCaml.
-\end{itemize}
-
-\dcas{Rewritten up to here}
-In
-\ocaml{quicksort} above, a pen-and-paper technique to prove termination would
-be to show that \ocaml{divide} is a \emph{well-founded coalgebra} on the
-functor $(1 + A \times - \times -)$ for a fixed type $A$. Here the term
-``well-founded'' refers to the fact that the infinite iteration of this
-coalgebra will produce a \emph{finite} binary tree with nodes in $A$.
-
-More generally, this is an instance of a \emph{divide-and-conquer} algorithm.
-Such algorithms are structured in three parts: first, the input is broken down
-in smaller sub-problems by means of a well-founded coalgebra. Secondly, these
-sub-problems are computed recursively and, finally, the results of the recursive
-call are put together by a an algebra. This particular way of doing recursion
-has been named in the functional programming literature as a
-hylomorphism~\cite{MeijerFP91, HuIT96}.
-
-Many years later Hinze et al.~\cite{HinzeWG15} showed that every recursive
-program can be turned into an hylomorphism by means of the conjugate rule. The
-essence of this result is that every complex recursion scheme can be explained
-by its associated \emph{adjunction} which establishes a way of reducing the
-proof obligation for the complex recursion scheme into the proof obligation for
-a basic hylomorphism. This can be paraphrased by saying
-\begin{center}
-  ``\emph{every recursion scheme can be formalised as an hylomorphism}''
-\end{center}
-
-This result has given a much needed unifying view of recursion scheme which
-helped understanding their nature and also in providing a unified way of
-reasoning about recursive programs by means of fusion laws.
-
-\dcas{My take would be different here from now on. See below:}
-
-Recursion schemes enjoy generic algebraic laws for program manipulation. Many
-common optimisations and transformations are instances of such laws. Examples of
-this include shortcut deforestation~\cite{TakanoM95}, and
-parallelisations~\cite{Gibbons96:Third}. \dcas{TODO: add more examples?}
-
-With this work, we study the formalisation of hylomorphism laws in proof
-assistants like Coq to derive efficient and correct recursive algorithms by
-program calculation techniques. Furthermore, our encoding of hylomorphisms
-allows Coq's code extraction mechanism to produce OCaml code that resembles what
-a programmer would have writen manually.
-
-\dcas{END (TODO: merge with Marco's. I would omit (or take parts of the) below)}
-
-There is a tension:
-\begin{itemize}
-  \item On one hand we can allow general recursion and the employ recursion
-schemes to help the programmer structuring programs so that they terminate
-  \item on the other hand we could forbid non-termination, but at this point
-there is no algorithm to tell termination and non-termination apart.
-\end{itemize}
-
-The type theoretical approach instead is to reject all definitions which cannot
-be proven well-defined. It is safer, but incomplete.
-
-With this work we intend to begin a study of the applications of recursion
-schemes that goes beyond functional programming, that is to dependent type
-theories such as a the ones employed in proof assistants like Agda or Coq where
-non-termination is outright rejected.
-
-This has been a long-standing challenge to come up with a sound algorithm which
-rejects all programs that do not terminate (soundness) and accepts as many
-well-defined programs as possible (completeness). Our aim is to provide an
-additional tool to the every-day programmer to be able to extend the range of
-recursive programs we can write.
-\mvol{What is the ``every-day programmer'' here? Are they working in Coq/Agda?}
-
-\dcas{Note: the below is useful to keep in the intro}
-One of the challenges here is to be able to keep the theory of recursions
-schemes as general as possible while maintaining its practical usefulness. In
-order to do this, we have to make some design choices which differ a bit from
-the general theory.  For instance, formalising the conjugate rule would require
-us to mechanise a non-trivial fragment of category theory inside the prover's
-logic.  Secondly, the general conjugate rule is a statement of correspondence of
-arrows rather than a proof that these programs actually exists. In functional
-programming languages like for instance Haskell this is not a problem, since
-every functor is strong and has a fixed-point.  On the other hand, in a type
-theory we have to restrict to those so-called polynomial functors. In this work
-we do not make use of these assumptions. By restricting ourselves to polynomial
-functors given in terms of containers~\cite{AbbottAG05} we are able to prove the
-recursion schemes within the proof assistant and at the same time to maintain a
-certain level of data abstraction.
-
-First we prove that the hylo equation has a unique solution
-\[
-  x = a \comp F(x) \comp c
-\]
-for every recursive coalgebra $c$ for a polynomial $F$. Then we prove each
-recursion scheme by reducing it to a hylo. The following recursion schemes can
-all be proven by reducing them to an hylo: folds, mutual recursive functions,
-recursion schemes from comonads and their respective coinductive dual.
-\mpav{Moved this from Section 3}
-
-\mvol{Need some transiiton here to build up to listing the goals and contributions of the paper.
-Also, ``axiom-free'' and ``obj.magic'' come out of nowhere here, so both need to be established earlier in the introduction.}
-The primary goals of this work are:
-\begin{enumerate}
-  \item that the
-  mechanisation of hylomorphisms and their laws should be axiom-free; and
-  \item that Coq
-hylomorphisms should be extracted to recursive OCaml functions that use no
-instances of \ocaml{Obj.magic}.
-\end{enumerate}
-In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
-Although the use of \ocaml{Obj.magic} should be safe in the code
-extracted by Coq, it has been proven that, for higher-order programs, simple
-interoperations can lead to incorrect behaviour or even
-segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
-\ocaml{Obj.magic} means that the extracted code will have stronger
-safety guarantees, and will allow the safe interoperation with other OCaml code.
-\mvol{This is good!}
-
-The contributions we make in this paper are as follows:
-\begin{itemize}
-  \item We provide a framework for \emph{generic programming} with recursion
-schemes in Coq
-  \item We formalise algorithms for divide and conquer, dynamic programming and
-mutual recursion
-  \item We validated optimisations in Coq and extract the resulting code to
-OCaml.
+    schemes in Coq, with a proof of their algebraic laws and 
+    program equivalences, that allows clean code extraction.
+  \item We formalise divide and conquer, dynamic programming, and
+    mutual recursion algorithms.
+  \item We verified the short-cut fusion optimisation in Coq and
+    extract optimised code to OCaml.
 \end{itemize}

### 2024-03-14 paper/paper.tex
minor
Containers

@@
 \section{Mechanising Extractable Container Functors}
-\dcas{Subsections are tentative topics to remember what I need to fill in here,
-but these are not final}
-
 In order to abstract recursion patters we need to be able to abstract away from
 the particular shape of the data.  This is achieved by introducing the concept
-of \emph{functors} which have the suitable fixed-point properties, i.e. those
+of \emph{functors} which have the suitable fixed-point properties, i.e.\ those
 who have a initial algebras. A common approach to construct such functors is to
-use \emph{container}~\cite{AbbottAG05} which is what we introduce in this
-Section~\ref{sec:containers}.  However, reasoning about containers equality will
-require us to consider both functional extensionality and heterogeneous
-equality. To avoid this in Section~\ref{sec:types-morphisms} we introduce a
-custom equivalence relation on types.
+use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:containers}).
+However, reasoning about container equality
+will require us to consider both functional extensionality and heterogeneous
+equality. We avoid these axioms by introducing a
+custom equivalence relation on types (Section~\ref{sec:types-morphisms}).
 
@@
 \label{sec:containers}
-To abstract away the shape of a particular data structure we use functors.
-These are functions \coq{F : Type -> Type} which additionally have a
-map \coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
-functor represents a container for abstract data that can be manipulated without
-changing the outer structure. For example, the type \coq{List : Type -> Type} is
-a functor and \coq{List A} is the type of lists over elements of type \coq{A}
-with the obvious \coq{fmap} function recursive over a list and applying the
-function \coq{A -> B} for each element of the list.
-Additionally the type of lists arises as the least fixed-point for the
-functor \coq{F X = unit + (A * X) }.
-
-In general not every functor has a fixed-point, so we are going to restrict
-ourselves to \emph{polynomial functors} as represented by containers.  A
-container is defined by a type of \emph{shapes} and a \emph{family of position
-types} indexed by shapes.
+Functors are functions \coq{F : Type -> Type} which additionally have a map
+\coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
+functor represents a container for abstract data that can be manipulated
+without changing the outer structure. For example, the type 
+\coq{List : Type -> Type} is a functor and \coq{List A} is the type of lists
+over elements of type \coq{A} with the obvious \coq{fmap} function recursive
+over a list and applying the function \coq{A -> B} for each element of the
+list. Additionally, the type of lists \coq{List A} arises as the least
+fixed-point of the functor \coq{F X = unit + (A * X) }.
+
+In general not every functor has a fixed-point, and it is not possible to build
+the fixed-point of a functor such as \coq{F} in Coq due to not being strictly
+positive.  Due to this, we are going to restrict to \emph{polynomial functors}
+as represented by containers.  A container is defined by a type of
+\emph{shapes} and a \emph{family of position types} indexed by shapes.
 \begin{coqcode}
@@
 \end{coqcode}
-For the functor \coq{F} we introduced in the previous paragraph we would
-define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
-the data type and one of these contains a piece of data of type \coq{A}.
-Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
-and \coq{Pos (inr a) = unit} indicating that the first constructor does not have
-any recursive type variables and the second constructor has one type variable.
+For the functor \coq{F} we introduced in the previous paragraph we would define
+\coq{Shape} as \coq{unit + A} indicating there are two constructors in the data
+type and one of these contains a piece of data of type \coq{A}.  Moreover, we
+would define the positions as \coq{Pos (inl tt) = Empty_set} and 
+\coq{Pos (inr a) = unit} indicating that the first constructor does not have
+any type variables and the second constructor has one type variable.
 
@@
 \begin{coqcode}
-Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
+Record App (X : Type) := MkApp { shape : Shape; contents : Pos shape -> X }.
 \end{coqcode}
 with the obvious action on morphisms given by post-composition
-with \coq{contents}.  In our running example, the type \coq{App X}, for
+with \coq{contents}.
+\begin{coqcode}
+Definition fmap (f : A -> B) (x : App C A) : App C B
+  := {| shape := shape x; contents := fun e => f (contents x e) |}
+\end{coqcode}
+In our running example, the type \coq{App X}, for
 some \coq{X} has two inhabitants.  The first is a pair composed
 by \coq{inl tt : Shape} and a function \coq{Pos (inl tt) -> X}. This latter is
-in fact a function of type we \coq{Empty_set -> X}  by definitions which in turn
-is isomorphic to the type \coq{unit}, the type with only one inhabitant.  The
+in fact a function of type \coq{Empty_set -> X}. This pair is, therefore
+isomorphic to the type \coq{unit}, the type with only one inhabitant.  The
 second inhabitant is the pair composed by \coq{inr a : Shape}, for
@@
 is isomorphic to \coq{X}.  This particular pair of shapes and positions is 
-isomorphic to the type \coq{unit + A * X}.
+therefore isomorphic to the type \coq{unit + A * X}.
 
 The correspondence between containers and polynomial functors has been
-formalised in this work and has been moved to the
-Appendix~\ref{sec:pol-containers} for space reasons. Generally, any functor that
-is isomorphic to a container extension is called a \emph{container functor}.
-Here we will call \emph{container functor} the functors defined by a specific
-container extension.
+formalised in this work and can be found in the accompanying code (file
+Container.v). 
+% 
+%Generally, any functor that
+%is isomorphic to a container extension is called a \emph{container functor}.
+%Here we will call \emph{container functor} the functors defined by a specific
+%container extension.
 
@@
 In the previous section we defined containers using dependent types.  However,
-Ocaml's type system is not equipped to handle these and to solve this problem
+Ocaml's type system is not equipped to handle these. To solve this problem
 Coq's code extraction mechanism will insert unsafe casts.
 
-Here consider instead representing the positions of a container using a validity
-predicate \coq{Valid : Shape -> Pos -> bool} which assigns shapes to positions.
-Moreover, it is crucial for our purposes of remaining axiom-free that this
-predicate is \emph{decidable} in order to be able to reason about equality of
-container extensions and, as a result, the contents of the container extensions
-need to be accessed via a function \coq{contents} whose domain depends on the
-shape of the container.  This means that for a given type of shapes and
-positions, the equality of two objects of a container extension will need to
-deal with heterogeneous equality.
-\mpav{This part needs more explanation.}
-
-\begin{coqcode}
-Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
-Coercion is_true (b : bool) : Prop := b = true.
-Record PosOf `(C : Cont Shape Pos) (shape : Shape) :=
-  ValidPos { val : Pos; IsValid : valid shape pos }.
-\end{coqcode}
-where \coq{valid} is the \emph{decidable} validity predicate. Note that booleans
-can be lifted to propositions similarly to the implementation provided by
-SSReflect~\cite{GonthierL09}. This allows us to define the dependent pair of
-valid positions as a record which will be extracted to a singleton type since
-the validity proof is erased by Coq's code extraction mechanism.
-
-We now define \emph{container extensions} as follows:
-\begin{coqcode}
-Record App `(C : Cont Shape Pos) (X : Type) :=
-  MkApp { shape : Shape; contents : PosOf C shape -> X }.
-\end{coqcode}
-The functorial action for container extensions on morphisms boils down to
-function composition, which trivially preserves identity and composition.
-\begin{coqcode}
-Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
-  := {| shape := shape x; contents := fun e => f (contents x e) |}
-\end{coqcode}
+Consider instead representing the positions of a container using a decidable
+validity predicate (\coq{valid}) which assigns shapes to positions. We use
+boolean functions and coercions from \coq{bool} to \coq{Prop} to represent
+decidable predicates, similarly to SSReflect~\cite{GonthierL09}. The family of
+positions for a given shape in a container can be recovered by defining:
+\begin{coqcode}
+Class Cont (Shape : Type) (APos : Type) := { valid : Shape -> APos -> bool }.
+Record Pos `(C : Cont Shape APos) (shape : Shape) :=
+  ValidPos { val : APos; IsValid : valid shape pos }.
+\end{coqcode}
+Coq's code extraction will now be able to erase the validity predicate, and
+generate OCaml code that is free of unsafe casts. The OCaml code extracted for
+\coq{Pos} will now be exactly the code extracted for \coq{APos}.
+The decidability of the validity predicate is crucial for our purposes of 
+remaining axiom-free. To illustrate this, suppose that we need to show
+the equality of two container extensions. We will need to show that, for
+the same positions, they will produce the same result. In Coq, the goal
+would look as follows: 
+\begin{coqcode}
+ k : Pos C s -> X
+ P1, P2 : valid s p
+ ---------------------------
+ k (ValidPos p P1) = k (ValidPos p P2)
+\end{coqcode}
+If \coq{valid} was a regular proposition in \coq{Prop}, it would not possible
+to prove the equality of \coq{P1} and \coq{P2}. However, by using a decidable
+predicate, if we know that \coq{P1} and \coq{P2} are of type 
+\coq{valid s p = true}, then we can prove without any axioms that 
+\coq{P1 = P2 = eq_refl}.
 
@@
 
-Finally we can define the type of containers as follows:
-
 \subsubsection{Equality of Container Extensions}
-The equality of container extensions needs to be extensional, since the
-contents are obtained by applying a function. We want to equate container
-extensions that have the same shape, and that, for equal shape and position,
-they return the same element. To avoid the use of functional extensionality
-axioms, we capture this relation with the following inductive proposition in
-Coq:
+Reasoning about the equality of container extensions is not entirely solved by
+using decidable validity predicates to define the families of positions. In
+general, we want to equate container extensions that have the same shape, and
+that, for equal shape and position, they return the same element. To avoid the
+use of the functional extensionality axiom, we capture this relation with the
+following inductive proposition in Coq:
 \begin{coqcode}
@@
 Structure morph A {eA : setoid A} B {eB : setoid B} :=
-  MkMorph
-    { app :> A -> B;
-      app_eq : forall x y, x =e y -> app x =e app y
-    }.
+  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
 Notation "A ~> B" := (@morph A _ B _).
@@
 equivalence relations. Since Coq's code extraction mechanism will erase any
-occurrence of \coq{Prop} in the code, so objects of type
-\coq{A ~> B} will be extracted to the OCaml equivalent to
-\coq{A -> B}. Furthermore, we add a coercion from \coq{A ~> B} to \coq{A -> B}.
+occurrence of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be
+extracted to the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
+from \coq{A ~> B} to \coq{A -> B}.
 On top of this, we define basic function composition and identity functions:
@@
 \end{coqcode}
-
 Using custom equivalences and proper morphisms, we redefine the definitions of
@@
 however, we chose to use Coq's propositional equality, since these would leads
-to simpler code. We explain why in Section~\ref{sec:cata}, and the
-mechanisation of initial algebras of container extensions. The definitions in
-Figure~\ref{fig:poly} are redefined to account for these changes in a
-straightforward way, i.e. we prove that validity predicates are indeed proper
-morphisms.  \dcas{Should we show, instead, the actual full definition? It
-becomes a bit more complicated ...}
+to simpler code. We explain why in Section~\ref{sec:algebras}, and the
+mechanisation of initial algebras of container extensions.
+%The definitions in
+%Figure~\ref{fig:poly} are redefined to account for these changes in a
+%straightforward way, i.e. we prove that validity predicates are indeed proper
+%morphisms.
 
@@
 \begin{coqcode}
-Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D), 
-  h \o g =e id -> h \o (g \o f) =e f.
-Proof.
-  intros H.  rewrite compA H.  reflexivity.
-Qed.
+Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D) (H : h \o g =e id), 
+  h \o (g \o f) =e f.
+Proof.  rewrite compA, H.  reflexivity. Qed.
 \end{coqcode}
@@
 \subsubsection{Polynomial Types}
-We define a number of equivalences for polynomial types. These equivalences are
-be used when building proper morphisms that manipulate their respective types.
+We define a number of equivalences for polynomial types.
+%These equivalences are
+%be used when building proper morphisms that manipulate their respective types.
 \begin{coqcode}
@@
 \end{coqcode}
-
 Most of the definitions that involve functions and polynomial types are
@@
 The proofs that these morphisms are proper are straightforward.  Finally, we
-also provide functions for currying/uncurrying, and partially applying a proper
-morphism. We avoid using definitions, or unfold them, to help Coq's code
-extraction mechanism to inline as many of these small definitions as possible.
+also provide functions for currying/uncurrying, and  flipping the arguments of
+a proper morphism. We force most of our definitions to be inlined, to help
+Coq's code extraction mechanism to inline as many of these combinators as
+possible.
 
@@
                     | inr p => fun H => cont (snd x) (ValidPos p H)
-                    end (IsValid e)
-|}.
+                    end (IsValid e) |}.
 \end{coqcode}
@@
 \section{Extraction}
-This section covers how to use our approach to calculate recursive programs
-from simple, non-recursive specifications. The steps are the following:
+We now proceed to calculate recursive programs from simple, non-recursive specifications.
+The steps are the following:
 \begin{enumerate}

### 2024-03-15 paper/paper.tex
edits to section 2

@@
 \label{sec:containers}
-Functors are functions \coq{F : Type -> Type} which additionally have a map
-\coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
-functor represents a container for abstract data that can be manipulated
-without changing the outer structure. For example, the type 
-\coq{List : Type -> Type} is a functor and \coq{List A} is the type of lists
-over elements of type \coq{A} with the obvious \coq{fmap} function recursive
-over a list and applying the function \coq{A -> B} for each element of the
-list. Additionally, the type of lists \coq{List A} arises as the least
-fixed-point of the functor \coq{F X = unit + (A * X) }.
+Functors are functions \coq{F : Type -> Type} which additionally have a
+map \coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
+functor represents a container for abstract data that can be manipulated without
+changing the outer structure. For example, the type \coq{List : Type -> Type} is
+a functor and \coq{List A} is the type of lists over elements of type \coq{A}
+with the obvious \coq{fmap} function recursively traversing a list and applying
+the function \coq{A -> B} for each element. Additionally, the type of
+lists \coq{List A} arises as the least fixed-point of the
+functor \coq{F X = unit + (A * X) }.
 
@@
 \label{sec:types-morphisms}
-We address the different forms of equality by defining a class of setoids,
-i.e.\ types with na associated equivalence relation, and considering only
-functions that respect the associated equivalences, i.e. \emph{proper}
+We address the different forms of equality by defining a class of
+\emph{setoids}, types with an associated equivalence relation, and considering
+only functions that respect the associated equivalences, or \emph{proper}
 morphisms with respect to the function \emph{respectfulness} relation. We use
@@
 Coq's code extraction mechanism remove any occurence of custom equivalence
-relations in the extracted OCaml code.  We use \coq{=e} to denote
-the equivalence relation of a setoid. By default, we associate every Coq type
-with the standard propositional equality, unless a different equivalence is
-specified (i.e.\ we allow overlapping instances, and Coq's propositional
-equality takes the lowest priority). 
+relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
+relation of a setoid. By default, we associate every Coq type with the standard
+propositional equality, unless a different equivalence is specified (we allow
+overlapping instances, and Coq's propositional equality takes the lowest
+priority).
 \begin{coqcode}
@@
 \end{coqcode}
-Note that we leverage on Coq's type class mechanism to fill in the necessary
-equivalence relations. Since Coq's code extraction mechanism will erase any
-occurrence of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be
-extracted to the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
+We rely on Coq's type class mechanism to fill in the necessary equivalence
+relations. Coq's code extraction mechanism will erase any occurrence
+of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
+the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
 from \coq{A ~> B} to \coq{A -> B}.
@@
 equality, or a combination of propositional and extensional equality. However,
-our compositional approach allows to build morphisms by plugging in other
-morphisms to our combinator, In our framework, our goal is that the
+our compositional approach allows us to build morphisms by plugging in other
+morphisms to our combinator. In our framework, our expectation is that the
 user-provided functions remain small, with relatively straightforward proofs of
@@
 Coq's \emph{Generalised Rewriting}. Since every morphism \coq{f : A ~> B}
-satisfies that if \coq{x =e y}, then \coq{f x =e f y}, we can add every
-morphism as a proper element of Coq's respectfulness relation. In practice,
-this means that we can use the \coq{rewrite} tactic on proofs of type 
-\coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
+satisfies the property that if \coq{x =e y}, then \coq{f x =e f y}, we can add
+every morphism as a proper element of Coq's respectfulness relation. In
+practice, this means that we can use the \coq{rewrite} tactic on proofs of
+type \coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
 arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},

### 2024-03-15 paper/paper.tex
Updates

@@
 priority).
-\begin{coqcode}
-Reserved Notation "f =e g" (at level 70, no associativity).
-Class setoid A : Type :=
-  MkEquiv
-    { eqRel : A -> A -> Prop;
-      e_refl : forall x, eqRel x x;
-      e_sym : forall x y, eqRel x y -> eqRel y x;
-      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
-    }.
-Notation "f =e g" := (eqRel f g).
-\end{coqcode}
+%\begin{coqcode}
+%Reserved Notation "f =e g" (at level 70, no associativity).
+%Class setoid A : Type :=
+%  MkEquiv
+%    { eqRel : A -> A -> Prop;
+%      e_refl : forall x, eqRel x x;
+%      e_sym : forall x y, eqRel x y -> eqRel y x;
+%      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
+%    }.
+%Notation "f =e g" := (eqRel f g).
+%\end{coqcode}
 Given types \coq{A} and  \coq{B}, with their respective
@@
 \coq{eB : B -> B -> Prop}, the we define the type
-\coq{A ~> B} to represents \emph{proper} morphisms of the
+\coq{A ~> B} to represent \emph{proper} morphisms of the
 respectfulness relation of \coq{R} to \coq{R'}.
@@
 \begin{coqcode}
-Class Cont (Shape : Type) (Pos : Type) := { valid : Shape * Pos ~> bool }.
-Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
+Class Cont `{setoid Shape} (Pos : Type) := { valid : Shape * Pos ~> bool }.
+Inductive AppR `{F : Cont Sf Pf} `{setoid X} (x y : App F X) : Prop :=
   | AppR_ext (Es : shape x =e shape y)
@@
 our compositional approach allows us to build morphisms by plugging in other
-morphisms to our combinator. In our framework, our expectation is that the
+morphisms to our combinators. In our framework, our expectation is that the
 user-provided functions remain small, with relatively straightforward proofs of
@@
 
-We prove the isomorphisms of polynomial types and the equivalent (redefined)
+We prove the isomorphisms of polynomial types and the equivalent 
 container extensions. As an example, we discuss (informally) the isomorphisms
@@
 \coq{App F X * App G X}, and then construct the necessary container
-extension. Given \coq{p_inl}/\coq{p_inr} that act as
-\coq{inl/inr} on valid product positions, the relevant
-functions that define the isomorphism are:
+extension. Given \coq{p_inl : Pos l -> Pos (l * r)} (resp. \coq{p_inr}) that
+act as \coq{inl} (resp.\ \coq{inr}) on product positions, and
+\coq{case_pos : (Pos l -> X) -> (Pos r -> X) -> Pos (l * r) -> X} that pattern
+matches on the product positions, the functions that witness the isomorphism
+are:
 \begin{coqcode}
 Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
-  ({| shape := shape (fst x); cont := fun e => cont x (e_inl e) |}, 
-   {| shape := shape (snd x); cont := fun e => cont x (e_inr e) |}).
+  ({| shape := shape (fst x); cont := fun e => cont x (p_inl e) |}, 
+   {| shape := shape (snd x); cont := fun e => cont x (p_inr e) |}).
 Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
 {| shape := (shape (fst x), shape (snd x));
-   cont := fun e => match val e with 
-                    | inl p => fun H => cont (fst x) (ValidPos p H)
-                    | inr p => fun H => cont (snd x) (ValidPos p H)
-                    end (IsValid e) |}.
-\end{coqcode}
-Proving that these functions indeed form an isomorphism is straightforward,
-using the fact that \coq{Prod} containers only have two valid positions.
+   cont := case_pos (cont (fst x)) (cont (snd x)) |}.
+\end{coqcode}
+Proving that the composition of these functions is the identity is
+straightforward using the fact that \coq{Prod} containers only have two valid
+positions.
 
@@
 formalise algebras for container extensions (Section~\ref{sec:algebras}), then
-we formalise recursive coalgebras (Section~\ref{sec:rec-coalgebras}) and then we
-put together these notions to formalise hylomorphisms (Section~\ref{sec:hylo}).
+we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
+these notions to formalise recursive coalgebras and hylomorphisms
+(Section~\ref{sec:rec-coalgebras}).
 
 \subsection{Algebras and Catamorphisms}
-\label{sec:algebras} An algebra is a set \coq{X} together with some operations
-on it\footnote[1]{Sometimes and algebra comes equipped with an equational
-theory, e.g. the theory of monoids, but here we focus on algerbraically free
-algebras}.  For example, give such a type \coq{X} we can define the monoid
-operations as \coq{u : unit -> X} for unit and \coq{m : X * X -> X} for
-multiplication. Notice that \coq{FX = unit + X * X} is a functor on \coq{X}
-which means we can equivalently describe these two maps as a single one of
-type \coq{F X -> X} which we call an algebra for the functor \coq{F}
-satisfying the unit and associative laws of the monoid.  In this work we will be
-using algebras to describe the operations of a data type and so we will not be
-needing additional equations on these operations.
+\label{sec:algebras} An algebra is a set \coq{X} (the \emph{carrier} of the
+algebra) together with some operations on it\footnote[1]{Sometimes and algebra
+comes equipped with an equational theory, e.g. the theory of monoids, but here
+we focus on algebraically free algebras}.  For example, given such a type
+\coq{X} we can define the monoid operations as \coq{u : unit -> X} for unit and
+\coq{m : X * X -> X} for multiplication. Notice that \coq{F X = unit + X * X}
+is a functor on \coq{X} which means we can equivalently describe these two maps
+as a single one of type \coq{F X -> X} which we call an algebra for the functor
+\coq{F} satisfying the unit and associative laws of the monoid.  In this work
+we will be using algebras to describe the operations of a data type and so we
+will not be needing additional equations on these operations.
 
@@
 formalise algebras in terms of containers and containers extensions. Given a
-type `\coq{A}` and a container \coq{F} an algebra for \coq{F} over the carrier
-`\coq{A}` is a map of the following type:
+type \coq{A} and a container \coq{F} an algebra for \coq{F} over the carrier
+\coq{A} is a map of the following type:
 \begin{coqcode}
@@
 
+\dcas{The writing below is confusing, but I am not sure how to fix it ... Can't
+we start with the list-generating functor? The monoid example is great, but as 
+soon as we move to talking about least-fixed points, it feels as if we'd better
+go straight into F X = unit + A * X}
 Now, the \emph{easiest} and most effective way to turn any type $A$ into a
@@
 % which does precisely what we described in the previous paragraph.
-
+%
 % where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
@@
 % $\ana{c}$ respectively.
-
+%
 %%\dcas{Boring stuff}
@@
 %\end{coqcode}
-We define smart constructors for the isomorphism of least fixed points
-\coq{LFix_out}.
-\begin{coqcode}
-l_in : App F (LFix F) ~> LFix F
-l_out : LFix F ~> App F (LFix F)
+We define \coq{LFix} as a setoid, where its equivalence relation can be
+described as the least fixed point of \coq{AppR}.  We define smart constructors
+for the isomorphism of least fixed points as respectful morphisms.
+\begin{coqcode}
+l_in : App F (LFix F) ~> LFix F                l_out : LFix F ~> App F (LFix F)
 \end{coqcode}
@@
 \begin{coqcode}
-Definition cata_f (alg : Alg F A) : LFix -> A
-:= fix f (x : LFix) :=
-  match x with
-  | LFix_in ax =>
-      let (sx, kx) := ax in
-      alg (MkApp sx (fun e => f (kx e)))
-  end.
-\end{coqcode}
-It is easy to show that this function is a morphism of \coq{F}-algebras. In fact,
-it is possible to define it as a map of the following type:
-\begin{coqcode}
-Alg F A ~> LFix ~> A
-\end{coqcode}
-
+  Definition cata_f `{F : Cont Sh P} (alg : Alg F A) : LFix -> A
+:= fix f (x : LFix) := match x with
+  | LFix_in ax => let (sx, kx) := ax in alg (MkApp sx (fun e => f (kx e))) end.
+\end{coqcode}
+It is easy to show that this function is a respectful morphism of
+\coq{F}-algebras. In fact, it is possible to define it as a map of the
+following type:
+\begin{coqcode}
+  cata : forall `{F : Cont Sh P} `{setoid A}, Alg F A ~> LFix ~> A
+\end{coqcode}
 We prove that catamorphisms satisfy the universal law, that is that there is a
@@
 \subsection{Coalgebras and Anamorphisms}
+\label{sec:coalg}
 While folds \emph{destruct} data, their dual, the unfolds, \emph{construct}
@@
 corecursively  ``runs'' the observations starting from the initial seed.
-
 % Consider the transition system below composed by three states \coq{x1}, \coq{x2}
@@
 % \end{tikzcd}\]
-
 The seed function can be modelled by a coalgebra. For example, for a type of
@@
 Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
-carrier \coq{X} is a map \coq{X -> FX}. In the previous example \coq{c} is
-an \coq{F}-coalgebra on the carrier \coq{X} with \coq{FX = L * X}.
+carrier \coq{X} is a map \coq{X -> F X}. In the previous example \coq{c} is
+an \coq{F}-coalgebra on the carrier \coq{X} with \coq{F X = L * X}.
 In our development we use the following notation for coalgebras:
 \begin{coqcode}
-Notation Coalg F A := (A ~> App F A) (only parsing).
-\end{coqcode}
-
+Notation Coalg F A := (A ~> App F A).
+\end{coqcode}
 Now an anamorphism takes such a seed function and runs it while collecting all
@@
 where \coq{GFix_in}  and \coq{GFix_out} are the two inverses witnessing the
-isomorphism. Additionally, we define smart constructors for the isomorphism of greatest fixed points
-\begin{coqcode}
-g_in : App F GFix ~> GFix
-g_out : GFix ~> App F GFix
-\end{coqcode}
-
+isomorphism. Similarly to \coq{LFix}, \coq{GFix} is defined as a setoid, with
+an equivalence relation that is the greatest fixpoint of \coq{AppR}.
+Additionally, we define smart constructors for the isomorphism of greatest
+fixed points
+\begin{coqcode}
+g_in : App F (GFix F) ~> GFix F                g_out : GFix F ~> App F (GFix F)
+\end{coqcode}
+\dcas{The explanation below is weird ... We have not defined 'ana', but state
+their universal property?}
 What witnesses the fact that this is the greatest fixed-point is the universal
@@
 map that runs the seed function and composes the information from the first step
-the result
-
 \begin{coqcode}
@@
 \end{coqcode}
-Equivalently, we could have define an anamorphism as the
-unique \coq{F}-coalgebra homomorphism, that is a map of the following type
-\begin{coqcode}
-Coalg F A ~> A ~> GFix
+In our Coq formalisation, we define anamorphisms as the
+unique \coq{F}-coalgebra homomorphisms, that are maps of the following type:
+\begin{coqcode}
+ana : forall `{F : Cont Sh P} `{setoid A}, Coalg F A ~> A ~> GFix F
 \end{coqcode}
@@
 \label{sec:rec-coalgebras}
-Hylomorphism capture the concept of \emph{divide-and-conquer} algorithms.  First
-the input is destructured (\emph{divide}) in smaller parts. This operation may
-or may not be structural. Next the results from the subparts are computed
-recursively and finally the results are composed together (\emph{conquer}) by
-means of an algebra.
+Hylomorphisms capture the concept of \emph{divide-and-conquer} algorithms.
+First the input is destructured (\emph{divide}) in smaller parts. This
+operation may or may not be structural. Next the results from the subparts are
+computed recursively and finally the results are composed together
+(\emph{conquer}) by means of an algebra.
 
@@
 
+\dcas{Marco, shouldn't the below be "given an F-algebra and a recursive
+F-coalgebra"? We contradict ourselves below where we say that the hylomorphism
+equation does not necessarily have unique solutions, or solutions at all. As
+far as I understand, these are also called hylomorphisms. E.g. one can define a
+general fixpoint combinator as a hylomorphism, which does not have a unique
+solution}
 Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
@@
 \end{equation}
-
-However this solution does not necessarily exists for an arbitrary pair of
+However this solution does not necessarily exist for an arbitrary pair of
 algebras and coalgebras. In fact, this definition cannot be accepted by Coq,
@@
 hylomorphism equation. These are hylomoprhisms where the coalgebra is
-\emph{recursive}.
-
-Recursive coalgebras are coalgebras whose anamorphism produces finite data which
-what the following predicate captures:
+\emph{recursive}, i.e. coalgebras that terminate on all inputs. The following
+predicate captures that a coalgebra terminates on an input: 
 \begin{coqcode}
@@
 \end{coqcode}
-Recursive anamorphisms are implemented in Coq recursively on the structure of
+A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra 
+\coq{c : Coalg F A} such that \coq{forall x, RecF c x}.
+Recursive hylomorphisms are implemented in Coq recursively on the structure of
 the proof for the type (\coq{RecF}) as follows:
@@
   : forall (x : A), RecF c x -> B := fix f x H
-  := match c x as h0 return
-       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
-     with
-     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
+  := match c x as h0 return (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
+     with | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
      end (RecF_inv H).
 \end{coqcode}
-We again use \coq{RecF_inv} to obtain the structurally smaller proof to use in
-the recursive calls. As we did with catamorphisms and recursive anamorphisms, we
-prove that \coq{hylo_def} is respectful, and use this proof to built the
+We use \coq{RecF_inv} to obtain the structurally smaller proof to use in
+the recursive calls. As we did with catamorphisms and anamorphisms, we
+prove that \coq{hylo_def} is respectful, and use this proof to build the
 corresponding higher-order proper morphism:
 \begin{coqcode}
+  hylo : forall `{F : Cont Sh P} `{setoid A} `{setoid B}, 
     Alg F B ~> RCoalg F A ~> A ~> B
@@
 The following lemmas prove the isomorphism:
-\mpav{David, check this code works}
 \begin{coqcode}

### 2024-03-16 paper/paper.tex
minor

--- 2024-03-14 paper/paper.tex
@@
 \section{Mechanising Extractable Container Functors}
+\dcas{Subsections are tentative topics to remember what I need to fill in here,
+but these are not final}
+
 In order to abstract recursion patters we need to be able to abstract away from
 the particular shape of the data.  This is achieved by introducing the concept
-of \emph{functors} which have the suitable fixed-point properties, i.e.\ those
+of \emph{functors} which have the suitable fixed-point properties, i.e. those
 who have a initial algebras. A common approach to construct such functors is to
-use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:containers}).
-However, reasoning about container equality
-will require us to consider both functional extensionality and heterogeneous
-equality. We avoid these axioms by introducing a
-custom equivalence relation on types (Section~\ref{sec:types-morphisms}).
+use \emph{container}~\cite{AbbottAG05} which is what we introduce in this
+Section~\ref{sec:containers}.  However, reasoning about containers equality will
+require us to consider both functional extensionality and heterogeneous
+equality. To avoid this in Section~\ref{sec:types-morphisms} we introduce a
+custom equivalence relation on types.
 
@@
 \label{sec:containers}
-Functors are functions \coq{F : Type -> Type} which additionally have a map
-\coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
-functor represents a container for abstract data that can be manipulated
-without changing the outer structure. For example, the type 
-\coq{List : Type -> Type} is a functor and \coq{List A} is the type of lists
-over elements of type \coq{A} with the obvious \coq{fmap} function recursive
-over a list and applying the function \coq{A -> B} for each element of the
-list. Additionally, the type of lists \coq{List A} arises as the least
-fixed-point of the functor \coq{F X = unit + (A * X) }.
-
-In general not every functor has a fixed-point, and it is not possible to build
-the fixed-point of a functor such as \coq{F} in Coq due to not being strictly
-positive.  Due to this, we are going to restrict to \emph{polynomial functors}
-as represented by containers.  A container is defined by a type of
-\emph{shapes} and a \emph{family of position types} indexed by shapes.
+To abstract away the shape of a particular data structure we use functors.
+These are functions \coq{F : Type -> Type} which additionally have a
+map \coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
+functor represents a container for abstract data that can be manipulated without
+changing the outer structure. For example, the type \coq{List : Type -> Type} is
+a functor and \coq{List A} is the type of lists over elements of type \coq{A}
+with the obvious \coq{fmap} function recursive over a list and applying the
+function \coq{A -> B} for each element of the list.
+Additionally the type of lists arises as the least fixed-point for the
+functor \coq{F X = unit + (A * X) }.
+
+In general not every functor has a fixed-point, so we are going to restrict
+ourselves to \emph{polynomial functors} as represented by containers.  A
+container is defined by a type of \emph{shapes} and a \emph{family of position
+types} indexed by shapes.
 \begin{coqcode}
@@
 \end{coqcode}
-For the functor \coq{F} we introduced in the previous paragraph we would define
-\coq{Shape} as \coq{unit + A} indicating there are two constructors in the data
-type and one of these contains a piece of data of type \coq{A}.  Moreover, we
-would define the positions as \coq{Pos (inl tt) = Empty_set} and 
-\coq{Pos (inr a) = unit} indicating that the first constructor does not have
-any type variables and the second constructor has one type variable.
+For the functor \coq{F} we introduced in the previous paragraph we would
+define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
+the data type and one of these contains a piece of data of type \coq{A}.
+Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
+and \coq{Pos (inr a) = unit} indicating that the first constructor does not have
+any recursive type variables and the second constructor has one type variable.
 
@@
 \begin{coqcode}
-Record App (X : Type) := MkApp { shape : Shape; contents : Pos shape -> X }.
+Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
 \end{coqcode}
 with the obvious action on morphisms given by post-composition
-with \coq{contents}.
-\begin{coqcode}
-Definition fmap (f : A -> B) (x : App C A) : App C B
-  := {| shape := shape x; contents := fun e => f (contents x e) |}
-\end{coqcode}
-In our running example, the type \coq{App X}, for
+with \coq{contents}.  In our running example, the type \coq{App X}, for
 some \coq{X} has two inhabitants.  The first is a pair composed
 by \coq{inl tt : Shape} and a function \coq{Pos (inl tt) -> X}. This latter is
-in fact a function of type \coq{Empty_set -> X}. This pair is, therefore
-isomorphic to the type \coq{unit}, the type with only one inhabitant.  The
+in fact a function of type we \coq{Empty_set -> X}  by definitions which in turn
+is isomorphic to the type \coq{unit}, the type with only one inhabitant.  The
 second inhabitant is the pair composed by \coq{inr a : Shape}, for
@@
 is isomorphic to \coq{X}.  This particular pair of shapes and positions is
-therefore isomorphic to the type \coq{unit + A * X}.
+isomorphic to the type \coq{unit + A * X}.
 
 The correspondence between containers and polynomial functors has been
-formalised in this work and can be found in the accompanying code (file
-Container.v). 
-% 
-%Generally, any functor that
-%is isomorphic to a container extension is called a \emph{container functor}.
-%Here we will call \emph{container functor} the functors defined by a specific
-%container extension.
+formalised in this work and has been moved to the
+Appendix~\ref{sec:pol-containers} for space reasons. Generally, any functor that
+is isomorphic to a container extension is called a \emph{container functor}.
+Here we will call \emph{container functor} the functors defined by a specific
+container extension.
 
@@
 In the previous section we defined containers using dependent types.  However,
-Ocaml's type system is not equipped to handle these. To solve this problem
+Ocaml's type system is not equipped to handle these and to solve this problem
 Coq's code extraction mechanism will insert unsafe casts.
 
-Consider instead representing the positions of a container using a decidable
-validity predicate (\coq{valid}) which assigns shapes to positions. We use
-boolean functions and coercions from \coq{bool} to \coq{Prop} to represent
-decidable predicates, similarly to SSReflect~\cite{GonthierL09}. The family of
-positions for a given shape in a container can be recovered by defining:
-\begin{coqcode}
-Class Cont (Shape : Type) (APos : Type) := { valid : Shape -> APos -> bool }.
-Record Pos `(C : Cont Shape APos) (shape : Shape) :=
-  ValidPos { val : APos; IsValid : valid shape pos }.
-\end{coqcode}
-Coq's code extraction will now be able to erase the validity predicate, and
-generate OCaml code that is free of unsafe casts. The OCaml code extracted for
-\coq{Pos} will now be exactly the code extracted for \coq{APos}.
-The decidability of the validity predicate is crucial for our purposes of 
-remaining axiom-free. To illustrate this, suppose that we need to show
-the equality of two container extensions. We will need to show that, for
-the same positions, they will produce the same result. In Coq, the goal
-would look as follows: 
-\begin{coqcode}
- k : Pos C s -> X
- P1, P2 : valid s p
- ---------------------------
- k (ValidPos p P1) = k (ValidPos p P2)
-\end{coqcode}
-If \coq{valid} was a regular proposition in \coq{Prop}, it would not possible
-to prove the equality of \coq{P1} and \coq{P2}. However, by using a decidable
-predicate, if we know that \coq{P1} and \coq{P2} are of type 
-\coq{valid s p = true}, then we can prove without any axioms that 
-\coq{P1 = P2 = eq_refl}.
+Here consider instead representing the positions of a container using a validity
+predicate \coq{Valid : Shape -> Pos -> bool} which assigns shapes to positions.
+Moreover, it is crucial for our purposes of remaining axiom-free that this
+predicate is \emph{decidable} in order to be able to reason about equality of
+container extensions and, as a result, the contents of the container extensions
+need to be accessed via a function \coq{contents} whose domain depends on the
+shape of the container.  This means that for a given type of shapes and
+positions, the equality of two objects of a container extension will need to
+deal with heterogeneous equality.
+\mpav{This part needs more explanation.}
+
+\begin{coqcode}
+Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
+Coercion is_true (b : bool) : Prop := b = true.
+Record PosOf `(C : Cont Shape Pos) (shape : Shape) :=
+  ValidPos { val : Pos; IsValid : valid shape pos }.
+\end{coqcode}
+where \coq{valid} is the \emph{decidable} validity predicate. Note that booleans
+can be lifted to propositions similarly to the implementation provided by
+SSReflect~\cite{GonthierL09}. This allows us to define the dependent pair of
+valid positions as a record which will be extracted to a singleton type since
+the validity proof is erased by Coq's code extraction mechanism.
+
+We now define \emph{container extensions} as follows:
+\begin{coqcode}
+Record App `(C : Cont Shape Pos) (X : Type) :=
+  MkApp { shape : Shape; contents : PosOf C shape -> X }.
+\end{coqcode}
+The functorial action for container extensions on morphisms boils down to
+function composition, which trivially preserves identity and composition.
+\begin{coqcode}
+Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
+  := {| shape := shape x; contents := fun e => f (contents x e) |}
+\end{coqcode}
 
@@
 
+Finally we can define the type of containers as follows:
+
 \subsubsection{Equality of Container Extensions}
-Reasoning about the equality of container extensions is not entirely solved by
-using decidable validity predicates to define the families of positions. In
-general, we want to equate container extensions that have the same shape, and
-that, for equal shape and position, they return the same element. To avoid the
-use of the functional extensionality axiom, we capture this relation with the
-following inductive proposition in Coq:
+The equality of container extensions needs to be extensional, since the
+contents are obtained by applying a function. We want to equate container
+extensions that have the same shape, and that, for equal shape and position,
+they return the same element. To avoid the use of functional extensionality
+axioms, we capture this relation with the following inductive proposition in
+Coq:
 \begin{coqcode}
@@
 Structure morph A {eA : setoid A} B {eB : setoid B} :=
-  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
+  MkMorph
+    { app :> A -> B;
+      app_eq : forall x y, x =e y -> app x =e app y
+    }.
 Notation "A ~> B" := (@morph A _ B _).
@@
 equivalence relations. Since Coq's code extraction mechanism will erase any
-occurrence of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be
-extracted to the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
-from \coq{A ~> B} to \coq{A -> B}.
+occurrence of \coq{Prop} in the code, so objects of type
+\coq{A ~> B} will be extracted to the OCaml equivalent to
+\coq{A -> B}. Furthermore, we add a coercion from \coq{A ~> B} to \coq{A -> B}.
 On top of this, we define basic function composition and identity functions:
@@
 \end{coqcode}
+
 Using custom equivalences and proper morphisms, we redefine the definitions of
@@
 however, we chose to use Coq's propositional equality, since these would leads
-to simpler code. We explain why in Section~\ref{sec:algebras}, and the
-mechanisation of initial algebras of container extensions.
-%The definitions in
-%Figure~\ref{fig:poly} are redefined to account for these changes in a
-%straightforward way, i.e. we prove that validity predicates are indeed proper
-%morphisms.
+to simpler code. We explain why in Section~\ref{sec:cata}, and the
+mechanisation of initial algebras of container extensions. The definitions in
+Figure~\ref{fig:poly} are redefined to account for these changes in a
+straightforward way, i.e. we prove that validity predicates are indeed proper
+morphisms.  \dcas{Should we show, instead, the actual full definition? It
+becomes a bit more complicated ...}
 
@@
 \begin{coqcode}
-Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D) (H : h \o g =e id), 
-  h \o (g \o f) =e f.
-Proof.  rewrite compA, H.  reflexivity. Qed.
+Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D), 
+  h \o g =e id -> h \o (g \o f) =e f.
+Proof.
+  intros H.  rewrite compA H.  reflexivity.
+Qed.
 \end{coqcode}
@@
 \subsubsection{Polynomial Types}
-We define a number of equivalences for polynomial types.
-%These equivalences are
-%be used when building proper morphisms that manipulate their respective types.
+We define a number of equivalences for polynomial types. These equivalences are
+be used when building proper morphisms that manipulate their respective types.
 \begin{coqcode}
@@
 \end{coqcode}
+
 Most of the definitions that involve functions and polynomial types are
@@
 The proofs that these morphisms are proper are straightforward.  Finally, we
-also provide functions for currying/uncurrying, and  flipping the arguments of
-a proper morphism. We force most of our definitions to be inlined, to help
-Coq's code extraction mechanism to inline as many of these combinators as
-possible.
+also provide functions for currying/uncurrying, and partially applying a proper
+morphism. We avoid using definitions, or unfold them, to help Coq's code
+extraction mechanism to inline as many of these small definitions as possible.
 
@@
                     | inr p => fun H => cont (snd x) (ValidPos p H)
-                    end (IsValid e) |}.
+                    end (IsValid e)
+|}.
 \end{coqcode}
@@
 
-\subsubsection{Example 1}
 Shortcut deforestation can be expressed succintly in terms of hylomorphisms and
@@
 
-\subsubsection{Example 2} We then prove that the length function fuses with
+We then prove that the length function fuses with
 the naive quadratic \emph{reverse} function:

### 2024-03-18 paper/paper.tex
RW

@@
 
-\subsection{Recursive Anamorphisms}
-Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
-inductive data type in the following way.  Intuitively, a recursive coalgebra
-can be applied only finitely many times, therefore when this is applied to an
-anamorphism the only possible traces it can produce from the seed function are
-the finite ones. We denote this special kind of recursion scheme \emph{recursive
-anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
-can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
-exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
-algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
-from the universality property of the hylomorphism and the fact that recursive
-anamorphisms satisfy the same equation.
-
-We now show how this is proven in the Coq development. First we define a
-recursive anamorphism as a function into \coq{LFix F}:
-\begin{coqcode}
-Definition rana_f__ `{setoid A} (c : Coalg F A)
-  : forall x : A, RecF c x -> LFix F
-  := fix f (x : A) (H : RecF c x) :=
-       let hx := c x in
-       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
-
-Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F := ...
-\end{coqcode}
-We omit the boilerplate details to obtain the setoid morphism.
-The recursive anamorphism has the following universality property:
-\begin{coqcode}
-Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
-    : f =e rana h <-> f =e l_in \o fmap f \o h.
-\end{coqcode}
-We can now show that every hylomorphism is composed by a recursive anamorphism
-followed by a catamorphism:
-\begin{coqcode}
-Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
-  (g : Alg F B) (h : RCoalg F A)
-: cata g \o rana h =e hylo g h.
-\end{coqcode}
-This result will allow us to \emph{calculate} programs by providing small, easy
-to verify implementations, and combining them into efficient recursive functions
-by using Coq's proofs of hylomorphism laws as optimisation passes.
+% \subsection{Recursive Anamorphisms}
+% Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
+% inductive data type in the following way.  Intuitively, a recursive coalgebra
+% can be applied only finitely many times, therefore when this is applied to an
+% anamorphism the only possible traces it can produce from the seed function are
+% the finite ones. We denote this special kind of recursion scheme \emph{recursive
+% anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
+% can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
+% exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
+% algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
+% from the universality property of the hylomorphism and the fact that recursive
+% anamorphisms satisfy the same equation.
+% 
+% We now show how this is proven in the Coq development. First we define a
+% recursive anamorphism as a function into \coq{LFix F}:
+% \begin{coqcode}
+% Definition rana_f__ `{setoid A} (c : Coalg F A)
+%   : forall x : A, RecF c x -> LFix F
+%   := fix f (x : A) (H : RecF c x) :=
+%        let hx := c x in
+%        LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
+% 
+% Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F := ...
+% \end{coqcode}
+% We omit the boilerplate details to obtain the setoid morphism.
+% The recursive anamorphism has the following universality property:
+% \begin{coqcode}
+% Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
+%     : f =e rana h <-> f =e l_in \o fmap f \o h.
+% \end{coqcode}
+% We can now show that every hylomorphism is composed by a recursive anamorphism
+% followed by a catamorphism:
+% \begin{coqcode}
+% Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
+%   (g : Alg F B) (h : RCoalg F A)
+% : cata g \o rana h =e hylo g h.
+% \end{coqcode}
+% This result will allow us to \emph{calculate} programs by providing small, easy
+% to verify implementations, and combining them into efficient recursive functions
+% by using Coq's proofs of hylomorphism laws as optimisation passes.
 
@@
 Within the domain of program optimization, program calculation serves as a
-well-established programming technique aimed at deriving efficient programs from
-their naive counterparts through systematic program transformation. This area
-has been extensively explored over the years. Tesson et al.\ demonstrated the
-efficacy of leveraging Coq to establish an approach for implementing a robust
-system dedicated to verifying the correctness of program transformations for
-functions that manipulate lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and
-Emoto went further and\ldots~\cite{MurataE19}
+well-established programming technique aimed at deriving efficient programs
+from their naive counterparts through systematic program transformation. This
+area has been extensively explored over the years. Tesson et al.\ demonstrated
+the efficacy of leveraging Coq to establish an approach for implementing a
+robust system dedicated to verifying the correctness of program transformations
+for functions that manipulate lists~\cite{10.1007/978-3-642-17796-5_10}. Murata
+and Emoto went further and\ldots~\cite{MurataE19}
 
@@
 
-Abreu et al.\ propose\ldots~\cite{AbreuDHJMS23}. Their approach has similarities to what is presented here.
-
-They differ in requiring the assumption of the functional extensionality axiom.
-Additionally, they are concerned with the erasure of termination proofs but have
-little discussion of what the resulting extracted code \emph{looks like} (that
-is, whether the extracted OCaml code resembles the natural formulation of the
-recursive function).
-\mvol{We should find out for sure if they use \texttt{obj.magic}.}
+Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
+using a recursion scheme in which termination is entirely enforced by its
+typing.  This is a significant advance, since it avoids \emph{completely} the
+need for termination proofs. Their work differs from ours in that they require
+the functional extensionality axiom, and the use of impredicative \coq{Set}.
+The authors justify well the use of impredicative \coq{Set} and its
+compatibility with the functional extensionality axiom. In contrast, our
+development remains entirely \emph{axiom-free}. Another key difference with our
+approach is that they do not discuss what the resulting extracted code
+\emph{looks like} (that is, whether the extracted OCaml code resembles the
+natural formulation of the recursive function). Through experiments, we found
+that their formalisation leads to \ocaml{Obj.magic}, and code with a complex
+structure that may be hard to understand or interface with other OCaml code.
+Due to the great benefit of entirely avoiding termination proofs, it would be
+interesting to extend their approach to improve code extraction.
 

### 2024-03-18 paper/paper.tex
minor

--- 2024-03-16 paper/paper.tex
@@
-\documentclass[anonymous, a4paper, UKenglish, cleveref, autoref, thm-restate]{lipics-v2021}
+\documentclass[ a4paper, UKenglish, cleveref, autoref, thm-restate]{lipics-v2021}
 \usepackage[utf8]{inputenc} % Required for inputting international characters
@@
 the proof assistant. Examples of this include common sorting algorithms, such
-as Quicksort and Mergesort.
+as quicksort and mergesort.
 %One example is Quicksort, written in OCaml
@@
 This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
-well as their algebraic laws that: (1) is \textbf{fully axiom-free}; and (2)
+well as their algebraic laws that: (1) is \emph{axiom-free}; and (2)
 allows the extraction of naturally-looking, clean OCaml code. While programmers
@@
 
-\dcas{Marco, shouldn't the below be "given an F-algebra and a recursive
-F-coalgebra"? We contradict ourselves below where we say that the hylomorphism
-equation does not necessarily have unique solutions, or solutions at all. As
-far as I understand, these are also called hylomorphisms. E.g. one can define a
-general fixpoint combinator as a hylomorphism, which does not have a unique
-solution}
 Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
-unique solution to the equation
+unique solution (when it exists) to the equation
 \begin{equation}
@@
 no corecursive definition that is guaranteed to be syntactically guarded -- as
-required by Coq -- due to the use of an arbitrary algebra $a$. Furthermore,
-solutions to the hylomorphism equation are \emph{not} guaranteed to be unique,
-or to even exist. To address these issues, we mechanise \emph{recursive
+required by Coq -- due to the use of an arbitrary algebra $a$.
+To address these issues, we mechanise \emph{recursive
 hylomorphisms} which are guaranteed to have a unique solution to the
@@
 % where \coq{h} is a recursive coalgebra.
+
 \section{Extraction}
-We now proceed to calculate recursive programs from simple, non-recursive specifications.
 The steps are the following:
@@
 
-\subsection{Shortcut Deforestation}
-
-Shortcut deforestation can be expressed succintly in terms of hylomorphisms and
-their laws~\cite{TakanoM95}. In particular, given a function:
-\begin{coqcode}
-  s  : forall A. (App F A -> A) -> (App F A -> A)
-\end{coqcode}
-We can conclude, by parametricity, that
-\begin{coqcode}
-  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
-\end{coqcode}
-This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
-\textbf{not} provable in Coq if we want to remain axiom-free, since we would
-need to add the necessary parametricity
-axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
-
-However, we prove a specific version of this theorem for the list generating
-container (i.e.\ the container whose least fixed point is a list), and use it
-to encode the example from Takano and Meijer~\cite{TakanoM95}:
-\begin{coqcode}
-Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
-\end{coqcode}
-Here, we are defining function \coq{sf1} as the composition of \coq{length},
-\coq{Lmap f} and \coq{append ys}. Functions \coq{length} and \coq{Lmap} are
-catamorphisms. Function \coq{append ys} is also a catamorphism that
-appends \coq{ys} to an input list. It is defined by applying an algebra to
-every cons node of a list, and applying a catamorphism with the input algebra
-to \coq{ys} in the nil case:
-\begin{coqcode}
-Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
-  fun x => match x with | MkCont sx kx => match sx with
-  | s_nil => fun _ => (hylo a ilist_coalg) l
-  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
-  end kx end.
-Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
-\end{coqcode}
-Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
-\coq{ListF} container. We apply the hylo fusion law repeatedly, unfold
-definitions, and simplify in our specification for \coq{sf1}: 
-\begin{coqcode}
-Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
-  rewrite hylo_map_fusion, <- acid_rain.
-  simpl; reflexivity.
-Defined.
-\end{coqcode}
-From this, we extract the following OCaml code:
-\begin{ocamlcode}
-let rec sf1 f ys = function | [] -> let rec f0 = function
-                                      | [] -> (Uint63.of_int (0))
-                                      | _ :: t -> add (Uint63.of_int (1)) (f0 t)
-                                    in f0 ys
-                            | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
-\end{ocamlcode}
-
-We then prove that the length function fuses with
-the naive quadratic \emph{reverse} function:
-\begin{coqcode}
-Definition sf2 : Ext (length \o reverse). 
-  calculate.  unfold length, reverse. rewrite hylo_fusion_l. 
-  2:{ (* Rewrite into the fused version *) }
-  simpl; reflexivity.
-Defined.
-\end{coqcode}
-This code extracts to the optimised length function on the input list:
-\begin{ocamlcode}
-let rec ex2 = function | [] -> (Uint63.of_int (0))
-                       | _ :: t -> add (Uint63.of_int (1)) (ex2 t)
-\end{ocamlcode}
-
 \subsection{Sorting Algorithms}
-
-So far, our examples could be modelled exclusively using catamorphisms. Here,
-we show examples that require recursive hylomorphisms and termination proofs.
-We complete the sorting algorithm examples by applying fusion optimisation.
-
-\subsubsection{Binary Trees}
-
-Both Mergesort and Quicksort are hylomorphisms whose recursive structure is
-described by a binary tree. In particular, the container used in their
-definition is isomorphic to \coq{TreeF (list nat) unit} and 
-\coq{TreeF unit nat} respectively, where \coq{TreeF} is defined as follows:
+So far, our examples could be modelled exclusively using catamorphisms. Here, we
+show examples that require recursive hylomorphisms and termination proofs.  We
+complete the sorting algorithm examples by applying fusion optimisation.
+
+Both mergesort and quicksort are divide-and-conquer algorithms that can be
+captured by the structure of an hylomorphisms. The structure of the recursion
+that of a binary tree.  For example, in the case of quicksort, a list is split
+into a pivot, the label of the node, and two sublists. We define the data
+functor of trees as follows:
 \begin{coqcode}
@@
 \end{coqcode}
-The definition as a container uses the following shapes and positions:
+We define the functor as a container using the following shapes and positions:
 \begin{coqcode}
@@
 destructors:
+\mpav{The following algebra should be called a\_in or l\_in whatever}
 \begin{coqcode}
 Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.
+
 (* Converting TreeF to ITreeF *)
@@
 
-\subsubsection{Mergesort}
-
-Mergesort is a hylomorphism with the recursive structure defined by the
-container \coq{(TreeF (list int) unit)}. 
-The computationally expensive part of the Mergesort implementation is the
-algebra, since this is where most of the comparisons happen. The recursive
-coalgebra for Mergesort is defined as the following Coq functions:
-\begin{coqcode}
-Fixpoint splitL (x : list int) (accL accR : list int) :=
-  match x with
-  | nil => (accL, accR)
-  | cons x xs => splitL xs accR (cons x accL)
-  end.
-
-Definition c_split : Coalg (TreeF (list int) unit) (list int) :=
-|{ x ~> match x with
-        | nil | cons _ nil => a_leaf x
-        | _ => let (l, r) := splitL x nil nil in
-               a_node tt l r
-        end
- }|.
-Defined.
-\end{coqcode}
-Here, the notation \coq{|{ x ~> E }|} defines the morphism \coq{fun x => E},
-and uses a custom tactic to prove that it is respectful. The recursive
-coalgebra \coq{c_split} takes a list, and it returns a leaf with the input
-list, if its size is less or equal than 1. Otherwise, it applies the function
-\coq{splitL}, and divides the list in two halves.
-
-It is easy to prove that this coalgebra is recursive, by showing that
-it respects the relation \coq{<} on the length of the lists:
-\begin{coqcode}
-Lemma split_fin : respects_relation c_split (@length int) lt.
-  Proof. (* *) Qed.
-
-Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
-  := mk_wf_coalg wf_lt split_fin.
-\end{coqcode}
-
-The algebra simply applies the \coq{mergeL} function in order to the sublists
-in every node:
-\begin{coqcode}
-Fixpoint mergeL l1 l2 {struct l1}  : list int := (* *)
-
-Definition merge : App (TreeF (list int) unit) (list int) ~> list int.
-|{ x ~> match a_out x with | node _ l r => mergeL l r | leaf e => e end }|.
-Defined.
-\end{coqcode}
-
-Putting everything together, Mergesort is defined as the following
-hylomorphism:
-\begin{coqcode}
-Definition msort := hylo merge tsplit.
-\end{coqcode}
-Note, however, that we could have composed the catamorphism and anamorphism,
-and used the deforestation result to optimise the intermediate binary tree
-away:
-\begin{coqcode}
-Definition msort : Ext (cata merge \o rana tsplit).
-calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
-\end{coqcode}
-In both cases, the result is the same, and it extracts to the usual divide and
-conquer definition of Mergesort in OCaml:
-\begin{ocamlcode}
-let rec msort x = match x with
-| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
-     let (l0, r) = splitL x [] [] in
-     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
-     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
-\end{ocamlcode}
-
-\subsubsection{Quicksort}
+% \subsubsection{Quicksort}
 
@@
 \end{coqcode}
-As with the Mergesort case, we prove that the coalgebra is recursive by showing
+We prove that the coalgebra \coq{c_split} is recursive by showing
 that it respects the ``less-than'' relation on the length of the lists.
@@
 
+The mergesort algorithm can be defined analogously and can be found in the formalisation. [REF LINK].
+
+% \subsubsection{Mergesort}
+
+% Mergesort is a hylomorphism with the recursive structure defined by the
+% container \coq{(TreeF (list int) unit)}.
+% The computationally expensive part of the Mergesort implementation is the
+% algebra, since this is where most of the comparisons happen. The recursive
+% coalgebra for Mergesort is defined as the following Coq functions:
+% \begin{coqcode}
+% Fixpoint splitL (x : list int) (accL accR : list int) :=
+%   match x with
+%   | nil => (accL, accR)
+%   | cons x xs => splitL xs accR (cons x accL)
+%   end.
+
+% Definition c_split : Coalg (TreeF (list int) unit) (list int) :=
+% |{ x ~> match x with
+%         | nil | cons _ nil => a_leaf x
+%         | _ => let (l, r) := splitL x nil nil in
+%                a_node tt l r
+%         end
+%  }|.
+% Defined.
+% \end{coqcode}
+% Here, the notation \coq{|{ x ~> E }|} defines the morphism \coq{fun x => E},
+% and uses a custom tactic to prove that it is respectful. The recursive
+% coalgebra \coq{c_split} takes a list, and it returns a leaf with the input
+% list, if its size is less or equal than 1. Otherwise, it applies the function
+% \coq{splitL}, and divides the list in two halves.
+
+% It is easy to prove that this coalgebra is recursive, by showing that
+% it respects the relation \coq{<} on the length of the lists:
+% \begin{coqcode}
+% Lemma split_fin : respects_relation c_split (@length int) lt.
+%   Proof. (* *) Qed.
+
+% Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
+%   := mk_wf_coalg wf_lt split_fin.
+% \end{coqcode}
+
+% The algebra simply applies the \coq{mergeL} function in order to the sublists
+% in every node:
+% \begin{coqcode}
+% Fixpoint mergeL l1 l2 {struct l1}  : list int := (* *)
+
+% Definition merge : App (TreeF (list int) unit) (list int) ~> list int.
+% |{ x ~> match a_out x with | node _ l r => mergeL l r | leaf e => e end }|.
+% Defined.
+% \end{coqcode}
+
+% Putting everything together, Mergesort is defined as the following
+% hylomorphism:
+% \begin{coqcode}
+% Definition msort := hylo merge tsplit.
+% \end{coqcode}
+% Note, however, that we could have composed the catamorphism and anamorphism,
+% and used the deforestation result to optimise the intermediate binary tree
+% away:
+% \begin{coqcode}
+% Definition msort : Ext (cata merge \o rana tsplit).
+% calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
+% \end{coqcode}
+% In both cases, the result is the same, and it extracts to the usual divide and
+% conquer definition of Mergesort in OCaml:
+% \begin{ocamlcode}
+% let rec msort x = match x with
+% | [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
+%      let (l0, r) = splitL x [] [] in
+%      let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
+%      mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
+% \end{ocamlcode}
 \subsubsection{Fusing a divide-and-conquer computation}
@@
 \ocaml{memo_knap}, which is finally discarded to produce the final result.
+
+\subsection{Shortcut Deforestation}
+
+Shortcut deforestation can be expressed succintly in terms of hylomorphisms and
+their laws~\cite{TakanoM95}. In particular, given a function:
+\begin{coqcode}
+  s  : forall A. (App F A -> A) -> (App F A -> A)
+\end{coqcode}
+We can conclude, by parametricity, that
+\begin{coqcode}
+  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
+\end{coqcode}
+This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
+\textbf{not} provable in Coq if we want to remain axiom-free, since we would
+need to add the necessary parametricity
+axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
+
+However, we prove a specific version of this theorem for the list generating
+container (i.e.\ the container whose least fixed point is a list), and use it
+to encode the example from Takano and Meijer~\cite{TakanoM95}:
+\begin{coqcode}
+Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
+\end{coqcode}
+Here, we are defining function \coq{sf1} as the composition of \coq{length},
+\coq{Lmap f} and \coq{append ys}. Functions \coq{length} and \coq{Lmap} are
+catamorphisms. Function \coq{append ys} is also a catamorphism that
+appends \coq{ys} to an input list. It is defined by applying an algebra to
+every cons node of a list, and applying a catamorphism with the input algebra
+to \coq{ys} in the nil case:
+\begin{coqcode}
+Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
+  fun x => match x with | MkCont sx kx => match sx with
+  | s_nil => fun _ => (hylo a ilist_coalg) l
+  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
+  end kx end.
+Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
+\end{coqcode}
+Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
+\coq{ListF} container. We apply the hylo fusion law repeatedly, unfold
+definitions, and simplify in our specification for \coq{sf1}:
+\begin{coqcode}
+Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
+  rewrite hylo_map_fusion, <- acid_rain.
+  simpl; reflexivity.
+Defined.
+\end{coqcode}
+From this, we extract the following OCaml code:
+\begin{ocamlcode}
+let rec sf1 f ys = function | [] -> let rec f0 = function
+                                      | [] -> (Uint63.of_int (0))
+                                      | _ :: t -> add (Uint63.of_int (1)) (f0 t)
+                                    in f0 ys
+                            | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
+\end{ocamlcode}
+
+We then prove that the length function fuses with
+the naive quadratic \emph{reverse} function:
+\begin{coqcode}
+Definition sf2 : Ext (length \o reverse).
+  calculate.  unfold length, reverse. rewrite hylo_fusion_l.
+  2:{ (* Rewrite into the fused version *) }
+  simpl; reflexivity.
+Defined.
+\end{coqcode}
+This code extracts to the optimised length function on the input list:
+\begin{ocamlcode}
+let rec ex2 = function | [] -> (Uint63.of_int (0))
+                       | _ :: t -> add (Uint63.of_int (1)) (ex2 t)
+\end{ocamlcode}
 

### 2024-03-18 paper/paper.tex
stuff

@@
 \section{Related Work}
-
-\begin{itemize}
-  \item McBride devised a generalisation of Bove and Capretta's general
-recursive function monad~\cite{McBride15, BoveC01}
-  \item Guarded recursion for productive definitions and general recursion
-~\cite{AtkeyM13, PaviottiMB15}
-  \item PACO library~\cite{HurNDV13}
-  \item Sized Types \cite{HughesPS96, abs-1202-3496}
-  \item Copatterns \cite{AbelPTS13, AbelP16}
-  \item IMPORTANT!!!!! Coq extraction from recursion
-    schemes~\cite{larchey2022braga} \dcas{Check bibtex}
-  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
-  \item Recursion schemes in Coq (but not hylos)~\cite{MurataE19}
-  \item EVEN MORE IMPORTANT!!! 2023 POPL paper on divide and conquer code in
-    Coq~\cite{AbreuDHJMS23}.  Some differences to point out: they require
-    functional-extensionality axiom. We are axiom-free. Little discussion about
-    extraction. They focus mostly on erasing termination proofs, but they do
-    not discuss "what does their extracted code look like". Their
-    representation of Functors leads me to think that they will produce loads
-        of "Obj.magic", but we'd need to check this.
-\mvol{We have a story for this now: obj.magic harms interaction with other ocaml code, the resulting ocaml code is not natural to read, and in general more complex/strange extracted code is likely to be less efficient and harder to optimise.}
-        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
-\end{itemize}
+%
+%\begin{itemize}
+%  \item McBride devised a generalisation of Bove and Capretta's general
+%recursive function monad~\cite{McBride15, BoveC01}
+%  \item Guarded recursion for productive definitions and general recursion
+%~\cite{AtkeyM13, PaviottiMB15}
+%  \item PACO library~\cite{HurNDV13}
+%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
+%  \item Copatterns \cite{AbelPTS13, AbelP16}
+%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
+%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
+%\end{itemize}
 
@@
 well-established programming technique aimed at deriving efficient programs
-from their naive counterparts through systematic program transformation. This
-area has been extensively explored over the years. Tesson et al.\ demonstrated
-the efficacy of leveraging Coq to establish an approach for implementing a
-robust system dedicated to verifying the correctness of program transformations
-for functions that manipulate lists~\cite{10.1007/978-3-642-17796-5_10}. Murata
-and Emoto went further and\ldots~\cite{MurataE19}
-
-\mvol{Need more general citations for program calculation.}
+from their naive counterparts through systematic program
+transformation~\cite{gibbonsSquiggol}. This area has been extensively explored
+over the years. Tesson et al.\ demonstrated the efficacy of leveraging Coq to
+establish an approach for implementing a robust system dedicated to verifying
+the correctness of program transformations for functions that manipulate
+lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
+formalised recursion schemes in Coq~\cite{MurataE19}. Their development does
+not include hylomorphisms and dynamorphisms, and relies on the functional
+extensionality axiom, as well as further extensionality axioms for each
+coinductive datatype that they use. They do not discuss the extracted OCaml
+code from their formalisation.  Larchey-Wendling and Monin encode recursion
+schemes in Coq, by formalising computational graphs of
+algorithms~\cite{larchey2022braga}. Their work does not focus on encoding
+generic recursion schemes, and proving their algebraic laws.  Castro et
+al.~\cite{farmsCastro} encode the laws of hylomorphisms as part of the type
+system of a functional language to calculate parallel programs from
+specifications. Their work focuses on parallelism, and they do not formalise
+their approach in a proof assistant, and the laws of hylomorphisms are axioms
+in their system.
+
+%\mvol{Need more general citations for program calculation.}
 

### 2024-03-18 paper/paper.tex
Title
Clean -> idiomatic

@@
 
-\title{A Mechanised Library for Generic (Co)Programming}
+\title{Magic-free Code Extraction of Generic Programs and Recursion Schemes in
+Coq}
 
@@
   laws. The key contribution of this paper is that this formalisation is fully
-  axiom-free, and it allows for the extraction of safe, clean, idiomatic OCaml
+  axiom-free, and it allows for the extraction of safe, idiomatic OCaml
   code for a variety of recursive programs. We formalise a series of divide-and
@@
 the performance of the code that is run \emph{within Coq}, it no longer allows
-the extraction of clean, naturally-looking code.  By this, we refer to code
+the extraction of idiomatic OCaml code.  By this, we refer to code
 that (1) does not preserve the recursive structure of common implementations;
@@
 well as their algebraic laws that: (1) is \emph{axiom-free}; and (2)
-allows the extraction of naturally-looking, clean OCaml code. While programmers
+allows the extraction of idiomatic OCaml code. While programmers
 still need to reason about the termination of their programs, our mechanisation
 will allow the use of the algebraic laws of hylomorphisms for program
-reasoning, as well as extracting to clean, naturally-looking code. As an
+reasoning, as well as extracting to idiomatic code. As an
 example our mechanisation of \ocaml{qsort} as a hylomorphism will produce

### 2024-03-18 paper/paper.tex
intro

@@
 safety guarantees, and will allow the safe interoperation with other OCaml
-code.
+code. The absence of \ocaml{Obj.magic} provides even stronger guarantees to the
+extracted code. In particular, since the code that we extract from Coq is
+implemented in a total language, we know that there can be no $\bot$ in its
+definition. This allows the programmer to use ``fast-and-loose'' reasoning on
+the extracted code~\cite{DanielssonHJG06}, which would otherwise be unsafe if
+the code contained \ocaml{Obj.magic}. For example, the programmer would need to
+manually check that the inputs satisfy the necessary assumptions which would be
+impossible to enforce in the presence of unsafe casts. Overall, the absence of
+\ocaml{Obj.magic} makes it easer to reason about, and interoperate with
+extracted code.
 
@@
 algorithms~\cite{larchey2022braga}. Their work does not focus on encoding
-generic recursion schemes, and proving their algebraic laws.  Castro et
+generic recursion schemes, and proving their algebraic laws.  Castro-Perez et
 al.~\cite{farmsCastro} encode the laws of hylomorphisms as part of the type

### 2024-03-18 paper/paper.tex
RelW

@@
 %\end{itemize}
-
-\subsection{Program Calculation}
+Encoding recursion schemes in Coq is not new. We compare our work with other
+encodings of program calculation techniques in Coq (Section~\ref{sec:calccoq}),
+recursion schemes in Coq (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}).
+Finally, we compare our encoding of recursion schemes to other forms of
+termination checking (Section~\ref{sec:termcheck}, and sized types
+(Section~\ref{sec:sizedT}, as a way to guarantee termination of nonstructural
+recursion.
+
+\subsection{Program Calculation}\label{sec:calccoq}
 
@@
 
-\subsection{Termination Checking}
+\subsubsection{Divide and Conquer Recursion}\label{sec:calccoq}
+
+Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
+using a recursion scheme in which termination is entirely enforced by its
+typing.  This is a significant advance, since it avoids \emph{completely} the
+need for termination proofs. Their work differs from ours in that they require
+the functional extensionality axiom, and the use of impredicative \coq{Set}.
+The authors justify well the use of impredicative \coq{Set} and its
+compatibility with the functional extensionality axiom. In contrast, our
+development remains entirely \emph{axiom-free}. Another key difference with our
+approach is that they do not discuss what the resulting extracted code
+\emph{looks like} (that is, whether the extracted OCaml code resembles the
+natural formulation of the recursive function). Through experiments, we found
+that their formalisation leads to \ocaml{Obj.magic}, and code with a complex
+structure that may be hard to understand or interface with other OCaml code.
+Due to the great benefit of entirely avoiding termination proofs, it would be
+interesting to extend their approach to improve code extraction.
+
+\subsection{Termination Checking}\label{sec:termcheck}
 
@@
 
-\subsubsection{Sized Types}
+\subsubsection{Sized Types}\label{sec:sizedT}
 
@@
 Coq as well, though it has not yet been adopted~\cite{CHAN_2023}.
-
-\subsubsection{Divide and Conquer Recursion}
-
-Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
-using a recursion scheme in which termination is entirely enforced by its
-typing.  This is a significant advance, since it avoids \emph{completely} the
-need for termination proofs. Their work differs from ours in that they require
-the functional extensionality axiom, and the use of impredicative \coq{Set}.
-The authors justify well the use of impredicative \coq{Set} and its
-compatibility with the functional extensionality axiom. In contrast, our
-development remains entirely \emph{axiom-free}. Another key difference with our
-approach is that they do not discuss what the resulting extracted code
-\emph{looks like} (that is, whether the extracted OCaml code resembles the
-natural formulation of the recursive function). Through experiments, we found
-that their formalisation leads to \ocaml{Obj.magic}, and code with a complex
-structure that may be hard to understand or interface with other OCaml code.
-Due to the great benefit of entirely avoiding termination proofs, it would be
-interesting to extend their approach to improve code extraction.
 

### 2024-03-18 paper/paper.tex
RelW
notes for section 4

--- 2024-03-18 paper/paper.tex
@@
 metavariable \coq{?Goal0} to our desired function. We will use this extensively
-in our examples.
+in our examples.\mvol{This is a very dry overview that doesn't need to be here. We can just jump
+  right into the examples.}
+
+\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
+  be used for a little overview of the rest of the section, like: ``We will go over a series
+  of case studies of various recursive algorithms, showing how they can be encoded
+  in our framework and extracted to idiomatic OCaml code.''}
+
+\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
+  Each of these is written like a case study, with a particular program worked out all the way
+  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
+  a sentence at the beginning of each section re-iterating this to make the whole thing
+  flow more smoothly.}
 
 \subsection{Sorting Algorithms}
+\mvol{Some glue text here, like: ``our first case study is divide-and-conquer sorting
+  algorithms''.}
 So far, our examples could be modelled exclusively using catamorphisms. Here, we
@@
 captured by the structure of an hylomorphisms. The structure of the recursion
-that of a binary tree.  For example, in the case of quicksort, a list is split
+is that of a binary tree.  For example, in the case of quicksort, a list is split
 into a pivot, the label of the node, and two sublists. We define the data
@@
 \subsubsection{Fusing a divide-and-conquer computation}
-
+\mvol{This is good, but it needs some text to connect it to the parent sub-section,
+  like: ``As an aside, we can show how aonther traversal can be fused into the divide-and-conquer
+  algorithm.''}
 Suppose that we map a function to the result of sorting the list. We can
@@
 \subsection{Knapsack}
-
+\mvol{Don't say ``to close this section'' if it is not the last sub-section!}
 To close this section, we show how our framework can represent more advanced
@@
 \subsection{Shortcut Deforestation}
-
+\mvol{Again, this needs a glue sentence, like: ``The final case study we consider is shortcut deforestation on lists.''}
 Shortcut deforestation can be expressed succintly in terms of hylomorphisms and

### 2024-03-18 paper/paper.tex
RelW

--- 2024-03-18 paper/paper.tex
@@
 termination checking (Section~\ref{sec:termcheck}, and sized types
-(Section~\ref{sec:sizedT}, as a way to guarantee termination of nonstructural
+(Section~\ref{sec:sizedT}), as a way to guarantee termination of nonstructural
 recursion.
@@
 
-\subsubsection{Divide and Conquer Recursion}\label{sec:calccoq}
-
-Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
-using a recursion scheme in which termination is entirely enforced by its
-typing.  This is a significant advance, since it avoids \emph{completely} the
-need for termination proofs. Their work differs from ours in that they require
-the functional extensionality axiom, and the use of impredicative \coq{Set}.
-The authors justify well the use of impredicative \coq{Set} and its
-compatibility with the functional extensionality axiom. In contrast, our
-development remains entirely \emph{axiom-free}. Another key difference with our
-approach is that they do not discuss what the resulting extracted code
-\emph{looks like} (that is, whether the extracted OCaml code resembles the
-natural formulation of the recursive function). Through experiments, we found
-that their formalisation leads to \ocaml{Obj.magic}, and code with a complex
-structure that may be hard to understand or interface with other OCaml code.
-Due to the great benefit of entirely avoiding termination proofs, it would be
-interesting to extend their approach to improve code extraction.
-
 \subsection{Termination Checking}\label{sec:termcheck}
@@
 mergesort''~\cite{AbreuDHJMS23}.
+
+\subsubsection{Divide and Conquer Recursion}\label{sec:reccoq}
+
+Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
+using a recursion scheme in which termination is entirely enforced by its
+typing.  This is a significant advance, since it avoids \emph{completely} the
+need for termination proofs. Their work differs from ours in that they require
+the functional extensionality axiom, and the use of impredicative \coq{Set}.
+The authors justify well the use of impredicative \coq{Set} and its
+compatibility with the functional extensionality axiom. In contrast, our
+development remains entirely \emph{axiom-free}. Another key difference with our
+approach is that they do not discuss what the resulting extracted code
+\emph{looks like} (that is, whether the extracted OCaml code resembles the
+natural formulation of the recursive function). Through experiments, we found
+that their formalisation leads to \ocaml{Obj.magic}, and code with a complex
+structure that may be hard to understand or interface with other OCaml code.
+Due to the great benefit of entirely avoiding termination proofs, it would be
+interesting to extend their approach to improve code extraction.
 

### 2024-03-18 paper/paper.tex

--- 2024-03-18 paper/paper.tex
@@
 %\end{itemize}
-
-\subsection{Program Calculation}
+Encoding recursion schemes in Coq is not new. We compare our work with other
+encodings of program calculation techniques in Coq (Section~\ref{sec:calccoq}),
+recursion schemes in Coq (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}).
+Finally, we compare our encoding of recursion schemes to other forms of
+termination checking (Section~\ref{sec:termcheck}, and sized types
+(Section~\ref{sec:sizedT}), as a way to guarantee termination of nonstructural
+recursion.
+
+\subsection{Program Calculation}\label{sec:calccoq}
 
@@
 
-\subsection{Termination Checking}
+\subsection{Termination Checking}\label{sec:termcheck}
 
@@
 
-\subsubsection{Sized Types}
-
-Another approach to certifying termination of recursive functions is the use of
-\emph{sized types}. Sized types were introduced by Hughes et al.\ as a way to
-track/verify various properties of recursive programs, including productivity
-and termination~\cite{HughesPS96}. The core idea of sized types is that types
-express bounds on the sizes of recursive data structures. With this approach,
-algorithms are implemented in the standard functional way, following the
-divide-and-conquer pattern of splitting, recurring, and merging, with the
-addition of the data types being indexed by a static approximation of the
-relative \emph{size}. An advantage of this approach is that it allows for code
-to be written in a natural way. There are costs, however. Programs must be
-written in a way that accounts for the handling of size indexes, and support for
-the size types must be added to the language.
-
-This approach has been used to express nonstructurally recursive
-algorithms in Agda: Copello et al.\ used sized types in a straightforward
-formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
-in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
-Coq as well, though it has not yet been adopted~\cite{CHAN_2023}.
-
-\subsubsection{Divide and Conquer Recursion}
+\subsubsection{Divide and Conquer Recursion}\label{sec:reccoq}
 
@@
 interesting to extend their approach to improve code extraction.
+
+\subsubsection{Sized Types}\label{sec:sizedT}
+
+Another approach to certifying termination of recursive functions is the use of
+\emph{sized types}. Sized types were introduced by Hughes et al.\ as a way to
+track/verify various properties of recursive programs, including productivity
+and termination~\cite{HughesPS96}. The core idea of sized types is that types
+express bounds on the sizes of recursive data structures. With this approach,
+algorithms are implemented in the standard functional way, following the
+divide-and-conquer pattern of splitting, recurring, and merging, with the
+addition of the data types being indexed by a static approximation of the
+relative \emph{size}. An advantage of this approach is that it allows for code
+to be written in a natural way. There are costs, however. Programs must be
+written in a way that accounts for the handling of size indexes, and support for
+the size types must be added to the language.
+
+This approach has been used to express nonstructurally recursive
+algorithms in Agda: Copello et al.\ used sized types in a straightforward
+formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
+in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
+Coq as well, though it has not yet been adopted~\cite{CHAN_2023}.
 

### 2024-03-18 paper/paper.tex
Revision of catamorphisms

--- 2024-03-18 paper/paper.tex
@@
   laws. The key contribution of this paper is that this formalisation is fully
-  axiom-free, and it allows for the extraction of safe, idiomatic OCaml
+  axiom-free, and it allows for the extraction of safe, clean, idiomatic OCaml
   code for a variety of recursive programs. We formalise a series of divide-and
@@
 the performance of the code that is run \emph{within Coq}, it no longer allows
-the extraction of idiomatic OCaml code.  By this, we refer to code
+the extraction of clean, naturally-looking code.  By this, we refer to code
 that (1) does not preserve the recursive structure of common implementations;
@@
 safety guarantees, and will allow the safe interoperation with other OCaml
-code. The absence of \ocaml{Obj.magic} provides even stronger guarantees to the
-extracted code. In particular, since the code that we extract from Coq is
-implemented in a total language, we know that there can be no $\bot$ in its
-definition. This allows the programmer to use ``fast-and-loose'' reasoning on
-the extracted code~\cite{DanielssonHJG06}, which would otherwise be unsafe if
-the code contained \ocaml{Obj.magic}. For example, the programmer would need to
-manually check that the inputs satisfy the necessary assumptions which would be
-impossible to enforce in the presence of unsafe casts. Overall, the absence of
-\ocaml{Obj.magic} makes it easer to reason about, and interoperate with
-extracted code.
+code.
 
@@
 well as their algebraic laws that: (1) is \emph{axiom-free}; and (2)
-allows the extraction of idiomatic OCaml code. While programmers
+allows the extraction of naturally-looking, clean OCaml code. While programmers
 still need to reason about the termination of their programs, our mechanisation
 will allow the use of the algebraic laws of hylomorphisms for program
-reasoning, as well as extracting to idiomatic code. As an
+reasoning, as well as extracting to clean, naturally-looking code. As an
 example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
@@
 use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:containers}).
-However, reasoning about container equality
-will require us to consider both functional extensionality and heterogeneous
-equality. We avoid these axioms by introducing a
-custom equivalence relation on types (Section~\ref{sec:types-morphisms}).
+However, reasoning about container equality will require us to consider both
+functional extensionality and heterogeneous equality. We avoid these axioms by
+introducing a custom equivalence relation on types
+(Section~\ref{sec:types-morphisms}).
 
@@
 positive.  Due to this, we are going to restrict to \emph{polynomial functors}
-as represented by containers.  A container is defined by a type of
-\emph{shapes} and a \emph{family of position types} indexed by shapes.
+as represented by containers. A container is defined by a type of \emph{shapes}
+and a \emph{family of position types} indexed by shapes.
 \begin{coqcode}
@@
 \end{coqcode}
-For the functor \coq{F} we introduced in the previous paragraph we would define
-\coq{Shape} as \coq{unit + A} indicating there are two constructors in the data
-type and one of these contains a piece of data of type \coq{A}.  Moreover, we
-would define the positions as \coq{Pos (inl tt) = Empty_set} and 
-\coq{Pos (inr a) = unit} indicating that the first constructor does not have
-any type variables and the second constructor has one type variable.
+For the functor \coq{F} we introduced in the previous paragraph we would
+define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
+the data type and one of these contains a piece of data of type \coq{A}.
+Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
+and  \coq{Pos (inr a) = unit} indicating that the first constructor does not
+have any type variables and the second constructor has one type variable.
 
@@
 \end{coqcode}
-In our running example, the type \coq{App X}, for
-some \coq{X} has two inhabitants.  The first is a pair composed
-by \coq{inl tt : Shape} and a function \coq{Pos (inl tt) -> X}. This latter is
-in fact a function of type \coq{Empty_set -> X}. This pair is, therefore
-isomorphic to the type \coq{unit}, the type with only one inhabitant.  The
-second inhabitant is the pair composed by \coq{inr a : Shape}, for
-some \coq{a : A}, and a function \coq{Pos (inr a) -> X}.  This latter type is
-equal to \coq{unit -> X} which represents the elements of \coq{X} and therefore
-is isomorphic to \coq{X}.  This particular pair of shapes and positions is 
-therefore isomorphic to the type \coq{unit + A * X}.
+In our running example, the type \coq{App X}, for some \coq{X} has two
+inhabitants.  The first is a pair composed by \coq{inl tt : Shape} and a
+function \coq{Pos (inl tt) -> X}. This latter is in fact a function of
+type \coq{Empty_set -> X}. This pair is, therefore isomorphic to the
+type \coq{unit}, the type with only one inhabitant.  The second inhabitant is
+the pair composed by \coq{inr a : Shape}, for some \coq{a : A}, and a
+function \coq{Pos (inr a) -> X}.  This latter type is equal to \coq{unit -> X}
+which represents the elements of \coq{X} and therefore is isomorphic to \coq{X}.
+This particular pair of shapes and positions is  therefore isomorphic to the
+type \coq{unit + A * X}.
 
@@
 \subsection{Algebras and Catamorphisms}
-\label{sec:algebras} An algebra is a set \coq{X} (the \emph{carrier} of the
-algebra) together with some operations on it\footnote[1]{Sometimes and algebra
-comes equipped with an equational theory, e.g. the theory of monoids, but here
-we focus on algebraically free algebras}.  For example, given such a type
-\coq{X} we can define the monoid operations as \coq{u : unit -> X} for unit and
-\coq{m : X * X -> X} for multiplication. Notice that \coq{F X = unit + X * X}
-is a functor on \coq{X} which means we can equivalently describe these two maps
-as a single one of type \coq{F X -> X} which we call an algebra for the functor
-\coq{F} satisfying the unit and associative laws of the monoid.  In this work
-we will be using algebras to describe the operations of a data type and so we
-will not be needing additional equations on these operations.
-
-Since in this work we will be considering only polynomial functors we need to
-formalise algebras in terms of containers and containers extensions. Given a
-type \coq{A} and a container \coq{F} an algebra for \coq{F} over the carrier
-\coq{A} is a map of the following type:
+\label{sec:algebras} An algebra is a set \coq{A} (the
+\emph{carrier} of the algebra) together with some operations on it sometimes
+subject to some equational laws. For example, given such a type \coq{A} we can
+define the monoid operations as \coq{u : unit -> A} for unit
+and \coq{m : A * A -> A} for multiplication. Notice
+that \coq{F X = unit + X * X} is a functor on \coq{X} which means we can
+equivalently describe these two maps as a single one of type \coq{F X -> X}
+which we call the algebra for the functor \coq{F} satisfying the unit and
+associative laws of the monoid. In this work we will be using algebras to
+describe the operations of a data type and so we will not be needing additional
+equations on these operations.
+
+Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
+by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of
+type \coq{App F A ~> A}:
 \begin{coqcode}
@@
 \end{coqcode}
-
-\dcas{The writing below is confusing, but I am not sure how to fix it ... Can't
-we start with the list-generating functor? The monoid example is great, but as 
-soon as we move to talking about least-fixed points, it feels as if we'd better
-go straight into F X = unit + A * X}
-Now, the \emph{easiest} and most effective way to turn any type $A$ into a
-monoid is to define it as the least solution to the equation
-\begin{coqcode}
-Free A = A + unit + Free A * Free A
-\end{coqcode}
-with \coq{Free : Type -> Type}. In Coq, this is the least fixed-point of the
-functor \coq{unit + A + X * X} on \coq{X}.  This is isomorphic to the type of
-lists over \coq{A} hence it is the least fixed-point for the
-functor \coq{unit + A * X} as well. This is the least ``set'' generated by the
-constructors \coq{Empty : unit -> List A} and \coq{Cons : A * List A -> List A}.
-These constructors give an algebra \coq{LFix_in : unit + A * List A -> List A}
-which has and inverse which we call \coq{LFix_out}.
-
-We now formalise this idea using container extensions.  For a container
-extension a data type is formalised as an inductive data type parametrised by
-the functor represented by \coq{F}:
+The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
+where the structure map is an isomorphism. This is sometimes referred to as the
+\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind this name shortly. \footnote[1]{To be more precise,
+initial algebras are isomorphisms by Lambek's lemma, but isomorphisms of
+functors do not necessarily correspond to initial algebras.}  We define the
+least fixed-point of \coq{F} as an inductive type:
 \begin{coqcode}
@@
 \end{coqcode}
-% To generalise recursion schemes we abstract the shape of the data using functors.
-% Assuming a base functor \coq{F} a catamorphism is then the unique
-% solution to the equation
-% \begin{minted}{coq}
-% cata alg = alg . fmap (cata alg) . InOp
-% \end{minted}
-% which does precisely what we described in the previous paragraph.
-%
-% where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
-% can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
-% $F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
-% $F$-coalgebra), $\InOp$ is the isomoprhism on the least fixed-point of $F$,
-% namely $\InOp : \mu F \to F\mu F$, and $\OutOp$ is the isomorphism on the
-% greatest fixed-point of $F$, namely $F \nu F \to \nu F$. When the unique
-% solutions to the above equations exist we denote them as $\cata{a}$ and
-% $\ana{c}$ respectively.
-%
-%%\dcas{Boring stuff}
-%The definition of \coq{LFix} in terms of \coq{App} requires us to define our
-%own custom induction scheme for initial algebras:
-%\begin{coqcode}
-%Lemma LFix_ind [P : LFix -> Prop]
-%  : (forall x : App F LFix,
-%        (forall e : Pos (shape x), P (cont x e)) ->
-%        P (LFix_in x))
-%    -> forall l : LFix, P l.
-%\end{coqcode}
+\mpav{David, make it more Coq syntax like}
+where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
+As an example, the initial \coq{F}-algebra for the
+functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
+being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
+operation \coq{Cons : A * LFix F -> LFix F}.
+
 We define \coq{LFix} as a setoid, where its equivalence relation can be
-described as the least fixed point of \coq{AppR}.  We define smart constructors
-for the isomorphism of least fixed points as respectful morphisms.
+described as the least fixed point of \coq{AppR} and we define smart
+constructors for the isomorphism of least fixed points as respectful morphisms:
 \begin{coqcode}
@@
 
-The most simple and well-known recursion scheme is the \emph{fold} or
-\emph{catamorphism} which consumes data. First a fold \emph{destructs} a data
-type using \coq{LFix_out}, then calls itself recursively and then folds the
-results of the recursive call using an \coq{F}-algebra.
+The least fixed-point for \coq{F} is the initial \coq{F}-algebra in the sense
+that it gives rise to an inductive recursion scheme.  Specifically, for any
+other \coq{F}-algebra there exists a \emph{unique} map, known as a \emph{fold} or
+\emph{catamorphism}, such that it structurally deconstructs the data type
+using \coq{LFix_out}, calls itself recursively and then composes the result of
+the recursive call using the given an \coq{F}-algebra. In other words, for any
+give \coq{F}-algebra there exists a unique \coq{F}-algebra homomorphism from the
+initial one. We define it in Coq as follows:
 \begin{coqcode}
@@
 \end{coqcode}
-We prove that catamorphisms satisfy the universal law, that is that there is a
-unique morphism of \coq{F}-algebras from the least fixed-point of \coq{F}:
+We prove that catamorphisms satisfy the universality property we explained previously:
 \begin{coqcode}
@@
 \end{coqcode}
+In other words, if there is any other \coq{f} with the same structural recursive
+shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
+catamorphism.
 
@@
 %\end{itemize}
-Encoding recursion schemes in Coq is not new. We compare our work with other
-encodings of program calculation techniques in Coq (Section~\ref{sec:calccoq}),
-recursion schemes in Coq (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}).
-Finally, we compare our encoding of recursion schemes to other forms of
-termination checking (Section~\ref{sec:termcheck}, and sized types
-(Section~\ref{sec:sizedT}), as a way to guarantee termination of nonstructural
-recursion.
-
-\subsection{Program Calculation}\label{sec:calccoq}
+
+\subsection{Program Calculation}
 
@@
 algorithms~\cite{larchey2022braga}. Their work does not focus on encoding
-generic recursion schemes, and proving their algebraic laws.  Castro-Perez et
+generic recursion schemes, and proving their algebraic laws.  Castro et
 al.~\cite{farmsCastro} encode the laws of hylomorphisms as part of the type
@@
 
-\subsection{Termination Checking}\label{sec:termcheck}
+\subsection{Termination Checking}
 
@@
 
-\subsubsection{Divide and Conquer Recursion}\label{sec:reccoq}
+\subsubsection{Sized Types}
+
+Another approach to certifying termination of recursive functions is the use of
+\emph{sized types}. Sized types were introduced by Hughes et al.\ as a way to
+track/verify various properties of recursive programs, including productivity
+and termination~\cite{HughesPS96}. The core idea of sized types is that types
+express bounds on the sizes of recursive data structures. With this approach,
+algorithms are implemented in the standard functional way, following the
+divide-and-conquer pattern of splitting, recurring, and merging, with the
+addition of the data types being indexed by a static approximation of the
+relative \emph{size}. An advantage of this approach is that it allows for code
+to be written in a natural way. There are costs, however. Programs must be
+written in a way that accounts for the handling of size indexes, and support for
+the size types must be added to the language.
+
+This approach has been used to express nonstructurally recursive
+algorithms in Agda: Copello et al.\ used sized types in a straightforward
+formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
+in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
+Coq as well, though it has not yet been adopted~\cite{CHAN_2023}.
+
+\subsubsection{Divide and Conquer Recursion}
 
@@
 interesting to extend their approach to improve code extraction.
-
-\subsubsection{Sized Types}\label{sec:sizedT}
-
-Another approach to certifying termination of recursive functions is the use of
-\emph{sized types}. Sized types were introduced by Hughes et al.\ as a way to
-track/verify various properties of recursive programs, including productivity
-and termination~\cite{HughesPS96}. The core idea of sized types is that types
-express bounds on the sizes of recursive data structures. With this approach,
-algorithms are implemented in the standard functional way, following the
-divide-and-conquer pattern of splitting, recurring, and merging, with the
-addition of the data types being indexed by a static approximation of the
-relative \emph{size}. An advantage of this approach is that it allows for code
-to be written in a natural way. There are costs, however. Programs must be
-written in a way that accounts for the handling of size indexes, and support for
-the size types must be added to the language.
-
-This approach has been used to express nonstructurally recursive
-algorithms in Agda: Copello et al.\ used sized types in a straightforward
-formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
-in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
-Coq as well, though it has not yet been adopted~\cite{CHAN_2023}.
 

### 2024-03-18 paper/paper.tex
Examples

@@
 \section{Extraction}
-The steps are the following:
-\begin{enumerate}
-  \item Write a specification as a morphism \coq{f : A ~> B}.
-  \item Use Coq's generalised rewriting and the laws of morphisms to 
-    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
-  \item Project the function \coq{h : A -> B} from \coq{g}
-    (i.e. \coq{h = app g}).
-  \item Extract \coq{h} to OCaml.
-\end{enumerate}
-Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
-\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
-proofs from the code, and inline our definitions. To ease the application of
-these steps, we provide the following helper type:
-\begin{coqcode}
-  Record Ext (f : A ~> B) := 
-    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
-\end{coqcode}
-Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
-since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
-will be the function \coq{target} that is extensionally equal to \coq{f}.  We
-use a shortcut tactic, \coq{calculate}, that is a synonym to 
-\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
-\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
-complete the definition with \coq{reflexivity}, and this will instantiate the
-metavariable \coq{?Goal0} to our desired function. We will use this extensively
-in our examples.\mvol{This is a very dry overview that doesn't need to be here. We can just jump
-  right into the examples.}
-
-\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
-  be used for a little overview of the rest of the section, like: ``We will go over a series
-  of case studies of various recursive algorithms, showing how they can be encoded
-  in our framework and extracted to idiomatic OCaml code.''}
-
-\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
-  Each of these is written like a case study, with a particular program worked out all the way
-  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
-  a sentence at the beginning of each section re-iterating this to make the whole thing
-  flow more smoothly.}
+We go in this section through a series of case studies of various recursive
+algorithms.  We show how they can be encoded in our framework, how can we do
+program calculation techniques for optimisation, and how can they be extracted
+to idiomatic OCaml code. Our examples are the Quicksort and Mergesort
+algorithms (Section~\ref{sec:sorting}), dynammic programming and Knapsack
+(Section~\ref{sec:knap}), and examples of the shortcut deforestation
+optimisation in our framework (Section~\ref{sec:shortcut}).
+%The steps are the following:
+%\begin{enumerate}
+%  \item Write a specification as a morphism \coq{f : A ~> B}.
+%  \item Use Coq's generalised rewriting and the laws of morphisms to 
+%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
+%  \item Project the function \coq{h : A -> B} from \coq{g}
+%    (i.e. \coq{h = app g}).
+%  \item Extract \coq{h} to OCaml.
+%\end{enumerate}
+%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
+%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
+%proofs from the code, and inline our definitions. To ease the application of
+%these steps, we provide the following helper type:
+%\begin{coqcode}
+%  Record Ext (f : A ~> B) := 
+%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
+%\end{coqcode}
+%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
+%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
+%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
+%use a shortcut tactic, \coq{calculate}, that is a synonym to 
+%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
+%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
+%complete the definition with \coq{reflexivity}, and this will instantiate the
+%metavariable \coq{?Goal0} to our desired function. We will use this extensively
+%in our examples.
+%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
+%  right into the examples.}
+%
+%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
+%  be used for a little overview of the rest of the section, like: ``We will go over a series
+%  of case studies of various recursive algorithms, showing how they can be encoded
+%  in our framework and extracted to idiomatic OCaml code.''}
+%
+%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
+%  Each of these is written like a case study, with a particular program worked out all the way
+%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
+%  a sentence at the beginning of each section re-iterating this to make the whole thing
+%  flow more smoothly.}
 
 \subsection{Sorting Algorithms}
-\mvol{Some glue text here, like: ``our first case study is divide-and-conquer sorting
-  algorithms''.}
-So far, our examples could be modelled exclusively using catamorphisms. Here, we
-show examples that require recursive hylomorphisms and termination proofs.  We
-complete the sorting algorithm examples by applying fusion optimisation.
+Our first case study is divide-and-conquer sorting algorithms.  Encoding them in
+our framework will require the use of recursive hylomorphisms and termination
+proofs.  We complete the sorting algorithm examples by applying fusion
+optimisation.
 

### 2024-03-18 paper/paper.tex
anamorphisms

--- 2024-03-18 paper/paper.tex
@@
 \label{sec:coalg}
-While folds \emph{destruct} data, their dual, the unfolds, \emph{construct}
-data. In particular, given a seed value and an \emph{observation}, the unfold
-corecursively  ``runs'' the observations starting from the initial seed.
+Algebras and catamorphisms dualise nicely to the coinductive setting.  The dual
+of an algebra is a coalgebra. A coalgebra can be thought of as an
+\emph{observation} map. For example, for a type of states \coq{X} and an
+alphabet type \coq{L} we can define a labelled transition system (LTS)
+on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{FX = L * X}
+implementing the \emph{transition} map. In particular, for a
+state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
+where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
+
+In general, for a functor \coq{F}, an \coq{F}-coalgebra is a pair of a
+type \coq{X} called the \emph{carrier} of the coalgebra and a structure
+map \coq{X -> FX}.
+
+In our development we use the following notation for coalgebras:
+\begin{coqcode}
+Notation Coalg F A := (A ~> App F A).
+\end{coqcode}
+
+Dually to the initial \coq{F}-algebra, a final \coq{F}-coalgebra is the greatest
+fixed-point for a functor \coq{F}. We define it using a coinductive data type:
+\begin{coqcode}
+CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
+\end{coqcode}
+where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix-in} is its
+inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
+defined as a setoid, with an equivalence relation that is the greatest fixpoint
+of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
+greatest fixed points
+\begin{coqcode}
+g_in : App F (GFix F) ~> GFix F                g_out : GFix F ~> App F (GFix F)
+\end{coqcode}
+
+The greatest fixed-point is a terminal \coq{F}-coalgebra in the sense that it
+yields a coinductive recursion scheme. Specifically, for any
+other \coq{F}-coalgebra there exists a \emph{unique} map, called the
+\emph{unfold} or \emph{anamorphism}, such that it applies the observation map,
+corecursively generates the rest of the computation and composes the result of
+the corecursive call by using the algebra \coq{GFix_in}. In other words, for any
+given \coq{F}-coalgebra there exists a unique \coq{F}-coalgebra homomorphism
+into the terminal \coq{F}-coalgebra.
+
 % Consider the transition system below composed by three states \coq{x1}, \coq{x2}
@@
 % \end{tikzcd}\]
-The seed function can be modelled by a coalgebra. For example, for a type of
-states \coq{X} and an alphabet type \coq{L} we can define a (labelled)
-transition system (LTS) on \coq{X} as a function \coq{c : X -> L * X}
-implementing the \emph{transition} map. In particular, for a
-state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
-where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
-Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
-carrier \coq{X} is a map \coq{X -> F X}. In the previous example \coq{c} is
-an \coq{F}-coalgebra on the carrier \coq{X} with \coq{F X = L * X}.
-In our development we use the following notation for coalgebras:
-\begin{coqcode}
-Notation Coalg F A := (A ~> App F A).
-\end{coqcode}
-Now an anamorphism takes such a seed function and runs it while collecting all
-the observable behaviours. In order to do so we define the type of infinite
-streams of labels which is the greatest fixed-point for the functor \coq{A * X}.
-We define this in Coq as follows:
-\begin{coqcode}
-CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
-\end{coqcode}
-where \coq{GFix_in}  and \coq{GFix_out} are the two inverses witnessing the
-isomorphism. Similarly to \coq{LFix}, \coq{GFix} is defined as a setoid, with
-an equivalence relation that is the greatest fixpoint of \coq{AppR}.
-Additionally, we define smart constructors for the isomorphism of greatest
-fixed points
-\begin{coqcode}
-g_in : App F (GFix F) ~> GFix F                g_out : GFix F ~> App F (GFix F)
-\end{coqcode}
-\dcas{The explanation below is weird ... We have not defined 'ana', but state
-their universal property?}
-What witnesses the fact that this is the greatest fixed-point is the universal
-property of the anamorphism. For a given \coq{F}-coalgebra this is the unique
-map that runs the seed function and composes the information from the first step
+% Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
+% carrier \coq{X} is a map \coq{X -> F X}. In the previous example \coq{c} is
+% an \coq{F}-coalgebra on the carrier \coq{X} with \coq{F X = L * X}.
+
+\mpav{David, I need full def of this ana}
+We define anamorphisms as follows:
+\begin{coqcode}
+ana : forall `{F : Cont Sh P} `{setoid A}, Coalg F A ~> A ~> GFix F
+\end{coqcode}
+From this definition the universality property falls out:
 \begin{coqcode}
@@
 \end{coqcode}
-In our Coq formalisation, we define anamorphisms as the
-unique \coq{F}-coalgebra homomorphisms, that are maps of the following type:
-\begin{coqcode}
-ana : forall `{F : Cont Sh P} `{setoid A}, Coalg F A ~> A ~> GFix F
-\end{coqcode}
+In words, for any \coq{F}-coalgebra, if there is any other function \coq{f} that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the same coalgebra.
 

### 2024-03-18 paper/paper.tex
Examples

@@
 \subsection{Sorting Algorithms}
-Our first case study is divide-and-conquer sorting algorithms.  Encoding them in
-our framework will require the use of recursive hylomorphisms and termination
-proofs.  We complete the sorting algorithm examples by applying fusion
-optimisation.
+Our first case study is divide-and-conquer sorting algorithms.  Encoding them
+in our framework will require the use of recursive hylomorphisms and
+termination proofs.  We complete the sorting algorithm examples by applying
+fusion optimisation.
 
@@
 captured by the structure of an hylomorphisms. The structure of the recursion
-is that of a binary tree.  For example, in the case of quicksort, a list is split
-into a pivot, the label of the node, and two sublists. We define the data
+is that of a binary tree.  For example, in the case of quicksort, a list is
+split into a pivot, the label of the node, and two sublists. We define the data
 functor of trees as follows:
@@
 \end{coqcode}
-We define the tree container, \coq{TreeF}, in a straightforward way, by making
+These define a container, \coq{TreeF}, in a straightforward way, by making
 the positions of type \coq{Tpos} only valid in \coq{Node}.
@@
 that it respects the ``less-than'' relation on the length of the lists.
-We then define, and extract, the hylomorphism \coq{hylo merge r_split}, which
-extracts to the following OCaml code:
+%\begin{coqcode}
+%Lemma split_fin : respects_relation c_split (@length int) lt.  (* *) Qed.
+%
+%Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
+%  := mk_wf_coalg wf_lt split_fin.
+%\end{coqcode}
+The code that we extract for \coq{hylo merge c_split} is the following:
 \begin{ocamlcode}
@@
 code is similar to a hand-written \coq{qsort}.
-
-The mergesort algorithm can be defined analogously and can be found in the formalisation. [REF LINK].
+The mergesort algorithm can be defined analogously and can be found in the
+formalisation. % [REF LINK].
 
@@
 \subsubsection{Fusing a divide-and-conquer computation}
-\mvol{This is good, but it needs some text to connect it to the parent sub-section,
-  like: ``As an aside, we can show how aonther traversal can be fused into the divide-and-conquer
-  algorithm.''}
+As an example of how can we use program calculation techniques in our 
+framework, we show how another traversal can be fused into the
+divide-and-conquer algorithm using the laws of hylomorphisms.
 Suppose that we map a function to the result of sorting the list. We can
@@
 \begin{coqcode}
-Definition qsort_times_two : Ext (Lmap times_two \o hylo merge tsplit).
-  calculate.  apply hylo_fusion_l. 
-  (* calculate the fused stage *)
-Defined.
+Definition qsort_times_two := Lmap times_two \o hylo merge tsplit.
 \end{coqcode}
@@
 and \coq{times_two} multiplies every element of the list by two.
+We can use Coq's generalised rewriting, and \coq{hylo_fusion_l} to fuse
+\coq{times_two} into the RHS hylomorphism in \coq{qsort_times_two}.
 After applying hylo fusion and the necessary rewrites, the hylomorphism that we
-extract is 
-\coq{hylo (merge \o natural times_two) tsplit}. Here, \coq{natural} defines
-a natural transformation between container extensions by applying a function
-to the shapes that preserves the validity of the positions. In this case,
-the function is \coq{times_two} that multiplies every pivot in the Quicksort
-tree by two. The extracted OCaml code is a single recursive traversal:
+extract is \coq{hylo (merge \o natural times_two) tsplit}. In this definition,
+\coq{natural} defines a natural transformation by applying \coq{times_two} to
+the shapes, and \coq{times_two} multiplies every pivot in the Quicksort tree by
+two. Our formalisation contains a proof that \coq{natural} is indeed a natural
+transformation, which relies on the fact that it preserves the structure of the
+shapes and, therefore, the validity of the positions.  The extracted OCaml code
+is a single recursive traversal:
 \begin{ocamlcode}
@@
 \subsection{Knapsack}
-\mvol{Don't say ``to close this section'' if it is not the last sub-section!}
-To close this section, we show how our framework can represent more advanced
-recursion schemes, such as \emph{dynamorphisms} for dynamic programming, by
-using their encoding as a hylomorphism. We use the knapsack example
-from~\cite{HinzeWG15}. First, we define memoisation tables in terms
-of containers. 
+We focus now on the formalisation and extraction of \emph{dynamorphisms} for
+dynamic programming, by using their encoding as a hylomorphism. We use the
+knapsack example from~\cite{HinzeWG15}. Dynamorphisms build a memoisation table
+that stores intermediate results, alongside the current computation. The
+algebra used to define a dynamorphism can access this memoisation table to
+speed up computation.  First, we define memoisation tables in terms of
+containers. 
 \begin{coqcode}
@@
 \end{coqcode}
+These tables map ``paths'' in the least fixed point of \coq{Memo} to elements
+of type \coq{A}. For example, if \coq{G} is a list-generating functor, these
+paths will be natural numbers. 
 Using these definitions, a dynamorphism is defined as follows:
@@
 \end{coqcode}
-Coq's code extraction mechanism is unable to inline several definitions in
-this case, but the extracted code can be trivially inlined to produce the
-following:
+Coq's code extraction mechanism is unable to inline several definitions in this
+case. We have manually inlined the extracted code for simplicity. The reader
+can check in our \textbf{Artefact} that the extracted code can be trivially
+inlined to produce the following:
 \begin{ocamlcode}
@@
 \subsection{Shortcut Deforestation}
-\mvol{Again, this needs a glue sentence, like: ``The final case study we consider is shortcut deforestation on lists.''}
-Shortcut deforestation can be expressed succintly in terms of hylomorphisms and
-their laws~\cite{TakanoM95}. In particular, given a function:
+The final case study we consider is shortcut deforestation on lists.  Shortcut
+deforestation can be expressed succintly in terms of hylomorphisms and their
+laws~\cite{TakanoM95}. In particular, given a function:
 \begin{coqcode}
@@
 Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
-  rewrite hylo_map_fusion, <- acid_rain.
-  simpl; reflexivity.
+  rewrite hylo_map_fusion, <- acid_rain. simpl; reflexivity.
 Defined.

### 2024-03-18 paper/paper.tex
minor

--- 2024-03-18 paper/paper.tex
@@
 Functors are functions \coq{F : Type -> Type} which additionally have a
-map \coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
+map
+\begin{coqcode}
+fmap : forall A B, (A -> B) -> F A -> F B
+\end{coqcode}
+witnessing the idea that a
 functor represents a container for abstract data that can be manipulated without

### 2024-03-18 paper/paper.tex
Examples

@@
 
-\title{Magic-free Code Extraction of Generic Programs and Recursion Schemes in
-Coq}
+\title{Mechanising Recursion Schemes with Magic-Free Coq Extraction}
 
@@
 
-\subsection{Sorting Algorithms}
+\subsection{Sorting Algorithms}\label{sec:sorting}
 Our first case study is divide-and-conquer sorting algorithms.  Encoding them
@@
 
-\subsection{Knapsack}
+\subsection{Knapsack}\label{sec:knap}
 We focus now on the formalisation and extraction of \emph{dynamorphisms} for
@@
 
-\subsection{Shortcut Deforestation}
+\subsection{Shortcut Deforestation}\label{sec:shortcut}
 The final case study we consider is shortcut deforestation on lists.  Shortcut

### 2024-03-18 paper/paper.tex
New Abstrac

--- 2024-03-18 paper/paper.tex
@@
 \begin{abstract}
-  Generic programming with recursion schemes provides a powerful abstraction
-  for reasoning about program equivalences and deriving optimisations which has
-  been successfully applied to functional programming. However, formalising
-  this technique in a type theory, (e.g. Coq) often requires compromises, such
-  as imposing performance penalties, requiring the assumption of additional
-  axioms, or introducing unsafe casts into extracted code (for example,
-  \ocaml{Obj.magic} in OCaml).
-
-  This paper presents a formalisation of hylomorphisms (a recursion scheme
-  capturing divide-and-conquer algorithms) in Coq, along with their algebraic
-  laws. The key contribution of this paper is that this formalisation is fully
-  axiom-free, and it allows for the extraction of safe, idiomatic OCaml
-  code for a variety of recursive programs. We formalise a series of divide-and
-  conquer, dynamic programming, and mutual recursion programs and demonstrate
-  that the extracted OCaml code for the programs formalised in our framework is
-  efficient, resembles code that a human programmer would write, and contains
-  no occurrences of \ocaml{Obj.magic}. We also present a machine-checked proof
-  of the well-known short-cut fusion optimisation.
+  Generic programming with recursion schemes provides a powerful abstraction for
+structuring recursion while ensuring termination and providing reasoning about
+program equivalences as well as deriving optimisations which has been
+successfully applied to functional programming. Formalising recursion schemes in
+a type theory offers additional termination guarantees, but it often requires
+compromises affecting the resulting code, such as imposing performance
+penalties, requiring the assumption of additional axioms, or introducing unsafe
+casts into extracted code (for example, \ocaml{Obj.magic} in OCaml).
+
+This paper presents the first Coq formalisation of a recursion scheme, called
+the \emph{hylomorphism}, along with its algebraic laws allowing for the
+mechanisation of all recognised (terminating) recursive algorithms. The key
+contribution of this paper is that this formalisation is fully axiom-free
+allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
+framework by formalising a series of algorithms based on different recursive paradigms such as divide-and conquer, dynamic programming,
+and mutual recursion and demonstrate that the extracted OCaml code for
+the programs formalised in our framework is efficient, resembles code that a
+human programmer would write, and contains no occurrences of \ocaml{Obj.magic}.
+We also present a machine-checked proof of the well-known short-cut fusion
+optimisation.
 \end{abstract}

### 2024-03-18 paper/paper.tex
Stuff

@@
 \section{Formalising Recursion Schemes}
-Recursion schemes provide an abstract way to consume and generate data. We know
+Recursion schemes provide an abstract way to consume and generate data. We now
 proceed onto describing how to formalise hylomorphisms in Coq. We first
@@
 \end{coqcode}
-\mpav{David, make it more Coq syntax like}
 where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
@@
 alphabet type \coq{L} we can define a labelled transition system (LTS)
-on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{FX = L * X}
+on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{F X = L * X}
 implementing the \emph{transition} map. In particular, for a
@@
 type \coq{X} called the \emph{carrier} of the coalgebra and a structure
-map \coq{X -> FX}.
+map \coq{X -> F X}.
 
@@
 \end{coqcode}
-where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix-in} is its
+where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix_in} is its
 inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
@@
 
-\mpav{David, I need full def of this ana}
 We define anamorphisms as follows:
 \begin{coqcode}
-ana : forall `{F : Cont Sh P} `{setoid A}, Coalg F A ~> A ~> GFix F
+Definition ana_f_ (c : Coalg F A) :=
+  cofix f x := 
+    match c x with | MkApp sx kx => GFix_in (MkApp sx (fun e => f (kx e))) end.
+
+Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*... ana_f_ ... *)
 \end{coqcode}
@@
 \begin{coqcode}
-Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
+Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix F)
 : f =e ana h <-> f =e g_in \o fmap f \o h.
 \end{coqcode}
-In words, for any \coq{F}-coalgebra, if there is any other function \coq{f} that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the same coalgebra.
+In words, for any \coq{F}-coalgebra, if there is any other function \coq{f}
+that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the
+same coalgebra.
 
@@
 A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra 
-\coq{c : Coalg F A} such that \coq{forall x, RecF c x}.
+\coq{c} such that \coq{forall x, RecF c x}.
 Recursive hylomorphisms are implemented in Coq recursively on the structure of
@@
 \end{coqcode}
-We now prove this is isomorphic to the least fixed-point of the functor \coq{F}
+We now prove this is isomorphic to the least fixed-point of the functor
+\coq{F}.
 We take the catamorphism from the finite elements of \coq{GFix F} to the
-inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}.
-Its inverse is the catamorphism on the restriction of \coq{g_in} to the finite elements of \coq{GFix}, which we denote by \coq{lg_in}.
-The following lemmas prove the isomorphism:
+inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
+inverse is the catamorphism on the restriction of \coq{g_in} to the finite
+elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
+prove the isomorphism:
 \begin{coqcode}
 Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
-Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata l_in \o cata lg_in =e id.
-\end{coqcode}
+Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
+\end{coqcode}
+The finite subtype of \coq{GFix F} allows us to compose
+catamorphisms and anamorphisms, by using the above isomorphism. In our work,
+however, we use \emph{recursive} anamorphisms, defined as \coq{hylo l_in c} for
+a recursive coalgebra \coq{c}, which compose easily with
+catamorphisms.
 % This predicate is equivalent to the proposition that
@@
 destructors:
-\mpav{The following algebra should be called a\_in or l\_in whatever}
+% \mpav{The following algebra should be called a\_in or l\_in whatever}
 \begin{coqcode}
 Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.
-
-(* Converting TreeF to ITreeF *)
 Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
@@
 \end{coqcode}
-
-% \subsubsection{Quicksort}
-
 The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with
@@
 \coq{A * Sg} and positions \coq{Pg}, given a a container \coq{G : Cont Sg Pg}.
-We define a function to insert elements into the memoisation tables:
+We define a function to insert elements into the memo tables:
 \begin{coqcode}
@@
 \end{coqcode}
-And two functions to inspect the head of a memoisation table, and remove an
+And two functions to inspect the head of a memo table, and remove an
 element from it:
@@
 Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
-memoisation table. The definition of the algebra can use this table to lookup
+memo table. The definition of the algebra can use this table to lookup
 elements, instead of triggering a further recursive call. Elements are inserted
@@
 axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
-
 However, we prove a specific version of this theorem for the list generating
@@
 This definition highlights why we cannot define equality of the positions and
-the contained elements generically (recall Section~\ref{sec:TODO}): the
+the contained elements generically: the
 contained types may not have (yet) an associated equivalence. In other words, we
@@
 by traversing the structure and equating the shapes and positions. This
-definition is equivalent to taking the fixpoint of \coq{AppR} in
-Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
-positions, the contents must be related by \coq{LFixR}.
+definition is equivalent to taking the fixpoint of \coq{AppR} where the shapes
+must be equivalent and, for equal positions, the contents must be related by
+\coq{LFixR}.
 \begin{coqcode}

### 2024-03-18 paper/paper.tex
Stuff

@@
 
+\section{Conclusions and Future Work}
+
+Hylomorphisms are a general recursion scheme that can encode any other 
+recursion scheme, and that satisfy a number of algebraic laws that can be used
+to reason about program equivalences. Despite their useful properties, not many 
+pieces of work formalise them in a proof assistant, partly due to the difficulty
+of dealing with termination, and reasoning about program equivalences. In this work,
+we tackle these problems by
+
 \end{document}

### 2024-03-18 paper/paper.tex
Stuff

--- 2024-03-18 paper/paper.tex
@@
 
+\section{Conclusions and Future Work}
+
+Hylomorphisms are a general recursion scheme that can encode any other
+recursion scheme, and that satisfy a number of algebraic laws that can be used
+to reason about program equivalences. Not many
+pieces of work formalise them in a proof assistant, partly due to the
+difficulty of dealing with termination, and reasoning about functional
+extensionality. In this work, we tackle these problems in a \emph{fully
+axiom-free} way that targets the extraction of idiomatic OCaml code.  This
+formalisation allows the use of program calculation techniques in Coq to derive
+formally optimised implementations from naive specifications. Furthermore, the
+rewritings that are applied to specifications are formal, machine-checked
+proofs that the resulting program is extensionally equal to the input
+specification.
+
+Remaining axiom-free forces us to deal with the well-known
+\emph{setoid hell}. As part of the future improvements, we will study how to
+mitigate this problem. At the moment, we use a short ad-hoc tactic that is able
+to automatically discharge many of these proofs in simple settings.  We will
+study the more thorough and systematic use of proof automation for respectful
+morphisms.  Generalisd rewriting in proofs involving setoids tends to be quite
+slow, due to the large size of the terms that need to be rewritten. Sometimes,
+this size is hidden in implicit arguments and coercions. We will study
+alternative formulations to try to improve the performance of the rewriting
+tactics (e.g.  canonical structures).  Currently, Coq is unable to inline a
+number of trivially inlineable definitions.  We will study alternative
+definitions, or extensions to Coq's code extraction mechanisms to force the
+full inlining of all container code that is used in hylomorphisms.  Finally,
+proving termination still remains a hurdle. In our framework this reduces to
+proving that the anamorphism
+terminates in all inputs, and we provide a convenient connection to
+well-founded recursion. Furthermore, recursive coalgebras compose with natural
+transformations, which allows the reuse
+of a number of core recursive coalgebras. A possible interesting future line of work
+is the use of the approach by Abreu et al.~\cite{AbreuDHJMS23} in combination with ours to
+improve code extraction from divide-and-conquer computations whose termination
+does not require an external proof.
+
 \bibliography{references.bib}

### 2024-03-18 paper/paper.tex
Intro

--- 2024-03-18 paper/paper.tex
@@
 penalties, requiring the assumption of additional axioms, or introducing unsafe
-casts into extracted code (for example, \ocaml{Obj.magic} in OCaml).
+casts into extracted code (e.g. \ocaml{Obj.magic} in OCaml).
 
@@
 allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
-framework by formalising a series of algorithms based on different recursive paradigms such as divide-and conquer, dynamic programming,
-and mutual recursion and demonstrate that the extracted OCaml code for
-the programs formalised in our framework is efficient, resembles code that a
-human programmer would write, and contains no occurrences of \ocaml{Obj.magic}.
-We also present a machine-checked proof of the well-known short-cut fusion
-optimisation.
+framework by formalising a series of algorithms based on different recursive
+paradigms such as divide-and conquer, dynamic programming, and mutual recursion
+and demonstrate that the extracted OCaml code for the programs formalised in our
+framework is efficient, resembles code that a human programmer would write, and
+contains no occurrences of \ocaml{Obj.magic}.  We also present a machine-checked
+proof of the well-known short-cut fusion optimisation.
 \end{abstract}
@@
 will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
-one approach is using the \coq{Fix} combinator, in which recursion is done on
-an \emph{accesibility} predicate on the input, which is a proof that there are
-no infinitely decreasing chains.}. A benefit of this approach is that,
-extracting verified nonstructural recursive functions to OCaml will erase the
-termination proofs, and so produce code that will be closer to what a
-programmer may have written directly in OCaml. However, reasoning about program
-equivalences requires dealing with such termination proofs, and it is common
-practice to use custom reduction lemmas that will be used extensively when
-proving properties about nonstructural recursive functions.
-
-\emph{Structured recursion schemes} and their associated algebraic laws provide
-an easier framework for reasoning about program equivalences.  In particular,
-\ocaml{qsort} is an instance of a \emph{divide-and-conquer}
-algorithm which, in the functional programming literature is known as a
-hylomorphism~\cite{MeijerFP91, HuIT96}.  Hinze et al.~\cite{HinzeWG15} showed
-that every recursive program (recursion scheme) can be formalised as a hylomorphism.
-Furthermore, hylomorphism laws can capture a number of useful equivalences,
-ranging from common optimisations such as \emph{short-cut
+one approach is using the \coq{Fix} combinator, in which recursion is done on an
+\emph{accesibility} predicate on the input, which is a proof that there are no
+infinitely decreasing chains.}. A benefit of this approach is that, extracting
+verified nonstructural recursive functions to OCaml will erase the termination
+proofs, and so produce code that will be closer to what a programmer may have
+written directly in OCaml. However, reasoning about program equivalences
+requires dealing with such termination proofs, and it is common practice to use
+custom reduction lemmas that will be used extensively when proving properties
+about nonstructural recursive functions.
+
+\emph{Structured recursion schemes} have been successfully employed in
+functional programming to structure recursive programs, for example, quicksort
+and mergesort are both instances of a divide-and-conquer algorithm which can be
+formalised using hylomorphisms~\cite{MeijerFP91, HuIT96}. In fact,  Hinze et
+al.~\cite{HinzeWG15} showed that \emph{every recursive program can be formalised
+as a hylomorphism}.  Furthermore, hylomorphism laws can capture a number of
+useful equivalences, ranging from common optimisations such as \emph{short-cut
 fusion}~\cite{TakanoM95}, to semi-automatic
 parallelisations~\cite{Gibbons96:Third, farmsCastro}.
-
-Not many approaches mechanise structured recursion
-schemes in proof assistants.  Recently, Abreu et al.~\cite{AbreuDHJMS23} encode
-an algebraic approach to divide-and-conquer computations, in which termination
-is entirely enforced by the typing discipline of the recursion scheme. However,
-although their approach solves the problem of termination proofs, as well as
-the performance of the code that is run \emph{within Coq}, it no longer allows
-the extraction of idiomatic OCaml code.  By this, we refer to code
-that (1) does not preserve the recursive structure of common implementations;
-and (2) lead to unsafe casts \ocaml{Obj.magic} in the generated code. 
-In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
-Although the use of \ocaml{Obj.magic} should be safe in the code
-extracted by Coq, it has been proven that, for higher-order programs, simple
-interoperations can lead to incorrect behaviour or even
-segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
-\ocaml{Obj.magic} means that the extracted code will have stronger
-safety guarantees, and will allow the safe interoperation with other OCaml
-code. The absence of \ocaml{Obj.magic} provides even stronger guarantees to the
-extracted code. In particular, since the code that we extract from Coq is
-implemented in a total language, we know that there can be no $\bot$ in its
-definition. This allows the programmer to use ``fast-and-loose'' reasoning on
-the extracted code~\cite{DanielssonHJG06}, which would otherwise be unsafe if
-the code contained \ocaml{Obj.magic}. For example, the programmer would need to
-manually check that the inputs satisfy the necessary assumptions which would be
-impossible to enforce in the presence of unsafe casts. Overall, the absence of
-\ocaml{Obj.magic} makes it easer to reason about, and interoperate with
-extracted code.
+% Recursion schemes have successfully been applied to functional programming to
+% structure recursive programs, however, because of the presence of
+% non-termination in languages like (e.g.) Haskell, they cannot guarantee
+% recursive definitions well-defined. % This is due to the possibility of applying a
+% %
+% recursion scheme to a non-terminating (co)algebra, but also to the unrestricted
+% % use of data functors.
+
+However, when used in the context of languages with general recursion like
+(e.g.) Haskell, recursion schemes cannot ensure termination, but while formalising
+recursion schemes in a type theory does provide stronger termination guarantees, to the best of our knowledge, not many attempts have been made at
+mechanise structured recursion schemes in a proof assistant.
+
+Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to divide-and-conquer
+computations in which termination is entirely enforced by the typing discipline. Their approach solves the problem of
+termination proofs as well as the performance of the code that is run
+\emph{within Coq}, but it does not allow the extraction of idiomatic OCaml code.
+By this we refer to code that (1) does not preserve the recursive structure of
+common implementations; and (2) lead to unsafe casts like \ocaml{Obj.magic} in the
+generated code.  In OCaml, \ocaml{Obj.magic} is problematic in that it has been proven that,
+for higher-order programs, simple interoperations can lead to incorrect
+behaviour or even segfaults~\cite{forster:hal-04329663} and, more generally, it invalidates the fast-and-loose principle~\cite{DanielssonHJG06}.
+% The absence of \ocaml{Obj.magic} provides even stronger
+% guarantees to the extracted code. In particular, the extracted code from Coq is
+% well-defined allowing the programmer to use the so-called \emph{fast-and-loose
+% reasoning}~\cite{DanielssonHJG06} on the extracted code. % For example, the
+% % programmer would need to manually
+% %
+% check that the inputs satisfy the necessary assumptions which would be
+% % impossible to enforce in the presence of unsafe casts. Overall, the absence
+% % of \ocaml{Obj.magic} makes it easer to reason about, and interoperate with
+% % extracted code.
 
 This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
-well as their algebraic laws that: (1) is \emph{axiom-free}; and (2)
-allows the extraction of idiomatic OCaml code. While programmers
-still need to reason about the termination of their programs, our mechanisation
-will allow the use of the algebraic laws of hylomorphisms for program
-reasoning, as well as extracting to idiomatic code. As an
-example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
-the following code:
+well as their algebraic laws that: (1) is \emph{axiom-free}; and (2) allows the
+extraction of idiomatic OCaml code. While programmers still need to reason about
+the termination of their programs, our mechanisation will allow the use of the
+algebraic laws of hylomorphisms for program reasoning, as well as extracting to
+idiomatic code. As an example our mechanisation of \ocaml{qsort} as a
+hylomorphism will produce the following code:
 \begin{coqcode}
@@
 \section{Formalising Recursion Schemes}
-Recursion schemes provide an abstract way to consume and generate data. We now
+Recursion schemes provide an abstract way to consume and generate data. We know
 proceed onto describing how to formalise hylomorphisms in Coq. We first
@@
 \end{coqcode}
+\mpav{David, make it more Coq syntax like}
 where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
@@
 alphabet type \coq{L} we can define a labelled transition system (LTS)
-on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{F X = L * X}
+on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{FX = L * X}
 implementing the \emph{transition} map. In particular, for a
@@
 type \coq{X} called the \emph{carrier} of the coalgebra and a structure
-map \coq{X -> F X}.
+map \coq{X -> FX}.
 
@@
 \end{coqcode}
-where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix_in} is its
+where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix-in} is its
 inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
@@
 
+\mpav{David, I need full def of this ana}
 We define anamorphisms as follows:
 \begin{coqcode}
-Definition ana_f_ (c : Coalg F A) :=
-  cofix f x := 
-    match c x with | MkApp sx kx => GFix_in (MkApp sx (fun e => f (kx e))) end.
-
-Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*... ana_f_ ... *)
+ana : forall `{F : Cont Sh P} `{setoid A}, Coalg F A ~> A ~> GFix F
 \end{coqcode}
@@
 \begin{coqcode}
-Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix F)
+Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix)
 : f =e ana h <-> f =e g_in \o fmap f \o h.
 \end{coqcode}
-In words, for any \coq{F}-coalgebra, if there is any other function \coq{f}
-that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the
-same coalgebra.
+In words, for any \coq{F}-coalgebra, if there is any other function \coq{f} that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the same coalgebra.
 
@@
 A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra 
-\coq{c} such that \coq{forall x, RecF c x}.
+\coq{c : Coalg F A} such that \coq{forall x, RecF c x}.
 Recursive hylomorphisms are implemented in Coq recursively on the structure of
@@
 \end{coqcode}
-We now prove this is isomorphic to the least fixed-point of the functor
-\coq{F}.
+We now prove this is isomorphic to the least fixed-point of the functor \coq{F}
 We take the catamorphism from the finite elements of \coq{GFix F} to the
-inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
-inverse is the catamorphism on the restriction of \coq{g_in} to the finite
-elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
-prove the isomorphism:
+inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}.
+Its inverse is the catamorphism on the restriction of \coq{g_in} to the finite elements of \coq{GFix}, which we denote by \coq{lg_in}.
+The following lemmas prove the isomorphism:
 \begin{coqcode}
 Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
-Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
-\end{coqcode}
-The finite subtype of \coq{GFix F} allows us to compose
-catamorphisms and anamorphisms, by using the above isomorphism. In our work,
-however, we use \emph{recursive} anamorphisms, defined as \coq{hylo l_in c} for
-a recursive coalgebra \coq{c}, which compose easily with
-catamorphisms.
+Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata l_in \o cata lg_in =e id.
+\end{coqcode}
 % This predicate is equivalent to the proposition that
@@
 destructors:
-% \mpav{The following algebra should be called a\_in or l\_in whatever}
+\mpav{The following algebra should be called a\_in or l\_in whatever}
 \begin{coqcode}
 Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.
+
+(* Converting TreeF to ITreeF *)
 Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
@@
 \end{coqcode}
+
+% \subsubsection{Quicksort}
+
 The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with
@@
 \coq{A * Sg} and positions \coq{Pg}, given a a container \coq{G : Cont Sg Pg}.
-We define a function to insert elements into the memo tables:
+We define a function to insert elements into the memoisation tables:
 \begin{coqcode}
@@
 \end{coqcode}
-And two functions to inspect the head of a memo table, and remove an
+And two functions to inspect the head of a memoisation table, and remove an
 element from it:
@@
 Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
-memo table. The definition of the algebra can use this table to lookup
+memoisation table. The definition of the algebra can use this table to lookup
 elements, instead of triggering a further recursive call. Elements are inserted
@@
 axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
+
 However, we prove a specific version of this theorem for the list generating
@@
 This definition highlights why we cannot define equality of the positions and
-the contained elements generically: the
+the contained elements generically (recall Section~\ref{sec:TODO}): the
 contained types may not have (yet) an associated equivalence. In other words, we
@@
 by traversing the structure and equating the shapes and positions. This
-definition is equivalent to taking the fixpoint of \coq{AppR} where the shapes
-must be equivalent and, for equal positions, the contents must be related by
-\coq{LFixR}.
+definition is equivalent to taking the fixpoint of \coq{AppR} in
+Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
+positions, the contents must be related by \coq{LFixR}.
 \begin{coqcode}

### 2024-03-18 paper/paper.tex
Refs

@@
 methods have been proposed for checking that these properties hold. A common
-approach is guardedness checking~\cite{10.5555/646535.695850}, which is a
+approach is guardedness checking~\cite{Eduardo94}, which is a
 \emph{syntactic} check that definitions avoid the introduction of
@@
 In particular, the problem of \emph{nonstructural recursion} (including
-divide-and-conquer algorithms) is well-studied~\cite{BOVE_KRAUSS_SOZEAU_2016}.
+divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
 % A distinction between standard and
@@
 in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
-Coq as well, though it has not yet been adopted~\cite{CHAN_2023}.
+Coq as well, though it has not yet been adopted~\cite{ChanLB23}.
 
@@
 \bibliography{references.bib}
+
+\end{document}
+
+\endinput
 \appendix

### 2024-03-18 paper/paper.tex
Introduction fits in the page

--- 2024-03-18 paper/paper.tex
@@
 However, when used in the context of languages with general recursion like
-(e.g.) Haskell, recursion schemes cannot ensure termination, but while formalising
-recursion schemes in a type theory does provide stronger termination guarantees, to the best of our knowledge, not many attempts have been made at
-mechanise structured recursion schemes in a proof assistant.
-
-Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to divide-and-conquer
-computations in which termination is entirely enforced by the typing discipline. Their approach solves the problem of
-termination proofs as well as the performance of the code that is run
-\emph{within Coq}, but it does not allow the extraction of idiomatic OCaml code.
-By this we refer to code that (1) does not preserve the recursive structure of
-common implementations; and (2) lead to unsafe casts like \ocaml{Obj.magic} in the
-generated code.  In OCaml, \ocaml{Obj.magic} is problematic in that it has been proven that,
-for higher-order programs, simple interoperations can lead to incorrect
-behaviour or even segfaults~\cite{forster:hal-04329663} and, more generally, it invalidates the fast-and-loose principle~\cite{DanielssonHJG06}.
-% The absence of \ocaml{Obj.magic} provides even stronger
+(e.g.) Haskell, recursion schemes cannot ensure termination, but while
+formalising recursion schemes in a type theory does provide stronger termination
+guarantees, to the best of our knowledge, not many attempts have been made at
+mechanising structured recursion schemes.
+
+Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
+divide-and-conquer computations in which termination is entirely enforced by the
+typing discipline. Their approach solves the problem of termination proofs as
+well as the performance of the code that is run \emph{within Coq}, but it does
+not allow the extraction of idiomatic OCaml code.  By this we refer to code that
+(1) does not preserve the recursive structure of common implementations; and (2)
+lead to unsafe casts like \ocaml{Obj.magic} in the generated code.  This latter is problematic in that, for higher-order programs, simple interoperations can lead to incorrect behaviour or
+even segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates
+the fast-and-loose principle~\cite{DanielssonHJG06}. % The absence
+% of \ocaml{Obj.magic} provides even stronger
 % guarantees to the extracted code. In particular, the extracted code from Coq is
@@
 \end{coqcode}
-The contributions we make in this paper are as follows:
+
+Finally, in this paper we provide a framework for \emph{generic programming} with recursion
+schemes in Coq, with a proof of their algebraic laws and program equivalences for clean code extraction:
 \begin{itemize}
-  \item We provide a framework for \emph{generic programming} with recursion
-    schemes in Coq, with a proof of their algebraic laws and 
-    program equivalences, that allows clean code extraction.
-  \item We formalise divide and conquer, dynamic programming, and
-    mutual recursion algorithms.
-  \item We verified the short-cut fusion optimisation in Coq and
-    extract optimised code to OCaml.
+  \item In Section~\ref{sec:containers} we formalise the type of container functors suitably modified for program extraction
+  \item In Section~\ref{sec:recursion-schemes} we formalise hylomorhisms and prove their universal property
+  \item In Section~\ref{sec:examples} we provide examples of divide and conquer, dynamic programming, and mutual recursion algorithms and we verify the short-cut fusion optimisation and show the extracted optimised code to OCaml.
 \end{itemize}
@@
 \section{Mechanising Extractable Container Functors}
+\label{sec:containers}
 In order to abstract recursion patters we need to be able to abstract away from
@@
 who have a initial algebras. A common approach to construct such functors is to
-use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:containers}).
+use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:functors}).
 However, reasoning about container equality will require us to consider both
@@
 \subsection{Functors and Containers}
-\label{sec:containers}
+\label{sec:functors}
 Functors are functions \coq{F : Type -> Type} which additionally have a
@@
 \section{Formalising Recursion Schemes}
+\label{sec:recursion-schemes}
 Recursion schemes provide an abstract way to consume and generate data. We now
@@
 \section{Extraction}
+\label{sec:examples}
 We go in this section through a series of case studies of various recursive
@@
 \section{Related Work}
+\label{sec:related-work}
 %

### 2024-03-18 paper/paper.tex
Refs
Introduction with my ref

@@
 the proof assistant. Examples of this include common sorting algorithms, such
-as quicksort and mergesort.
+as quicksort:
 %One example is Quicksort, written in OCaml
@@
 accepted by a proof assistant.  The reason is that the \ocaml{divide}
-function \emph{destructuring} the input ``dives'' deeper in the input returning
+function \emph{destructuring} the input dives deeper in the input returning
 two sublists with the head as a pivot.
 
-The main approach for implementing nonstructural recursion in Coq is to use
+The main approach for implementing non-structural recursion in Coq is to use
 \emph{well-founded recursion}, where recursive definitions are coupled with
@@
 functional programming to structure recursive programs, for example, quicksort
-and mergesort are both instances of a divide-and-conquer algorithm which can be
-formalised using hylomorphisms~\cite{MeijerFP91, HuIT96}. In fact,  Hinze et
-al.~\cite{HinzeWG15} showed that \emph{every recursive program can be formalised
-as a hylomorphism}.  Furthermore, hylomorphism laws can capture a number of
-useful equivalences, ranging from common optimisations such as \emph{short-cut
+and mergesort are both instances of a divide-and-conquer algorithm which in
+terms of recursion schemes can be can be formalised as an
+hylomorphism~\cite{MeijerFP91, HuIT96}. Furthermore, it has been shown by Hinze
+et al.~\cite{HinzeWG15} that hylomorphisms provide the basic building block of
+every recursion scheme. In particular, any complex recursion scheme can be
+transmogrified down into an hylomorphisms by means of an adjunction.
+Furthermore, hylomorphism laws can capture a number of useful equivalences,
+ranging from common optimisations such as \emph{short-cut
 fusion}~\cite{TakanoM95}, to semi-automatic
@@
 well as the performance of the code that is run \emph{within Coq}, but it does
-not allow the extraction of idiomatic OCaml code.  By this we refer to code that
-(1) does not preserve the recursive structure of common implementations; and (2)
-lead to unsafe casts like \ocaml{Obj.magic} in the generated code.  This latter is problematic in that, for higher-order programs, simple interoperations can lead to incorrect behaviour or
-even segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates
-the fast-and-loose principle~\cite{DanielssonHJG06}. % The absence
-% of \ocaml{Obj.magic} provides even stronger
-% guarantees to the extracted code. In particular, the extracted code from Coq is
-% well-defined allowing the programmer to use the so-called \emph{fast-and-loose
-% reasoning}~\cite{DanielssonHJG06} on the extracted code. % For example, the
-% % programmer would need to manually
-% %
-% check that the inputs satisfy the necessary assumptions which would be
-% % impossible to enforce in the presence of unsafe casts. Overall, the absence
-% % of \ocaml{Obj.magic} makes it easer to reason about, and interoperate with
-% % extracted code.
-
-This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
-well as their algebraic laws that: (1) is \emph{axiom-free}; and (2) allows the
+not allow for extraction of idiomatic OCaml code. This is problematic since code
+extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
+Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, most uses of extraction
+(1) do not preserve the recursive structure of common implementations; and (2)
+lead to unsafe casts like \ocaml{Obj.magic} in the generated code. This latter
+is also problematic in that, for higher-order programs, simple interoperations
+can lead to incorrect behaviour or even segfaults~\cite{forster:hal-04329663}
+and, moreover, it invalidates the fast-and-loose
+principle~\cite{DanielssonHJG06}.
+
+This work presents the first formalisation in Coq of \emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
 extraction of idiomatic OCaml code. While programmers still need to reason about
@@
 algebraic laws of hylomorphisms for program reasoning, as well as extracting to
-idiomatic code. As an example our mechanisation of \ocaml{qsort} as a
-hylomorphism will produce the following code:
-\begin{coqcode}
-let rec qsort = function | [] -> []
-  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
-              let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
-              app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
-\end{coqcode}
-
-Finally, in this paper we provide a framework for \emph{generic programming} with recursion
-schemes in Coq, with a proof of their algebraic laws and program equivalences for clean code extraction:
+idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}). % As an example, our mechanisation of \ocaml{qsort} as a
+% hylomorphism will produce the following code:
+% \begin{coqcode}
+% let rec qsort = function | [] -> []
+%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
+%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
+%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
+% \end{coqcode}
+
+To summarise, our contributions are as follows. In this paper we provide a
+framework for \emph{generic programming} with recursion schemes in Coq, with a
+proof of their algebraic laws and program equivalences for clean code
+extraction:
 \begin{itemize}
-  \item In Section~\ref{sec:containers} we formalise the type of container functors suitably modified for program extraction
-  \item In Section~\ref{sec:recursion-schemes} we formalise hylomorhisms and prove their universal property
-  \item In Section~\ref{sec:examples} we provide examples of divide and conquer, dynamic programming, and mutual recursion algorithms and we verify the short-cut fusion optimisation and show the extracted optimised code to OCaml.
+  \item In Section~\ref{sec:containers} we formalise the type of container functors ensuring the presence of least and greatest fixed-points for functors and suitably adapted for program extraction
+  \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and hylomorhisms and their universal properties
+  \item In Section~\ref{sec:examples} we use the framework to formalise examples of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
+Furthermore we verify the short-cut fusion optimisation and show the extracted optimised code to OCaml.
 \end{itemize}

### 2024-03-18 paper/paper.tex
Gitattributes

@@
 
-This paper presents the first Coq formalisation of a recursion scheme, called
+This paper presents the first Coq formalisation to our knowledge of a recursion scheme, called
 the \emph{hylomorphism}, along with its algebraic laws allowing for the
@@
 
-This work presents the first formalisation in Coq of \emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
+This work presents the first Coq formalisation to our knowledge of
+\emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
 extraction of idiomatic OCaml code. While programmers still need to reason about
@@
 recursion scheme, and that satisfy a number of algebraic laws that can be used
-to reason about program equivalences. Not many
-pieces of work formalise them in a proof assistant, partly due to the
+to reason about program equivalences. To our knowledge, this is their first formalisation
+in Coq. This is partly due to the
 difficulty of dealing with termination, and reasoning about functional
@@
 study the more thorough and systematic use of proof automation for respectful
-morphisms.  Generalisd rewriting in proofs involving setoids tends to be quite
+morphisms.  Generalised rewriting in proofs involving setoids tends to be quite
 slow, due to the large size of the terms that need to be rewritten. Sometimes,

### 2024-03-19 paper/paper.tex
fixes

@@
 \emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
-extraction of idiomatic OCaml code. While programmers still need to reason about
+extraction of idiomatic OCaml code. The full mechanisation can be found on Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While programmers still need to reason about
 the termination of their programs, our mechanisation will allow the use of the
 algebraic laws of hylomorphisms for program reasoning, as well as extracting to
-idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}). % As an example, our mechanisation of \ocaml{qsort} as a
+idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
+
+% As an example, our mechanisation of \ocaml{qsort} as a
 % hylomorphism will produce the following code:
@@
 The mergesort algorithm can be defined analogously and can be found in the
-formalisation. % [REF LINK].
+formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].
 
@@
 \begin{ocamlcode}
-let rec qsort_times_two = function
-| [] -> []
-| h :: t ->
-  let (l, r) = partition (fun x0 -> leb x0 h) t in
+let rec qsort_times_two = function | [] -> []
+| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
             let x0 = fun p -> qsort_times_two (match p with
-                                     | Lbranch -> l
-                                     | Rbranch -> r) in
+                                 | Lbranch -> l | Rbranch -> r) in
             app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
@@
 case. We have manually inlined the extracted code for simplicity. The reader
-can check in our \textbf{Artefact} that the extracted code can be trivially
+can check in our artefact that the extracted code can be trivially
 inlined to produce the following:
 \begin{ocamlcode}
-let knapsack wvs x =
-  let (y, _) =
+let knapsack wvs x = let (y, _) =
     (let rec f x0 =
@@
            cont = fun _ -> fn } }
-     in f x).lFix_out.shape
-  in y
+     in f x).lFix_out.shape in y
 \end{ocamlcode}

### 2024-03-19 paper/paper.tex
Recursive Hylomorphisms

--- 2024-03-18 paper/paper.tex
@@
 \label{sec:rec-coalgebras}
-Hylomorphisms capture the concept of \emph{divide-and-conquer} algorithms.
-First the input is destructured (\emph{divide}) in smaller parts. This
-operation may or may not be structural. Next the results from the subparts are
-computed recursively and finally the results are composed together
+Hylomorphisms capture the concept of \emph{divide-and-conquer} algorithms where
+the input is first destructured (\emph{divide}) in smaller parts by means of a
+coalgebra which are computed recursively and then composed back together
 (\emph{conquer}) by means of an algebra.
 
-A catamorphism is a special kind of hylomorphism where the coalgebra
-map \coq{InOp} only destructures the input structurally, for example,
-decomposing a list into its head and tail. A \emph{recursive
-coalgebra}~\cite{AdamekMM19,CaprettaUV04} is a coalgebra which does not
-necessarily decompose the input structurally, still maintaining the size of the
-input smaller. This is what, for example, happens in quicksort where the
-partition function, given a list, returns two smaller sublists and a pivot
-element.
-
-Thus, in general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
+In general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
 unique solution (when it exists) to the equation
@@
 \end{equation}
-However this solution does not necessarily exist for an arbitrary pair of
-algebras and coalgebras. In fact, this definition cannot be accepted by Coq,
-since there is no structurally smaller argument to do recursion on and there is
-no corecursive definition that is guaranteed to be syntactically guarded -- as
-required by Coq -- due to the use of an arbitrary algebra $a$.
-To address these issues, we mechanise \emph{recursive
+As we stated earlier, a solution to this equation does not exist for an arbitrary  algebra and coalgebra pair and, in fact, this definition cannot be accepted by Coq.
+
+In order to find a solution we restrict ourselves to the so-called
+\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  An example of a
+recursive coalgebra is the \coq{partition} function in quicksort which
+destructures a list into a pivot and two sublist and as long as the sublists are
+smaller the partitioning function still yields a unique solution to the
+recursion scheme.
+
+We mechanise \emph{recursive
 hylomorphisms} which are guaranteed to have a unique solution to the

### 2024-03-19 paper/paper.tex

@@
 
-This work presents the first formalisation in Coq of \emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
-extraction of idiomatic OCaml code. While programmers still need to reason about
+This work presents the first Coq formalisation to our knowledge of
+\emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
+extraction of idiomatic OCaml code. The full mechanisation can be found on Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While programmers still need to reason about
 the termination of their programs, our mechanisation will allow the use of the
 algebraic laws of hylomorphisms for program reasoning, as well as extracting to
-idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}). % As an example, our mechanisation of \ocaml{qsort} as a
+idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
+
+% As an example, our mechanisation of \ocaml{qsort} as a
 % hylomorphism will produce the following code:

### 2024-03-19 paper/paper.tex
stuff

@@
 structuring recursion while ensuring termination and providing reasoning about
-program equivalences as well as deriving optimisations which has been
+program equivalences, as well as deriving optimisations which have been
 successfully applied to functional programming. Formalising recursion schemes in
@@
 However, when used in the context of languages with general recursion like
-(e.g.) Haskell, recursion schemes cannot ensure termination, but while
+ Haskell, recursion schemes cannot ensure termination, but while
 formalising recursion schemes in a type theory does provide stronger termination
@@
 extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
-Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, most uses of extraction
+Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, in Abreu et al.'s approach, extraction
 (1) do not preserve the recursive structure of common implementations; and (2)
@@
 algebraic laws of hylomorphisms for program reasoning, as well as extracting to
-idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
+idiomatic code (see \ocaml{qsort} in Section~\ref{sec:examples}).
 
@@
 \begin{coqcode}
-Definition qsort_times_two := Lmap times_two \o hylo merge tsplit.
+Definition qsort_times_two := Lmap times_two \o hylo merge c_split.
 \end{coqcode}
@@
 After applying hylo fusion and the necessary rewrites, the hylomorphism that we
-extract is \coq{hylo (merge \o natural times_two) tsplit}. In this definition,
+extract is \coq{hylo (merge \o natural times_two) c_split}. In this definition,
 \coq{natural} defines a natural transformation by applying \coq{times_two} to

### 2024-03-25 paper/paper.tex
Edited this intro. I couldn't leave it in an non-optimal form.

--- 2024-03-19 paper/paper.tex
@@
 
-This paper presents the first Coq formalisation to our knowledge of a recursion scheme, called
-the \emph{hylomorphism}, along with its algebraic laws allowing for the
-mechanisation of all recognised (terminating) recursive algorithms. The key
-contribution of this paper is that this formalisation is fully axiom-free
-allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
-framework by formalising a series of algorithms based on different recursive
-paradigms such as divide-and conquer, dynamic programming, and mutual recursion
-and demonstrate that the extracted OCaml code for the programs formalised in our
-framework is efficient, resembles code that a human programmer would write, and
-contains no occurrences of \ocaml{Obj.magic}.  We also present a machine-checked
-proof of the well-known short-cut fusion optimisation.
+To the best of our knowledge, this paper presents the first Coq formalisation of
+a recursion scheme, called the \emph{hylomorphism}, along with its algebraic
+laws allowing for the mechanisation of all recognised (terminating) recursive
+algorithms. The key contribution of this paper is that this formalisation is
+fully axiom-free allowing for the extraction of safe, idiomatic OCaml code. We
+exemplify the framework by formalising a series of algorithms based on different
+recursive paradigms such as divide-and conquer, dynamic programming, and mutual
+recursion and demonstrate that the extracted OCaml code for the programs
+formalised in our framework is efficient, resembles code that a human programmer
+would write, and contains no occurrences of \ocaml{Obj.magic}.  We also present
+a machine-checked proof of the well-known short-cut fusion optimisation.
 \end{abstract}
@@
 The main approach for implementing non-structural recursion in Coq is to use
-\emph{well-founded recursion}, where recursive definitions are coupled with
+\emph{well-founded recursion}, where recursive definitions are accompanied by
 \emph{termination proofs}. Using well-founded recursion, the recursive calls
@@
 \emph{accesibility} predicate on the input, which is a proof that there are no
-infinitely decreasing chains.}. A benefit of this approach is that, extracting
-verified nonstructural recursive functions to OCaml will erase the termination
-proofs, and so produce code that will be closer to what a programmer may have
-written directly in OCaml. However, reasoning about program equivalences
-requires dealing with such termination proofs, and it is common practice to use
-custom reduction lemmas that will be used extensively when proving properties
-about nonstructural recursive functions.
-
-\emph{Structured recursion schemes} have been successfully employed in
-functional programming to structure recursive programs, for example, quicksort
-and mergesort are both instances of a divide-and-conquer algorithm which in
-terms of recursion schemes can be can be formalised as an
-hylomorphism~\cite{MeijerFP91, HuIT96}. Furthermore, it has been shown by Hinze
-et al.~\cite{HinzeWG15} that hylomorphisms provide the basic building block of
-every recursion scheme. In particular, any complex recursion scheme can be
-transmogrified down into an hylomorphisms by means of an adjunction.
-Furthermore, hylomorphism laws can capture a number of useful equivalences,
-ranging from common optimisations such as \emph{short-cut
-fusion}~\cite{TakanoM95}, to semi-automatic
-parallelisations~\cite{Gibbons96:Third, farmsCastro}.
+infinitely decreasing chains.}.
+A benefit of this strategy is that when extracting verified non-structural
+recursive functions to OCaml, the additional proofs are discarded, resulting in
+code that closely resembles what a programmer might have written directly in
+OCaml. However, reasoning about program equivalences needs  such termination
+proofs to be kept around, leading to the common practice of employing additional
+reduction lemmas extensively.
+
+% A benefit of this approach is that, extracting
+% verified nonstructural recursive functions to OCaml will erase the additional
+% proofs, thus producing code that will be closer to what a programmer may have
+% written directly in OCaml. However, reasoning about program equivalences
+% requires dealing with such termination proofs, and it is common practice to use
+% custom reduction lemmas that will be used extensively when proving properties
+% about nonstructural recursive functions.
+
+\emph{Structured recursion schemes}~\cite{HinzeW16, HinzeWG15} is a technique
+for structuring recursive definitions that has found a lot of success in
+functional programming. The idea is that, rather than writing a self-refential
+definition, we use a function library, the recursion scheme, which takes as
+input the non-recursive parts of the algorithm and returns the desired recursive
+definition which is guaranteed to terminate by the properties of the recursion
+scheme itself. More complex recursion schemes do exist but it has been
+proven by Hinze et al.~\cite{HinzeWG15} that they are all instances of a divide-and-conquer
+algorithm which goes under the name of hylomorphism~\cite{MeijerFP91,HuIT96}.
+This result, thus, yields a general framework to reason about the algebraic properties of these
+algorithms, e.g. short-cut fusion and semi-automatic parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}.
+
+% A more complex recursion scheme is the hylomorphism~\cite{MeijerFP91, HuIT96} which has a divide-and-conquer
+% structure. It has been shown by Hinze et
+% al.~\cite{HinzeWG15} that any complex recursion scheme can be transmogrified
+% into an hylomorphisms by means of an adjunction. Furthermore, hylomorphism laws
+% can capture a number of useful equivalences, ranging from common optimisations
+% such as \emph{short-cut fusion}~\cite{TakanoM95}, to semi-automatic
+% parallelisations~\cite{Gibbons96:Third, farmsCastro}.
+
 % Recursion schemes have successfully been applied to functional programming to
@@
 However, when used in the context of languages with general recursion like
-(e.g.) Haskell, recursion schemes cannot ensure termination, but while
+(e.g.) Haskell and ML, recursion schemes cannot ensure termination. This is because the
+input information itself could be non-terminating. On the other hand, while
 formalising recursion schemes in a type theory does provide stronger termination
 guarantees, to the best of our knowledge, not many attempts have been made at
-mechanising structured recursion schemes.
+mechanising them.
 
@@
 well as the performance of the code that is run \emph{within Coq}, but it does
-not allow for extraction of idiomatic OCaml code. This is problematic since code
+not allow for extraction of idiomatic OCaml code. This is unfortunate since code
 extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
@@
 
-This work presents the first Coq formalisation to our knowledge of
-\emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
-extraction of idiomatic OCaml code. The full mechanisation can be found on Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While programmers still need to reason about
-the termination of their programs, our mechanisation will allow the use of the
-algebraic laws of hylomorphisms for program reasoning, as well as extracting to
-idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
+This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
+\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code. The
+full mechanisation can be found on
+Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While
+programmers still need to reason about the termination of their programs, our
+mechanisation will allow the use of the algebraic laws of hylomorphisms for
+program reasoning, as well as extracting to idiomatic code (see
+extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
 

### 2024-03-25 paper/paper.tex

@@
 However, when used in the context of languages with general recursion like
-(e.g.) Haskell and ML, recursion schemes cannot ensure termination. This is because the
-input information itself could be non-terminating. On the other hand, while
-formalising recursion schemes in a type theory does provide stronger termination
-guarantees, to the best of our knowledge, not many attempts have been made at
-mechanising them.
+(e.g.) Haskell and ML, recursion schemes cannot ensure termination. This is
+because the input information itself could be non-terminating. On the other
+hand, while formalising recursion schemes in a type theory does provide stronger
+termination guarantees, to the best of our knowledge, not many attempts have
+been made at mechanising them.
 
@@
 extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
-
 % As an example, our mechanisation of \ocaml{qsort} as a

### 2024-10-07 paper/paper.tex
I moved stuff in the background:

@@
 % % % \subsubsection{Histomorphism}
-
 % % % The definition of $\lambda$ is:
@@
 % % % \end{align*}
-
 % % % % \begin{align*}
@@
 % % % % \end{align*}
-
 % % % % Defining the algebra
@@
 % % % % \end{align*}
-
-\section{Mechanising Extractable Container Functors}
-\label{sec:containers}
-In order to abstract recursion patters we need to be able to abstract away from
-the particular shape of the data.  This is achieved by introducing the concept
-of \emph{functors} which have the suitable fixed-point properties, i.e.\ those
-who have a initial algebras. A common approach to construct such functors is to
-use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:functors}).
-However, reasoning about container equality will require us to consider both
-functional extensionality and heterogeneous equality. We avoid these axioms by
-introducing a custom equivalence relation on types
-(Section~\ref{sec:types-morphisms}).
-
-\subsection{Functors and Containers}
-\label{sec:functors}
+\section{Recursion Schemes}
+\subsection{Functors}
 Functors are functions \coq{F : Type -> Type} which additionally have a
@@
 \end{coqcode}
-witnessing the idea that a
-functor represents a container for abstract data that can be manipulated without
-changing the outer structure. For example, the type \coq{List : Type -> Type} is
-a functor and \coq{List A} is the type of lists over elements of type \coq{A}
-with the obvious \coq{fmap} function recursively traversing a list and applying
-the function \coq{A -> B} for each element. Additionally, the type of
-lists \coq{List A} arises as the least fixed-point of the
-functor \coq{F X = unit + (A * X) }.
+witnessing the idea that a functor represents a container for abstract data that
+can be manipulated without changing the outer structure. For example, the type
+\coq{List : Type -> Type} is a functor and \coq{List A} is the type of lists
+over elements of type \coq{A} with the obvious \coq{fmap} function recursively
+traversing a list and applying the function \coq{A -> B} for each element.
+Additionally, the type of lists \coq{List A} arises as the least fixed-point of
+the functor \coq{F X = unit + (A * X) }.
 
@@
 positive. Due to this, we are going to restrict to \emph{polynomial functors}
-as represented by containers. A container is defined by a type of \emph{shapes}
-and a \emph{family of position types} indexed by shapes.
-\begin{coqcode}
-Context (Shape : Type) (Pos : Shape -> Type).
-\end{coqcode}
-For the functor \coq{F} we introduced in the previous paragraph we would
-define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
-the data type and one of these contains a piece of data of type \coq{A}.
-Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
-and  \coq{Pos (inr a) = unit} indicating that the first constructor does not
-have any type variables and the second constructor has one type variable.
-
-At this point an \emph{extension} of this container is a functor defined as
-follows:
-\begin{coqcode}
-Record App (X : Type) := MkApp { shape : Shape; contents : Pos shape -> X }.
-\end{coqcode}
-with the obvious action on morphisms given by post-composition
-with \coq{contents}.
-\begin{coqcode}
-Definition fmap (f : A -> B) (x : App C A) : App C B
-  := {| shape := shape x; contents := fun e => f (contents x e) |}
-\end{coqcode}
-In our running example, the type \coq{App X}, for some \coq{X} has two
-inhabitants.  The first is a pair composed by \coq{inl tt : Shape} and a
-function \coq{Pos (inl tt) -> X}. This latter is in fact a function of
-type \coq{Empty_set -> X}. This pair is, therefore isomorphic to the
-type \coq{unit}, the type with only one inhabitant.  The second inhabitant is
-the pair composed by \coq{inr a : Shape}, for some \coq{a : A}, and a
-function \coq{Pos (inr a) -> X}.  This latter type is equal to \coq{unit -> X}
-which represents the elements of \coq{X} and therefore is isomorphic to \coq{X}.
-This particular pair of shapes and positions is  therefore isomorphic to the
-type \coq{unit + A * X}.
-
-The correspondence between containers and polynomial functors has been
-formalised in this work and can be found in the accompanying code (file
-Container.v). 
-% 
-%Generally, any functor that
-%is isomorphic to a container extension is called a \emph{container functor}.
-%Here we will call \emph{container functor} the functors defined by a specific
-%container extension.
-
-\subsection{Extractable Containers}
-In the previous section we defined containers using dependent types.  However,
-Ocaml's type system is not equipped to handle these. To solve this problem
-Coq's code extraction mechanism will insert unsafe casts.
-
-Consider instead representing the positions of a container using a decidable
-validity predicate (\coq{valid}) which assigns shapes to positions. We use
-boolean functions and coercions from \coq{bool} to \coq{Prop} to represent
-decidable predicates, similarly to SSReflect~\cite{GonthierL09}. The family of
-positions for a given shape in a container can be recovered by defining:
-\begin{coqcode}
-Class Cont (Shape : Type) (APos : Type) := { valid : Shape -> APos -> bool }.
-Record Pos `(C : Cont Shape APos) (shape : Shape) :=
-  ValidPos { val : APos; IsValid : valid shape pos }.
-\end{coqcode}
-Coq's code extraction will now be able to erase the validity predicate, and
-generate OCaml code that is free of unsafe casts. The OCaml code extracted for
-\coq{Pos} will now be exactly the code extracted for \coq{APos}.
-The decidability of the validity predicate is crucial for our purposes of 
-remaining axiom-free. To illustrate this, suppose that we need to show
-the equality of two container extensions. We will need to show that, for
-the same positions, they will produce the same result. In Coq, the goal
-would look as follows: 
-\begin{coqcode}
- k : Pos C s -> X
- P1, P2 : valid s p
- ---------------------------
- k (ValidPos p P1) = k (ValidPos p P2)
-\end{coqcode}
-If \coq{valid} was a regular proposition in \coq{Prop}, it would not possible
-to prove the equality of \coq{P1} and \coq{P2}. However, by using a decidable
-predicate, if we know that \coq{P1} and \coq{P2} are of type 
-\coq{valid s p = true}, then we can prove without any axioms that 
-\coq{P1 = P2 = eq_refl}.
-
-% A container extension is then defined as follows:
-% \begin{coqcode}
-% Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
-% Record App (X : Type) := { shape    : Shape;
-%                           contents : {p | Valid shape p} -> X }.
-% \end{coqcode}
-% This approach solves the problem of the unsafe casts, since the code extracted
-% for \coq{{p | Valid shape pos}} will be an OCaml singleton type defined as the
-% OCaml equivalent to \coq{Pos}.
-
-\subsubsection{Equality of Container Extensions}
-Reasoning about the equality of container extensions is not entirely solved by
-using decidable validity predicates to define the families of positions. In
-general, we want to equate container extensions that have the same shape, and
-that, for equal shape and position, they return the same element. To avoid the
-use of the functional extensionality axiom, we capture this relation with the
-following inductive proposition in Coq:
-\begin{coqcode}
-Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
-| AppR_ext (Es : shape x = shape y)
-           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
-\end{coqcode}
-Note that  we do not care about the validity proof of the positions, only their
-value. This is to simplify (slightly) our proofs. This relation is
-trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.
-
-However, the use of a different equality for container extensions now forces us
-to deal with the fact that some types have different definitions of equality.
-In particular, we want to reason about the equality of functions of types such
-as \coq{App F A -> B} (or \coq{B -> App F A}). Since 
-these types now come with their own equivalence, any function that manipulates
-them needs to be \emph{respectful}. I.e.\ given 
-\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
-we want functions (morphisms) that satisfy the following property:
-\begin{coqcode}
-forall (x y : X), R x y -> R' (f x) (f y)
-\end{coqcode}
-
-\subsection{Types and Morphisms}
-\label{sec:types-morphisms}
-We address the different forms of equality by defining a class of
-\emph{setoids}, types with an associated equivalence relation, and considering
-only functions that respect the associated equivalences, or \emph{proper}
-morphisms with respect to the function \emph{respectfulness} relation. We use
-the type-class mechanism, instead of setoids in Coq's standard library, to help
-Coq's code extraction mechanism remove any occurence of custom equivalence
-relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
-relation of a setoid. By default, we associate every Coq type with the standard
-propositional equality, unless a different equivalence is specified (we allow
-overlapping instances, and Coq's propositional equality takes the lowest
-priority).
-%\begin{coqcode}
-%Reserved Notation "f =e g" (at level 70, no associativity).
-%Class setoid A : Type :=
-%  MkEquiv
-%    { eqRel : A -> A -> Prop;
-%      e_refl : forall x, eqRel x x;
-%      e_sym : forall x y, eqRel x y -> eqRel y x;
-%      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
-%    }.
-%Notation "f =e g" := (eqRel f g).
-%\end{coqcode}
-Given types \coq{A} and  \coq{B}, with their respective
-equivalence relations \coq{eA : A -> A -> Prop} and
-\coq{eB : B -> B -> Prop}, the we define the type
-\coq{A ~> B} to represent \emph{proper} morphisms of the
-respectfulness relation of \coq{R} to \coq{R'}.
-\begin{coqcode}
-Structure morph A {eA : setoid A} B {eB : setoid B} :=
-  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
-Notation "A ~> B" := (@morph A _ B _).
-\end{coqcode}
-We rely on Coq's type class mechanism to fill in the necessary equivalence
-relations. Coq's code extraction mechanism will erase any occurrence
-of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
-the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
-from \coq{A ~> B} to \coq{A -> B}.
-On top of this, we define basic function composition and identity functions:
-\begin{coqcode}
-Notation "f \o g" = (comp f g).
-Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
-Definition id : A ~> A := ...
-\end{coqcode}
-Using custom equivalences and proper morphisms, we redefine the definitions of
-container extensions and container equality. In particular, container
-extensions require a proper morphism to check the validity of positions in
-shapes, and container equality now uses equivalences of shapes and contained
-elements: 
-\begin{coqcode}
-Class Cont `{setoid Shape} (Pos : Type) := { valid : Shape * Pos ~> bool }.
-Inductive AppR `{F : Cont Sf Pf} `{setoid X} (x y : App F X) : Prop :=
-  | AppR_ext (Es : shape x =e shape y)
-             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
-\end{coqcode}
-Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
-however, we chose to use Coq's propositional equality, since these would leads
-to simpler code. We explain why in Section~\ref{sec:algebras}, and the
-mechanisation of initial algebras of container extensions.
-%The definitions in
-%Figure~\ref{fig:poly} are redefined to account for these changes in a
-%straightforward way, i.e. we prove that validity predicates are indeed proper
-%morphisms.
-
-This definition leads to the well-known ``setoid hell'', which we mitigate by
-providing tactics and notations to automatically discharge proofs of
-\coq{app_eq} for morphisms, whenever the types use the standard propositional
-equality, or a combination of propositional and extensional equality. However,
-our compositional approach allows us to build morphisms by plugging in other
-morphisms to our combinators. In our framework, our expectation is that the
-user-provided functions remain small, with relatively straightforward proofs of
-\coq{app_eq}.
-
-However, by using this mechanisation, we gain simplified proofs via the use of
-Coq's \emph{Generalised Rewriting}. Since every morphism \coq{f : A ~> B}
-satisfies the property that if \coq{x =e y}, then \coq{f x =e f y}, we can add
-every morphism as a proper element of Coq's respectfulness relation. In
-practice, this means that we can use the \coq{rewrite} tactic on proofs of
-type \coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
-arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},
-and \coq{transitivity} tactics. For example:
-\begin{coqcode}
-Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D) (H : h \o g =e id), 
-  h \o (g \o f) =e f.
-Proof.  rewrite compA, H.  reflexivity. Qed.
-\end{coqcode}
-
-\subsubsection{Polynomial Types}
-We define a number of equivalences for polynomial types.
-%These equivalences are
-%be used when building proper morphisms that manipulate their respective types.
-\begin{coqcode}
-Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
-Instance pair_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A * B).
-Instance sum_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A + B).
-Instance prop_eq : setoid Prop.
-Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
-\end{coqcode}
-Most of the definitions that involve functions and polynomial types are
-straightworward.  Identity and composition are defined as 
-\coq{fun x => x} and \coq{fun f g x => f (g x)} 
-respectively, and the proofs that they are proper morphisms is straightforward,
-and automatically discharged by Coq. 
-Products are built using function \coq{fun f g x => (f x, g x)},
-with the projections being the standard Coq  \coq{fst} and
-\coq{snd} functions. Similarly, sum injections are encoded using
-Coq's \coq{inl} and \coq{inr} constructors, and
-pattern matching on them uses the function 
-\coq{fun f g x => match x with | inl y => f y | inr y => g y end}.
-The proofs that these morphisms are proper are straightforward.  Finally, we
-also provide functions for currying/uncurrying, and  flipping the arguments of
-a proper morphism. We force most of our definitions to be inlined, to help
-Coq's code extraction mechanism to inline as many of these combinators as
-possible.
-
-We prove the isomorphisms of polynomial types and the equivalent 
-container extensions. As an example, we discuss (informally) the isomorphisms
-of pairs with their equivalent container extensions. Suppose that we
-know that \coq{App F X} is isomorphic to \coq{A}, and
-\coq{App G X} is isomorphic to \coq{B}. Then we can
-show that \coq{App (Prod F G) X} is isomorphic to
-\coq{A * B}.  If we have an element of type 
-\coq{App (Prod F G) X},
-using the \coq{inl} position, we can obtain
-\coq{App F X}.  Similarly, using \coq{inr}, we can
-obtain \coq{App G X}. Since these are the only two valid positions
-in the shape of pairs, we have finished. It is now sufficient to use the
-isomorphisms of \coq{App F X} and \coq{App G X} to
-obtain \coq{A * B}. Similarly if we have \coq{A * B},
-we can first use the isomorphisms of 
-\coq{A} and
-\coq{B} to obtain
-\coq{App F X * App G X}, and then construct the necessary container
-extension. Given \coq{p_inl : Pos l -> Pos (l * r)} (resp. \coq{p_inr}) that
-act as \coq{inl} (resp.\ \coq{inr}) on product positions, and
-\coq{case_pos : (Pos l -> X) -> (Pos r -> X) -> Pos (l * r) -> X} that pattern
-matches on the product positions, the functions that witness the isomorphism
-are:
-\begin{coqcode}
-Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
-  ({| shape := shape (fst x); cont := fun e => cont x (p_inl e) |}, 
-   {| shape := shape (snd x); cont := fun e => cont x (p_inr e) |}).
-Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
-{| shape := (shape (fst x), shape (snd x));
-   cont := case_pos (cont (fst x)) (cont (snd x)) |}.
-\end{coqcode}
-Proving that the composition of these functions is the identity is
-straightforward using the fact that \coq{Prod} containers only have two valid
-positions.
-
-\section{Formalising Recursion Schemes}
-\label{sec:recursion-schemes}
+as represented by containers.
+\subsection{Algebras and Coalgebras}
+\subsection{Hylomorphisms}
 Recursion schemes provide an abstract way to consume and generate data. We now
@@
 formalise algebras for container extensions (Section~\ref{sec:algebras}), then
-we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
-these notions to formalise recursive coalgebras and hylomorphisms
+we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together these
+notions to formalise recursive coalgebras and hylomorphisms
 (Section~\ref{sec:rec-coalgebras}).
@@
 \subsection{Algebras and Catamorphisms}
-\label{sec:algebras} An algebra is a set \coq{A} (the
-\emph{carrier} of the algebra) together with some operations on it sometimes
-subject to some equational laws. For example, given such a type \coq{A} we can
-define the monoid operations as \coq{u : unit -> A} for unit
-and \coq{m : A * A -> A} for multiplication. Notice
-that \coq{F X = unit + X * X} is a functor on \coq{X} which means we can
-equivalently describe these two maps as a single one of type \coq{F X -> X}
-which we call the algebra for the functor \coq{F} satisfying the unit and
-associative laws of the monoid. In this work we will be using algebras to
-describe the operations of a data type and so we will not be needing additional
-equations on these operations.
+\label{sec:algebras} An algebra is a set \coq{A} (the \emph{carrier} of the
+algebra) together with some operations on it sometimes subject to some
+equational laws. For example, given such a type \coq{A} we can define the monoid
+operations as \coq{u : unit -> A} for unit and \coq{m : A * A -> A} for
+multiplication. Notice that \coq{F X = unit + X * X} is a functor on \coq{X}
+which means we can equivalently describe these two maps as a single one of type
+\coq{F X -> X} which we call the algebra for the functor \coq{F} satisfying the
+unit and associative laws of the monoid. In this work we will be using algebras
+to describe the operations of a data type and so we will not be needing
+additional equations on these operations.
 
 Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
-by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of
-type \coq{App F A ~> A}:
+by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of type
+\coq{App F A ~> A}:
 \begin{coqcode}
@@
 where the structure map is an isomorphism. This is sometimes referred to as the
-\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind this name shortly. \footnote[1]{To be more precise,
-initial algebras are isomorphisms by Lambek's lemma, but isomorphisms of
-functors do not necessarily correspond to initial algebras.}  We define the
-least fixed-point of \coq{F} as an inductive type:
+\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind
+this name shortly. \footnote[1]{To be more precise, initial algebras are
+isomorphisms by Lambek's lemma, but isomorphisms of functors do not necessarily
+correspond to initial algebras.}  We define the least fixed-point of \coq{F} as
+an inductive type:
 \begin{coqcode}
@@
 
+\section{Mechanising Extractable Container Functors}
+\label{sec:containers}
+In order to abstract recursion patters we need to be able to abstract away from
+the particular shape of the data.  This is achieved by introducing the concept
+of \emph{functors} which have the suitable fixed-point properties, i.e.\ those
+who have a initial algebras. A common approach to construct such functors is to
+use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:functors}).
+However, reasoning about container equality will require us to consider both
+functional extensionality and heterogeneous equality. We avoid these axioms by
+introducing a custom equivalence relation on types
+(Section~\ref{sec:types-morphisms}).
+
+\subsection{Functors and Containers}
+\label{sec:functors}
+[CUT]
+A container is defined by a type of \emph{shapes}
+and a \emph{family of position types} indexed by shapes.
+\begin{coqcode}
+Context (Shape : Type) (Pos : Shape -> Type).
+\end{coqcode}
+For the functor \coq{F} we introduced in the previous paragraph we would
+define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
+the data type and one of these contains a piece of data of type \coq{A}.
+Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
+and  \coq{Pos (inr a) = unit} indicating that the first constructor does not
+have any type variables and the second constructor has one type variable.
+
+At this point an \emph{extension} of this container is a functor defined as
+follows:
+\begin{coqcode}
+Record App (X : Type) := MkApp { shape : Shape; contents : Pos shape -> X }.
+\end{coqcode}
+with the obvious action on morphisms given by post-composition
+with \coq{contents}.
+\begin{coqcode}
+Definition fmap (f : A -> B) (x : App C A) : App C B
+  := {| shape := shape x; contents := fun e => f (contents x e) |}
+\end{coqcode}
+In our running example, the type \coq{App X}, for some \coq{X} has two
+inhabitants.  The first is a pair composed by \coq{inl tt : Shape} and a
+function \coq{Pos (inl tt) -> X}. This latter is in fact a function of
+type \coq{Empty_set -> X}. This pair is, therefore isomorphic to the
+type \coq{unit}, the type with only one inhabitant.  The second inhabitant is
+the pair composed by \coq{inr a : Shape}, for some \coq{a : A}, and a
+function \coq{Pos (inr a) -> X}.  This latter type is equal to \coq{unit -> X}
+which represents the elements of \coq{X} and therefore is isomorphic to \coq{X}.
+This particular pair of shapes and positions is  therefore isomorphic to the
+type \coq{unit + A * X}.
+
+The correspondence between containers and polynomial functors has been
+formalised in this work and can be found in the accompanying code (file
+Container.v). 
+% 
+%Generally, any functor that
+%is isomorphic to a container extension is called a \emph{container functor}.
+%Here we will call \emph{container functor} the functors defined by a specific
+%container extension.
+
+\subsection{Extractable Containers}
+In the previous section we defined containers using dependent types.  However,
+Ocaml's type system is not equipped to handle these. To solve this problem
+Coq's code extraction mechanism will insert unsafe casts.
+
+Consider instead representing the positions of a container using a decidable
+validity predicate (\coq{valid}) which assigns shapes to positions. We use
+boolean functions and coercions from \coq{bool} to \coq{Prop} to represent
+decidable predicates, similarly to SSReflect~\cite{GonthierL09}. The family of
+positions for a given shape in a container can be recovered by defining:
+\begin{coqcode}
+Class Cont (Shape : Type) (APos : Type) := { valid : Shape -> APos -> bool }.
+Record Pos `(C : Cont Shape APos) (shape : Shape) :=
+  ValidPos { val : APos; IsValid : valid shape pos }.
+\end{coqcode}
+Coq's code extraction will now be able to erase the validity predicate, and
+generate OCaml code that is free of unsafe casts. The OCaml code extracted for
+\coq{Pos} will now be exactly the code extracted for \coq{APos}.
+The decidability of the validity predicate is crucial for our purposes of 
+remaining axiom-free. To illustrate this, suppose that we need to show
+the equality of two container extensions. We will need to show that, for
+the same positions, they will produce the same result. In Coq, the goal
+would look as follows: 
+\begin{coqcode}
+ k : Pos C s -> X
+ P1, P2 : valid s p
+ ---------------------------
+ k (ValidPos p P1) = k (ValidPos p P2)
+\end{coqcode}
+If \coq{valid} was a regular proposition in \coq{Prop}, it would not possible
+to prove the equality of \coq{P1} and \coq{P2}. However, by using a decidable
+predicate, if we know that \coq{P1} and \coq{P2} are of type 
+\coq{valid s p = true}, then we can prove without any axioms that 
+\coq{P1 = P2 = eq_refl}.
+
+% A container extension is then defined as follows:
+% \begin{coqcode}
+% Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
+% Record App (X : Type) := { shape    : Shape;
+%                           contents : {p | Valid shape p} -> X }.
+% \end{coqcode}
+% This approach solves the problem of the unsafe casts, since the code extracted
+% for \coq{{p | Valid shape pos}} will be an OCaml singleton type defined as the
+% OCaml equivalent to \coq{Pos}.
+
+\subsubsection{Equality of Container Extensions}
+Reasoning about the equality of container extensions is not entirely solved by
+using decidable validity predicates to define the families of positions. In
+general, we want to equate container extensions that have the same shape, and
+that, for equal shape and position, they return the same element. To avoid the
+use of the functional extensionality axiom, we capture this relation with the
+following inductive proposition in Coq:
+\begin{coqcode}
+Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
+| AppR_ext (Es : shape x = shape y)
+           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
+\end{coqcode}
+Note that  we do not care about the validity proof of the positions, only their
+value. This is to simplify (slightly) our proofs. This relation is
+trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.
+
+However, the use of a different equality for container extensions now forces us
+to deal with the fact that some types have different definitions of equality.
+In particular, we want to reason about the equality of functions of types such
+as \coq{App F A -> B} (or \coq{B -> App F A}). Since 
+these types now come with their own equivalence, any function that manipulates
+them needs to be \emph{respectful}. I.e.\ given 
+\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
+we want functions (morphisms) that satisfy the following property:
+\begin{coqcode}
+forall (x y : X), R x y -> R' (f x) (f y)
+\end{coqcode}
+
+\subsection{Types and Morphisms}
+\label{sec:types-morphisms}
+We address the different forms of equality by defining a class of
+\emph{setoids}, types with an associated equivalence relation, and considering
+only functions that respect the associated equivalences, or \emph{proper}
+morphisms with respect to the function \emph{respectfulness} relation. We use
+the type-class mechanism, instead of setoids in Coq's standard library, to help
+Coq's code extraction mechanism remove any occurence of custom equivalence
+relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
+relation of a setoid. By default, we associate every Coq type with the standard
+propositional equality, unless a different equivalence is specified (we allow
+overlapping instances, and Coq's propositional equality takes the lowest
+priority).
+%\begin{coqcode}
+%Reserved Notation "f =e g" (at level 70, no associativity).
+%Class setoid A : Type :=
+%  MkEquiv
+%    { eqRel : A -> A -> Prop;
+%      e_refl : forall x, eqRel x x;
+%      e_sym : forall x y, eqRel x y -> eqRel y x;
+%      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
+%    }.
+%Notation "f =e g" := (eqRel f g).
+%\end{coqcode}
+Given types \coq{A} and  \coq{B}, with their respective
+equivalence relations \coq{eA : A -> A -> Prop} and
+\coq{eB : B -> B -> Prop}, the we define the type
+\coq{A ~> B} to represent \emph{proper} morphisms of the
+respectfulness relation of \coq{R} to \coq{R'}.
+\begin{coqcode}
+Structure morph A {eA : setoid A} B {eB : setoid B} :=
+  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
+Notation "A ~> B" := (@morph A _ B _).
+\end{coqcode}
+We rely on Coq's type class mechanism to fill in the necessary equivalence
+relations. Coq's code extraction mechanism will erase any occurrence
+of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
+the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
+from \coq{A ~> B} to \coq{A -> B}.
+On top of this, we define basic function composition and identity functions:
+\begin{coqcode}
+Notation "f \o g" = (comp f g).
+Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
+Definition id : A ~> A := ...
+\end{coqcode}
+Using custom equivalences and proper morphisms, we redefine the definitions of
+container extensions and container equality. In particular, container
+extensions require a proper morphism to check the validity of positions in
+shapes, and container equality now uses equivalences of shapes and contained
+elements: 
+\begin{coqcode}
+Class Cont `{setoid Shape} (Pos : Type) := { valid : Shape * Pos ~> bool }.
+Inductive AppR `{F : Cont Sf Pf} `{setoid X} (x y : App F X) : Prop :=
+  | AppR_ext (Es : shape x =e shape y)
+             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
+\end{coqcode}
+Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
+however, we chose to use Coq's propositional equality, since these would leads
+to simpler code. We explain why in Section~\ref{sec:algebras}, and the
+mechanisation of initial algebras of container extensions.
+%The definitions in
+%Figure~\ref{fig:poly} are redefined to account for these changes in a
+%straightforward way, i.e. we prove that validity predicates are indeed proper
+%morphisms.
+
+This definition leads to the well-known ``setoid hell'', which we mitigate by
+providing tactics and notations to automatically discharge proofs of
+\coq{app_eq} for morphisms, whenever the types use the standard propositional
+equality, or a combination of propositional and extensional equality. However,
+our compositional approach allows us to build morphisms by plugging in other
+morphisms to our combinators. In our framework, our expectation is that the
+user-provided functions remain small, with relatively straightforward proofs of
+\coq{app_eq}.
+
+However, by using this mechanisation, we gain simplified proofs via the use of
+Coq's \emph{Generalised Rewriting}. Since every morphism \coq{f : A ~> B}
+satisfies the property that if \coq{x =e y}, then \coq{f x =e f y}, we can add
+every morphism as a proper element of Coq's respectfulness relation. In
+practice, this means that we can use the \coq{rewrite} tactic on proofs of
+type \coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
+arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},
+and \coq{transitivity} tactics. For example:
+\begin{coqcode}
+Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D) (H : h \o g =e id), 
+  h \o (g \o f) =e f.
+Proof.  rewrite compA, H.  reflexivity. Qed.
+\end{coqcode}
+
+\subsubsection{Polynomial Types}
+We define a number of equivalences for polynomial types.
+%These equivalences are
+%be used when building proper morphisms that manipulate their respective types.
+\begin{coqcode}
+Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
+Instance pair_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A * B).
+Instance sum_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A + B).
+Instance prop_eq : setoid Prop.
+Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
+\end{coqcode}
+Most of the definitions that involve functions and polynomial types are
+straightworward.  Identity and composition are defined as 
+\coq{fun x => x} and \coq{fun f g x => f (g x)} 
+respectively, and the proofs that they are proper morphisms is straightforward,
+and automatically discharged by Coq. 
+Products are built using function \coq{fun f g x => (f x, g x)},
+with the projections being the standard Coq  \coq{fst} and
+\coq{snd} functions. Similarly, sum injections are encoded using
+Coq's \coq{inl} and \coq{inr} constructors, and
+pattern matching on them uses the function 
+\coq{fun f g x => match x with | inl y => f y | inr y => g y end}.
+The proofs that these morphisms are proper are straightforward.  Finally, we
+also provide functions for currying/uncurrying, and  flipping the arguments of
+a proper morphism. We force most of our definitions to be inlined, to help
+Coq's code extraction mechanism to inline as many of these combinators as
+possible.
+
+We prove the isomorphisms of polynomial types and the equivalent 
+container extensions. As an example, we discuss (informally) the isomorphisms
+of pairs with their equivalent container extensions. Suppose that we
+know that \coq{App F X} is isomorphic to \coq{A}, and
+\coq{App G X} is isomorphic to \coq{B}. Then we can
+show that \coq{App (Prod F G) X} is isomorphic to
+\coq{A * B}.  If we have an element of type 
+\coq{App (Prod F G) X},
+using the \coq{inl} position, we can obtain
+\coq{App F X}.  Similarly, using \coq{inr}, we can
+obtain \coq{App G X}. Since these are the only two valid positions
+in the shape of pairs, we have finished. It is now sufficient to use the
+isomorphisms of \coq{App F X} and \coq{App G X} to
+obtain \coq{A * B}. Similarly if we have \coq{A * B},
+we can first use the isomorphisms of 
+\coq{A} and
+\coq{B} to obtain
+\coq{App F X * App G X}, and then construct the necessary container
+extension. Given \coq{p_inl : Pos l -> Pos (l * r)} (resp. \coq{p_inr}) that
+act as \coq{inl} (resp.\ \coq{inr}) on product positions, and
+\coq{case_pos : (Pos l -> X) -> (Pos r -> X) -> Pos (l * r) -> X} that pattern
+matches on the product positions, the functions that witness the isomorphism
+are:
+\begin{coqcode}
+Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
+  ({| shape := shape (fst x); cont := fun e => cont x (p_inl e) |}, 
+   {| shape := shape (snd x); cont := fun e => cont x (p_inr e) |}).
+Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
+{| shape := (shape (fst x), shape (snd x));
+   cont := case_pos (cont (fst x)) (cont (snd x)) |}.
+\end{coqcode}
+Proving that the composition of these functions is the identity is
+straightforward using the fact that \coq{Prod} containers only have two valid
+positions.
+
+\section{Formalising Recursion Schemes}
+\label{sec:recursion-schemes}
+[MOVED to background]
 \section{Extraction}

### 2024-10-07 paper/paper.tex
Introduction

@@
 \label{sec:intro}
-Recursive definitions cannot be proven well-defined automatically due to the
-halting problem. Modern proof assistants like Coq or Agda provide a sound, but
-incomplete algorithm which syntactically checks for termination or
-productivity. For recursion, this is done by automatically inferring
-which argument in the recursive call gets smaller with respects to the original
-input argument. For productivity, the algorithm checks that the corecursive
-call appears directly under a constructor to make sure that this function
-always produces at least one element after each recursive step.
-This implies that some functions, though well-defined, cannot be accepted by
-the proof assistant. Examples of this include common sorting algorithms, such
-as quicksort:
-%One example is Quicksort, written in OCaml
-%below:
-\begin{minted}{OCaml}
-let rec qsort xs = match divide xs with | None -> []
- | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
-\end{minted}
-While \ocaml{qsort} is a well-defined mathematical function it cannot be
-accepted by a proof assistant. The reason is that the \ocaml{divide}
-function \emph{destructuring} the input dives deeper in the input returning
-two sublists with the head as a pivot.
-
-The main approach for implementing non-structural recursion in Coq is to use
-\emph{well-founded recursion}, where recursive definitions are accompanied by
-\emph{termination proofs}. Using well-founded recursion, the recursive calls
-will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
-one approach is using the \coq{Fix} combinator, in which recursion is done on an
-\emph{accesibility} predicate on the input, which is a proof that there are no
-infinitely decreasing chains.}.
-A benefit of this strategy is that when extracting verified non-structural
-recursive functions to OCaml, the additional proofs are discarded, resulting in
-code that closely resembles what a programmer might have written directly in
-OCaml. However, reasoning about program equivalences needs  such termination
-proofs to be kept around, leading to the common practice of employing additional
-reduction lemmas extensively.
-
-% A benefit of this approach is that, extracting
-% verified nonstructural recursive functions to OCaml will erase the additional
-% proofs, thus producing code that will be closer to what a programmer may have
-% written directly in OCaml. However, reasoning about program equivalences
-% requires dealing with such termination proofs, and it is common practice to use
-% custom reduction lemmas that will be used extensively when proving properties
-% about nonstructural recursive functions.
-
-\emph{Structured recursion schemes}~\cite{HinzeW16, HinzeWG15} is a technique
-for structuring recursive definitions that has found a lot of success in
-functional programming. The idea is that, rather than writing a self-refential
-definition, we use a function library, the recursion scheme, which takes as
-input the non-recursive parts of the algorithm and returns the desired recursive
-definition which is guaranteed to terminate by the properties of the recursion
-scheme itself. More complex recursion schemes do exist but it has been
-proven by Hinze et al.~\cite{HinzeWG15} that they are all instances of a divide-and-conquer
-algorithm which goes under the name of hylomorphism~\cite{MeijerFP91,HuIT96}.
-This result, thus, yields a general framework to reason about the algebraic properties of these
-algorithms, e.g. short-cut fusion and semi-automatic parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}.
+Structured recursion schemes~\cite{HinzeW16, HinzeWG15} are powerful
+abstractions that capture common patterns of recursion. The main benefit of
+structuring computation using recursion schemes, is that they enjoy
+well-estabished \emph{algebraic properties} that can serve as a foundation for
+program optimisation (e.g.\ \emph{fusion} laws, or semi-automatic
+parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}).  These
+algebraic properties led to their use in the context of \emph{program
+calculation}, where programmers would describe their code using simple,
+inefficient specifications in an \emph{algebra of programming}~\cite{}, and then
+use the algebraic laws of this algebra of programming to \emph{calculate} an
+efficient version of the same algorithm.
+
+\dcas{TODO: Introduce example}
+
+Despite the rigorous algebraic laws that are satisfied by recursion schemes,
+there is a lack of tool support for their use in the context of program
+calculation. In fact, most of the work on applying program calculation is done
+by performing pen-and-paper proofs, and then translating the result to specific
+instances of recursion schemes, implemented in a programming language (generally
+Haskell)~\cite{TODO:INeedToFindSomeMoreReferences}. Most of the work on
+automating program calculation techniques either build implementations of the
+algebraic laws, together with some form of syntax or
+EDSL~\cite{TODO:INeedToFindSomeMoreReferences}. Such implementations are not
+verified, and can introduce bugs in the way they apply the algebraic laws as
+rewriting rules.
+
+A more rigorous approach to implementing program calculation techniques is to
+\emph{mechanise} them as part of a proof assistant. However, most of the work in
+mechanising recursion schemes focuses on a narrow subset of the known recursion
+schemes due to termination
+issues~\cite{TODO:FillInReferencesForCoqMechanisationsFromRelatedWorkSection}.
+Indeed, most of the mechanisations focus on \emph{maps} and \emph{folds}, and
+few authors focus on the \emph{unfolds}. In general, Hinze et
+al.~\cite{HinzeWG15} has proven that all known  recursion schemes are instances
+of a  divide-and-conquer algorithm which goes under the name of
+\emph{hylomorphism}~\cite{MeijerFP91,HuIT96}.  To date, no work has mechanised
+hylomorphisms in the Coq proof assistant.
+
+%Recursive definitions cannot be proven well-defined automatically due to the
+%halting problem. Modern proof assistants like Coq or Agda provide a sound, but
+%incomplete algorithm which syntactically checks for termination or
+%productivity. For recursion, this is done by automatically inferring
+%which argument in the recursive call gets smaller with respects to the original
+%input argument. For productivity, the algorithm checks that the corecursive
+%call appears directly under a constructor to make sure that this function
+%always produces at least one element after each recursive step.
+%This implies that some functions, though well-defined, cannot be accepted by
+%the proof assistant. Examples of this include common sorting algorithms, such
+%as quicksort:
+%%One example is Quicksort, written in OCaml
+%%below:
+%\begin{minted}{OCaml}
+%let rec qsort xs = match divide xs with | None -> []
+% | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
+%\end{minted}
+%While \ocaml{qsort} is a well-defined mathematical function it cannot be
+%accepted by a proof assistant. The reason is that the \ocaml{divide}
+%function \emph{destructuring} the input dives deeper in the input returning
+%two sublists with the head as a pivot.
+%
+%The main approach for implementing non-structural recursion in Coq is to use
+%\emph{well-founded recursion}, where recursive definitions are accompanied by
+%\emph{termination proofs}. Using well-founded recursion, the recursive calls
+%will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
+%one approach is using the \coq{Fix} combinator, in which recursion is done on an
+%\emph{accesibility} predicate on the input, which is a proof that there are no
+%infinitely decreasing chains.}.
+%A benefit of this strategy is that when extracting verified non-structural
+%recursive functions to OCaml, the additional proofs are discarded, resulting in
+%code that closely resembles what a programmer might have written directly in
+%OCaml. However, reasoning about program equivalences needs  such termination
+%proofs to be kept around, leading to the common practice of employing additional
+%reduction lemmas extensively.
+%
+%% A benefit of this approach is that, extracting
+%% verified nonstructural recursive functions to OCaml will erase the additional
+%% proofs, thus producing code that will be closer to what a programmer may have
+%% written directly in OCaml. However, reasoning about program equivalences
+%% requires dealing with such termination proofs, and it is common practice to use
+%% custom reduction lemmas that will be used extensively when proving properties
+%% about nonstructural recursive functions.
+
+%\emph{Structured recursion schemes}~\cite{HinzeW16, HinzeWG15} is a technique
+%for structuring recursive definitions that has found a lot of success in
+%functional programming. The idea is that, rather than writing a self-refential
+%definition, we use a function library, the recursion scheme, which takes as
+%input the non-recursive parts of the algorithm and returns the desired recursive
+%definition which is guaranteed to terminate by the properties of the recursion
+%scheme itself. More complex recursion schemes do exist but it has been
+%proven by Hinze et al.~\cite{HinzeWG15} that they are all instances of a divide-and-conquer
+%algorithm which goes under the name of hylomorphism~\cite{MeijerFP91,HuIT96}.
+%This result, thus, yields a general framework to reason about the algebraic properties of these
+%algorithms, e.g. short-cut fusion and semi-automatic parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}.
 
@@
 
-However, when used in the context of languages with general recursion like
-(e.g.) Haskell and ML, recursion schemes cannot ensure termination. This is
-because the input information itself could be non-terminating. On the other
-hand, while formalising recursion schemes in a type theory does provide stronger
-termination guarantees, to the best of our knowledge, not many attempts have
-been made at mechanising them.
+%However, when used in the context of languages with general recursion like
+%(e.g.) Haskell and ML, recursion schemes cannot ensure termination. This is
+%because the input information itself could be non-terminating. On the other
+%hand, while formalising recursion schemes in a type theory does provide stronger
+%termination guarantees, to the best of our knowledge, not many attempts have
+%been made at mechanising them.
 
@@
 not allow for extraction of idiomatic OCaml code. This is unfortunate since code
-extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
-Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, in Abreu et al.'s approach, extraction
-(1) do not preserve the recursive structure of common implementations; and (2)
-lead to unsafe casts like \ocaml{Obj.magic} in the generated code. This latter
-is also problematic in that, for higher-order programs, simple interoperations
-can lead to incorrect behaviour or even segfaults~\cite{forster:hal-04329663}
-and, moreover, it invalidates the fast-and-loose
-principle~\cite{DanielssonHJG06}.
+extration is what allows the execution of code that has been verified in
+Coq~\cite{OnoHTNH11, Larchey-Wendling23, MiculanP12, Sakaguchi20}.  In Abreu et
+al.'s approach, extraction (1) do not preserve the recursive structure of common
+implementations; and (2) lead to unsafe casts like \ocaml{Obj.magic} in the
+generated code. This latter is also problematic in that, for higher-order
+programs, simple interoperations can lead to incorrect behaviour or even
+segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates the
+fast-and-loose principle~\cite{DanielssonHJG06}.
 
@@
 mechanisation will allow the use of the algebraic laws of hylomorphisms for
-program reasoning, as well as extracting to idiomatic code (see
+\emph{program calculation}, as well as extracting to idiomatic code (see
 extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
@@
 To summarise, our contributions are as follows. In this paper we provide a
-framework for \emph{generic programming} with recursion schemes in Coq, with a
-proof of their algebraic laws and program equivalences for clean code
-extraction:
+framework for \emph{program calculation} with recursion schemes in Coq, where
+program calculation is done by applying regular Coq equalities, based on  proofs
+of the algebraic laws of hylomorphisms, and that allows the extraction of safe,
+idiomatic OCaml code:
 \begin{itemize}

### 2024-10-08 paper/paper.tex
minor
Stuff

@@
 program optimisation (e.g.\ \emph{fusion} laws, or semi-automatic
-parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}).  These
-algebraic properties led to their use in the context of \emph{program
-calculation}, where programmers would describe their code using simple,
-inefficient specifications in an \emph{algebra of programming}~\cite{}, and then
-use the algebraic laws of this algebra of programming to \emph{calculate} an
-efficient version of the same algorithm.
-
-\dcas{TODO: Introduce example}
-
-Despite the rigorous algebraic laws that are satisfied by recursion schemes,
-there is a lack of tool support for their use in the context of program
+parallelisations~\cite{TakanoM95,Gibbons96:Third, Morihata09:Third,
+farmsCastro}).  These algebraic properties led to their use in the context of
+\emph{program calculation}, where programmers would describe their code using
+simple, inefficient specifications in an \emph{algebra of
+programming}~\cite{BirddeMoor96:Algebra}, and then use the algebraic laws of
+this algebra of programming to \emph{calculate} an efficient version of the same
+algorithm. Suppose, for example, that we want to write a program sorts a list of
+integers, and then multiplies by 2 all its elements. In OCaml, we may write this
+function directly:
+\begin{minted}{ocaml}
+let rec sort_times_two = function
+  | [] -> []
+  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
+  sort_times_two l @ (h * 2) :: sort_times_two t
+\end{minted}
+Instead of writing this function directly, we may observe that we can derive it
+by \emph{fusing} a regular \emph{quickort} OCaml implementation, with
+\ocaml{map (fun x -> x * 2)}. This is the main idea behind program calculation:
+start
+from a simple specification, e.g.\
+$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
+equivalences and algebraic laws to rewrite it to an optimised version  (e.g.\
+the fused OCaml implementation above). We will revisit a similar example
+in Section~\ref{sec:sorting}.
+
+Despite the rigorous set of algebraic laws that are satisfied by recursion
+schemes, there is a lack of tool support for their use in the context of program
 calculation. In fact, most of the work on applying program calculation is done
@@
 instances of recursion schemes, implemented in a programming language (generally
-Haskell)~\cite{TODO:INeedToFindSomeMoreReferences}. Most of the work on
-automating program calculation techniques either build implementations of the
-algebraic laws, together with some form of syntax or
-EDSL~\cite{TODO:INeedToFindSomeMoreReferences}. Such implementations are not
-verified, and can introduce bugs in the way they apply the algebraic laws as
-rewriting rules.
-
-A more rigorous approach to implementing program calculation techniques is to
-\emph{mechanise} them as part of a proof assistant. However, most of the work in
-mechanising recursion schemes focuses on a narrow subset of the known recursion
-schemes due to termination
-issues~\cite{TODO:FillInReferencesForCoqMechanisationsFromRelatedWorkSection}.
-Indeed, most of the mechanisations focus on \emph{maps} and \emph{folds}, and
-few authors focus on the \emph{unfolds}. In general, Hinze et
-al.~\cite{HinzeWG15} has proven that all known  recursion schemes are instances
-of a  divide-and-conquer algorithm which goes under the name of
-\emph{hylomorphism}~\cite{MeijerFP91,HuIT96}.  To date, no work has mechanised
-hylomorphisms in the Coq proof assistant.
+Haskell). There are some examples of implementations of program calculation
+techniques, but these implementations are scarce, not up to date, and not
+verified. For example, Cunha et al. automated program calculation techniques by
+a custom Haskell implementation~\cite{DrHylo}.
+
+Several authors build tools for applying program calculation techniques by
+\emph{mechanising} them as part of a proof assistant. However, most of the work
+in mechanising recursion schemes focuses on a narrow subset of the known
+recursion schemes due to termination issues, or do not focus on proving
+algebraic laws of generic recursion schemes~\cite{10.1007/978-3-642-17796-5_10,
+MurataE19, larchey2022braga}.  Indeed, most of the mechanisations of generic
+recursion schemes focus on \emph{maps} and \emph{folds}, and few authors focus
+on the \emph{unfolds}. In general, Hinze et al.~\cite{HinzeWG15} has proven that
+all known  recursion schemes are instances of a  divide-and-conquer algorithm
+which goes under the name of \emph{hylomorphism}~\cite{MeijerFP91,HuIT96}.  To
+date, no work has mechanised hylomorphisms in the Coq proof assistant, together
+with their algebraic laws.
 
@@
 Coq~\cite{OnoHTNH11, Larchey-Wendling23, MiculanP12, Sakaguchi20}.  In Abreu et
-al.'s approach, extraction (1) do not preserve the recursive structure of common
-implementations; and (2) lead to unsafe casts like \ocaml{Obj.magic} in the
-generated code. This latter is also problematic in that, for higher-order
+al.'s approach, extraction (1) does not preserve the recursive structure of
+common implementations; and (2) leads to unsafe casts like \ocaml{Obj.magic} in
+the generated code. This latter is also problematic in that, for higher-order
 programs, simple interoperations can lead to incorrect behaviour or even
@@
 full mechanisation can be found on
-Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While
-programmers still need to reason about the termination of their programs, our
-mechanisation will allow the use of the algebraic laws of hylomorphisms for
-\emph{program calculation}, as well as extracting to idiomatic code (see
-extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
+Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}\dcas{I SHOULD
+ANONYMISE THIS}. While programmers still need to reason about the termination of
+their programs, our mechanisation will allow the use of the algebraic laws of
+hylomorphisms for \emph{program calculation}, as well as extracting to idiomatic
+code (see examples of extracted code in Section~\ref{sec:examples}).
 % As an example, our mechanisation of \ocaml{qsort} as a
@@
 \begin{itemize}
-  \item In Section~\ref{sec:containers} we formalise the type of container functors ensuring the presence of least and greatest fixed-points for functors and suitably adapted for program extraction
-  \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and hylomorhisms and their universal properties
-  \item In Section~\ref{sec:examples} we use the framework to formalise examples of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
-Furthermore we verify the short-cut fusion optimisation and show the extracted optimised code to OCaml.
+  \item In Section~\ref{sec:containers} we formalise the type of container
+functors ensuring the presence of least and greatest fixed-points for functors
+and suitably adapted for program extraction
+  \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and
+hylomorhisms and their universal properties
+  \item In Section~\ref{sec:examples} we use the framework to formalise examples
+of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
+Furthermore we verify the short-cut fusion optimisation and show the extracted
+optimised code to OCaml.
 \end{itemize}
@@
 \end{equation}
-As we stated earlier, a solution to this equation does not exist for an arbitrary  algebra and coalgebra pair and, in fact, this definition cannot be accepted by Coq.
+As we stated earlier, a solution to this equation does not exist for an
+arbitrary  algebra and coalgebra pair and, in fact, this definition cannot be
+accepted by Coq.
 

### 2024-10-08 paper/paper.tex
pass over related work section

@@
 Encoding recursion schemes in Coq is not new. We compare our work with other
-encodings of program calculation techniques in Coq (Section~\ref{sec:calccoq}),
-recursion schemes in Coq (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}).
-Finally, we compare our encoding of recursion schemes to other forms of
-termination checking (Section~\ref{sec:termcheck}, and sized types
-(Section~\ref{sec:sizedT}), as a way to guarantee termination of nonstructural
+encodings of program calculation techniques in Coq and recursion schemes in Coq
+(Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}). Additionally, we compare our
+encoding of recursion schemes to other forms of termination checking
+(Section~\ref{sec:termcheck}, as a way to guarantee termination of nonstructural
 recursion.
@@
 Within the domain of program optimization, program calculation serves as a
-well-established programming technique aimed at deriving efficient programs
-from their naive counterparts through systematic program
-transformation~\cite{gibbonsSquiggol}. This area has been extensively explored
-over the years. Tesson et al.\ demonstrated the efficacy of leveraging Coq to
-establish an approach for implementing a robust system dedicated to verifying
-the correctness of program transformations for functions that manipulate
+well-established programming technique aimed at deriving efficient programs from
+their naive counterparts through systematic program
+transformation~\cite{gibbonsSquiggol}. Programmers may specify algorithms
+abstractly at a high level, and then derive an efficient implementation using
+the process of program calculation~\cite{BirddeMoor96:Algebra}. This area has
+been extensively explored over the years.
+
+Tesson et al.\ demonstrated the efficacy of leveraging Coq to establish an
+approach for implementing a robust system dedicated to verifying the correctness
+of program transformations for functions that manipulate
 lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
-formalised recursion schemes in Coq~\cite{MurataE19}. Their development does
-not include hylomorphisms and dynamorphisms, and relies on the functional
+formalised recursion schemes in Coq~\cite{MurataE19}. Their development does not
+include hylomorphisms and dynamorphisms, and relies on the functional
 extensionality axiom, as well as further extensionality axioms for each
-coinductive datatype that they use. They do not discuss the extracted OCaml
-code from their formalisation.  Larchey-Wendling and Monin encode recursion
-schemes in Coq, by formalising computational graphs of
-algorithms~\cite{larchey2022braga}. Their work does not focus on encoding
-generic recursion schemes, and proving their algebraic laws.  Castro-Perez et
-al.~\cite{farmsCastro} encode the laws of hylomorphisms as part of the type
-system of a functional language to calculate parallel programs from
-specifications. Their work focuses on parallelism, and they do not formalise
-their approach in a proof assistant, and the laws of hylomorphisms are axioms
-in their system.
+coinductive datatype that they use. They do not discuss the extracted OCaml code
+from their formalisation. Larchey-Wendling and Monin encode recursion schemes in
+Coq, by formalising computational graphs of algorithms~\cite{larchey2022braga}.
+Their work does not focus on encoding generic recursion schemes, and proving
+their algebraic laws. Castro-Perez et al.~\cite{farmsCastro} encode the laws of
+hylomorphisms as part of the type system of a functional language to calculate
+parallel programs from specifications. Their work focuses on parallelism, and
+they do not formalise their approach in a proof assistant, and the laws of
+hylomorphisms are axioms in their system.
 
 %\mvol{Need more general citations for program calculation.}
+\subsection{Divide and Conquer Recursion}\label{sec:reccoq}
+
+Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
+using a recursion scheme in which termination is entirely enforced by its
+typing.  This is a significant advance, since it avoids \emph{completely} the
+need for termination proofs. Their work differs from ours in that they require
+the functional extensionality axiom, and the use of impredicative \coq{Set}.
+The authors justify well the use of impredicative \coq{Set} and its
+compatibility with the functional extensionality axiom. In contrast, our
+development remains entirely \emph{axiom-free}. Another key difference with our
+approach is that they do not discuss what the resulting extracted code
+\emph{looks like} (that is, whether the extracted OCaml code resembles the
+natural formulation of the recursive function). Through experiments, we found
+that their formalisation leads to \ocaml{Obj.magic}, and code with a complex
+structure that may be hard to understand or interface with other OCaml code.
+Due to the great benefit of entirely avoiding termination proofs, it would be
+interesting to extend their approach to improve code extraction.
 
@@
 definitions will not be accepted by such checks.
+
+\mvol{Do we need to discuss this or can we cut it?}
 
@@
 
-\subsubsection{Divide and Conquer Recursion}\label{sec:reccoq}
-
-Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
-using a recursion scheme in which termination is entirely enforced by its
-typing.  This is a significant advance, since it avoids \emph{completely} the
-need for termination proofs. Their work differs from ours in that they require
-the functional extensionality axiom, and the use of impredicative \coq{Set}.
-The authors justify well the use of impredicative \coq{Set} and its
-compatibility with the functional extensionality axiom. In contrast, our
-development remains entirely \emph{axiom-free}. Another key difference with our
-approach is that they do not discuss what the resulting extracted code
-\emph{looks like} (that is, whether the extracted OCaml code resembles the
-natural formulation of the recursive function). Through experiments, we found
-that their formalisation leads to \ocaml{Obj.magic}, and code with a complex
-structure that may be hard to understand or interface with other OCaml code.
-Due to the great benefit of entirely avoiding termination proofs, it would be
-interesting to extend their approach to improve code extraction.
-
-\subsubsection{Sized Types}\label{sec:sizedT}
-
-Another approach to certifying termination of recursive functions is the use of
-\emph{sized types}. Sized types were introduced by Hughes et al.\ as a way to
-track/verify various properties of recursive programs, including productivity
-and termination~\cite{HughesPS96}. The core idea of sized types is that types
-express bounds on the sizes of recursive data structures. With this approach,
-algorithms are implemented in the standard functional way, following the
-divide-and-conquer pattern of splitting, recurring, and merging, with the
-addition of the data types being indexed by a static approximation of the
-relative \emph{size}. An advantage of this approach is that it allows for code
-to be written in a natural way. There are costs, however. Programs must be
-written in a way that accounts for the handling of size indexes, and support for
-the size types must be added to the language.
-
-This approach has been used to express nonstructurally recursive
-algorithms in Agda: Copello et al.\ used sized types in a straightforward
-formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
-in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
-Coq as well, though it has not yet been adopted~\cite{ChanLB23}.
-
 \section{Conclusions and Future Work}
@@
 alternative formulations to try to improve the performance of the rewriting
-tactics (e.g.  canonical structures).  Currently, Coq is unable to inline a
+tactics (e.g.\ canonical structures).  Currently, Coq is unable to inline a
 number of trivially inlineable definitions.  We will study alternative

### 2024-10-08 paper/paper.tex
ESOP

@@
-\documentclass[ a4paper, UKenglish, cleveref, autoref, thm-restate]{lipics-v2021}
+\documentclass{llncs}
+% \documentclass[llncs, a4paper, UKenglish, cleveref, autoref, thm-restate]{article}
 \usepackage[utf8]{inputenc} % Required for inputting international characters
@@
 % \addbibresource{references.bib}
-\usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
+% \usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
 \usepackage{caption}
@@
 \usepackage{enumitem}
-\usepackage{mathpartir}
-\usepackage{bussproofs}
+% \usepackage{mathpartir}
+% \usepackage{bussproofs}
 \usepackage{xparse}
@@
 \newcommand{\mvol}[1]{\textcolor{blue}{\textsc{Michael}: #1}}
-\newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
-\newenvironment{proofof}[1] {\begin{proof}[Proof of {#1}]}{\end{proof}}
+% \newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
+% \newenvironment{proofof}[1] {\begin{proof}[Proof of {#1}]}{\end{proof}}
 \newcommand{\eqdef}{\stackrel{\mathrm{\Delta}}{=}}
@@
 
-\author{David Castro-Perez}{School of Computing, University of Kent}{d-castro-perez@kent.ac.uk}{}{}
-\author{Marco Paviotti}{School of Computing, University of Kent}{m.paviotti@kent.ac.uk}{}{}
-\author{Michael Vollmer}{School of Computing, University of Kent}{m.vollmer@kent.ac.uk}{}{}
-
-%\author{Jane {Open Access}}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
-Please provide at least the name of the affiliation and the country.
-The full address is optional.
-Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.
-
-%\author{Joan R.
-Public\footnote{Optional footnote, e.g.
-to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
-
-\authorrunning{D. Castro-Perez, M. Paviotti, and M. Vollmer} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{D. Castro-Perez, M. Paviotti, and M. Vollmer} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
-
-\keywords{hylomorphisms, Coq, code extraction, verification} %TODO mandatory; please add comma-separated list of keywords
-
-\category{} %optional, e.g. invited paper
-
-\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
-%\relatedversiondetails[linktext={opt.
-text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g.
-Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional
-
-%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-%\supplementdetails[linktext={opt.
-text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
-Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-%\acknowledgements{I want to thank \dots}%optional
-
-%\nolinenumbers %uncomment to disable line numbering
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \author{David Castro-Perez}{School of Computing, University of Kent}{d-castro-perez@kent.ac.uk}{}{}
+% \author{Marco Paviotti}{School of Computing, University of Kent}{m.paviotti@kent.ac.uk}{}{}
+% \author{Michael Vollmer}{School of Computing, University of Kent}{m.vollmer@kent.ac.uk}{}{}
 

### 2024-10-08 paper/paper.tex
Elements of Category Theory

@@
 % \addbibresource{references.bib}
-% \usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
+\pagestyle{plain}
+
+\usepackage{amsmath, mathpartir}
+%, amssymb, amsfonts, stmaryrd}
+% \usepackage{mathpartir}
+% \usepackage{bussproofs}
+
 \usepackage{caption}
@@
 \usepackage{enumitem}
-% \usepackage{mathpartir}
-% \usepackage{bussproofs}
 \usepackage{xparse}
@@
 %\newcommand{\coq}[1]{\mintinline{coq}{#1}}
+
+\newcommand{\Hom}{\text{Hom}}
+\newcommand{\Obj}{\text{Obj}}
+\newcommand{\Arr}{\text{Arr}}
+
+\newcommand{\List}[1]{\operator{List}(#1)}
+\newcommand{\nil}{\operator{nil}}
+\newcommand{\cons}{\operator{cons}}
+
 \newmintinline[coq]{coq}{fontsize=\small}
@@
 \section{Recursion Schemes}
-\subsection{Functors}
-Functors are functions \coq{F : Type -> Type} which additionally have a
-map
-\begin{coqcode}
-fmap : forall A B, (A -> B) -> F A -> F B
-\end{coqcode}
-witnessing the idea that a functor represents a container for abstract data that
-can be manipulated without changing the outer structure. For example, the type
-\coq{List : Type -> Type} is a functor and \coq{List A} is the type of lists
-over elements of type \coq{A} with the obvious \coq{fmap} function recursively
-traversing a list and applying the function \coq{A -> B} for each element.
-Additionally, the type of lists \coq{List A} arises as the least fixed-point of
-the functor \coq{F X = unit + (A * X) }.
-
-In general not every functor has a fixed-point, and it is not possible to build
-the fixed-point of a functor such as \coq{F} in Coq due to not being strictly
-positive. Due to this, we are going to restrict to \emph{polynomial functors}
-as represented by containers.
-\subsection{Algebras and Coalgebras}
-\subsection{Hylomorphisms}
-Recursion schemes provide an abstract way to consume and generate data. We now
-proceed onto describing how to formalise hylomorphisms in Coq. We first
-formalise algebras for container extensions (Section~\ref{sec:algebras}), then
-we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together these
-notions to formalise recursive coalgebras and hylomorphisms
-(Section~\ref{sec:rec-coalgebras}).
+The structure of the a certain data is very similar to the structure of an
+algorithm which processes that data. Recursion schemes aim at analyzing these
+analogies.
+
+\subsection{Elements of Category Theory}
+A \emph{category}, denoted with $\CatC, \CatD, \CatE$,  is a collection of
+objects $A, B, C$, denoted by $\Obj(\CatC)$ and a collection of arrows $f, g, h$
+between these objects, denoted by $\Arr{\CatC}$, such that there always exists
+an identity arrow $id_{A} : A \to A$ for each object $A$ and for two arrows
+$A \xrightarrow{f} B$ and $B \xrightarrow{g} C$ there always exists an arrow
+$A \xrightarrow{g\circ f} C$ obeying the associativity law. We denote
+$\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to $B$\footnote[1]{For
+presentation purposes we shall not deal with size issues and assume all the
+categories are small.}. The initial object in a category, denoted by $0$, is the
+object such that for any other object $A$ there is a unique arrow
+$0 \xrightarrow{!} A$.  Dually, the terminal object, denoted by $1$, is the
+object such that for any other object $A$ there is a unique arrow
+$A \xrightarrow{!} 1$. Initial and terminal object are unique up-to isomorphism.
+
+For example, the category of sets, denoted by $\Set$, is the category where
+objects are sets and arrows are functions between sets. The initial object $0$
+in $\Set$ is the empty set $\emptyset$ and the terminal object $1$ is any
+singleton set. In this section, we are going to assume types are sets, hence
+this category can also be viewed as the category of types and programs between
+them.
+
+A \emph{functor} $F : \CatC \to \CatD$ is a map between categories mapping both
+objects and arrows from one category to another. Hence a functor has two
+components, one which maps objects into objects
+$F : \Obj(\CatC) \to \Obj(\CatD)$ and one which maps arrows into arrows
+$F : \Hom_{\CatC}(A, B) \to \Hom_{\CatD} (FA,FB)$ such that identity and
+composition of arrows are preserved:
+\begin{mathpar}
+  F(id_A) = id_{FA} \and F(g \circ f) = F(g) \circ F(f)
+\end{mathpar}
+This latter component is also called the \emph{functorial action} and can be
+thought of as the $\operator{fmap}$ program in functional programming.
+
+For example, in $\Set$, we can define the set of lists
+$\List{A} \iso 1 + A \times \List{A}$ as the set inductively generated by the
+constructors $\nil : 1 \to \List{A}$ and
+$\cons : A \times \List{A} \to \List{A}$. This type is a functor with functorial
+action $\List{f} : \List{A} \to \List{B}$ given by $\List{f}(*) = *$ and
+$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
+$\List{f}$ is well-defined as it recursively calls on a smaller argument,
+however, there is a more general way of proving these type of definitions are
+well-defined.
 
 \subsection{Algebras and Catamorphisms}
-\label{sec:algebras} An algebra is a set \coq{A} (the \emph{carrier} of the
-algebra) together with some operations on it sometimes subject to some
-equational laws. For example, given such a type \coq{A} we can define the monoid
-operations as \coq{u : unit -> A} for unit and \coq{m : A * A -> A} for
-multiplication. Notice that \coq{F X = unit + X * X} is a functor on \coq{X}
-which means we can equivalently describe these two maps as a single one of type
-\coq{F X -> X} which we call the algebra for the functor \coq{F} satisfying the
-unit and associative laws of the monoid. In this work we will be using algebras
-to describe the operations of a data type and so we will not be needing
-additional equations on these operations.
+\label{sec:algebras}
+For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(A,\alpha)$
+where $A$ is an object of the category called the \emph{carrier} of the algebra
+and $\alpha$ is an arrow of type $FA \to A$ called the \emph{structure map}.
+The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
+objects are $F$-algebras and arrows $f : (A, \alpha) \to (B, \beta)$ in
+$F\Alg(\CatC)$ are arrows $f : A \to B$ in $\CatC$ such that they respect the
+structure of the algebra, that is $f \circ \alpha = \beta \circ F(f)$.  The
+initial object in this category is called the \emph{initial $F$-algebra}, that
+is the $F$-algebra which has a unique $F$-algebra homomorphism into any other
+$F$-algebra. If $(A, \alpha)$ is initial this implies $A$ is the \emph{least
+fixed-point} for the functor $F$, i.e. $A \iso FA$ is the least solution of the
+domain equation $X \iso F X$ by Lambek's lemma.
+
+For example, for a set $A$ and a given the functor $F : \Set \to \Set$ mapping
+$X \mapsto 1 + A \times X$ an algebra is a set $X$ together with an arrow
+$\alpha : 1 + A \times X \to X$.in $\Set$ given such a type
+\coq{A} we can define the monoid operations as \coq{u : unit -> A} for unit and
+\coq{m : A * A -> A} for multiplication. Notice that \coq{F X = unit + X * X} is
+a functor on \coq{X} which means we can equivalently describe these two maps as
+a single one of type \coq{F X -> X} which we call the algebra for the functor
+\coq{F} satisfying the unit and associative laws of the monoid. In this work we
+will be using algebras to describe the operations of a data type and so we will
+not be needing additional equations on these operations.
 
@@
 same coalgebra.
+
+\subsection{Hylomorphisms}
+Recursion schemes provide an abstract way to consume and generate data. We now
+proceed onto describing how to formalise hylomorphisms in Coq. We first
+formalise algebras for container extensions (Section~\ref{sec:algebras}), then
+we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together these
+notions to formalise recursive coalgebras and hylomorphisms
+(Section~\ref{sec:rec-coalgebras}).
 

### 2024-10-08 paper/paper.tex
Algebras and Catamorphisms

@@
 \newcommand{\cons}{\operator{cons}}
+\newcommand{\foldr}{\operator{foldr}}
 
@@
 \subsection{Elements of Category Theory}
-A \emph{category}, denoted with $\CatC, \CatD, \CatE$,  is a collection of
-objects $A, B, C$, denoted by $\Obj(\CatC)$ and a collection of arrows $f, g, h$
-between these objects, denoted by $\Arr{\CatC}$, such that there always exists
-an identity arrow $id_{A} : A \to A$ for each object $A$ and for two arrows
-$A \xrightarrow{f} B$ and $B \xrightarrow{g} C$ there always exists an arrow
+A \emph{category} is a collection of objects $A, B, C$, denoted by $\Obj(\CatC)$
+and a collection of arrows $f, g, h$ between these objects, denoted by
+$\Arr{\CatC}$, such that there always exists an identity arrow
+$id_{A} : A \to A$ for each object $A$ and for two arrows $A \xrightarrow{f} B$
+and $B \xrightarrow{g} C$ there always exists an arrow
 $A \xrightarrow{g\circ f} C$ obeying the associativity law. We denote
-$\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to $B$\footnote[1]{For
-presentation purposes we shall not deal with size issues and assume all the
-categories are small.}. The initial object in a category, denoted by $0$, is the
-object such that for any other object $A$ there is a unique arrow
-$0 \xrightarrow{!} A$.  Dually, the terminal object, denoted by $1$, is the
-object such that for any other object $A$ there is a unique arrow
-$A \xrightarrow{!} 1$. Initial and terminal object are unique up-to isomorphism.
+$\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to $B$ and we use the letters
+$\CatC$, $\CatD$, $\CatE\dots$ for categories\footnote[1]{For presentation
+purposes we shall not deal with size issues and assume all the categories are
+small.}. The initial object in a category, denoted by $0$, is the object such
+that for any other object $A$ there is a unique arrow $0 \xrightarrow{!} A$.
+Dually, the terminal object, denoted by $1$, is the object such that for any
+other object $A$ there is a unique arrow $A \xrightarrow{!} 1$. Initial and
+terminal object are unique up-to isomorphism.
 
@@
 The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
-objects are $F$-algebras and arrows $f : (A, \alpha) \to (B, \beta)$ in
-$F\Alg(\CatC)$ are arrows $f : A \to B$ in $\CatC$ such that they respect the
-structure of the algebra, that is $f \circ \alpha = \beta \circ F(f)$.  The
-initial object in this category is called the \emph{initial $F$-algebra}, that
-is the $F$-algebra which has a unique $F$-algebra homomorphism into any other
-$F$-algebra. If $(A, \alpha)$ is initial this implies $A$ is the \emph{least
-fixed-point} for the functor $F$, i.e. $A \iso FA$ is the least solution of the
-domain equation $X \iso F X$ by Lambek's lemma.
-
-For example, for a set $A$ and a given the functor $F : \Set \to \Set$ mapping
-$X \mapsto 1 + A \times X$ an algebra is a set $X$ together with an arrow
-$\alpha : 1 + A \times X \to X$.in $\Set$ given such a type
-\coq{A} we can define the monoid operations as \coq{u : unit -> A} for unit and
-\coq{m : A * A -> A} for multiplication. Notice that \coq{F X = unit + X * X} is
-a functor on \coq{X} which means we can equivalently describe these two maps as
-a single one of type \coq{F X -> X} which we call the algebra for the functor
-\coq{F} satisfying the unit and associative laws of the monoid. In this work we
-will be using algebras to describe the operations of a data type and so we will
-not be needing additional equations on these operations.
-
+objects are $F$-algebras and arrows $f : (A, \alpha) \to (B, \beta)$ are
+\emph{$F$-algebra homomorphisms}  $f : A \to B$ in $\CatC$ such that they
+respect the structure of the algebra, that is
+$f \circ \alpha = \beta \circ F(f)$. The initial object in this category is
+called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
+unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
+the initial $F$-algebra is precisely the \emph{least fixed-point} for the
+functor $F$ which we denote by $(\mu F, \InIso)$ with
+$\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
+isomorphism $F\mu F \iso \mu F$. By the property of initial $F$-algebras, for
+any other $F$-algebra $(B, \beta)$ there exists a unique $F$-algebra
+homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (B, \beta)$ and
+pronounced ``\emph{catamorphism}''.
+
+For example, for a set $A$ we define the functor $F : \Set \to \Set$ mapping
+$X \mapsto 1 + A \times X$. An $F$-algebra is a set $B$ together with a
+structure map $[base, ind] : 1 + A \times B \to B$. The initial $F$-algebra is
+clearly the set of lists $\List{A}$. The catamorphism associated with the type
+of lists is the unique arrow recursively translating the initial algebra
+$[\nil,\cons]$ into the algebra $[base, ind]$:
+\begin{align*}
+  & \cata{[base,ind]}(\nil) = base\\
+  & \cata{[base, ind]}(\cons(a, xs)) = ind(a, \cata{[base, ind]}(xs))
+\end{align*}
+In functional programming this is commonly referred to as
+$\foldr : (1 \to B) \to (A \times B \to B) \to B$. We can in fact set
+$\foldr\;base\;ind = \cata{[base, ind]}$.
+
+\subsection{Dump}
 Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given

### 2024-10-09 paper/paper.tex
Algebras and Coalgebras

@@
 \newcommand{\Alg}{\text{-Alg}}
+\newcommand{\Coalg}{\text{-CoAlg}}
 \newcommand{\Free}{\text{Free\xspace}}
@@
 
+\newcommand{\Str}[1]{\operator{Str}(#1)}
 \newcommand{\List}[1]{\operator{List}(#1)}
@@
 \label{sec:intro}
-Structured recursion schemes~\cite{HinzeW16, HinzeWG15} are powerful
+Structured recursion schemes~\cite{HinzeW16,HinzeWG15} are powerful
 abstractions that capture common patterns of recursion. The main benefit of
@@
 program optimisation (e.g.\ \emph{fusion} laws, or semi-automatic
-parallelisations~\cite{TakanoM95,Gibbons96:Third, Morihata09:Third,
-farmsCastro}).  These algebraic properties led to their use in the context of
+parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
+These algebraic properties led to their use in the context of
 \emph{program calculation}, where programmers would describe their code using
@@
 recursion schemes due to termination issues, or do not focus on proving
-algebraic laws of generic recursion schemes~\cite{10.1007/978-3-642-17796-5_10,
-MurataE19, larchey2022braga}.  Indeed, most of the mechanisations of generic
+algebraic laws of generic recursion schemes~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.
+Indeed, most of the mechanisations of generic
 recursion schemes focus on \emph{maps} and \emph{folds}, and few authors focus
@@
 extration is what allows the execution of code that has been verified in
-Coq~\cite{OnoHTNH11, Larchey-Wendling23, MiculanP12, Sakaguchi20}.  In Abreu et
+Coq~\cite{OnoHTNH11,Larchey-Wendling23,MiculanP12,Sakaguchi20}.  In Abreu et
 al.'s approach, extraction (1) does not preserve the recursive structure of
@@
 
-% \section{Background on Recursion Schemes}
-% Recursion schemes provide a systematic way to work with data types, such as
-% trees or lists, by abstracting away the recursion into higher-order functions.
-% Canonical recursion schemes include \emph{folds} (catamorphisms) which consume
-% data and \emph{unfolds} (anamorphisms) which produce data. These are the unique
-% solutions (when they exist) to the following two equations respectively
-% \[
-%   h  = a \comp F h \comp \InOp \qquad k = \OutOp \comp F k \comp c
-% \]
-% where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
-% can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
-% $F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
-% $F$-coalgebra), $\InOp$ is the isomoprhism on the least fixed-point of $F$,
-% namely $\InOp : \mu F \to F\mu F$, and $\OutOp$ is the isomorphism on
-% the greatest fixed-point of $F$, namely $F \nu F \to \nu F$. When
-% the unique solutions to the above equations exist we denote them as $\cata{a}$
-% and $\ana{c}$ respectively.
-
-% \subsection{Hylomorphisms}
-% Catamorphisms (dually anamorphisms) do not capture every recursion
-% scheme we can think of.  As an example, consider an example of a classic
-% \emph{divide-and-conquer} program such as quicksort algorithm:\footnote[1]{This
-% program is written in Coq notation to be consistent, but it is not a program
-% that Coq can accept.}
-
-% \mpav{I need this to be written in Coq notation}
-% \begin{coqcode}
-% qs :: [a] -> [a]
-% qs [] = []
-% qs (x:xs) = qs (filter (x <=) xs) ++ [x] ++ (qs (filter (x >) xs))
-% \end{coqcode}
-% This program is formed of essentially two parts. The first one is what is
-% commonly thought as the \emph{partition} function destructuring the input list
-% in three parts, the pivot and the other two sublists, and the second part
-% recursing over the sublists and composing the results back together.
-
-% Notice that if we define the functor \coq{TreeF} $X = X \times A \times X$ with
-% the obvious functorial action then \coq{partition} can be viewed as a
-% \coq{TreeF}-coalgebra on lists over the type $A$
-% \begin{coqcode}
-% c : [A] -> TreeF [A]
-% c (x:xs) = (filter (x <=) xs, x, filter (x >) xs)
-% \end{coqcode}
-
-% While we can define an \coq{TreeF}-algebra on lists
-% over $A$ as well, that is a map
-% \begin{coqcode}
-% a : TreeF [A] -> [A]
-% a (left, x, right) = left ++ [x] ++ right
-% \end{coqcode}
-
-% It should be easy to see that the program \coq{qs} can now be defined as the
-% unique program that runs the coalgebra $c$ recurses over the sublists produced
-% and then composes the results together. The structure exemplified by this
-% program is a typical divide-and-conquer algorithm which can be implemented using
-% an \emph{hylomorphism}. An hylomorphism is the unique solution $x : C \to A$ to
-% the hylo equation
-% \begin{equation}
-%   \label{eq:hylo}
-%   x =  a \comp Fx \comp c
-% \end{equation}
-% for a $F$-algebra on $A$, an $F$-coalgebra on $C$ and a functor $F$.  If for
-% every $F$-algebra there exists a unique solution to (\ref{eq:hylo}) then we say
-% the $F$-coalgebra $c$ is \emph{recursive}. Dually if for every $F$-coalgebra $c$
-% there exists a unique solution to (\ref{eq:hylo}) we say the $F$-algebra $a$ is
-% \emph{corecursive}. We denote the unique solution to the hylo equation
-% (\ref{eq:hylo}) as $\hylo{a}{c}$.
-
-% Moreover, it shoud be easy to see that catamorphisms and anamorphism arise as a
-% special case of hylomorphisms
-% \begin{equation}
-%   \label{eq:cata}
-%   \cata{a} = \hylo{\InOp}{a} \qquad \ana{c} = \hylo{c}{\OutOp}
-% \end{equation}
-% since $\InOp$ is indeed a recursive coalgebra and $\OutOp$ is a corecursive
-% algebra.
-
-% \subsection{Implementing Recursion Schemes}
-% Hylomorphisms can implement all structured recursion schemes~\cite{HinzeWG15,
-% HinzeW16}.  Here we show the result by means of a very basic example: mutual
-% recursion. We refer the reader to Hinze et al.~\cite{HinzeWG15} for the
-% theoretical result and to Yang and Wu~\cite{YangW22} for more examples of
-% recursion schemes.
-
-% % More specifically, in the context of recursive data types, such as in Haskell,
-% % where induction and coinduction is conflated, an hylomorphism can be implemented
-% % in terms of a catarmorphism (folding) and an anamorphism (unfolding)
-% % \[
-% %   \hylo{a}{c} = \cata{a} \comp \ana{c}
-% % \]
-% % They first generate a data structure using an anamorphism and then immediately
-% % fold or reduce it using a catamorphism.
-
-% % The hylomorphism pattern is particularly useful for defining recursive
-% % algorithms in a functional programming context, where recursion is a fundamental
-% % tool for solving problems. By separating the generation of data structures from
-% % their consumption, hylomorphisms enable clearer and more modular code, making it
-% % easier to reason about and maintain recursive algorithms.
-
-% % However, in a type theory inductive and coinductive data types are separate.
-
-% % In this section we will introduce the notion of hylomorphism. By way of examples
-% % we will also give high level intution as to way all recursion schemes can be
-% % defined in terms of a basic
-
-% Say that we want to prove that the two mutually recursive functions below are
-% well-defined
-
-% \begin{tabular}{LL}
-%   \operator{even} : \Nat \to \Bool                  &  \operator{odd} : \Nat \to \Bool\\
-%   \operator{even} (0) = \top                        &  \operator{odd}(0) = \bot\\
-%   \operator{even} (n+1) = \neg \operator{odd}(n)    &  \operator{odd}(n+1) = \neg (\operator{even}(n))
-% \end{tabular}
-
-% One option is to prove that the following function is
-% equivalent to the above
-% \begin{align*}
-%   & \operator{even-odd} : \Nat \to \Bool \times \Bool\\
-%   & \operator{even-odd} (0) = (\top, \bot)\\
-%   & \operator{even-odd} (n+1) = \operator{let } (x,y) \leftarrow \operator{even-odd}(n) \operator{ in } (\neg x, \neg y)
-% \end{align*}
-% which is clearly a catamorphism $\cata{a}$ with $a (\Nil) = (\top, \bot)$ and
-% $a(\operator{x,y}) = (\neg x, \neg y)$ and hence an hylomorphism as defined in
-% (\ref{eq:cata}).  Now we are left to show they are equivalent. Well, it is easy
-% to do so since $\operator{even} = \pi_{1} \comp \operator{even-odd}$
-% $\operator{odd} = \pi_{2} \comp \operator{even-odd}$. The proof that these two
-% pairs of functions are the same is an easy induction over the natural numbers.
-
-% This is an instance of a more general phenonema whose argument uses the concept
-% of an adjunction~\cite{MacLane71}.  For this particular example, the argument is
-% as follows.  There is an isomorphism between pairs of maps
-% $(A \to B_{1}, A \to B_{2})$ and maps $(A \to B_{1} \times B_{2})$.  This is
-% given by $\Phi (f,g) = \pair{f}{g}$ and
-% $\Psi(f) = (\pi_{1} \comp f, \pi_{2} \comp f)$ where the projections $\pi$ are
-% part of the counit of the adjunction between the diagonal functor
-% $\Delta X = (X,X)$ and the product $\times$.  Once this correspondence is
-% established we can implement the map $A \to B_{1} \times B_{2}$ as an
-% hylomorphism and derive the original and more complex recursion scheme.
-
-% % In general an adjunction is a natural correspondence of maps
-% % \[
-% %   LA \to B \iff A \to RB
-% % \]
-% % where $L$ and $R$ are called adjoints.
-
-% However, formalising adjunctions in Coq would require to formalise a
-% non-negligible amount of category theory which goes beyond the scope of the
-% present work. In this paper, we will instead derive the recursion schemes as
-% needed and implement them as hylos.
-% % \subsubsection{Course-of-Value Recursion}
-% % The knapsack algorithm is defined as following in pseudo-Haskell code:
-% % \begin{align*}
-% %   & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
-% %   & \operator{knapsack}\;wvs\; c = \operator{maximum}_{0}
-% %     [v + \operator{knapsack}\;wvs\; (c - w) \mid (w,v) \leftarrow wvs, w \le c]
-% % \end{align*}
-% % %
-% % with dynamic programming knapsack can be rephrased like this
-% % \begin{align*}
-% %   & \operator{knapsack} : [(\Nat, \R)] \to \Nat \to \R\\
-% %   & \operator{knapsack}\;wvs\;c= table !! c \\
-% %   & \qquad table  = [ks\;i \mid i \leftarrow [0..c]]\\
-% %   & \qquad ks\; i = maximum_{0} [v + table!! (i - w) \mid (w,v) \leftarrow wvs, w \le i]
-% % \end{align*}
-
-% % The recursion scheme is justified by solving a recursive equation.  First we
-% % define an operation to create a memory table representing the list of capacities
-% % $[0..c]$. We first define a functor $D X = 1 + X$ representing the base functor
-% % for constructing the naturals $\Nat = \mu D$ the we define the generating
-% % function by induction on the naturals
-% % \begin{align*}
-% %   & \operator{fan} : \Nat \to D_{\infty}\Nat\\
-% %   & \operator{fan} (0)    = \pair{0}{\inl}\\
-% %   & \operator{fan} (n+1)  = \pair{n+1}{\inr (\operator{fan}(n))}
-% % \end{align*}
-% % The recursion scheme from a comonad is then defined as follows
-% % \begin{align*}
-% %   & \operator{rsfc} : \Nat \to A\\
-% %   & \operator{rsfc} = alg \comp DD_{\infty}(\operator{rsfc}) \comp D(\operator{fan}) \comp \InOp
-% % \end{align*}
-% % At his point we can given the definition of the algebra $alg$ which effectively implements knapsack.
-% % Firs we define the $\operator{lookup}$ function which returns (if present) the element indexed in the memory table
-% % \begin{align*}
-% %   & \operator{lookup} : \Nat \to D_{\infty} \R \to \R_{\bot}\\
-% %   & \operator{lookup} (0,   \pair{r}{\textunderscore})  = r\\
-% %   & \operator{lookup} (n+1, \pair{r}{\inl})         = \bot\\
-% %   & \operator{lookup} (n+1, \pair{r}{\inr(table)})  = \operator{lookup} (n, table)
-% % \end{align*}
-% % and now we can define the algebra over the reals
-% % \begin{align*}
-% %   & \operator{alg} : DD_{\infty} \R \to \R\\
-% %   & \operator{alg}\; (\inl)  = 0\\
-% %   & \operator{alg}\; (\inr(table))  = \operator{maximun}_{0} [ v + \operator{lookup}\; table\; (i - w) \mid (w,v) \leftarrow wvs, w \le i]
-% % \end{align*}
-
-% % It remains to show how an $\operator{rsfc}$ is implemented in terms of
-% % hylomorhisms.
-
-% % \paragraph{Implementation into hylomorphisms}
-% % In order to implement rsfcs as hylos we need a parametric distributive law
-% % \[
-% %   \lambda_{X} : GG_{\infty}X \to G_{\infty}GX
-% % \]
-% % This can be defined as
-% % $\lambda = G_{\infty}G(\head) \comp \anacofree{G(\tail)}$.  The distributive law
-% % is instrumental to be able to lift coalgebras in the following sense. Give a
-% % $G$-coalgebra on $X$, a lifting $H$ is a functor which lifts this coalgebra to a
-% % $G$-coalgebra on $HX$.  In particular, the lifting $G_{\lambda}$ takes
-% % $G_{\infty}$-coalgebras on $X$ and transforms them into $G_{\infty}$-coalgebras
-% % on $GX$
-% % \begin{align*}
-% %   & G_{\lambda}(c) : GA \to G_{\infty} GA\\
-% %   & G_{\lambda}(c) = \lambda \comp G (c)
-% % \end{align*}
-% % where $c : C \to GC$.
-
-% % Now, given an algebra $a : GG_{\infty} A \to A$ on the carrier $A$ we can
-% % convert it into an algebra for the carrier $G_{\infty} A$ by
-% % \begin{align*}
-% %   & \floor{a} : GG_{\infty} A \to G_{\infty} A\\
-% %   & \floor{a} = G_{\infty}(a) \comp \anacofree{G_{\lambda}(\delta)}
-% % \end{align*}
-% % where $\eta_{A} : A \to G_{\infty} A$ is the unit of the comonad creating an
-% % infinite constant list.
-
-% % At this point the hylo is simply defined by
-% % $\hylo{\InOp}{a} : \mu F \to G_{\infty}A$  and the rsfc is derived as follows
-% % \begin{align*}
-% %   & \operator{rsfc} : \mu G \to A \\
-% %   & \operator{rsfc} = \head \comp \hylo{\InOp}{a}
-% % \end{align*}
-
-% % % \subsection{Catamorphisms and Anamorphism}
-% % % A catamorphism is the unique program defined as
-% % % \[
-% % %   x = a \comp \fmap{x} \comp \InOp
-% % % \]
-% % % Can be implemented:
-% % % \[
-% % %   \cata{a} = \hylo{\InOp}{a}
-% % % \]
-
-% % % An anamorphism is the unique program satisfying the equation
-% % % \[
-% % %   x = \OutOp \comp \fmap{x} \comp{c}
-% % % \]
-% % % The anamorphism is indicated by $\ana{c}$. Recursion schemes given by a
-% % % corecursive algebra such as $\OutIso$ are also called hylomorphism but they are
-% % % indicated with the symbol for the anamorphism, $\cohylo{c}{\OutOp}$.
-
-% % \paragraph{Free Monads}
-% % Given an endofunctor $D$ and a type $A$ the free monad is defined as the least
-% % solution to the following domain equation
-% % \[
-% %   D^{*}A = A + DD^{*}A
-% % \]
-% % where $\eta_{A} : A \to D^{*}A$ is the left injection and
-% % $\call_{A} : DD^{*}A \to D^{*}A$ is the right injection. Given an algebra
-% % $a : DA \to A$ there is a unique solution to the following recursive equation
-% % \[
-% %   x \comp \call_{A} = a \comp D(x)
-% % \]
-% % that is $\catafree{a} = \cata{[id_{A}; a]}$. With this operation we can extend
-% % the free algebra $\call$ to the multiplication $\mu : D^{*}D^{*}A \to D^{*}A$
-% % defined as $\mu = \catafree{\call}$.
-
-% % \paragraph{Cofree Comonads}
-% % Given an endofunctor $D$ and a type $A$ the cofree comonad is defined as the
-% % greatest solution to the following domain equation
-% % \[
-% %   D_{\infty}A = A + DD_{\infty}A
-% % \]
-% % where $\head : D_{\infty}A \to A$ is the first projection and
-% % $\tail_{A} : D_{\infty}A \to DD_{\infty}A$ is the second projection.
-% % Given an algebra $c : A \to DA$ there is a unique solution to the following
-% % recursive equation
-% % \[
-% %   \tail_{A} \comp x = D(x) \comp c
-% % \]
-% % that is $\anacofree{c} = \ana{\pair{id_{A}}{c}}$. With the latter operation we
-% % can extend $\tail$ to the comultiplication for the comonad
-% % $\delta : D_{\infty}A \to D_{\infty} D_{\infty}A$ defined by
-% % $\delta = \anacofree{\tail}$.
-
-% % \subsection{Accumulators}
-% % For an algebra $a : D(A^{P})\times P \to A$:
-% % \begin{align*}
-% %   & g : \mu D \times P \to B\\
-% %   & g (d, p) = a (\fmap{(\curry g)} \comp \InOp(d), p)
-% % \end{align*}
-
-% % \paragraph{Hylo Implementation}
-% % Given an algebra $a : D(A^{P}) \to A^{P}$ the accumorphism corresponds to the unique
-% % solution to the following equation
-% % \begin{align*}
-% %   & f : \mu D \to A^{P}\\
-% %   & f = a \comp \fmap (f) \comp \InOp
-% % \end{align*}
-% % thus $f = \hylo{\InOp}{a}$.
-
-% % The complex recursion scheme is then
-% % $g = \uncurry\hylo{\InOp}{a} : \mu D \times P \to B$.
-
-% % \subsection{Mutu-Hylo}
-% % For an algebra $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and an algebra
-% % $a_{2} : D(A_{1} \times A_{2}) \to A_{2}$:
-% % \begin{align*}
-% %   & (g_{1}, g_{2}) : (\mu D, \mu D) \to (A_{1}, A_{2})\\
-% %   & (g_{1}, g_{2}) = (a_{1}, a_{2}) \comp (\fmap{\pair{g_{1}}{g_{2}}},\fmap{\pair{g_{1}}{g_{2}}}) \comp (\InOp, \InOp)
-% % \end{align*}
-
-% % \paragraph{Hylo Implementation}
-% % Given an algebra $a : D(A_{1} \times A_{2}) \to A_{1} \times A_{2}$
-% % \begin{align*}
-% %   & f : \mu D \to A_{1} \times A_{2}\\
-% %   & f = a \comp \fmap{f} \comp \InOp
-% % \end{align*}
-% % for which the solution is an hylo $f = \hylo{\InOp}{a}$.
-
-% % The mutumorphism can be implemented as
-% % $(g_{1}, g_{2}) = (\pi_{1} \comp f, \pi_{2} \comp f)$.
-
-% % \paragraph{Coq Implementation}
-% % Given the two algebras  $a_{1} : D(A_{1} \times A_{2}) \to A_{1}$ and
-% % $a_{2} : D(A_{1} \times A_{2}) \to A_{2}$.
-% % We define the two recursive functions  $f_{1}$ and $f_{2}$ as
-
-% % \[
-% %   f_{1} = \pi_{1} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}} \qquad f_{2} = \pi_{2} \comp \hylo{\InIso}{\pair{a_{1}}{a_{2}}}
-% % \]
-% % % A recursion scheme from a comonad $D_{\infty}$ for a data functor $F$
-
-% % % For example, for the functor $(1 + -)$ an algebra over this functor is a
-% % % function $1 + X \to X$ is a pair of functions $\Nil : 1 \to X$ and
-% % % $\Succ : X \to X$. The cofree comonad over $D$ is defined as follows
-% % % \[
-% % %   D_{\infty} A \iso A \times (1 + D_{\infty})
-% % % \]
-% % % which is the the type of non-empty possibly infinite lists with elements in $A$
-
-% % % First we have to assume a distributive law
-% % % $\lambda : DD_{\infty} \to D_{\infty}D$.  For an algebra
-% % % $a : D(D_{\infty} A) \to A$,
-% % % \begin{align*}
-% % %   & g : \mu D \to A\\
-% % %   & g = a \comp D(G_{\infty}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
-% % % \end{align*}
-% % % where
-% % % $G^{\lambda} (alg : D\mu F \to \mu F) = DGX \xrightarrow{\lambda} GDX \xrightarrow{G(alg)} GX$
-% % % is a $D$-algebra on $GX$. The catamorphism of this algebra has the following
-% % % type
-% % % \[
-% % %   \cata{G^{\lambda}(\InIso)} : \mu D \to G\mu D
-% % % \]
-% % % which is the type of a $G$-algebra used to initialise the memory table.
-
-% % % Given an algebra $a : DG_{\infty} A \to G_{\infty} A$ we can equivalently ask
-% % % what is the unique solution of the following equation
-% % % \begin{align*}
-% % %   & f : \mu F \to G_{\infty} A\\
-% % %   & f = a \comp \fmap{f} \comp \InOp
-% % % \end{align*}
-% % % The solution to this equation can be found again using an hylo
-% % % $f = \hylo{\InOp}{a} : \mu F \to G_{\infty} A$.  The course-of-value recursion
-% % % can be recovered from this by setting $g = \head \comp f$
-
-% % % \subsubsection{Histomorphism}
-% % % The definition of $\lambda$ is:
-% % % \begin{align*}
-% % %   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
-% % %   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
-% % % \end{align*}
-% % % For example, for $GX = 1 + X$ we have to define
-% % % \begin{align*}
-% % %   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
-% % %   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
-% % %   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
-% % % \end{align*}
-% % % % \begin{align*}
-% % % %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
-% % % %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
-% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
-% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
-% % % % \end{align*}
-% % % % Defining the algebra
-% % % % \begin{align*}
-% % % %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
-% % % %   & \operator{alg}\; Zero           = 0\\
-% % % %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
-            % % % % \end{align*}
 \section{Recursion Schemes}
 The structure of the a certain data is very similar to the structure of an
-algorithm which processes that data. Recursion schemes aim at analyzing these
-analogies.
+algorithm which processes that data. This relationship manifests in the form of
+structured recursion schemes which are widely used in functional languages such
+Haskell. Canonical examples are folds (catamorphisms), which consume data, and
+unfolds (anamorphisms), which produce it. While some implementations of
+recursion schemes like $\operator{foldr}$ in Haskell as specific to a particular
+data structure, in this case $\operator{List}$s, we can generalise further these
+ideas to account for generic (co)inductive data types and generic algorithms
+operating on them. This is done by looking at recursion schemes from the point
+of view of category theory.
 
@@
 however, there is a more general way of proving these type of definitions are
-well-defined.
+well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
+the greatest set generated by the constructor
+$\cons : A \times \Str{A} \to \Str{A}$.
 
@@
 homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (B, \beta)$ and
-pronounced ``\emph{catamorphism}''.
+pronounced ``\emph{catamorphism}'' satisfying:
+\begin{equation}
+  \label{eq:cata:unique}
+  f = \cata{\alpha} \iff f = \alpha \circ F(f) \circ \InIso
+\end{equation}
+In words, any $F$-algebra homomorphism from the initial $F$-algebra is a
+catamorphism.
 
@@
 
-\subsection{Dump}
-Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
-by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of type
-\coq{App F A ~> A}:
-\begin{coqcode}
-Notation Alg F A := (App F A ~> A).
-\end{coqcode}
-The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
-where the structure map is an isomorphism. This is sometimes referred to as the
-\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind
-this name shortly. \footnote[1]{To be more precise, initial algebras are
-isomorphisms by Lambek's lemma, but isomorphisms of functors do not necessarily
-correspond to initial algebras.}  We define the least fixed-point of \coq{F} as
-an inductive type:
-\begin{coqcode}
-Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
-\end{coqcode}
-where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
-As an example, the initial \coq{F}-algebra for the
-functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
-being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
-operation \coq{Cons : A * LFix F -> LFix F}.
-
-We define \coq{LFix} as a setoid, where its equivalence relation can be
-described as the least fixed point of \coq{AppR} and we define smart
-constructors for the isomorphism of least fixed points as respectful morphisms:
-\begin{coqcode}
-l_in : App F (LFix F) ~> LFix F                l_out : LFix F ~> App F (LFix F)
-\end{coqcode}
-
-The least fixed-point for \coq{F} is the initial \coq{F}-algebra in the sense
-that it gives rise to an inductive recursion scheme.  Specifically, for any
-other \coq{F}-algebra there exists a \emph{unique} map, known as a \emph{fold} or
-\emph{catamorphism}, such that it structurally deconstructs the data type
-using \coq{LFix_out}, calls itself recursively and then composes the result of
-the recursive call using the given an \coq{F}-algebra. In other words, for any
-give \coq{F}-algebra there exists a unique \coq{F}-algebra homomorphism from the
-initial one. We define it in Coq as follows:
-\begin{coqcode}
-  Definition cata_f `{F : Cont Sh P} (alg : Alg F A) : LFix -> A
-:= fix f (x : LFix) := match x with
-  | LFix_in ax => let (sx, kx) := ax in alg (MkApp sx (fun e => f (kx e))) end.
-\end{coqcode}
-It is easy to show that this function is a respectful morphism of
-\coq{F}-algebras. In fact, it is possible to define it as a map of the
-following type:
-\begin{coqcode}
-  cata : forall `{F : Cont Sh P} `{setoid A}, Alg F A ~> LFix ~> A
-\end{coqcode}
-We prove that catamorphisms satisfy the universality property we explained previously:
-\begin{coqcode}
-Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
-    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
-\end{coqcode}
-In other words, if there is any other \coq{f} with the same structural recursive
-shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
-catamorphism.
-
 \subsection{Coalgebras and Anamorphisms}
 \label{sec:coalg}
-Algebras and catamorphisms dualise nicely to the coinductive setting.  The dual
-of an algebra is a coalgebra. A coalgebra can be thought of as an
-\emph{observation} map. For example, for a type of states \coq{X} and an
-alphabet type \coq{L} we can define a labelled transition system (LTS)
-on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{F X = L * X}
-implementing the \emph{transition} map. In particular, for a
-state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
-where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
-
-In general, for a functor \coq{F}, an \coq{F}-coalgebra is a pair of a
-type \coq{X} called the \emph{carrier} of the coalgebra and a structure
-map \coq{X -> F X}.
-
-In our development we use the following notation for coalgebras:
-\begin{coqcode}
-Notation Coalg F A := (A ~> App F A).
-\end{coqcode}
-
-Dually to the initial \coq{F}-algebra, a final \coq{F}-coalgebra is the greatest
-fixed-point for a functor \coq{F}. We define it using a coinductive data type:
-\begin{coqcode}
-CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
-\end{coqcode}
-where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix_in} is its
-inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
-defined as a setoid, with an equivalence relation that is the greatest fixpoint
-of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
-greatest fixed points
-\begin{coqcode}
-g_in : App F (GFix F) ~> GFix F                g_out : GFix F ~> App F (GFix F)
-\end{coqcode}
-
-The greatest fixed-point is a terminal \coq{F}-coalgebra in the sense that it
-yields a coinductive recursion scheme. Specifically, for any
-other \coq{F}-coalgebra there exists a \emph{unique} map, called the
-\emph{unfold} or \emph{anamorphism}, such that it applies the observation map,
-corecursively generates the rest of the computation and composes the result of
-the corecursive call by using the algebra \coq{GFix_in}. In other words, for any
-given \coq{F}-coalgebra there exists a unique \coq{F}-coalgebra homomorphism
-into the terminal \coq{F}-coalgebra.
-
-% Consider the transition system below composed by three states \coq{x1}, \coq{x2}
-% and \coq{x3} with the transitions labelled by the set of labels \coq{\{a, b\}} and define the map \coq{c} such that it implements the transition system below
-% \[\begin{tikzcd}
-%     {x_1} & {x_2} & {x_3} \arrow["b"', loop, distance=2em, in=35, out=325]
-% 	\arrow["a", from=1-1, to=1-2]
-% 	\arrow["b", from=1-2, to=1-3]
-% \end{tikzcd}\]
-% Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
-% carrier \coq{X} is a map \coq{X -> F X}. In the previous example \coq{c} is
-% an \coq{F}-coalgebra on the carrier \coq{X} with \coq{F X = L * X}.
-
-We define anamorphisms as follows:
-\begin{coqcode}
-Definition ana_f_ (c : Coalg F A) :=
-  cofix f x :=
-    match c x with | MkApp sx kx => GFix_in (MkApp sx (fun e => f (kx e))) end.
-
-Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*... ana_f_ ... *)
-\end{coqcode}
-From this definition the universality property falls out:
-\begin{coqcode}
-Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix F)
-: f =e ana h <-> f =e g_in \o fmap f \o h.
-\end{coqcode}
-In words, for any \coq{F}-coalgebra, if there is any other function \coq{f}
-that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the
+The dual of an algebra is a coalgebra. For an endofunctor $B : \CatC \to \CatC$,
+a $B$-\emph{coalgebra} is a pair $(X, c)$ where $X \in \Obj(\CatC)$ is the
+carrier of the coalgebra and $c : X \to BX$ is a morphism.
+
+For example, for a set of states $X$ and a finite set of labels $L$ we can
+define a labelled transition system (LTS) on $X$ as a function $X \to BX$
+implementing the \emph{transition system} with $BX = L \times X$.  In
+particular, for a state $x \in X$, $c(x_{1})$ returns a pair $(l, x_{2})$ where
+$l \in L$ is the observable action and $x_{2} \in X$ is the next state.  In this
+case, we can even set some notation for the transition map:
+\[
+  x \xrightarrow{l} y \eqdef c(x) = (l, y)
+\]
+The category of $B$-coalgebras, denoted $B\Coalg$, is the category where objects
+are $B$-coalgebras and morphisms $f : (X, \gamma_{X}) \to (Y, \gamma_{Y})$ are
+$B$-coalgebra hommorphisms $f : A \to B$, that is
+$ \gamma_{Y} \circ f = F(f) \circ \gamma_{X}$. Terminal object in this category
+$B$-coalgebra corresponds to the greatest fixed-point for the functor $B$,
+denoted by $(\nu B, \OutIso)$ where $\OutIso$ denotes the final $B$-coalgebra
+$\OutIso : \nu B \to B\nu B$. For example, the terminal coalgebra for the
+functor $BX = L \times X$ is the set of of infinite streams, that is the
+greatest solution to the domain equation $\Str{A} \iso A \times \Str{A}$.  As a
+result, for any $B$-coalgebra $(B, \gamma)$ there exists a unique $B$-coalgebra
+homomorphism into the terminal coalgebra $(\nu B, \OutIso)$ which is denoted by
+$\ana{\gamma}$ and pronounced \emph{catamorphism}. We spell out the uniqueness
+property:
+\begin{equation}
+  \label{eq:ana:unique}
+  f = \ana{\gamma} \iff f = \OutOp \circ B(f) \circ \gamma
+\end{equation}
+In words, for any $B$-coalgebra, if there is any other $B$-coalgebra
+homomorphism $f$ into the terminal object then it must be the anamorphism on the
 same coalgebra.
 
-\subsection{Hylomorphisms}
+\subsection{Recursive Coalgebras and Hylomorphism}
+\label{sec:rec-coalgebras}
 Recursion schemes provide an abstract way to consume and generate data. We now
@@
 
-\subsection{Recursive Coalgebras and Hylomorphism}
-\label{sec:rec-coalgebras}
 Hylomorphisms capture the concept of \emph{divide-and-conquer} algorithms where
@@
 % where \coq{h} is a recursive coalgebra.
+
+\subsection{Dump}
+Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
+by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of type
+\coq{App F A ~> A}:
+\begin{coqcode}
+Notation Alg F A := (App F A ~> A).
+\end{coqcode}
+The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
+where the structure map is an isomorphism. This is sometimes referred to as the
+\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind
+this name shortly. \footnote[1]{To be more precise, initial algebras are
+isomorphisms by Lambek's lemma, but isomorphisms of functors do not necessarily
+correspond to initial algebras.}  We define the least fixed-point of \coq{F} as
+an inductive type:
+\begin{coqcode}
+Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
+\end{coqcode}
+where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
+As an example, the initial \coq{F}-algebra for the
+functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
+being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
+operation \coq{Cons : A * LFix F -> LFix F}.
+
+We define \coq{LFix} as a setoid, where its equivalence relation can be
+described as the least fixed point of \coq{AppR} and we define smart
+constructors for the isomorphism of least fixed points as respectful morphisms:
+\begin{coqcode}
+l_in : App F (LFix F) ~> LFix F                l_out : LFix F ~> App F (LFix F)
+\end{coqcode}
+The least fixed-point for \coq{F} is the initial \coq{F}-algebra in the sense
+that it gives rise to an inductive recursion scheme.  Specifically, for any
+other \coq{F}-algebra there exists a \emph{unique} map, known as a \emph{fold} or
+\emph{catamorphism}, such that it structurally deconstructs the data type
+using \coq{LFix_out}, calls itself recursively and then composes the result of
+the recursive call using the given an \coq{F}-algebra. In other words, for any
+give \coq{F}-algebra there exists a unique \coq{F}-algebra homomorphism from the
+initial one. We define it in Coq as follows:
+\begin{coqcode}
+  Definition cata_f `{F : Cont Sh P} (alg : Alg F A) : LFix -> A
+:= fix f (x : LFix) := match x with
+  | LFix_in ax => let (sx, kx) := ax in alg (MkApp sx (fun e => f (kx e))) end.
+\end{coqcode}
+It is easy to show that this function is a respectful morphism of
+\coq{F}-algebras. In fact, it is possible to define it as a map of the
+following type:
+\begin{coqcode}
+  cata : forall `{F : Cont Sh P} `{setoid A}, Alg F A ~> LFix ~> A
+\end{coqcode}
+We prove that catamorphisms satisfy the universality property we explained previously:
+\begin{coqcode}
+Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
+    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
+\end{coqcode}
+In other words, if there is any other \coq{f} with the same structural recursive
+shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
+catamorphism.
 

### 2024-10-09 paper/paper.tex
Some edits to section 3

@@
+%%% TeX-command-extra-options: "-shell-escape"
 \documentclass{llncs}
@@
   Generic programming with recursion schemes provides a powerful abstraction for
-structuring recursion while ensuring termination and providing reasoning about
-program equivalences, as well as deriving optimisations which have been
-successfully applied to functional programming. Formalising recursion schemes in
-a type theory offers additional termination guarantees, but it often requires
+structuring recursion, in part due to the rigorous set of algebraic laws that
+they satisfy. These laws are the basis for reasoning about program equivalences
+and, therefore , they can be used for reasoning about program correctness and
+optimisations.  Some of these optimisations are successfully applied by
+compilers of (functional) languages.  Formalising recursion schemes in a type
+theory offers additional termination guarantees, but it often requires
 compromises affecting the resulting code, such as imposing performance
@@
 well-estabished \emph{algebraic properties} that can serve as a foundation for
-program optimisation (e.g.\ \emph{fusion} laws, or semi-automatic
+reasoning about program equivalences, transformations, and
+optimisations (e.g.\ \emph{fusion} laws, or semi-automatic
 parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
@@
 this algebra of programming to \emph{calculate} an efficient version of the same
-algorithm. Suppose, for example, that we want to write a program sorts a list of
+algorithm. Suppose, for example, that we want to write a program that sorts a list of
 integers, and then multiplies by 2 all its elements. In OCaml, we may write this
@@
 techniques, but these implementations are scarce, not up to date, and not
-verified. For example, Cunha et al. automated program calculation techniques by
-a custom Haskell implementation~\cite{DrHylo}.
+verified in a proof assistant. For example, Cunha et al.~\cite{DrHylo} automated
+program calculation techniques by a custom Haskell implementation.
 
@@
 recursion schemes due to termination issues, or do not focus on proving
-algebraic laws of generic recursion schemes~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.
-Indeed, most of the mechanisations of generic
-recursion schemes focus on \emph{maps} and \emph{folds}, and few authors focus
-on the \emph{unfolds}. In general, Hinze et al.~\cite{HinzeWG15} has proven that
-all known  recursion schemes are instances of a  divide-and-conquer algorithm
-which goes under the name of \emph{hylomorphism}~\cite{MeijerFP91,HuIT96}.  To
-date, no work has mechanised hylomorphisms in the Coq proof assistant, together
-with their algebraic laws.
+algebraic laws of generic recursion
+schemes~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.  Indeed,
+most of the mechanisations of generic recursion schemes focus on \emph{maps} and
+\emph{folds}, and few authors focus on the \emph{unfolds}. Omitting
+\emph{unfolds} severely limits the expressivity of the resulting mechanisations,
+and prevent them from being able to mechanise the most general recursion scheme:
+a divide-and-conquer algorithm which goes under the name of
+\emph{hylomorphism}~\cite{MeijerFP91,HuIT96}. This generality has been proven by
+Hinze et al.~\cite{HinzeWG15}, by showing that \emph{all known  recursion
+schemes are instances of hylomorphisms}.  To date, no work has mechanised
+hylomorphisms in the Coq proof assistant, together with their algebraic laws.
 
@@
 well as the performance of the code that is run \emph{within Coq}, but it does
-not allow for extraction of idiomatic OCaml code. This is unfortunate since code
-extration is what allows the execution of code that has been verified in
+not allow for extraction of idiomatic OCaml code, and it is not well-suited for
+program calculation. This is unfortunate since code extration is what allows the
+execution of code that has been verified in
 Coq~\cite{OnoHTNH11,Larchey-Wendling23,MiculanP12,Sakaguchi20}.  In Abreu et
-al.'s approach, extraction (1) does not preserve the recursive structure of
+al's approach, extraction (1) does not preserve the recursive structure of
 common implementations; and (2) leads to unsafe casts like \ocaml{Obj.magic} in
@@
 This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
-\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code. The
-full mechanisation can be found on
-Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}\dcas{I SHOULD
-ANONYMISE THIS}. While programmers still need to reason about the termination of
+\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code.
+The
+full mechanisation is open source.
+%can be found on
+%Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}\dcas{I SHOULD
+%  ANONYMISE THIS}.
+While programmers still need to reason about the termination of
 their programs, our mechanisation will allow the use of the algebraic laws of
@@
 \begin{itemize}
+  \item In Section~\ref{sec:rec-schemes} we give an overview about recursion
+schemes.
   \item In Section~\ref{sec:containers} we formalise the type of container
 functors ensuring the presence of least and greatest fixed-points for functors
-and suitably adapted for program extraction
+and suitably adapted for program extraction.
   \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and
-hylomorhisms and their universal properties
+hylomorhisms and their universal properties.
   \item In Section~\ref{sec:examples} we use the framework to formalise examples
@@
 
-\section{Recursion Schemes}
-The structure of the a certain data is very similar to the structure of an
+\section{Recursion Schemes}\label{sec:rec-schemes}
+The structure of data is very similar to the structure of an
 algorithm which processes that data. This relationship manifests in the form of
@@
 unfolds (anamorphisms), which produce it. While some implementations of
-recursion schemes like $\operator{foldr}$ in Haskell as specific to a particular
+recursion schemes like $\operator{foldr}$ in Haskell are specific to a particular
 data structure, in this case $\operator{List}$s, we can generalise further these
@@
 This latter component is also called the \emph{functorial action} and can be
-thought of as the $\operator{fmap}$ program in functional programming.
+thought of as the $\operator{fmap}$ higher-order function in functional
+programming.
 
@@
 constructors $\nil : 1 \to \List{A}$ and
-$\cons : A \times \List{A} \to \List{A}$. This type is a functor with functorial
-action $\List{f} : \List{A} \to \List{B}$ given by $\List{f}(*) = *$ and
+$\cons : A \times \List{A} \to \List{A}$.
+The terminal object in $\Set$ is the \emph{unit} type, that only has one
+inhabitant, $* : 1$, analogous to the unit type in Haskell: \haskell{() :: ()}.
+The $\List{-}$ type is a functor. Its action on objects is to take any set
+$A$ to $\List{A}$, and its functorial
+action $\List{f} : \List{A} \to \List{B}$ is given by $\List{f}(*) = *$ and
 $\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
-$\List{f}$ is well-defined as it recursively calls on a smaller argument,
-however, there is a more general way of proving these type of definitions are
+$\List{f}$ is well-defined as it recursively calls on a smaller argument.
+However, there is a more general way of proving these type of definitions are
 well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
@@
 \label{sec:containers}
-In order to abstract recursion patters we need to be able to abstract away from
-the particular shape of the data.  This is achieved by introducing the concept
-of \emph{functors} which have the suitable fixed-point properties, i.e.\ those
-who have a initial algebras. A common approach to construct such functors is to
-use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:functors}).
-However, reasoning about container equality will require us to consider both
-functional extensionality and heterogeneous equality. We avoid these axioms by
-introducing a custom equivalence relation on types
-(Section~\ref{sec:types-morphisms}).
+
+In this section, we focus on mechanising \emph{container functors} in a way that
+is suitable for code extraction. Recall from Section~\ref{sec:rec-schemes} that
+we need to be able toabstract away from the particular shape of the data, and
+that this is achieved by introducing the concept of \emph{functors} which have
+the suitable fixed-point properties, i.e.\ those which have a initial algebras.
+A common approach to construct such functors is to use
+\emph{containers}~\cite{AbbottAG05}.  However, reasoning about container
+equality will require us to consider both functional extensionality and
+heterogeneous equality. We avoid these axioms by introducing a custom
+equivalence relation on types (Section~\ref{sec:types-morphisms}).
 
@@
 \label{sec:functors}
-[CUT]
-A container is defined by a type of \emph{shapes}
-and a \emph{family of position types} indexed by shapes.
-\begin{coqcode}
-Context (Shape : Type) (Pos : Shape -> Type).
-\end{coqcode}
+
+A straightforward mechanisation of functors as defined in
+Section~\ref{sec:rec-schemes} does not work in Coq due to the strict positivity
+condition. In fact, if we simply require that functors are functions
+\coq{F : Type -> Type}
+satisfying the necessary identity and composition properties, then we will not
+be able to construct least/greatest fixed points of this type, and we will not
+be able to mechanise recursion schemes that follow this structure:
+\begin{coqcode}
+(* REJECTED due to negative occurrence of (LFix F) *)
+Inductive LFix (F : Type -> Type) := LFix_in (lfix_out : F (LFix F)).
+\end{coqcode}
+One common workaround for this is the use of Mendler-style
+recursion~\cite{Mendler91}. Roughly, the idea is to ``pack'' an abstract type
+together with the functor that captures the structure of the recursion. From
+this abstract type, we will be able to ``extract'' the contents of the functor.
+This is, for example, the solution used by the PaCo library~\cite{PaCo}, and a
+simplified version of it is the following:
+\begin{coqcode}
+Inductive MendlerLFix (F : Type -> Type) :=
+  In Y (next : Y -> MendlerLFix F) (in_op : F Y).
+\end{coqcode}
+While this definition solves the strict positivity problem, it does not solve
+the problems of: (1) reasoning about data/program equalities; and (2) extraction
+to idiomatic OCaml code. To reason about program equalities, we would need to
+reason about equalities of functions with \emph{some abstract} \coq{Y} as their
+domain.  Furthermore, the extraction of \coq{F Y} will lead to unsafe casts in
+the extracted OCaml code.
+
+\paragraph{Containers} Another common solution is to the strict positivity
+problem is the use \emph{containers}~\cite{AbbottAG05}.  A container is defined
+by a type of \emph{shapes} and a \emph{family of position types} indexed by
+shapes.
+\begin{coqcode}
+(Shape : Type) (Pos : Shape -> Type)
+\end{coqcode}
+
 For the functor \coq{F} we introduced in the previous paragraph we would

### 2024-10-09 paper/paper.tex
Restored cut out contents

@@
 \label{sec:recursion-schemes}
-[MOVED to background]
+\dcas{Do not cut below}
+Recursion schemes provide an abstract way to consume and generate data. We now
+proceed onto describing how to formalise hylomorphisms in Coq. We first
+formalise algebras for container extensions (Section~\ref{sec:algebras}), then
+we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
+these notions to formalise recursive coalgebras and hylomorphisms
+(Section~\ref{sec:rec-coalgebras}).
+
+\subsection{Algebras and Catamorphisms}
+\label{sec:algebras} An algebra is a set \coq{A} (the
+\emph{carrier} of the algebra) together with some operations on it sometimes
+subject to some equational laws. For example, given such a type \coq{A} we can
+define the monoid operations as \coq{u : unit -> A} for unit
+and \coq{m : A * A -> A} for multiplication. Notice
+that \coq{F X = unit + X * X} is a functor on \coq{X} which means we can
+equivalently describe these two maps as a single one of type \coq{F X -> X}
+which we call the algebra for the functor \coq{F} satisfying the unit and
+associative laws of the monoid. In this work we will be using algebras to
+describe the operations of a data type and so we will not be needing additional
+equations on these operations.
+
+Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
+by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of
+type \coq{App F A ~> A}:
+\begin{coqcode}
+Notation Alg F A := (App F A ~> A).
+\end{coqcode}
+The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
+where the structure map is an isomorphism. This is sometimes referred to as the
+\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind this name shortly. \footnote[1]{To be more precise,
+initial algebras are isomorphisms by Lambek's lemma, but isomorphisms of
+functors do not necessarily correspond to initial algebras.}  We define the
+least fixed-point of \coq{F} as an inductive type:
+\begin{coqcode}
+Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
+\end{coqcode}
+where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
+As an example, the initial \coq{F}-algebra for the
+functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
+being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
+operation \coq{Cons : A * LFix F -> LFix F}.
+
+We define \coq{LFix} as a setoid, where its equivalence relation can be
+described as the least fixed point of \coq{AppR} and we define smart
+constructors for the isomorphism of least fixed points as respectful morphisms:
+\begin{coqcode}
+l_in : App F (LFix F) ~> LFix F                l_out : LFix F ~> App F (LFix F)
+\end{coqcode}
+
+The least fixed-point for \coq{F} is the initial \coq{F}-algebra in the sense
+that it gives rise to an inductive recursion scheme.  Specifically, for any
+other \coq{F}-algebra there exists a \emph{unique} map, known as a \emph{fold} or
+\emph{catamorphism}, such that it structurally deconstructs the data type
+using \coq{LFix_out}, calls itself recursively and then composes the result of
+the recursive call using the given an \coq{F}-algebra. In other words, for any
+give \coq{F}-algebra there exists a unique \coq{F}-algebra homomorphism from the
+initial one. We define it in Coq as follows:
+\begin{coqcode}
+  Definition cata_f `{F : Cont Sh P} (alg : Alg F A) : LFix -> A
+:= fix f (x : LFix) := match x with
+  | LFix_in ax => let (sx, kx) := ax in alg (MkApp sx (fun e => f (kx e))) end.
+\end{coqcode}
+It is easy to show that this function is a respectful morphism of
+\coq{F}-algebras. In fact, it is possible to define it as a map of the
+following type:
+\begin{coqcode}
+  cata : forall `{F : Cont Sh P} `{setoid A}, Alg F A ~> LFix ~> A
+\end{coqcode}
+We prove that catamorphisms satisfy the universality property we explained previously:
+\begin{coqcode}
+Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
+    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
+\end{coqcode}
+In other words, if there is any other \coq{f} with the same structural recursive
+shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
+catamorphism.
+
+\subsection{Coalgebras and Anamorphisms}
+\label{sec:coalg}
+Algebras and catamorphisms dualise nicely to the coinductive setting.  The dual
+of an algebra is a coalgebra. A coalgebra can be thought of as an
+\emph{observation} map. For example, for a type of states \coq{X} and an
+alphabet type \coq{L} we can define a labelled transition system (LTS)
+on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{F X = L * X}
+implementing the \emph{transition} map. In particular, for a
+state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
+where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
+
+In general, for a functor \coq{F}, an \coq{F}-coalgebra is a pair of a
+type \coq{X} called the \emph{carrier} of the coalgebra and a structure
+map \coq{X -> F X}.
+
+In our development we use the following notation for coalgebras:
+\begin{coqcode}
+Notation Coalg F A := (A ~> App F A).
+\end{coqcode}
+
+Dually to the initial \coq{F}-algebra, a final \coq{F}-coalgebra is the greatest
+fixed-point for a functor \coq{F}. We define it using a coinductive data type:
+\begin{coqcode}
+CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
+\end{coqcode}
+where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix_in} is its
+inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
+defined as a setoid, with an equivalence relation that is the greatest fixpoint
+of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
+greatest fixed points
+\begin{coqcode}
+g_in : App F (GFix F) ~> GFix F                g_out : GFix F ~> App F (GFix F)
+\end{coqcode}
+
+The greatest fixed-point is a terminal \coq{F}-coalgebra in the sense that it
+yields a coinductive recursion scheme. Specifically, for any
+other \coq{F}-coalgebra there exists a \emph{unique} map, called the
+\emph{unfold} or \emph{anamorphism}, such that it applies the observation map,
+corecursively generates the rest of the computation and composes the result of
+the corecursive call by using the algebra \coq{GFix_in}. In other words, for any
+given \coq{F}-coalgebra there exists a unique \coq{F}-coalgebra homomorphism
+into the terminal \coq{F}-coalgebra.
+
+% Consider the transition system below composed by three states \coq{x1}, \coq{x2}
+% and \coq{x3} with the transitions labelled by the set of labels \coq{\{a, b\}} and define the map \coq{c} such that it implements the transition system below
+% \[\begin{tikzcd}
+%     {x_1} & {x_2} & {x_3} \arrow["b"', loop, distance=2em, in=35, out=325]
+% 	\arrow["a", from=1-1, to=1-2]
+% 	\arrow["b", from=1-2, to=1-3]
+% \end{tikzcd}\]
+% Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
+% carrier \coq{X} is a map \coq{X -> F X}. In the previous example \coq{c} is
+% an \coq{F}-coalgebra on the carrier \coq{X} with \coq{F X = L * X}.
+
+We define anamorphisms as follows:
+\begin{coqcode}
+Definition ana_f_ (c : Coalg F A) :=
+  cofix f x :=
+    match c x with | MkApp sx kx => GFix_in (MkApp sx (fun e => f (kx e))) end.
+
+Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*... ana_f_ ... *)
+\end{coqcode}
+From this definition the universality property falls out:
+\begin{coqcode}
+Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix F)
+: f =e ana h <-> f =e g_in \o fmap f \o h.
+\end{coqcode}
+In words, for any \coq{F}-coalgebra, if there is any other function \coq{f}
+that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the
+same coalgebra.
+
+\subsection{Recursive Coalgebras and Hylomorphism}
+\label{sec:rec-coalgebras}
+Hylomorphisms capture the concept of \emph{divide-and-conquer} algorithms where
+the input is first destructured (\emph{divide}) in smaller parts by means of a
+coalgebra which are computed recursively and then composed back together
+(\emph{conquer}) by means of an algebra.
+
+In general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
+unique solution (when it exists) to the equation
+\begin{equation}
+  \label{eq:hylo}
+  f = a \circ F\; f \circ c
+\end{equation}
+As we stated earlier, a solution to this equation does not exist for an arbitrary  algebra and coalgebra pair and, in fact, this definition cannot be accepted by Coq.
+
+In order to find a solution we restrict ourselves to the so-called
+\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  An example of a
+recursive coalgebra is the \coq{partition} function in quicksort which
+destructures a list into a pivot and two sublist and as long as the sublists are
+smaller the partitioning function still yields a unique solution to the
+recursion scheme.
+
+We mechanise \emph{recursive
+hylomorphisms} which are guaranteed to have a unique solution to the
+hylomorphism equation. These are hylomoprhisms where the coalgebra is
+\emph{recursive}, i.e. coalgebras that terminate on all inputs. The following
+predicate captures that a coalgebra terminates on an input:
+\begin{coqcode}
+Inductive RecF (h : Coalg F A) : A -> Prop :=
+| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
+\end{coqcode}
+A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra
+\coq{c} such that \coq{forall x, RecF c x}.
+Recursive hylomorphisms are implemented in Coq recursively on the structure of
+the proof for the type (\coq{RecF}) as follows:
+\begin{coqcode}
+Definition hylo_def (a : Alg F B) (c : Coalg F A)
+  : forall (x : A), RecF c x -> B := fix f x H
+  := match c x as h0 return (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
+     with | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
+     end (RecF_inv H).
+\end{coqcode}
+We use \coq{RecF_inv} to obtain the structurally smaller proof to use in
+the recursive calls. As we did with catamorphisms and anamorphisms, we
+prove that \coq{hylo_def} is respectful, and use this proof to build the
+corresponding higher-order proper morphism:
+\begin{coqcode}
+  hylo : forall `{F : Cont Sh P} `{setoid A} `{setoid B},
+    Alg F B ~> RCoalg F A ~> A ~> B
+\end{coqcode}
+Finally, we show that recursive hylomorphisms are the unique solution to the
+hylomorphism equation.
+\begin{coqcode}
+Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
+    : f =e hylo g h <-> f =e g \o fmap f \o h.
+\end{coqcode}
+Hylomorphisms fusion falls out from the universal property:
+\begin{coqcode}
+Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
+  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
+\end{coqcode}
+% Qed.
+%   rewrite hylo_univ, fmap_comp, compA, <- E2,
+%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
+
+  % rewrite hylo_univ, fmap_comp, compA, <- E2,
+  %         <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
+
+  % (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
+  % (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
+  % (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
+  % (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
+  % (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
+  % (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
+  % (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
+  % (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
+  % (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
+Using the hylo fusion law, we can prove the well-known \emph{deforestation
+optimisation}. This is when two consecutive recursive computations, one that
+builds a data structure, and another one that consumes it, can be fused together
+into a single recursive definition. This, in turn, allows us to prove that a
+recursive hylomorphism is the composition of a catamorphism and a recursive
+anamorphism.
+\begin{coqcode}
+Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
+  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
+  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
+\end{coqcode}
+
+% \subsection{Recursive Anamorphisms}
+% Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
+% inductive data type in the following way.  Intuitively, a recursive coalgebra
+% can be applied only finitely many times, therefore when this is applied to an
+% anamorphism the only possible traces it can produce from the seed function are
+% the finite ones. We denote this special kind of recursion scheme \emph{recursive
+% anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
+% can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
+% exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
+% algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
+% from the universality property of the hylomorphism and the fact that recursive
+% anamorphisms satisfy the same equation.
+%
+% We now show how this is proven in the Coq development. First we define a
+% recursive anamorphism as a function into \coq{LFix F}:
+% \begin{coqcode}
+% Definition rana_f__ `{setoid A} (c : Coalg F A)
+%   : forall x : A, RecF c x -> LFix F
+%   := fix f (x : A) (H : RecF c x) :=
+%        let hx := c x in
+%        LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
+%
+% Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F := ...
+% \end{coqcode}
+% We omit the boilerplate details to obtain the setoid morphism.
+% The recursive anamorphism has the following universality property:
+% \begin{coqcode}
+% Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
+%     : f =e rana h <-> f =e l_in \o fmap f \o h.
+% \end{coqcode}
+% We can now show that every hylomorphism is composed by a recursive anamorphism
+% followed by a catamorphism:
+% \begin{coqcode}
+% Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
+%   (g : Alg F B) (h : RCoalg F A)
+% : cata g \o rana h =e hylo g h.
+% \end{coqcode}
+% This result will allow us to \emph{calculate} programs by providing small, easy
+% to verify implementations, and combining them into efficient recursive functions
+% by using Coq's proofs of hylomorphism laws as optimisation passes.
+
+\subsubsection{On the subtype of finite elements}
+In this development we have defined recursive anamorphisms on inductive data
+types.  We might have as well defined them on the subtype of finite elements of
+coinductive data types using a predicate which states when an element of a
+coinductive data type is finite:
+\begin{coqcode}
+Inductive FinF : GFix F -> Prop :=
+| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
+\end{coqcode}
+Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
+is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
+easy to see. We first define a catamorphism \coq{ccata_f_} from the
+subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
+any \coq{F}-algebra.
+\begin{coqcode}
+Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
+  : forall x : GFix F, FinF x -> A := fix f x H :=
+    let hx := g_out x in
+      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
+\end{coqcode}
+We now prove this is isomorphic to the least fixed-point of the functor
+\coq{F}.
+We take the catamorphism from the finite elements of \coq{GFix F} to the
+inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
+inverse is the catamorphism on the restriction of \coq{g_in} to the finite
+elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
+prove the isomorphism:
+\begin{coqcode}
+Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
+Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
+\end{coqcode}
+The finite subtype of \coq{GFix F} allows us to compose
+catamorphisms and anamorphisms, by using the above isomorphism. In our work,
+however, we use \emph{recursive} anamorphisms, defined as \coq{hylo l_in c} for
+a recursive coalgebra \coq{c}, which compose easily with
+catamorphisms.
+% This predicate is equivalent to the proposition that
+% the anamorphism of such coalgebras produce only finite trees:
+% \begin{coqcode}
+% Lemma ana_rec_term `{setoid A} (h : Coalg F A)
+%   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
+% \end{coqcode}
+
+% \mpav{Moved here}
+
+% Note that such proofs will be erased when extracting code to OCaml, and this is
+% standard practice when building general recursive functions in
+% Coq\footnote{\dcas{This technique had a name in the community, but I cannot
+% remember. Something predicates}}. The core definitions are the following:
+% \begin{coqcode}
+% Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
+
+% \end{coqcode}
+% Here, \coq{RecF_inv} provides the structurally smaller element for
+% the Coq recursive definition. We then couple coalgebras with a proof that they
+% are recursive, and add a coercion from recursive coalgebras to coalgebras.
+% \begin{coqcode}
+% Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
+% \end{coqcode}
+% In some cases, proving that coalgebras can only be applied finitely many
+% times is straightforward (e.g.\ they are recursive on some inductive type in
+% Coq). However, in many cases, proving that a coalgebra is recursive directly
+% can be tedious and cumbersome. To ease development, we prove that if a
+% given coalgebra respects a \emph{well-founded relation}, then it is
+% necessarily recursive.
+% \begin{coqcode}
+% Definition respects_relation `{setoid A} (c : Coalg F A)
+%       {B} (m : A -> B) (R : B -> B -> Prop)
+%   := forall x (p : Pos (shape (c x))), R (m (cont (c x) p)) (m x).
+
+% Lemma wf_coalg_rec `{setoid A} {B}
+%   (m : A -> B) (R : B -> B -> Prop) (WF : well_founded R)
+%   (c : Coalg F A) (RR : respects_relation c m R) : RecP c.
+% \end{coqcode}
+% In the above definition, \coq{m} is used as a form of ``metric''
+% for termination, and the well-founded relation is on \coq{B}. This
+% works equivalently to transporting a well-founded relation
+% \coq{R}
+% on
+% \coq{B} to \coq{A} as
+% \coq{R' := fun x y => R (m x) (m y)}, which is also trivially
+% well-founded. Using these definitions, we can construct recursive coalgebras by
+% proving that they respect a well-founded relation, which generally leads to
+% simpler proofs.
+
+% Similarly to the rest of the definitions, we prove that recursive anamorphisms
+% are respectful, and use this fact to build them as higher order proper
+% morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
+% the unique solution to the equation
+% \coq{f =e l_in \o fmap f \o h},
+% where \coq{h} is a recursive coalgebra.
+
 \section{Extraction}

### 2024-10-09 paper/paper.tex
Separate Haskell from math

--- 2024-10-09 paper/paper.tex
@@
 \newcommand{\List}[1]{\operator{List}(#1)}
+\newcommand{\Strbare}{\operator{Str}}
+\newcommand{\Listbare}{\operator{List}}
 \newcommand{\nil}{\operator{nil}}
@@
 Instead of writing this function directly, we may observe that we can derive it
-by \emph{fusing} a regular \emph{quickort} OCaml implementation, with
+by \emph{fusing} a regular \emph{quicksort} OCaml implementation, with
 \ocaml{map (fun x -> x * 2)}. This is the main idea behind program calculation:
@@
 with their algebraic laws.
-
-%Recursive definitions cannot be proven well-defined automatically due to the
-%halting problem. Modern proof assistants like Coq or Agda provide a sound, but
-%incomplete algorithm which syntactically checks for termination or
-%productivity. For recursion, this is done by automatically inferring
-%which argument in the recursive call gets smaller with respects to the original
-%input argument. For productivity, the algorithm checks that the corecursive
-%call appears directly under a constructor to make sure that this function
-%always produces at least one element after each recursive step.
-%This implies that some functions, though well-defined, cannot be accepted by
-%the proof assistant. Examples of this include common sorting algorithms, such
-%as quicksort:
-%%One example is Quicksort, written in OCaml
-%%below:
-%\begin{minted}{OCaml}
-%let rec qsort xs = match divide xs with | None -> []
-% | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
-%\end{minted}
-%While \ocaml{qsort} is a well-defined mathematical function it cannot be
-%accepted by a proof assistant. The reason is that the \ocaml{divide}
-%function \emph{destructuring} the input dives deeper in the input returning
-%two sublists with the head as a pivot.
-%
-%The main approach for implementing non-structural recursion in Coq is to use
-%\emph{well-founded recursion}, where recursive definitions are accompanied by
-%\emph{termination proofs}. Using well-founded recursion, the recursive calls
-%will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
-%one approach is using the \coq{Fix} combinator, in which recursion is done on an
-%\emph{accesibility} predicate on the input, which is a proof that there are no
-%infinitely decreasing chains.}.
-%A benefit of this strategy is that when extracting verified non-structural
-%recursive functions to OCaml, the additional proofs are discarded, resulting in
-%code that closely resembles what a programmer might have written directly in
-%OCaml. However, reasoning about program equivalences needs  such termination
-%proofs to be kept around, leading to the common practice of employing additional
-%reduction lemmas extensively.
-%
-%% A benefit of this approach is that, extracting
-%% verified nonstructural recursive functions to OCaml will erase the additional
-%% proofs, thus producing code that will be closer to what a programmer may have
-%% written directly in OCaml. However, reasoning about program equivalences
-%% requires dealing with such termination proofs, and it is common practice to use
-%% custom reduction lemmas that will be used extensively when proving properties
-%% about nonstructural recursive functions.
-
-%\emph{Structured recursion schemes}~\cite{HinzeW16, HinzeWG15} is a technique
-%for structuring recursive definitions that has found a lot of success in
-%functional programming. The idea is that, rather than writing a self-refential
-%definition, we use a function library, the recursion scheme, which takes as
-%input the non-recursive parts of the algorithm and returns the desired recursive
-%definition which is guaranteed to terminate by the properties of the recursion
-%scheme itself. More complex recursion schemes do exist but it has been
-%proven by Hinze et al.~\cite{HinzeWG15} that they are all instances of a divide-and-conquer
-%algorithm which goes under the name of hylomorphism~\cite{MeijerFP91,HuIT96}.
-%This result, thus, yields a general framework to reason about the algebraic properties of these
-%algorithms, e.g. short-cut fusion and semi-automatic parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}.
-
-% A more complex recursion scheme is the hylomorphism~\cite{MeijerFP91, HuIT96} which has a divide-and-conquer
-% structure. It has been shown by Hinze et
-% al.~\cite{HinzeWG15} that any complex recursion scheme can be transmogrified
-% into an hylomorphisms by means of an adjunction. Furthermore, hylomorphism laws
-% can capture a number of useful equivalences, ranging from common optimisations
-% such as \emph{short-cut fusion}~\cite{TakanoM95}, to semi-automatic
-% parallelisations~\cite{Gibbons96:Third, farmsCastro}.
-
-% Recursion schemes have successfully been applied to functional programming to
-% structure recursive programs, however, because of the presence of
-% non-termination in languages like (e.g.) Haskell, they cannot guarantee
-% recursive definitions well-defined. % This is due to the possibility of applying a
-% %
-% recursion scheme to a non-terminating (co)algebra, but also to the unrestricted
-% % use of data functors.
-
-%However, when used in the context of languages with general recursion like
-%(e.g.) Haskell and ML, recursion schemes cannot ensure termination. This is
-%because the input information itself could be non-terminating. On the other
-%hand, while formalising recursion schemes in a type theory does provide stronger
-%termination guarantees, to the best of our knowledge, not many attempts have
-%been made at mechanising them.
 
@@
 ideas to account for generic (co)inductive data types and generic algorithms
-operating on them. This is done by looking at recursion schemes from the point
-of view of category theory.
+operating on them.
+
+Furthermore folds and unfolds can be shown to capture a wide
+range of recursion schemes such as primitive recursion, mutual recursion,
+dynamic programming algorithms, polymorphic recursion, recursion with
+accumulators and so on. However, for divide-and-conquer algorithms folds need to
+be generalised to hylomorphisms which provide the ultimate basic building block
+for any other recursion scheme. This is done by looking at recursion schemes
+from the point of view of category theory.
 
@@
 in $\Set$ is the empty set $\emptyset$ and the terminal object $1$ is any
-singleton set. In this section, we are going to assume types are sets, hence
-this category can also be viewed as the category of types and programs between
-them.
+singleton set. The reader who is not accustomed with category theory can assume
+types are sets, giving the intuition that the category $\Set$ can also be viewed
+as the category of (simple) types and programs between them.
 
@@
 \end{mathpar}
-This latter component is also called the \emph{functorial action} and can be
-thought of as the $\operator{fmap}$ program in functional programming.
-
-For example, in $\Set$, we can define the set of lists
-$\List{A} \iso 1 + A \times \List{A}$ as the set inductively generated by the
+This latter component is also called the \emph{functorial action}.
+
+In $\Set$, we can define the set of lists
+\[
+  \List{A} \iso 1 + A \times \List{A}
+\]
+as the set inductively generated by the
 constructors $\nil : 1 \to \List{A}$ and
-$\cons : A \times \List{A} \to \List{A}$. This type is a functor with functorial
-action $\List{f} : \List{A} \to \List{B}$ given by $\List{f}(*) = *$ and
-$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
-$\List{f}$ is well-defined as it recursively calls on a smaller argument,
-however, there is a more general way of proving these type of definitions are
-well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
-the greatest set generated by the constructor
-$\cons : A \times \Str{A} \to \Str{A}$.
+$\cons : A \times \List{A} \to \List{A}$. This set extends to a functor
+$\Listbare : \Set \to \Set$. Its functorial action has type
+$\List{f} : \List{A} \to \List{B}$ for a morphism $f : A \to B$ and it is given
+by $\List{f}(*) = *$ and $\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$.
+Notice that the definition $\List{f}$ is well-defined as it recursively calls on
+a smaller argument.
+
+Similarly, the set of streams
+\[
+  \Str{A} \iso A \times \Str{A}
+\]
+is the greatest set generated by the constructor
+$\cons : A \times \Str{A} \to \Str{A}$. This set also extends to a functor
+$\Strbare : \Set \to \Set$ whose functorial action is simply
+$\Str{f}(\cons(a,s)) = \cons (f (a), \Str{f}(s))$ defined coinductively.
 
@@
 $\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
-isomorphism $F\mu F \iso \mu F$. By the property of initial $F$-algebras, for
+isomorphism $F\mu F \iso \mu F$.
+By the property of initial $F$-algebras, for
 any other $F$-algebra $(B, \beta)$ there exists a unique $F$-algebra
@@
 $[\nil,\cons]$ into the algebra $[base, ind]$:
-\begin{align*}
-  & \cata{[base,ind]}(\nil) = base\\
-  & \cata{[base, ind]}(\cons(a, xs)) = ind(a, \cata{[base, ind]}(xs))
-\end{align*}
+
 In functional programming this is commonly referred to as
@@
 are $B$-coalgebras and morphisms $f : (X, \gamma_{X}) \to (Y, \gamma_{Y})$ are
-$B$-coalgebra hommorphisms $f : A \to B$, that is
+$B$-coalgebra homomorphisms $f : A \to B$, that is
 $ \gamma_{Y} \circ f = F(f) \circ \gamma_{X}$. Terminal object in this category
@@
 denoted by $(\nu B, \OutIso)$ where $\OutIso$ denotes the final $B$-coalgebra
-$\OutIso : \nu B \to B\nu B$. For example, the terminal coalgebra for the
-functor $BX = L \times X$ is the set of of infinite streams, that is the
-greatest solution to the domain equation $\Str{A} \iso A \times \Str{A}$.  As a
-result, for any $B$-coalgebra $(B, \gamma)$ there exists a unique $B$-coalgebra
-homomorphism into the terminal coalgebra $(\nu B, \OutIso)$ which is denoted by
-$\ana{\gamma}$ and pronounced \emph{catamorphism}. We spell out the uniqueness
-property:
+$\OutIso : \nu B \to B\nu B$. For
+
+For example, the terminal coalgebra for the functor $BX = L \times X$ is the set
+of of infinite streams, that is the greatest solution to the domain equation
+$\Str{A} \iso A \times \Str{A}$.  As a result, for any $B$-coalgebra
+$(B, \gamma)$ there exists a unique $B$-coalgebra homomorphism into the terminal
+coalgebra $(\nu B, \OutIso)$ which is denoted by $\ana{\gamma}$ and pronounced
+\emph{anamorphism}. We spell out the uniqueness property:
 \begin{equation}
@@
 \label{sec:rec-coalgebras}
-Recursion schemes provide an abstract way to consume and generate data. We now
-proceed onto describing how to formalise hylomorphisms in Coq. We first
-formalise algebras for container extensions (Section~\ref{sec:algebras}), then
-we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together these
-notions to formalise recursive coalgebras and hylomorphisms
-(Section~\ref{sec:rec-coalgebras}).
-
-Hylomorphisms capture the concept of \emph{divide-and-conquer} algorithms where
-the input is first destructured (\emph{divide}) in smaller parts by means of a
-coalgebra which are computed recursively and then composed back together
-(\emph{conquer}) by means of an algebra.
+Recursion schemes provide an abstract way to consume and generate data capturing
+\emph{divide-and-conquer} algorithms where the input is first destructured
+(\emph{divide}) in smaller parts by means of a coalgebra which are computed
+recursively and then composed back together (\emph{conquer}) by means of an
+algebra.
 
@@
 In order to find a solution we restrict ourselves to the so-called
-\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  An example of a
+\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.
+A $F$-coalgebra $(X, c)$ is recursive if
+
+An example of a
 recursive coalgebra is the \coq{partition} function in quicksort which
@@
 
-We mechanise \emph{recursive
-hylomorphisms} which are guaranteed to have a unique solution to the
-hylomorphism equation. These are hylomoprhisms where the coalgebra is
-\emph{recursive}, i.e. coalgebras that terminate on all inputs. The following
-predicate captures that a coalgebra terminates on an input:
-\begin{coqcode}
-Inductive RecF (h : Coalg F A) : A -> Prop :=
-| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
-\end{coqcode}
-A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra
-\coq{c} such that \coq{forall x, RecF c x}.
-Recursive hylomorphisms are implemented in Coq recursively on the structure of
-the proof for the type (\coq{RecF}) as follows:
-\begin{coqcode}
-Definition hylo_def (a : Alg F B) (c : Coalg F A)
-  : forall (x : A), RecF c x -> B := fix f x H
-  := match c x as h0 return (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
-     with | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
-     end (RecF_inv H).
-\end{coqcode}
-We use \coq{RecF_inv} to obtain the structurally smaller proof to use in
-the recursive calls. As we did with catamorphisms and anamorphisms, we
-prove that \coq{hylo_def} is respectful, and use this proof to build the
-corresponding higher-order proper morphism:
-\begin{coqcode}
-  hylo : forall `{F : Cont Sh P} `{setoid A} `{setoid B},
-    Alg F B ~> RCoalg F A ~> A ~> B
-\end{coqcode}
-Finally, we show that recursive hylomorphisms are the unique solution to the
-hylomorphism equation.
-\begin{coqcode}
-Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
-    : f =e hylo g h <-> f =e g \o fmap f \o h.
-\end{coqcode}
-Hylomorphisms fusion falls out from the universal property:
 \begin{coqcode}
@@
 \end{coqcode}
-% Qed.
-%   rewrite hylo_univ, fmap_comp, compA, <- E2,
-%           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
-
-  % rewrite hylo_univ, fmap_comp, compA, <- E2,
-  %         <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
-
-  % (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
-  % (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
-  % (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
-  % (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
-  % (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
-  % (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
-  % (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
-  % (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
-  % (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
 Using the hylo fusion law, we can prove the well-known \emph{deforestation
@@
 
-% \subsection{Recursive Anamorphisms}
-% Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
-% inductive data type in the following way.  Intuitively, a recursive coalgebra
-% can be applied only finitely many times, therefore when this is applied to an
-% anamorphism the only possible traces it can produce from the seed function are
-% the finite ones. We denote this special kind of recursion scheme \emph{recursive
-% anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
-% can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
-% exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
-% algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
-% from the universality property of the hylomorphism and the fact that recursive
-% anamorphisms satisfy the same equation.
-%
-% We now show how this is proven in the Coq development. First we define a
-% recursive anamorphism as a function into \coq{LFix F}:
-% \begin{coqcode}
-% Definition rana_f__ `{setoid A} (c : Coalg F A)
-%   : forall x : A, RecF c x -> LFix F
-%   := fix f (x : A) (H : RecF c x) :=
-%        let hx := c x in
-%        LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
-%
-% Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F := ...
-% \end{coqcode}
-% We omit the boilerplate details to obtain the setoid morphism.
-% The recursive anamorphism has the following universality property:
-% \begin{coqcode}
-% Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
-%     : f =e rana h <-> f =e l_in \o fmap f \o h.
-% \end{coqcode}
-% We can now show that every hylomorphism is composed by a recursive anamorphism
-% followed by a catamorphism:
-% \begin{coqcode}
-% Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
-%   (g : Alg F B) (h : RCoalg F A)
-% : cata g \o rana h =e hylo g h.
-% \end{coqcode}
-% This result will allow us to \emph{calculate} programs by providing small, easy
-% to verify implementations, and combining them into efficient recursive functions
-% by using Coq's proofs of hylomorphism laws as optimisation passes.
-
-\subsubsection{On the subtype of finite elements}
-In this development we have defined recursive anamorphisms on inductive data
-types. We might have as well defined them on the subtype of finite elements of
-coinductive data types using a predicate which states when an element of a
-coinductive data type is finite:
-\begin{coqcode}
-Inductive FinF : GFix F -> Prop :=
-| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
-\end{coqcode}
-Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
-is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
-easy to see. We first define a catamorphism \coq{ccata_f_} from the
-subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
-any \coq{F}-algebra.
-\begin{coqcode}
-Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
-  : forall x : GFix F, FinF x -> A := fix f x H :=
-    let hx := g_out x in
-      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
-\end{coqcode}
-We now prove this is isomorphic to the least fixed-point of the functor
-\coq{F}.
-We take the catamorphism from the finite elements of \coq{GFix F} to the
-inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
-inverse is the catamorphism on the restriction of \coq{g_in} to the finite
-elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
-prove the isomorphism:
-\begin{coqcode}
-Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
-Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
-\end{coqcode}
-The finite subtype of \coq{GFix F} allows us to compose
-catamorphisms and anamorphisms, by using the above isomorphism. In our work,
-however, we use \emph{recursive} anamorphisms, defined as \coq{hylo l_in c} for
-a recursive coalgebra \coq{c}, which compose easily with
-catamorphisms.
-% This predicate is equivalent to the proposition that
-% the anamorphism of such coalgebras produce only finite trees:
-% \begin{coqcode}
-% Lemma ana_rec_term `{setoid A} (h : Coalg F A)
-%   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
-% \end{coqcode}
-
-% \mpav{Moved here}
-
-% Note that such proofs will be erased when extracting code to OCaml, and this is
-% standard practice when building general recursive functions in
-% Coq\footnote{\dcas{This technique had a name in the community, but I cannot
-% remember. Something predicates}}. The core definitions are the following:
-% \begin{coqcode}
-% Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).
-
-% \end{coqcode}
-% Here, \coq{RecF_inv} provides the structurally smaller element for
-% the Coq recursive definition. We then couple coalgebras with a proof that they
-% are recursive, and add a coercion from recursive coalgebras to coalgebras.
-% \begin{coqcode}
-% Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
-% \end{coqcode}
-% In some cases, proving that coalgebras can only be applied finitely many
-% times is straightforward (e.g.\ they are recursive on some inductive type in
-% Coq). However, in many cases, proving that a coalgebra is recursive directly
-% can be tedious and cumbersome. To ease development, we prove that if a
-% given coalgebra respects a \emph{well-founded relation}, then it is
-% necessarily recursive.
-% \begin{coqcode}
-% Definition respects_relation `{setoid A} (c : Coalg F A)
-%       {B} (m : A -> B) (R : B -> B -> Prop)
-%   := forall x (p : Pos (shape (c x))), R (m (cont (c x) p)) (m x).
-
-% Lemma wf_coalg_rec `{setoid A} {B}
-%   (m : A -> B) (R : B -> B -> Prop) (WF : well_founded R)
-%   (c : Coalg F A) (RR : respects_relation c m R) : RecP c.
-% \end{coqcode}
-% In the above definition, \coq{m} is used as a form of ``metric''
-% for termination, and the well-founded relation is on \coq{B}. This
-% works equivalently to transporting a well-founded relation
-% \coq{R}
-% on
-% \coq{B} to \coq{A} as
-% \coq{R' := fun x y => R (m x) (m y)}, which is also trivially
-% well-founded. Using these definitions, we can construct recursive coalgebras by
-% proving that they respect a well-founded relation, which generally leads to
-% simpler proofs.
-
-% Similarly to the rest of the definitions, we prove that recursive anamorphisms
-% are respectful, and use this fact to build them as higher order proper
-% morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
-% the unique solution to the equation
-% \coq{f =e l_in \o fmap f \o h},
-% where \coq{h} is a recursive coalgebra.
+\subsection{Recursive Anamorphisms}
+Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
+inductive data type in the following way.  Intuitively, a recursive coalgebra
+can be applied only finitely many times, therefore when this is applied to an
+anamorphism the only possible traces it can produce from the seed function are
+the finite ones. We denote this special kind of recursion scheme \emph{recursive
+anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
+can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
+exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
+algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
+from the universality property of the hylomorphism and the fact that recursive
+anamorphisms satisfy the same equation.
+
+\subsection{Haskell Implementation of Recursion Schemes}
+\begin{minted}{haskell}
+newtype Fix f = In {inOp :: f (Fix f )}
+\end{minted}
+
+\begin{minted}{haskell}
+cata :: Functor f => (f a -> a) -> Fix f -> a
+cata alg = alg . fmap (cata alg) . In
+\end{minted}
+\begin{minted}{haskell}
+foldr :: (a -> b -> b) -> b -> [ a ] -> b
+foldr f e [ ]    = e
+foldr f e (x:xs) = f x (foldr f e xs)
+\end{minted}
+which is very useful in list processing. As a fold, The frequently used function map is also a fold:
+\begin{minted}{haskell}
+map :: (a -> b) -> [a] -> [b]
+map f = foldr (\x xs -> (f x):xs) []
+\end{minted}
+
+\begin{minted}{haskell}
+newtype CoFix f = OutOp {out :: f (Fix f )}
+\end{minted}
+
+\begin{minted}{haskell}
+ana :: Functor b => (a -> b a) -> a -> CoFix b
+ana coalg = OutOp . fmap (ana coalg) . coalg
+\end{minted}
+
+ and can be
+thought of as the $\operator{fmap}$ program in functional programming.
+\begin{minted}{haskell}
+class Functor f where
+  fmap :: (a -> b) -> f a -> f b
+\end{minted}
 
@@
 catamorphism.
+
+\subsubsection{On the subtype of finite elements}
+In this development we have defined recursive anamorphisms on inductive data
+types. We might have as well defined them on the subtype of finite elements of
+coinductive data types using a predicate which states when an element of a
+coinductive data type is finite:
+\begin{coqcode}
+Inductive FinF : GFix F -> Prop :=
+| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
+\end{coqcode}
+Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
+is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
+easy to see. We first define a catamorphism \coq{ccata_f_} from the
+subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
+any \coq{F}-algebra.
+\begin{coqcode}
+Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
+  : forall x : GFix F, FinF x -> A := fix f x H :=
+    let hx := g_out x in
+      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
+\end{coqcode}
+We now prove this is isomorphic to the least fixed-point of the functor
+\coq{F}.
+We take the catamorphism from the finite elements of \coq{GFix F} to the
+inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
+inverse is the catamorphism on the restriction of \coq{g_in} to the finite
+elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
+prove the isomorphism:
+\begin{coqcode}
+Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
+Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
+\end{coqcode}
+The finite subtype of \coq{GFix F} allows us to compose
+catamorphisms and anamorphisms, by using the above isomorphism. In our work,
+however, we use \emph{recursive} anamorphisms, defined as \coq{hylo l_in c} for
+a recursive coalgebra \coq{c}, which compose easily with
+catamorphisms.
 

### 2024-10-09 paper/paper.tex

--- 2024-10-09 paper/paper.tex
@@
 hylomorphisms in the Coq proof assistant, together with their algebraic laws.
-
-%Recursive definitions cannot be proven well-defined automatically due to the
-%halting problem. Modern proof assistants like Coq or Agda provide a sound, but
-%incomplete algorithm which syntactically checks for termination or
-%productivity. For recursion, this is done by automatically inferring
-%which argument in the recursive call gets smaller with respects to the original
-%input argument. For productivity, the algorithm checks that the corecursive
-%call appears directly under a constructor to make sure that this function
-%always produces at least one element after each recursive step.
-%This implies that some functions, though well-defined, cannot be accepted by
-%the proof assistant. Examples of this include common sorting algorithms, such
-%as quicksort:
-%%One example is Quicksort, written in OCaml
-%%below:
-%\begin{minted}{OCaml}
-%let rec qsort xs = match divide xs with | None -> []
-% | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
-%\end{minted}
-%While \ocaml{qsort} is a well-defined mathematical function it cannot be
-%accepted by a proof assistant. The reason is that the \ocaml{divide}
-%function \emph{destructuring} the input dives deeper in the input returning
-%two sublists with the head as a pivot.
-%
-%The main approach for implementing non-structural recursion in Coq is to use
-%\emph{well-founded recursion}, where recursive definitions are accompanied by
-%\emph{termination proofs}. Using well-founded recursion, the recursive calls
-%will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
-%one approach is using the \coq{Fix} combinator, in which recursion is done on an
-%\emph{accesibility} predicate on the input, which is a proof that there are no
-%infinitely decreasing chains.}.
-%A benefit of this strategy is that when extracting verified non-structural
-%recursive functions to OCaml, the additional proofs are discarded, resulting in
-%code that closely resembles what a programmer might have written directly in
-%OCaml. However, reasoning about program equivalences needs  such termination
-%proofs to be kept around, leading to the common practice of employing additional
-%reduction lemmas extensively.
-%
-%% A benefit of this approach is that, extracting
-%% verified nonstructural recursive functions to OCaml will erase the additional
-%% proofs, thus producing code that will be closer to what a programmer may have
-%% written directly in OCaml. However, reasoning about program equivalences
-%% requires dealing with such termination proofs, and it is common practice to use
-%% custom reduction lemmas that will be used extensively when proving properties
-%% about nonstructural recursive functions.
-
-%\emph{Structured recursion schemes}~\cite{HinzeW16, HinzeWG15} is a technique
-%for structuring recursive definitions that has found a lot of success in
-%functional programming. The idea is that, rather than writing a self-refential
-%definition, we use a function library, the recursion scheme, which takes as
-%input the non-recursive parts of the algorithm and returns the desired recursive
-%definition which is guaranteed to terminate by the properties of the recursion
-%scheme itself. More complex recursion schemes do exist but it has been
-%proven by Hinze et al.~\cite{HinzeWG15} that they are all instances of a divide-and-conquer
-%algorithm which goes under the name of hylomorphism~\cite{MeijerFP91,HuIT96}.
-%This result, thus, yields a general framework to reason about the algebraic properties of these
-%algorithms, e.g. short-cut fusion and semi-automatic parallelisations~\cite{TakanoM95,Gibbons96:Third, farmsCastro}.
-
-% A more complex recursion scheme is the hylomorphism~\cite{MeijerFP91, HuIT96} which has a divide-and-conquer
-% structure. It has been shown by Hinze et
-% al.~\cite{HinzeWG15} that any complex recursion scheme can be transmogrified
-% into an hylomorphisms by means of an adjunction. Furthermore, hylomorphism laws
-% can capture a number of useful equivalences, ranging from common optimisations
-% such as \emph{short-cut fusion}~\cite{TakanoM95}, to semi-automatic
-% parallelisations~\cite{Gibbons96:Third, farmsCastro}.
-
-% Recursion schemes have successfully been applied to functional programming to
-% structure recursive programs, however, because of the presence of
-% non-termination in languages like (e.g.) Haskell, they cannot guarantee
-% recursive definitions well-defined. % This is due to the possibility of applying a
-% %
-% recursion scheme to a non-terminating (co)algebra, but also to the unrestricted
-% % use of data functors.
-
-%However, when used in the context of languages with general recursion like
-%(e.g.) Haskell and ML, recursion schemes cannot ensure termination. This is
-%because the input information itself could be non-terminating. On the other
-%hand, while formalising recursion schemes in a type theory does provide stronger
-%termination guarantees, to the best of our knowledge, not many attempts have
-%been made at mechanising them.
 
@@
 
-For example, in $\Set$, we can define the set of lists
-$\List{A} \iso 1 + A \times \List{A}$ as the set inductively generated by the
+In $\Set$, we can define the set of lists
+\[
+  \List{A} \iso 1 + A \times \List{A}
+\]
+as the set inductively generated by the
 constructors $\nil : 1 \to \List{A}$ and

### 2024-10-09 paper/paper.tex
Section 3

--- 2024-10-09 paper/paper.tex
@@
 
-\usepackage{amsmath, mathpartir}
+\usepackage{amsmath, mathpartir, stmaryrd}
 %, amssymb, amsfonts, stmaryrd}
@@
 \newcommand{\Set}{\mathbf{Set}}
+\newcommand{\Type}{\mathbf{Type}}
 \newcommand{\iso}{\cong}
@@
 \end{coqcode}
+
 where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
@@
 \end{coqcode}
+
+\noindent
 One common workaround for this is the use of Mendler-style
@@
 \end{coqcode}
+
+\noindent
 While this definition solves the strict positivity problem, it does not solve
@@
 \paragraph{Containers} Another common solution is to the strict positivity
-problem is the use \emph{containers}~\cite{AbbottAG05}.  A container is defined
-by a type of \emph{shapes} and a \emph{family of position types} indexed by
-shapes.
-\begin{coqcode}
-(Shape : Type) (Pos : Shape -> Type)
-\end{coqcode}
-
-For the functor \coq{F} we introduced in the previous paragraph we would
-define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
-the data type and one of these contains a piece of data of type \coq{A}.
-Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
-and  \coq{Pos (inr a) = unit} indicating that the first constructor does not
-have any type variables and the second constructor has one type variable.
-
-At this point an \emph{extension} of this container is a functor defined as
-follows:
-\begin{coqcode}
-Record App (X : Type) := MkApp { shape : Shape; contents : Pos shape -> X }.
-\end{coqcode}
-with the obvious action on morphisms given by post-composition
-with \coq{contents}.
-\begin{coqcode}
-Definition fmap (f : A -> B) (x : App C A) : App C B
-  := {| shape := shape x; contents := fun e => f (contents x e) |}
-\end{coqcode}
-In our running example, the type \coq{App X}, for some \coq{X} has two
-inhabitants.  The first is a pair composed by \coq{inl tt : Shape} and a
-function \coq{Pos (inl tt) -> X}. This latter is in fact a function of
-type \coq{Empty_set -> X}. This pair is, therefore isomorphic to the
-type \coq{unit}, the type with only one inhabitant.  The second inhabitant is
-the pair composed by \coq{inr a : Shape}, for some \coq{a : A}, and a
-function \coq{Pos (inr a) -> X}.  This latter type is equal to \coq{unit -> X}
-which represents the elements of \coq{X} and therefore is isomorphic to \coq{X}.
-This particular pair of shapes and positions is  therefore isomorphic to the
-type \coq{unit + A * X}.
-
-The correspondence between containers and polynomial functors has been
-formalised in this work and can be found in the accompanying code (file
-Container.v). 
+problem is the use \emph{containers}~\cite{AbbottAG05}.  A container
+$S \triangleright P$ is defined by a type of \emph{shapes} $S : \Type$ and a
+\emph{family of position types} indexed by shapes $P : S \to \Type$.  An
+\emph{extension} of this container is a functor
+$\llbracket S \triangleright P \rrbracket$, whose action on objects is given by
+$\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
+and on morphisms is given by post-composition, keeping the shape the same
+$\llbracket S \triangleright P \rrbracket \; f = \lambda (s,\;g).\;(s,\;f\circ g)$.
+
+To explain the role of shapes and positions in containers, consider the functor
+$F\; X = 1 + A \times X$.  An equivalent container, i.e.\ a container with an
+extension that is isomorphic to $F$ requires a shape that can distinguish two
+cases, indicating that there are two constructors to build an object of type
+$F\;X$. For example, we could define $S_{F} = 1 + A$. We need to consider two
+cases for the family of positions:  $P_{F}(\mathsf{inj}_{1}\;*) = 0$, because
+there are no occurrences of $X$ in the left case; and
+$P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence of $X$ in the
+right hand side. It is easy to see that
+$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.
+
+In general, every polynomial functor has an equivalent container, with an
+isomorphic extension. We have mechanised this result, and this can  be found in
+the accompanying code (file Container.v).
 % 
@@
 \subsection{Extractable Containers}
-In the previous section we defined containers using dependent types.  However,
-Ocaml's type system is not equipped to handle these. To solve this problem
-Coq's code extraction mechanism will insert unsafe casts.
+A straightforward encoding of containers in Coq would use dependent types  to
+represent families of positions.  However, Ocaml's type system is not equipped
+to handle these, which will lead to extracting OCaml code with unsafe casts.
 
 Consider instead representing the positions of a container using a decidable
-validity predicate (\coq{valid}) which assigns shapes to positions. We use
-boolean functions and coercions from \coq{bool} to \coq{Prop} to represent
-decidable predicates, similarly to SSReflect~\cite{GonthierL09}. The family of
-positions for a given shape in a container can be recovered by defining:
-\begin{coqcode}
-Class Cont (Shape : Type) (APos : Type) := { valid : Shape -> APos -> bool }.
-Record Pos `(C : Cont Shape APos) (shape : Shape) :=
-  ValidPos { val : APos; IsValid : valid shape pos }.
+validity predicate which assigns shapes to positions. In Coq, we use
+a boolean function and a coercion from \coq{bool} to \coq{Prop} to represent
+the decidable predicate \coq{valid}, similarly to SSReflect~\cite{GonthierL09}. Given
+a \coq{Shape : Type}, and a type of \emph{all possible positions} \coq{APos : Type},
+the family of
+positions for a given shape can be defined as a dependent pair:
+\begin{coqcode}
+Definition Pos (s : Shape) = { p : APos | valid s p }.
 \end{coqcode}
@@
 \begin{coqcode}
- k : Pos C s -> X
+ k : Pos s -> X
  P1, P2 : valid s p
  ---------------------------
- k (ValidPos p P1) = k (ValidPos p P2)
+ k (existT _ p P1) = k (existT _ p P2)
 \end{coqcode}
@@
 
-% A container extension is then defined as follows:
-% \begin{coqcode}
-% Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
-% Record App (X : Type) := { shape    : Shape;
-%                           contents : {p | Valid shape p} -> X }.
+A container extension, \coq{App} in Coq, is then defined as follows:
+\begin{coqcode}
+Record App (X : Type)
+   := { shape : Shape; contents : {p | valid shape p} -> X }.
 % \end{coqcode}
-% This approach solves the problem of the unsafe casts, since the code extracted
-% for \coq{{p | Valid shape pos}} will be an OCaml singleton type defined as the
-% OCaml equivalent to \coq{Pos}.
+
+This approach solves the problem of the unsafe casts, since the code extracted
+for \coq{{p | valid shape p}} will be an OCaml singleton type defined as the
+OCaml equivalent to \coq{Pos}.
 
@@
 \begin{coqcode}
-Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
-| AppR_ext (Es : shape x = shape y)
-           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
+Inductive AppR (x y : App F X) : Prop :=
+| AppR_ext (shape x = shape y)
+           (forall p1 p2, projT1 e1 = projT1 e2 -> contents p1 = contents p2).
 \end{coqcode}
@@
 In particular, we want to reason about the equality of functions of types such
-as \coq{App F A -> B} (or \coq{B -> App F A}). Since 
+as \coq{App A -> B} (or \coq{B -> App A}). Since
 these types now come with their own equivalence, any function that manipulates
@@
 Coq's \coq{inl} and \coq{inr} constructors, and
-pattern matching on them uses the function 
-\coq{fun f g x => match x with | inl y => f y | inr y => g y end}.
+pattern matching on them uses the function:
+\begin{coqcode}
+fun f g x => match x with | inl y => f y | inr y => g y end
+\end{coqcode}
 The proofs that these morphisms are proper are straightforward.  Finally, we

### 2024-10-09 paper/paper.tex
On Hylomorphism

--- 2024-10-09 paper/paper.tex
@@
 purposes we shall not deal with size issues and assume all the categories are
-small.}. The initial object in a category, denoted by $0$, is the object such
+locally small.}. The initial object in a category, denoted by $0$, is the object such
 that for any other object $A$ there is a unique arrow $0 \xrightarrow{!} A$.
@@
 The terminal object in $\Set$ is the \emph{unit} type, that only has one
-inhabitant, $* : 1$, analogous to the unit type in Haskell: \haskell{() :: ()}.
+inhabitant, $* : 1$.
 The $\List{-}$ type is a functor. Its action on objects is to take any set
@@
 \label{sec:algebras}
-For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(A,\alpha)$
-where $A$ is an object of the category called the \emph{carrier} of the algebra
-and $\alpha$ is an arrow of type $FA \to A$ called the \emph{structure map}.
+For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(X,a_{X})$
+where $X$ is an object of the category called the \emph{carrier} of the algebra
+and $a_{X}$ is an arrow of type $FX \to X$ called the \emph{structure map}.
 The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
-objects are $F$-algebras and arrows $f : (A, \alpha) \to (B, \beta)$ are
-\emph{$F$-algebra homomorphisms}  $f : A \to B$ in $\CatC$ such that they
+objects are $F$-algebras and arrows $f : (X, a_{X}) \to (Y, a_{Y})$ are
+\emph{$F$-algebra homomorphisms}  $f : X \to Y$ in $\CatC$ such that they
 respect the structure of the algebra, that is
-$f \circ \alpha = \beta \circ F(f)$. The initial object in this category is
+$f \circ a_{X} = A_{Y} \circ F(f)$. The initial object in this category is
 called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
@@
 $\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
-isomorphism $F\mu F \iso \mu F$.
-By the property of initial $F$-algebras, for
-any other $F$-algebra $(B, \beta)$ there exists a unique $F$-algebra
-homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (B, \beta)$ and
+isomorphism $F\mu F \iso \mu F$.  By the property of initial $F$-algebras, for
+any other $F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra
+homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (X, a_{X})$ and
 pronounced ``\emph{catamorphism}'' satisfying:
@@
   \label{eq:cata:unique}
-  f = \cata{\alpha} \iff f = \alpha \circ F(f) \circ \InIso
+  f = \cata{a_{X}} \iff f = a_{X} \circ F(f) \circ \InIso
 \end{equation}
 In words, any $F$-algebra homomorphism from the initial $F$-algebra is a
-catamorphism.
+catamorphism. As a result of the uniqueness property we can derive the fusion
+law. For all $F$-algebra homomorphisms $f : (X, a_{X}) \to (Y, a_{Y})$ we have
+\begin{equation}
+  \label{eq:cata:fusion}
+  f \circ \cata{a_{X}} = \cata{a_{Y}}
+\end{equation}
+which means that the composition of a program $f$ with a catamorphism recursing
+once over the data structure is the same as performing that recursion once using
+the algebra $a_{Y}$ instead of $a_{X}$. This is a useful result for program
+optimisation as we shall see. \mpav{We shall see? }
 
@@
 $X \mapsto 1 + A \times X$. An $F$-algebra is a set $B$ together with a
-structure map $[base, ind] : 1 + A \times B \to B$. The initial $F$-algebra is
+structure map $[b, i] : 1 + A \times B \to B$. The initial $F$-algebra is
 clearly the set of lists $\List{A}$. The catamorphism associated with the type
 of lists is the unique arrow recursively translating the initial algebra
-$[\nil,\cons]$ into the algebra $[base, ind]$:
-
-In functional programming this is commonly referred to as
+$[\nil,\cons]$ into the algebra $[b, i]$.
+
+\mpav{TODO}In functional programming this is commonly referred to as
 $\foldr : (1 \to B) \to (A \times B \to B) \to B$. We can in fact set
@@
 The dual of an algebra is a coalgebra. For an endofunctor $B : \CatC \to \CatC$,
-a $B$-\emph{coalgebra} is a pair $(X, c)$ where $X \in \Obj(\CatC)$ is the
-carrier of the coalgebra and $c : X \to BX$ is a morphism.
+a $B$-\emph{coalgebra} is a pair $(X, c_{X})$ where $X \in \Obj(\CatC)$ is the
+carrier of the coalgebra and $c_{X} : X \to BX$ is a morphism.
 
@@
 implementing the \emph{transition system} with $BX = L \times X$.  In
-particular, for a state $x \in X$, $c(x_{1})$ returns a pair $(l, x_{2})$ where
-$l \in L$ is the observable action and $x_{2} \in X$ is the next state.  In this
-case, we can even set some notation for the transition map:
+particular, for a state $x_{1} \in X$, $c(x_{1})$ returns a pair $(l, x_{2})$
+where $l \in L$ is the observable action and $x_{2} \in X$ is the next state.
+In this case, we can even set some notation for the transition map:
 \[
@@
 denoted by $(\nu B, \OutIso)$ where $\OutIso$ denotes the final $B$-coalgebra
-$\OutIso : \nu B \to B\nu B$. For
+$\OutIso : \nu B \to B\nu B$.
 
@@
 coalgebra $(\nu B, \OutIso)$ which is denoted by $\ana{\gamma}$ and pronounced
-\emph{anamorphism}. We spell out the uniqueness property:
+``\emph{anamorphism}''. We spell out the uniqueness property:
 \begin{equation}
@@
 homomorphism $f$ into the terminal object then it must be the anamorphism on the
-same coalgebra.
+same coalgebra. As a result of the uniqueness property we can derive the fusion
+laws. For all $B$-coalgebra homomorphisms
+$f : (X, \gamma_{X}) \to (Y, \gamma_{Y})$ we have
+\begin{equation}
+  \label{eq:ana:fusion}
+  \ana{\gamma_{Y}} \circ f = \ana{\gamma_{X}}
+\end{equation}
 
@@
 
-In general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
-unique solution (when it exists) to the equation
+Let $(A,a)$ be an $F$-algebra and $(C,c)$ be an $F$-coalgebra. An arrow
+$C \to A$ is an \emph{hylomorphism}, written $h : (C,c) \rightarrowtail (A,a)$ if
+it satisfies
 \begin{equation}
   \label{eq:hylo}
-  f = a \circ F\; f \circ c
+  h = a \circ F(h) \circ c
 \end{equation}
@@
 
-In order to find a solution we restrict ourselves to the so-called
-\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.
-A $F$-coalgebra $(X, c)$ is recursive if
-
-An example of a
-recursive coalgebra is the \coq{partition} function in quicksort which
-destructures a list into a pivot and two sublist and as long as the sublists are
-smaller the partitioning function still yields a unique solution to the
-recursion scheme.
-
-\begin{coqcode}
-Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
-  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
-\end{coqcode}
+A coalgebra $(C,c)$ is \emph{recursive} if for every algebra $(A, a)$ there is a
+\emph{unique} hylo $(C,c ) \rightarrowtail (A, a)$. We denote these type of
+hylos by $\hylo{c}{a}$.
+
+An example of a recursive coalgebra is the \coq{partition} function in quicksort
+which destructures a list into a pivot and two sublist and as long as the
+sublists are smaller the partitioning function still yields a unique solution to
+the recursion scheme.
+The uniqueness property of the hylo yields the following fusion law:
+\begin{equation}
+  \label{eq:hylo:fusion}
+  f \circ \hylo{c}{a}= \hylo{c'}{a}
+\end{equation}
 Using the hylo fusion law, we can prove the well-known \emph{deforestation
@@
 anamorphism.
+
+$(A, c_{1})$ is an recursive coalgebra
+$(B, c_{2})$ is an recursive coalgebra
+
+$(C, a_{2})$ is an algebra
+$(B, a_{1})$ is an algebra
+
+$c_{2} \circ a_{1} = id$
+
+\begin{equation}
+  \label{eq:hylo:deforestation}
+  \hylo{a_{2}}{c_{2}} \circ \hylo{a_{1}}{c_{1}} = \hylo{a_{2}}{c_{1}}
+\end{equation}
+
 \begin{coqcode}
@@
 Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
-inductive data type in the following way.  Intuitively, a recursive coalgebra
+inductive data type in the following way. A recursive coalgebra
 can be applied only finitely many times, therefore when this is applied to an
@@
 the finite ones. We denote this special kind of recursion scheme \emph{recursive
-anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
-can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
-exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
-algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
+anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
+can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
+exactly hylomorphisms on the recursive $F$-coalgebra and the
+algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
 from the universality property of the hylomorphism and the fact that recursive
@@
 newtype Fix f = In {inOp :: f (Fix f )}
+\end{minted}
+
+\begin{minted}{haskell}
+data ListF a x = Nil | Cons a x deriving Functor
+data TreeF a x = Empty | Node x a x deriving Functor
 \end{minted}

### 2024-10-09 paper/paper.tex
Removed dump

@@
 \end{minted}
-
-\subsection{Dump}
-Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
-by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of type
-\coq{App F A ~> A}:
-\begin{coqcode}
-Notation Alg F A := (App F A ~> A).
-\end{coqcode}
-The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
-where the structure map is an isomorphism. This is sometimes referred to as the
-\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind
-this name shortly. \footnote[1]{To be more precise, initial algebras are
-isomorphisms by Lambek's lemma, but isomorphisms of functors do not necessarily
-correspond to initial algebras.}  We define the least fixed-point of \coq{F} as
-an inductive type:
-\begin{coqcode}
-Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
-\end{coqcode}
-
-where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
-As an example, the initial \coq{F}-algebra for the
-functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
-being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
-operation \coq{Cons : A * LFix F -> LFix F}.
-
-We define \coq{LFix} as a setoid, where its equivalence relation can be
-described as the least fixed point of \coq{AppR} and we define smart
-constructors for the isomorphism of least fixed points as respectful morphisms:
-\begin{coqcode}
-l_in : App F (LFix F) ~> LFix F                l_out : LFix F ~> App F (LFix F)
-\end{coqcode}
-The least fixed-point for \coq{F} is the initial \coq{F}-algebra in the sense
-that it gives rise to an inductive recursion scheme.  Specifically, for any
-other \coq{F}-algebra there exists a \emph{unique} map, known as a \emph{fold} or
-\emph{catamorphism}, such that it structurally deconstructs the data type
-using \coq{LFix_out}, calls itself recursively and then composes the result of
-the recursive call using the given an \coq{F}-algebra. In other words, for any
-give \coq{F}-algebra there exists a unique \coq{F}-algebra homomorphism from the
-initial one. We define it in Coq as follows:
-\begin{coqcode}
-  Definition cata_f `{F : Cont Sh P} (alg : Alg F A) : LFix -> A
-:= fix f (x : LFix) := match x with
-  | LFix_in ax => let (sx, kx) := ax in alg (MkApp sx (fun e => f (kx e))) end.
-\end{coqcode}
-It is easy to show that this function is a respectful morphism of
-\coq{F}-algebras. In fact, it is possible to define it as a map of the
-following type:
-\begin{coqcode}
-  cata : forall `{F : Cont Sh P} `{setoid A}, Alg F A ~> LFix ~> A
-\end{coqcode}
-We prove that catamorphisms satisfy the universality property we explained previously:
-\begin{coqcode}
-Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
-    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
-\end{coqcode}
-In other words, if there is any other \coq{f} with the same structural recursive
-shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
-catamorphism.
-
-\subsubsection{On the subtype of finite elements}
-In this development we have defined recursive anamorphisms on inductive data
-types. We might have as well defined them on the subtype of finite elements of
-coinductive data types using a predicate which states when an element of a
-coinductive data type is finite:
-\begin{coqcode}
-Inductive FinF : GFix F -> Prop :=
-| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
-\end{coqcode}
-Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
-is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
-easy to see. We first define a catamorphism \coq{ccata_f_} from the
-subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
-any \coq{F}-algebra.
-\begin{coqcode}
-Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
-  : forall x : GFix F, FinF x -> A := fix f x H :=
-    let hx := g_out x in
-      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
-\end{coqcode}
-We now prove this is isomorphic to the least fixed-point of the functor
-\coq{F}.
-We take the catamorphism from the finite elements of \coq{GFix F} to the
-inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
-inverse is the catamorphism on the restriction of \coq{g_in} to the finite
-elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
-prove the isomorphism:
-\begin{coqcode}
-Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
-Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
-\end{coqcode}
-The finite subtype of \coq{GFix F} allows us to compose
-catamorphisms and anamorphisms, by using the above isomorphism. In our work,
-however, we use \emph{recursive} anamorphisms, defined as \coq{hylo l_in c} for
-a recursive coalgebra \coq{c}, which compose easily with
-catamorphisms.
 

### 2024-10-09 paper/paper.tex
Edits to Section 3

--- 2024-10-09 paper/paper.tex
@@
 purposes we shall not deal with size issues and assume all the categories are
-locally small.}. The initial object in a category, denoted by $0$, is the object such
+small.}. The initial object in a category, denoted by $0$, is the object such
 that for any other object $A$ there is a unique arrow $0 \xrightarrow{!} A$.
@@
 The terminal object in $\Set$ is the \emph{unit} type, that only has one
-inhabitant, $* : 1$.
+inhabitant, $* : 1$, analogous to the unit type in Haskell: \haskell{() :: ()}.
 The $\List{-}$ type is a functor. Its action on objects is to take any set
@@
 \label{sec:algebras}
-For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(X,a_{X})$
-where $X$ is an object of the category called the \emph{carrier} of the algebra
-and $a_{X}$ is an arrow of type $FX \to X$ called the \emph{structure map}.
+For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(A,\alpha)$
+where $A$ is an object of the category called the \emph{carrier} of the algebra
+and $\alpha$ is an arrow of type $FA \to A$ called the \emph{structure map}.
 The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
-objects are $F$-algebras and arrows $f : (X, a_{X}) \to (Y, a_{Y})$ are
-\emph{$F$-algebra homomorphisms}  $f : X \to Y$ in $\CatC$ such that they
+objects are $F$-algebras and arrows $f : (A, \alpha) \to (B, \beta)$ are
+\emph{$F$-algebra homomorphisms}  $f : A \to B$ in $\CatC$ such that they
 respect the structure of the algebra, that is
-$f \circ a_{X} = A_{Y} \circ F(f)$. The initial object in this category is
+$f \circ \alpha = \beta \circ F(f)$. The initial object in this category is
 called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
@@
 $\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
-isomorphism $F\mu F \iso \mu F$.  By the property of initial $F$-algebras, for
-any other $F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra
-homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (X, a_{X})$ and
+isomorphism $F\mu F \iso \mu F$.
+By the property of initial $F$-algebras, for
+any other $F$-algebra $(B, \beta)$ there exists a unique $F$-algebra
+homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (B, \beta)$ and
 pronounced ``\emph{catamorphism}'' satisfying:
@@
   \label{eq:cata:unique}
-  f = \cata{a_{X}} \iff f = a_{X} \circ F(f) \circ \InIso
+  f = \cata{\alpha} \iff f = \alpha \circ F(f) \circ \InIso
 \end{equation}
 In words, any $F$-algebra homomorphism from the initial $F$-algebra is a
-catamorphism. As a result of the uniqueness property we can derive the fusion
-law. For all $F$-algebra homomorphisms $f : (X, a_{X}) \to (Y, a_{Y})$ we have
-\begin{equation}
-  \label{eq:cata:fusion}
-  f \circ \cata{a_{X}} = \cata{a_{Y}}
-\end{equation}
-which means that the composition of a program $f$ with a catamorphism recursing
-once over the data structure is the same as performing that recursion once using
-the algebra $a_{Y}$ instead of $a_{X}$. This is a useful result for program
-optimisation as we shall see. \mpav{We shall see? }
+catamorphism.
 
@@
 $X \mapsto 1 + A \times X$. An $F$-algebra is a set $B$ together with a
-structure map $[b, i] : 1 + A \times B \to B$. The initial $F$-algebra is
+structure map $[base, ind] : 1 + A \times B \to B$. The initial $F$-algebra is
 clearly the set of lists $\List{A}$. The catamorphism associated with the type
 of lists is the unique arrow recursively translating the initial algebra
-$[\nil,\cons]$ into the algebra $[b, i]$.
-
-\mpav{TODO}In functional programming this is commonly referred to as
+$[\nil,\cons]$ into the algebra $[base, ind]$:
+
+In functional programming this is commonly referred to as
 $\foldr : (1 \to B) \to (A \times B \to B) \to B$. We can in fact set
@@
 The dual of an algebra is a coalgebra. For an endofunctor $B : \CatC \to \CatC$,
-a $B$-\emph{coalgebra} is a pair $(X, c_{X})$ where $X \in \Obj(\CatC)$ is the
-carrier of the coalgebra and $c_{X} : X \to BX$ is a morphism.
+a $B$-\emph{coalgebra} is a pair $(X, c)$ where $X \in \Obj(\CatC)$ is the
+carrier of the coalgebra and $c : X \to BX$ is a morphism.
 
@@
 implementing the \emph{transition system} with $BX = L \times X$.  In
-particular, for a state $x_{1} \in X$, $c(x_{1})$ returns a pair $(l, x_{2})$
-where $l \in L$ is the observable action and $x_{2} \in X$ is the next state.
-In this case, we can even set some notation for the transition map:
+particular, for a state $x \in X$, $c(x_{1})$ returns a pair $(l, x_{2})$ where
+$l \in L$ is the observable action and $x_{2} \in X$ is the next state.  In this
+case, we can even set some notation for the transition map:
 \[
@@
 denoted by $(\nu B, \OutIso)$ where $\OutIso$ denotes the final $B$-coalgebra
-$\OutIso : \nu B \to B\nu B$.
+$\OutIso : \nu B \to B\nu B$. For
 
@@
 coalgebra $(\nu B, \OutIso)$ which is denoted by $\ana{\gamma}$ and pronounced
-``\emph{anamorphism}''. We spell out the uniqueness property:
+\emph{anamorphism}. We spell out the uniqueness property:
 \begin{equation}
@@
 homomorphism $f$ into the terminal object then it must be the anamorphism on the
-same coalgebra. As a result of the uniqueness property we can derive the fusion
-laws. For all $B$-coalgebra homomorphisms
-$f : (X, \gamma_{X}) \to (Y, \gamma_{Y})$ we have
-\begin{equation}
-  \label{eq:ana:fusion}
-  \ana{\gamma_{Y}} \circ f = \ana{\gamma_{X}}
-\end{equation}
+same coalgebra.
 
@@
 
-Let $(A,a)$ be an $F$-algebra and $(C,c)$ be an $F$-coalgebra. An arrow
-$C \to A$ is an \emph{hylomorphism}, written $h : (C,c) \rightarrowtail (A,a)$ if
-it satisfies
+In general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
+unique solution (when it exists) to the equation
 \begin{equation}
   \label{eq:hylo}
-  h = a \circ F(h) \circ c
+  f = a \circ F\; f \circ c
 \end{equation}
@@
 
-A coalgebra $(C,c)$ is \emph{recursive} if for every algebra $(A, a)$ there is a
-\emph{unique} hylo $(C,c ) \rightarrowtail (A, a)$. We denote these type of
-hylos by $\hylo{c}{a}$.
-
-An example of a recursive coalgebra is the \coq{partition} function in quicksort
-which destructures a list into a pivot and two sublist and as long as the
-sublists are smaller the partitioning function still yields a unique solution to
-the recursion scheme.
-The uniqueness property of the hylo yields the following fusion law:
-\begin{equation}
-  \label{eq:hylo:fusion}
-  f \circ \hylo{c}{a}= \hylo{c'}{a}
-\end{equation}
+In order to find a solution we restrict ourselves to the so-called
+\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.
+A $F$-coalgebra $(X, c)$ is recursive if
+
+An example of a
+recursive coalgebra is the \coq{partition} function in quicksort which
+destructures a list into a pivot and two sublist and as long as the sublists are
+smaller the partitioning function still yields a unique solution to the
+recursion scheme.
+
+\begin{coqcode}
+Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
+  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
+\end{coqcode}
 Using the hylo fusion law, we can prove the well-known \emph{deforestation
@@
 anamorphism.
-
-$(A, c_{1})$ is an recursive coalgebra
-$(B, c_{2})$ is an recursive coalgebra
-
-$(C, a_{2})$ is an algebra
-$(B, a_{1})$ is an algebra
-
-$c_{2} \circ a_{1} = id$
-
-\begin{equation}
-  \label{eq:hylo:deforestation}
-  \hylo{a_{2}}{c_{2}} \circ \hylo{a_{1}}{c_{1}} = \hylo{a_{2}}{c_{1}}
-\end{equation}
-
 \begin{coqcode}
@@
 Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
-inductive data type in the following way. A recursive coalgebra
+inductive data type in the following way.  Intuitively, a recursive coalgebra
 can be applied only finitely many times, therefore when this is applied to an
@@
 the finite ones. We denote this special kind of recursion scheme \emph{recursive
-anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
-can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
-exactly hylomorphisms on the recursive $F$-coalgebra and the
-algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
+anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
+can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
+exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
+algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
 from the universality property of the hylomorphism and the fact that recursive
@@
 newtype Fix f = In {inOp :: f (Fix f )}
-\end{minted}
-
-\begin{minted}{haskell}
-data ListF a x = Nil | Cons a x deriving Functor
-data TreeF a x = Empty | Node x a x deriving Functor
 \end{minted}
@@
 is suitable for code extraction. Recall from Section~\ref{sec:rec-schemes} that
-we need to be able toabstract away from the particular shape of the data, and
-that this is achieved by introducing the concept of \emph{functors} which have
-the suitable fixed-point properties, i.e.\ those which have a initial algebras.
+we need to be able to abstract away from particular shape of the data, and we
+can do this using \emph{functors} that have suitable fixed-point properties;
+i.e.\ those which have a initial algebras.
+%
 A common approach to construct such functors is to use
-\emph{containers}~\cite{AbbottAG05}.  However, reasoning about container
-equality will require us to consider both functional extensionality and
-heterogeneous equality. We avoid these axioms by introducing a custom
-equivalence relation on types (Section~\ref{sec:types-morphisms}).
+\emph{containers}~\cite{AbbottAG05}. However, reasoning about container equality
+will require us to consider both functional extensionality and heterogeneous
+equality. We avoid these axioms by introducing a custom equivalence relation on
+types (Section~\ref{sec:types-morphisms}).
 
@@
 
-\paragraph{Containers} Another common solution is to the strict positivity
+% \paragraph{Containers}
+Another common solution is to the strict positivity
 problem is the use \emph{containers}~\cite{AbbottAG05}.  A container
@@
 $F\; X = 1 + A \times X$.  An equivalent container, i.e.\ a container with an
-extension that is isomorphic to $F$ requires a shape that can distinguish two
+extension that is isomorphic to $F$, requires a shape that can distinguish two
 cases, indicating that there are two constructors to build an object of type
@@
 In general, every polynomial functor has an equivalent container, with an
-isomorphic extension. We have mechanised this result, and this can  be found in
-the accompanying code (file Container.v).
+isomorphic extension, and we have mechanised this result.
+% and this can  be found in the accompanying code (file Container.v).
 % 
@@
 A straightforward encoding of containers in Coq would use dependent types to
-represent families of positions.  However, Ocaml's type system is not equipped
-to handle these, which will lead to extracting OCaml code with unsafe casts.
+represent families of positions. However, Ocaml's type system is not equipped to
+handle these, which will lead to extracting OCaml code with unsafe casts.
 
@@
 use of the functional extensionality axiom, we capture this relation with the
-following inductive proposition in Coq:
+following inductive proposition:
 \begin{coqcode}
@@
 Note that we do not care about the validity proof of the positions, only their
-value. This is to simplify (slightly) our proofs. This relation is
-trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.
+value. This is to simplify (slightly) our proofs. This relation is trivially
+\emph{reflexive}, \emph{transitive}, and \emph{symmetric}.
 
@@
 In particular, we want to reason about the equality of functions of types such
-as \coq{App A -> B} (or \coq{B -> App A}). Since
-these types now come with their own equivalence, any function that manipulates
-them needs to be \emph{respectful}. I.e.\ given 
+as \coq{App A -> B} (or \coq{B -> App A}).
+%
+Since these types now come with their own equivalence, any function that
+manipulates them needs to be \emph{respectful}; i.e.\ given
 \coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
@@
 We prove the isomorphisms of polynomial types and the equivalent 
-container extensions. As an example, we discuss (informally) the isomorphisms
+container extensions. As an example, we will consider the isomorphisms
 of pairs with their equivalent container extensions. Suppose that we

### 2024-10-09 paper/paper.tex
small changes to notation

--- 2024-10-09 paper/paper.tex
@@
 \[
-  x \xrightarrow{l} y \eqdef c(x) = (l, y)
+  x \xrightarrow{l} y \eqdef c_{X}(x) = (l, y)
 \]
 The category of $B$-coalgebras, denoted $B\Coalg$, is the category where objects
-are $B$-coalgebras and morphisms $f : (X, \gamma_{X}) \to (Y, \gamma_{Y})$ are
+are $B$-coalgebras and morphisms $f : (X, c_{X}) \to (Y, c_{Y})$ are
 $B$-coalgebra homomorphisms $f : A \to B$, that is
-$ \gamma_{Y} \circ f = F(f) \circ \gamma_{X}$. Terminal object in this category
+$ c_{Y} \circ f = F(f) \circ c_{X}$. Terminal object in this category
 $B$-coalgebra corresponds to the greatest fixed-point for the functor $B$,
@@
 $\Str{A} \iso A \times \Str{A}$.  As a result, for any $B$-coalgebra
-$(B, \gamma)$ there exists a unique $B$-coalgebra homomorphism into the terminal
-coalgebra $(\nu B, \OutIso)$ which is denoted by $\ana{\gamma}$ and pronounced
+$(X, c_{X})$ there exists a unique $B$-coalgebra homomorphism into the terminal
+coalgebra $(\nu B, \OutIso)$ which is denoted by $\ana{c_{X}}$ and pronounced
 ``\emph{anamorphism}''. We spell out the uniqueness property:
@@
   \label{eq:ana:unique}
-  f = \ana{\gamma} \iff f = \OutOp \circ B(f) \circ \gamma
+  f = \ana{c_{X}} \iff f = \OutOp \circ B(f) \circ c_{X}
 \end{equation}
@@
 laws. For all $B$-coalgebra homomorphisms
-$f : (X, \gamma_{X}) \to (Y, \gamma_{Y})$ we have
+$f : (X, c_{X}) \to (Y, c_{Y})$ we have
 \begin{equation}
   \label{eq:ana:fusion}
-  \ana{\gamma_{Y}} \circ f = \ana{\gamma_{X}}
+  \ana{c_{Y}} \circ f = \ana{c_{X}}
 \end{equation}

### 2024-10-10 paper/paper.tex
Fixes to hylo laws

@@
 
+\usepackage{inconsolata}
 % \usepackage{draftwatermark}
@@
 the recursion scheme.
-The uniqueness property of the hylo yields the following fusion law:
-\begin{equation}
+The uniqueness property of the hylo yields the following fusion laws:
+\begin{align}
   \label{eq:hylo:fusion}
-  f \circ \hylo{c}{a}= \hylo{c'}{a}
-\end{equation}
-Using the hylo fusion law, we can prove the well-known \emph{deforestation
-optimisation}. This is when two consecutive recursive computations, one that
-builds a data structure, and another one that consumes it, can be fused together
-into a single recursive definition. This, in turn, allows us to prove that a
-recursive hylomorphism is the composition of a catamorphism and a recursive
-anamorphism.
-
-$(A, c_{1})$ is an recursive coalgebra
-$(B, c_{2})$ is an recursive coalgebra
+  f \circ \hylo{c}{a} & = \hylo{c}{a'}
+  \hspace{1cm}
+  \Longleftarrow
+  \hspace{1cm}
+  f \circ a = a' \circ F(f) \\
+  \hylo{c}{a} \circ f & = \hylo{c'}{a}
+  \hspace{1cm}
+  \Longleftarrow
+  \hspace{1cm}
+  c \circ f = F(f) \circ c'
+\end{align}
+Using the hylo fusion laws, we can prove the well-known \emph{deforestation
+optimisation}, also known as the \emph{composition
+law}~\cite{DBLP:conf/ifl/HinzeHJ10}. This is when two consecutive recursive
+computations, one that builds a data structure, and another one that consumes
+it, can be fused together into a single recursive definition. This, in turn,
+allows us to prove that a recursive hylomorphism is the composition of a
+catamorphism and a recursive anamorphism.
+
+$(A, c_{1})$ is a recursive coalgebra
+$(B, c_{2})$ is a recursive coalgebra
 
@@
 \end{equation}
-
-\begin{coqcode}
-Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
-  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
-  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
-\end{coqcode}
 
@@
 right hand side. It is easy to see that
-$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.
-
-In general, every polynomial functor has an equivalent container, with an
-isomorphic extension, and we have mechanised this result.
-% and this can  be found in the accompanying code (file Container.v).
+$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.  In general, every
+polynomial functor has an equivalent container, and we have mechanised this
+result.
 %
@@
 \begin{coqcode}
-Structure morph A {eA : setoid A} B {eB : setoid B} :=
+Record morph A {eA : setoid A} B {eB : setoid B} :=
   MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
@@
 
-This definition leads to the well-known ``setoid hell'', which we mitigate by
-providing tactics and notations to automatically discharge proofs of
-\coq{app_eq} for morphisms, whenever the types use the standard propositional
-equality, or a combination of propositional and extensional equality. However,
-our compositional approach allows us to build morphisms by plugging in other
-morphisms to our combinators. In our framework, our expectation is that the
-user-provided functions remain small, with relatively straightforward proofs of
-\coq{app_eq}.
-
-However, by using this mechanisation, we gain simplified proofs via the use of
+Our definition of morphisms, and the use of different equivalences leads to the
+well-known ``setoid hell''. We mitigate this problem by providing tactics and
+notations to automatically discharge proofs of \coq{app_eq} for morphisms,
+whenever the types use the standard propositional equality, or a combination of
+propositional and extensional equality. However, our compositional approach
+allows us to build morphisms by plugging in other morphisms to our combinators.
+In our framework, our expectation is that the user-provided functions remain
+small, with relatively straightforward proofs of \coq{app_eq}.
+
+However, by using this mechanisation, we gain simplified proofs via
 Coq's \emph{Generalised Rewriting}. Since every morphism \coq{f : A ~> B}

### 2024-10-10 paper/paper.tex
Fixes

@@
 
-A container extension, \coq{App} in Coq, is then defined as follows:
-\begin{coqcode}
-Record App (X : Type)
-   := { shape : Shape; contents : {p | valid shape p} -> X }.
+An extension of a container \coq{F}, \coq{App F} in Coq, is then defined as
+follows:
+\begin{coqcode}
+Record App F (X : Type)
+   := { shape : Shape F; cont : {p | valid shape p} -> X }.
 % \end{coqcode}
@@
 | AppR_ext (shape x = shape y)
-           (forall p1 p2, projT1 e1 = projT1 e2 -> contents p1 = contents p2).
+           (forall p1 p2, projT1 e1 = projT1 e2 -> cont p1 = cont p2).
 \end{coqcode}
@@
 Class Cont `{setoid Shape} (Pos : Type) := { valid : Shape * Pos ~> bool }.
-Inductive AppR `{F : Cont Sf Pf} `{setoid X} (x y : App F X) : Prop :=
+Inductive AppR F `{setoid X} (x y : App F X) : Prop :=
   | AppR_ext (Es : shape x =e shape y)
@@
 
-\subsection{Algebras and Catamorphisms}
-\label{sec:algebras} An algebra is a set \coq{A} (the
-\emph{carrier} of the algebra) together with some operations on it sometimes
-subject to some equational laws. For example, given such a type \coq{A} we can
-define the monoid operations as \coq{u : unit -> A} for unit
-and \coq{m : A * A -> A} for multiplication. Notice
-that \coq{F X = unit + X * X} is a functor on \coq{X} which means we can
-equivalently describe these two maps as a single one of type \coq{F X -> X}
-which we call the algebra for the functor \coq{F} satisfying the unit and
-associative laws of the monoid. In this work we will be using algebras to
-describe the operations of a data type and so we will not be needing additional
-equations on these operations.
-
-Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
-by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of
-type \coq{App F A ~> A}:
+\subsection{Algebras and Catamorphisms for Containers}
+\label{sec:algebras}
+
+Recall that an algebra is a set $A$ together with a morphism that defines the
+operations of the algebra $F\;A \to A$.  Given a type \coq{A} and a container
+\coq{F}, an `\coq{App F}'-algebra is a pair given by the carrier \coq{A}, and the
+\emph{structure map} of type:
 \begin{coqcode}
@@
 \end{coqcode}
-The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
-where the structure map is an isomorphism. This is sometimes referred to as the
-\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind this name shortly. \footnote[1]{To be more precise,
-initial algebras are isomorphisms by Lambek's lemma, but isomorphisms of
-functors do not necessarily correspond to initial algebras.}  We define the
-least fixed-point of \coq{F} as an inductive type:
-\begin{coqcode}
-Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
-\end{coqcode}
-where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
+
+\noindent
+For simplicity, we will abuse terminology, and we calling these
+\coq{F}-algebras, instead of `\coq{App F}'-algebras.
+
+The initial \coq{F}-algebra can be defined as follows, thanks to the use of
+containers:
+\begin{coqcode}
+Inductive LFix F : Type := LFix_in { LFix_out : App F (LFix F) }.
+\end{coqcode}
+where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its inverse $\InOp$ (see Section~\ref{sec:rec-schemes}).
 As an example, the initial \coq{F}-algebra for the
@@
 We define \coq{LFix} as a setoid, where its equivalence relation can be
-described as the least fixed point of \coq{AppR} and we define smart
+described as the least fixed point of \coq{AppR}. We define smart
 constructors for the isomorphism of least fixed points as respectful morphisms:

### 2024-10-10 paper/paper.tex
Fixes

@@
 \begin{coqcode}
-Class Cont `{setoid Shape} (Pos : Type) := { valid : Shape * Pos ~> bool }.
 Inductive AppR F `{setoid X} (x y : App F X) : Prop :=
@@
 
-The least fixed-point for \coq{F} is the initial \coq{F}-algebra in the sense
-that it gives rise to an inductive recursion scheme.  Specifically, for any
-other \coq{F}-algebra there exists a \emph{unique} map, known as a \emph{fold} or
-\emph{catamorphism}, such that it structurally deconstructs the data type
-using \coq{LFix_out}, calls itself recursively and then composes the result of
-the recursive call using the given an \coq{F}-algebra. In other words, for any
-give \coq{F}-algebra there exists a unique \coq{F}-algebra homomorphism from the
-initial one. We define it in Coq as follows:
-\begin{coqcode}
-  Definition cata_f `{F : Cont Sh P} (alg : Alg F A) : LFix -> A
-:= fix f (x : LFix) := match x with
-  | LFix_in ax => let (sx, kx) := ax in alg (MkApp sx (fun e => f (kx e))) end.
+Catamorphisms are constructed, as expected, so that they structurally
+deconstruct the datatype, call themselves recursively, and then compose the
+result using an \coq{F}-algebra.
+\begin{coqcode}
+Definition cata_f F (alg : Alg F A) : LFix -> A
+  := fix f (x : LFix)
+     := match x with
+        | LFix_in ax =>
+             alg {shape := shape ax; cont := fun e => f (cont ax e)}
+        end.
 \end{coqcode}
@@
 \begin{coqcode}
-  cata : forall `{F : Cont Sh P} `{setoid A}, Alg F A ~> LFix ~> A
+  cata : forall F `{setoid A}, Alg F A ~> LFix ~> A
 \end{coqcode}
@@
 \label{sec:coalg}
-Algebras and catamorphisms dualise nicely to the coinductive setting.  The dual
-of an algebra is a coalgebra. A coalgebra can be thought of as an
-\emph{observation} map. For example, for a type of states \coq{X} and an
-alphabet type \coq{L} we can define a labelled transition system (LTS)
-on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{F X = L * X}
-implementing the \emph{transition} map. In particular, for a
-state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
-where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.
-
-In general, for a functor \coq{F}, an \coq{F}-coalgebra is a pair of a
-type \coq{X} called the \emph{carrier} of the coalgebra and a structure
-map \coq{X -> F X}.
-
-In our development we use the following notation for coalgebras:
+
+In general, for a container \coq{F}, an \coq{F}-coalgebra is a pair of a carrier
+\coq{X} and a structure map \coq{X -> App F X}.  In our development we use the
+following notation for coalgebras:
 \begin{coqcode}
@@
 Dually to the initial \coq{F}-algebra, a final \coq{F}-coalgebra is the greatest
-fixed-point for a functor \coq{F}. We define it using a coinductive data type:
-\begin{coqcode}
-CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
-\end{coqcode}
+fixed-point for \coq{App F}. We define it using a coinductive data type:
+\begin{coqcode}
+CoInductive GFix F : Type := GFix_in { GFix_out : App F GFix }.
+\end{coqcode}
+
+\noindent
 where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix_in} is its
@@
 The greatest fixed-point is a terminal \coq{F}-coalgebra in the sense that it
-yields a coinductive recursion scheme. Specifically, for any
-other \coq{F}-coalgebra there exists a \emph{unique} map, called the
-\emph{unfold} or \emph{anamorphism}, such that it applies the observation map,
-corecursively generates the rest of the computation and composes the result of
-the corecursive call by using the algebra \coq{GFix_in}. In other words, for any
-given \coq{F}-coalgebra there exists a unique \coq{F}-coalgebra homomorphism
-into the terminal \coq{F}-coalgebra.
-
-% Consider the transition system below composed by three states \coq{x1}, \coq{x2}
-% and \coq{x3} with the transitions labelled by the set of labels \coq{\{a, b\}} and define the map \coq{c} such that it implements the transition system below
-% \[\begin{tikzcd}
-%     {x_1} & {x_2} & {x_3} \arrow["b"', loop, distance=2em, in=35, out=325]
-% 	\arrow["a", from=1-1, to=1-2]
-% 	\arrow["b", from=1-2, to=1-3]
-% \end{tikzcd}\]
-% Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
-% carrier \coq{X} is a map \coq{X -> F X}. In the previous example \coq{c} is
-% an \coq{F}-coalgebra on the carrier \coq{X} with \coq{F X = L * X}.
-
-We define anamorphisms as follows:
+yields a coinductive recursion scheme: the \emph{anamorphism}.
 \begin{coqcode}
@@
   cofix f x :=
-    match c x with | MkApp sx kx => GFix_in (MkApp sx (fun e => f (kx e))) end.
-
-Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*... ana_f_ ... *)
+   let cx := c x in
+   GFix_in { shape := shape cx; cont := fun e => f (cont cx e) }.
+
+Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*...*)
 \end{coqcode}
@@
 \label{sec:rec-coalgebras}
-Hylomorphisms capture the concept of \emph{divide-and-conquer} algorithms where
-the input is first destructured (\emph{divide}) in smaller parts by means of a
-coalgebra which are computed recursively and then composed back together
-(\emph{conquer}) by means of an algebra.
+Recall that hylomorphisms capture the concept of \emph{divide-and-conquer}
+algorithms where the input is first destructured (\emph{divide}) in smaller
+parts by means of a coalgebra which are computed recursively and then composed
+back together (\emph{conquer}) by means of an algebra.
 
@@
 \end{equation}
-As we stated earlier, a solution to this equation does not exist for an arbitrary  algebra and coalgebra pair and, in fact, this definition cannot be accepted by Coq.
+As we stated earlier, a solution to this equation does not exist for an
+arbitrary  algebra/coalgebra pair and, in fact, this definition cannot be
+accepted by Coq.
 
 In order to find a solution we restrict ourselves to the so-called
-\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  An example of a
-recursive coalgebra is the \coq{partition} function in quicksort which
-destructures a list into a pivot and two sublist and as long as the sublists are
-smaller the partitioning function still yields a unique solution to the
-recursion scheme.
-
-We mechanise \emph{recursive
-hylomorphisms} which are guaranteed to have a unique solution to the
-hylomorphism equation. These are hylomoprhisms where the coalgebra is
-\emph{recursive}, i.e. coalgebras that terminate on all inputs. The following
-predicate captures that a coalgebra terminates on an input:
+\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  We mechanise
+\emph{recursive hylomorphisms} which are guaranteed to have a unique solution to
+the hylomorphism equation. These are hylomoprhisms where the coalgebra is
+\emph{recursive}, i.e. coalgebras that terminate on all inputs. We represent
+recursive coalgebras using a predicate that states that a coalgebra terminates
+on an input:
 \begin{coqcode}
@@
 \end{coqcode}
+
+\noindent
 A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra
@@
 Definition hylo_def (a : Alg F B) (c : Coalg F A)
-  : forall (x : A), RecF c x -> B := fix f x H
-  := match c x as h0 return (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
-     with | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
+  : forall (x : A), RecF c x -> B
+  := fix f x H
+     := match c x as cx
+          return (forall e : Pos (shape cx), RecF c (cont cx e)) -> B
+        with
+        | MkApp sx kx => fun H => a (MkApp sx (fun e => f (kx e) (H e)))
         end (RecF_inv H).
 \end{coqcode}
+
+\noindent
 We use \coq{RecF_inv} to obtain the structurally smaller proof to use in
-the recursive calls. As we did with catamorphisms and anamorphisms, we
+the recursive calls.
+
+For convenience, we package coalgebras with their proofs that they terminate for
+all inputs:
+\begin{coqcode}
+Notation RCoalg F A := ({ c : Coalg F A | forall x, RecF c x }).
+\end{coqcode}
+
+\noindent
+As we did with catamorphisms and anamorphisms, we
 prove that \coq{hylo_def} is respectful, and use this proof to build the
@@
 \begin{coqcode}
-  hylo : forall `{F : Cont Sh P} `{setoid A} `{setoid B},
-    Alg F B ~> RCoalg F A ~> A ~> B
+hylo : forall F `{setoid A} `{setoid B}, Alg F B ~> RCoalg F A ~> A ~> B
 \end{coqcode}
@@
 \end{coqcode}
-Hylomorphisms fusion falls out from the universal property:
-\begin{coqcode}
-Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
-  (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2) : f2 \o hylo g1 h1 =e hylo g2 h1.
-\end{coqcode}
+Hylomorphisms fusion falls out from this uniqueness property:
+\begin{coqcode}
+Lemma hylo_fusion_l
+  (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C) (f2 : B ~> C)
+  : f2 \o g1 =e g2 \o fmap f2 ->
+    f2 \o hylo g1 h1 =e hylo g2 h1.
+
+Lemma hylo_fusion_r
+  (h1 : RCoalg F B) (g1 : Alg F C) (h2 : RCoalg F A) (f1 : A ~> B)
+  :  h1 \o f1 =e fmap f1 \o h2 ->
+     hylo g1 h1 \o f1 =e hylo g1 h2.
+\end{coqcode}
+
+\noindent
+It is important to highlight that, in our mechanisation, these proofs follow
+\emph{exactly} the steps that one would do in a pen-and-paper proof.
+We show the series of rewrite steps for \coq{hylo_fusion_l} in Figure~\ref{fig:proof-fusion-l}:
+\begin{figure}
+\begin{coqcode}
+(* E2 : f2 \o g1 =e g2 \o fmap f2 *)
+(* ------------------------------------- *)
+(* Goal : f2 \o hylo g1 h1 =e hylo g2 h1 *)
+
+apply hylo_univ.
+       (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1        *)
+rewrite fmap_comp.
+       (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
+rewrite [...]. (* rearranging by associativity *)
+       (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
+rewrite <- E2.
+       (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1      *)
+rewrite [...]. (* rearranging by associativity *)
+       (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o fmap (hylo g1 h1)) \o h1)      *)
+rewrite <- hylo_unroll.
+       (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1                             *)
+\end{coqcode}
+  \caption{Rewrite steps to prove \coq{hylo_fusion_l} in our mechanisation. The
+steps are exactly the same that woudld be required in a manual pen-and-paper
+proof.}
+\label{fig:proof-fusion-l}
+\end{figure}
 % Qed.
@@
   % (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
-Using the hylo fusion law, we can prove the well-known \emph{deforestation
-optimisation}. This is when two consecutive recursive computations, one that
-builds a data structure, and another one that consumes it, can be fused together
-into a single recursive definition. This, in turn, allows us to prove that a
-recursive hylomorphism is the composition of a catamorphism and a recursive
-anamorphism.
+Proving the deforestation optimisation is a straightforward application
+of \coq{hylo_fusion_l} (or \coq{hylo_fusion_r}).,
 \begin{coqcode}
@@
 \subsubsection{On the subtype of finite elements}
-In this development we have defined recursive anamorphisms on inductive data
+As we mention in
+Section~\ref{sec:rec-schemes}, we define recursive anamorphisms as
+hylomorphisms built by using a recursive coalgebra, and the respective
+initial \coq{F}-algebra.
+In other words, in this development we have defined recursive anamorphisms on inductive data
 types.  We might have as well defined them on the subtype of finite elements of
@@
 catamorphisms and anamorphisms, by using the above isomorphism. In our work,
-however, we use \emph{recursive} anamorphisms, defined as \coq{hylo l_in c} for
-a recursive coalgebra \coq{c}, which compose easily with
-catamorphisms.
+however, we use \emph{recursive} anamorphisms, defined as follows for
+a recursive coalgebra \coq{c}:
+\begin{coqcode}
+Definition rana c := hylo l_in c.
+\end{coqcode}
+
+\noindent
+The main advantage of this definition is that recursive anamorphisms
+compose with catamorphisms without the need to reason about termination
+and finiteness of values of coinductive types.
 % This predicate is equivalent to the proposition that

### 2024-10-10 paper/paper.tex
Distributed Haskell code

--- 2024-10-10 paper/paper.tex
@@
 
-\usepackage{inconsolata}
 % \usepackage{draftwatermark}
@@
 and a collection of arrows $f, g, h$ between these objects, denoted by
-$\Arr{\CatC}$, such that there always exists an identity arrow
+$\Arr (\CatC)$, such that there always exists an identity arrow
 $id_{A} : A \to A$ for each object $A$ and for two arrows $A \xrightarrow{f} B$
@@
 $\cons : A \times \Str{A} \to \Str{A}$.
+
+\subsubsection{Functors in Haskell}
+We can declare the class of types which are functors.
+This is done by imposing that a function \mintinline{haskell}{f : * -> *} comes with a functorial action given by function \mintinline{haskell}{fmap}:
+\begin{minted}{haskell}
+class Functor f where
+  fmap :: (a -> b) -> f a -> f b
+\end{minted}
+
+Now the type of lists can be defined as follows along with their instance of the class \mintinline{haskell}{Functor}:
+\begin{minted}{haskell}
+data List a = Nil | Cons a (List a)
+
+instance Functor List where
+ fmap f Nil = Nil
+ fmap f (Cons x xs) = Cons (f x) (fmap f xs)
+\end{minted}
 
@@
 
+\subsubsection{Inductive Types and Catamorphisms in Haskell }
+\begin{minted}{haskell}
+newtype Fix f = In {inOp :: f (Fix f )}
+\end{minted}
+
+\begin{minted}{haskell}
+data ListF a x = Nil | Cons a x deriving Functor
+\end{minted}
+
+\begin{minted}{haskell}
+cata :: Functor f => (f a -> a) -> Fix f -> a
+cata alg = alg . fmap (cata alg) . In
+\end{minted}
+\begin{minted}{haskell}
+foldr :: (a -> b -> b) -> b -> [ a ] -> b
+foldr f e [ ]    = e
+foldr f e (x:xs) = f x (foldr f e xs)
+\end{minted}
+
 \subsection{Coalgebras and Anamorphisms}
@@
 
+\subsubsection{Coinductive Types and Anamorphisms in Haskell }
+
+\begin{minted}{haskell}
+newtype CoFix f = OutOp {out :: f (Fix f )}
+\end{minted}
+
+\begin{minted}{haskell}
+ana :: Functor b => (a -> b a) -> a -> CoFix b
+ana coalg = OutOp . fmap (ana coalg) . coalg
+\end{minted}
+
 \subsection{Recursive Coalgebras and Hylomorphism}
@@
 the recursion scheme.
-The uniqueness property of the hylo yields the following fusion laws:
-\begin{align}
+The uniqueness property of the hylo yields the following fusion law:
+\begin{equation}
   \label{eq:hylo:fusion}
-  f \circ \hylo{c}{a} & = \hylo{c}{a'}
-  \hspace{1cm}
-  \Longleftarrow
-  \hspace{1cm}
-  f \circ a = a' \circ F(f) \\
-  \hylo{c}{a} \circ f & = \hylo{c'}{a}
-  \hspace{1cm}
-  \Longleftarrow
-  \hspace{1cm}
-  c \circ f = F(f) \circ c'
-\end{align}
-Using the hylo fusion laws, we can prove the well-known \emph{deforestation
-optimisation}, also known as the \emph{composition
-law}~\cite{DBLP:conf/ifl/HinzeHJ10}. This is when two consecutive recursive
-computations, one that builds a data structure, and another one that consumes
-it, can be fused together into a single recursive definition. This, in turn,
-allows us to prove that a recursive hylomorphism is the composition of a
-catamorphism and a recursive anamorphism.
-
-$(A, c_{1})$ is a recursive coalgebra
-$(B, c_{2})$ is a recursive coalgebra
-
-$(C, a_{2})$ is an algebra
-$(B, a_{1})$ is an algebra
-
-$c_{2} \circ a_{1} = id$
-
-\begin{equation}
-  \label{eq:hylo:deforestation}
-  \hylo{a_{2}}{c_{2}} \circ \hylo{a_{1}}{c_{1}} = \hylo{a_{2}}{c_{1}}
+  f \circ \hylo{c}{a}= \hylo{c'}{a}
 \end{equation}
-
-\subsection{Recursive Anamorphisms}
+Using the hylo fusion law, we can prove the well-known \emph{deforestation
+optimisation}. This is when two consecutive recursive computations, one that
+builds a data structure, and another one that consumes it, can be fused together
+into a single recursive definition. This, in turn, allows us to prove that a
+recursive hylomorphism is the composition of a catamorphism and a recursive
+anamorphism.
+
+\subsubsection{Haskell Implementation of Hylomorphisms}
+
+which is very useful in list processing. As a fold, The frequently used function map is also a fold:
+\begin{minted}{haskell}
+map :: (a -> b) -> [a] -> [b]
+map f = foldr (\x xs -> (f x):xs) []
+\end{minted}
+ and can be
+
+\subsubsection{Recursive Anamorphisms}
 Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
@@
 
-\subsection{Haskell Implementation of Recursion Schemes}
-\begin{minted}{haskell}
-newtype Fix f = In {inOp :: f (Fix f )}
-\end{minted}
-
-\begin{minted}{haskell}
-data ListF a x = Nil | Cons a x deriving Functor
-data TreeF a x = Empty | Node x a x deriving Functor
-\end{minted}
-
-\begin{minted}{haskell}
-cata :: Functor f => (f a -> a) -> Fix f -> a
-cata alg = alg . fmap (cata alg) . In
-\end{minted}
-\begin{minted}{haskell}
-foldr :: (a -> b -> b) -> b -> [ a ] -> b
-foldr f e [ ]    = e
-foldr f e (x:xs) = f x (foldr f e xs)
-\end{minted}
-which is very useful in list processing. As a fold, The frequently used function map is also a fold:
-\begin{minted}{haskell}
-map :: (a -> b) -> [a] -> [b]
-map f = foldr (\x xs -> (f x):xs) []
-\end{minted}
-
-\begin{minted}{haskell}
-newtype CoFix f = OutOp {out :: f (Fix f )}
-\end{minted}
-
-\begin{minted}{haskell}
-ana :: Functor b => (a -> b a) -> a -> CoFix b
-ana coalg = OutOp . fmap (ana coalg) . coalg
-\end{minted}
-
- and can be
-thought of as the $\operator{fmap}$ program in functional programming.
-\begin{minted}{haskell}
-class Functor f where
-  fmap :: (a -> b) -> f a -> f b
-\end{minted}
-
 \section{Mechanising Extractable Container Functors}
@@
 right hand side. It is easy to see that
-$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.  In general, every
-polynomial functor has an equivalent container, and we have mechanised this
-result.
+$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.
+
+In general, every polynomial functor has an equivalent container, with an
+isomorphic extension, and we have mechanised this result.
+% and this can  be found in the accompanying code (file Container.v).
 % 
@@
 \begin{coqcode}
-Record morph A {eA : setoid A} B {eB : setoid B} :=
+Structure morph A {eA : setoid A} B {eB : setoid B} :=
   MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
@@
 
-Our definition of morphisms, and the use of different equivalences leads to the
-well-known ``setoid hell''. We mitigate this problem by providing tactics and
-notations to automatically discharge proofs of \coq{app_eq} for morphisms,
-whenever the types use the standard propositional equality, or a combination of
-propositional and extensional equality. However, our compositional approach
-allows us to build morphisms by plugging in other morphisms to our combinators.
-In our framework, our expectation is that the user-provided functions remain
-small, with relatively straightforward proofs of \coq{app_eq}.
-
-However, by using this mechanisation, we gain simplified proofs via
+This definition leads to the well-known ``setoid hell'', which we mitigate by
+providing tactics and notations to automatically discharge proofs of
+\coq{app_eq} for morphisms, whenever the types use the standard propositional
+equality, or a combination of propositional and extensional equality. However,
+our compositional approach allows us to build morphisms by plugging in other
+morphisms to our combinators. In our framework, our expectation is that the
+user-provided functions remain small, with relatively straightforward proofs of
+\coq{app_eq}.
+
+However, by using this mechanisation, we gain simplified proofs via the use of
 Coq's \emph{Generalised Rewriting}. Since every morphism \coq{f : A ~> B}

### 2024-10-10 paper/paper.tex

--- 2024-10-10 paper/paper.tex
@@
 \end{equation}
+
+\subsubsection{Coinductive Types and Anamorphisms in Haskell }
+
+\begin{minted}{haskell}
+newtype CoFix f = OutOp {out :: f (Fix f )}
+\end{minted}
+
+\begin{minted}{haskell}
+ana :: Functor b => (a -> b a) -> a -> CoFix b
+ana coalg = OutOp . fmap (ana coalg) . coalg
+\end{minted}
 
@@
 
-$(A, c_{1})$ is a recursive coalgebra
-$(B, c_{2})$ is a recursive coalgebra
-
-$(C, a_{2})$ is an algebra
-$(B, a_{1})$ is an algebra
-
-$c_{2} \circ a_{1} = id$
-
-\begin{equation}
-  \label{eq:hylo:deforestation}
-  \hylo{a_{2}}{c_{2}} \circ \hylo{a_{1}}{c_{1}} = \hylo{a_{2}}{c_{1}}
-\end{equation}
-
-\subsection{Recursive Anamorphisms}
+\subsubsection{Haskell Implementation of Hylomorphisms}
+
+which is very useful in list processing. As a fold, The frequently used function map is also a fold:
+\begin{minted}{haskell}
+map :: (a -> b) -> [a] -> [b]
+map f = foldr (\x xs -> (f x):xs) []
+\end{minted}
+ and can be
+
+\subsubsection{Recursive Anamorphisms}
 Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
@@
 anamorphisms satisfy the same equation.
-
-\subsection{Haskell Implementation of Recursion Schemes}
-\begin{minted}{haskell}
-newtype Fix f = In {inOp :: f (Fix f )}
-\end{minted}
-
-\begin{minted}{haskell}
-data ListF a x = Nil | Cons a x deriving Functor
-data TreeF a x = Empty | Node x a x deriving Functor
-\end{minted}
-
-\begin{minted}{haskell}
-cata :: Functor f => (f a -> a) -> Fix f -> a
-cata alg = alg . fmap (cata alg) . In
-\end{minted}
-\begin{minted}{haskell}
-foldr :: (a -> b -> b) -> b -> [ a ] -> b
-foldr f e [ ]    = e
-foldr f e (x:xs) = f x (foldr f e xs)
-\end{minted}
-which is very useful in list processing. As a fold, The frequently used function map is also a fold:
-\begin{minted}{haskell}
-map :: (a -> b) -> [a] -> [b]
-map f = foldr (\x xs -> (f x):xs) []
-\end{minted}
-
-\begin{minted}{haskell}
-newtype CoFix f = OutOp {out :: f (Fix f )}
-\end{minted}
-
-\begin{minted}{haskell}
-ana :: Functor b => (a -> b a) -> a -> CoFix b
-ana coalg = OutOp . fmap (ana coalg) . coalg
-\end{minted}
-
- and can be
-thought of as the $\operator{fmap}$ program in functional programming.
-\begin{minted}{haskell}
-class Functor f where
-  fmap :: (a -> b) -> f a -> f b
-\end{minted}
 

### 2024-10-10 paper/paper.tex
remove github footnote link
Haskell for inductive/coindutive type

@@
 \subsubsection{Functors in Haskell}
-We can declare the class of types which are functors.
-This is done by imposing that a function \mintinline{haskell}{f : * -> *} comes with a functorial action given by function \mintinline{haskell}{fmap}:
+It is often helpful for functional programmers to think of categorical
+constructions in terms of Haskell thereby assuming that Haskell types are sets
+and programs are functions~\cite{DanielssonHJG06}.
+
+We can declare the class of types which are functors. This is done by imposing
+that a function \mintinline{haskell}{f : * -> *} comes with a functorial action
+given by function \mintinline{haskell}{fmap}:
 \begin{minted}{haskell}
@@
 
-Now the type of lists can be defined as follows along with their instance of the class \mintinline{haskell}{Functor}:
+We now define the type of lists as usual and then show that it is an instance of the class \mintinline{haskell}{Functor} by implementing the \mintinline{haskell}{fmap} program:
 \begin{minted}{haskell}
@@
 \subsubsection{Inductive Types and Catamorphisms in Haskell }
+We now implment the least fixed-point of a functor \mintinline{haskell}{f}.
+Inuitively this is the type of \mintinline{haskell}{f}-branching trees of finite depth:
 \begin{minted}{haskell}
@@
 \end{minted}
-
-\begin{minted}{haskell}
-data ListF a x = Nil | Cons a x deriving Functor
-\end{minted}
-
+Note that since Haskell is a lazy language with general recursion this type can
+also represent infinite trees. However, the point of using recursion schemes is
+precisely that of avoid using general recursion in the first place.
+
+A catamorphism can be readily implemented as follows
 \begin{minted}{haskell}
@@
 \end{minted}
+
+As an example of how to construct inductive types we define the functor $X \mapsto 1 + A \times X$ as data type parameterized by a type:
 \begin{minted}{haskell}
-foldr :: (a -> b -> b) -> b -> [ a ] -> b
-foldr f e [ ]    = e
-foldr f e (x:xs) = f x (foldr f e xs)
+data ListF a x = Nil | Cons a deriving Functor
+newtype List a = (Fix ListF) a
 \end{minted}
+It should be easy now to see that a foldr is precisely a catamorphism over the
+type \mintinline{haskell}{List} which we have just defined.  For example,
+summing up the elements of a list can be implemented using a catamorphism as
+follows
+\begin{minted}{haskell}
+sum :: Fix (ListF Int) -> Int
+sum xs = cata (\case { Nil -> 0; Cons n m -> n + m } )  xs
+\end{minted}
+In the program above, the code inside the lambda case is precisely the \mintinline{haskell}{ListF Int}-algebra for the catamorphism.
 
@@
 \subsubsection{Coinductive Types and Anamorphisms in Haskell }
-
+Similarly to the inductive case, we can implement coinductive types and
+catamorphisms as follows:
 \begin{minted}{haskell}
@@
 \end{minted}
-
+As mentioned above, this type is similar to the data type
+\mintinline{haskell}{Fix} we already defined. However, this time we are going to
+use this assuming that it is the coinductive fixed-point by defining an
+anamorphism on it:
 \begin{minted}{haskell}
@@
 \end{minted}
+
+For example, the streams of natural numbers can now be defined using an anamorphism
+\begin{minted}{haskell}
+nat :: CoFix (ListF Int)
+nat = ana (\x -> Cons x (x+1)) 0
+\end{minted}
+where the argument to the anamorphism is the
+\mintinline{haskell}{ListF Int}-coalgebra.
 
@@
 sublists are smaller the partitioning function still yields a unique solution to
-the recursion scheme.
-The uniqueness property of the hylo yields the following fusion laws:
+the recursion scheme.  The uniqueness property of the hylo yields the following
+fusion laws:
 \begin{align}
@@
 \subsubsection{Haskell Implementation of Hylomorphisms}
-
-which is very useful in list processing. As a fold, The frequently used function map is also a fold:
+which is very useful in list processing. As a fold, The frequently used function
+map is also a fold:
 \begin{minted}{haskell}
@@
  and can be
+
+\begin{minted}{haskell}
+hylo :: (c -> f c) -> (a -> f a) -> c -> a
+hylo c a = a . fmap (hylo c a). c
+\end{minted}
 

### 2024-10-10 paper/paper.tex
Stuff

--- 2024-10-10 paper/paper.tex
@@
 
+Suppose that \coq{F} is a container represented in Coq. One can think of these
+\coq{F} as pairs containing the shapes and families of positions (which are, in
+turn, defined in terms of decidable validity predicates).
 An extension of a container \coq{F}, \coq{App F} in Coq, is then defined as
@@
    := { shape : Shape F; cont : {p | valid shape p} -> X }.
-% \end{coqcode}
+\end{coqcode}
 
@@
 In particular, we want to reason about the equality of functions of types such
-as \coq{App A -> B} (or \coq{B -> App A}).
+as \coq{App F A -> B} (or \coq{B -> App F A}).
 %
@@
 Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
-however, we chose to use Coq's propositional equality, since these would leads
+however, we chose to use Coq's propositional equality, since this leads
 to simpler code. We explain why in Section~\ref{sec:algebras}, and the
@@
 \end{coqcode}
+
+\noindent
 Proving that the composition of these functions is the identity is
@@
 \noindent
-For simplicity, we will abuse terminology, and we calling these
+For simplicity, we will abuse terminology, and call these
 \coq{F}-algebras, instead of `\coq{App F}'-algebras.
@@
 \end{coqcode}
+
+\noindent
 where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its inverse $\InOp$ (see Section~\ref{sec:rec-schemes}).
 As an example, the initial \coq{F}-algebra for the
-functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
+container that is isomorpic to the functor
+\coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
 being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
@@
 We define \coq{LFix} as a setoid, where its equivalence relation can be
-described as the least fixed point of \coq{AppR}. We define smart
-constructors for the isomorphism of least fixed points as respectful morphisms:
+described as the least fixed point of \coq{AppR} (see
+Section~\ref{sec:types-morphisms}). We define smart constructors for the
+isomorphism of least fixed points as respectful morphisms:
 \begin{coqcode}
@@
 \begin{coqcode}
-Definition cata_f F (alg : Alg F A) : LFix -> A
+Definition cata_f (alg : Alg F A) : LFix -> A
   := fix f (x : LFix)
@@
 \end{coqcode}
+
 It is easy to show that this function is a respectful morphism of
@@
 \begin{coqcode}
-  cata : forall F `{setoid A}, Alg F A ~> LFix ~> A
-\end{coqcode}
-We prove that catamorphisms satisfy the universality property we explained previously:
+  cata : forall `{setoid A}, Alg F A ~> LFix ~> A
+\end{coqcode}
+
+\noindent
+We prove that catamorphisms satisfy the universality property we explained in
+Section~\ref{sec:rec-schemes}:
 \begin{coqcode}
@@
 \end{coqcode}
+
+\noindent
 In other words, if there is any other \coq{f} with the same structural recursive
@@
 of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
-greatest fixed points
+greatest fixed points:
 \begin{coqcode}
@@
 \end{coqcode}
+
+\noindent
 From this definition the universality property falls out:
@@
 
-\subsection{Recursive Coalgebras and Hylomorphism}
+\subsection{Mechanising Hylomorphisms}
 \label{sec:rec-coalgebras}
@@
 
-In general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
-unique solution (when it exists) to the equation
-\begin{equation}
-  \label{eq:hylo}
+As we mentioned in Section~\ref{sec:rec-schemes}, given an $F$-algebra and
+$F$-coalgebra, the hylomorphism is the unique solution (when it exists) to the
+equation
+\[
   f = a \circ F\; f \circ c
-\end{equation}
+\]
 As we stated earlier, a solution to this equation does not exist for an
@@
 \noindent
-A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra
-\coq{c} such that \coq{forall x, RecF c x}.
+For convenience, we package recursive coalgebras with their proofs that they
+terminate for all inputs:
+\begin{coqcode}
+Notation RCoalg F A := ({ c : Coalg F A | forall x, RecF c x }).
+\end{coqcode}
 Recursive hylomorphisms are implemented in Coq recursively on the structure of
@@
         with
-        | MkApp sx kx => fun H => a (MkApp sx (fun e => f (kx e) (H e)))
+        | cx => fun H =>
+            a { shape := shape cx ; cont := fun e => f (cont cx e) (H e) }
         end (RecF_inv H).
@@
 the recursive calls.
-
-For convenience, we package coalgebras with their proofs that they terminate for
-all inputs:
-\begin{coqcode}
-Notation RCoalg F A := ({ c : Coalg F A | forall x, RecF c x }).
-\end{coqcode}
 
@@
 code is similar to a hand-written \coq{qsort}.
-The mergesort algorithm can be defined analogously and can be found in the
-formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].
+The mergesort algorithm can be defined analogously.
+% and can be found in the
+% formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].
 

### 2024-10-10 paper/paper.tex
Stuff

@@
 It is important to highlight that, in our mechanisation, these proofs follow
-\emph{exactly} the steps that one would do in a pen-and-paper proof.
-We show the series of rewrite steps for \coq{hylo_fusion_l} in Figure~\ref{fig:proof-fusion-l}:
+\emph{exactly} the steps that one would do in a pen-and-paper proof.  We show
+the series of rewrite steps for \coq{hylo_fusion_l} in
+Figure~\ref{fig:proof-fusion-l}. The steps of this proof are: (1) apply the
+uniqueness law of recursive hylomorphisms; (2) rewrite the goal using the
+property that functors preserve composition; (3) rewrite the goal using the
+condition of \coq{hylo_fusion_l}; (4) use the uniqueness law of hylomorphisms to
+fold $a \circ F \; f \circ c$ into $f$.
 \begin{figure}
@@
 Definition MemoPos := Pg.
-Instance Memo : Cont MemoShape MemoPos := { valid := valid \o pair (snd \o fst) snd }.
+Instance Memo : Cont MemoShape MemoPos
+  := { valid := valid \o pair (snd \o fst) snd }.
 
@@
     (let rec f x0 =
-      if x0=0 then Uint63.of_int (0)
+      if x0=0 then
+        { lFix_out = {
+             shape = Uint63.of_int (0);
+             cont  = fun _ -> f 0 } }
       else let fn := f (x0-1) in { lFix_out = { 
@@
 \begin{ocamlcode}
-let rec sf1 f ys = function | [] -> let rec f0 = function
+let rec sf1 f ys =
+  function | [] -> let rec f0 = function
                    | [] -> (Uint63.of_int (0))
@@
 
-\mvol{Do we need to discuss this or can we cut it?}
-
 In particular, the problem of \emph{nonstructural recursion} (including
@@
 mergesort''~\cite{AbreuDHJMS23}.
+There are common approaches in Coq to deal with termination, but none of these
+approaches address program calculation techniques, and the mechanisation of
+fusion laws.
 

### 2024-10-10 paper/paper.tex
quicksort in Haskell

--- 2024-10-10 paper/paper.tex
@@
 
-As an example of how to construct inductive types we define the functor $X \mapsto 1 + A \times X$ as data type parameterized by a type:
+As an example of how to construct inductive types we define the functor
+$X \mapsto 1 + A \times X$ as data type parameterized by a type:
 \begin{minted}{haskell}
@@
 \begin{minted}{haskell}
-hylo :: (c -> f c) -> (a -> f a) -> c -> a
-hylo c a = a . fmap (hylo c a). c
+hylo :: Functor f => (c -> f c) -> (f a -> a) -> c -> a
+hylo c a = a . fmap (hylo c a) . c
+\end{minted}
+
+ As an example, we the quicksort program can be written as an hylomorphism
+
+% quicksort :: [Int] -> [Int]
+% quicksort (x:xs) = let (xs, x, ys) = partition x xs
+%                in (quicksort xs) ++ [x] ++ (quicksort ys)
+%   where
+%     partition x xs = ([l | l <- xs, l < x], x, [r | r <- xs, r > x])
+
+\begin{minted}{haskell}
+data TreeF a k = Leaf | Branch k a k deriving Functor
+\end{minted}
+
+\begin{minted}{haskell}
+divide :: [Int] -> TreeF Int [Int]
+divide [] = Leaf
+divide (p:xs) = Branch [l | l <- xs, l < p] p [r | r <- xs, r > p]
+\end{minted}
+
+\begin{minted}{haskell}
+conquer :: TreeF Int [Int] -> [Int]
+conquer Leaf = []
+conquer (Branch ls x rs) = ls ++ [x] ++ rs
+\end{minted}
+\begin{minted}{haskell}
+quicksort' = hylo divide conquer
 \end{minted}
@@
 
-Suppose that \coq{F} is a container represented in Coq. One can think of these
-\coq{F} as pairs containing the shapes and families of positions (which are, in
-turn, defined in terms of decidable validity predicates).
 An extension of a container \coq{F}, \coq{App F} in Coq, is then defined as
@@
    := { shape : Shape F; cont : {p | valid shape p} -> X }.
-\end{coqcode}
+% \end{coqcode}
 
@@
 In particular, we want to reason about the equality of functions of types such
-as \coq{App F A -> B} (or \coq{B -> App F A}).
+as \coq{App A -> B} (or \coq{B -> App A}).
 %
@@
 Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
-however, we chose to use Coq's propositional equality, since this leads
+however, we chose to use Coq's propositional equality, since these would leads
 to simpler code. We explain why in Section~\ref{sec:algebras}, and the
@@
 \end{coqcode}
-
-\noindent
 Proving that the composition of these functions is the identity is
@@
 \noindent
-For simplicity, we will abuse terminology, and call these
+For simplicity, we will abuse terminology, and we calling these
 \coq{F}-algebras, instead of `\coq{App F}'-algebras.
@@
 \end{coqcode}
-
-\noindent
 where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its inverse $\InOp$ (see Section~\ref{sec:rec-schemes}).
 As an example, the initial \coq{F}-algebra for the
-container that is isomorpic to the functor
-\coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
+functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
 being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
@@
 We define \coq{LFix} as a setoid, where its equivalence relation can be
-described as the least fixed point of \coq{AppR} (see
-Section~\ref{sec:types-morphisms}). We define smart constructors for the
-isomorphism of least fixed points as respectful morphisms:
+described as the least fixed point of \coq{AppR}. We define smart
+constructors for the isomorphism of least fixed points as respectful morphisms:
 \begin{coqcode}
@@
 \begin{coqcode}
-Definition cata_f (alg : Alg F A) : LFix -> A
+Definition cata_f F (alg : Alg F A) : LFix -> A
   := fix f (x : LFix)
@@
 \end{coqcode}
-
 It is easy to show that this function is a respectful morphism of
@@
 \begin{coqcode}
-  cata : forall `{setoid A}, Alg F A ~> LFix ~> A
-\end{coqcode}
-
-\noindent
-We prove that catamorphisms satisfy the universality property we explained in
-Section~\ref{sec:rec-schemes}:
+  cata : forall F `{setoid A}, Alg F A ~> LFix ~> A
+\end{coqcode}
+We prove that catamorphisms satisfy the universality property we explained previously:
 \begin{coqcode}
@@
 \end{coqcode}
-
-\noindent
 In other words, if there is any other \coq{f} with the same structural recursive
@@
 of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
-greatest fixed points:
+greatest fixed points
 \begin{coqcode}
@@
 \end{coqcode}
-
-\noindent
 From this definition the universality property falls out:
@@
 
-\subsection{Mechanising Hylomorphisms}
+\subsection{Recursive Coalgebras and Hylomorphism}
 \label{sec:rec-coalgebras}
@@
 
-As we mentioned in Section~\ref{sec:rec-schemes}, given an $F$-algebra and
-$F$-coalgebra, the hylomorphism is the unique solution (when it exists) to the
-equation
-\[
+In general given an $F$-algebra and $F$-coalgebra, the hylomorphism is the
+unique solution (when it exists) to the equation
+\begin{equation}
+  \label{eq:hylo}
   f = a \circ F\; f \circ c
-\]
+\end{equation}
 As we stated earlier, a solution to this equation does not exist for an
@@
 \noindent
-For convenience, we package recursive coalgebras with their proofs that they
-terminate for all inputs:
-\begin{coqcode}
-Notation RCoalg F A := ({ c : Coalg F A | forall x, RecF c x }).
-\end{coqcode}
+A recursive coalgebra of type \coq{RCoalg F A} is a coalgebra
+\coq{c} such that \coq{forall x, RecF c x}.
 Recursive hylomorphisms are implemented in Coq recursively on the structure of
@@
         with
-        | cx => fun H =>
-            a { shape := shape cx ; cont := fun e => f (cont cx e) (H e) }
+        | MkApp sx kx => fun H => a (MkApp sx (fun e => f (kx e) (H e)))
         end (RecF_inv H).
@@
 the recursive calls.
+
+For convenience, we package coalgebras with their proofs that they terminate for
+all inputs:
+\begin{coqcode}
+Notation RCoalg F A := ({ c : Coalg F A | forall x, RecF c x }).
+\end{coqcode}
 

### 2024-10-10 paper/paper.tex
Stuff

--- 2024-10-10 paper/paper.tex
@@
 
-This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
-\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code.
-The
-full mechanisation is open source.
+The contributions of this paper are as follows.  This work presents the first
+Coq formalisation \emph{hylomorphisms} that (1) is \emph{fully axiom-free}; (2)
+allows the extraction of idiomatic OCaml code; and (3) can use regular Coq
+equalities to do program calculation, derive correct implementations, and apply
+optimisations.
+The full mechanisation is open source.
 %can be found on
@@
 %  ANONYMISE THIS}.
-While programmers still need to reason about the termination of
-their programs, our mechanisation will allow the use of the algebraic laws of
-hylomorphisms for \emph{program calculation}, as well as extracting to idiomatic
-code (see examples of extracted code in Section~\ref{sec:examples}).
+While programmers still need to reason about the termination of their programs,
+the use \emph{program calculation} can ease this task by allowing the use of
+fusion on programs that have already been proven to terminate.
 % As an example, our mechanisation of \ocaml{qsort} as a
@@
 % \end{coqcode}
-
-To summarise, our contributions are as follows. In this paper we provide a
-framework for \emph{program calculation} with recursion schemes in Coq, where
-program calculation is done by applying regular Coq equalities, based on  proofs
-of the algebraic laws of hylomorphisms, and that allows the extraction of safe,
-idiomatic OCaml code:
+%
+
+The remainder of the paper is structured as follows:
 \begin{itemize}
@@
 the algebra $a_{Y}$ instead of $a_{X}$. This is a useful result for program
-optimisation as we shall see. \mpav{We shall see? }
+optimisation as we shall see.
 
@@
 $[\nil,\cons]$ into the algebra $[b, i]$.
-
-\mpav{TODO}In functional programming this is commonly referred to as
+In functional programming this is commonly referred to as
 $\foldr : (1 \to B) \to (A \times B \to B) \to B$. We can in fact set

### 2024-10-10 paper/paper.tex
Hylomorphisms

--- 2024-10-10 paper/paper.tex
@@
 \subsubsection{Haskell Implementation of Hylomorphisms}
-which is very useful in list processing. As a fold, The frequently used function
-map is also a fold:
-\begin{minted}{haskell}
-map :: (a -> b) -> [a] -> [b]
-map f = foldr (\x xs -> (f x):xs) []
-\end{minted}
- and can be
-
+% which is very useful in list processing. As a fold, The frequently used function
+% map is also a fold:
+% \begin{minted}{haskell}
+% map :: (a -> b) -> [a] -> [b]
+% map f = foldr (\x xs -> (f x):xs) []
+% \end{minted}
+We implement the hylomorphism tightly following the theory explained above:
 \begin{minted}{haskell}
@@
 
- As an example, we the quicksort program can be written as an hylomorphism
-
-% quicksort :: [Int] -> [Int]
-% quicksort (x:xs) = let (xs, x, ys) = partition x xs
-%                in (quicksort xs) ++ [x] ++ (quicksort ys)
-%   where
-%     partition x xs = ([l | l <- xs, l < x], x, [r | r <- xs, r > x])
-
+A classic example of a divide-and-conquer algorithm that can be written as an hylomorphism is the quicksort program below:
+
+\begin{minded}{haskell}
+partition x xs = ([l | l <- xs, l < x], x, [r | r <- xs, r >= x])
+
+quicksort :: [Int] -> [Int]
+quicksort [] = []
+quicksort (x:xs) = let (xs, x, ys) = partition x xs
+                   in (quicksort xs) ++ [x] ++ (quicksort ys)
+\end{minted}
+The partition program takes an element \mintinline{haskell}{x} and a list and returns a triple consisting of a list with all the elements in the list smaller than \mintinline{haskell}{x}, the element itself and a list with all the elements in the list greater or equal than  \mintinline{haskell}{x}.
+The quicksort program first runs the partition program, which divides the list in three parts, calls itself recursively on the sublists and then puts the results together using the concatenation function.
+
+To implement this as an hylo we define the functor representing trees where leaves are just empty and each node contains a value:
 \begin{minted}{haskell}
@@
 \end{minted}
-
+Now the partition program can be written as a  \mintinline{haskell}{TreeF}-coalgebra
 \begin{minted}{haskell}
@@
 \end{minted}
-
+and the concatenation function can be written as  \mintinline{haskell}{TreeF}-algebra
 \begin{minted}{haskell}
@@
 \end{minted}
+Finally, quicksort can be rewritten as an hylomorphism
 \begin{minted}{haskell}
-quicksort' = hylo divide conquer
+quicksort = hylo divide conquer
 \end{minted}
+As an important remark we have used the in-built list type for keeping the code as tidy as possible, but it would have been possible to use the inductive \mintinline{haskell}{List} type we defined above and define both  \mintinline{haskell}{divide} and \mintinline{haskell}{conquer} via catamorphisms.
+
+Finally, it is important to note that both  \mintinline{haskell}{divide} and  \mintinline{haskell}{conquer} are not initial algebras or terminal coalgebras on \mintinline{haskell}{ListF}.
+These are rather (co)algebras on \mintinline{haskell}{ListF}.
+Hence the terminating argument relies solely on the fact that \mintinline{haskell}{divide} is a recursive coalgebra thus there is an unique solution to \mintinline{haskell}{hylo divide conquer}.
 

### 2024-10-10 paper/paper.tex
Finished Background

@@
 
-\begin{minded}{haskell}
+\begin{minted}{haskell}
 partition x xs = ([l | l <- xs, l < x], x, [r | r <- xs, r >= x])
@@
 \end{minted}
-As an important remark we have used the in-built list type for keeping the code as tidy as possible, but it would have been possible to use the inductive \mintinline{haskell}{List} type we defined above and define both  \mintinline{haskell}{divide} and \mintinline{haskell}{conquer} via catamorphisms.
-
-Finally, it is important to note that both  \mintinline{haskell}{divide} and  \mintinline{haskell}{conquer} are not initial algebras or terminal coalgebras on \mintinline{haskell}{ListF}.
-These are rather (co)algebras on \mintinline{haskell}{ListF}.
-Hence the terminating argument relies solely on the fact that \mintinline{haskell}{divide} is a recursive coalgebra thus there is an unique solution to \mintinline{haskell}{hylo divide conquer}.
+As an important remark we have used the in-built list type for keeping the code
+as tidy as possible, but it would have been possible to use the inductive
+\mintinline{haskell}{List} type we defined above and define both
+\mintinline{haskell}{divide} and \mintinline{haskell}{conquer} via
+catamorphisms.
+
+Finally, it is important to note that both  \mintinline{haskell}{divide} and
+\mintinline{haskell}{conquer} are not initial algebras or terminal coalgebras on
+\mintinline{haskell}{ListF}. These are rather (co)algebras on
+\mintinline{haskell}{ListF}.  Hence the terminating argument relies solely on
+the fact that \mintinline{haskell}{divide} is a recursive coalgebra thus there
+is an unique solution to \mintinline{haskell}{hylo divide conquer}.
 

### 2024-10-10 paper/paper.tex
Uniqueness of hylos

@@
 and suitably adapted for program extraction.
-  \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and
-hylomorhisms and their universal properties.
+  \item In Section~\ref{sec:recursion-schemes} we mechanise folds, unfolds, and
+hylomorphisms, as well as proving their uniqueness properties in Coq.
   \item In Section~\ref{sec:examples} we use the framework to formalise examples
@@
 \subsubsection{Haskell Implementation of Hylomorphisms}
-% which is very useful in list processing. As a fold, The frequently used function
-% map is also a fold:
-% \begin{minted}{haskell}
-% map :: (a -> b) -> [a] -> [b]
-% map f = foldr (\x xs -> (f x):xs) []
-% \end{minted}
-We implement the hylomorphism tightly following the theory explained above:
+which is very useful in list processing. As a fold, The frequently used function
+map is also a fold:
+\begin{minted}{haskell}
+map :: (a -> b) -> [a] -> [b]
+map f = foldr (\x xs -> (f x):xs) []
+\end{minted}
+ and can be
+
 \begin{minted}{haskell}
@@
 
-A classic example of a divide-and-conquer algorithm that can be written as an hylomorphism is the quicksort program below:
-
-\begin{minted}{haskell}
-partition x xs = ([l | l <- xs, l < x], x, [r | r <- xs, r >= x])
-
-quicksort :: [Int] -> [Int]
-quicksort [] = []
-quicksort (x:xs) = let (xs, x, ys) = partition x xs
-                   in (quicksort xs) ++ [x] ++ (quicksort ys)
-\end{minted}
-The partition program takes an element \mintinline{haskell}{x} and a list and returns a triple consisting of a list with all the elements in the list smaller than \mintinline{haskell}{x}, the element itself and a list with all the elements in the list greater or equal than  \mintinline{haskell}{x}.
-The quicksort program first runs the partition program, which divides the list in three parts, calls itself recursively on the sublists and then puts the results together using the concatenation function.
-
-To implement this as an hylo we define the functor representing trees where leaves are just empty and each node contains a value:
+ As an example, we the quicksort program can be written as an hylomorphism
+
+% quicksort :: [Int] -> [Int]
+% quicksort (x:xs) = let (xs, x, ys) = partition x xs
+%                in (quicksort xs) ++ [x] ++ (quicksort ys)
+%   where
+%     partition x xs = ([l | l <- xs, l < x], x, [r | r <- xs, r > x])
+
 \begin{minted}{haskell}
@@
 \end{minted}
-Now the partition program can be written as a  \mintinline{haskell}{TreeF}-coalgebra
+
 \begin{minted}{haskell}
@@
 \end{minted}
-and the concatenation function can be written as  \mintinline{haskell}{TreeF}-algebra
+
 \begin{minted}{haskell}
@@
 \end{minted}
-Finally, quicksort can be rewritten as an hylomorphism
 \begin{minted}{haskell}
-quicksort = hylo divide conquer
+quicksort' = hylo divide conquer
 \end{minted}
-As an important remark we have used the in-built list type for keeping the code
-as tidy as possible, but it would have been possible to use the inductive
-\mintinline{haskell}{List} type we defined above and define both
-\mintinline{haskell}{divide} and \mintinline{haskell}{conquer} via
-catamorphisms.
-
-Finally, it is important to note that both  \mintinline{haskell}{divide} and
-\mintinline{haskell}{conquer} are not initial algebras or terminal coalgebras on
-\mintinline{haskell}{ListF}. These are rather (co)algebras on
-\mintinline{haskell}{ListF}.  Hence the terminating argument relies solely on
-the fact that \mintinline{haskell}{divide} is a recursive coalgebra thus there
-is an unique solution to \mintinline{haskell}{hylo divide conquer}.
 
@@
 algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
-from the universality property of the hylomorphism and the fact that recursive
+from the uniqueness property of the hylomorphism and the fact that recursive
 anamorphisms satisfy the same equation.
@@
 \begin{coqcode}
-Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
+Lemma hylo_uniq (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
     : f =e hylo g h <-> f =e g \o fmap f \o h.
@@
 
-apply hylo_univ.
+apply hylo_uniq.
        (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1        *)
@@
 % Qed.
-%   rewrite hylo_univ, fmap_comp, compA, <- E2,
+%   rewrite hylo_uniq, fmap_comp, compA, <- E2,
 %           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
 
-  % rewrite hylo_univ, fmap_comp, compA, <- E2,
+  % rewrite hylo_uniq, fmap_comp, compA, <- E2,
   %         <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.

### 2024-10-10 paper/paper.tex
minor
add related work on fusion

--- 2024-10-10 paper/paper.tex
@@
 and suitably adapted for program extraction.
-  \item In Section~\ref{sec:recursion-schemes} we mechanise folds, unfolds, and
-hylomorphisms, as well as proving their uniqueness properties in Coq.
+  \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and
+hylomorhisms and their universal properties.
   \item In Section~\ref{sec:examples} we use the framework to formalise examples
@@
 algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
-from the uniqueness property of the hylomorphism and the fact that recursive
+from the universality property of the hylomorphism and the fact that recursive
 anamorphisms satisfy the same equation.
@@
 \begin{coqcode}
-Lemma hylo_uniq (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
+Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
     : f =e hylo g h <-> f =e g \o fmap f \o h.
@@
 
-apply hylo_uniq.
+apply hylo_univ.
        (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1        *)
@@
 % Qed.
-%   rewrite hylo_uniq, fmap_comp, compA, <- E2,
+%   rewrite hylo_univ, fmap_comp, compA, <- E2,
 %           <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
 
-  % rewrite hylo_uniq, fmap_comp, compA, <- E2,
+  % rewrite hylo_univ, fmap_comp, compA, <- E2,
   %         <- compA, <- compA, (compA g1), <- hylo_unr.  reflexivity.
@@
 (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}). Additionally, we compare our
-encoding of recursion schemes to other forms of termination checking
-(Section~\ref{sec:termcheck}, as a way to guarantee termination of nonstructural
-recursion.
+encoding of recursion schemes to other forms of termination checking for
+nonstructural recursion (Section~\ref{sec:termcheck}), and we discuss other
+approaches to fusion in functional programs (Section~\ref{sec:fusion}).
 
@@
 fusion laws.
+
+\subsection{Fusion}\label{sec:fusion}
+
+Fusion in functional programming is a general technique for combining multiple
+computations into a single pass to eliminate intermediate data structures and
+improve efficiency. The fusion we have discussed in this paper relating to
+program calculation is one form of fusion, but there are other approaches.
+
+\emph{Deforestation}, introduced by Wadler~\cite{wadler88}, is a specific
+approach to fusion that aims to remove intermediate trees (such as lists)
+created between chained operations. A well-known method of deforestation is
+\haskell{foldr/build} fusion~\cite{gill93}, which optimizes list-processing by
+fusing list-producing functions (like \haskell{map} or \haskell{filter}) with
+list-consuming ones (like \haskell{foldr}), avoiding the creation of temporary
+lists.
+
+Extending this idea, stream fusion generalizes the concept to handle other data
+types, using a lazy stream abstraction to represent sequences and fuse
+transformations like \haskell{map} and \haskell{fold}~\cite{coutts07}. Both
+approaches eliminate redundant structures and ensure computations happen in a
+single efficient pass over the data.
+
+Another dimension of fusion is \emph{tupling}, a technique used to combine
+multiple recursive functions into a single function by grouping their results
+into a tuple~\cite{hughes89}. This approach helps avoid redundant traversals of
+the same data structure when multiple operations are applied. In addition to
+techniques like \haskell{foldr/build} fusion and stream fusion, tupling
+complements deforestation by allowing multiple recursive functions to share a
+single traversal, further optimizing computations and minimizing intermediate
+data structures.
 

### 2024-10-10 paper/paper.tex

--- 2024-10-10 paper/paper.tex
@@
 \mintinline{haskell}{ListF}. These are rather (co)algebras on
-\mintinline{haskell}{ListF}.  Hence the terminating argument relies solely on
+\mintinline{haskell}{TreeF}.  Hence the terminating argument relies solely on
 the fact that \mintinline{haskell}{divide} is a recursive coalgebra thus there

### 2024-10-10 paper/paper.tex
Stuff

--- 2024-10-10 paper/paper.tex
@@
   (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C) (f2 : B ~> C)
-  : f2 \o g1 =e g2 \o fmap f2 ->
-    f2 \o hylo g1 h1 =e hylo g2 h1.
+  : f2 \o g1 =e g2 \o fmap f2 -> f2 \o hylo g1 h1 =e hylo g2 h1.
 
@@
   (h1 : RCoalg F B) (g1 : Alg F C) (h2 : RCoalg F A) (f1 : A ~> B)
-  :  h1 \o f1 =e fmap f1 \o h2 ->
-     hylo g1 h1 \o f1 =e hylo g1 h2.
+  :  h1 \o f1 =e fmap f1 \o h2 -> hylo g1 h1 \o f1 =e hylo g1 h2.
 \end{coqcode}
@@
 \end{coqcode}
+
+  \vspace{-.5cm}
   \caption{Rewrite steps to prove \coq{hylo_fusion_l} in our mechanisation. The
@@
 \end{coqcode}
-And two functions to inspect the head of a memo table, and remove an
-element from it:
+And two functions, to inspect the head of a table, and remove an
+element:
 \begin{coqcode}
@@
   end kx end.
-Definition knapsackA wvs : App NatF (Table NatF int) ~> int := 
-  (* [knapsack_alg wvs] as a respecful morphism *)
+Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
 \end{coqcode}
@@
 \begin{ocamlcode}
-let knapsack wvs x = let (y, _) =
-    (let rec f x0 =
+let knapsack wvs x = let (y, _) = (let rec f x0 =
       if x0=0 then
-        { lFix_out = {
-             shape = Uint63.of_int (0);
-             cont  = fun _ -> f 0 } }
+        { lFix_out = {shape = Uint63.of_int (0); cont  = fun _ -> f 0 } }
       else let fn := f (x0-1) in { lFix_out = {
@@
 
-We then prove that the length function fuses with
-the naive quadratic \emph{reverse} function:
+\noindent
+We then prove that \ocaml{length} fuses with
+the naive quadratic \ocaml{reverse} function:
 \begin{coqcode}

### 2024-10-10 paper/paper.tex
shrink related work slightly

--- 2024-10-10 paper/paper.tex
@@
 Encoding recursion schemes in Coq is not new. We compare our work with other
-encodings of program calculation techniques in Coq and recursion schemes in Coq
-(Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}). Additionally, we compare our
-encoding of recursion schemes to other forms of termination checking for
-nonstructural recursion (Section~\ref{sec:termcheck}), and we discuss other
-approaches to fusion in functional programs (Section~\ref{sec:fusion}).
+encodings of program calculation techniques and recursion schemes in Coq, and
+discuss other approaches to fusion and to termination of nonstructural
+recursion.
+% and recursion schemes in Coq
+% (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}). Additionally, we compare our
+% encoding of recursion schemes to other forms of termination checking for
+% nonstructural recursion (Section~\ref{sec:termcheck}), and we discuss other
+% approaches to fusion in functional programs (Section~\ref{sec:fusion}).
 
@@
 lists.
-
+%
 Extending this idea, stream fusion generalizes the concept to handle other data
@@
 into a tuple~\cite{hughes89}. This approach helps avoid redundant traversals of
-the same data structure when multiple operations are applied. In addition to
-techniques like \haskell{foldr/build} fusion and stream fusion, tupling
-complements deforestation by allowing multiple recursive functions to share a
-single traversal, further optimizing computations and minimizing intermediate
-data structures.
+the same data structure when multiple operations are applied.
+% In addition to
+% techniques like \haskell{foldr/build} fusion and stream fusion, tupling
+% complements deforestation by allowing multiple recursive functions to share a
+% single traversal, further optimizing computations and minimizing intermediate
+% data structures.
 

### 2024-10-10 paper/paper.tex
minor

--- 2024-10-10 paper/paper.tex
@@
 \end{minted}
-
-For example, the streams of natural numbers can now be defined using an anamorphism
+For example, the stream of natural numbers can now be defined using an anamorphism
 \begin{minted}{haskell}
@@
 
-A classic example of a divide-and-conquer algorithm that can be written as an hylomorphism is the quicksort program below:
-
+A classic example of a divide-and-conquer algorithm that can be written as an
+hylomorphism is the quicksort program below:
 \begin{minted}{haskell}
@@
 \end{minted}
-The partition program takes an element \mintinline{haskell}{x} and a list and returns a triple consisting of a list with all the elements in the list smaller than \mintinline{haskell}{x}, the element itself and a list with all the elements in the list greater or equal than  \mintinline{haskell}{x}.
-The quicksort program first runs the partition program, which divides the list in three parts, calls itself recursively on the sublists and then puts the results together using the concatenation function.
-
-To implement this as an hylo we define the functor representing trees where leaves are just empty and each node contains a value:
+The partition program takes an element \mintinline{haskell}{x} and a list and
+returns a triple consisting of a list with all the elements in the list smaller
+than \mintinline{haskell}{x}, the element itself and a list with all the
+elements in the list greater or equal than  \mintinline{haskell}{x}. The
+quicksort program first runs the partition program, which divides the list in
+three parts, calls itself recursively on the sublists and then puts the results
+together using the concatenation function.
+
+To implement this as an hylo we define the functor representing trees where
+leaves are just empty and each node contains a value:
 \begin{minted}{haskell}
@@
   (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C) (f2 : B ~> C)
-  : f2 \o g1 =e g2 \o fmap f2 -> f2 \o hylo g1 h1 =e hylo g2 h1.
+  : f2 \o g1 =e g2 \o fmap f2 ->
+    f2 \o hylo g1 h1 =e hylo g2 h1.
 
@@
   (h1 : RCoalg F B) (g1 : Alg F C) (h2 : RCoalg F A) (f1 : A ~> B)
-  :  h1 \o f1 =e fmap f1 \o h2 -> hylo g1 h1 \o f1 =e hylo g1 h2.
+  :  h1 \o f1 =e fmap f1 \o h2 ->
+     hylo g1 h1 \o f1 =e hylo g1 h2.
 \end{coqcode}
@@
 \end{coqcode}
-
-  \vspace{-.5cm}
   \caption{Rewrite steps to prove \coq{hylo_fusion_l} in our mechanisation. The
@@
 \end{coqcode}
-And two functions, to inspect the head of a table, and remove an
-element:
+And two functions to inspect the head of a memo table, and remove an
+element from it:
 \begin{coqcode}
@@
   end kx end.
-Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
+Definition knapsackA wvs : App NatF (Table NatF int) ~> int := 
+  (* [knapsack_alg wvs] as a respecful morphism *)
 \end{coqcode}
@@
 \begin{ocamlcode}
-let knapsack wvs x = let (y, _) = (let rec f x0 =
+let knapsack wvs x = let (y, _) =
+    (let rec f x0 =
       if x0=0 then
-        { lFix_out = {shape = Uint63.of_int (0); cont  = fun _ -> f 0 } }
+        { lFix_out = {
+             shape = Uint63.of_int (0);
+             cont  = fun _ -> f 0 } }
       else let fn := f (x0-1) in { lFix_out = { 
@@
 
-\noindent
-We then prove that \ocaml{length} fuses with
-the naive quadratic \ocaml{reverse} function:
+We then prove that the length function fuses with
+the naive quadratic \emph{reverse} function:
 \begin{coqcode}

### 2024-10-10 paper/paper.tex
minor

--- 2024-10-10 paper/paper.tex
@@
 
-We now define the type of lists as usual and then show that it is an instance of the class \mintinline{haskell}{Functor} by implementing the \mintinline{haskell}{fmap} program:
+We now define the type of lists as usual and then show that it is an instance of
+the class \mintinline{haskell}{Functor} by implementing the
+\mintinline{haskell}{fmap} program:
 \begin{minted}{haskell}
@@
 \end{minted}
-In the program above, the code inside the lambda case is precisely the \mintinline{haskell}{ListF Int}-algebra for the catamorphism.
+In the program above, the code inside the lambda case is precisely the
+\mintinline{haskell}{ListF Int}-algebra for the catamorphism.
 
@@
 Consider instead representing the positions of a container using a decidable
-validity predicate which assigns shapes to positions. In Coq, we use
-a boolean function and a coercion from \coq{bool} to \coq{Prop} to represent
-the decidable predicate \coq{valid}, similarly to SSReflect~\cite{GonthierL09}. Given
-a \coq{Shape : Type}, and a type of \emph{all possible positions} \coq{APos : Type},
-the family of
-positions for a given shape can be defined as a dependent pair:
+validity predicate which assigns shapes to positions. In Coq, we use a boolean
+function and a coercion from \coq{bool} to \coq{Prop} to represent the decidable
+predicate \coq{valid}, similarly to SSReflect~\cite{GonthierL09}. Given a
+\coq{Shape : Type}, and a type of \emph{all possible positions} \coq{APos : Type},
+the family of positions for a given shape can be defined as a dependent pair:
 \begin{coqcode}
@@
 generate OCaml code that is free of unsafe casts. The OCaml code extracted for
-\coq{Pos} will now be exactly the code extracted for \coq{APos}.
-The decidability of the validity predicate is crucial for our purposes of 
-remaining axiom-free. To illustrate this, suppose that we need to show
-the equality of two container extensions. We will need to show that, for
-the same positions, they will produce the same result. In Coq, the goal
-would look as follows: 
+\coq{Pos} will now be exactly the code extracted for \coq{APos}.  The
+decidability of the validity predicate is crucial for our purposes of  remaining
+axiom-free. To illustrate this, suppose that we need to show the equality of two
+container extensions. We will need to show that, for the same positions, they
+will produce the same result. In Coq, the goal would look as follows:
 \begin{coqcode}

### 2024-10-10 paper/paper.tex
related work

--- 2024-10-10 paper/paper.tex
@@
 approach is that they do not discuss what the resulting extracted code
-\emph{looks like} (that is, whether the extracted OCaml code resembles the
-natural formulation of the recursive function). Through experiments, we found
-that their formalisation leads to \ocaml{Obj.magic}, and code with a complex
-structure that may be hard to understand or interface with other OCaml code.
-Due to the great benefit of entirely avoiding termination proofs, it would be
-interesting to extend their approach to improve code extraction.
-
-\subsection{Termination Checking}\label{sec:termcheck}
-
-Termination and productivity are non-trivial properties of programs, and various
-methods have been proposed for checking that these properties hold. A common
-approach is guardedness checking~\cite{Eduardo94}, which is a
-\emph{syntactic} check that definitions avoid the introduction of
-non-normalizable terms. This sort of check generally looks for a
-\emph{structural decrease} of arguments in the recursive calls of a function.
-Coq uses such a check, and it works for many classic functional programming
-patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
-definitions will not be accepted by such checks.
-
-In particular, the problem of \emph{nonstructural recursion} (including
+\emph{looks like}.
+% (that is, whether the extracted OCaml code resembles the
+% natural formulation of the recursive function).
+Through experiments, we found that their formalisation leads to
+\ocaml{Obj.magic}, and code with a complex structure that may be hard to
+understand or interface with other OCaml code. Due to the great benefit of
+entirely avoiding termination proofs, it would be interesting to extend their
+approach to improve code extraction.
+
+\subsection{Termination Checking of Nonstructural Recursion}\label{sec:termcheck}
+
+% Termination and productivity are non-trivial properties of programs, and various
+% methods have been proposed for checking that these properties hold. A common
+% approach is guardedness checking~\cite{Eduardo94}, which is a
+% \emph{syntactic} check that definitions avoid the introduction of
+% non-normalizable terms. This sort of check generally looks for a
+% \emph{structural decrease} of arguments in the recursive calls of a function.
+% Coq uses such a check, and it works for many classic functional programming
+% patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
+% definitions will not be accepted by such checks.
+
+%In particular,
+The problem of \emph{nonstructural recursion} (including
 divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
@@
 types, using a lazy stream abstraction to represent sequences and fuse
-transformations like \haskell{map} and \haskell{fold}~\cite{coutts07}. Both
-approaches eliminate redundant structures and ensure computations happen in a
-single efficient pass over the data.
-
-Another dimension of fusion is \emph{tupling}, a technique used to combine
-multiple recursive functions into a single function by grouping their results
-into a tuple~\cite{hughes89}. This approach helps avoid redundant traversals of
-the same data structure when multiple operations are applied.
+transformations like \haskell{map} and \haskell{fold}~\cite{coutts07}.
+% Both approaches eliminate redundant structures and ensure computations happen in
+% a single efficient pass over the data.
+
+% Another dimension of fusion is \emph{tupling}, a technique used to combine
+% multiple recursive functions into a single function by grouping their results
+% into a tuple~\cite{hughes89}. This approach helps avoid redundant traversals of
+% the same data structure when multiple operations are applied.
 % In addition to
@@
 % data structures.
+
+In the context of Coq, Danvy explores the use of fold/unfold lemmas to mechanize
+equational reasoning about recursive programs in the Coq proof
+assistant~\cite{DANVY_2022}. By formalizing recursive program transformations,
+fold-unfold lemmas enable optimizations like fusion by allowing programs to be
+systematically simplified, eliminating intermediate structures.
 

### 2024-10-10 paper/paper.tex
minro

--- 2024-10-10 paper/paper.tex
@@
 
-To the best of our knowledge, this paper presents the first Coq formalisation of
+This paper presents the first Coq formalisation of
 a recursion scheme, called the \emph{hylomorphism}, along with its algebraic
@@
 \label{sec:recursion-schemes}
-\dcas{Do not cut below}
 Recursion schemes provide an abstract way to consume and generate data. We now

### 2024-10-10 paper/paper.tex
more related work on fusion

--- 2024-10-10 paper/paper.tex
@@
 
-In the context of Coq, Danvy explores the use of fold/unfold lemmas to mechanize
+In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
 equational reasoning about recursive programs in the Coq proof
@@
 fold-unfold lemmas enable optimizations like fusion by allowing programs to be
-systematically simplified, eliminating intermediate structures.
+systematically simplified, eliminating intermediate structures. Additionally, an
+approach to deforestation called the worker/wrapper transformation~\cite{gill09}
+was formalised and mechanised in Agda~\cite{SCULTHORPE_HUTTON_2014}.
 

### 2024-10-10 paper/paper.tex
Stuff

--- 2024-10-10 paper/paper.tex
@@
 (* Goal : f2 \o hylo g1 h1 =e hylo g2 h1 *)
-
 apply hylo_uniq.
@@
 rewrite fmap_comp.
-       (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
-rewrite [...]. (* rearranging by associativity *)
+(* rearranging by associativity *)
        (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
 rewrite <- E2.
-       (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1      *)
-rewrite [...]. (* rearranging by associativity *)
+(* rearranging by associativity *)
        (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o fmap (hylo g1 h1)) \o h1)      *)
@@
 Defined.
-
 Definition c_split : Coalg (TreeF unit int) (list int).
@@
   end.
-
 Definition knapsack_alg (wvs : list (nat * int))
@@
 
-In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
+In the context of Coq, Danvy explores the use of fold/unfold lemmas to mechanize
 equational reasoning about recursive programs in the Coq proof
@@
 fold-unfold lemmas enable optimizations like fusion by allowing programs to be
-systematically simplified, eliminating intermediate structures. Additionally, an
-approach to deforestation called the worker/wrapper transformation~\cite{gill09}
-was formalised and mechanised in Agda~\cite{SCULTHORPE_HUTTON_2014}.
+systematically simplified, eliminating intermediate structures.
 

### 2024-10-10 paper/paper.tex
improve fusion subsection

--- 2024-10-10 paper/paper.tex
@@
 In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
-equational reasoning about recursive programs in the Coq proof
-assistant~\cite{DANVY_2022}. By formalizing recursive program transformations,
-fold-unfold lemmas enable optimizations like fusion by allowing programs to be
-systematically simplified, eliminating intermediate structures. Additionally, an
-approach to deforestation called the worker/wrapper transformation~\cite{gill09}
-was formalised and mechanised in Agda~\cite{SCULTHORPE_HUTTON_2014}.
+equational reasoning about recursive programs~\cite{DANVY_2022}. Fold-unfold
+lemmas enable optimizations like fusion by allowing programs to be
+systematically simplified, including eliminating intermediate structures.
+Additionally, an approach to deforestation called the worker/wrapper
+transformation~\cite{gill09} was formalised and mechanised in
+Agda~\cite{SCULTHORPE_HUTTON_2014}.
 

### 2024-10-10 paper/paper.tex

--- 2024-10-10 paper/paper.tex
@@
 
-In the context of Coq, Danvy explores the use of fold/unfold lemmas to mechanize
-equational reasoning about recursive programs in the Coq proof
-assistant~\cite{DANVY_2022}. By formalizing recursive program transformations,
-fold-unfold lemmas enable optimizations like fusion by allowing programs to be
-systematically simplified, eliminating intermediate structures.
+In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
+equational reasoning about recursive programs~\cite{DANVY_2022}. Fold-unfold
+lemmas enable optimizations like fusion by allowing programs to be
+systematically simplified, including eliminating intermediate structures.
+Additionally, an approach to deforestation called the worker/wrapper
+transformation~\cite{gill09} was formalised and mechanised in
+Agda~\cite{SCULTHORPE_HUTTON_2014}.
 

### 2024-10-10 paper/paper.tex
Stuff

--- 2024-10-10 paper/paper.tex
@@
 however, we chose to use Coq's propositional equality, since this leads
-to simpler code. We explain why in Section~\ref{sec:algebras}, and the
+to simpler code. We explain why in Section~\ref{sec:coq-algebras}, and the
 mechanisation of initial algebras of container extensions.
@@
 proceed onto describing how to formalise hylomorphisms in Coq. We first
-formalise algebras for container extensions (Section~\ref{sec:algebras}), then
-we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
+formalise algebras for container extensions (Section~\ref{sec:coq-algebras}), then
+we formalise coalgebras (Section~\ref{sec:coq-coalg}) and then we put together
 these notions to formalise recursive coalgebras and hylomorphisms
-(Section~\ref{sec:rec-coalgebras}).
+(Section~\ref{sec:coq-rec-coalgebras}).
 
 \subsection{Algebras and Catamorphisms for Containers}
-\label{sec:algebras}
+\label{sec:coq-algebras}
 
@@
 \subsection{Coalgebras and Anamorphisms}
-\label{sec:coalg}
+\label{sec:coq-coalg}
 
@@
 \subsection{Mechanising Hylomorphisms}
-\label{sec:rec-coalgebras}
+\label{sec:coq-rec-coalgebras}
 Recall that hylomorphisms capture the concept of \emph{divide-and-conquer}
@@
 fold $a \circ F \; f \circ c$ into $f$.
-\begin{figure}
+\begin{figure}[t!]
 \begin{coqcode}
@@
 rewrite fmap_comp.
-(* rearranging by associativity *)
+rewrite ... (* rearranging by associativity *)
        (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
 rewrite <- E2.
-(* rearranging by associativity *)
+rewrite ... (* rearranging by associativity *)
        (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o fmap (hylo g1 h1)) \o h1)      *)
@@
 \coq{true}, and the right position \coq{false}.
-\begin{figure}
+\begin{figure}[t!]
 \begin{minipage}{.52\textwidth}

### 2024-10-10 paper/paper.tex
minor

--- 2024-10-10 paper/paper.tex
@@
 systematically simplified, including eliminating intermediate structures.
-Additionally, an approach to deforestation called the worker/wrapper
+Additionally, an approach to fusion called the worker/wrapper
 transformation~\cite{gill09} was formalised and mechanised in

### 2024-10-10 paper/paper.tex
title

--- 2024-10-10 paper/paper.tex
@@
 
-\title{Mechanising Recursion Schemes with Magic-Free Coq Extraction}
+\title{Mechanising Hylomorphisms for Extraction of Executable Code}
 
@@
 (* Goal : f2 \o hylo g1 h1 =e hylo g2 h1 *)
+
 apply hylo_uniq.
@@
 rewrite fmap_comp.
-(* rearranging by associativity *)
+       (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
+rewrite [...]. (* rearranging by associativity *)
        (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
 rewrite <- E2.
-(* rearranging by associativity *)
+       (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1      *)
+rewrite [...]. (* rearranging by associativity *)
        (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o fmap (hylo g1 h1)) \o h1)      *)
@@
 Defined.
+
 Definition c_split : Coalg (TreeF unit int) (list int).
@@
   end.
+
 Definition knapsack_alg (wvs : list (nat * int))
@@
 
-In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
-equational reasoning about recursive programs~\cite{DANVY_2022}. Fold-unfold
-lemmas enable optimizations like fusion by allowing programs to be
-systematically simplified, including eliminating intermediate structures.
-Additionally, an approach to fusion called the worker/wrapper
-transformation~\cite{gill09} was formalised and mechanised in
-Agda~\cite{SCULTHORPE_HUTTON_2014}.
+In the context of Coq, Danvy explores the use of fold/unfold lemmas to mechanize
+equational reasoning about recursive programs in the Coq proof
+assistant~\cite{DANVY_2022}. By formalizing recursive program transformations,
+fold-unfold lemmas enable optimizations like fusion by allowing programs to be
+systematically simplified, eliminating intermediate structures.
 

### 2024-12-04 ESOP/rebuttal.txt
rebuttal

Reviewer A
------------------------------------
I didn't get a clear sense from the paper of the depth and signiicance of the
contribution.  Setoid based reasoning is well understood (if painful).
Containers have been studied and also formalized before, though I think mostly
in Agda and perhaps mostly in settings assuming axioms that might complicate
code extraction.  The material on hylomorphims and exampls itself is not new.
So it seems what is new is how to put the pieces together to avoid the various
pitfalls in formalizing hylomorphisms in Coq to maintain the ability to generate
OCaml code.  The code that is generated still looks like it would need
additional work to incorporate into a program dealing with say ordinary lists or
trees defined using OCaml datatypes rather than as container equivalents.  So
I'm not fully persuaded that the benefits of the end results make the paper
compelling on their own either.

Reviewer B
------------------------------------
A criticism is that I am struggling to position this work in a particular area.
Is this an elaborate new Coq library that's going to be useful to lots of
people? Does it provide any major take-away lessons for Coq programming? I'm not
sure. Above I have guessed what the take-aways are, but this is not summarized
in the introduction.

Reviewer C
------------------------------------
I could not understand what is the real motivation of this study. Is it a case
study of various Coq features, including the OCaml-code extraction function? Or,
in a real situation, does an OCaml programmer use this to write a program? For
the latter, why does using Coq help with which problem of hylomorphisms?

Although the paper gives gentle explanations here and there, I felt that there
is little self-contained and detailed explanation on the termination issue.

Overall, I felt that the explanation of the motivation is insufficient, while
there seems to be no serious problem in the paper, and the authors at least work
out some amount of a case study on OCaml extraction from Coq. I have no strong
opinion for rejection or acceptance, and I think that there seems to be no
problem if this paper is accepted.

### 2024-12-04 ESOP/rebuttal.txt
Draft

@@
+We would like to thank the reviewers for their constructive feedback, which has
+helped us clarify our contributions and refine the positioning of our work.
+Below, we address the concerns raised by each reviewer, emphasizing the novelty,
+motivation, and significance of our contributions.
+
+Overall we think the reviewers have a sound understanding of the paper. We would
+like to emphasize that our approach uniquely combines a formal and sound OCaml
+extraction of recursion schemes within Coq, addressing termination challenges
+without relying on axioms. This is the first formalisation of recursion schemes
+within a type theory which ensures termination of definitions as opposed to
+current presentations of recursion schemes in, e.g., Haskell which can be broken
+by using non-terminating arguments to the recursion scheme.
+
+Using Coq extraction mechanism we can provide the functional programmer with a
+termination guarantee and at the same time provide the Coq programmer with a
+library which accepts more recursive definitions.
+
+Revisions will highlight practical insights for Coq programming, expand on
+termination handling, and articulate the broader significance of bridging formal
+reasoning with practical code generation.
+
+We will refine the introduction to clearly summarize these
+contributions and position the work within the broader landscape.
+
+Summary of Revisions
+---------------------------------------------------------------------------
+In response to the reviewers' comments, we will make the following changes:
+
+- Clarify Motivation and Contributions: Revise the introduction to clearly
+articulate the motivation, contributions, and broader significance of the work,
+positioning it as both a methodological advancement and a practical framework.
+
+- Highlight Novelty: Emphasize how our work goes beyond prior formalizations by
+addressing specific challenges in Coq related to termination and extraction,
+which have not been explored in depth in Agda or other systems.
+
+- Explanation of Termination: Include a detailed and self-contained discussion of
+termination issues and how our framework resolves them.  Future Work: Add a
+discussion on extending the framework for better integration with standard OCaml
+datatypes and libraries.
+
+We hope these clarifications address the reviewers' concerns and demonstrate the
+value of our contributions. Thank you for your thoughtful and detailed feedback.
+
 Reviewer A
 ------------------------------------
-I didn't get a clear sense from the paper of the depth and signiicance of the
-contribution.  Setoid based reasoning is well understood (if painful).
-Containers have been studied and also formalized before, though I think mostly
-in Agda and perhaps mostly in settings assuming axioms that might complicate
-code extraction.  The material on hylomorphims and exampls itself is not new.
-So it seems what is new is how to put the pieces together to avoid the various
-pitfalls in formalizing hylomorphisms in Coq to maintain the ability to generate
-OCaml code.  The code that is generated still looks like it would need
-additional work to incorporate into a program dealing with say ordinary lists or
-trees defined using OCaml datatypes rather than as container equivalents.  So
-I'm not fully persuaded that the benefits of the end results make the paper
-compelling on their own either.
+Comment:
+> Setoid-based reasoning is well understood (if painful). Containers have been
+> studied and also formalized before, though I think mostly in Agda...The material
+> on hylomorphisms and examples itself is not new...So it seems what is new is how
+> to put the pieces together to avoid the various pitfalls in formalizing
+> hylomorphisms in Coq to maintain the ability to generate OCaml code.
+
+We agree with the reviewer that setoid reasoning, hylomorphisms and containers
+are established topics. As correctly pointed out by the reviewer however they
+have never been pieced together. In particular, the whole area of recursion
+schemes has only been considered within the broader realm of functional
+programming. In this setting, it is still possible to introduce non-termination,
+e.g. by feeding the recursion scheme with invalid non-terminating algebras.
+Hence recursion scheme have mostly been used to structure programs for reasoning
+rather than ensuring formally that they define terminating functions.
+
+This work can be viewed as a first step towards considering the use of recursion
+ schemes to help proof assistants to accept more recursive definitions.
+
+The issue of extracting runnable code is perhaps tangential but still helpful
+from a programming point of view as 1) OCaml does not have full access to
+recursion schemes since there is no way of declaring a functor as easily as for
+example in Haskell and 2) it provides
+
+Comment: "The code that is generated still looks like it would need additional
+work to incorporate into a program dealing with, say, ordinary lists or trees
+defined using OCaml datatypes rather than as container equivalents."
+
+[DAVID TO DO]
 
@@
 ------------------------------------
-A criticism is that I am struggling to position this work in a particular area.
-Is this an elaborate new Coq library that's going to be useful to lots of
-people? Does it provide any major take-away lessons for Coq programming? I'm not
-sure. Above I have guessed what the take-aways are, but this is not summarized
-in the introduction.
+Comment:
+> I am struggling to position this work in a particular area...Is this
+> an elaborate new Coq library that’s going to be useful to lots of people? Does
+> it provide any major take-away lessons for Coq programming?
+
+Our primary goal is to provide a framework that advances the formalization and
+practical usability of recursive schemes like hylomorphisms in Coq, particularly
+with a focus on sound OCaml code extraction. While this work includes the
+implementation of a Coq library, its broader significance lies in showcasing how
+termination-sensitive constructs like hylomorphisms can be handled rigorously
+within Coq's framework, offering insights into programming techniques that
+balance formal reasoning with practical usability.
+
+Comment:
+> Above I have guessed what the take-aways are, but this is not
+> summarized in the introduction.
+
+Thank you for highlighting this. We will revise the introduction to explicitly
+summarize the key contributions and take-away lessons from our work, including:
+
+A novel approach to formalizing hylomorphisms in Coq while preserving code
+extraction capabilities.  A case study demonstrating practical challenges and
+solutions in Coq's termination handling.  Insights into bridging the gap between
+formal proof systems and practical programming.
 
@@
 ------------------------------------
-I could not understand what is the real motivation of this study. Is it a case
-study of various Coq features, including the OCaml-code extraction function? Or,
-in a real situation, does an OCaml programmer use this to write a program? For
-the latter, why does using Coq help with which problem of hylomorphisms?
+Comment:
+> I could not understand what is the real motivation of this study. Is
+> it a case study of various Coq features, including the OCaml-code extraction
+> function?
 
-Although the paper gives gentle explanations here and there, I felt that there
-is little self-contained and detailed explanation on the termination issue.
+The primary motivation for this work is to address the challenge of formalizing
+recursive schemes like hylomorphisms in a manner compatible with Coq’s rigorous
+type system and termination requirements, while enabling sound and practical
+code extraction to OCaml. This is not merely a case study of Coq features but
+rather a contribution to the methodology of combining formal reasoning with
+executable code generation, particularly for functional programming constructs.
 
-Overall, I felt that the explanation of the motivation is insufficient, while
-there seems to be no serious problem in the paper, and the authors at least work
-out some amount of a case study on OCaml extraction from Coq. I have no strong
-opinion for rejection or acceptance, and I think that there seems to be no
-problem if this paper is accepted.
+Comment:
+> In a real situation, does an OCaml programmer use this to write a
+> program? For the latter, why does using Coq help with which problem of
+> hylomorphisms?
+
+Using Coq to formalize and extract hylomorphisms addresses critical issues of
+correctness and termination that are often left informal in direct OCaml
+implementations. By using Coq, we ensure that recursive schemes are well-founded
+and free from subtle bugs or inconsistencies. While the extracted code might
+require adaptation for direct use by OCaml programmers, the guarantees provided
+by Coq add significant value in contexts requiring high assurance.
+
+Comment:
+> I felt that there is little self-contained and detailed explanation on
+> the termination issue.
+
+We acknowledge this and will expand the paper to include a more self-contained
+explanation of the termination handling in our approach, detailing how Coq's
+type system and proof obligations are leveraged to ensure well-foundedness in
+recursive constructs.

### 2024-12-04 paper/paper.tex
Draft

--- 2024-10-10 paper/paper.tex
@@
 
-% \author{David Castro-Perez}{School of Computing, University of Kent}{d-castro-perez@kent.ac.uk}{}{}
-% \author{Marco Paviotti}{School of Computing, University of Kent}{m.paviotti@kent.ac.uk}{}{}
-% \author{Michael Vollmer}{School of Computing, University of Kent}{m.vollmer@kent.ac.uk}{}{}
+\author{David Castro Perez\inst{1}\orcidID{0000-0002-6939-4189} \and
+Marco Paviotti\inst{2,3}\orcidID{0000-0002-1513-0807} \and
+Michael Vollmer\inst{3}\orcidID{0000-0002-0496-8268}}
+%
+\authorrunning{D. Castro Perez et al.}
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+%
+\institute{University Of Kent, Canterbury, CT2 7NZ, United Kingdom\\
+\email{\{D.Castro-Perez, M.Paviotti, M.Vollmer\}@kent.ac.uk}}
 
@@
 %%% TeX-command-extra-options: "-shell-escape"
+%%% TeX-engine: luatex
 %%% End:

### 2024-12-04 ESOP/rebuttal.txt
deleted some responses.

@@
 
-Thank you for highlighting this. We will revise the introduction to explicitly
-summarize the key contributions and take-away lessons from our work, including:
-
-A novel approach to formalizing hylomorphisms in Coq while preserving code
-extraction capabilities.  A case study demonstrating practical challenges and
-solutions in Coq's termination handling.  Insights into bridging the gap between
-formal proof systems and practical programming.
+Thank you for highlighting this. We will revise the introduction accordingly.
 
@@
 
-The primary motivation for this work is to address the challenge of formalizing
-recursive schemes like hylomorphisms in a manner compatible with Coq’s rigorous
-type system and termination requirements, while enabling sound and practical
-code extraction to OCaml. This is not merely a case study of Coq features but
-rather a contribution to the methodology of combining formal reasoning with
-executable code generation, particularly for functional programming constructs.
-
 Comment:
@@
 
-Using Coq to formalize and extract hylomorphisms addresses critical issues of
-correctness and termination that are often left informal in direct OCaml
-implementations. By using Coq, we ensure that recursive schemes are well-founded
-and free from subtle bugs or inconsistencies. While the extracted code might
-require adaptation for direct use by OCaml programmers, the guarantees provided
-by Coq add significant value in contexts requiring high assurance.
-
 Comment:
@@
 > the termination issue.
-
-We acknowledge this and will expand the paper to include a more self-contained
-explanation of the termination handling in our approach, detailing how Coq's
-type system and proof obligations are leveraged to ensure well-foundedness in
-recursive constructs.

### 2024-12-04 ESOP/rebuttal.txt
minor

@@
 
-Summary of Revisions
+Summary of Planned Revisions
 ---------------------------------------------------------------------------

### 2024-12-04 ESOP/rebuttal.txt
Summary

--- 2024-12-04 ESOP/rebuttal.txt
@@
 helped us clarify our contributions and refine the positioning of our work.
-Below, we address the concerns raised by each reviewer, emphasizing the novelty,
-motivation, and significance of our contributions.
 
-Overall we think the reviewers have a sound understanding of the paper. We would
-like to emphasize that our approach uniquely combines a formal and sound OCaml
-extraction of recursion schemes within Coq, addressing termination challenges
-without relying on axioms. This is the first formalisation of recursion schemes
-within a type theory which ensures termination of definitions as opposed to
-current presentations of recursion schemes in, e.g., Haskell which can be broken
-by using non-terminating arguments to the recursion scheme.
+A common theme among the reviewers is the need for a clearer presentation of the
+motivation behind our work, which we address here in summary.
 
-Using Coq extraction mechanism we can provide the functional programmer with a
-termination guarantee and at the same time provide the Coq programmer with a
-library which accepts more recursive definitions.
+While the individual components of this work are well-known, as correctly noted
+by Reviewer A, their integration into a cohesive framework has not been
+previously studied. Although this might initially seem like a limitation, we
+argue that it represents an essential necessary step forward in the area of
+recursion schemes in the context of type theory.
 
-Revisions will highlight practical insights for Coq programming, expand on
-termination handling, and articulate the broader significance of bridging formal
-reasoning with practical code generation.
+In fact, recursion schemes have been explored extensively in the context of
+functional programming languages like Haskell, where their theoretical soundness
+has been shown by Hinze, Wu and Gibbons. However, in practice, these schemes
+have never been used in practice to guarantee termination, simply because
+Haskell allows invalid algebras and functors to be used with them.
 
-We will refine the introduction to clearly summarize these
-contributions and position the work within the broader landscape.
+To address this gap, our work encodes recursion schemes within a proof assistant
+-- Coq -- ensuring termination by design and enabling extraction to executable
+code in a programming language. Containers and setoids are used as standard
+tools to achieve this goal.
 
-Summary of Revisions
----------------------------------------------------------------------------
-In response to the reviewers' comments, we will make the following changes:
+We believe this approach benefits both the programming and type theory
+communities. For programmers, it offers a practical library that ensures
+termination and correctness. For type theorists, it contributes to addressing
+the long-standing challenge of defining recursive constructions in systems like
+Coq or Agda.
 
-- Clarify Motivation and Contributions: Revise the introduction to clearly
-articulate the motivation, contributions, and broader significance of the work,
-positioning it as both a methodological advancement and a practical framework.
-
-- Highlight Novelty: Emphasize how our work goes beyond prior formalizations by
-addressing specific challenges in Coq related to termination and extraction,
-which have not been explored in depth in Agda or other systems.
-
-- Explanation of Termination: Include a detailed and self-contained discussion of
-termination issues and how our framework resolves them.  Future Work: Add a
-discussion on extending the framework for better integration with standard OCaml
-datatypes and libraries.
-
-We hope these clarifications address the reviewers' concerns and demonstrate the
-value of our contributions. Thank you for your thoughtful and detailed feedback.
+We believe our work is a first step in try to understand how recursion schemes
+can be integrated in type theories to tackle termination challenges. We intend
+to have a more theoretical follow up addressing corecursive algorithms.
 

### 2024-12-04 ESOP/dcastrop_rebuttal.txt
Rebuttal Version Dcastrop

We would like to thank the reviewers for their constructive feedback, which has
helped us clarify our contributions and refine the positioning of our work.

A common theme among the reviewers is the need for a clearer presentation of the
motivation behind our work, which we address here in summary.

The main motivation of our work is to build a verified framework for program
calculation and code extraction. The state-of-the art does not address
divide-and-conquer algorithms, and does not address the extraction of idiomatic
code.

While the individual components of this work are well-known, as correctly noted
by Reviewer A, their integration into a cohesive framework has not been
previously studied. Moreover, their use to formalise a divide-and-conquer
recursion scheme (hylomorphisms) in a type theory has also never been explored.
One key benefit of our approach is that we allow the use of regular Coq proofs
to do program calculation, i.e. derive efficient implementations from simple
specifications. Unlike previous work, we can encode divide-and-conquer
computations, and our mechanisation extracts idiomatic OCaml code. We note in
Section 6.1 that previous formalisations of recursion schemes in proof
assistants focus on maps and folds. Only Abreu et al. (2023) encode divide and
conquer computations, but in a way that does not lead to idiomatic code
extraction (Section 6.2). Code extraction is not straightforward, and it
requires the careful selection of the representations of containers, functors,
etc, such as the use of decidable predicates to define indexed families of
types.

Furthermore, the use of hylomorphisms in a type theory has some benefits
regarding termination proofs. Unlike common approaches to encode non-structural
recursion in a proof assistant, the use of hylomorphisms allows to focus the
proofs on showing that the coalgebra is recursive. This means that termination
proofs become more modular, and can be plugged in to build different
divide-and-conquer computations. Our framework allows programmers to build
libraries of recursive coalgebras for different datatypes, which can be reused
when building other non-structural recursive functions. The encoding of
recursive coalgebras to prove termination has not been explored before.

Our work allows to encode program calculation proofs in a natural way that,
until today, are not machine-checked. Furthermore, we believe our work is a
first step in try to understand how recursion schemes can be integrated in type
theories to tackle termination challenges. We intend to have a more theoretical
follow up addressing corecursive algorithms.

Review A
===========================================================================

> This paper investigates the challenges of formalizing reasoning about
> recursion schemes (cata, ana, hylomorphisms) in Coq in a way that is
> compatible with code extraction.  Other approaches have run into problems,
> first because [...]

It is important to note that, as we point out in Section 6.1, other approaches
do not formalise divide-and-conquer computations. We allow them due to our
encoding of recursive coalgebras.

> The code that is generated still looks like it would need additional work to
> incorporate into a program dealing with say ordinary lists or trees defined
> using OCaml datatypes rather than as container equivalents.

The use of ordinary lists and trees is not only possible, but easy. In fact,
containers are only used internally in our examples, the input/output types use
regular OCaml datatypes. We do that by defining them as initial algebras of the
respective container extension (e.g. ListF, or TreeF in Section 5). By doing
that, we get (for free) that OCaml lists/trees/etc are isomorphic to `Fix
ListF` or `Fix TreeF`.  To further simplify interacting with these container
extensions, it should be straightforward (even if tedious) to automate the
translation between a strictly positive datatype and its container equivalent
(e.g. Section 6 of Abott et al. (2005), or our own encoding of polynomial
functors as containers in Coq).

> p12.  I the definition of comp, I was surprised that the arrows between (B ~>
> C) and (A ~> B) and (A ~> C) are not ordinary function arrows.  I see this is
> also the way it is in the formalization, is there a particular reason it has
> to be this way or just for convenience?

It simplifies the proofs. In particular, if we have that `f =e g`, then we can
trivially conclude that `f \o h =e g \o h`. Otherwise, we would need to do this
proof manually every time it is required.

Review B
===========================================================================

> * "least" fixed point, what does "least" mean here? and anyway that is not what Lambek's lemma says.

[TODO MARCO]

> * "pronounced catamorphism" I think this is functional programming terminology, not category theory

[TODO MARCO]

> * "Note that since Haskell". I didn't understand the point of this paragraph. I do think Haskell is a bad language for demonstrating this.

> Section 2.3. Again, the terminal object might not exist, "greatest" doesn't make sense, etc..

[TODO MARCO: does "greatest" not make sense?? I thought it was standard terminology]

> Page 8. "TreeF". This is not a tree, and it's never used as a tree. It's just a pair of lists with a midpoint.

The least/greatest fixpoint of TreeF is an inductive/coinductive tree.

> Page 9. Quicksort is not a solution to this equation, rather, "sort" is a
> solution. These are functions between sets, and so cannot distinguish
> intensional algorithmic specifics.

The solution to this equation behaves exactly as the naive implementation of
Quicksort in Haskell/OCaml.

I did not understand the recursive anamorphisms paragraph, but it seems to be a digression.

> Does the extra variable x0 have a performance cost?

The extra `x0` can be trivially inlined in a compiler pass. We are also
considering extending Coq's code extraction with manual inlining hints.

Review C
===========================================================================

> I could not understand what is the real motivation of this study.
> does an OCaml programmer use this to write a program?

Our end goal is to build a framework for program verification based on program
calculation techniques. In that sense, our framework is meant to make the use
of program calculation more accessible to programmers. Prior to this work,
using program calculation for divide-and-conquer computations would require
pen-and-paper proofs and a manual translation to Haskell/OCaml code.

> For the latter, why does using Coq help with which problem of hylomorphisms?

Hylomorphisms are not guaranteed to exist, or to be unique, due to termination
issues. If a hylomorphism can be encoded in our framework in Coq, it must exist
and be unique.

### 2024-12-05 ESOP/dcastrop_rebuttal.txt
Marco's TODOs

@@
 
-[TODO MARCO]
+We don't quite understand this comment.
+
+We define least fixed-point for F as the initial object in the category F-iso of
+fixed-points for F and F-iso homomorphisms.  By Lambek's lemma an initial
+F-algebra is a fixed-point, but in particular also the least fixed-point,
+because we can take any iso X ~ FX, and prove that there exists a unique F-iso
+morphism (mu F ---> X) simply because X is an F-algebra.
+
+An initial F-algebra is also a fixed-point for F and in particular is also the least.
 
@@
 
-[TODO MARCO]
+we agree, although despite the FP community has adopted their own terminology
+what they're indeed referring to the exact same thing.
 
-> * "Note that since Haskell". I didn't understand the point of this paragraph. I do think Haskell is a bad language for demonstrating this.
+> * "Note that since Haskell". I didn't understand the point of this paragraph.
+    I do think Haskell is a bad language for demonstrating this.
+
+The point here is that Haskell has equi-recursive types which are both inductive
+and coinductive data types.  However, when programming with recursion schemes it
+is common to make inductive/coinductive types distinguished by defining a Fix
+and CoFix operator and then imposing that catamorphisms are used on Fix and
+anamorphisms are used on CoFix.
 
@@
 
-[TODO MARCO: does "greatest" not make sense?? I thought it was standard terminology]
+The argument might just be the dual of what we explained earlier.
+The terminal F-coalgebra might not exist, but if it does it corresponds to the greatest fixed-point of the functor F.
 
@@
 
-The least/greatest fixpoint of TreeF is an inductive/coinductive tree.
+The least/greatest fixpoint of TreeF is an inductive/coinductive tree. The name TreeF it refers
+to the fact that it is the functor giving rise to the a tree via the fixed-point operator.
 
@@
 
-I did not understand the recursive anamorphisms paragraph, but it seems to be a digression.
+> I did not understand the recursive anamorphisms paragraph, but it seems to be a digression.
+
+We do use recursive anamorphisms in page 18, paragraph "On the subtype of finite
+elements".  In particular an anamorphism applied to a recursive coalgebra yields
+a finite F-tree (for a functor F).
+
+Consider the Coq type
+
+Inductive FinF : GFix F -> Prop
+
+Then {x : GFix F | FinF x} is the subtype of finite elements (of a coinductive
+data type) which is isomorphic to its inductive counterpart.
 

### 2024-12-05 ESOP/dcastrop_rebuttal.txt
Including Mike's comments'

@@
 
-The use of ordinary lists and trees is not only possible, but easy. In fact,
+@@MIKE: This is a valid observation, in that the use of containers in our examples
+results in a slight mismatch between the extracted OCaml code and, for example,
+existing code/libraries that are expecting more ordinary datatypes for trees or
+lists. This could be ameliorated with either a minor amount of marshalling, or
+even a straightforward automatic translation. However, the fact that this
+extracted code contains no unsafe coercions, and contains minimal noise imposed
+by the details of the Coq formalism, make it possible at all for the resulting
+OCaml code to be usable. As we detail in the paper, prior related work does not
+meet those criteria.
+
+@@David: The use of ordinary lists and trees is not only possible, but easy. In fact,
 containers are only used internally in our examples, the input/output types use
@@
 ===========================================================================
+We appreciate Reviewer C's thoughtful question regarding the motivation of our
+study. To clarify, the central contribution of our work lies in proposing a
+cohesive framework for recursion schemes in a proof assistant such as Coq, and
+the primary *application* of this framework is the verification of
+divide-and-conquer algorithms in Coq with clean extraction to readable and
+idiomatic OCaml. In contrast, prior work on recursion schemes in Haskell has
+largely relied on informal, pen-and-paper proofs and manual, unverified
+implementations in Haskell, leaving a gap in the ability to produce formally
+verified and automatically extracted implementations. Our framework addresses
+this gap by integrating formal verification and code generation into a single,
+cohesive process.
 

### 2024-12-05 ESOP/dcastrop_rebuttal.txt
Rebuttal

@@
 
-@@MIKE: This is a valid observation, in that the use of containers in our examples
-results in a slight mismatch between the extracted OCaml code and, for example,
-existing code/libraries that are expecting more ordinary datatypes for trees or
-lists. This could be ameliorated with either a minor amount of marshalling, or
-even a straightforward automatic translation. However, the fact that this
-extracted code contains no unsafe coercions, and contains minimal noise imposed
-by the details of the Coq formalism, make it possible at all for the resulting
-OCaml code to be usable. As we detail in the paper, prior related work does not
-meet those criteria.
-
-@@David: The use of ordinary lists and trees is not only possible, but easy. In fact,
+The use of ordinary lists and trees is not only possible, but easy. In fact,
 containers are only used internally in our examples, the input/output types use
@@
 that, we get (for free) that OCaml lists/trees/etc are isomorphic to `Fix
-ListF` or `Fix TreeF`.  To further simplify interacting with these container
-extensions, it should be straightforward (even if tedious) to automate the
-translation between a strictly positive datatype and its container equivalent
-(e.g. Section 6 of Abott et al. (2005), or our own encoding of polynomial
-functors as containers in Coq).
+ListF` or `Fix TreeF`.
+
+It is, however, a valid observation, in that the use of containers may result
+in a slight mismatch between the extracted OCaml code and, for example,
+existing code/libraries that are expecting more ordinary datatypes for trees or
+lists. This could be ameliorated with either a minor amount of marshalling, or
+even a straightforward automatic translation (e.g. relying on an encoding of
+polynomial functors as containers -- we provide one in our implementation).
+However, the fact that this extracted code contains no unsafe coercions, and
+contains minimal noise imposed by the details of the Coq formalism, make it
+possible at all for the resulting OCaml code to be usable. As we detail in the
+paper, prior related work does not meet those criteria.
 
@@
 
-> * "least" fixed point, what does "least" mean here? and anyway that is not what Lambek's lemma says.
+> * "least" fixed point, what does "least" mean here? and anyway that is not
+> what Lambek's lemma says.
 
@@
 
-we agree, although despite the FP community has adopted their own terminology
+We agree, although despite the FP community has adopted their own terminology
 what they're indeed referring to the exact same thing.

### 2024-12-05 ESOP/rebuttal.txt
Update

@@
 
+The main motivation of our work is to build a verified framework for program
+calculation and code extraction. The state-of-the art does not address
+divide-and-conquer algorithms, and does not address the extraction of idiomatic
+code.
+
 While the individual components of this work are well-known, as correctly noted
 by Reviewer A, their integration into a cohesive framework has not been
-previously studied. Although this might initially seem like a limitation, we
-argue that it represents an essential necessary step forward in the area of
-recursion schemes in the context of type theory.
+previously studied. Moreover, their use to formalise a divide-and-conquer
+recursion scheme (hylomorphisms) in a type theory has also never been explored.
+One key benefit of our approach is that we allow the use of regular Coq proofs
+to do program calculation, i.e. derive efficient implementations from simple
+specifications. Unlike previous work, we can encode divide-and-conquer
+computations, and our mechanisation extracts idiomatic OCaml code. We note in
+Section 6.1 that previous formalisations of recursion schemes in proof
+assistants focus on maps and folds. Only Abreu et al. (2023) encode divide and
+conquer computations, but in a way that does not lead to idiomatic code
+extraction (Section 6.2). Code extraction is not straightforward, and it
+requires the careful selection of the representations of containers, functors,
+etc, such as the use of decidable predicates to define indexed families of
+types.
 
-In fact, recursion schemes have been explored extensively in the context of
-functional programming languages like Haskell, where their theoretical soundness
-has been shown by Hinze, Wu and Gibbons. However, in practice, these schemes
-have never been used in practice to guarantee termination, simply because
-Haskell allows invalid algebras and functors to be used with them.
+Furthermore, the use of hylomorphisms in a type theory has some benefits
+regarding termination proofs. Unlike common approaches to encode non-structural
+recursion in a proof assistant, the use of hylomorphisms allows to focus the
+proofs on showing that the coalgebra is recursive. This means that termination
+proofs become more modular, and can be plugged in to build different
+divide-and-conquer computations. Our framework allows programmers to build
+libraries of recursive coalgebras for different datatypes, which can be reused
+when building other non-structural recursive functions. The encoding of
+recursive coalgebras to prove termination in a proof assistant has not been
+explored before.
 
-To address this gap, our work encodes recursion schemes within a proof assistant
--- Coq -- ensuring termination by design and enabling extraction to executable
-code in a programming language. Containers and setoids are used as standard
-tools to achieve this goal.
+Our work allows to encode program calculation proofs in a natural way (i.e.
+equivalent to pen-and-paper proofs) that, until today, are not machine-checked.
+Furthermore, we believe our work is a first step in try to understand how
+recursion schemes can be integrated in type theories to tackle termination
+challenges. We intend to have a more theoretical follow up addressing
+corecursive algorithms.
 
-We believe this approach benefits both the programming and type theory
-communities. For programmers, it offers a practical library that ensures
-termination and correctness. For type theorists, it contributes to addressing
-the long-standing challenge of defining recursive constructions in systems like
-Coq or Agda.
+Review A
+===========================================================================
 
-We believe our work is a first step in try to understand how recursion schemes
-can be integrated in type theories to tackle termination challenges. We intend
-to have a more theoretical follow up addressing corecursive algorithms.
+> This paper investigates the challenges of formalizing reasoning about
+> recursion schemes (cata, ana, hylomorphisms) in Coq in a way that is
+> compatible with code extraction.  Other approaches have run into problems,
+> first because [...]
 
-Reviewer A
-------------------------------------
-Comment:
-> Setoid-based reasoning is well understood (if painful). Containers have been
-> studied and also formalized before, though I think mostly in Agda...The material
-> on hylomorphisms and examples itself is not new...So it seems what is new is how
-> to put the pieces together to avoid the various pitfalls in formalizing
-> hylomorphisms in Coq to maintain the ability to generate OCaml code.
+It is important to note that, as we point out in Section 6.1, other approaches
+do not formalise divide-and-conquer computations. We allow them due to our
+encoding of recursive coalgebras.
 
-We agree with the reviewer that setoid reasoning, hylomorphisms and containers
-are established topics. As correctly pointed out by the reviewer however they
-have never been pieced together. In particular, the whole area of recursion
-schemes has only been considered within the broader realm of functional
-programming. In this setting, it is still possible to introduce non-termination,
-e.g. by feeding the recursion scheme with invalid non-terminating algebras.
-Hence recursion scheme have mostly been used to structure programs for reasoning
-rather than ensuring formally that they define terminating functions.
+> The code that is generated still looks like it would need additional work to
+> incorporate into a program dealing with say ordinary lists or trees defined
+> using OCaml datatypes rather than as container equivalents.
 
-This work can be viewed as a first step towards considering the use of recursion
- schemes to help proof assistants to accept more recursive definitions.
+The use of ordinary lists and trees is not only possible, but easy. In fact,
+containers are only used internally in our examples, the input/output types use
+regular OCaml datatypes. We do that by defining them as initial algebras of the
+respective container extension (e.g. ListF, or TreeF in Section 5). By doing
+that, we get (for free) that OCaml lists/trees/etc are isomorphic to `Fix
+ListF` or `Fix TreeF`.
 
-The issue of extracting runnable code is perhaps tangential but still helpful
-from a programming point of view as 1) OCaml does not have full access to
-recursion schemes since there is no way of declaring a functor as easily as for
-example in Haskell and 2) it provides
+It is, however, a valid observation, in that the use of containers may result
+in a slight mismatch between the extracted OCaml code and, for example,
+existing code/libraries that are expecting more ordinary datatypes for trees or
+lists. This could be ameliorated with either a minor amount of marshalling, or
+even a straightforward automatic translation (e.g. relying on an encoding of
+polynomial functors as containers -- we provide one in our implementation).
+However, the fact that this extracted code contains no unsafe coercions, and
+contains minimal noise imposed by the details of the Coq formalism, make it
+possible at all for the resulting OCaml code to be usable. As we detail in the
+paper, prior related work does not meet those criteria.
 
-Comment: "The code that is generated still looks like it would need additional
-work to incorporate into a program dealing with, say, ordinary lists or trees
-defined using OCaml datatypes rather than as container equivalents."
+> p12.  I the definition of comp, I was surprised that the arrows between (B ~>
+> C) and (A ~> B) and (A ~> C) are not ordinary function arrows.  I see this is
+> also the way it is in the formalization, is there a particular reason it has
+> to be this way or just for convenience?
 
-[DAVID TO DO]
+It simplifies the proofs. In particular, if we have that `f =e g`, then we can
+trivially conclude that `f \o h =e g \o h`. Otherwise, we would need to do this
+proof manually every time it is required.
 
-Reviewer B
-------------------------------------
-Comment:
-> I am struggling to position this work in a particular area...Is this
-> an elaborate new Coq library that’s going to be useful to lots of people? Does
-> it provide any major take-away lessons for Coq programming?
+Review B
+===========================================================================
 
-Our primary goal is to provide a framework that advances the formalization and
-practical usability of recursive schemes like hylomorphisms in Coq, particularly
-with a focus on sound OCaml code extraction. While this work includes the
-implementation of a Coq library, its broader significance lies in showcasing how
-termination-sensitive constructs like hylomorphisms can be handled rigorously
-within Coq's framework, offering insights into programming techniques that
-balance formal reasoning with practical usability.
+> * "least" fixed point, what does "least" mean here? and anyway that is not
+> what Lambek's lemma says.
 
-Comment:
-> Above I have guessed what the take-aways are, but this is not
-> summarized in the introduction.
+We don't quite understand this comment.
 
-Thank you for highlighting this. We will revise the introduction accordingly.
+We define least fixed-point for F as the initial object in the category F-iso of
+fixed-points for F and F-iso homomorphisms.  By Lambek's lemma an initial
+F-algebra is a fixed-point, but in particular also the least fixed-point,
+because we can take any iso X ~ FX, and prove that there exists a unique F-iso
+morphism (mu F ---> X) simply because X is an F-algebra.
 
-Reviewer C
-------------------------------------
-Comment:
-> I could not understand what is the real motivation of this study. Is
-> it a case study of various Coq features, including the OCaml-code extraction
-> function?
+An initial F-algebra is also a fixed-point for F and in particular is also the least.
 
-Comment:
-> In a real situation, does an OCaml programmer use this to write a
-> program? For the latter, why does using Coq help with which problem of
-> hylomorphisms?
+> * "pronounced catamorphism" I think this is functional programming terminology, not category theory
 
-Comment:
-> I felt that there is little self-contained and detailed explanation on
-> the termination issue.
+We agree, although despite the FP community has adopted their own terminology
+what they're indeed referring to the exact same thing.
+
+> * "Note that since Haskell". I didn't understand the point of this paragraph.
+    I do think Haskell is a bad language for demonstrating this.
+
+The point here is that Haskell has equi-recursive types which are both inductive
+and coinductive data types.  However, when programming with recursion schemes it
+is common to make inductive/coinductive types distinguished by defining a Fix
+and CoFix operator and then imposing that catamorphisms are used on Fix and
+anamorphisms are used on CoFix.
+
+> Section 2.3. Again, the terminal object might not exist, "greatest" doesn't make sense, etc..
+
+The argument might just be the dual of what we explained earlier.
+The terminal F-coalgebra might not exist, but if it does it corresponds to the greatest fixed-point of the functor F.
+
+> Page 8. "TreeF". This is not a tree, and it's never used as a tree. It's just a pair of lists with a midpoint.
+
+The least/greatest fixpoint of TreeF is an inductive/coinductive tree. The name TreeF it refers
+to the fact that it is the functor giving rise to the a tree via the fixed-point operator.
+
+> Page 9. Quicksort is not a solution to this equation, rather, "sort" is a
+> solution. These are functions between sets, and so cannot distinguish
+> intensional algorithmic specifics.
+
+The solution to this equation behaves exactly as the naive implementation of
+Quicksort in Haskell/OCaml.
+
+> I did not understand the recursive anamorphisms paragraph, but it seems to be a digression.
+
+We do use recursive anamorphisms in page 18, paragraph "On the subtype of finite
+elements".  In particular an anamorphism applied to a recursive coalgebra yields
+a finite F-tree (for a functor F).
+
+Consider the Coq type
+
+Inductive FinF : GFix F -> Prop
+
+Then {x : GFix F | FinF x} is the subtype of finite elements (of a coinductive
+data type) which is isomorphic to its inductive counterpart.
+
+> Does the extra variable x0 have a performance cost?
+
+The extra `x0` can be trivially inlined in a compiler pass. We are also
+considering extending Coq's code extraction with manual inlining hints.
+
+Review C
+===========================================================================
+We appreciate Reviewer C's thoughtful question regarding the motivation of our
+study. To clarify, the central contribution of our work lies in proposing a
+cohesive framework for recursion schemes in a proof assistant such as Coq, and
+the primary *application* of this framework is the verification of
+divide-and-conquer algorithms in Coq with clean extraction to readable and
+idiomatic OCaml. In contrast, prior work on recursion schemes in Haskell has
+largely relied on informal, pen-and-paper proofs and manual, unverified
+implementations in Haskell, leaving a gap in the ability to produce formally
+verified and automatically extracted implementations. Our framework addresses
+this gap by integrating formal verification and code generation into a single,
+cohesive process.
+
+> I could not understand what is the real motivation of this study.
+> does an OCaml programmer use this to write a program?
+
+Our end goal is to build a framework for program verification based on program
+calculation techniques. In that sense, our framework is meant to make the use
+of program calculation more accessible to programmers. Prior to this work,
+using program calculation for divide-and-conquer computations would require
+pen-and-paper proofs and a manual translation to Haskell/OCaml code.
+
+> For the latter, why does using Coq help with which problem of hylomorphisms?
+
+Hylomorphisms are not guaranteed to exist, or to be unique, due to termination
+issues. If a hylomorphism can be encoded in our framework in Coq, it must exist
+and be unique.

### 2025-03-13 paper/paper.tex
LIPICs style

@@
 %%% TeX-command-extra-options: "-shell-escape"
-\documentclass{llncs}
+%\documentclass{llncs}
+\documentclass[anonymous]{lipics-v2021}
 % \documentclass[llncs, a4paper, UKenglish, cleveref, autoref, thm-restate]{article}
@@
 
-\author{David Castro Perez\inst{1}\orcidID{0000-0002-6939-4189} \and
-Marco Paviotti\inst{2,3}\orcidID{0000-0002-1513-0807} \and
-Michael Vollmer\inst{3}\orcidID{0000-0002-0496-8268}}
-%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% LNCS author information
+% \author{David Castro Perez\inst{1}\orcidID{0000-0002-6939-4189} \and
+% Marco Paviotti\inst{2,3}\orcidID{0000-0002-1513-0807} \and
+% Michael Vollmer\inst{3}\orcidID{0000-0002-0496-8268}}
+% %
+% \authorrunning{D. Castro Perez et al.}
+% % First names are abbreviated in the running head.
+% % If there are more than two authors, 'et al.' is used.
+% %
+% \institute{University Of Kent, Canterbury, CT2 7NZ, United Kingdom\\
+% \email{\{D.Castro-Perez, M.Paviotti, M.Vollmer\}@kent.ac.uk}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% LIPICS author information
+\author{David {Castro Perez}}%
+{University Of Kent, Canterbury, CT2 7NZ, United Kingdom}%
+{D.Castro-Perez@kent.ac.uk}%
+{0000-0002-6939-4189}%
+{}
+\author{Marco Paviotti}%
+{University Of Kent, Canterbury, CT2 7NZ, United Kingdom}%
+{M.Paviotti@kent.ac.uk}%
+{0000-0002-1513-0807}%
+{}
+\author{Michael Vollmer}%
+{University Of Kent, Canterbury, CT2 7NZ, United Kingdom}%
+{M.Vollmer@kent.ac.uk}%
+{0000-0002-0496-8268}%
+{}
+
 \authorrunning{D. Castro Perez et al.}
-% First names are abbreviated in the running head.
-% If there are more than two authors, 'et al.' is used.
-%
-\institute{University Of Kent, Canterbury, CT2 7NZ, United Kingdom\\
-\email{\{D.Castro-Perez, M.Paviotti, M.Vollmer\}@kent.ac.uk}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\ccsdesc{}
+\keywords{}
 

### 2025-03-14 paper/paper.tex
Current version

@@
 
-This paper presents the first Coq formalisation of
-a recursion scheme, called the \emph{hylomorphism}, along with its algebraic
-laws allowing for the mechanisation of all recognised (terminating) recursive
-algorithms. The key contribution of this paper is that this formalisation is
-fully axiom-free allowing for the extraction of safe, idiomatic OCaml code. We
-exemplify the framework by formalising a series of algorithms based on different
-recursive paradigms such as divide-and conquer, dynamic programming, and mutual
-recursion and demonstrate that the extracted OCaml code for the programs
-formalised in our framework is efficient, resembles code that a human programmer
-would write, and contains no occurrences of \ocaml{Obj.magic}.  We also present
-a machine-checked proof of the well-known short-cut fusion optimisation.
+This paper presents the first Coq formalisation of a recursion scheme, called
+the \emph{hylomorphism}, along with its algebraic laws allowing for the
+mechanisation of all recognised (terminating) recursive algorithms. The key
+contribution of this paper is that this formalisation is fully axiom-free
+allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
+framework by formalising a series of algorithms based on different recursive
+paradigms such as divide-and conquer, dynamic programming, and mutual recursion
+and demonstrate that the extracted OCaml code for the programs formalised in our
+framework is efficient, resembles code that a human programmer would write, and
+contains no occurrences of \ocaml{Obj.magic}.  Furthermore, we show the use of
+the machine-checked proofs of the laws of hylomorphisms to do program
+optimisations.
 \end{abstract}
@@
 
-Despite the rigorous set of algebraic laws that are satisfied by recursion
-schemes, there is a lack of tool support for their use in the context of program
-calculation. In fact, most of the work on applying program calculation is done
-by performing pen-and-paper proofs, and then translating the result to specific
-instances of recursion schemes, implemented in a programming language (generally
-Haskell). There are some examples of implementations of program calculation
-techniques, but these implementations are scarce, not up to date, and not
-verified in a proof assistant. For example, Cunha et al.~\cite{DrHylo} automated
-program calculation techniques by a custom Haskell implementation.
+There is not much tool support to use program calculation techniques. In fact,
+uses of program calculation are done by  pen-and-paper proofs, and then
+translating the result to specific instances of recursion schemes, implemented
+in a programming language (generally Haskell).  There are some examples of
+implementations of program calculation techniques, but these implementations are
+scarce, not up to date, and not verified in a proof assistant. For example,
+Cunha et al.~\cite{DrHylo} automated program calculation techniques by a custom
+Haskell implementation.
 
 Several authors build tools for applying program calculation techniques by
-\emph{mechanising} them as part of a proof assistant. However, most of the work
-in mechanising recursion schemes focuses on a narrow subset of the known
-recursion schemes due to termination issues, or do not focus on proving
-algebraic laws of generic recursion
-schemes~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.  Indeed,
-most of the mechanisations of generic recursion schemes focus on \emph{maps} and
-\emph{folds}, and few authors focus on the \emph{unfolds}. Omitting
+\emph{mechanising} them as part of a proof assistant. However, most of the
+existing work only mechanises a narrow subset of the know recursion schemes, or
+do not prove their algebraic
+laws~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.  Indeed,
+most of the mechanisations of generic recursion schemes cover \emph{maps} and
+\emph{folds}, and few authors focus implement \emph{unfolds}. Omitting
 \emph{unfolds} severely limits the expressivity of the resulting mechanisations,
@@
 Coq~\cite{OnoHTNH11,Larchey-Wendling23,MiculanP12,Sakaguchi20}.  In Abreu et
-al's approach, extraction (1) does not preserve the recursive structure of
-common implementations; and (2) leads to unsafe casts like \ocaml{Obj.magic} in
-the generated code. This latter is also problematic in that, for higher-order
-programs, simple interoperations can lead to incorrect behaviour or even
-segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates the
-fast-and-loose principle~\cite{DanielssonHJG06}.
-
-The contributions of this paper are as follows.  This work presents the first
-Coq formalisation \emph{hylomorphisms} that (1) is \emph{fully axiom-free}; (2)
+al's approach, extraction (1) does not produce code with the same recursive
+structure  as existing manual implementations; and (2) leads to unsafe casts
+like \ocaml{Obj.magic} in the generated code. This latter is also problematic in
+that, for higher-order programs, simple interoperations can lead to incorrect
+behaviour or even segfaults~\cite{forster:hal-04329663} and, moreover, it
+invalidates the fast-and-loose principle~\cite{DanielssonHJG06}.
+
+The contributions of this paper are as follows.  This work presents a Coq
+formalisation of \emph{hylomorphisms} that (1) is \emph{fully axiom-free}; (2)
 allows the extraction of idiomatic OCaml code; and (3) can use regular Coq
 equalities to do program calculation, derive correct implementations, and apply
-optimisations.
-The full mechanisation is open source.
+optimisations.  The full mechanisation is open source.
 %can be found on
@@
 fusion on programs that have already been proven to terminate.
-% As an example, our mechanisation of \ocaml{qsort} as a
-% hylomorphism will produce the following code:
-% \begin{coqcode}
-% let rec qsort = function | [] -> []
-%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
-%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
-%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
-% \end{coqcode}
-%
 
@@
 \section{Recursion Schemes}\label{sec:rec-schemes}
-The structure of data is very similar to the structure of an
-algorithm which processes that data. This relationship manifests in the form of
-structured recursion schemes which are widely used in functional languages such
-Haskell. Canonical examples are folds (catamorphisms), which consume data, and
-unfolds (anamorphisms), which produce it. While some implementations of
-recursion schemes like $\operator{foldr}$ in Haskell are specific to a particular
-data structure, in this case $\operator{List}$s, we can generalise further these
-ideas to account for generic (co)inductive data types and generic algorithms
-operating on them.
+The structure of data is very similar to the structure of an algorithm which
+processes that data. This relationship manifests in the form of structured
+recursion schemes which are widely used in functional languages such Haskell.
+Canonical examples are folds (catamorphisms), which consume data, and unfolds
+(anamorphisms), which produce it. While some implementations of recursion
+schemes like $\operator{foldr}$/$\operator{unfoldr}$ in Haskell are specific to
+a particular data structure, in this case $\operator{List}$s, we can generalise
+further these ideas to account for generic (co)inductive data types and generic
+algorithms operating on them.
 
@@
 \]
-as the set inductively generated by the
-constructors $\nil : 1 \to \List{A}$ and
-$\cons : A \times \List{A} \to \List{A}$.
-The terminal object in $\Set$ is the \emph{unit} type, that only has one
-inhabitant, $* : 1$.
-The $\List{-}$ type is a functor. Its action on objects is to take any set
-$A$ to $\List{A}$, and its functorial
-action $\List{f} : \List{A} \to \List{B}$ is given by $\List{f}(*) = *$ and
+as the set inductively generated by the constructors $\nil : 1 \to \List{A}$ and
+$\cons : A \times \List{A} \to \List{A}$.  The terminal object in $\Set$ is the
+\emph{unit} type, that only has one inhabitant, $* : 1$.  The $\List{-}$ type is
+a functor from  $\Set \to \Set$. When a functor maps a category to itself, it is
+called an \textbf{endofunctor}. The action of $\List{-}$ is to take any set $A$
+to the set $\List{A}$, and its functorial action
+$\List{f} : \List{A} \to \List{B}$ is given by $\List{f}(\nil(*)) = \nil(*)$ and
 $\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
@@
 $\cons : A \times \Str{A} \to \Str{A}$.
-
-\subsubsection{Functors in Haskell}
-It is often helpful for functional programmers to think of categorical
-constructions in terms of Haskell thereby assuming that Haskell types are sets
-and programs are functions~\cite{DanielssonHJG06}.
-
-We can declare the class of types which are functors. This is done by imposing
-that a function \mintinline{haskell}{f : * -> *} comes with a functorial action
-given by function \mintinline{haskell}{fmap}:
-\begin{minted}{haskell}
-class Functor f where
-  fmap :: (a -> b) -> f a -> f b
-\end{minted}
-
-We now define the type of lists as usual and then show that it is an instance of
-the class \mintinline{haskell}{Functor} by implementing the
-\mintinline{haskell}{fmap} program:
-\begin{minted}{haskell}
-data List a = Nil | Cons a (List a)
-
-instance Functor List where
- fmap f Nil = Nil
- fmap f (Cons x xs) = Cons (f x) (fmap f xs)
-\end{minted}
 
@@
 respect the structure of the algebra, that is
-$f \circ a_{X} = A_{Y} \circ F(f)$. The initial object in this category is
+$f \circ a_{X} = a_{Y} \circ F(f)$. The initial object in this category is
 called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
@@
   \label{eq:cata:unique}
-  f = \cata{a_{X}} \iff f = a_{X} \circ F(f) \circ \InIso
+  f = \cata{a_{X}} \iff f = a_{X} \circ F(f) \circ \InOp
 \end{equation}
@@
 $\foldr\;base\;ind = \cata{[base, ind]}$.
-
-\subsubsection{Inductive Types and Catamorphisms in Haskell }
-We now implment the least fixed-point of a functor \mintinline{haskell}{f}.
-Inuitively this is the type of \mintinline{haskell}{f}-branching trees of finite depth:
-\begin{minted}{haskell}
-newtype Fix f = In {inOp :: f (Fix f )}
-\end{minted}
-Note that since Haskell is a lazy language with general recursion this type can
-also represent infinite trees. However, the point of using recursion schemes is
-precisely that of avoid using general recursion in the first place.
-
-A catamorphism can be readily implemented as follows
-\begin{minted}{haskell}
-cata :: Functor f => (f a -> a) -> Fix f -> a
-cata alg = alg . fmap (cata alg) . In
-\end{minted}
-
-As an example of how to construct inductive types we define the functor
-$X \mapsto 1 + A \times X$ as data type parameterized by a type:
-\begin{minted}{haskell}
-data ListF a x = Nil | Cons a deriving Functor
-newtype List a = (Fix ListF) a
-\end{minted}
-It should be easy now to see that a foldr is precisely a catamorphism over the
-type \mintinline{haskell}{List} which we have just defined.  For example,
-summing up the elements of a list can be implemented using a catamorphism as
-follows
-\begin{minted}{haskell}
-sum :: Fix (ListF Int) -> Int
-sum xs = cata (\case { Nil -> 0; Cons n m -> n + m } )  xs
-\end{minted}
-In the program above, the code inside the lambda case is precisely the
-\mintinline{haskell}{ListF Int}-algebra for the catamorphism.
 
@@
 
-\subsubsection{Coinductive Types and Anamorphisms in Haskell }
-Similarly to the inductive case, we can implement coinductive types and
-catamorphisms as follows:
-\begin{minted}{haskell}
-newtype CoFix f = OutOp {out :: f (Fix f )}
-\end{minted}
-As mentioned above, this type is similar to the data type
-\mintinline{haskell}{Fix} we already defined. However, this time we are going to
-use this assuming that it is the coinductive fixed-point by defining an
-anamorphism on it:
-\begin{minted}{haskell}
-ana :: Functor b => (a -> b a) -> a -> CoFix b
-ana coalg = OutOp . fmap (ana coalg) . coalg
-\end{minted}
-For example, the stream of natural numbers can now be defined using an anamorphism
-\begin{minted}{haskell}
-nat :: CoFix (ListF Int)
-nat = ana (\x -> Cons x (x+1)) 0
-\end{minted}
-where the argument to the anamorphism is the
-\mintinline{haskell}{ListF Int}-coalgebra.
-
 \subsection{Recursive Coalgebras and Hylomorphism}
@@
 \end{equation}
-As we stated earlier, a solution to this equation does not exist for an
-arbitrary algebra and coalgebra pair and, in fact, this definition cannot be
-accepted by Coq.
+A solution to this equation does not exist for an arbitrary algebra and
+coalgebra pair and, in fact, a definition like this cannot be accepted by Coq.
 
@@
 
-\subsubsection{Haskell Implementation of Hylomorphisms}
-% which is very useful in list processing. As a fold, The frequently used function
-% map is also a fold:
-% \begin{minted}{haskell}
-% map :: (a -> b) -> [a] -> [b]
-% map f = foldr (\x xs -> (f x):xs) []
-% \end{minted}
-We implement the hylomorphism tightly following the theory explained above:
-\begin{minted}{haskell}
-hylo :: Functor f => (c -> f c) -> (f a -> a) -> c -> a
-hylo c a = a . fmap (hylo c a) . c
-\end{minted}
-
-A classic example of a divide-and-conquer algorithm that can be written as an
-hylomorphism is the quicksort program below:
-\begin{minted}{haskell}
-partition x xs = ([l | l <- xs, l < x], x, [r | r <- xs, r >= x])
-
-quicksort :: [Int] -> [Int]
-quicksort [] = []
-quicksort (x:xs) = let (xs, x, ys) = partition x xs
-                   in (quicksort xs) ++ [x] ++ (quicksort ys)
-\end{minted}
-The partition program takes an element \mintinline{haskell}{x} and a list and
-returns a triple consisting of a list with all the elements in the list smaller
-than \mintinline{haskell}{x}, the element itself and a list with all the
-elements in the list greater or equal than  \mintinline{haskell}{x}. The
-quicksort program first runs the partition program, which divides the list in
-three parts, calls itself recursively on the sublists and then puts the results
-together using the concatenation function.
-
-To implement this as an hylo we define the functor representing trees where
-leaves are just empty and each node contains a value:
-\begin{minted}{haskell}
-data TreeF a k = Leaf | Branch k a k deriving Functor
-\end{minted}
-Now the partition program can be written as a  \mintinline{haskell}{TreeF}-coalgebra
-\begin{minted}{haskell}
-divide :: [Int] -> TreeF Int [Int]
-divide [] = Leaf
-divide (p:xs) = Branch [l | l <- xs, l < p] p [r | r <- xs, r > p]
-\end{minted}
-and the concatenation function can be written as  \mintinline{haskell}{TreeF}-algebra
-\begin{minted}{haskell}
-conquer :: TreeF Int [Int] -> [Int]
-conquer Leaf = []
-conquer (Branch ls x rs) = ls ++ [x] ++ rs
-\end{minted}
-Finally, quicksort can be rewritten as an hylomorphism
-\begin{minted}{haskell}
-quicksort = hylo divide conquer
-\end{minted}
-As an important remark we have used the in-built list type for keeping the code
-as tidy as possible, but it would have been possible to use the inductive
-\mintinline{haskell}{List} type we defined above and define both
-\mintinline{haskell}{divide} and \mintinline{haskell}{conquer} via
-catamorphisms.
-
-Finally, it is important to note that both  \mintinline{haskell}{divide} and
-\mintinline{haskell}{conquer} are not initial algebras or terminal coalgebras on
-\mintinline{haskell}{ListF}. These are rather (co)algebras on
-\mintinline{haskell}{TreeF}.  Hence the terminating argument relies solely on
-the fact that \mintinline{haskell}{divide} is a recursive coalgebra thus there
-is an unique solution to \mintinline{haskell}{hylo divide conquer}.
-
-\subsubsection{Recursive Anamorphisms}
+\paragraph{Recursive Anamorphisms}
 Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
@@
 can be applied only finitely many times, therefore when this is applied to an
-anamorphism the only possible traces it can produce from the seed function are
-the finite ones. We denote this special kind of recursion scheme \emph{recursive
+anamorphism the only possible types it can produce from the seed are
+the finite ones. We call this special kind of recursion scheme \emph{recursive
 anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
@@
 exactly hylomorphisms on the recursive $F$-coalgebra and the
-algebra $\InIso$ for the inductive data type $\nu F$. This fact falls out
+algebra $\InIso$ for the inductive data type $\mu F$. This fact falls out
 from the uniqueness property of the hylomorphism and the fact that recursive
@@
 \label{sec:containers}
-
-In this section, we focus on mechanising \emph{container functors} in a way that
-is suitable for code extraction. Recall from Section~\ref{sec:rec-schemes} that
-we need to be able to abstract away from particular shape of the data, and we
-can do this using \emph{functors} that have suitable fixed-point properties;
-i.e.\ those which have a initial algebras.
-%
-A common approach to construct such functors is to use
-\emph{containers}~\cite{AbbottAG05}. However, reasoning about container equality
-will require us to consider both functional extensionality and heterogeneous
-equality. We avoid these axioms by introducing a custom equivalence relation on
-types (Section~\ref{sec:types-morphisms}).
-
-\subsection{Functors and Containers}
-\label{sec:functors}
 
@@
 \coq{F : Type -> Type}
-satisfying the necessary identity and composition properties, then we will not
-be able to construct least/greatest fixed points of this type, and we will not
-be able to mechanise recursion schemes that follow this structure:
-\begin{coqcode}
-(* REJECTED due to negative occurrence of (LFix F) *)
-Inductive LFix (F : Type -> Type) := LFix_in (lfix_out : F (LFix F)).
-\end{coqcode}
-
-\noindent
-One common workaround for this is the use of Mendler-style
-recursion~\cite{Mendler91}. Roughly, the idea is to ``pack'' an abstract type
-together with the functor that captures the structure of the recursion. From
-this abstract type, we will be able to ``extract'' the contents of the functor.
-This is, for example, the solution used by the PaCo library~\cite{PaCo}, and a
-simplified version of it is the following:
-\begin{coqcode}
-Inductive MendlerLFix (F : Type -> Type) :=
-  In Y (next : Y -> MendlerLFix F) (in_op : F Y).
-\end{coqcode}
-
-\noindent
-While this definition solves the strict positivity problem, it does not solve
-the problems of: (1) reasoning about data/program equalities; and (2) extraction
-to idiomatic OCaml code. To reason about program equalities, we would need to
-reason about equalities of functions with \emph{some abstract} \coq{Y} as their
-domain.  Furthermore, the extraction of \coq{F Y} will lead to unsafe casts in
-the extracted OCaml code.
-
-% \paragraph{Containers}
-Another common solution is to the strict positivity
-problem is the use \emph{containers}~\cite{AbbottAG05}.  A container
-$S \triangleright P$ is defined by a type of \emph{shapes} $S : \Type$ and a
-\emph{family of position types} indexed by shapes $P : S \to \Type$.  An
-\emph{extension} of this container is a functor
-$\llbracket S \triangleright P \rrbracket$, whose action on objects is given by
+satisfying the necessary identity and composition properties, then
+we will not be able to construct least/greatest fixed points of this type.  A
+common solution to the strict positivity problem is the use
+\emph{containers}~\cite{AbbottAG05}.  A container $S \triangleright P$ is
+defined by a type of \emph{shapes} $S : \Type$ and a \emph{family of position
+types} indexed by shapes $P : S \to \Type$.  An \emph{extension} of this
+container is a functor $\llbracket S \triangleright P \rrbracket$, whose action
+on objects is given by
 $\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
@@
 $\llbracket S \triangleright P \rrbracket \; f = \lambda (s,\;g).\;(s,\;f\circ g)$.
-
-To explain the role of shapes and positions in containers, consider the functor
-$F\; X = 1 + A \times X$.  An equivalent container, i.e.\ a container with an
-extension that is isomorphic to $F$, requires a shape that can distinguish two
-cases, indicating that there are two constructors to build an object of type
-$F\;X$. For example, we could define $S_{F} = 1 + A$. We need to consider two
-cases for the family of positions:  $P_{F}(\mathsf{inj}_{1}\;*) = 0$, because
-there are no occurrences of $X$ in the left case; and
-$P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence of $X$ in the
-right hand side. It is easy to see that
-$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.  In general, every
-polynomial functor has an equivalent container, and we have mechanised this
-result.
-%
-%Generally, any functor that
-%is isomorphic to a container extension is called a \emph{container functor}.
-%Here we will call \emph{container functor} the functors defined by a specific
-%container extension.
-
-\subsection{Extractable Containers}
-A straightforward encoding of containers in Coq would use dependent types to
-represent families of positions. However, Ocaml's type system is not equipped to
-handle these, which will lead to extracting OCaml code with unsafe casts.
-
-Consider instead representing the positions of a container using a decidable
-validity predicate which assigns shapes to positions. In Coq, we use a boolean
-function and a coercion from \coq{bool} to \coq{Prop} to represent the decidable
-predicate \coq{valid}, similarly to SSReflect~\cite{GonthierL09}. Given a
-\coq{Shape : Type}, and a type of \emph{all possible positions} \coq{APos : Type},
-the family of positions for a given shape can be defined as a dependent pair:
-\begin{coqcode}
-Definition Pos (s : Shape) = { p : APos | valid s p }.
-\end{coqcode}
-Coq's code extraction will now be able to erase the validity predicate, and
-generate OCaml code that is free of unsafe casts. The OCaml code extracted for
-\coq{Pos} will now be exactly the code extracted for \coq{APos}.  The
-decidability of the validity predicate is crucial for our purposes of  remaining
-axiom-free. To illustrate this, suppose that we need to show the equality of two
-container extensions. We will need to show that, for the same positions, they
-will produce the same result. In Coq, the goal would look as follows:
-\begin{coqcode}
- k : Pos s -> X
- P1, P2 : valid s p
- ---------------------------
- k (existT _ p P1) = k (existT _ p P2)
-\end{coqcode}
-If \coq{valid} was a regular proposition in \coq{Prop}, it would not possible
-to prove the equality of \coq{P1} and \coq{P2}. However, by using a decidable
-predicate, if we know that \coq{P1} and \coq{P2} are of type 
-\coq{valid s p = true}, then we can prove without any axioms that 
-\coq{P1 = P2 = eq_refl}.
-
-Suppose that \coq{F} is a container represented in Coq. One can think of these
-\coq{F} as pairs containing the shapes and families of positions (which are, in
-turn, defined in terms of decidable validity predicates).
-An extension of a container \coq{F}, \coq{App F} in Coq, is then defined as
-follows:
-\begin{coqcode}
-Record App F (X : Type)
-   := { shape : Shape F; cont : {p | valid shape p} -> X }.
-\end{coqcode}
-
-This approach solves the problem of the unsafe casts, since the code extracted
-for \coq{{p | valid shape p}} will be an OCaml singleton type defined as the
-OCaml equivalent to \coq{Pos}.
-
-\subsubsection{Equality of Container Extensions}
-Reasoning about the equality of container extensions is not entirely solved by
-using decidable validity predicates to define the families of positions. In
-general, we want to equate container extensions that have the same shape, and
-that, for equal shape and position, they return the same element. To avoid the
-use of the functional extensionality axiom, we capture this relation with the
-following inductive proposition:
-\begin{coqcode}
-Inductive AppR (x y : App F X) : Prop :=
-| AppR_ext (shape x = shape y)
-           (forall p1 p2, projT1 e1 = projT1 e2 -> cont p1 = cont p2).
-\end{coqcode}
-Note that we do not care about the validity proof of the positions, only their
-value. This is to simplify (slightly) our proofs. This relation is trivially
-\emph{reflexive}, \emph{transitive}, and \emph{symmetric}.
-
-However, the use of a different equality for container extensions now forces us
-to deal with the fact that some types have different definitions of equality.
-In particular, we want to reason about the equality of functions of types such
-as \coq{App F A -> B} (or \coq{B -> App F A}).
-%
-Since these types now come with their own equivalence, any function that
-manipulates them needs to be \emph{respectful}; i.e.\ given
-\coq{R : X -> X -> Prop} and \coq{R' : Y -> Y -> Prop},
-we want functions (morphisms) that satisfy the following property:
-\begin{coqcode}
-forall (x y : X), R x y -> R' (f x) (f y)
-\end{coqcode}
-
-\subsection{Types and Morphisms}
-\label{sec:types-morphisms}
-We address the different forms of equality by defining a class of
-\emph{setoids}, types with an associated equivalence relation, and considering
-only functions that respect the associated equivalences, or \emph{proper}
-morphisms with respect to the function \emph{respectfulness} relation. We use
-the type-class mechanism, instead of setoids in Coq's standard library, to help
-Coq's code extraction mechanism remove any occurence of custom equivalence
-relations in the extracted OCaml code. We use \coq{=e} to denote the equivalence
-relation of a setoid. By default, we associate every Coq type with the standard
-propositional equality, unless a different equivalence is specified (we allow
-overlapping instances, and Coq's propositional equality takes the lowest
-priority).
-%\begin{coqcode}
-%Reserved Notation "f =e g" (at level 70, no associativity).
-%Class setoid A : Type :=
-%  MkEquiv
-%    { eqRel : A -> A -> Prop;
-%      e_refl : forall x, eqRel x x;
-%      e_sym : forall x y, eqRel x y -> eqRel y x;
-%      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
-%    }.
-%Notation "f =e g" := (eqRel f g).
-%\end{coqcode}
-Given types \coq{A} and  \coq{B}, with their respective
-equivalence relations \coq{eA : A -> A -> Prop} and
-\coq{eB : B -> B -> Prop}, the we define the type
-\coq{A ~> B} to represent \emph{proper} morphisms of the
-respectfulness relation of \coq{R} to \coq{R'}.
-\begin{coqcode}
-Record morph A {eA : setoid A} B {eB : setoid B} :=
-  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
-Notation "A ~> B" := (@morph A _ B _).
-\end{coqcode}
-We rely on Coq's type class mechanism to fill in the necessary equivalence
-relations. Coq's code extraction mechanism will erase any occurrence
-of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be extracted to
-the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
-from \coq{A ~> B} to \coq{A -> B}.
-On top of this, we define basic function composition and identity functions:
-\begin{coqcode}
-Notation "f \o g" = (comp f g).
-Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
-Definition id : A ~> A := ...
-\end{coqcode}
-Using custom equivalences and proper morphisms, we redefine the definitions of
-container extensions and container equality. In particular, container
-extensions require a proper morphism to check the validity of positions in
-shapes, and container equality now uses equivalences of shapes and contained
-elements: 
-\begin{coqcode}
-Inductive AppR F `{setoid X} (x y : App F X) : Prop :=
-  | AppR_ext (Es : shape x =e shape y)
-             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
-\end{coqcode}
-Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
-however, we chose to use Coq's propositional equality, since this leads
-to simpler code. We explain why in Section~\ref{sec:coq-algebras}, and the
-mechanisation of initial algebras of container extensions.
-%The definitions in
-%Figure~\ref{fig:poly} are redefined to account for these changes in a
-%straightforward way, i.e. we prove that validity predicates are indeed proper
-%morphisms.
-
-Our definition of morphisms, and the use of different equivalences leads to the
-well-known ``setoid hell''. We mitigate this problem by providing tactics and
-notations to automatically discharge proofs of \coq{app_eq} for morphisms,
-whenever the types use the standard propositional equality, or a combination of
-propositional and extensional equality. However, our compositional approach
-allows us to build morphisms by plugging in other morphisms to our combinators.
-In our framework, our expectation is that the user-provided functions remain
-small, with relatively straightforward proofs of \coq{app_eq}.
-
-However, by using this mechanisation, we gain simplified proofs via
-Coq's \emph{Generalised Rewriting}. Since every morphism \coq{f : A ~> B}
-satisfies the property that if \coq{x =e y}, then \coq{f x =e f y}, we can add
-every morphism as a proper element of Coq's respectfulness relation. In
-practice, this means that we can use the \coq{rewrite} tactic on proofs of
-type \coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
-arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},
-and \coq{transitivity} tactics. For example:
-\begin{coqcode}
-Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D) (H : h \o g =e id), 
-  h \o (g \o f) =e f.
-Proof.  rewrite compA, H.  reflexivity. Qed.
-\end{coqcode}
-
-\subsubsection{Polynomial Types}
-We define a number of equivalences for polynomial types.
-%These equivalences are
-%be used when building proper morphisms that manipulate their respective types.
-\begin{coqcode}
-Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
-Instance pair_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A * B).
-Instance sum_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A + B).
-Instance prop_eq : setoid Prop.
-Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
-\end{coqcode}
-Most of the definitions that involve functions and polynomial types are
-straightworward.  Identity and composition are defined as 
-\coq{fun x => x} and \coq{fun f g x => f (g x)} 
-respectively, and the proofs that they are proper morphisms is straightforward,
-and automatically discharged by Coq. 
-Products are built using function \coq{fun f g x => (f x, g x)},
-with the projections being the standard Coq  \coq{fst} and
-\coq{snd} functions. Similarly, sum injections are encoded using
-Coq's \coq{inl} and \coq{inr} constructors, and
-pattern matching on them uses the function:
-\begin{coqcode}
-fun f g x => match x with | inl y => f y | inr y => g y end
-\end{coqcode}
-The proofs that these morphisms are proper are straightforward.  Finally, we
-also provide functions for currying/uncurrying, and  flipping the arguments of
-a proper morphism. We force most of our definitions to be inlined, to help
-Coq's code extraction mechanism to inline as many of these combinators as
-possible.
-
-We prove the isomorphisms of polynomial types and the equivalent 
-container extensions. As an example, we will consider the isomorphisms
-of pairs with their equivalent container extensions. Suppose that we
-know that \coq{App F X} is isomorphic to \coq{A}, and
-\coq{App G X} is isomorphic to \coq{B}. Then we can
-show that \coq{App (Prod F G) X} is isomorphic to
-\coq{A * B}.  If we have an element of type 
-\coq{App (Prod F G) X},
-using the \coq{inl} position, we can obtain
-\coq{App F X}.  Similarly, using \coq{inr}, we can
-obtain \coq{App G X}. Since these are the only two valid positions
-in the shape of pairs, we have finished. It is now sufficient to use the
-isomorphisms of \coq{App F X} and \coq{App G X} to
-obtain \coq{A * B}. Similarly if we have \coq{A * B},
-we can first use the isomorphisms of 
-\coq{A} and
-\coq{B} to obtain
-\coq{App F X * App G X}, and then construct the necessary container
-extension. Given \coq{p_inl : Pos l -> Pos (l * r)} (resp. \coq{p_inr}) that
-act as \coq{inl} (resp.\ \coq{inr}) on product positions, and
-\coq{case_pos : (Pos l -> X) -> (Pos r -> X) -> Pos (l * r) -> X} that pattern
-matches on the product positions, the functions that witness the isomorphism
-are:
-\begin{coqcode}
-Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
-  ({| shape := shape (fst x); cont := fun e => cont x (p_inl e) |}, 
-   {| shape := shape (snd x); cont := fun e => cont x (p_inr e) |}).
-Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
-{| shape := (shape (fst x), shape (snd x));
-   cont := case_pos (cont (fst x)) (cont (snd x)) |}.
-\end{coqcode}
-
-\noindent
-Proving that the composition of these functions is the identity is
-straightforward using the fact that \coq{Prod} containers only have two valid
-positions.
+Intuitively, the type of shapes captures the different ``cases'' of an algebraic
+datatype, and the positions of a shape are used to identify the elements inside
+a container of the given shape.
+
+Any polynomial functor can be represented as a container (see accompanying code
+for a mechanisation).  For example, a container that is equivalent to
+$F\; X = 1 + A \times X$ requires a shape that can distinguish two cases,
+indicating that there are two constructors to build an object of type $F\;X$,
+e.g. $S_{F} = 1 + A$. We need to consider two cases for the family of positions:
+$P_{F}(\mathsf{inj}_{1}\;*) = 0$, because there are no occurrences of $X$ in the
+left case; and $P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence
+of $X$ in the right hand side. It is easy to see that
+$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.
+
+The straightforward mechanisation of containers and container extensions is
+still not enough for our goals of remaining axiom-free, extracting to idiomatic
+OCaml code, and avoiding \ocaml{Obj.magic} in the extracted code. The main
+problem is that a container extension requires a function from a family of
+positions to a type: $\forall s : S.\; P\;s \to X$. On the Coq side, dealing
+with the equality of objects that contain (dependent) function types would force
+us to add the functional extensionality axiom, and axiom K (or equivalent) to
+deal with the heterogeneous equality of position.  For extraction, OCaml's type
+system is not powerful enough to represent such dependencies in the general
+case, which will lead to \ocaml{Obj.magic}.
+
+To avoid the functional extensionality axiom, we require that all the types used
+in our definitions are \emph{setoids}, where we will use \coq{x =e y} to
+represent setoid equality. All morphisms need to be \emph{proper} morphisms with respect
+to the function \emph{respectfulness} relation. In our development, we use
+\begin{coqcode}
+Notation "A ~> B" := (* proper morphisms  s.t. forall x y, x =e y -> f x =e f y *).
+\end{coqcode}
+We add an automatic coercion from \coq{A ~> B} to \coq{A -> B}, provide tactics
+to automatically discharge proofs that \coq{forall x y, x =e y -> f x =e f y}
+for the simplest cases of \coq{f}, and we provide a library with
+definitions for basic building blocks: sum and product types, exponentials,
+composition, identity,
+\ldots
+
+To avoid axiom K, and extract containers to idiomatic OCaml code, we
+define containters in terms of shapes, \emph{base} positions (\coq{APos : Set}),
+and \emph{decidable validity predicates} that
+specify when a base position is valid in a shape. Since validity predicates are in
+\coq{Prop}, they will be erased during extraction, and their decidability
+will make unnecessary the use of axiom K.
+\begin{coqcode}
+Record App F (X : Type) := {shape : Shape F; cont : {p : APos | valid s p = true} -> X}.
+\end{coqcode}
+
+Since we require every type to be a setoid, we need to define a setoid equality
+(\coq{=e}) for container functors. Given two \coq{x y : App F X}, we define
+\coq{x =e y} as the setoid equality of the shapes, and the extensional equality of
+the contents, given equal shapes. I.e.\ \coq{x =e y} iff
+\begin{coqcode}
+shape x =e shape y /\ (forall p p', projT1 p = projT1 p' -> cont x p =e cont y p')
+\end{coqcode}
 

### 2025-03-15 paper/paper.tex
pass over section 6

@@
 
+\subsection{Fusion Optimisation}\label{sec:fusion}
+
+Fusion in functional programming is a general technique for combining multiple
+computations into a single pass to eliminate intermediate data structures or
+repeated traversals, improving efficiency. The fusion we have discussed in this
+paper relating to program calculation is one form of fusion, but there are other
+approaches, and prior research has explored the mechanisation and verification
+of fusion transformations.
+
+Fusion has been applied to recursive tree traversals, in order to reduce the
+number of times each section of the tree must be visited. Research has
+demonstrated that this sort of fusion can be performed automatically by a
+compiler~\cite{10.1145/3133900}, and the correctness of tree traversal fusion
+has been verified mechanically in Coq~\cite{10.1145/3377555.3377884}.
+
+Another related family transformations are worker/wrapper
+transformations~\cite{gill09}, which have been used in the context of optimising
+compilers, and have been mechanised in Agda~\cite{SCULTHORPE_HUTTON_2014}. This
+technique has been applied to improving the performance of programs defined
+using folds via fusion~\cite{hutton10}. Their work does not focus on the use of
+an interactive proof assistant, but rather on the manual use of equational
+reasoning to re-write programs and on the automatic transformation of functional
+programs by a compiler.
+
 %\mvol{Need more general citations for program calculation.}
@@
 need for termination proofs. Their work differs from ours in that they require
-the functional extensionality axiom, and the use of impredicative \coq{Set}.
-The authors justify well the use of impredicative \coq{Set} and its
-compatibility with the functional extensionality axiom. In contrast, our
-development remains entirely \emph{axiom-free}. Another key difference with our
-approach is that they do not discuss what the resulting extracted code
-\emph{looks like}.
+the functional extensionality axiom, and the use of impredicative \coq{Set}. The
+authors justify well the use of impredicative \coq{Set} and its compatibility
+with the functional extensionality axiom. In contrast, our development remains
+entirely \emph{axiom-free}, and we retain the ability to extract natural-looking
+OCaml code.
+% Another key difference with our
+% approach is that they do not discuss what the resulting extracted code
+% \emph{looks like}.
 % (that is, whether the extracted OCaml code resembles the
 % natural formulation of the recursive function).
-Through experiments, we found that their formalisation leads to
-\ocaml{Obj.magic}, and code with a complex structure that may be hard to
-understand or interface with other OCaml code. Due to the great benefit of
-entirely avoiding termination proofs, it would be interesting to extend their
+Through experiments, we found that code extracted from their formalisation makes
+heavy use of \ocaml{Obj.magic} and has a generally complex structure that may be
+hard to understand and integrate with other OCaml code. Due to the great benefit
+of entirely avoiding termination proofs, it would be interesting to extend their
 approach to improve code extraction.
@@
 
-\subsection{Fusion}\label{sec:fusion}
-
-Fusion in functional programming is a general technique for combining multiple
-computations into a single pass to eliminate intermediate data structures and
-improve efficiency. The fusion we have discussed in this paper relating to
-program calculation is one form of fusion, but there are other approaches.
-
-\emph{Deforestation}, introduced by Wadler~\cite{wadler88}, is a specific
-approach to fusion that aims to remove intermediate trees (such as lists)
-created between chained operations. A well-known method of deforestation is
-\haskell{foldr/build} fusion~\cite{gill93}, which optimizes list-processing by
-fusing list-producing functions (like \haskell{map} or \haskell{filter}) with
-list-consuming ones (like \haskell{foldr}), avoiding the creation of temporary
-lists.
-%
-Extending this idea, stream fusion generalizes the concept to handle other data
-types, using a lazy stream abstraction to represent sequences and fuse
-transformations like \haskell{map} and \haskell{fold}~\cite{coutts07}.
+% \subsection{Fusion}\label{sec:fusion}
+
+% Fusion in functional programming is a general technique for combining multiple
+% computations into a single pass to eliminate intermediate data structures and
+% improve efficiency. The fusion we have discussed in this paper relating to
+% program calculation is one form of fusion, but there are other approaches.
+
+% \emph{Deforestation}, introduced by Wadler~\cite{wadler88}, is a specific
+% approach to fusion that aims to remove intermediate trees (such as lists)
+% created between chained operations. A well-known method of deforestation is
+% \haskell{foldr/build} fusion~\cite{gill93}, which optimizes list-processing by
+% fusing list-producing functions (like \haskell{map} or \haskell{filter}) with
+% list-consuming ones (like \haskell{foldr}), avoiding the creation of temporary
+% lists.
+% %
+% Extending this idea, stream fusion generalizes the concept to handle other data
+% types, using a lazy stream abstraction to represent sequences and fuse
+% transformations like \haskell{map} and \haskell{fold}~\cite{coutts07}.
 % Both approaches eliminate redundant structures and ensure computations happen in
@@
 
-In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
-equational reasoning about recursive programs~\cite{DANVY_2022}. Fold-unfold
-lemmas enable optimizations like fusion by allowing programs to be
-systematically simplified, including eliminating intermediate structures.
-Additionally, an approach to fusion called the worker/wrapper
-transformation~\cite{gill09} was formalised and mechanised in
-Agda~\cite{SCULTHORPE_HUTTON_2014}.
+% In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
+% equational reasoning about recursive programs~\cite{DANVY_2022}. Fold-unfold
+% lemmas enable optimizations like fusion by allowing programs to be
+% systematically simplified, including eliminating intermediate structures.
+% Additionally, an approach to fusion called the worker/wrapper
+% transformation~\cite{gill09} was formalised and mechanised in
+% Agda~\cite{SCULTHORPE_HUTTON_2014}.
 

### 2025-03-15 paper/paper.tex
There was a pagestyle plain which erased page numbers

@@
 %%% TeX-command-extra-options: "-shell-escape"
-%\documentclass{llncs}
-\documentclass[anonymous]{lipics-v2021}
+% \documentclass{llncs}
+\documentclass[a4paper,anonymous, UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
 % \documentclass[llncs, a4paper, UKenglish, cleveref, autoref, thm-restate]{article}
@@
 % \addbibresource{references.bib}
-\pagestyle{plain}
 
@@
 
-\ccsdesc{}
-\keywords{}
+% \author{Jane {Open Access}}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+Please provide at least the name of the affiliation and the country.
+The full address is optional.
+Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.
+
+% \author{Joan R.
+Public\footnote{Optional footnote, e.g.
+to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
+
+\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+
+\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+
+\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
+
+\category{} %optional, e.g. invited paper
+
+\acknowledgements{I want to thank \dots}%optional
+
+%\nolinenumbers %uncomment to disable line numbering
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\EventEditors{}
+\EventNoEds{2}
+\EventLongTitle{ITP 2025)}
+\EventShortTitle{ITP 2025}
+\EventAcronym{ITP}
+\EventYear{2025}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2025-03-17 paper/paper.tex
Moving Section 3 into Section 4

@@
 \end{mathpar}
-This latter component is also called the \emph{functorial action} and can be
-thought of as the $\operator{fmap}$ higher-order function in functional
-programming.
+This latter component is also called the \emph{functorial action} and, if types
+are $\Set$s, this can be thought of as the $\operator{fmap}$ higher-order
+function in functional programming.
 
@@
 as the set inductively generated by the constructors $\nil : 1 \to \List{A}$ and
-$\cons : A \times \List{A} \to \List{A}$.  The terminal object in $\Set$ is the
-\emph{unit} type, that only has one inhabitant, $* : 1$.  The $\List{-}$ type is
-a functor from  $\Set \to \Set$. When a functor maps a category to itself, it is
-called an \textbf{endofunctor}. The action of $\List{-}$ is to take any set $A$
-to the set $\List{A}$, and its functorial action
-$\List{f} : \List{A} \to \List{B}$ is given by $\List{f}(\nil(*)) = \nil(*)$ and
+$\cons : A \times \List{A} \to \List{A}$. The $\List{-}$ type is an \emph{endofunctor}
+$\Set \to \Set$ mapping objects and arrows in $\Set$ to $\Set$ itself.
+Its functorial action $\List{f} : \List{A} \to \List{B}$ is
+given by $\List{f}(\nil(*)) = \nil(*)$ and
 $\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
 $\List{f}$ is well-defined as it recursively calls on a smaller argument.
-However, there is a more general way of proving these type of definitions are
-well-defined. Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
+Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
 the greatest set generated by the constructor
-$\cons : A \times \Str{A} \to \Str{A}$.
+$\cons : A \times \Str{A} \to \Str{A}$. The maps $\head : \Str{A} \to A$ and
+$\tail : \Str{A} \to \Str{A}$ can be easily constructed from the isomorphism.
 
@@
 same coalgebra. As a result of the uniqueness property we can derive the fusion
-laws. For all $B$-coalgebra homomorphisms
-$f : (X, c_{X}) \to (Y, c_{Y})$ we have
-\begin{equation}
+law:\begin{equation}
   \label{eq:ana:fusion}
@@
 \end{equation}
+for all $B$-coalgebra homomorphisms $f : (X, c_{X}) \to (Y, c_{Y})$.
 
@@
 
-\section{Mechanising Extractable Container Functors}
+\section{Formalising Recursion Schemes}
+\label{sec:recursion-schemes}
+Recursion schemes provide an abstract way to consume and generate data. We now
+proceed onto describing how to formalise hylomorphisms in Coq. We first
+formalise container functors as a tool to represent polynomial functors (Section~\ref{sec:containers}).
+Then we formalise algebras for container extensions (Section~\ref{sec:coq-algebras}), then
+we formalise coalgebras (Section~\ref{sec:coq-coalg}) and then we put together
+these notions to formalise recursive coalgebras and hylomorphisms
+(Section~\ref{sec:coq-rec-coalgebras}).
+
+\subsection{Mechanising Extractable Container Functors}
 \label{sec:containers}
@@
 \end{coqcode}
-
-\section{Formalising Recursion Schemes}
-\label{sec:recursion-schemes}
-Recursion schemes provide an abstract way to consume and generate data. We now
-proceed onto describing how to formalise hylomorphisms in Coq. We first
-formalise algebras for container extensions (Section~\ref{sec:coq-algebras}), then
-we formalise coalgebras (Section~\ref{sec:coq-coalg}) and then we put together
-these notions to formalise recursive coalgebras and hylomorphisms
-(Section~\ref{sec:coq-rec-coalgebras}).
 

### 2025-03-17 paper/paper.tex
Massaged section 3.1

@@
 \begin{itemize}
-  \item In Section~\ref{sec:rec-schemes} we give an overview about recursion
+  \item In Section~\ref{sec:category-theory} we give an overview about recursion
 schemes.
@@
 
-\section{Recursion Schemes}\label{sec:rec-schemes}
+\section{Recursion Schemes}\label{sec:category-theory}
 The structure of data is very similar to the structure of an algorithm which
@@
 Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
-inductive data type in the following way. A recursive coalgebra
-can be applied only finitely many times, therefore when this is applied to an
-anamorphism the only possible types it can produce from the seed are
-the finite ones. We call this special kind of recursion scheme \emph{recursive
-anamorphism}. We can show that recursive anamorphisms of type $X \to \nu F$
-can also be given the type $X \to \mu F$. Moreover, these anamorphisms are
-exactly hylomorphisms on the recursive $F$-coalgebra and the
-algebra $\InIso$ for the inductive data type $\mu F$. This fact falls out
-from the uniqueness property of the hylomorphism and the fact that recursive
-anamorphisms satisfy the same equation.
+inductive data type in the following way. A recursive coalgebra can be applied
+only finitely many times, therefore when this is applied to an anamorphism the
+only possible types it can produce from the seed are the finite ones. We call
+this special kind of recursion scheme \emph{recursive anamorphism}. We can show
+that recursive anamorphisms of type $X \to \nu F$ can also be given the type
+$X \to \mu F$. Moreover, these anamorphisms are exactly hylomorphisms on the
+recursive $F$-coalgebra and the algebra $\InIso$ for the inductive data type
+$\mu F$. This fact falls out from the uniqueness property of the hylomorphism
+and the fact that recursive anamorphisms satisfy the same equation.
 
@@
 proceed onto describing how to formalise hylomorphisms in Coq. We first
-formalise container functors as a tool to represent polynomial functors (Section~\ref{sec:containers}).
-Then we formalise algebras for container extensions (Section~\ref{sec:coq-algebras}), then
-we formalise coalgebras (Section~\ref{sec:coq-coalg}) and then we put together
-these notions to formalise recursive coalgebras and hylomorphisms
+formalise container functors as a tool to represent polynomial functors
+(Section~\ref{sec:containers}). Then we formalise algebras for container
+extensions (Section~\ref{sec:coq-algebras}), then we formalise coalgebras
+(Section~\ref{sec:coq-coalg}) and then we put together these notions to
+formalise recursive coalgebras and hylomorphisms
 (Section~\ref{sec:coq-rec-coalgebras}).
@@
 \label{sec:containers}
-
-A straightforward mechanisation of functors as defined in
-Section~\ref{sec:rec-schemes} does not work in Coq due to the strict positivity
-condition. In fact, if we simply require that functors are functions
-\coq{F : Type -> Type}
-satisfying the necessary identity and composition properties, then
-we will not be able to construct least/greatest fixed points of this type.  A
-common solution to the strict positivity problem is the use
-\emph{containers}~\cite{AbbottAG05}.  A container $S \triangleright P$ is
-defined by a type of \emph{shapes} $S : \Type$ and a \emph{family of position
-types} indexed by shapes $P : S \to \Type$.  An \emph{extension} of this
-container is a functor $\llbracket S \triangleright P \rrbracket$, whose action
-on objects is given by
+Recursion schemes are defined using generic functors representing the scheme
+behind the recursive operator.  However, as not all functors possess enough
+fixed-points we need to be able to restrict to those mappings \coq{F : Type -> Type}
+that are so called \emph{strictly positive}.
+
+A common solution is to consider \emph{containers}~\cite{AbbottAG05}.  A
+container $S \triangleright P$ is defined by a type of \emph{shapes} $S : \Type$
+and a \emph{family of position types} indexed by shapes $P : S \to \Type$.  An
+\emph{extension} of this container is a functor
+$\llbracket S \triangleright P \rrbracket$, whose action on objects is given by
 $\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
@@
 
-The straightforward mechanisation of containers and container extensions is
-still not enough for our goals of remaining axiom-free, extracting to idiomatic
-OCaml code, and avoiding \ocaml{Obj.magic} in the extracted code. The main
-problem is that a container extension requires a function from a family of
-positions to a type: $\forall s : S.\; P\;s \to X$. On the Coq side, dealing
-with the equality of objects that contain (dependent) function types would force
-us to add the functional extensionality axiom, and axiom K (or equivalent) to
-deal with the heterogeneous equality of position.  For extraction, OCaml's type
-system is not powerful enough to represent such dependencies in the general
-case, which will lead to \ocaml{Obj.magic}.
-
-To avoid the functional extensionality axiom, we require that all the types used
-in our definitions are \emph{setoids}, where we will use \coq{x =e y} to
-represent setoid equality. All morphisms need to be \emph{proper} morphisms with respect
-to the function \emph{respectfulness} relation. In our development, we use
+Note however, that the type $\forall s : S.\; P\;s \to X$ would force us to add
+the functional extensionality axiom to deal with the heterogeneous equality of
+position. Thus, we require that all the types used in our definitions are
+\emph{setoids}, where we will use \coq{x =e y} to represent setoid equality. All
+morphisms need to be \emph{proper} morphisms with respect to the function
+\emph{respectfulness} relation. In our development, we use
 \begin{coqcode}
@@
 \end{coqcode}
-We add an automatic coercion from \coq{A ~> B} to \coq{A -> B}, provide tactics
+We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide tactics
 to automatically discharge proofs that \coq{forall x y, x =e y -> f x =e f y}
-for the simplest cases of \coq{f}, and we provide a library with
-definitions for basic building blocks: sum and product types, exponentials,
-composition, identity,
-\ldots
-
-To avoid axiom K, and extract containers to idiomatic OCaml code, we
-define containters in terms of shapes, \emph{base} positions (\coq{APos : Set}),
+for the simplest cases of \coq{f}.
+
+We define containers in terms of shapes, \emph{base} positions (\coq{APos : Set}),
 and \emph{decidable validity predicates} that
 specify when a base position is valid in a shape. Since validity predicates are in
-\coq{Prop}, they will be erased during extraction, and their decidability
-will make unnecessary the use of axiom K.
+\coq{Prop}, they will be erased during extraction:
 \begin{coqcode}
@@
 
-Since we require every type to be a setoid, we need to define a setoid equality
-(\coq{=e}) for container functors. Given two \coq{x y : App F X}, we define
-\coq{x =e y} as the setoid equality of the shapes, and the extensional equality of
-the contents, given equal shapes. I.e.\ \coq{x =e y} iff
+Given two \coq{x y : App F X}, we define
+\coq{x =e y} as the setoid equality of the shapes as
 \begin{coqcode}
@@
 \noindent
-where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its inverse $\InOp$ (see Section~\ref{sec:rec-schemes}).
+where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its inverse $\InOp$ (see Section~\ref{sec:category-theory}).
 As an example, the initial \coq{F}-algebra for the
@@
 We prove that catamorphisms satisfy the universality property we explained in
-Section~\ref{sec:rec-schemes}:
+Section~\ref{sec:category-theory}:
 \begin{coqcode}
@@
 
-As we mentioned in Section~\ref{sec:rec-schemes}, given an $F$-algebra and
+As we mentioned in Section~\ref{sec:category-theory}, given an $F$-algebra and
 $F$-coalgebra, the hylomorphism is the unique solution (when it exists) to the
@@
 As we mention in
-Section~\ref{sec:rec-schemes}, we define recursive anamorphisms as
+Section~\ref{sec:category-theory}, we define recursive anamorphisms as
 hylomorphisms built by using a recursive coalgebra, and the respective

### 2025-03-17 paper/paper.tex
shrink related work even more

@@
 
-Tesson et al.\ demonstrated the efficacy of leveraging Coq to establish an
-approach for implementing a robust system dedicated to verifying the correctness
-of program transformations for functions that manipulate
-lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
+% Tesson et al.\ demonstrated the efficacy of leveraging Coq to establish an
+% approach for implementing a robust system dedicated to verifying the correctness
+% of program transformations for functions that manipulate
+% lists~\cite{10.1007/978-3-642-17796-5_10}.
+Tesson et al.\ used Coq to mechanically verify program transformations for functions that manipulate lists~\cite{10.1007/978-3-642-17796-5_10}.
+Murata and Emoto went further and
 formalised recursion schemes in Coq~\cite{MurataE19}. Their development does not
@@
 
-Fusion in functional programming is a general technique for combining multiple
-computations into a single pass to eliminate intermediate data structures or
-repeated traversals, improving efficiency. The fusion we have discussed in this
+% Fusion in functional programming is a general technique for combining multiple
+% computations into a single pass to eliminate intermediate data structures or
+% repeated traversals, improving efficiency.
+The fusion we have discussed in this
 paper relating to program calculation is one form of fusion, but there are other
-approaches, and prior research has explored the mechanisation and verification
-of fusion transformations.
-
+approaches.
+%
 Fusion has been applied to recursive tree traversals, in order to reduce the
@@
 has been verified mechanically in Coq~\cite{10.1145/3377555.3377884}.
-
+%
 Another related family transformations are worker/wrapper
@@
 using folds via fusion~\cite{hutton10}. Their work does not focus on the use of
-an interactive proof assistant, but rather on the manual use of equational
-reasoning to re-write programs and on the automatic transformation of functional
+an interactive proof assistant, but rather on the automatic optimisation of
 programs by a compiler.
@@
 a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
-Take, for instance, division by iterated subtraction, which is inherently
-non-structurally recursive since it involves recursion based on the result of a
-subtraction. There is a nonstandard implementation of divivion found on Coq's
-standard library, which involves a four-argument function that effectively
-combines subtraction and division. Similarly, the mergesort in Coq's standard
+% Take, for instance, division by iterated subtraction, which is inherently
+% non-structurally recursive since it involves recursion based on the result of a
+% subtraction. There is a nonstandard implementation of divivion found on Coq's
+% standard library, which involves a four-argument function that effectively
+% combines subtraction and division.
+% Similarly,
+The mergesort in Coq's standard
 library uses an ``explicit stack of pending merges'' in order to avoid issues
@@
 fusion laws.
-
-% \subsection{Fusion}\label{sec:fusion}
-
-% Fusion in functional programming is a general technique for combining multiple
-% computations into a single pass to eliminate intermediate data structures and
-% improve efficiency. The fusion we have discussed in this paper relating to
-% program calculation is one form of fusion, but there are other approaches.
-
-% \emph{Deforestation}, introduced by Wadler~\cite{wadler88}, is a specific
-% approach to fusion that aims to remove intermediate trees (such as lists)
-% created between chained operations. A well-known method of deforestation is
-% \haskell{foldr/build} fusion~\cite{gill93}, which optimizes list-processing by
-% fusing list-producing functions (like \haskell{map} or \haskell{filter}) with
-% list-consuming ones (like \haskell{foldr}), avoiding the creation of temporary
-% lists.
-% %
-% Extending this idea, stream fusion generalizes the concept to handle other data
-% types, using a lazy stream abstraction to represent sequences and fuse
-% transformations like \haskell{map} and \haskell{fold}~\cite{coutts07}.
-% Both approaches eliminate redundant structures and ensure computations happen in
-% a single efficient pass over the data.
-
-% Another dimension of fusion is \emph{tupling}, a technique used to combine
-% multiple recursive functions into a single function by grouping their results
-% into a tuple~\cite{hughes89}. This approach helps avoid redundant traversals of
-% the same data structure when multiple operations are applied.
-% In addition to
-% techniques like \haskell{foldr/build} fusion and stream fusion, tupling
-% complements deforestation by allowing multiple recursive functions to share a
-% single traversal, further optimizing computations and minimizing intermediate
-% data structures.
-
-% In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
-% equational reasoning about recursive programs~\cite{DANVY_2022}. Fold-unfold
-% lemmas enable optimizations like fusion by allowing programs to be
-% systematically simplified, including eliminating intermediate structures.
-% Additionally, an approach to fusion called the worker/wrapper
-% transformation~\cite{gill09} was formalised and mechanised in
-% Agda~\cite{SCULTHORPE_HUTTON_2014}.
 

### 2025-03-17 paper/paper.tex
shrink related work more

@@
 formalised recursion schemes in Coq~\cite{MurataE19}. Their development does not
-include hylomorphisms and dynamorphisms, and relies on the functional
-extensionality axiom, as well as further extensionality axioms for each
-coinductive datatype that they use. They do not discuss the extracted OCaml code
-from their formalisation. Larchey-Wendling and Monin encode recursion schemes in
+include hylomorphisms and dynamorphisms, relies on various additional axioms, and does not discuss the extracted OCaml code.
+% the functional
+% extensionality axiom, as well as further extensionality axioms for each
+% coinductive datatype that they use.
+% They do not discuss the extracted OCaml code
+% from their formalisation.
+Larchey-Wendling and Monin encode recursion schemes in
 Coq, by formalising computational graphs of algorithms~\cite{larchey2022braga}.

### 2025-03-18 paper/paper.tex
bananas
Minor changes

--- 2025-03-17 paper/paper.tex
@@
 
-\paragraph{Recursive Anamorphisms}
+\paragraph*{Recursive Anamorphisms}
 Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
@@
 the functional extensionality axiom to deal with the heterogeneous equality of
-position. Thus, we require that all the types used in our definitions are
+positions. Thus, we require that all the types used in our definitions are
 \emph{setoids}, where we will use \coq{x =e y} to represent setoid equality. All
@@
 \end{coqcode}
-We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide tactics
-to automatically discharge proofs that \coq{forall x y, x =e y -> f x =e f y}
-for the simplest cases of \coq{f}.
+We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide
+tactics to automatically discharge proofs that morphisms are proper,  for some
+simple cases.
 

### 2025-03-18 paper/lensbans.tex

@@
 
-\newcommand{\lbanana}{%
-  \mathopen{\tikz[baseline=0.25em,xscale=0.006em,yscale=0.006em]
-  \draw[join=round] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
-              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;\mspace{1mu}}}
-\newcommand{\rbanana}{%
-  \mathclose{\mspace{1mu}\tikz[baseline=0.25em,xscale=-0.006em,yscale=0.006em]
-  \draw[join=round] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
-              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;}}
 % filled versions

### 2025-03-18 paper/paper.tex

--- 2025-03-17 paper/paper.tex
@@
 \newcommand{\anamor}[1]{{\llens\, #1\, \rlens}}
-\newcommand{\catamor}[1]{\lbans\, #1\, \rbans}
-\newcommand{\cata}[1]{\lbans #1 \rbans}
+\newcommand{\catamor}[1]{\lbanana\, #1\, \rbanana}
+\newcommand{\fold}[1]{\catamor{#1}}
+\newcommand{\cata}[1]{\catamor{#1}}
 \newcommand{\ana}[1]{\llens #1 \rlens}
@@
 \newcommand{\cohylo}[2]{\ana{#1 \to #2}}
-\newcommand{\fold}[1]{\catamor{#1}}
 \newcommand{\unfold}[1]{\anamor{#1}}
@@
 \newminted{ocaml}{fontsize=\small}
+
+\newsavebox{\lbananabox}
+\newcommand{\lbananamacro}{%
+  \begin{tikzpicture}[baseline=0.25em,xscale=0.005em,yscale=0.005em]
+  \draw[solid, join=round] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
+              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;
+\end{tikzpicture}
+}
+\savebox{\lbananabox}{\lbananamacro}
+\newcommand{\lbanana}{\mathopen{\usebox{\lbananabox}\hspace{-0.6ex}}}
+
+\newsavebox{\rbananabox}
+\newcommand{\rbananamacro}{%
+  \begin{tikzpicture}[baseline=0.25em,xscale=-0.005em,yscale=0.005em]
+  \draw[solid, join=round] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
+              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;
+  \end{tikzpicture}
+}
+\savebox{\rbananabox}{\rbananamacro}
+\newcommand{\rbanana}{\mathclose{\usebox{\rbananabox}}}
 

### 2025-03-18 paper/paper.tex
Some notes on correctness

@@
 initial \coq{F}-algebra.
-In other words, in this development we have defined recursive anamorphisms on inductive data
-types.  We might have as well defined them on the subtype of finite elements of
-coinductive data types using a predicate which states when an element of a
-coinductive data type is finite:
+In other words, in this development we have defined recursive anamorphisms on
+inductive data types.  We might have as well defined them on the subtype of
+finite elements of coinductive data types using a predicate which states when an
+element of a coinductive data type is finite:
 \begin{coqcode}
@@
 catamorphisms and anamorphisms, by using the above isomorphism. In our work,
-however, we use \emph{recursive} anamorphisms, defined as follows for
-a recursive coalgebra \coq{c}:
+however, we use \emph{recursive} anamorphisms, defined as hylomorphisms on the
+recursive recursive coalgebra \coq{c}, and the initial \coq{F}-algebra:
 \begin{coqcode}
@@
 % where \coq{h} is a recursive coalgebra.
+%
+
+\subsubsection{Proving Correctness}
+
+Suppose that we have an algebra \coq{a : Alg F B}, a recursive coalgebra
+\coq{c : RCoalg F A}, and a property \coq{P : A -> B -> Prop}, and we want to
+guarantee that the hylomorphism satisfies this property:
+\coq{forall x, P x (hylo a c x)}.
+Such proofs in our framework are done by induction on the proof that the
+recursive coalgebra terminates. In particular, the induction hypothesis would be
+that \coq{P} is satisfied by any recursive call of the hylomorphism
+\begin{coqcode}
+forall e, P (cont (c x) e) (hylo a c (cont (c x) e))
+\end{coqcode}
+\noindent
+Then, by the uniqueness property of hylomorphisms, it is enough to prove that
+\begin{coqcode}
+P x ((a \o fmap (hylo a c) \o c) x)
+\end{coqcode}
+We will see in Section~\ref{sec:examples} one example where we will prove that
+the output of a sorting algorithm is a sorted permutation of the input, and we
+will show that these proofs are comparable to other techniques for
+non-structural recursion.
 
@@
 Defined.
-Definition c_split : Coalg (TreeF unit int) (list int).
+Definition split : Coalg (TreeF unit int) (list int).
 |{ x ~> match x with
@@
 \end{coqcode}
-We prove that the coalgebra \coq{c_split} is recursive by showing
+We prove that the coalgebra \coq{split} is recursive by showing
 that it respects the ``less-than'' relation on the length of the lists.
 %\begin{coqcode}
-%Lemma split_fin : respects_relation c_split (@length int) lt.  (* *) Qed.
+%Lemma split_fin : respects_relation split (@length int) lt.  (* *) Qed.
 %
@@
 %\end{coqcode}
-The code that we extract for \coq{hylo merge c_split} is the following:
+The code that we extract for \coq{hylo merge split} is the following:
 \begin{ocamlcode}
@@
 The mergesort algorithm can be defined analogously.
+
+\paragraph*{Proving Quicksort Correct}
+
+We show that the output of \coq{hylo merge split} produces a sorted
+permutation of the input:
+\begin{coqcode}
+forall (l : list int), Sorted (hylo merge split l) /\ Perm l (hylo merge split l).
+\end{coqcode}
+The proof in our framework starts by induction on the proof that \coq{split}
+is recursive, and that it terminates on \coq{l}, i.e. using tactic
+\coq{induction (recP split l)}. Then, we can unfold
+\coq{hylo merge split} to \coq{merge \o fmap (hylo merge split) \o split} using
+hylomorphism uniqueness, and the proof would then proceed equally to any other
+similar proof. Overall, our proof is shorter than alternative implementations, and of comparable complexity\footnote{The QSort definition and
+correctness proof in the equations package is under 200LOC --
+\href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
+and other online proofs of correctness are of similar complexity
+\href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
 % and can be found in the
@@
 \begin{coqcode}
-Definition qsort_times_two := Lmap times_two \o hylo merge c_split.
+Definition qsort_times_two := Lmap times_two \o hylo merge split.
 \end{coqcode}
@@
 After applying hylo fusion and the necessary rewrites, the hylomorphism that we
-extract is \coq{hylo (merge \o natural times_two) c_split}. In this definition,
+extract is \coq{hylo (merge \o natural times_two) split}. In this definition,
 \coq{natural} defines a natural transformation by applying \coq{times_two} to

### 2025-03-18 paper/paper.tex
minor

--- 2025-03-18 paper/paper.tex
@@
 
-\paragraph*{Recursive Anamorphisms}
+\paragraph{Recursive Anamorphisms}
 Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
@@
 the functional extensionality axiom to deal with the heterogeneous equality of
-positions. Thus, we require that all the types used in our definitions are
+position. Thus, we require that all the types used in our definitions are
 \emph{setoids}, where we will use \coq{x =e y} to represent setoid equality. All
@@
 \end{coqcode}
-We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide
-tactics to automatically discharge proofs that morphisms are proper,  for some
-simple cases.
+We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide tactics
+to automatically discharge proofs that \coq{forall x y, x =e y -> f x =e f y}
+for the simplest cases of \coq{f}.
 
@@
 \noindent
-where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its inverse $\InOp$ (see Section~\ref{sec:category-theory}).
-As an example, the initial \coq{F}-algebra for the
-container that is isomorpic to the functor
-\coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
-being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
-operation \coq{Cons : A * LFix F -> LFix F}.
+where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its
+inverse $\InOp$ (see Section~\ref{sec:category-theory}).  As an example, the
+initial \coq{F}-algebra for the container that is isomorpic to the functor
+\coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra being
+defined by the empty list \coq{Empty : unit -> LFix F} and the cons operation
+\coq{Cons : A * LFix F -> LFix F}.
 

### 2025-03-18 paper/paper.tex
Changes

--- 2025-03-18 paper/paper.tex
@@
 
-Any polynomial functor can be represented as a container (see accompanying code
-for a mechanisation).  For example, a container that is equivalent to
-$F\; X = 1 + A \times X$ requires a shape that can distinguish two cases,
-indicating that there are two constructors to build an object of type $F\;X$,
-e.g. $S_{F} = 1 + A$. We need to consider two cases for the family of positions:
+Any polynomial functor can be represented as a container (see accompanying
+code).  For example, a container that is equivalent to $F\; X = 1 + A \times X$
+requires a shape that can distinguish two cases, indicating that there are two
+constructors to build an object of type $F\;X$, e.g. $S_{F} = 1 + A$. We need to
+consider two cases for the family of positions:
 $P_{F}(\mathsf{inj}_{1}\;*) = 0$, because there are no occurrences of $X$ in the
@@
 
-Note however, that the type $\forall s : S.\; P\;s \to X$ would force us to add
-the functional extensionality axiom to deal with the heterogeneous equality of
-positions. Thus, we require that all the types used in our definitions are
-\emph{setoids}, where we will use \coq{x =e y} to represent setoid equality. All
-morphisms need to be \emph{proper} morphisms with respect to the function
-\emph{respectfulness} relation. In our development, we use
+The type $\forall s : S.\; P\;s \to X$ would force us to add
+the functional extensionality axiom. Thus, we require that all the types used in
+our definitions are \emph{setoids}, where we use \coq{x =e y} to represent
+setoid equality. All morphisms need to be \emph{proper} morphisms that respect
+setoid equalities. In our development, we use
 \begin{coqcode}
@@
 
-We define containers in terms of shapes, \emph{base} positions (\coq{APos : Set}),
-and \emph{decidable validity predicates} that
-specify when a base position is valid in a shape. Since validity predicates are in
-\coq{Prop}, they will be erased during extraction:
+%% WARNING: removing the point about axiom K leads to leaving the use of
+%% decidable validity predicates unexplained -- do not remove!
+We define containers in terms of shapes, \emph{base} positions
+(\coq{APos : Set}), and \emph{decidable validity predicates} that specify when a
+base position is valid in a shape. Since validity predicates are in \coq{Prop},
+they will be erased during extraction, and since they are decidable, they
+prevent the use of axiom K to deal with heterogeneous equalities of positions:
 \begin{coqcode}
@@
 
-Given two \coq{x y : App F X}, we define
-\coq{x =e y} as the setoid equality of the shapes as
+%% Make sure you do not cut text so much that it ends up being inconsistent with
+%% the code! I just restored an old explanation and cut some of it in a consistent way
+We define the setoid equality of two container extensions \coq{x y : App F X} as
 \begin{coqcode}

### 2025-03-18 paper/paper.tex
Changes

@@
 \begin{coqcode}
-Record App F (X : Type) := {shape : Shape F; cont : {p : APos | valid s p = true} -> X}.
+Record App F X := {shape : Shape F; cont : {p : APos F | valid s p = true} -> X}.
 \end{coqcode}
@@
 operations of the algebra $F\;A \to A$.  Given a type \coq{A} and a container
-\coq{F}, an `\coq{App F}'-algebra is a pair given by the carrier \coq{A}, and the
-\emph{structure map} of type:
+\coq{F}, an `\coq{App F}'-algebra (or `\coq{F}-algebra' for short) is a pair
+given by the carrier \coq{A}, and the \emph{structure map} of type:
 \begin{coqcode}
@@
 \end{coqcode}
-
-\noindent
-For simplicity, we will abuse terminology, and call these
-\coq{F}-algebras, instead of `\coq{App F}'-algebras.
-
-The initial \coq{F}-algebra can be defined as follows, thanks to the use of
-containers:
+The initial \coq{F}-algebra can be defined as the inductive type with a finite
+number of uses of \coq{App F}
 \begin{coqcode}
@@
 \end{coqcode}
-
-\noindent
-where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its
+\coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its
 inverse $\InOp$ (see Section~\ref{sec:category-theory}).  As an example, the
@@
 
-We define \coq{LFix} as a setoid, where its equivalence relation can be
-described as the least fixed point of \coq{AppR} (see
-Section~\ref{sec:types-morphisms}). We define smart constructors for the
-isomorphism of least fixed points as respectful morphisms:
+The \coq{LFix F} setoid equality is the least fixed point of the \coq{App F}
+setoid equality. We define smart constructors for the isomorphism of least fixed
+points as proper morphisms:
 \begin{coqcode}
@@
 
-Catamorphisms are constructed, as expected, so that they structurally
-deconstruct the datatype, call themselves recursively, and then compose the
-result using an \coq{F}-algebra.
-\begin{coqcode}
-Definition cata_f (alg : Alg F A) : LFix -> A
-  := fix f (x : LFix)
-     := match x with
-        | LFix_in ax =>
-             alg {shape := shape ax; cont := fun e => f (cont ax e)}
+Catamorphisms are constructed so that they structurally deconstruct the
+datatype, call themselves recursively, and then compose the result using an
+\coq{F}-algebra.
+\begin{coqcode}
+Definition cata_f (a : Alg F A) : LFix -> A
+  := fix f (x : LFix) := match x with
+        | LFix_in x => a {shape := shape x; cont := fun e => f (cont x e)}
         end.
@@
 
-It is easy to show that this function is a respectful morphism of
-\coq{F}-algebras. In fact, it is possible to define it as a map of the
-following type:
+It is easy to show that this function is a proper morphism of \coq{F}-algebras.
+In fact, it is possible to define it as a map of the following type:
 \begin{coqcode}

### 2025-03-18 paper/lensbans.tex
banana
lenses

@@
+\newsavebox{\lbananabox}
+\newcommand{\lbananamacro}{%
+  \begin{tikzpicture}[baseline=0.25em,xscale=0.005em,yscale=0.005em]
+  \draw[solid, join=round] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
+              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;
+\end{tikzpicture}
+}
+\savebox{\lbananabox}{\lbananamacro}
+\newcommand{\lbanana}{\mathopen{\usebox{\lbananabox}\hspace{-0.6ex}}}
+
+\newsavebox{\rbananabox}
+\newcommand{\rbananamacro}{%
+  \begin{tikzpicture}[baseline=0.25em,xscale=-0.005em,yscale=0.005em]
+  \draw[solid, join=round] (2,0) to[out=140,in=-90] (0,3) to[out=90,in=-140] (2,6) -- (2.1,5.9)
+              to[out=-120,in=90] (1.2,3) to[out=-90,in=120] (2.1,0.1) -- cycle;
+  \end{tikzpicture}
+}
+\savebox{\rbananabox}{\rbananamacro}
+\newcommand{\rbanana}{\mathclose{\usebox{\rbananabox}}}
+
 \newsavebox{\lbansbox}
@@
 \newcommand{\llensmacro}{%
-\begin{tikzpicture}[baseline=0.45ex,xscale=0.006em,yscale=0.012ex]%
-\draw[solid,join=round] (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9) to[out=-120, in=90] (0.7,3) to[out=-90, in=120] (1.5,0.1) -- cycle;%
+\begin{tikzpicture}[baseline=0.25em,xscale=0.006em,yscale=0.006em]
+\draw[solid, join=round] (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9) to[out=-120,in=90] (0.7,3) to[out=-90,in=120] (1.5,0.1) -- cycle;%
 \end{tikzpicture}}
@@
 \newcommand{\rlensmacro}{%
-\begin{tikzpicture}[baseline=0.45ex,xscale=-0.006em,yscale=0.012ex]%
-\draw[solid,join=round] (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9) to[out=-120, in=90] (0.7,3) to[out=-90, in=120] (1.5,0.1) -- cycle;%
+\begin{tikzpicture}[baseline=0.25em,xscale=-0.006em,yscale=0.006em]
+  \draw[join=round] (1.4,0) -- (0,0) -- (0,6) -- (1.4,6) -- (1.5,5.9)
+  to[out=-120,in=90] (0.7,3) to[out=-90,in=120] (1.5,0.1) -- cycle;
 \end{tikzpicture}}

### 2025-03-18 paper/paper.tex
banana
minor

@@
 \newcommand{\anamor}[1]{{\llens\, #1\, \rlens}}
-\newcommand{\catamor}[1]{\lbans\, #1\, \rbans}
-\newcommand{\cata}[1]{\lbans #1 \rbans}
+\newcommand{\catamor}[1]{\lbanana\, #1\, \rbanana}
+\newcommand{\cata}[1]{\catamor{#1}}
 \newcommand{\ana}[1]{\llens #1 \rlens}
@@
 \label{sec:containers}
-Recursion schemes are defined using generic functors representing the scheme
-behind the recursive operator.  However, as not all functors possess enough
-fixed-points we need to be able to restrict to those mappings \coq{F : Type -> Type}
-that are so called \emph{strictly positive}.
+Recursion schemes are formulated using generic functors that encapsulate the
+structure of the recursive operator. However, since not all functors have the
+suitable fixed points, it is necessary to restrict our focus to those mappings
+of the form \coq{F : Type -> Type} which are additionally \emph{strictly positive}.
 

### 2025-03-18 paper/paper.tex
Updates

@@
 The initial \coq{F}-algebra can be defined as the inductive type with a finite
-number of uses of \coq{App F}
+number of \coq{App F}
 \begin{coqcode}
@@
 \end{coqcode}
-
-It is easy to show that this function is a proper morphism of \coq{F}-algebras.
-In fact, it is possible to define it as a map of the following type:
+It is easy to prove that catamorphisms respect setoid equalities.  In fact, we
+define it as a proper morphism:
 \begin{coqcode}
@@
 \end{coqcode}
-
-\noindent
-We prove that catamorphisms satisfy the universality property we explained in
-Section~\ref{sec:category-theory}:
+Finally, we prove that catamorphisms satisfy their universality property (see
+Section~\ref{sec:category-theory}):
 \begin{coqcode}
@@
 \end{coqcode}
-
 Dually to the initial \coq{F}-algebra, a final \coq{F}-coalgebra is the greatest
-fixed-point for \coq{App F}. We define it using a coinductive data type:
+fixed-point of \coq{App F}. We define it using a coinductive data type:
 \begin{coqcode}
@@
 \end{coqcode}
-
-\noindent
-where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix_in} is its
+\coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix_in} is its
 inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
 defined as a setoid, with an equivalence relation that is the greatest fixpoint
-of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
-greatest fixed points:
+of the \coq{App F} setoid equality. Additionally, we define smart constructors
+for the isomorphism of greatest fixed points:
 \begin{coqcode}
@@
 \end{coqcode}
-
 The greatest fixed-point is a terminal \coq{F}-coalgebra in the sense that it
@@
 Definition ana_f_ (c : Coalg F A) :=
-  cofix f x :=
-   let cx := c x in
+  cofix f x := let cx := c x in
    GFix_in { shape := shape cx; cont := fun e => f (cont cx e) }.
@@
 \end{coqcode}
-In words, for any \coq{F}-coalgebra, if there is any other function \coq{f}
-that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the
-same coalgebra.
+In words, for any \coq{F}-coalgebra, if there is any other function \coq{f} that
+is an \coq{F}-coalgebra homomorphism then it must be the anamorphism on the same
+coalgebra.
 
@@
 the hylomorphism equation. These are hylomoprhisms where the coalgebra is
-\emph{recursive}, i.e. coalgebras that terminate on all inputs. We represent
+\emph{recursive}, i.e.\ coalgebras that terminate on all inputs. We represent
 recursive coalgebras using a predicate that states that a coalgebra terminates
@@
 \end{coqcode}
-
-\noindent
-For convenience, we package recursive coalgebras with their proofs that they
-terminate for all inputs:
+For convenience, we package recursive coalgebras with proofs that they terminate
+for all inputs:
 \begin{coqcode}
@@
 Recursive hylomorphisms are implemented in Coq recursively on the structure of
-the proof for the type (\coq{RecF}) as follows:
+the proof that the anamorphism terminates on an input:
 \begin{coqcode}
@@
   := fix f x H
-     := match c x as cx
-          return (forall e : Pos (shape cx), RecF c (cont cx e)) -> B
-        with
-        | cx => fun H =>
-            a { shape := shape cx ; cont := fun e => f (cont cx e) (H e) }
+     := match c x as cx return (forall e : Pos (shape cx), RecF c (cont cx e)) -> B with
+        | cx => fun H => a { shape := shape cx ; cont := fun e => f (cont cx e) (H e) }
         end (RecF_inv H).
@@
 the recursive calls.
-
-\noindent
 As we did with catamorphisms and anamorphisms, we
-prove that \coq{hylo_def} is respectful, and use this proof to build the
+prove that \coq{hylo_def} is a proper morphism, and use this proof to build the
 corresponding higher-order proper morphism:
@@
 Figure~\ref{fig:proof-fusion-l}. The steps of this proof are: (1) apply the
-uniqueness law of recursive hylomorphisms; (2) rewrite the goal using the
-property that functors preserve composition; (3) rewrite the goal using the
+uniqueness law of recursive hylomorphisms; (2) rewrite using that
+functors preserve composition; (3) rewrite using the
 condition of \coq{hylo_fusion_l}; (4) use the uniqueness law of hylomorphisms to
@@
 Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
-is isomorphic its corresponding the inductive data type \coq{LFix F}. This is
-easy to see. We first define a catamorphism \coq{ccata_f_} from the
+is isomorphic its corresponding the inductive data type \coq{LFix F}.
+We show this by
+defining a catamorphism \coq{ccata_f_} from the
 subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
@@
 the output of a sorting algorithm is a sorted permutation of the input, and we
-will show that these proofs are comparable to other techniques for
+show that these proofs are comparable to other techniques for
 non-structural recursion.

### 2025-03-19 paper/paper.tex
Rewriting section 4

@@
 \end{coqcode}
-
-\noindent
-The main advantage of this definition is that recursive anamorphisms
-compose with catamorphisms without the need to reason about termination
-and finiteness of values of coinductive types.
+The main advantage of this definition is that recursive anamorphisms compose
+with catamorphisms without the need to reason about termination and finiteness
+of values.
 % This predicate is equivalent to the proposition that
@@
 Then, by the uniqueness property of hylomorphisms, it is enough to prove that
+\coq{P} is satisfied by the unfolding of the hylomorphism:
 \begin{coqcode}
@@
 We will see in Section~\ref{sec:examples} one example where we will prove that
-the output of a sorting algorithm is a sorted permutation of the input, and we
-show that these proofs are comparable to other techniques for
-non-structural recursion.
+the output of a sorting algorithm is a sorted permutation of the input, and
+show that these proofs are comparable to other techniques for non-structural
+recursion, with the additional advantage that any result that is proven on a
+hylomorphism will also hold for any result of doing program calculation.
 
@@
 \subsection{Sorting Algorithms}\label{sec:sorting}
-Our first case study is divide-and-conquer sorting algorithms.  Encoding them
-in our framework will require the use of recursive hylomorphisms and
-termination proofs.  We complete the sorting algorithm examples by applying
-fusion optimisation.
-
-Both mergesort and quicksort are divide-and-conquer algorithms that can be
-captured by the structure of an hylomorphisms. The structure of the recursion
-is that of a binary tree.  For example, in the case of quicksort, a list is
-split into a pivot, the label of the node, and two sublists. We define the data
-functor of trees as follows:
+Our first case study is divide-and-conquer sorting algorithms encoded as
+recursive hylomorphisms. Our accompanying code contains both mergesort and
+quicksort as examples, but we will focus on quicksort for conciseness. We will
+also show how to use program calculation to apply a fusion optimisation.  The
+recursive structure of both mergesort and quicksort is a binary tree. For
+example, in the case of quicksort, a list is split into a pivot, the label of
+the node, and two sublists. We define the data functor of trees as follows:
 \begin{coqcode}
@@
 \end{coqcode}
-We define the functor as a container using the following shapes and positions:
+To define an equivalent container, we notice that this functor has two shapes,
+one for leaves, and one for nodes, and that nodes have two positions (akin to
+\coq{l r : X} in \coq{i_node}). We also note that, in this container, positions
+are only valid in nodes.
 \begin{coqcode}
@@
 \end{coqcode}
-These define a container, \coq{TreeF}, in a straightforward way, by making
-the positions of type \coq{Tpos} only valid in \coq{Node}.
-We define a series of definitions for tree container constructors and 
-destructors:
+We name \coq{TreeF} the container that has these shapes and (base) poisitions.
+For convenience, we define a series of wrappers to construct these containers:
 % \mpav{The following algebra should be called a\_in or l\_in whatever}
 \begin{coqcode}
-Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.
 Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
@@
 \end{coqcode}
-The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with
-the following algebra and coalgebra.
-\begin{coqcode}
-  Definition merge : App (TreeF unit int) (list int) ~> list int.
-|{ x : (App (TreeF unit int) (list int)) ~> (
-           match x with
-           | MkCont sx kx =>
+The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with the
+following algebra and coalgebra. In the below definition, we use \coq{posL} and
+\coq{posR} as wrappers for \coq{Lbranch} and \coq{Rbranch} with validity proofs.
+\begin{coqcode}
+Definition merge_f (x : App (TreeF unit int) (list int)) : list int :=
+  let (sx, kx) := x in
   match sx return (Container.Pos sx -> _) -> _ with
@@
   | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
-               end kx
-           end
-  )}|.
-Defined.
-Definition split : Coalg (TreeF unit int) (list int).
-|{ x ~> match x with
+  end kx.
+Definition qsplit_f (x : list int) : App (TreeF unit int) (list int) :=
+  match x with
   | nil => a_leaf tt
-        | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in
-                      a_node h l r
-        end}|.
-Defined.
-\end{coqcode}
-We prove that the coalgebra \coq{split} is recursive by showing
-that it respects the ``less-than'' relation on the length of the lists.
+  | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in a_node h l r
+  end.
+\end{coqcode}
+Our tactics automatically discharge the proof that these are proper morphisms
+(called \coq{merge} and \coq{qsplit}), and we prove that the coalgebra
+\coq{qsplit} is recursive by showing that it respects the ``less-than'' relation
+on the length of the lists.
 %\begin{coqcode}
@@
 %\end{coqcode}
-The code that we extract for \coq{hylo merge split} is the following:
+The code that we extract for \coq{hylo merge qsplit} is similar to a hand-written implementation:
 \begin{ocamlcode}
 let rec qsort = function
-| [] -> [] | h :: t ->
-  let (l, r) = partition (fun x0 -> leb x0 h) t in
+  | [] -> []
+  | h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
     let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
@@
 \end{ocamlcode}
-Note that Coq's code extraction is unable to inline \coq{x0}, but the resulting
-code is similar to a hand-written \coq{qsort}.
-The mergesort algorithm can be defined analogously.
 
@@
 
-We show that the output of \coq{hylo merge split} produces a sorted
+We show that the output of \coq{hylo merge qsplit} produces a sorted
 permutation of the input:
 \begin{coqcode}
-forall (l : list int), Sorted (hylo merge split l) /\ Perm l (hylo merge split l).
-\end{coqcode}
-The proof in our framework starts by induction on the proof that \coq{split}
+forall (l : list int), Sorted (hylo merge qsplit l) /\ Perm l (hylo merge qsplit l).
+\end{coqcode}
+The proof in our framework starts by induction on the proof that \coq{qsplit}
 is recursive, and that it terminates on \coq{l}, i.e. using tactic
-\coq{induction (recP split l)}. Then, we can unfold
-\coq{hylo merge split} to \coq{merge \o fmap (hylo merge split) \o split} using
-hylomorphism uniqueness, and the proof would then proceed equally to any other
-similar proof. Overall, our proof is shorter than alternative implementations, and of comparable complexity\footnote{The QSort definition and
-correctness proof in the equations package is under 200LOC --
+\coq{induction (recP qsplit l)}. Then, we can unfold
+\coq{hylo merge qsplit} to \coq{merge \o fmap (hylo merge qsplit) \o qsplit} using
+hylomorphism uniqueness, and the rest of the proof is standard. Overall, our
+proof is shorter than alternative implementations, and of comparable
+complexity\footnote{The QSort definition and correctness proof in the equations
+package is under 200LOC --
 \href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
@@
 \begin{coqcode}
-Definition qsort_times_two := Lmap times_two \o hylo merge split.
+Definition qsort_times_two := Lmap times_two \o hylo merge qsplit.
 \end{coqcode}
@@
 and \coq{times_two} multiplies every element of the list by two.
-We can use Coq's generalised rewriting, and \coq{hylo_fusion_l} to fuse
-\coq{times_two} into the RHS hylomorphism in \coq{qsort_times_two}.
-After applying hylo fusion and the necessary rewrites, the hylomorphism that we
-extract is \coq{hylo (merge \o natural times_two) split}. In this definition,
+We can use Coq's generalised rewriting and \coq{hylo_fusion_l} to fuse
+\coq{times_two} into \coq{hylo merge qsplit}.
+After the necessary rewrites, we obtain \coq{hylo (merge \o natural times_two) qsplit}. In this definition,
 \coq{natural} defines a natural transformation by applying \coq{times_two} to
-the shapes, and \coq{times_two} multiplies every pivot in the Quicksort tree by
-two. Our formalisation contains a proof that \coq{natural} is indeed a natural
-transformation, which relies on the fact that it preserves the structure of the
-shapes and, therefore, the validity of the positions.  The extracted OCaml code
+the shapes, thus multiplying every pivot by
+two.
+%% Our formalisation contains a proof that \coq{natural} is indeed a natural
+%% transformation, which relies on the fact that it preserves the structure of the
+%% shapes and, therefore, the validity of the positions.
+The extracted OCaml code
 is a single recursive traversal:

### 2025-03-19 paper/paper.tex
Changes

@@
 \coq{times_two} into \coq{hylo merge qsplit}.
-After the necessary rewrites, we obtain \coq{hylo (merge \o natural times_two) qsplit}. In this definition,
-\coq{natural} defines a natural transformation by applying \coq{times_two} to
-the shapes, thus multiplying every pivot by
-two.
+After the necessary rewrites, we obtain
+\coq{hylo (merge \o natural times_two) qsplit}.
+In this definition, \coq{natural} defines a natural transformation by
+applying \coq{times_two} to the shapes, thus multiplying every pivot by two.
 %% Our formalisation contains a proof that \coq{natural} is indeed a natural

### 2025-03-19 paper/paper.tex
Updates to section 4

@@
 \subsection{Knapsack}\label{sec:knap}
-We focus now on the formalisation and extraction of \emph{dynamorphisms} for
-dynamic programming, by using their encoding as a hylomorphism. We use the
+We show now the formalisation and extraction of \emph{dynamorphisms} for
+dynamic programming. We use the
 knapsack example from~\cite{HinzeWG15}. Dynamorphisms build a memoisation table
@@
 Definition MemoPos := Pg.
-Instance Memo : Cont MemoShape MemoPos
-  := { valid := valid \o pair (snd \o fst) snd }.
-
+Instance Memo : Cont MemoShape MemoPos := { valid := valid \o pair (snd \o fst) snd }.
 Definition Table := LFix Memo.
@@
 \end{coqcode}
-The hylomorphism for knapsack is as follows, where \coq{out_nat} is the
-recursive coalgebra for \coq{nat}.
+Given the recursive coalgebra for \coq{nat}, \coq{out_nat}, knapsack is defined
+as
 \begin{coqcode}
@@
 axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
-However, we prove a specific version of this theorem for the list generating
-container (i.e.\ the container whose least fixed point is a list), and use it
-to encode the example from Takano and Meijer~\cite{TakanoM95}:
+However, we prove a specific version of this theorem for the container whose
+least fixed point is a list, and use it to encode the example from Takano and
+Meijer~\cite{TakanoM95}:
 \begin{coqcode}

### 2025-03-19 paper/paper.tex
Abstract

@@
   Generic programming with recursion schemes provides a powerful abstraction for
-structuring recursion, in part due to the rigorous set of algebraic laws that
-they satisfy. These laws are the basis for reasoning about program equivalences
-and, therefore , they can be used for reasoning about program correctness and
-optimisations.  Some of these optimisations are successfully applied by
-compilers of (functional) languages.  Formalising recursion schemes in a type
-theory offers additional termination guarantees, but it often requires
-compromises affecting the resulting code, such as imposing performance
-penalties, requiring the assumption of additional axioms, or introducing unsafe
-casts into extracted code (e.g. \ocaml{Obj.magic} in OCaml).
-
-This paper presents the first Coq formalisation of a recursion scheme, called
-the \emph{hylomorphism}, along with its algebraic laws allowing for the
-mechanisation of all recognised (terminating) recursive algorithms. The key
-contribution of this paper is that this formalisation is fully axiom-free
+structuring recursion and provide a rigorous reasoning principle for program
+optimisations which have been successfully applied to compilers for functional
+languages. Formalising recursion schemes in a type theory offers additional
+termination guarantees, but it often requires compromising on performance,
+requiring the assumption of additional axioms, and/or introducing unsafe casts
+into extracted code.
+
+This paper presents the first Coq formalisation of \emph{hylomorphisms} allowing for the
+mechanisation of all recognised recursive algorithms. The key
+contribution of this paper is that this formalisation is fully \emph{axiom-free}
 allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
@@
 and demonstrate that the extracted OCaml code for the programs formalised in our
-framework is efficient, resembles code that a human programmer would write, and
-contains no occurrences of \ocaml{Obj.magic}.  Furthermore, we show the use of
+framework is efficient, humanly readable, and runnable. Furthermore, we show the use of
 the machine-checked proofs of the laws of hylomorphisms to do program

### 2025-03-20 paper/paper.tex
First draft, new intro, massaged related work.

@@
 contribution of this paper is that this formalisation is fully \emph{axiom-free}
-allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
+allowing for the extraction of safe, idiomatic functional code. We exemplify the
 framework by formalising a series of algorithms based on different recursive
 paradigms such as divide-and conquer, dynamic programming, and mutual recursion
-and demonstrate that the extracted OCaml code for the programs formalised in our
+and demonstrate that the extracted functional code for the programs formalised in our
 framework is efficient, humanly readable, and runnable. Furthermore, we show the use of
@@
 \label{sec:intro}
-Structured recursion schemes~\cite{HinzeW16,HinzeWG15} are powerful
+Structured recursion schemes are powerful
 abstractions that capture common patterns of recursion. The main benefit of
@@
 \end{minted}
-Instead of writing this function directly, we may observe that we can derive it
-by \emph{fusing} a regular \emph{quicksort} OCaml implementation, with
-\ocaml{map (fun x -> x * 2)}. This is the main idea behind program calculation:
-start
-from a simple specification, e.g.\
+This program is derived by \emph{fusing} a regular \emph{quicksort} OCaml implementation, with
+\ocaml{map (fun x -> x * 2)}. A lot of program calculations have this pattern:
+start from a simple specification, e.g.\
 $\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
-equivalences and algebraic laws to rewrite it to an optimised version  (e.g.\
-the fused OCaml implementation above). We will revisit a similar example
+equivalences and algebraic laws to rewrite it to an optimised version. We will revisit a similar example
 in Section~\ref{sec:sorting}.
 
-There is not much tool support to use program calculation techniques. In fact,
-uses of program calculation are done by  pen-and-paper proofs, and then
-translating the result to specific instances of recursion schemes, implemented
-in a programming language (generally Haskell).  There are some examples of
-implementations of program calculation techniques, but these implementations are
-scarce, not up to date, and not verified in a proof assistant. For example,
-Cunha et al.~\cite{DrHylo} automated program calculation techniques by a custom
-Haskell implementation.
-
-Several authors build tools for applying program calculation techniques by
-\emph{mechanising} them as part of a proof assistant. However, most of the
-existing work only mechanises a narrow subset of the know recursion schemes, or
-do not prove their algebraic
-laws~\cite{10.1007/978-3-642-17796-5_10,MurataE19,larchey2022braga}.  Indeed,
-most of the mechanisations of generic recursion schemes cover \emph{maps} and
-\emph{folds}, and few authors focus implement \emph{unfolds}. Omitting
-\emph{unfolds} severely limits the expressivity of the resulting mechanisations,
-and prevent them from being able to mechanise the most general recursion scheme:
-a divide-and-conquer algorithm which goes under the name of
-\emph{hylomorphism}~\cite{MeijerFP91,HuIT96}. This generality has been proven by
-Hinze et al.~\cite{HinzeWG15}, by showing that \emph{all known  recursion
-schemes are instances of hylomorphisms}.  To date, no work has mechanised
-hylomorphisms in the Coq proof assistant, together with their algebraic laws.
-
-Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
-divide-and-conquer computations in which termination is entirely enforced by the
-typing discipline. Their approach solves the problem of termination proofs as
-well as the performance of the code that is run \emph{within Coq}, but it does
-not allow for extraction of idiomatic OCaml code, and it is not well-suited for
-program calculation. This is unfortunate since code extration is what allows the
-execution of code that has been verified in
-Coq~\cite{OnoHTNH11,Larchey-Wendling23,MiculanP12,Sakaguchi20}.  In Abreu et
-al's approach, extraction (1) does not produce code with the same recursive
-structure  as existing manual implementations; and (2) leads to unsafe casts
-like \ocaml{Obj.magic} in the generated code. This latter is also problematic in
-that, for higher-order programs, simple interoperations can lead to incorrect
-behaviour or even segfaults~\cite{forster:hal-04329663} and, moreover, it
-invalidates the fast-and-loose principle~\cite{DanielssonHJG06}.
+This program calculation stems from the more general theory of hylomorphisms.  A
+hylomorphism is a structured recursion scheme representing the idea of a
+divide-and-conquer algorithm where, first, the problem is split into smaller
+subproblems,  (divide), then subsolutions are computed recursively, and,
+finally, these subsolutions are put together to compute the original solution
+(conquer). This general pattern of recursion induces a reasoning principle which
+is called fusion law which has been exemplified in the previous paragraph.
+
+Hylomorphism are very powerful non-structural recursive algorithms that can be
+used to implement structural recursion, as well as many other known recursion
+schemes such as mutual recursion, accumulators, primitive recursion and
+course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. This power
+comes at a price. Since hylomorphisms are not necessarily structural they need
+extra-care is needed when encoding these in a type theory. In particular, the
+divide part of the algorithm must be \emph{recursive} that is the input needs to
+be split into smaller parts (not necessarily following the structure of the
+input data).
+
+Implementations of recursion schemes currently focus on non-terminating
+languages (e.g. Haskell) and they would benefit from a type-theoretical
+formalisation which leverages the underlying prover's logic to ensure
+correctness of definitions.  While existing mechanisations do exist they either
+do not cover a large subset of the recursion schemes or are not suitable for
+program calculation or code extraction (see Related
+Work~\ref{sec:related-work}).
+
+In this work we mechanise hylomorphisms thus reaping the benefits of their
+generality and their algebraic properties. We do this in a way which allows the
+user to write specifications of their code, reason about program optimisations
+and eventually be able to extract humanly readable functional code which can be
+run which means it cannot contain unsafe casts. These are, in particular,
+problematic  for higher-order programs as simple interoperations with them can
+lead to incorrect behaviour or even segmentation
+faults~\cite{forster:hal-04329663} and, moreover, it invalidates the
+fast-and-loose principle~\cite{DanielssonHJG06}.
+
+To retain the hylomorphism's generality we encode data structures using
+polynomial functors and to ensure our extracted code is runnable we make a point
+to remain axiom-free and avoid indexed types in the definition of our recursion
+schemes.
 
@@
 \label{sec:related-work}
+
 %
@@
 %\end{itemize}
-Encoding recursion schemes in Coq is not new. We compare our work with other
+We compare our work with other
 encodings of program calculation techniques and recursion schemes in Coq, and
@@
 
-\subsection{Program Calculation}\label{sec:calccoq}
-
+% \textbf{Program Calculation}.
+% \label{sec:calccoq}
 Within the domain of program optimization, program calculation serves as a
@@
 
-\subsection{Fusion Optimisation}\label{sec:fusion}
-
+% \textbf{Fusion Optimisation}.
+% %
+% \label{sec:fusion}
 Fusion in functional programming is a general technique for combining multiple
@@
 %\mvol{Need more general citations for program calculation.}
-\subsection{Divide and Conquer Recursion}\label{sec:reccoq}
-
+% \textbf{Divide and Conquer Recursion}.%
+% \label{sec:reccoq}
 Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
@@
 Through experiments, we found that code extracted from their formalisation makes
-heavy use of \ocaml{Obj.magic} and has a generally complex structure that may be
+heavy use of unsafe casts and has a generally complex structure that may be
 hard to understand and integrate with other OCaml code. Due to the great benefit
@@
 
-\subsection{Termination Checking of Nonstructural Recursion}\label{sec:termcheck}
-
+%\textbf{Termination Checking of Nonstructural Recursion}.% \label{sec:termcheck}
 % Termination and productivity are non-trivial properties of programs, and various
@@
 % definitions will not be accepted by such checks.
-
 %In particular,

### 2025-03-20 paper/paper.tex
Fixes & typos

@@
 
-\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
+\keywords{hylomorphisms, program calculation, divide and conquer, fusion} %TODO mandatory; please add comma-separated list of keywords
 
@@
   Generic programming with recursion schemes provides a powerful abstraction for
-structuring recursion and provide a rigorous reasoning principle for program
+  structuring recursion and provides a rigorous reasoning principle for program
   optimisations which have been successfully applied to compilers for functional
@@
 
-This paper presents the first Coq formalisation of \emph{hylomorphisms} allowing for the
-mechanisation of all recognised recursive algorithms. The key
-contribution of this paper is that this formalisation is fully \emph{axiom-free}
-allowing for the extraction of safe, idiomatic functional code. We exemplify the
-framework by formalising a series of algorithms based on different recursive
-paradigms such as divide-and conquer, dynamic programming, and mutual recursion
-and demonstrate that the extracted functional code for the programs formalised in our
-framework is efficient, humanly readable, and runnable. Furthermore, we show the use of
-the machine-checked proofs of the laws of hylomorphisms to do program
+  This paper presents the first Coq formalisation of \emph{hylomorphisms}
+  allowing for the mechanisation of all recognised recursive algorithms. The key
+  contribution of this paper is that this formalisation is fully
+  \emph{axiom-free} allowing for the extraction of safe, idiomatic functional
+  code. We exemplify the framework by formalising a series of algorithms based
+  on different recursive paradigms such as divide-and conquer, dynamic
+  programming, and mutual recursion and demonstrate that the extracted
+  functional code for the programs formalised in our framework is efficient,
+  humanly readable, and runnable. Furthermore, we show the use of the
+  machine-checked proofs of the laws of hylomorphisms to do program
   optimisations.
@@
 \label{sec:intro}
-Structured recursion schemes are powerful
-abstractions that capture common patterns of recursion. The main benefit of
-structuring computation using recursion schemes, is that they enjoy
-well-estabished \emph{algebraic properties} that can serve as a foundation for
-reasoning about program equivalences, transformations, and
-optimisations (e.g.\ \emph{fusion} laws, or semi-automatic
+Structured recursion schemes are powerful abstractions that capture common
+patterns of recursion. The main benefit of structuring computation using
+recursion schemes, is that they enjoy well-estabished \emph{algebraic
+properties} that can serve as a foundation for reasoning about program
+equivalences, transformations, and optimisations (e.g.\ \emph{fusion} laws, or
+semi-automatic
 parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
-These algebraic properties led to their use in the context of
-\emph{program calculation}, where programmers would describe their code using
-simple, inefficient specifications in an \emph{algebra of
-programming}~\cite{BirddeMoor96:Algebra}, and then use the algebraic laws of
-this algebra of programming to \emph{calculate} an efficient version of the same
-algorithm. Suppose, for example, that we want to write a program that sorts a list of
+%
+These algebraic properties are widely used in the context of \emph{program
+calculation}, where programmers describe their code using simple, inefficient
+specifications in an \emph{algebra of programming}~\cite{BirddeMoor96:Algebra},
+and then use the algebraic laws of this algebra of programming to
+\emph{calculate} an efficient version of the same algorithm.
+%
+Suppose, for example, that we want to write a program that sorts a list of
 integers, and then multiplies by 2 all its elements. In OCaml, we may write this
@@
 \end{minted}
-This program is derived by \emph{fusing} a regular \emph{quicksort} OCaml implementation, with
-\ocaml{map (fun x -> x * 2)}. A lot of program calculations have this pattern:
-start from a simple specification, e.g.\
-$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
-equivalences and algebraic laws to rewrite it to an optimised version. We will revisit a similar example
-in Section~\ref{sec:sorting}.
+This program is derived by \emph{fusing} a regular \emph{quicksort} OCaml
+implementation, with
+\ocaml{map (fun x -> x * 2)}.
+A lot of program calculations have this pattern: start from a simple
+specification, e.g.\ $\mathsf{map}\;(\lambda x.\;x \times 2) \circ
+\mathsf{sort}$, and use program equivalences and algebraic laws to rewrite it to
+an optimised version. We will revisit a similar example in
+Section~\ref{sec:sorting}.
 
@@
 divide-and-conquer algorithm where, first, the problem is split into smaller
-subproblems,  (divide), then subsolutions are computed recursively, and,
-finally, these subsolutions are put together to compute the original solution
-(conquer). This general pattern of recursion induces a reasoning principle which
-is called fusion law which has been exemplified in the previous paragraph.
-
-Hylomorphism are very powerful non-structural recursive algorithms that can be
+subproblems (divide), then subsolutions are computed recursively, and, finally,
+these subsolutions are put together to compute the solution to the original
+problem (conquer).  This general pattern of recursion induces a reasoning
+principle, called \emph{fusion law}, which has been exemplified in the previous
+paragraph.
+
+Hylomorphisms are very powerful non-structural recursive algorithms that can be
 used to implement structural recursion, as well as many other known recursion
@@
 course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. This power
-comes at a price. Since hylomorphisms are not necessarily structural they need
-extra-care is needed when encoding these in a type theory. In particular, the
-divide part of the algorithm must be \emph{recursive} that is the input needs to
-be split into smaller parts (not necessarily following the structure of the
-input data).
+comes at a price. Since hylomorphisms are not necessarily structural, extra-care
+is needed when encoding them in a type theory. In particular, the divide part of
+the algorithm must be \emph{recursive} that is the input needs to be split into
+smaller parts (not necessarily following the structure of the input data).
 
@@
 do not cover a large subset of the recursion schemes or are not suitable for
-program calculation or code extraction (see Related
-Work~\ref{sec:related-work}).
-
-In this work we mechanise hylomorphisms thus reaping the benefits of their
+program calculation or code extraction (see Section~\ref{sec:related-work},
+Related Work).
+
+In this work we mechanise hylomorphisms, thus reaping the benefits of their
 generality and their algebraic properties. We do this in a way which allows the
-user to write specifications of their code, reason about program optimisations
-and eventually be able to extract humanly readable functional code which can be
-run which means it cannot contain unsafe casts. These are, in particular,
-problematic  for higher-order programs as simple interoperations with them can
-lead to incorrect behaviour or even segmentation
+user to write specifications of their code, reason about program optimisations,
+and eventually extract humanly readable functional code which can be run. Our
+extracted code does not contain unsafe casts. We focus strongly on avoiding
+unsafe casts because, even if the extracted code has been verified, simple
+interoperations with them can lead to incorrect behaviour or even segmentation
 faults~\cite{forster:hal-04329663} and, moreover, it invalidates the
@@
 
-To retain the hylomorphism's generality we encode data structures using
-polynomial functors and to ensure our extracted code is runnable we make a point
-to remain axiom-free and avoid indexed types in the definition of our recursion
-schemes.
+To preserve the generality of hylomorphisms, we encode data structures using
+polynomial functors. To ensure that the extracted code is runnable, we make a
+conscious effort to stay axiom-free and avoid using indexed types in the
+definition of our recursion schemes.
 
@@
 formalisation of \emph{hylomorphisms} that (1) is \emph{fully axiom-free}; (2)
-allows the extraction of idiomatic OCaml code; and (3) can use regular Coq
+allows the extraction of idiomatic code; and (3) can use regular Coq
 equalities to do program calculation, derive correct implementations, and apply
@@
   \item In Section~\ref{sec:containers} we formalise the type of container
-functors ensuring the presence of least and greatest fixed-points for functors
-and suitably adapted for program extraction.
+    functors, ensuring the presence of least and greatest fixed-points for
+    functors, and suitably adapted for program extraction.
   \item In Section~\ref{sec:recursion-schemes} we mechanise folds, unfolds, and
@@
     of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
-Furthermore we verify the short-cut fusion optimisation and show the extracted
-optimised code to OCaml.
+    Furthermore we verify the short-cut fusion optimisation and show the
+    extracted optimised code.
 \end{itemize}
@@
 A \emph{category} is a collection of objects $A, B, C$, denoted by $\Obj(\CatC)$
-and a collection of arrows $f, g, h$ between these objects, denoted by
-$\Arr (\CatC)$, such that there always exists an identity arrow
-$id_{A} : A \to A$ for each object $A$ and for two arrows $A \xrightarrow{f} B$
-and $B \xrightarrow{g} C$ there always exists an arrow
-$A \xrightarrow{g\circ f} C$ obeying the associativity law. We denote
-$\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to $B$ and we use the letters
-$\CatC$, $\CatD$, $\CatE\dots$ for categories\footnote[1]{For presentation
-purposes we shall not deal with size issues and assume all the categories are
-locally small.}. The initial object in a category, denoted by $0$, is the object such
-that for any other object $A$ there is a unique arrow $0 \xrightarrow{!} A$.
-Dually, the terminal object, denoted by $1$, is the object such that for any
-other object $A$ there is a unique arrow $A \xrightarrow{!} 1$. Initial and
-terminal object are unique up-to isomorphism.
+and a collection of arrows $f, g, h$ between these objects, denoted by $\Arr
+(\CatC)$, such that there always exists an identity arrow $id_{A} : A \to A$ for
+each object $A$ and for two arrows $A \xrightarrow{f} B$ and $B \xrightarrow{g}
+C$ there always exists an arrow $A \xrightarrow{g\circ f} C$ obeying the
+associativity law. We denote $\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to
+$B$ and we use the letters $\CatC$, $\CatD$, $\CatE\dots$ for
+categories\footnote[1]{For presentation purposes we shall not deal with size
+  issues and assume all the categories are locally small.}. The initial object
+in a category, denoted by $0$, is the object such that for any other object $A$
+there is a unique arrow $0 \xrightarrow{!} A$.  Dually, the terminal object,
+denoted by $1$, is the object such that for any other object $A$ there is a
+unique arrow $A \xrightarrow{!} 1$. Initial and terminal object are unique up-to
+isomorphism.
 
@@
 as the set inductively generated by the constructors $\nil : 1 \to \List{A}$ and
-$\cons : A \times \List{A} \to \List{A}$. The $\List{-}$ type is an \emph{endofunctor}
-$\Set \to \Set$ mapping objects and arrows in $\Set$ to $\Set$ itself.
-Its functorial action $\List{f} : \List{A} \to \List{B}$ is
-given by $\List{f}(\nil(*)) = \nil(*)$ and
-$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
-$\List{f}$ is well-defined as it recursively calls on a smaller argument.
-Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is
-the greatest set generated by the constructor
-$\cons : A \times \Str{A} \to \Str{A}$. The maps $\head : \Str{A} \to A$ and
-$\tail : \Str{A} \to \Str{A}$ can be easily constructed from the isomorphism.
+$\cons : A \times \List{A} \to \List{A}$. The $\List{-}$ type is a
+\emph{functor} $\Set \to \Set$ mapping objects and arrows in $\Set$ to $\Set$
+itself (such functors from a category to itself are called \emph{endofunctors}).
+Its functorial action $\List{f} : \List{A} \to \List{B}$ is given by
+$\List{f}(\nil(*)) = \nil(*)$ and $\List{f}(\cons(a,xs)) = \cons(f(a),
+\List{f}(xs))$. Notice that the definition $\List{f}$ is well-defined as it
+recursively calls on a smaller argument.  Similarly, the set of streams $\Str{A}
+\iso A \times \Str{A}$ is the greatest set generated by the constructor $\cons :
+A \times \Str{A} \to \Str{A}$. The maps $\head : \Str{A} \to A$ and $\tail :
+\Str{A} \to \Str{A}$ can be easily constructed from the isomorphism.
 
@@
 
-Let $(A,a)$ be an $F$-algebra and $(C,c)$ be an $F$-coalgebra. An arrow
-$C \to A$ is an \emph{hylomorphism}, written $h : (C,c) \rightarrowtail (A,a)$ if
-it satisfies
+Let $(A,a)$ be an $F$-algebra and $(C,c)$ be an $F$-coalgebra. An arrow $C \to
+A$ is an \emph{hylomorphism}, written $h : (C,c) \rightarrowtail (A,a)$ if it
+satisfies
 \begin{equation}
@@
 they will be erased during extraction, and since they are decidable, they
-prevent the use of axiom K to deal with heterogeneous equalities of positions:
+do not require axiom K to deal with heterogeneous equalities.
+Container extensions are defined as
 \begin{coqcode}
@@
 \begin{coqcode}
-Definition hylo_def (a : Alg F B) (c : Coalg F A)
-  : forall (x : A), RecF c x -> B
+Definition hylo_def (a : Alg F B) (c : Coalg F A) : forall (x : A), RecF c x -> B
   := fix f x H
@@
 \begin{coqcode}
-Lemma hylo_fusion_l
-  (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C) (f2 : B ~> C)
+Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C) (f2 : B ~> C)
   : f2 \o g1 =e g2 \o fmap f2 -> f2 \o hylo g1 h1 =e hylo g2 h1.
 
-Lemma hylo_fusion_r
-  (h1 : RCoalg F B) (g1 : Alg F C) (h2 : RCoalg F A) (f1 : A ~> B)
+Lemma hylo_fusion_r (h1 : RCoalg F B) (g1 : Alg F C) (h2 : RCoalg F A) (f1 : A ~> B)
   :  h1 \o f1 =e fmap f1 \o h2 -> hylo g1 h1 \o f1 =e hylo g1 h2.
@@
   \caption{Rewrite steps to prove \coq{hylo_fusion_l} in our mechanisation. The
-steps are exactly the same that woudld be required in a manual pen-and-paper
+steps are exactly the same that would be required in a manual pen-and-paper
 proof.}
@@
 \begin{coqcode}
-Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
-  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
-  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
+Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C) (g1 : Alg F B) (h2 : RCoalg F B)
+  : h2 \o g1 =e id -> hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.
 \end{coqcode}
@@
 \begin{coqcode}
-Inductive FinF : GFix F -> Prop :=
-| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
+Definition FinF : GFix F -> Prop := RecF g_out.
 \end{coqcode}
@@
     let hx := g_out x in
-      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
+      g (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
 \end{coqcode}
@@
 program calculation techniques for optimisation, and how can they be extracted
-to idiomatic OCaml code. Our examples are the Quicksort and Mergesort
-algorithms (Section~\ref{sec:sorting}), dynammic programming and Knapsack
-(Section~\ref{sec:knap}), and examples of the shortcut deforestation
-optimisation in our framework (Section~\ref{sec:shortcut}).
+to idiomatic functional code. Although our examples use OCaml as target
+language, extraction works equally well to other target languages. Our examples
+are the Quicksort and Mergesort algorithms (Section~\ref{sec:sorting}), dynammic
+programming and Knapsack (Section~\ref{sec:knap}), and examples of the shortcut
+deforestation optimisation in our framework (Section~\ref{sec:shortcut}).
 %The steps are the following:
@@
 extensionality axiom, as well as further extensionality axioms for each
-coinductive datatype that they use. They do not discuss the extracted OCaml code
+coinductive datatype that they use. They do not discuss the extracted code
 from their formalisation. Larchey-Wendling and Monin encode recursion schemes in
@@
 heavy use of unsafe casts and has a generally complex structure that may be
-hard to understand and integrate with other OCaml code. Due to the great benefit
+hard to understand and integrate with other code. Due to the great benefit
 of entirely avoiding termination proofs, it would be interesting to extend their
@@
 extensionality. In this work, we tackle these problems in a \emph{fully
-axiom-free} way that targets the extraction of idiomatic OCaml code.  This
+axiom-free} way that targets the extraction of idiomatic code.  This
 formalisation allows the use of program calculation techniques in Coq to derive

### 2025-03-20 paper/paper.tex
Typos

@@
 patterns of recursion. The main benefit of structuring computation using
-recursion schemes, is that they enjoy well-estabished \emph{algebraic
+recursion schemes, is that they enjoy well-established \emph{algebraic
 properties} that can serve as a foundation for reasoning about program
@@
 divide-and-conquer algorithm where, first, the problem is split into smaller
-subproblems (divide), then subsolutions are computed recursively, and, finally,
-these subsolutions are put together to compute the solution to the original
+sub-problems (divide), then sub-solutions are computed recursively, and, finally,
+these sub-solutions are put together to compute the solution to the original
 problem (conquer).  This general pattern of recursion induces a reasoning
@@
 An example of a recursive coalgebra is the \coq{partition} function in quicksort
-which destructures a list into a pivot and two sublist and as long as the
-sublists are smaller the partitioning function still yields a unique solution to
-the recursion scheme.  The uniqueness property of the hylo yields the following
+which destructures a list into a pivot and two sub-list and as long as the
+sub-lists are smaller the partitioning function still yields a unique solution to
+the recursion scheme.  The uniqueness property of the hylomorphisms yields the following
 fusion laws:
@@
 \end{align}
-Using the hylo fusion laws, we can prove the well-known \emph{deforestation
+Using the hylomorphism fusion laws, we can prove the well-known \emph{deforestation
 optimisation}, also known as the \emph{composition
@@
 inverse $\InOp$ (see Section~\ref{sec:category-theory}).  As an example, the
-initial \coq{F}-algebra for the container that is isomorpic to the functor
+initial \coq{F}-algebra for the container that is isomorphic to the functor
 \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra being
@@
 \emph{recursive hylomorphisms} which are guaranteed to have a unique solution to
-the hylomorphism equation. These are hylomoprhisms where the coalgebra is
+the hylomorphism equation. These are hylomorphisms where the coalgebra is
 \emph{recursive}, i.e.\ coalgebras that terminate on all inputs. We represent
@@
 language, extraction works equally well to other target languages. Our examples
-are the Quicksort and Mergesort algorithms (Section~\ref{sec:sorting}), dynammic
+are the Quicksort and Mergesort algorithms (Section~\ref{sec:sorting}), dynamic
 programming and Knapsack (Section~\ref{sec:knap}), and examples of the shortcut
@@
 The final case study we consider is shortcut deforestation on lists.  Shortcut
-deforestation can be expressed succintly in terms of hylomorphisms and their
+deforestation can be expressed succinctly in terms of hylomorphisms and their
 laws~\cite{TakanoM95}. In particular, given a function:
@@
 % \label{sec:calccoq}
-Within the domain of program optimization, program calculation serves as a
+Within the domain of program optimisation, program calculation serves as a
 well-established programming technique aimed at deriving efficient programs from
@@
 non-structurally recursive since it involves recursion based on the result of a
-subtraction. There is a nonstandard implementation of divivion found on Coq's
+subtraction. There is a nonstandard implementation of division found on Coq's
 standard library, which involves a four-argument function that effectively
@@
 with nonstructural definitions. There is a major downside, however; as noted by
-Abreu et al., the result is ``barely recognizable as a form of
+Abreu et al., the result is ``barely recognisable as a form of
 mergesort''~\cite{AbreuDHJMS23}.

### 2025-03-21 paper/paper.tex
Typos
Introduction is done.

@@
 
-This program calculation stems from the more general theory of hylomorphisms.  A
-hylomorphism is a structured recursion scheme representing the idea of a
-divide-and-conquer algorithm where, first, the problem is split into smaller
-sub-problems (divide), then sub-solutions are computed recursively, and, finally,
-these sub-solutions are put together to compute the solution to the original
-problem (conquer).  This general pattern of recursion induces a reasoning
-principle, called \emph{fusion law}, which has been exemplified in the previous
-paragraph.
+This program calculation stems from the more general theory of hylomorphisms.
+A hylomorphism is a structured recursion scheme representing the idea of a divide-and-conquer algorithm where, first, the problem is split into smaller sub-problems (divide), then sub-solutions are computed recursively, and, finally, these sub-solutions are put together to compute the solution to the original problem (conquer).
+This general pattern of recursion induces a reasoning principle, called the \emph{fusion law}, which has been exemplified in the previous paragraph.
+By leveraging this principle, we can systematically transform recursive definitions into more efficient versions while preserving correctness.
+% This program calculation stems from the more general theory of hylomorphisms.  A
+% hylomorphism is a structured recursion scheme representing the idea of a
+% divide-and-conquer algorithm where, first, the problem is split into smaller
+% sub-problems (divide), then sub-solutions are computed recursively, and, finally,
+% these sub-solutions are put together to compute the solution to the original
+% problem (conquer).  This general pattern of recursion induces a reasoning
+% principle, called \emph{fusion law}, which has been exemplified in the previous
+% paragraph.
 
@@
 smaller parts (not necessarily following the structure of the input data).
+On the other hand, a key benefit is the reusability of termination proofs: once
+the correctness of the divide phase is established, the same proof can be reused
+for any other conquer phase we choose. Moreover, in the case of structural
+recursion, the divide phase naturally follows the structure of the data, making
+it correct by definition and eliminating the need for additional proofs.
 
@@
 
-The contributions of this paper are as follows. This work presents a Coq
-formalisation of \emph{hylomorphisms} that (1) is \emph{fully axiom-free}; (2)
-allows the extraction of idiomatic code; and (3) can use regular Coq
-equalities to do program calculation, derive correct implementations, and apply
-optimisations.  The full mechanisation is open source.
-%can be found on
+We list the contributions of this paper.
+First, in Section~\ref{sec:category-theory}, we provide the theoretical background on recursion schemes and hylomorphisms.
+We then provide a Coq formalisation of \emph{hylomorphisms} (Section~\ref{sec:recursion-schemes}) that (1) is \emph{fully axiom-free}; (2) allows the extraction of idiomatic and efficient code; and (3) can use regular Coq equalities to do program calculation, derive correct implementations, and apply optimisations.
+In Section~\ref{sec:extraction}, we apply this framework to formalise practical examples of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
+Additionally, we verify the short-cut fusion optimisation and present the extracted optimised code.
+
 %Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}\dcas{I SHOULD
 %  ANONYMISE THIS}.
-While programmers still need to reason about the termination of their programs,
-the use \emph{program calculation} can ease this task by allowing the use of
-fusion on programs that have already been proven to terminate.
-
-The remainder of the paper is structured as follows:
-\begin{itemize}
-  \item In Section~\ref{sec:category-theory} we give an overview about recursion
-schemes.
-  \item In Section~\ref{sec:containers} we formalise the type of container
-    functors, ensuring the presence of least and greatest fixed-points for
-    functors, and suitably adapted for program extraction.
-  \item In Section~\ref{sec:recursion-schemes} we mechanise folds, unfolds, and
-    hylomorphisms, as well as proving their uniqueness properties in Coq.
-  \item In Section~\ref{sec:examples} we use the framework to formalise examples
-    of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
-    Furthermore we verify the short-cut fusion optimisation and show the
-    extracted optimised code.
-\end{itemize}
 
@@
 \end{coqcode}
-Hylomorphisms fusion falls out from this uniqueness property:
+Fusing a hylomorphism with any algebra or coalgebra homomorphism (in the code
+below, \coq{f2} and \coq{f1} respectively) falls out from this uniqueness
+property:
 \begin{coqcode}
@@
 \end{coqcode}
-We will see in Section~\ref{sec:examples} one example where we will prove that
+We will see in Section~\ref{sec:extraction} one example where we will prove that
 the output of a sorting algorithm is a sorted permutation of the input, and
@@
 \section{Extraction}
-\label{sec:examples}
+\label{sec:extraction}
 We go in this section through a series of case studies of various recursive
@@
 \section{Conclusions and Future Work}
-
-Hylomorphisms are a general recursion scheme that can encode any other
-recursion scheme, and that satisfy a number of algebraic laws that can be used
-to reason about program equivalences. To our knowledge, this is their first formalisation
-in Coq. This is partly due to the
-difficulty of dealing with termination, and reasoning about functional
-extensionality. In this work, we tackle these problems in a \emph{fully
-axiom-free} way that targets the extraction of idiomatic code.  This
-formalisation allows the use of program calculation techniques in Coq to derive
+In this work we mechanise hylomorphisms and their algebraic laws and use them
+to perform optimisations for programs than can be later extracted into human-readable and runnable code.
+We do this in a \emph{fully
+  axiom-free}.
+
+This formalisation allows the use of program calculation techniques in Coq to derive
 formally optimised implementations from naive specifications. Furthermore, the
@@
 
-Remaining axiom-free forces us to deal with the well-known
-\emph{setoid hell}. As part of the future improvements, we will study how to
-mitigate this problem. At the moment, we use a short ad-hoc tactic that is able
-to automatically discharge many of these proofs in simple settings.  We will
-study the more thorough and systematic use of proof automation for respectful
+As part of the future improvements, we will study how to mitigate the problem of
+setoids. At the moment, we use a short ad-hoc tactic that is able to
+automatically discharge many of these proofs in simple settings.  We will study
+the more thorough and systematic use of proof automation for respectful
 morphisms. Generalised rewriting in proofs involving setoids tends to be quite

### 2025-03-21 paper/paper.tex
Intro redone.... again

@@
 \label{sec:intro}
-Structured recursion schemes are powerful abstractions that capture common
-patterns of recursion. The main benefit of structuring computation using
-recursion schemes, is that they enjoy well-established \emph{algebraic
-properties} that can serve as a foundation for reasoning about program
-equivalences, transformations, and optimisations (e.g.\ \emph{fusion} laws, or
-semi-automatic
-parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
-%
-These algebraic properties are widely used in the context of \emph{program
-calculation}, where programmers describe their code using simple, inefficient
-specifications in an \emph{algebra of programming}~\cite{BirddeMoor96:Algebra},
-and then use the algebraic laws of this algebra of programming to
-\emph{calculate} an efficient version of the same algorithm.
-%
+Over the years, extensive research has been conducted on program calculation
+techniques, a well-established approach for deriving efficient programs from
+simpler specifications through systematic program
+transformation~\cite{gibbonsSquiggol, BirddeMoor96:Algebra}. A key aspect of
+this approach is the use of structured recursion schemes, which serve as
+powerful abstractions that capture common patterns of recursion. By structuring
+computation in this way, programs benefit from well-established \emph{algebraic
+properties} that provide a solid foundation for reasoning about program
+equivalences, transformations, and optimisations -- such as \emph{fusion} laws or
+semi-automatic parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}.
+In the context of \emph{program calculation}, these algebraic properties allow
+programmers to describe code using simple, inefficient specifications within an
+\emph{algebra of programming}~\cite{BirddeMoor96:Algebra}. They can then apply
+algebraic laws to systematically \emph{calculate} more efficient versions of the
+same algorithms.
+
 Suppose, for example, that we want to write a program that sorts a list of
@@
 
-This program calculation stems from the more general theory of hylomorphisms.
-A hylomorphism is a structured recursion scheme representing the idea of a divide-and-conquer algorithm where, first, the problem is split into smaller sub-problems (divide), then sub-solutions are computed recursively, and, finally, these sub-solutions are put together to compute the solution to the original problem (conquer).
-This general pattern of recursion induces a reasoning principle, called the \emph{fusion law}, which has been exemplified in the previous paragraph.
-By leveraging this principle, we can systematically transform recursive definitions into more efficient versions while preserving correctness.
+This program calculation stems from the more general theory of hylomorphisms. A
+hylomorphism is a structured recursion scheme representing the idea of a
+divide-and-conquer algorithm where, first, the problem is split into smaller
+sub-problems (divide), then sub-solutions are computed recursively, and,
+finally, these sub-solutions are put together to compute the solution to the
+original problem (conquer). This general pattern of recursion induces a
+reasoning principle, called the \emph{fusion law}, which has been exemplified in
+the previous paragraph. By leveraging this principle, we can systematically
+transform recursive definitions into more efficient versions while preserving
+correctness.
 % This program calculation stems from the more general theory of hylomorphisms.  A
@@
 $\cons : A \times \List{A} \to \List{A}$. The $\List{-}$ type is a
-\emph{functor} $\Set \to \Set$ mapping objects and arrows in $\Set$ to $\Set$
-itself (such functors from a category to itself are called \emph{endofunctors}).
+\emph{functor} $\Set \to \Set$, in particular, an \emph{endofunctor}, mapping objects and arrows in $\Set$ to $\Set$
+itself.
 Its functorial action $\List{f} : \List{A} \to \List{B}$ is given by
@@
 \label{sec:extraction}
-We go in this section through a series of case studies of various recursive
-algorithms.  We show how they can be encoded in our framework, how can we do
-program calculation techniques for optimisation, and how can they be extracted
-to idiomatic functional code. Although our examples use OCaml as target
-language, extraction works equally well to other target languages. Our examples
-are the Quicksort and Mergesort algorithms (Section~\ref{sec:sorting}), dynamic
-programming and Knapsack (Section~\ref{sec:knap}), and examples of the shortcut
-deforestation optimisation in our framework (Section~\ref{sec:shortcut}).
+In this section we show how programs and their optimisations can be encoded in
+our framework and how can they be extracted to idiomatic functional code.
+Although our examples use OCaml as target language, extraction works equally
+well to other target languages (e.g. Haskell or Lisp). In particular, we show how to formalise
+ quicksort (Section~\ref{sec:sorting}), we give an example of dynamic
+programming algorithm, that is knapsack (Section~\ref{sec:knap}), and show examples of the shortcut
+deforestation optimisation (Section~\ref{sec:shortcut}).
 %The steps are the following:
@@
 %\end{itemize}
-We compare our work with other
-encodings of program calculation techniques and recursion schemes in Coq, and
-discuss other approaches to fusion and to termination of nonstructural
-recursion.
 % and recursion schemes in Coq
@@
 % \label{sec:calccoq}
-Within the domain of program optimisation, program calculation serves as a
-well-established programming technique aimed at deriving efficient programs from
-their naive counterparts through systematic program
-transformation~\cite{gibbonsSquiggol}. Programmers may specify algorithms
-abstractly at a high level, and then derive an efficient implementation using
-the process of program calculation~\cite{BirddeMoor96:Algebra}. This area has
-been extensively explored over the years.
 
@@
 % \label{sec:fusion}
-Fusion in functional programming is a general technique for combining multiple
+Fusion in functional programming is a more general technique for combining multiple
 computations into a single pass to eliminate intermediate data structures or

### 2025-03-21 paper/paper.tex
Section 4, minor

@@
 \subsection{Sorting Algorithms}\label{sec:sorting}
-Our first case study is divide-and-conquer sorting algorithms encoded as
-recursive hylomorphisms. Our accompanying code contains both mergesort and
-quicksort as examples, but we will focus on quicksort for conciseness. We will
-also show how to use program calculation to apply a fusion optimisation.  The
-recursive structure of both mergesort and quicksort is a binary tree. For
-example, in the case of quicksort, a list is split into a pivot, the label of
-the node, and two sublists. We define the data functor of trees as follows:
-\begin{coqcode}
-Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
-\end{coqcode}
+We now focus on divide-and-conquer sorting algorithms and demonstrate how to use
+program calculation to apply a fusion optimisation. Our accompanying code
+contains both mergesort and quicksort as examples, but we will focus on
+quicksort for conciseness.
+
+The recursive structure of quicksort and mergesort is given by the following
+functor
+\begin{coqcode}
+Inductive ITreeF N X := i_leaf | i_node (n : N) (l r : X)
+\end{coqcode}
+To explain this consider quicksort.
+The list is split into a pivot which is represented by the and two sublists. We define the data functor of trees as follows:
+
 To define an equivalent container, we notice that this functor has two shapes,
@@
 \end{coqcode}
-We name \coq{TreeF} the container that has these shapes and (base) poisitions.
+We name \coq{TreeF} the container that has these shapes and (base) positions.
 For convenience, we define a series of wrappers to construct these containers:

### 2025-03-21 paper/paper.tex
Rewrites to conclusions and related work

@@
 
-Tesson et al.\ demonstrated the efficacy of leveraging Coq to establish an
-approach for implementing a robust system dedicated to verifying the correctness
-of program transformations for functions that manipulate
+There is existing prior work on formalising recursion schemes in a proof
+assistant.  Tesson et al.\ demonstrated the efficacy of leveraging Coq to
+establish an approach for implementing a robust system dedicated to verifying
+the correctness of program transformations for functions that manipulate
 lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
@@
 extensionality axiom, as well as further extensionality axioms for each
-coinductive datatype that they use. They do not discuss the extracted code
-from their formalisation. Larchey-Wendling and Monin encode recursion schemes in
-Coq, by formalising computational graphs of algorithms~\cite{larchey2022braga}.
-Their work does not focus on encoding generic recursion schemes, and proving
-their algebraic laws. Castro-Perez et al.~\cite{farmsCastro} encode the laws of
-hylomorphisms as part of the type system of a functional language to calculate
-parallel programs from specifications. Their work focuses on parallelism, and
-they do not formalise their approach in a proof assistant, and the laws of
-hylomorphisms are axioms in their system.
-
-% \textbf{Fusion Optimisation}.
-% %
-% \label{sec:fusion}
-Fusion in functional programming is a more general technique for combining multiple
-computations into a single pass to eliminate intermediate data structures or
-repeated traversals, improving efficiency. The fusion we have discussed in this
-paper relating to program calculation is one form of fusion, but there are other
-approaches, and prior research has explored the mechanisation and verification
-of fusion transformations.
-
-Fusion has been applied to recursive tree traversals, in order to reduce the
-number of times each section of the tree must be visited. Research has
-demonstrated that this sort of fusion can be performed automatically by a
-compiler~\cite{10.1145/3133900}, and the correctness of tree traversal fusion
-has been verified mechanically in Coq~\cite{10.1145/3377555.3377884}.
-
-Another related family transformations are worker/wrapper
-transformations~\cite{gill09}, which have been used in the context of optimising
-compilers, and have been mechanised in Agda~\cite{SCULTHORPE_HUTTON_2014}. This
-technique has been applied to improving the performance of programs defined
-using folds via fusion~\cite{hutton10}. Their work does not focus on the use of
-an interactive proof assistant, but rather on the manual use of equational
-reasoning to re-write programs and on the automatic transformation of functional
-programs by a compiler.
+coinductive datatype that they use. They do not discuss the extracted code from
+their formalisation. Larchey-Wendling and Monin encode recursion schemes in Coq,
+by formalising computational graphs of algorithms~\cite{larchey2022braga}.
+Their work does not focus on encoding higher-order generic recursion schemes,
+and proving their algebraic laws. Castro-Perez et al.~\cite{farmsCastro} encode
+the laws of hylomorphisms as part of a type system to calculate parallel
+programs from specifications. Their work focuses on parallelism, and they do not
+formalise their approach in a proof assistant, and they treat the  laws of
+hylomorphisms as axioms.
 
@@
 Through experiments, we found that code extracted from their formalisation makes
-heavy use of unsafe casts and has a generally complex structure that may be
-hard to understand and integrate with other code. Due to the great benefit
+heavy use of unsafe casts and has a generally complex structure that may be hard
+to understand and integrate with other external code. Due to the great benefit
 of entirely avoiding termination proofs, it would be interesting to extend their
 approach to improve code extraction.
+
+The problem of \emph{nonstructural recursion} (including
+divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
+% A distinction between standard and
+% nonstandard recursion lies in the fact that standard recursion adheres strictly
+% to recursing on pattern variables, avoiding reliance on results from other
+% function calls. Nonstandard recursion may deviate from this convention, allowing
+% for more flexibility.
+Certain functions that are not structurally recursive can be reformulated using
+a nonstandard approach to achieve structural recursion.  For exammple, the
+mergesort in Coq's standard library uses an ``explicit stack of pending merges''
+in order to avoid issues with nonstructural definitions. There is a major
+downside, however; as noted by Abreu et al., the result is ``barely recognisable
+as a form of mergesort''~\cite{AbreuDHJMS23}.  There are common approaches in
+Coq to deal with termination~\cite{DBLP:journals/pacmpl/SozeauM19,
+  DBLP:conf/icfp/Sozeau07}, but none of these approaches address program
+calculation techniques, and the mechanisation of fusion laws.
+
+% \textbf{Fusion Optimisation}.
+% %
+% \label{sec:fusion}
+The fusion laws that we formalise in this paper are applied in the more general
+context of compilers for functional programming lanmguages.  Fusion has been
+applied to recursive tree traversals, in order to reduce the number of times
+each section of the tree must be visited. Research has demonstrated that this
+sort of fusion can be performed automatically by a
+compiler~\cite{10.1145/3133900}. The correctness of tree traversal fusion has
+been verified mechanically in Coq~\cite{10.1145/3377555.3377884}, but the
+authors do not formalise unfolds and hylomorphisms.
+
+Another related family transformations are worker/wrapper
+transformations~\cite{gill09}, which have been used in the context of optimising
+compilers, and have been mechanised in Agda~\cite{SCULTHORPE_HUTTON_2014}. This
+technique has been applied to improving the performance of programs defined
+using folds via fusion~\cite{hutton10}. Their work does not focus on the use of
+an interactive proof assistant, but rather on the manual use of equational
+reasoning to re-write programs and on the automatic transformation of functional
+programs by a compiler.
 
@@
 %In particular,
-The problem of \emph{nonstructural recursion} (including
-divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
-% A distinction between standard and
-% nonstandard recursion lies in the fact that standard recursion adheres strictly
-% to recursing on pattern variables, avoiding reliance on results from other
-% function calls. Nonstandard recursion may deviate from this convention, allowing
-% for more flexibility.
-Certain functions that are not structurally recursive can be reformulated using
-a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
-Take, for instance, division by iterated subtraction, which is inherently
-non-structurally recursive since it involves recursion based on the result of a
-subtraction. There is a nonstandard implementation of division found on Coq's
-standard library, which involves a four-argument function that effectively
-combines subtraction and division. Similarly, the mergesort in Coq's standard
-library uses an ``explicit stack of pending merges'' in order to avoid issues
-with nonstructural definitions. There is a major downside, however; as noted by
-Abreu et al., the result is ``barely recognisable as a form of
-mergesort''~\cite{AbreuDHJMS23}.
-There are common approaches in Coq to deal with termination, but none of these
-approaches address program calculation techniques, and the mechanisation of
-fusion laws.
 
@@
 \section{Conclusions and Future Work}
-In this work we mechanise hylomorphisms and their algebraic laws and use them
-to perform optimisations for programs than can be later extracted into human-readable and runnable code.
-We do this in a \emph{fully
-  axiom-free}.
-
-This formalisation allows the use of program calculation techniques in Coq to derive
-formally optimised implementations from naive specifications. Furthermore, the
-rewritings that are applied to specifications are formal, machine-checked
-proofs that the resulting program is extensionally equal to the input
-specification.
+In this work we mechanise hylomorphisms and their algebraic laws and use them to
+perform program calculation and optimisations. The resulting programs can be
+later extracted into idiomatic and runnable code. Our mechanisation is
+\emph{fully axiom-free}. This formalisation allows us to use Coq as a framework
+for program calculation in a way that is close to common practice in pen and
+paper program calculation proofs. This implies that every rewriting step is the
+result of applying a formal, machine-checked proof that the resulting program is
+extensionally equal to the input specification.
 
@@
 automatically discharge many of these proofs in simple settings.  We will study
-the more thorough and systematic use of proof automation for respectful
-morphisms. Generalised rewriting in proofs involving setoids tends to be quite
-slow, due to the large size of the terms that need to be rewritten. Sometimes,
-this size is hidden in implicit arguments and coercions. We will study
-alternative formulations to try to improve the performance of the rewriting
-tactics (e.g.\ canonical structures).  Currently, Coq is unable to inline a
-number of trivially inlineable definitions.  We will study alternative
-definitions, or extensions to Coq's code extraction mechanisms to force the
-full inlining of all container code that is used in hylomorphisms.  Finally,
-proving termination still remains a hurdle. In our framework this reduces to
-proving that the anamorphism
-terminates in all inputs, and we provide a convenient connection to
+the more thorough and systematic use of proof automation for proper morphisms.
+Generalised rewriting in proofs involving setoids tends to be quite slow, due to
+the large size of the terms that need to be rewritten. Sometimes, this size is
+hidden in implicit arguments and coercions. We will study alternative
+formulations to try to improve the performance of the rewriting tactics
+(e.g.\ canonical structures).  Currently, Coq is unable to inline a number of
+trivially inlineable definitions.  We will study alternative definitions, or
+extensions to Coq's code extraction mechanisms to force the full inlining of all
+container code that is used in hylomorphisms.  Finally, proving termination
+still remains a hurdle. In our framework this reduces to proving that the
+anamorphism terminates in all inputs, and we provide a convenient connection to
 well-founded recursion. Furthermore, recursive coalgebras compose with natural
-transformations, which allows the reuse
-of a number of core recursive coalgebras. A possible interesting future line of work
-is the use of the approach by Abreu et al.~\cite{AbreuDHJMS23} in combination with ours to
-improve code extraction from divide-and-conquer computations whose termination
-does not require an external proof.
+transformations, which allows the reuse of a number of core recursive
+coalgebras. A possible interesting future line of work is the use of the
+approach by Abreu et al.~\cite{AbreuDHJMS23} in combination with ours to improve
+code extraction from divide-and-conquer computations whose termination does not
+require an external proof.
 

### 2025-03-21 paper/paper.tex
Section 4.1 was a mess

--- 2025-03-21 paper/paper.tex
@@
 
-The recursive structure of quicksort and mergesort is given by the following
+The recursive structure of quicksort is given by the following
 functor
@@
 \end{coqcode}
-To explain this consider quicksort.
-The list is split into a pivot which is represented by the and two sublists. We define the data functor of trees as follows:
-
-To define an equivalent container, we notice that this functor has two shapes,
-one for leaves, and one for nodes, and that nodes have two positions (akin to
-\coq{l r : X} in \coq{i_node}). We also note that, in this container, positions
-are only valid in nodes.
+The idea is that a list is either empty or is split into a pivot which is
+represented by node of type \coq{N} and the two sublists of type \coq{X}.
+
+The container encoding \coq{ITreeF} has two shapes, one for leaves and one for
+nodes, and nodes have two positions, one for each sublist \coq{l r : X} in
+\coq{i_node}.
 \begin{coqcode}
@@
 \end{coqcode}
-We name \coq{TreeF} the container that has these shapes and (base) positions.
-For convenience, we define a series of wrappers to construct these containers:
-% \mpav{The following algebra should be called a\_in or l\_in whatever}
-\begin{coqcode}
-Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
-Notation a_node x l r := (MkCont (Node _ x)
-  (fun p => match val p with | Lbranch => l | Rbranch => r end)).
-\end{coqcode}
-The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with the
-following algebra and coalgebra. In the below definition, we use \coq{posL} and
-\coq{posR} as wrappers for \coq{Lbranch} and \coq{Rbranch} with validity proofs.
-\begin{coqcode}
+The container for \coq{ITreeF} is denoted by \coq{TreeF}. For convenience, we
+define a series of wrappers \coq{a_leaf} and \coq{a_node} to construct these
+containers.  Now, the container for quicksort is given by \coq{TreeF unit int}.
+In the definition below, we use \coq{posL} and \coq{posR} as wrappers for
+\coq{Lbranch} and \coq{Rbranch} with validity proofs.
+\begin{coqcode}
+Definition qsplit_f (x : list int) : App (TreeF unit int) (list int) :=
+  match x with
+  | nil => a_leaf tt
+  | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in a_node h l r
+  end.
+
 Definition merge_f (x : App (TreeF unit int) (list int)) : list int :=
@@
   end kx.
-Definition qsplit_f (x : list int) : App (TreeF unit int) (list int) :=
-  match x with
-  | nil => a_leaf tt
-  | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in a_node h l r
-  end.
-\end{coqcode}
-Our tactics automatically discharge the proof that these are proper morphisms
-(called \coq{merge} and \coq{qsplit}), and we prove that the coalgebra
-\coq{qsplit} is recursive by showing that it respects the ``less-than'' relation
-on the length of the lists.
-%\begin{coqcode}
-%Lemma split_fin : respects_relation split (@length int) lt.  (* *) Qed.
-%
-%Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
-%  := mk_wf_coalg wf_lt split_fin.
-%\end{coqcode}
-The code that we extract for \coq{hylo merge qsplit} is similar to a hand-written implementation:
+\end{coqcode}
+The proofs that these are proper morphisms (called \coq{merge} and \coq{qsplit})
+are automatically discharged by our tactics, but we still have to prove that the
+coalgebra \coq{qsplit} is recursive. In this specific case we have to prove that
+the two sublists are smaller than the input list. In general, we are required to
+prove that the input is related to the data inside the structure functor, that
+is that the element of the carrier of the coalgebra is related to the content of
+the functor by some total relation. Here we give a specialised property, the
+full general property can be found in the formalisation:
+\begin{coqcode}
+Definition respects_relation `{setoid A} (c : Coalg F B) B} (R : B -> B -> Prop)
+:= forall x (p : Pos (shape (c x))), R (length (cont (c x) p)) (length x).
+\end{coqcode}
+
+The resulting code extracted from \coq{hylo merge qsplit} is similar to a
+hand-written implementation:
 \begin{ocamlcode}
@@
 \end{ocamlcode}
-
-\paragraph*{Proving Quicksort Correct}
-
-We show that the output of \coq{hylo merge qsplit} produces a sorted
-permutation of the input:
+The correctness proof is done by proving the following statement:
 \begin{coqcode}
@@
 \end{coqcode}
-The proof in our framework starts by induction on the proof that \coq{qsplit}
-is recursive, and that it terminates on \coq{l}, i.e. using tactic
-\coq{induction (recP qsplit l)}. Then, we can unfold
-\coq{hylo merge qsplit} to \coq{merge \o fmap (hylo merge qsplit) \o qsplit} using
-hylomorphism uniqueness, and the rest of the proof is standard. Overall, our
-proof is shorter than alternative implementations, and of comparable
-complexity\footnote{The QSort definition and correctness proof in the equations
+that is any list produced by quicksort is sorted and is some permutation of the original one.
+
+The proof is by induction on the fact that \coq{qsplit} is recursive. This means
+that \coq{qsplit} terminates on \coq{l}.  In practice this is done by using the
+tactic \coq{induction (recP qsplit l)}. Then we can unfold \coq{hylo merge qsplit} to
+\coq{merge \o fmap (hylo merge qsplit) \o qsplit} using hylomorphism
+uniqueness~\ref{sec:coq-rec-coalgebras}. The rest of the proof is standard.
+Overall, our proof is shorter than alternative implementations, and of
+comparable complexity\footnote{The QSort definition and correctness proof in the
+equations
 package is under 200LOC --
@@
 % \end{ocamlcode}
+
 \subsubsection{Fusing a divide-and-conquer computation}
-As an example of how can we use program calculation techniques in our 
-framework, we show how another traversal can be fused into the
-divide-and-conquer algorithm using the laws of hylomorphisms.
-Suppose that we map a function to the result of sorting the list. We can
-use our framework to fuse both computations. In particular, consider
-the following definition:
+We now show how we can use program calculation techniques to fuse a traversal
+with a  divide-and-conquer algorithm to obtain a new program that only performs
+recursion once instead of twice. In particular, we can map a function to the
+result of quicksort and obtain a new program which orders the elements and
+computes the function on every element by traversing the list only once. The
+composition of these two programs is given by the following definition:
 \begin{coqcode}
@@
 \end{coqcode}
-Here, \coq{Lmap times_two} is a list map function defined as a hylomorphism,
+where \coq{Lmap times_two} is a list map function defined as a hylomorphism,
 and \coq{times_two} multiplies every element of the list by two.
 We can use Coq's generalised rewriting and \coq{hylo_fusion_l} to fuse
-\coq{times_two} into \coq{hylo merge qsplit}.
-After the necessary rewrites, we obtain
+\coq{times_two} into \coq{hylo merge qsplit} which gives the program
 \coq{hylo (merge \o natural times_two) qsplit}.
 In this definition, \coq{natural} defines a natural transformation by
-applying \coq{times_two} to the shapes, thus multiplying every pivot by two.
+applying \coq{times_two} to the shapes thus multiplying every pivot by two.
 %% Our formalisation contains a proof that \coq{natural} is indeed a natural
@@
 %% shapes and, therefore, the validity of the positions.
-The extracted OCaml code
-is a single recursive traversal:
+The extracted OCaml code is a single recursive traversal:
 \begin{ocamlcode}
@@
 \end{ocamlcode}
-
+\todo{Write something here, don't leave it empty}{}
 \subsection{Knapsack}\label{sec:knap}

### 2025-03-21 paper/paper.tex
Added hylos in agda
Consistency changes

@@
 \begin{coqcode}
-Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
+Inductive Tshape L A := | Leaf | Node (ELEM : A).
 Inductive Tpos := | Lbranch | Rbranch. 
@@
 define a series of wrappers \coq{a_leaf} and \coq{a_node} to construct these
-containers.  Now, the container for quicksort is given by \coq{TreeF unit int}.
+containers.  Now, the container for quicksort is given by \coq{TreeF int}.
 In the definition below, we use \coq{posL} and \coq{posR} as wrappers for
@@
 \begin{coqcode}
-Definition qsplit_f (x : list int) : App (TreeF unit int) (list int) :=
+Definition qsplit_f (x : list int) : App (TreeF int) (list int) :=
   match x with
@@
 
-Definition merge_f (x : App (TreeF unit int) (list int)) : list int :=
+Definition merge_f (x : App (TreeF int) (list int)) : list int :=
   let (sx, kx) := x in
@@
 coinductive datatype that they use. They do not discuss the extracted code from
-their formalisation. Larchey-Wendling and Monin encode recursion schemes in Coq,
+their formalisation.  Mu et al.\ formalise hylomorphisms in Agda, and can do
+relational program transformation~\cite{DBLP:journals/jfp/MuKJ09}. It is unclear
+how to extract runnable code from their definitions, and they do not seem to
+formalise hylomorphisms in terms of generic functors and datatypes.
+
+Larchey-Wendling and Monin encode recursion schemes in Coq,
 by formalising computational graphs of algorithms~\cite{larchey2022braga}.

### 2025-03-21 paper/paper.tex
Containers need more work!

--- 2025-03-21 paper/paper.tex
@@
 objects are $F$-algebras and arrows $f : (X, a_{X}) \to (Y, a_{Y})$ are
-\emph{$F$-algebra homomorphisms}  $f : X \to Y$ in $\CatC$ such that they
+\emph{$F$-algebra homomorphisms}. These are  arrows $f : X \to Y$ in the underlying category $\CatC$ such that they
 respect the structure of the algebra, that is
@@
 unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
-the initial $F$-algebra is precisely the \emph{least fixed-point} for the
-functor $F$ which we denote by $(\mu F, \InIso)$ with
-$\InIso : F \mu F \to \mu F$ and $\InOp : \mu F \to F \mu F$ witnessing the
-isomorphism $F\mu F \iso \mu F$.  By the property of initial $F$-algebras, for
-any other $F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra
-homomorphism, denoted by $\cata{\beta} : (\mu F, \InIso) \to (X, a_{X})$ and
-pronounced ``\emph{catamorphism}'' satisfying:
+if $F$ has an initial $F$-algebra then this is the \emph{least fixed-point} for the
+functor $F$ which we denote by $(\mu F, \InIso)$ where
+$\InIso : F \mu F \to \mu F$ is the $F$-algebra  witnessing the
+isomorphism $F\mu F \iso \mu F$, furthermore  $\InOp : \mu F \to F \mu F$ is the
+inverse of $\InIso$. The uniqueness property of initial $F$-algebras states that for any other
+$F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra homomorphism, denoted
+by $\cata{\beta} : (\mu F, \InIso) \to (X, a_{X})$ and pronounced
+``\emph{catamorphism}'' or ``\emph{fold}'' satisfying:
 \begin{equation}
@@
 \end{equation}
-In words, any $F$-algebra homomorphism from the initial $F$-algebra is a
-catamorphism. As a result of the uniqueness property we can derive the fusion
+As a result of the uniqueness property we can derive the fusion
 law. For all $F$-algebra homomorphisms $f : (X, a_{X}) \to (Y, a_{Y})$ we have
@@
 structure map $[b, i] : 1 + A \times B \to B$. The initial $F$-algebra is
-clearly the set of lists $\List{A}$. The catamorphism associated with the type
-of lists is the unique arrow recursively translating the initial algebra
-$[\nil,\cons]$ into the algebra $[b, i]$.
-In functional programming this is commonly referred to as
-$\foldr : (1 \to B) \to (A \times B \to B) \to B$. We can in fact set
-$\foldr\;base\;ind = \cata{[base, ind]}$.
+clearly the set of lists $\List{A}$ and the catamorphism associated with the
+type of lists is the unique arrow which recursively translates the initial
+algebra $[\nil,\cons]$ into the algebra $[base, step]$ while turning the operation
+$\nil$ into $base$ and $\cons$ into $step$.  In functional programming this is
+commonly referred to as $\foldr : (1 \to B) \to (A \times B \to B) \to \List{A} \to B$. We
+can in fact set $\foldr\;base\;step = \cata{[base, step]}$.
 
@@
 $B$-coalgebra homomorphisms $f : A \to B$, that is
-$ c_{Y} \circ f = F(f) \circ c_{X}$. Terminal object in this category
-$B$-coalgebra corresponds to the greatest fixed-point for the functor $B$,
-denoted by $(\nu B, \OutIso)$ where $\OutIso$ denotes the final $B$-coalgebra
-$\OutIso : \nu B \to B\nu B$.
+$ c_{Y} \circ f = F(f) \circ c_{X}$. The terminal object in this category is
+called the terminal, or final, $B$-coalgebra. The carrier of this coalgebra
+corresponds to the greatest fixed-point for the functor $B$, denoted by
+$(\nu B, \OutIso)$ with $\OutIso$ being the final $B$-coalgebra witnessing the
+isomorphism and $\OutIso : \nu B \to B\nu B$ being its inverse.
 
 For example, the terminal coalgebra for the functor $BX = L \times X$ is the set
-of of infinite streams, that is the greatest solution to the domain equation
-$\Str{A} \iso A \times \Str{A}$.  As a result, for any $B$-coalgebra
+of infinite streams, that is the greatest solution to the domain equation
+$\Str{A} \iso A \times \Str{A}$.  The uniqueness property of terminal $B$-coalgebras states that for any other $B$-coalgebra
 $(X, c_{X})$ there exists a unique $B$-coalgebra homomorphism into the terminal
 coalgebra $(\nu B, \OutIso)$ which is denoted by $\ana{c_{X}}$ and pronounced
-``\emph{anamorphism}''. We spell out the uniqueness property:
+``\emph{anamorphism}'' or ``\emph{unfold}''. We spell out the uniqueness property:
 \begin{equation}
@@
 \end{equation}
-In words, for any $B$-coalgebra, if there is any other $B$-coalgebra
-homomorphism $f$ into the terminal object then it must be the anamorphism on the
-same coalgebra. As a result of the uniqueness property we can derive the fusion
-law:\begin{equation}
+From the uniqueness property we can derive the fusion law for unfolds:
+\begin{equation}
   \label{eq:ana:fusion}
@@
 
+\subsection{Polynomial Functors}
+Recursion schemes are formulated using generic functors that encapsulate the
+structure of the recursive operator. However, since not all functors have the
+suitable fixed points, it is necessary to restrict our focus to \emph{strictly
+positive functors}.
+
+For example, in the category \textbf{Set}, consider the polynomial functor
+corresponding to
+\[
+F(X) = A + B \times X + C \times X^2.
+\]
+We can model this functor by using a set $S$ defined by $A \uplus B \uplus C$ which distinguishes the three
+cases, and a function $P$ which tells us the number of occurrences of $X$ in
+each case: for $s = (1,a) \mapsto \emptyset$; for $s = (2, b) \mapsto 1$ the singleton set and  for $s = (3, c) \mapsto 2$ the $2$-element set.
+
+To formalise this we need  a function $p : P \to S$ and a function
+$q : P \times X \to P$.  Here, $S$ serves as a \emph{shape object} that indexes
+the different cases of the functor, while $P$ represents the positions where
+elements of $X$ appear. The projection $p: P \to S$ assigns each position to its
+corresponding shape, and the map $q: P \to P \times X$ describes how these
+positions are instantiated using elements of $X$. The functor associated with
+this data is defined on objects using a dependent sum (to be precise a coend):
+% \[
+% F(X) = S \times_P (P \times X),
+% \]
+% which can also be written as a dependent sum (a coend to be precise)
+\[
+F(X) = \sum_{s \in S} (P(s) \to X)
+\]
+where  \( P(s) = p^{-1}(s) \) called the fiber of \( p \) over \( s \). Given a
+function \( f: X \to Y \), the functor acts by post-composition, mapping an
+element \( (s, p, x) \) in \( F(X) \) to \( (s, p, f(x)) \) in \( F(Y) \),
+ensuring functoriality.
+
+% First off, a Given a universe of types $\Type$, a container $S \triangleright P$
+% is defined by a type of \emph{shapes} $S : \Type$ and a \emph{family of position
+% types} indexed by shapes $P : S \to \Type$.  An \emph{extension} of this
+% container is a functor $\llbracket S \triangleright P \rrbracket$, whose action
+% on objects is given by
+% $\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
+% and on morphisms is given by post-composition, keeping the shape the same
+% $\llbracket S \triangleright P \rrbracket \; f = \lambda (s,\;g).\;(s,\;f\circ g)$.
+% Intuitively, the type of shapes captures the different ``cases'' of an algebraic
+% datatype, and the positions of a shape are used to identify the elements inside
+% a container of the given shape.
+
+% Any polynomial functor can be represented as a container (see accompanying
+% code).  For example, a container that is equivalent to $F\; X = 1 + A \times X$
+% requires a shape that can distinguish two cases, indicating that there are two
+% constructors to build an object of type $F\;X$, e.g. $S_{F} = 1 + A$. We need to
+% consider two cases for the family of positions:
+% $P_{F}(\mathsf{inj}_{1}\;*) = 0$, because there are no occurrences of $X$ in the
+% left case; and $P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence
+% of $X$ in the right hand side. It is easy to see that
+% $\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.
+
 \section{Formalising Recursion Schemes}
@@
 \label{sec:containers}
-Recursion schemes are formulated using generic functors that encapsulate the
-structure of the recursive operator. However, since not all functors have the
-suitable fixed points, it is necessary to restrict our focus to those mappings
-of the form \coq{F : Type -> Type} which are additionally \emph{strictly positive}.
-
-A common solution is to consider \emph{containers}~\cite{AbbottAG05}.  A
-container $S \triangleright P$ is defined by a type of \emph{shapes} $S : \Type$
-and a \emph{family of position types} indexed by shapes $P : S \to \Type$.  An
-\emph{extension} of this container is a functor
-$\llbracket S \triangleright P \rrbracket$, whose action on objects is given by
-$\llbracket S \triangleright P \rrbracket \; X = \Sigma s : S.\; P\;s \to X$,
-and on morphisms is given by post-composition, keeping the shape the same
-$\llbracket S \triangleright P \rrbracket \; f = \lambda (s,\;g).\;(s,\;f\circ g)$.
-Intuitively, the type of shapes captures the different ``cases'' of an algebraic
-datatype, and the positions of a shape are used to identify the elements inside
-a container of the given shape.
-
-Any polynomial functor can be represented as a container (see accompanying
-code).  For example, a container that is equivalent to $F\; X = 1 + A \times X$
-requires a shape that can distinguish two cases, indicating that there are two
-constructors to build an object of type $F\;X$, e.g. $S_{F} = 1 + A$. We need to
-consider two cases for the family of positions:
-$P_{F}(\mathsf{inj}_{1}\;*) = 0$, because there are no occurrences of $X$ in the
-left case; and $P_{F}(\mathsf{inj}_{2}\;a) = 1$, because there is one occurrence
-of $X$ in the right hand side. It is easy to see that
-$\llbracket S_{F} \triangleright P_{F} \rrbracket \cong F$.
-
+\todo{FILL OUT}{}
 The type $\forall s : S.\; P\;s \to X$ would force us to add
@@
 \todo{Write something here, don't leave it empty}{}
+
 \subsection{Knapsack}\label{sec:knap}

### 2025-03-21 paper/paper.tex
Fixes

@@
 
-\newmintinline[coq]{coq}{fontsize=\small}
+\newmintinline[coq]{coq}{fontsize=\small, breaklines}
 \newmintinline[ocaml]{OCaml}{fontsize=\small}
@@
 The container for \coq{ITreeF} is denoted by \coq{TreeF}. For convenience, we
-define a series of wrappers \coq{a_leaf} and \coq{a_node} to construct these
-containers.  Now, the container for quicksort is given by \coq{TreeF int}.
+define the wrappers \coq{a_leaf} and \coq{a_node} to construct values of the
+extension of this container.
+Now, the container for quicksort is given by \coq{TreeF int}.
 In the definition below, we use \coq{posL} and \coq{posR} as wrappers for
@@
 is that the element of the carrier of the coalgebra is related to the content of
-the functor by some total relation. Here we give a specialised property, the
+the functor by some well-founded relation. Here we give a specialised property, the
 full general property can be found in the formalisation:
@@
 \end{coqcode}
-
 The resulting code extracted from \coq{hylo merge qsplit} is similar to a
@@
 %% shapes and, therefore, the validity of the positions.
-The extracted OCaml code is a single recursive traversal:
 \begin{ocamlcode}
@@
 \end{ocamlcode}
-\todo{Write something here, don't leave it empty}{}
+The extracted OCaml code is a single recursive traversal. Note the similarity to
+a hand-written implementation, and that it has been derived by fusing two
+different recursive programs using regular Coq rewrite tactics.
 
@@
 This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
-\textbf{not} provable in Coq if we want to remain axiom-free, since we would
-need to add the necessary parametricity
+\textbf{not} provable in Coq without adding the parametricity
 axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.

### 2025-03-24 paper/paper.tex
edits for sections 1 and 2

@@
 
-Hylomorphisms are very powerful non-structural recursive algorithms that can be
-used to implement structural recursion, as well as many other known recursion
-schemes such as mutual recursion, accumulators, primitive recursion and
-course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. This power
-comes at a price. Since hylomorphisms are not necessarily structural, extra-care
-is needed when encoding them in a type theory. In particular, the divide part of
-the algorithm must be \emph{recursive} that is the input needs to be split into
-smaller parts (not necessarily following the structure of the input data).
-On the other hand, a key benefit is the reusability of termination proofs: once
-the correctness of the divide phase is established, the same proof can be reused
-for any other conquer phase we choose. Moreover, in the case of structural
+Hylomorphisms are highly versatile, non-structural recursive algorithms that are
+capable of implementing structural recursion as well as other known recursion
+schemes like mutual recursion, accumulators, primitive recursios, and
+course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. However, this
+versatility comes with a trade-off: because hylomorphisms are not inherently
+structurally recursive, special attention is required when encoding them in a
+type theory. Specifically, the ``divide'' phase of the algorithm must be
+recursive, meaning the input needs to be broken down into smaller parts, which
+do not necessarily align with the structure of the input data. On the other
+hand, a key benefit is the reusability of termination proofs: once the
+correctness of the divide phase is established, the same proof can be reused for
+any other conquer phase we choose. Moreover, in the case of structural
 recursion, the divide phase naturally follows the structure of the data, making
@@
 
-Implementations of recursion schemes currently focus on non-terminating
+Implementations of recursion schemes generally focus on non-terminating
 languages (e.g. Haskell) and they would benefit from a type-theoretical
@@
 
-In this work we mechanise hylomorphisms, thus reaping the benefits of their
-generality and their algebraic properties. We do this in a way which allows the
-user to write specifications of their code, reason about program optimisations,
-and eventually extract humanly readable functional code which can be run. Our
-extracted code does not contain unsafe casts. We focus strongly on avoiding
-unsafe casts because, even if the extracted code has been verified, simple
-interoperations with them can lead to incorrect behaviour or even segmentation
-faults~\cite{forster:hal-04329663} and, moreover, it invalidates the
-fast-and-loose principle~\cite{DanielssonHJG06}.
+In this work, we mechanise hylomorphisms, reaping the benefits of their
+generality and algebraic properties. Our approach enables users to write
+high-level specifications for their programs, reason about program
+optimisations, and ultimately extract human-readable, executable functional
+code. The extracted code is free from unsafe casts, a priority in our design. We
+focus strongly on avoiding unsafe casts because, even if the extracted code has
+been verified, simple interoperations with them can lead to incorrect behaviour
+or even segmentation faults~\cite{forster:hal-04329663} and, moreover, it
+invalidates the fast-and-loose principle~\cite{DanielssonHJG06}.
 
@@
 
-Furthermore folds and unfolds can be shown to capture a wide
-range of recursion schemes such as primitive recursion, mutual recursion,
-dynamic programming algorithms, polymorphic recursion, recursion with
-accumulators and so on. However, for divide-and-conquer algorithms folds need to
-be generalised to hylomorphisms which provide the ultimate basic building block
-for any other recursion scheme. This is done by looking at recursion schemes
-from the point of view of category theory.
+Furthermore, folds and unfolds can be shown to capture a wide range of recursion
+schemes such as primitive recursion, mutual recursion, dynamic programming
+algorithms, polymorphic recursion, and recursion with accumulators. However, for
+divide-and-conquer algorithms, folds need to be generalised to hylomorphisms,
+which provide the ultimate basic building block for any other recursion scheme.
+To do this, we look at recursion schemes from the point of view of category
+theory.
 

### 2025-03-24 paper/paper.tex
edits for section 4

@@
 Although our examples use OCaml as target language, extraction works equally
-well to other target languages (e.g. Haskell or Lisp). In particular, we show how to formalise
- quicksort (Section~\ref{sec:sorting}), we give an example of dynamic
-programming algorithm, that is knapsack (Section~\ref{sec:knap}), and show examples of the shortcut
-deforestation optimisation (Section~\ref{sec:shortcut}).
+well to other target languages (e.g. Haskell or Scheme). In particular, we show
+how to formalise sorting algorithms by looking at quicksort
+(Section~\ref{sec:sorting}), we show how to formalise dynamic programming
+algorithms by looking at knapsack (Section~\ref{sec:knap}), and, finally, we
+show an example of the shortcut deforestation optimisation
+(Section~\ref{sec:shortcut}).
 %The steps are the following:
@@
 We now focus on divide-and-conquer sorting algorithms and demonstrate how to use
-program calculation to apply a fusion optimisation. Our accompanying code
-contains both mergesort and quicksort as examples, but we will focus on
-quicksort for conciseness.
+program calculation to apply a fusion optimisation. In the supplemental material
+accompanying this paper, we formalise both mergesort and quicksort, but in this
+section, we focus solely on quicksort.
 
@@
 \end{coqcode}
-Coq's code extraction mechanism is unable to inline several definitions in this
-case. We have manually inlined the extracted code for simplicity. The reader
-can check in our artefact that the extracted code can be trivially
-inlined to produce the following:
+While the extracted code has the expected performance characteristics, Coq's
+code extraction mechanism is unable to inline several definitions in this case.
+We have manually inlined the extracted code for simplicity. The full extracted
+code is available in the artefact accompanying this paper, and an interested
+reader can check that simple inlining would produce the following function:
+% The reader can check
+% in our artefact that the extracted code can be trivially inlined to produce the
+% following:
 \begin{ocamlcode}

### 2025-03-24 paper/paper.tex
Cuts/edits to get paper within 16 page limit.

@@
 processes that data. This relationship manifests in the form of structured
-recursion schemes which are widely used in functional languages such Haskell.
+recursion schemes, which are widely used in functional languages such Haskell.
 Canonical examples are folds (catamorphisms), which consume data, and unfolds
@@
 Furthermore, folds and unfolds can be shown to capture a wide range of recursion
-schemes such as primitive recursion, mutual recursion, dynamic programming
+schemes, such as primitive recursion, mutual recursion, dynamic programming
 algorithms, polymorphic recursion, and recursion with accumulators. However, for
@@
 \label{sec:recursion-schemes}
-Recursion schemes provide an abstract way to consume and generate data. We now
-proceed onto describing how to formalise hylomorphisms in Coq. We first
-formalise container functors as a tool to represent polynomial functors
+%Recursion schemes provide an abstract way to consume and generate data.
+We now consider the formalisation hylomorphisms in Coq. We first formalise
+container functors as a tool to represent polynomial functors
 (Section~\ref{sec:containers}). Then we formalise algebras for container
@@
 Overall, our proof is shorter than alternative implementations, and of
-comparable complexity\footnote{The QSort definition and correctness proof in the
-equations
-package is under 200LOC --
-\href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
-and other online proofs of correctness are of similar complexity
-\href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
+comparable complexity.
+%% Commenting out footnote to save space, since we mention the Coq QSort in related work already...
+% \footnote{The QSort definition and correctness proof in the
+% equations
+% package is under 200LOC --
+% \href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
+% and other online proofs of correctness are of similar complexity
+% \href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
 % and can be found in the

### 2025-03-24 paper/paper.tex
minor

--- 2025-03-24 paper/paper.tex
@@
 
-Hylomorphisms are highly versatile, non-structural recursive algorithms that are
-capable of implementing structural recursion as well as other known recursion
-schemes like mutual recursion, accumulators, primitive recursios, and
-course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. However, this
-versatility comes with a trade-off: because hylomorphisms are not inherently
-structurally recursive, special attention is required when encoding them in a
-type theory. Specifically, the ``divide'' phase of the algorithm must be
-recursive, meaning the input needs to be broken down into smaller parts, which
-do not necessarily align with the structure of the input data. On the other
-hand, a key benefit is the reusability of termination proofs: once the
-correctness of the divide phase is established, the same proof can be reused for
-any other conquer phase we choose. Moreover, in the case of structural
+Hylomorphisms are very powerful non-structural recursive algorithms that can be
+used to implement structural recursion, as well as many other known recursion
+schemes such as mutual recursion, accumulators, primitive recursion and
+course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. This power
+comes at a price. Since hylomorphisms are not necessarily structural, extra-care
+is needed when encoding them in a type theory. In particular, the divide part of
+the algorithm must be \emph{recursive} that is the input needs to be split into
+smaller parts (not necessarily following the structure of the input data).
+On the other hand, a key benefit is the reusability of termination proofs: once
+the correctness of the divide phase is established, the same proof can be reused
+for any other conquer phase we choose. Moreover, in the case of structural
 recursion, the divide phase naturally follows the structure of the data, making
@@
 
-Implementations of recursion schemes generally focus on non-terminating
+Implementations of recursion schemes currently focus on non-terminating
 languages (e.g. Haskell) and they would benefit from a type-theoretical
@@
 
-In this work, we mechanise hylomorphisms, reaping the benefits of their
-generality and algebraic properties. Our approach enables users to write
-high-level specifications for their programs, reason about program
-optimisations, and ultimately extract human-readable, executable functional
-code. The extracted code is free from unsafe casts, a priority in our design. We
-focus strongly on avoiding unsafe casts because, even if the extracted code has
-been verified, simple interoperations with them can lead to incorrect behaviour
-or even segmentation faults~\cite{forster:hal-04329663} and, moreover, it
-invalidates the fast-and-loose principle~\cite{DanielssonHJG06}.
+In this work we mechanise hylomorphisms, thus reaping the benefits of their
+generality and their algebraic properties. We do this in a way which allows the
+user to write specifications of their code, reason about program optimisations,
+and eventually extract humanly readable functional code which can be run. Our
+extracted code does not contain unsafe casts. We focus strongly on avoiding
+unsafe casts because, even if the extracted code has been verified, simple
+interoperations with them can lead to incorrect behaviour or even segmentation
+faults~\cite{forster:hal-04329663} and, moreover, it invalidates the
+fast-and-loose principle~\cite{DanielssonHJG06}.
 
@@
 
-We list the contributions of this paper.
-First, in Section~\ref{sec:category-theory}, we provide the theoretical background on recursion schemes and hylomorphisms.
-We then provide a Coq formalisation of \emph{hylomorphisms} (Section~\ref{sec:recursion-schemes}) that (1) is \emph{fully axiom-free}; (2) allows the extraction of idiomatic and efficient code; and (3) can use regular Coq equalities to do program calculation, derive correct implementations, and apply optimisations.
-In Section~\ref{sec:extraction}, we apply this framework to formalise practical examples of divide-and-conquer, dynamic programming, and mutual recursion algorithms.
-Additionally, we verify the short-cut fusion optimisation and present the extracted optimised code.
-
-%Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}\dcas{I SHOULD
-%  ANONYMISE THIS}.
+We list the contributions of this paper. First, in
+Section~\ref{sec:category-theory}, we provide the theoretical background on
+recursion schemes and hylomorphisms. We then provide a Coq formalisation of
+\emph{hylomorphisms} (Section~\ref{sec:recursion-schemes}) that (1) is
+\emph{fully axiom-free}; (2) allows the extraction of idiomatic and efficient
+code; and (3) can use regular Coq equalities to do program calculation, derive
+correct implementations, and apply optimisations. In
+Section~\ref{sec:extraction}, we apply this framework to formalise practical
+examples of divide-and-conquer, dynamic programming, and mutual recursion
+algorithms. Additionally, we verify the short-cut fusion optimisation and
+present the extracted optimised code.
 
@@
 
-Furthermore, folds and unfolds can be shown to capture a wide range of recursion
-schemes such as primitive recursion, mutual recursion, dynamic programming
-algorithms, polymorphic recursion, and recursion with accumulators. However, for
-divide-and-conquer algorithms, folds need to be generalised to hylomorphisms,
-which provide the ultimate basic building block for any other recursion scheme.
-To do this, we look at recursion schemes from the point of view of category
-theory.
+Furthermore folds and unfolds can be shown to capture a wide
+range of recursion schemes such as primitive recursion, mutual recursion,
+dynamic programming algorithms, polymorphic recursion, recursion with
+accumulators and so on. However, for divide-and-conquer algorithms folds need to
+be generalised to hylomorphisms which provide the ultimate basic building block
+for any other recursion scheme. This is done by looking at recursion schemes
+from the point of view of category theory.
 
@@
 $\cons : A \times \List{A} \to \List{A}$. The $\List{-}$ type is a
-\emph{functor} $\Set \to \Set$, in particular, an \emph{endofunctor}, mapping objects and arrows in $\Set$ to $\Set$
-itself.
+\emph{functor} $\Set \to \Set$, in particular, an \emph{endofunctor}, mapping
+objects and arrows in $\Set$ to $\Set$ itself.
 Its functorial action $\List{f} : \List{A} \to \List{B}$ is given by
-$\List{f}(\nil(*)) = \nil(*)$ and $\List{f}(\cons(a,xs)) = \cons(f(a),
-\List{f}(xs))$. Notice that the definition $\List{f}$ is well-defined as it
-recursively calls on a smaller argument.  Similarly, the set of streams $\Str{A}
-\iso A \times \Str{A}$ is the greatest set generated by the constructor $\cons :
-A \times \Str{A} \to \Str{A}$. The maps $\head : \Str{A} \to A$ and $\tail :
-\Str{A} \to \Str{A}$ can be easily constructed from the isomorphism.
+$\List{f}(\nil(*)) = \nil(*)$ and
+$\List{f}(\cons(a,xs)) = \cons(f(a), \List{f}(xs))$. Notice that the definition
+$\List{f}$ is well-defined as it recursively calls on a smaller argument.
+Similarly, the set of streams $\Str{A} \iso A \times \Str{A}$ is the greatest
+set generated by the constructor $\cons : A \times \Str{A} \to \Str{A}$. The
+maps $\head : \Str{A} \to A$ and $\tail : \Str{A} \to \Str{A}$ can be easily
+constructed from the isomorphism.
 
@@
 of infinite streams, that is the greatest solution to the domain equation
-$\Str{A} \iso A \times \Str{A}$.  The uniqueness property of terminal $B$-coalgebras states that for any other $B$-coalgebra
-$(X, c_{X})$ there exists a unique $B$-coalgebra homomorphism into the terminal
-coalgebra $(\nu B, \OutIso)$ which is denoted by $\ana{c_{X}}$ and pronounced
-``\emph{anamorphism}'' or ``\emph{unfold}''. We spell out the uniqueness property:
+$\Str{A} \iso A \times \Str{A}$.  The uniqueness property of terminal
+$B$-coalgebras states that for any other $B$-coalgebra $(X, c_{X})$ there exists
+a unique $B$-coalgebra homomorphism into the terminal coalgebra
+$(\nu B, \OutIso)$ which is denoted by $\ana{c_{X}}$ and pronounced
+``\emph{anamorphism}'' or ``\emph{unfold}''. We spell out the uniqueness
+property:
 \begin{equation}
@@
 \]
-We can model this functor by using a set $S$ defined by $A \uplus B \uplus C$ which distinguishes the three
-cases, and a function $P$ which tells us the number of occurrences of $X$ in
-each case: for $s = (1,a) \mapsto \emptyset$; for $s = (2, b) \mapsto 1$ the singleton set and  for $s = (3, c) \mapsto 2$ the $2$-element set.
+We can model this functor by using a set $S$ defined by $A + B + C$
+
+which distinguishes the three cases, and a function $P$ which tells us the
+number of occurrences of $X$ in each case: for $s = (1,a) \mapsto \emptyset$;
+for $s = (2, b) \mapsto 1$ the singleton set and  for $s = (3, c) \mapsto 2$ the
+$2$-element set.
 

### 2025-03-24 paper/paper.tex
Natural transformations

@@
 constructed from the isomorphism.
+
+Given two functors $F, G : \CatC \to \CatD$, a \emph{natural transformation} is
+a map between functors is a family of morphisms $\phi_{X} : FX \to GX$ indexed
+by the objects $X \in \Obj(\CatC)$ and such that it is \emph{natural} in $X$,
+that is $ G(f) \cdot \phi_{X} = \phi_{Y} \cdot F(f)$. Intuitively, a natural
+transformation transforms the structure of a functor into the structure of
+another functor without assuming what is the type of data contained in them, for
+example, a program $f : \texttt{List }X \to \texttt{Maybe} X$ which returns
+nothing if the list is empty and the head of the list otherwise needs not to
+know the type of the data.
 
@@
 accompanying this paper, we formalise both mergesort and quicksort, but in this
-section, we focus solely on quicksort.
-
-The recursive structure of quicksort is given by the following
-functor
+section, we focus solely on quicksort. The recursive structure of quicksort is
+given by the following functor
 \begin{coqcode}
@@
 \end{coqcode}
-where \coq{Lmap times_two} is a list map function defined as a hylomorphism,
-and \coq{times_two} multiplies every element of the list by two.
-We can use Coq's generalised rewriting and \coq{hylo_fusion_l} to fuse
-\coq{times_two} into \coq{hylo merge qsplit} which gives the program
-\coq{hylo (merge \o natural times_two) qsplit}.
-In this definition, \coq{natural} defines a natural transformation by
-applying \coq{times_two} to the shapes thus multiplying every pivot by two.
+where \coq{Lmap times_two} is a list map function defined as a hylomorphism, and
+\coq{times_two} multiplies every element of the list by two. We can use Coq's
+generalised rewriting and \coq{hylo_fusion_l} to fuse \coq{times_two} into
+\coq{hylo merge qsplit} which gives the program \coq{hylo (merge \o natural times_two) qsplit}.
+In this definition, \coq{natural} defines a natural transformation by applying
+\coq{times_two} to the shapes thus multiplying every pivot by two.
+
 %% Our formalisation contains a proof that \coq{natural} is indeed a natural

### 2025-03-24 paper/paper.tex
Containers

@@
 \newcommand{\foldr}{\operator{foldr}}
+
+\newcommand{\vcomp}{\circ}
 
@@
 \]
-We can model this functor by using a set $S$ defined by $A + B + C$
-
-which distinguishes the three cases, and a function $P$ which tells us the
-number of occurrences of $X$ in each case: for $s = (1,a) \mapsto \emptyset$;
-for $s = (2, b) \mapsto 1$ the singleton set and  for $s = (3, c) \mapsto 2$ the
-$2$-element set.
-
-To formalise this we need  a function $p : P \to S$ and a function
-$q : P \times X \to P$.  Here, $S$ serves as a \emph{shape object} that indexes
-the different cases of the functor, while $P$ represents the positions where
-elements of $X$ appear. The projection $p: P \to S$ assigns each position to its
-corresponding shape, and the map $q: P \to P \times X$ describes how these
-positions are instantiated using elements of $X$. The functor associated with
-this data is defined on objects using a dependent sum (to be precise a coend):
-% \[
-% F(X) = S \times_P (P \times X),
-% \]
-% which can also be written as a dependent sum (a coend to be precise)
+We can model this functor by using a \emph{container}~\cite{AbbottAG05}. This is a set $S$
+of \emph{shapes}, for example, $S = A + B + C$ for the example above, together with a function $P(s)$ denoting the exponent of $X$ for each shape.
+Then a container extension is defined as the sum of all maps
+$P(s) \to X$
 \[
@@
 \]
-where  \( P(s) = p^{-1}(s) \) called the fiber of \( p \) over \( s \). Given a
-function \( f: X \to Y \), the functor acts by post-composition, mapping an
-element \( (s, p, x) \) in \( F(X) \) to \( (s, p, f(x)) \) in \( F(Y) \),
-ensuring functoriality.
+The functorial action
+of $F$ is given by post-composition, mapping an element \( (s, g) \) in
+\( F(X) \) to \( (s, f \vcomp g ) \) in \( F(Y) \), for $f : A \to B$.
 

### 2025-03-24 paper/paper.tex
Containers

@@
 \label{sec:containers}
-\todo{FILL OUT}{}
-The type $\forall s : S.\; P\;s \to X$ would force us to add
-the functional extensionality axiom. Thus, we require that all the types used in
-our definitions are \emph{setoids}, where we use \coq{x =e y} to represent
-setoid equality. All morphisms need to be \emph{proper} morphisms that respect
-setoid equalities. In our development, we use
-\begin{coqcode}
-Notation "A ~> B" := (* proper morphisms  s.t. forall x y, x =e y -> f x =e f y *).
-\end{coqcode}
-We add an automatic coercion from \coq{A ~> B} to \coq{A -> B} and provide
-tactics to automatically discharge proofs that morphisms are proper,  for some
-simple cases.
+
+A direct encoding of containers as presented in
+Section~\ref{sec:category-theory} requires the use of the functional
+extensionality axiom, and axiom K to deal with equalities of objects of type
+$P(s) \to X$. Furthermore, extraction will need to insert unsafe casts for
+anything of type $P(s)$.  To avoid these problems, we use \emph{setoids}, and
+encode families of positions using \emph{decidable validity predicates}.
+
+In our formalisation, we require that every type is a setoid, where \coq{x =e y}
+denotes setoid equality. All morphisms need to be \emph{proper} morphisms that
+respect setoid equalities. We use special notation for such morphisms, \coq{A ~>
+  B}, and we add an implicit coercion from \coq{A ~> B} to \coq{A -> B}.  We
+also provide tactics to automatically discharge proofs that morphisms are
+proper,  for some simple cases.
 
@@
 %% decidable validity predicates unexplained -- do not remove!
-We define containers in terms of shapes, \emph{base} positions
-(\coq{APos : Set}), and \emph{decidable validity predicates} that specify when a
-base position is valid in a shape. Since validity predicates are in \coq{Prop},
-they will be erased during extraction, and since they are decidable, they
-do not require axiom K to deal with heterogeneous equalities.
-Container extensions are defined as
+We define containers in terms of shapes, \emph{base} positions (\coq{APos :
+  Set}), and \emph{decidable validity predicates} that specify when a base
+position is valid in a shape. Since validity predicates are in \coq{Prop}, they
+will be erased during extraction, and since they are decidable, they do not
+require axiom K to deal with heterogeneous equalities.  Container extensions are
+defined as
 \begin{coqcode}
@@
 \end{coqcode}
-
 %% Make sure you do not cut text so much that it ends up being inconsistent with

### 2025-03-24 paper/paper.tex
Knapsack

--- 2025-03-24 paper/paper.tex
@@
 \subsection{Knapsack}\label{sec:knap}
-We show now the formalisation and extraction of \emph{dynamorphisms} for
-dynamic programming. We use the
-knapsack example from~\cite{HinzeWG15}. Dynamorphisms build a memoisation table
-that stores intermediate results, alongside the current computation. The
-algebra used to define a dynamorphism can access this memoisation table to
-speed up computation.  First, we define memoisation tables in terms of
-containers.
-\begin{coqcode}
-Definition MemoShape : Type := A * Sg.
-Definition MemoPos := Pg.
-Instance Memo : Cont MemoShape MemoPos := { valid := valid \o pair (snd \o fst) snd }.
+We show now show the formalisation and extraction of \emph{dynamorphisms} for
+dynamic programming using the knapsack example from~\cite{HinzeWG15}. In
+general, dynamorphisms build a memoisation table which stores intermediate
+results alongside the current computation. The algebra used to define a
+dynamorphism can access this memoisation table to speed up computation.
+First, we define memoisation tables in terms of containers:
+\begin{coqcode}
+Instance Memo : Cont (A * Sg) Pg := { valid := valid \o pair (snd \o fst) snd }.
 Definition Table := LFix Memo.
 \end{coqcode}
-Memoisation tables are the least fixed point of the container defined by shapes
-\coq{A * Sg} and positions \coq{Pg}, given a a container \coq{G : Cont Sg Pg}.
+Given a type of shapes \coq{Sg} and positions \coq{Pg}, the container
+\coq{Memo : Cont (A * Sg) Pg}.  The least fixed-point of which is the least
+fixed-point of the
+
+given a container \coq{G : Cont Sg Pg}
+
 We define a function to insert elements into the memo tables:

### 2025-03-24 paper/paper.tex
Fix
minor

@@
 \label{sec:containers}
-
 A direct encoding of containers as presented in
@@
 extensionality axiom, and axiom K to deal with equalities of objects of type
-$P(s) \to X$. Furthermore, extraction will need to insert unsafe casts for
+$P(s) \to X$. Furthermore, upon extraction Coq will need to insert unsafe casts for
 anything of type $P(s)$.  To avoid these problems, we use \emph{setoids}, and
@@
 In our formalisation, we require that every type is a setoid, where \coq{x =e y}
-denotes setoid equality. All morphisms need to be \emph{proper} morphisms that
-respect setoid equalities. We use special notation for such morphisms, \coq{A ~>
-  B}, and we add an implicit coercion from \coq{A ~> B} to \coq{A -> B}.  We
-also provide tactics to automatically discharge proofs that morphisms are
-proper,  for some simple cases.
+denotes setoid equality and that all morphisms need to be \emph{proper}
+morphisms that respect setoid equalities. We use special notation  \coq{A ~> B}
+for proper morphisms and we add an implicit coercion from \coq{A ~> B} to \coq{A -> B}.  We also provide tactics to automatically discharge proofs that morphisms are proper,  for some simple cases.
 
@@
 %% decidable validity predicates unexplained -- do not remove!
-We define containers in terms of shapes, \emph{base} positions (\coq{APos :
-  Set}), and \emph{decidable validity predicates} that specify when a base
-position is valid in a shape. Since validity predicates are in \coq{Prop}, they
-will be erased during extraction, and since they are decidable, they do not
+We define containers in terms of shapes, \emph{base} positions
+(\coq{APos : Set}), and \emph{decidable validity predicates} that specify when a
+base position is valid in a shape. Since validity predicates are in \coq{Prop},
+they will be erased during extraction, and since they are decidable, they do not
 require axiom K to deal with heterogeneous equalities.  Container extensions are

### 2025-03-24 paper/paper.tex
Section 4.2

@@
 morphisms that respect setoid equalities. We use special notation  \coq{A ~> B}
-for proper morphisms and we add an implicit coercion from \coq{A ~> B} to \coq{A -> B}.  We also provide tactics to automatically discharge proofs that morphisms are proper,  for some simple cases.
+for proper morphisms and we add an implicit coercion from \coq{A ~> B} to
+\coq{A -> B}.  We also provide tactics to automatically discharge proofs that
+morphisms are proper,  for some simple cases.
 
@@
 different recursive programs using regular Coq rewrite tactics.
-
+\newcommand{\Id}{\text{Id}}
 \subsection{Knapsack}\label{sec:knap}
-We show now show the formalisation and extraction of \emph{dynamorphisms} for
-dynamic programming using the knapsack example from~\cite{HinzeWG15}. In
-general, dynamorphisms build a memoisation table which stores intermediate
-results alongside the current computation. The algebra used to define a
-dynamorphism can access this memoisation table to speed up computation.
-First, we define memoisation tables in terms of containers:
+Dynamorphisms are hylomorphisms where the algebra has access to the memory table
+of the result of all previous recursive calls. To implement this we need to
+formalise the idea of a memory table which is in our case simply represented by
+the cofree comonad. For a functor $G : \CatC \to \CatC$ the cofree comonad is
+defined as the greatest solution to the following equation
+\[
+  G_{\infty} A \iso A \times G (G_{\infty} A)
+\]
+In words, these are the (possibly) infinite trees which branch out with shape
+$G$. For example, for the identity functor $\Id_{\infty} A$ is the type of
+infinite streams. For $G X = 1 + X$, the type $G_{\infty} A$ is isomorphic to
+$i \times (1 + G_{\infty} A)$. This is the type representing our memory table
+with the operation $\head_{A} : G_{\infty} A$ being the result of the previous
+recursive call. Notice that in the case of a recursive $G$-coalgebra this type
+only contains finite $G$-trees and so we can view $G_{\infty}A$ as the finite
+$G$-trees thus being able to inspect all the previous recursive calls. This has
+been expounded thoroughly by Hinze et al.~\cite{HinzeWG15}.
+
+In our framework, we formalise $G$ as a container with shape \coq{Sg} and
+position \coq{Pg}, then \coq{Memo} is the least fixed-point of the functor
+$F X = A \times GX$:
 \begin{coqcode}
@@
 \end{coqcode}
-Given a type of shapes \coq{Sg} and positions \coq{Pg}, the container
-\coq{Memo : Cont (A * Sg) Pg}.  The least fixed-point of which is the least
-fixed-point of the
-
-given a container \coq{G : Cont Sg Pg}
-
-We define a function to insert elements into the memo tables:
-\begin{coqcode}
-Definition Cons : A * App G Table ~> App Memo Table := (* *)
-\end{coqcode}
-And two functions, to inspect the head of a table, and remove an
-element:
-\begin{coqcode}
+The operations of the cofree comonad \coq{Memo} are the \coq{Cons}, the \coq{headT} and the \coq{tailT}:
+\begin{coqcode}
+Definition consT : A * App G Table ~> Table := (* *)
 Definition headT : Table ~> A := (* *)
@@
 \end{coqcode}
-These tables map ``paths'' in the least fixed point of \coq{Memo} to elements
-of type \coq{A}. For example, if \coq{G} is a list-generating functor, these
-paths will be natural numbers. 
-Using these definitions, a dynamorphism is defined as follows:
+Finally, we define a dynamorphism as a hylomorphism from a type \coq{B} into a
+type \coq{Table} post-composed with the \coq{headT} map which outputs the last
+result from the memory table:
 \begin{coqcode}
 Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
-  := headT \o hylo (l_in \o Cons \o pair a id) c.
-\end{coqcode}
-Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
-memo table. The definition of the algebra can use this table to lookup
-elements, instead of triggering a further recursive call. Elements are inserted
-into the memoisation table by the use of \coq{Cons} to the result of applying
-the algebra. The algebra for the dynamorphism looks up the previously computed
-elements to produce the result, thus saving the corresponding recursive calls:
-\begin{coqcode}
-Fixpoint memo_knap table wvs :=
-  match wvs with | nil => nil | h :: t =>
-      match lookupT (Nat.pred (fst h)) table with
-      | Some u => (u + snd h)%sint63 :: memo_knapsack table t
-      | None => memo_knapsack table t
-      end
-  end.
-Definition knapsack_alg (wvs : list (nat * int))
-  (x : App NatF (Table NatF int)) : int :=
-  match x with | MkCont sx kx => match sx with
-  | inl tt => fun kx => 0%sint63
-  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
-  end kx end.
-Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
-\end{coqcode}
-Given the recursive coalgebra for \coq{nat}, \coq{out_nat}, knapsack is defined
-as
+  := headT \o hylo (consT \o pair a id) c.
+\end{coqcode}
+Note the map \coq{App G Table ~> A} corresponding to a map $G G_{\infty}A \to A$ is not a $G$-algebra.
+To recover the $G$-algebra we pair it with the identity map and  so we need to
+transmogrify it by pairing it with the identity map and post-composing it with
+\coq{consT}. This algebra uses this table to lookup
+elements, instead of triggering a further recursive call, then elements are inserted
+into the memoisation table by the use of \coq{consT} to the result of applying
+the algebra. In our example this algebra is called \coq{knapsackA} (see accompanying code).
+% \begin{coqcode}
+% Fixpoint memo_knap table wvs :=
+%   match wvs with | nil => nil | h :: t =>
+%       match lookupT (Nat.pred (fst h)) table with
+%       | Some u => (u + snd h)%sint63 :: memo_knapsack table t
+%       | None => memo_knapsack table t
+%       end
+%   end.
+% Definition knapsack_alg (wvs : list (nat * int))
+%   (x : App NatF (Table NatF int)) : int :=
+%   match x with | MkCont sx kx => match sx with
+%   | inl tt => fun kx => 0%sint63
+%   | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
+%   end kx end.
+% Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
+% \end{coqcode}
+Finally, given the recursive coalgebra \coq{out_nat} on the natural numbers, we can define knapsack using the recursion scheme for dynamorphisms:
 \begin{coqcode}
@@
 \end{coqcode}
-While the extracted code has the expected performance characteristics, Coq's
-code extraction mechanism is unable to inline several definitions in this case.
-We have manually inlined the extracted code for simplicity. The full extracted
-code is available in the artefact accompanying this paper, and an interested
-reader can check that simple inlining would produce the following function:
+The full extracted code coming from this specification
+is available in the artefact accompanying this paper, and an interested reader
+can check that simple inlining would produce the following function:
 % The reader can check
@@
 \end{ocamlcode}
-Note how the recursive calls of \ocaml{f} build the memoisation table, and how
+Note that in the extracted code, the recursive calls to \ocaml{f} build the memoisation table, and that
 this memoisation table is used to compute the intermediate results in

### 2025-03-24 paper/paper.tex
First draft

@@
 denoted by $1$, is the object such that for any other object $A$ there is a
-unique arrow $A \xrightarrow{!} 1$. Initial and terminal object are unique up-to
+unique arrow $A \xrightarrow{!} 1$. As a result of the uniqueness properties, initial and terminal objects are unique up-to
 isomorphism.
@@
 \end{coqcode}
-For convenience, we package recursive coalgebras with proofs that they terminate
-for all inputs:
+For convenience, we equip recursive coalgebras with an additional proof of termination:
 \begin{coqcode}
@@
 \end{coqcode}
-Finally, we show that recursive hylomorphisms are the unique solution to the
-hylomorphism equation.
+Finally, we show that recursive hylomorphisms are the unique solution to the following
+hylomorphism equation:
 \begin{coqcode}
@@
 computes the function on every element by traversing the list only once. The
-composition of these two programs is given by the following definition:
+composition of these two programs is given by the following:
 \begin{coqcode}
@@
 is available in the artefact accompanying this paper, and an interested reader
-can check that simple inlining would produce the following function:
+can check that simple inlining would produce the following:
 % The reader can check
@@
 \subsection{Shortcut Deforestation}\label{sec:shortcut}
-The final case study we consider is shortcut deforestation on lists.  Shortcut
-deforestation can be expressed succinctly in terms of hylomorphisms and their
-laws~\cite{TakanoM95}. In particular, given a function:
+We now showcase the shortcut deforestation optimisation on lists.  This is used
+to remove the creation of intermediate data structures by fusing multiple
+recursive traversals into one.  For example, we can use it to show the following
+code from Takano and
+Meijer~\cite{TakanoM95}
+\begin{coqcode}
+Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
+\end{coqcode}
+which uses three different hylomorphisms can be fused into one.
+Here   \coq{sf1} is the composition of \coq{length},
+\coq{Lmap f} and \coq{append ys} which are, in particular,
+catamorphisms.
+
+To prove our goal we use the so called \emph{acid rain} theorem which states
+that given a parametric function of the following type:
 \begin{coqcode}
@@
 \end{coqcode}
-We can conclude, by parametricity, that
-\begin{coqcode}
-  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
-\end{coqcode}
-This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
-\textbf{not} provable in Coq without adding the parametricity
-axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
-However, we prove a specific version of this theorem for the container whose
-least fixed point is a list, and use it to encode the example from Takano and
-Meijer~\cite{TakanoM95}:
-\begin{coqcode}
-Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
-\end{coqcode}
-Here, we are defining function \coq{sf1} as the composition of \coq{length},
-\coq{Lmap f} and \coq{append ys}. Functions \coq{length} and \coq{Lmap} are
-catamorphisms. Function \coq{append ys} is also a catamorphism that
-appends \coq{ys} to an input list. It is defined by applying an algebra to
-every cons node of a list, and applying a catamorphism with the input algebra
-to \coq{ys} in the nil case:
+we can conclude, by parametricity, that the following equation holds:
+\begin{coqcode}
+  hylo a l_out \o hylo (s l_in) c =e hylo (s a) c
+\end{coqcode}
+Unfortunately, this is not provable in Coq without adding the parametricity
+axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}, but we can prove it for specific
+functors, e.g. the functor of lists.
+
+Instead, we prove the acid rain theorem of the specific parametric function
+\coq{tau}. This essentially is a recursive function on lists which uses the
+algebra for both the recursive step and for the base case where it is used in
+the hylomorphism to continue the recursion:
 \begin{coqcode}
@@
   end kx end.
+\end{coqcode}
+Our acid rain theorem is stated as follows:
+\begin{coqcode}
+  hylo a l_out \o hylo (tau l l_in) c =e hylo (tau l a) c
+\end{coqcode}
+
+Now we use \coq{tau} to define the \coq{append} function as follows:
+\begin{coqcode}
 Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
@@
 Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
-\coq{ListF} container. We apply the hylo fusion law repeatedly, unfold
-definitions, and simplify in our specification for \coq{sf1}:
-\begin{coqcode}
-Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
-  rewrite hylo_map_fusion, <- acid_rain. simpl; reflexivity.
-Defined.
-\end{coqcode}
-From this, we extract the following OCaml code:
+\coq{ListF} container. The acid rain theorem allows us to fuse this definition
+of append with the maps \coq{Lmap} and \coq{length} which is the optimisation we
+wanted in the first place. The following is the extracted OCaml code from the
+optimised program:
 \begin{ocamlcode}
@@
 \end{ocamlcode}
-
-\noindent
-We then prove that \ocaml{length} fuses with
+Notice how the \coq{Lmap} is optimised away and the function \coq{length} is
+applied to both the arguments of the append function.
+
+As a second example, we prove that \ocaml{length} fuses with
 the naive quadratic \ocaml{reverse} function:
@@
 Definition sf2 : Ext (length \o reverse).
-  calculate.  unfold length, reverse. rewrite hylo_fusion_l.
-  2:{ (* Rewrite into the fused version *) }
-  simpl; reflexivity.
-Defined.
 \end{coqcode}
@@
 \end{ocamlcode}
+Here by fusing the hylomorphism for the reverse function with the length
+function we obtain the original length function.
 
@@
 require an external proof.
+
+\newpage
 

### 2025-03-24 paper/paper.tex
Fixes

@@
 capable of implementing structural recursion as well as other known recursion
-schemes like mutual recursion, accumulators, primitive recursios, and
+schemes like mutual recursion, accumulators, primitive recursion, and
 course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. However, this
@@
 
-Instead, we prove the acid rain theorem of the specific parametric function
-\coq{tau}. This essentially is a recursive function on lists which uses the
-algebra for both the recursive step and for the base case where it is used in
-the hylomorphism to continue the recursion:
+We prove a specialised version of the acid rain theorem that fuses
+hylomorphisms defined in terms of the following function \coq{tau} that maps
+list algebras to list algebras.  This essentially is a function on lists which
+uses the algebra for both the recursive step and for the base case where it is
+used in the hylomorphism to continue the recursion:
 \begin{coqcode}
@@
 Certain functions that are not structurally recursive can be reformulated using
-a nonstandard approach to achieve structural recursion.  For exammple, the
+a nonstandard approach to achieve structural recursion.  For example, the
 mergesort in Coq's standard library uses an ``explicit stack of pending merges''
@@
 The fusion laws that we formalise in this paper are applied in the more general
-context of compilers for functional programming lanmguages.  Fusion has been
+context of compilers for functional programming languages.  Fusion has been
 applied to recursive tree traversals, in order to reduce the number of times

### 2025-03-24 paper/paper.tex
minor

--- 2025-03-24 paper/paper.tex
@@
 \end{coqcode}
-Note the map \coq{App G Table ~> A} corresponding to a map $G G_{\infty}A \to A$ is not a $G$-algebra.
-To recover the $G$-algebra we pair it with the identity map and  so we need to
-transmogrify it by pairing it with the identity map and post-composing it with
-\coq{consT}. This algebra uses this table to lookup
-elements, instead of triggering a further recursive call, then elements are inserted
-into the memoisation table by the use of \coq{consT} to the result of applying
-the algebra. In our example this algebra is called \coq{knapsackA} (see accompanying code).
+Note the map \coq{App G Table ~> A} corresponding to a map $G G_{\infty}A \to A$
+is not a $G$-algebra.  To recover the $G$-algebra we pair it with the identity
+map and  so we need to transmogrify it by pairing it with the identity map and
+post-composing it with \coq{consT}. This algebra uses this table to lookup
+elements, instead of triggering a further recursive call, then elements are
+inserted into the memoisation table by the use of \coq{consT} to the result of
+applying the algebra. In our example this algebra is called \coq{knapsackA} (see
+accompanying code).
 % \begin{coqcode}
@@
 % \end{coqcode}
-Finally, given the recursive coalgebra \coq{out_nat} on the natural numbers, we can define knapsack using the recursion scheme for dynamorphisms:
+Finally, given the recursive coalgebra \coq{out_nat} on the natural numbers, we
+can define knapsack using the recursion scheme for dynamorphisms:
 \begin{coqcode}
@@
 \end{coqcode}
-The full extracted code coming from this specification
-is available in the artefact accompanying this paper, and an interested reader
-can check that simple inlining would produce the following:
+The full extracted code coming from this specification is available in the
+artefact accompanying this paper, and an interested reader can check that simple
+inlining would produce the following:
 % The reader can check
@@
 \end{ocamlcode}
-Note that in the extracted code, the recursive calls to \ocaml{f} build the memoisation table, and that
-this memoisation table is used to compute the intermediate results in
-\ocaml{memo_knap}, which is finally discarded to produce the final result.
+Note that in the extracted code, the recursive calls to \ocaml{f} build the
+memoisation table, and that this memoisation table is used to compute the
+intermediate results in \ocaml{memo_knap}, which is finally discarded to produce
+the final result.
 

### 2025-03-24 paper/paper.tex
Final submission to ITP

@@
 
-\acknowledgements{I want to thank \dots}%optional
+% \acknowledgements{I want to thank \dots}
+% optional
 
@@
 Suppose, for example, that we want to write a program that sorts a list of
-integers, and then multiplies by 2 all its elements. In OCaml, we may write this
+integers and multiplies them by 2 at the same time. In OCaml, we may write this
 function directly:
@@
 \end{minted}
-This program is derived by \emph{fusing} a regular \emph{quicksort} OCaml
-implementation, with
-\ocaml{map (fun x -> x * 2)}.
-A lot of program calculations have this pattern: start from a simple
-specification, e.g.\ $\mathsf{map}\;(\lambda x.\;x \times 2) \circ
-\mathsf{sort}$, and use program equivalences and algebraic laws to rewrite it to
-an optimised version. We will revisit a similar example in
-Section~\ref{sec:sorting}.
+
+This program is the same as composing a \emph{quicksort} OCaml implementation,
+with the function \ocaml{map (fun x -> x * 2)} which is a known program
+optimisation called \emph{fusion}.  A lot of program calculations have this
+pattern: start from a simple specification, e.g.\
+$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
+equivalences and algebraic laws to rewrite it to an optimised version. We will
+revisit a similar example in Section~\ref{sec:sorting}.
 

### 2025-03-24 paper/paper.tex
Title change

@@
 % \documentclass{llncs}
-\documentclass[a4paper,anonymous, UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
 % \documentclass[llncs, a4paper, UKenglish, cleveref, autoref, thm-restate]{article}
@@
 
-\title{Mechanising Hylomorphisms for Extraction of Executable Code}
+\title{Program Optimisations via Hylomorphisms for Extraction of Executable Code}
 

### 2025-03-25 paper/paper.tex
Typo fixes

--- 2025-03-24 paper/paper.tex
@@
 \begin{coqcode}
-Inductive Tshape L A := | Leaf | Node (ELEM : A).
+Inductive Tshape A := | Leaf | Node (ELEM : A).
 Inductive Tpos := | Lbranch | Rbranch. 
@@
   match x with
-  | nil => a_leaf tt
+  | nil => a_leaf
   | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in a_node h l r
@@
   match sx return (Container.Pos sx -> _) -> _ with
-  | Leaf _ _ => fun _ => nil
-  | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
+  | Leaf _ => fun _ => nil
+  | Node h => fun k => List.app (k (posL h)) (h :: k (posR h))
   end kx.
@@
 \end{coqcode}
-The resulting code extracted from \coq{hylo merge qsplit} is similar to a
-hand-written implementation:
+The resulting extracted code is similar to a hand-written implementation:
 \begin{ocamlcode}
@@
 infinite streams. For $G X = 1 + X$, the type $G_{\infty} A$ is isomorphic to
-$i \times (1 + G_{\infty} A)$. This is the type representing our memory table
+$A \times (1 + G_{\infty} A)$. This is the type representing our memory table
 with the operation $\head_{A} : G_{\infty} A$ being the result of the previous

### 2025-06-16 paper/paper.tex
Fixing typos in Section 2

@@
   | h :: t -> let (l, r) = partition (fun x -> x < h) t in
-  sort_times_two l @ (h * 2) :: sort_times_two t
+  sort_times_two l @ (h * 2) :: sort_times_two r
 \end{minted}
@@
 Given two functors $F, G : \CatC \to \CatD$, a \emph{natural transformation} is
-a map between functors is a family of morphisms $\phi_{X} : FX \to GX$ indexed
+is a family of morphisms $\phi_{X} : FX \to GX$ indexed
 by the objects $X \in \Obj(\CatC)$ and such that it is \emph{natural} in $X$,
@@
 $X \mapsto 1 + A \times X$. An $F$-algebra is a set $B$ together with a
-structure map $[b, i] : 1 + A \times B \to B$. The initial $F$-algebra is
+structure map $[base, step] : 1 + A \times B \to B$. The initial $F$-algebra is
 clearly the set of lists $\List{A}$ and the catamorphism associated with the
@@
 $(\nu B, \OutIso)$ with $\OutIso$ being the final $B$-coalgebra witnessing the
-isomorphism and $\OutIso : \nu B \to B\nu B$ being its inverse.
-
-For example, the terminal coalgebra for the functor $BX = L \times X$ is the set
-of infinite streams, that is the greatest solution to the domain equation
+isomorphism and $\OutIso^{\circ} : B\nu B \to \nu B$ being its inverse.
+
+For example, the terminal coalgebra for the functor $BX = A \times X$ is the set
+of infinite streams over the set $A$, that is the greatest solution to the domain equation
 $\Str{A} \iso A \times \Str{A}$.  The uniqueness property of terminal
@@
 
-An example of a recursive coalgebra is the \coq{partition} function in quicksort
-which destructures a list into a pivot and two sub-list and as long as the
-sub-lists are smaller the partitioning function still yields a unique solution to
-the recursion scheme.  The uniqueness property of the hylomorphisms yields the following
-fusion laws:
+An example of a recursive coalgebra is the \coq{partition} function used in
+quicksort, which has the type $\text{List } A \to B(\text{List } A)$, where the
+functor $B$ is defined as $BX = X \times A \times X$. This function deconstructs
+a list by selecting a pivot element of type $A$ and splitting the remaining
+elements into two sublists.
+
+Notice that the initial algebra for lists corresponds to the functor
+$F X = 1 + A \times X$ and its associated algebra map (or its inverse, viewed as
+a coalgebra), $\InOp$ has the type $\text{List } A \to F(\text{List
+} A)$. Importantly, \coq{partition} is not this map, which means that it does
+not arise from the standard initial algebra structure, and thus catamorphisms
+here cannot be used to define recursive functions.
+
+Nevertheless, since \coq{partition} always produces sublists that are strictly
+smaller than the input list, it still supports a well-founded recursion scheme,
+ensuring the existence of a unique solution. The uniqueness property of the
+hylomorphisms yields the following fusion laws:
 \begin{align}
@@
 %Recursion schemes provide an abstract way to consume and generate data.
-We now consider the formalisation hylomorphisms in Coq. We first formalise
+We now consider the formalisation of hylomorphisms in Coq. We first formalise
 container functors as a tool to represent polynomial functors
@@
 \end{coqcode}
-Finally, we prove that catamorphisms satisfy their universality property (see
+Finally, we prove that catamorphisms satisfy their universal property (see
 Section~\ref{sec:category-theory}):

### 2025-06-16 paper/paper.tex
I've made one pass

@@
 a particular data structure, in this case $\operator{List}$s, we can generalise
-further these ideas to account for generic (co)inductive data types and generic
+these ideas further to account for generic (co)inductive data types and generic
 algorithms operating on them.
@@
 categories\footnote[1]{For presentation purposes we shall not deal with size
-  issues and assume all the categories are locally small.}. The initial object
-in a category, denoted by $0$, is the object such that for any other object $A$
-there is a unique arrow $0 \xrightarrow{!} A$.  Dually, the terminal object,
+  issues and assume all the categories are locally small.}. The initial object (when it exists), denoted by $0$,
+is the object such that for any other object $A$
+there is a unique arrow $0 \xrightarrow{!} A$.  Dually, the terminal object (when it exists),
 denoted by $1$, is the object such that for any other object $A$ there is a
@@
 by the objects $X \in \Obj(\CatC)$ and such that it is \emph{natural} in $X$,
-that is $ G(f) \cdot \phi_{X} = \phi_{Y} \cdot F(f)$. Intuitively, a natural
-transformation transforms the structure of a functor into the structure of
-another functor without assuming what is the type of data contained in them, for
+that is $ G(f) \cdot \phi_{X} = \phi_{Y} \cdot F(f)$, for all morphisms
+$f : X \to Y$. Intuitively, a natural transformation is akin to a polymorphic
+function transforming the structure of a functor into the structure of another
+functor without assuming what is the type of data contained in them.  For
 example, a program $f : \texttt{List }X \to \texttt{Maybe} X$ which returns
-nothing if the list is empty and the head of the list otherwise needs not to
-know the type of the data.
+nothing if the list is empty and the head of the list otherwise can work for all
+types $X$ uniformly because it needs not to inspect the data inside the list.
 
@@
 \label{sec:algebras}
-For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(X,a_{X})$
+For an endofunctor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(X,a_{X})$
 where $X$ is an object of the category called the \emph{carrier} of the algebra
@@
 $F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra homomorphism, denoted
-by $\cata{\beta} : (\mu F, \InIso) \to (X, a_{X})$ and pronounced
+by $\cata{\alpha_{X}} : (\mu F, \InIso) \to (X, a_{X})$ and pronounced
 ``\emph{catamorphism}'' or ``\emph{fold}'' satisfying:
@@
 For example, for a set of states $X$ and a finite set of labels $L$ we can
-define a labelled transition system (LTS) on $X$ as a function $X \to BX$
+define a labelled transition system (LTS)~\cite{WinskelM95} on $X$ as a function $X \to BX$
 implementing the \emph{transition system} with $BX = L \times X$.  In
@@
 where $l \in L$ is the observable action and $x_{2} \in X$ is the next state.
-In this case, we can even set some notation for the transition map:
-\[
-  x \xrightarrow{l} y \eqdef c_{X}(x) = (l, y)
-\]
+
 The category of $B$-coalgebras, denoted $B\Coalg$, is the category where objects
@@
 For example, the terminal coalgebra for the functor $BX = A \times X$ is the set
-of infinite streams over the set $A$, that is the greatest solution to the domain equation
+of infinite streams over the set $A$, that is the greatest solution to the equation
 $\Str{A} \iso A \times \Str{A}$.  The uniqueness property of terminal
@@
 A coalgebra $(C,c)$ is \emph{recursive} if for every algebra $(A, a)$ there is a
-\emph{unique} hylo $(C,c ) \rightarrowtail (A, a)$. We denote these type of
-hylos by $\hylo{c}{a}$.
+\emph{unique} hylo $(C,c ) \rightarrowtail (A, a)$~\cite{CaprettaUV04,AdamekMM19}. We denote these
+type of hylos by $\hylo{c}{a}$.
 
@@
 Using the hylomorphism fusion laws, we can prove the well-known \emph{deforestation
-optimisation}, also known as the \emph{composition
+optimisation}~\cite{Wadler90}, also known as the \emph{composition
 law}~\cite{DBLP:conf/ifl/HinzeHJ10}. This is when two consecutive recursive
 computations, one that builds a data structure, and another one that consumes
-it, can be fused together into a single recursive definition. This, in turn,
-allows us to prove that a recursive hylomorphism is the composition of a
-catamorphism and a recursive anamorphism.
+it, can be fused together into a single recursive definition. % This, in turn,
+% allows us to prove that a recursive hylomorphism is the composition of a
+% catamorphism and a recursive anamorphism.
 
@@
 Recursion schemes are formulated using generic functors that encapsulate the
-structure of the recursive operator. However, since not all functors have the
+structure of the recursive operator. However, since not all functors have
 suitable fixed points, it is necessary to restrict our focus to \emph{strictly
@@
 (Section~\ref{sec:containers}). Then we formalise algebras for container
-extensions (Section~\ref{sec:coq-algebras}), then we formalise coalgebras
-(Section~\ref{sec:coq-coalg}) and then we put together these notions to
+extensions (Section~\ref{sec:coq-algebras}). In
+Section~\ref{sec:coq-coalg} we formalise coalgebras and put together these notions to
 formalise recursive coalgebras and hylomorphisms
@@
 \end{coqcode}
-From this definition the universality property falls out:
+From this definition the universal property falls out:
 \begin{coqcode}
@@
 
-% \subsection{Recursive Anamorphisms}
-% Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
-% inductive data type in the following way.  Intuitively, a recursive coalgebra
-% can be applied only finitely many times, therefore when this is applied to an
-% anamorphism the only possible traces it can produce from the seed function are
-% the finite ones. We denote this special kind of recursion scheme \emph{recursive
-% anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
-% can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
-% exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
-% algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
-% from the universality property of the hylomorphism and the fact that recursive
-% anamorphisms satisfy the same equation.
-%
-% We now show how this is proven in the Coq development. First we define a
-% recursive anamorphism as a function into \coq{LFix F}:
-% \begin{coqcode}
-% Definition rana_f__ `{setoid A} (c : Coalg F A)
-%   : forall x : A, RecF c x -> LFix F
-%   := fix f (x : A) (H : RecF c x) :=
-%        let hx := c x in
-%        LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
-%
-% Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F := ...
-% \end{coqcode}
-% We omit the boilerplate details to obtain the setoid morphism.
-% The recursive anamorphism has the following universality property:
-% \begin{coqcode}
-% Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
-%     : f =e rana h <-> f =e l_in \o fmap f \o h.
-% \end{coqcode}
-% We can now show that every hylomorphism is composed by a recursive anamorphism
-% followed by a catamorphism:
-% \begin{coqcode}
-% Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
-%   (g : Alg F B) (h : RCoalg F A)
-% : cata g \o rana h =e hylo g h.
-% \end{coqcode}
-% This result will allow us to \emph{calculate} programs by providing small, easy
-% to verify implementations, and combining them into efficient recursive functions
-% by using Coq's proofs of hylomorphism laws as optimisation passes.
-
 \subsubsection{On the subtype of finite elements}
@@
 Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
-is isomorphic its corresponding the inductive data type \coq{LFix F}.
+is isomorphic to its corresponding inductive data type \coq{LFix F}.
 We show this by
@@
 however, we use \emph{recursive} anamorphisms, defined as hylomorphisms on the
-recursive recursive coalgebra \coq{c}, and the initial \coq{F}-algebra:
+ recursive coalgebra \coq{c}, and the initial \coq{F}-algebra:
 \begin{coqcode}
@@
 \coq{times_two} multiplies every element of the list by two. We can use Coq's
-generalised rewriting and \coq{hylo_fusion_l} to fuse \coq{times_two} into
+generalised rewriting~\cite{Sozeau09} and \coq{hylo_fusion_l} to fuse \coq{times_two} into
 \coq{hylo merge qsplit} which gives the program \coq{hylo (merge \o natural times_two) qsplit}.
@@
 Note the map \coq{App G Table ~> A} corresponding to a map $G G_{\infty}A \to A$
-is not a $G$-algebra.  To recover the $G$-algebra we pair it with the identity
-map and  so we need to transmogrify it by pairing it with the identity map and
+is not a $G$-algebra.  To recover the $G$-algebra we need to transform it by pairing it with the identity map and
 post-composing it with \coq{consT}. This algebra uses this table to lookup
@@
 \end{coqcode}
-which uses three different hylomorphisms can be fused into one.
+which uses that three different hylomorphisms can be fused into one.
 Here \coq{sf1} is the composition of \coq{length},
@@
 using a recursion scheme in which termination is entirely enforced by its
-typing. This is a significant advance, since it avoids \emph{completely} the
+typing. This is a significant advance, since it \emph{completely} avoids the
 need for termination proofs. Their work differs from ours in that they require
@@
 
-Another related family transformations are worker/wrapper
+Another related family of transformations are worker/wrapper
 transformations~\cite{gill09}, which have been used in the context of optimising

### 2025-06-17 paper/paper.tex
Addressed most of the reviewers comments

@@
 {0000-0002-6939-4189}%
-{}
+{EP/Y00339X/1, EP/T014512/1}
 \author{Marco Paviotti}%
@@
 extensionality axiom, and axiom K to deal with equalities of objects of type
-$P(s) \to X$. Furthermore, upon extraction Coq will need to insert unsafe casts for
+$P(s) \to X$. Axiom K states that a predicate on equality proofs holds for all
+such proofs if it holds for reflexivity. This would be needed to reason about
+the equality of any two $f : P(s_1) \to X$ and $g : P(s_2) \to X$, when $s_1 =
+s_2$. In Coq, if we know $(s_1 , f) = (s_2 , g)$, we cannot extract a proof that
+$f = g$ unless we assume axiom K, or the equality on the type of $s_1$ and $s_2$
+is decidable.  Furthermore, upon extraction Coq will need to insert unsafe casts
+for
 anything of type $P(s)$.  To avoid these problems, we use \emph{setoids}, and
@@
 \end{coqcode}
-The initial \coq{F}-algebra can be defined as the inductive type with a finite
-number of \coq{App F}
+We define the initial \coq{F}-algebra as the inductive type which is constructed
+by applying \coq{App F} finitely many times.
 \begin{coqcode}
@@
 The \coq{LFix F} setoid equality is the least fixed point of the \coq{App F}
-setoid equality. We define smart constructors for the isomorphism of least fixed
+setoid equality, i.e. if \coq{LFixR F} represents the setoid equality of
+\coq{LFix F}, and we have two \coq{x y : LFix F}, then \coq{LFixR F x y} iff
+\begin{coqcode}
+shape x =e shape y /\ (forall p p', projT1 p = projT1 p' -> LFixR F (cont x p) (cont y p'))
+\end{coqcode}
+We define smart constructors for the isomorphism of least fixed
 points as proper morphisms:
@@
 \end{coqcode}
-
 Catamorphisms are constructed so that they structurally deconstruct the
@@
 \begin{coqcode}
-Definition cata_f (a : Alg F A) : LFix -> A
+Definition cata_f (a : Alg F A) : LFix F -> A
   := fix f (x : LFix) := match x with
@@
 It is easy to prove that catamorphisms respect setoid equalities.  In fact, we
-define it as a proper morphism:
-\begin{coqcode}
-cata : forall `{setoid A}, Alg F A ~> LFix ~> A
+define it as a setoid morphism between the setoid of \coq{F}-algebras and the
+setoid of functions from \coq{LFix F} to \coq{A}:
+\begin{coqcode}
+cata : forall `{setoid A}, Alg F A ~> (LFix F ~> A)
 \end{coqcode}
@@
 
-As we mentioned in Section~\ref{sec:category-theory}, given an $F$-algebra and
-$F$-coalgebra, the hylomorphism is the unique solution (when it exists) to the
-equation
+As we mentioned in Section~\ref{sec:category-theory}, given an $F$-algebra $a$ and
+$F$-coalgebra $c$, $f$ is a hylomorphism if it satisfies
 \[
@@
 \end{coqcode}
-For convenience, we equip recursive coalgebras with an additional proof of termination:
+\coq{RecF} represents that a coalgebra will eventually terminate on an input.
+The base case takes place when the set of valid positions in the
+container extension returned by \coq{h x} is empty.  For convenience, we equip
+recursive coalgebras with an additional proof of termination:
 \begin{coqcode}
@@
 \end{coqcode}
-Recursive hylomorphisms are implemented in Coq recursively on the structure of
-the proof that the anamorphism terminates on an input:
+Recursive hylomorphisms are implemented by structural recursion on the proof
+that coalgebra \coq{c} eventually terminates for some input \coq{x}
+(\coq{RecF c x}).
 \begin{coqcode}
@@
 In other words, in this development we have defined recursive anamorphisms on
-inductive data types.  We might have as well defined them on the subtype of
+inductive data types.  We might as well have defined them on the subtype of
 finite elements of coinductive data types using a predicate which states when an
@@
 \end{coqcode}
+\coq{FinF} represents finiteness since it must contain a base case with no
+positions, after a finite number of applications of \coq{g_out}.
 Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
@@
  recursive coalgebra \coq{c}, and the initial \coq{F}-algebra:
-\begin{coqcode}
-Definition rana c := hylo l_in c.
-\end{coqcode}
+\coq{hylo l_in c}.
 The main advantage of this definition is that recursive anamorphisms compose
@@
 \begin{coqcode}
-Inductive ITreeF N X := i_leaf | i_node (n : N) (l r : X)
+Inductive ITreeF A X := i_leaf | i_node (n : A) (l r : X)
 \end{coqcode}
@@
 Inductive Tpos := | Lbranch | Rbranch. 
+\end{coqcode}
+The validity predicate, \coq{valid_f} below, simply specifies that positions are
+only valid in \coq{Node}. Our tactics automatically discharge the necessary
+proofs to construct the respective setoid morphism.
+\begin{coqcode}
+Definition valid {A} (x : Tshape A * Tp) : bool :=
+  match x with | (Node _ _, _)  => true | _ => false end.
 \end{coqcode}
@@
 are automatically discharged by our tactics, but we still have to prove that the
-coalgebra \coq{qsplit} is recursive. In this specific case we have to prove that
-the two sublists are smaller than the input list. In general, we are required to
-prove that the input is related to the data inside the structure functor, that
-is that the element of the carrier of the coalgebra is related to the content of
-the functor by some well-founded relation. Here we give a specialised property, the
-full general property can be found in the formalisation:
-\begin{coqcode}
-Definition respects_relation `{setoid A} (c : Coalg F B) B} (R : B -> B -> Prop)
-:= forall x (p : Pos (shape (c x))), R (length (cont (c x) p)) (length x).
+coalgebra \coq{qsplit} is recursive. In general, we know that if the input is
+related to the data inside the structure functor returned by the coalgebra by
+some \emph{well-founded relation} \coq{R}, then the coalgebra is recursive. In
+this specific case, it is sufficient to show that the two sublists are smaller
+than the input list.
+\begin{coqcode}
+  forall x (p : Pos (shape (qsplit x))), length (cont (qsplit x) p) < length x.
 \end{coqcode}
@@
 %% Commenting out footnote to save space, since we mention the Coq QSort in related work already...
-% \footnote{The QSort definition and correctness proof in the
-% equations
-% package is under 200LOC --
-% \href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
-% and other online proofs of correctness are of similar complexity
-% \href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
-% and can be found in the
-% formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].
+\footnote{The QSort definition and correctness proof in the
+equations
+package is under 200LOC --
+\href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
+and other online proofs of correctness are of similar complexity
+\href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
+and can be found in the
+formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].
 
@@
 \begin{coqcode}
-Instance Memo : Cont (A * Sg) Pg := { valid := valid \o pair (snd \o fst) snd }.
-Definition Table := LFix Memo.
+Instance Memo A : Cont (A * Sg) Pg := { valid := valid \o pair (snd \o fst) snd }.
+Definition Table A := LFix (Memo A).
 \end{coqcode}
@@
 \begin{coqcode}
-Definition consT : A * App G Table ~> Table := (* *)
-Definition headT : Table ~> A := (* *)
-Definition tailT : Table ~> App G Table := (* *)
+Definition consT A : A * App G (Table A) ~> Table A := (* *)
+Definition headT A : Table A ~> A := (* *)
+Definition tailT A : TableA  ~> App G (Table A) := (* *)
 \end{coqcode}
@@
 \begin{coqcode}
-Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
+Definition dyna A (a : App G (Table A) ~> A) (c : RCoalg G B) : B ~> A
   := headT \o hylo (consT \o pair a id) c.
 \end{coqcode}
-Note the map \coq{App G Table ~> A} corresponding to a map $G G_{\infty}A \to A$
+Note the map \coq{App G (Table A) ~> A} corresponding to a map $G G_{\infty}A \to A$
 is not a $G$-algebra.  To recover the $G$-algebra we need to transform it by pairing it with the identity map and

### 2025-06-26 paper/paper.tex

@@
   contribution of this paper is that this formalisation is fully
-  \emph{axiom-free} allowing for the extraction of safe, idiomatic functional
+  \emph{axiom-free}, and it allows the extraction of safe, idiomatic functional
   code. We exemplify the framework by formalising a series of algorithms based
@@
 To preserve the generality of hylomorphisms, we encode data structures using
-polynomial functors. To ensure that the extracted code is runnable, we make a
-conscious effort to stay axiom-free and avoid using indexed types in the
-definition of our recursion schemes.
+polynomial functors. To ensure that the extracted code is runnable, avoid using
+indexed types in the definition of our recursion schemes. To strengthen the
+results of this paper, we avoid all axioms.
 
@@
 for
-anything of type $P(s)$.  To avoid these problems, we use \emph{setoids}, and
+anything of type $P(s)$, because OCaml requires the input value to have a type that is identical to the function parameter, and
+in the general case this cannot be done in OCaml.
+To avoid these problems, we use \emph{setoids}, and
 encode families of positions using \emph{decidable validity predicates}.
@@
 \begin{coqcode}
-shape x =e shape y /\ (forall p p', projT1 p = projT1 p' -> LFixR F (cont x p) (cont y p'))
+shape x =e shape y /\ (forall p p', projT1 p = projT1 p'
+   -> LFixR F (cont x p) (cont y p'))
 \end{coqcode}
@@
 As we stated earlier, a solution to this equation does not exist for an
-arbitrary  algebra/coalgebra pair and, in fact, this definition cannot be
-accepted by Coq.
-
-In order to find a solution we restrict ourselves to the so-called
+arbitrary  algebra/coalgebra pair and, in fact, a recursive function definition
+like this cannot be directly accepted by Coq.
+
+In order to find the unique solution we restrict ourselves to the so-called
 \emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  We mechanise
@@
 \end{coqcode}
+We can directly write mergesort as \coq{hylo merge qsplit}, or derive it by
+program calculation from the composition of \coq{cata merge}, and the
+recursive anamorphism \coq{hylo l_in qsplit}.
 The resulting extracted code is similar to a hand-written implementation:
@@
 \coq{merge \o fmap (hylo merge qsplit) \o qsplit} using hylomorphism
-uniqueness~\ref{sec:coq-rec-coalgebras}. The rest of the proof is standard.
+uniqueness (Section~\ref{sec:coq-rec-coalgebras}). The rest of the proof is standard.
 Overall, our proof is shorter than alternative implementations, and of
-comparable complexity.
+comparable complexity%
 %% Commenting out footnote to save space, since we mention the Coq QSort in related work already...
@@
 \href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
-and can be found in the
-formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].
+% and can be found in the
+% formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].
 
@@
 different recursive programs using regular Coq rewrite tactics.
+
+It would be straightforward to prove the correctness of the fused version, by
+simply proving that \coq{Lmap times_two} preserves the order of the elements.
 \newcommand{\Id}{\text{Id}}
@@
 Finally, given the recursive coalgebra \coq{out_nat} on the natural numbers, we
-can define knapsack using the recursion scheme for dynamorphisms:
+can define knapsack, given a list of pairs of weights and values \coq{wvs},
+using the recursion scheme for dynamorphisms:
 \begin{coqcode}
@@
 to remove the creation of intermediate data structures by fusing multiple
-recursive traversals into one.  For example, we can use it to show the following
+recursive traversals into one.  For example, we can use it to show that the following
 code from Takano and
-Meijer~\cite{TakanoM95}
+Meijer~\cite{TakanoM95}, which uses three different hylomorphisms, can be fused into one:
 \begin{coqcode}
@@
 \end{coqcode}
-which uses that three different hylomorphisms can be fused into one.
 Here \coq{sf1} is the composition of \coq{length},
@@
 catamorphisms.
-
-To prove our goal we use the so called \emph{acid rain} theorem which states
-that given a parametric function of the following type:
+To prove our goal we use the so called \emph{acid rain} theorem~\cite{TakanoM95}
+which states that given a parametric function of the following type:
 \begin{coqcode}
@@
 their formalisation.  Mu et al.\ formalise hylomorphisms in Agda, and can do
-relational program transformation~\cite{DBLP:journals/jfp/MuKJ09}. It is unclear
-how to extract runnable code from their definitions, and they do not seem to
-formalise hylomorphisms in terms of generic functors and datatypes.
+relational program transformation~\cite{DBLP:journals/jfp/MuKJ09}. Their paper
+does not discuss extracting runnable code from their encodings, and they do not
+seem to formalise hylomorphisms in terms of generic functors and datatypes.
 

### 2025-06-27 paper/paper.tex
Paper camera ready

@@
 
-\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+% \ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10003752.10010124.10010125.10010127</concept_id>
+       <concept_desc>Theory of computation~Functional constructs</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10003790.10011740</concept_id>
+       <concept_desc>Theory of computation~Type theory</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10010124.10010138.10010142</concept_id>
+       <concept_desc>Theory of computation~Program verification</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Functional constructs}
+\ccsdesc[300]{Theory of computation~Type theory}
+\ccsdesc[500]{Theory of computation~Program verification}
 
@@
 function directly:
+
+\pagebreak
 \begin{minted}{ocaml}

### 2025-06-27 paper/paper.tex
Types

@@
 
-  This paper presents the first Coq formalisation of \emph{hylomorphisms}
+  This paper presents the first Rocq formalisation of \emph{hylomorphisms}
   allowing for the mechanisation of all recognised recursive algorithms. The key
@@
 To preserve the generality of hylomorphisms, we encode data structures using
-polynomial functors. To ensure that the extracted code is runnable, avoid using
-indexed types in the definition of our recursion schemes. To strengthen the
-results of this paper, we avoid all axioms.
+polynomial functors. To ensure that the extracted code is idiomatic and does not
+contain unsafe casts, we avoid using indexed types in the definition of the
+recursion schemes, and make sure that the definitions can be inlined by Rocq's
+extraction mechanism. To strengthen the results of this paper, we avoid all
+axioms.
 
@@
 Section~\ref{sec:category-theory}, we provide the theoretical background on
-recursion schemes and hylomorphisms. We then provide a Coq formalisation of
+recursion schemes and hylomorphisms. We then provide a Rocq formalisation of
 \emph{hylomorphisms} (Section~\ref{sec:recursion-schemes}) that (1) is
 \emph{fully axiom-free}; (2) allows the extraction of idiomatic and efficient
-code; and (3) can use regular Coq equalities to do program calculation, derive
+code; and (3) can use regular Rocq equalities to do program calculation, derive
 correct implementations, and apply optimisations. In
@@
 A solution to this equation does not exist for an arbitrary algebra and
-coalgebra pair and, in fact, a definition like this cannot be accepted by Coq.
+coalgebra pair and, in fact, a recursive function definition like this is not
+accepted by Rocq.
 
@@
 quicksort, which has the type $\text{List } A \to B(\text{List } A)$, where the
-functor $B$ is defined as $BX = X \times A \times X$. This function deconstructs
+functor $B$ is defined as $BX = 1 + X \times A \times X$. This function deconstructs
 a list by selecting a pivot element of type $A$ and splitting the remaining
@@
 not arise from the standard initial algebra structure, and thus catamorphisms
-here cannot be used to define recursive functions.
+here cannot be used to define this recursive function.
 
@@
 of $F$ is given by post-composition, mapping an element \( (s, g) \) in
-\( F(X) \) to \( (s, f \vcomp g ) \) in \( F(Y) \), for $f : A \to B$.
+\( F(A) \) to \( (s, f \vcomp g ) \) in \( F(B) \), for $f : A \to B$.
 
@@
 %Recursion schemes provide an abstract way to consume and generate data.
-We now consider the formalisation of hylomorphisms in Coq. We first formalise
+We now consider the formalisation of hylomorphisms in Rocq. We first formalise
 container functors as a tool to represent polynomial functors
@@
 the equality of any two $f : P(s_1) \to X$ and $g : P(s_2) \to X$, when $s_1 =
-s_2$. In Coq, if we know $(s_1 , f) = (s_2 , g)$, we cannot extract a proof that
+s_2$. In Rocq, if we know $(s_1 , f) = (s_2 , g)$, we cannot extract a proof that
 $f = g$ unless we assume axiom K, or the equality on the type of $s_1$ and $s_2$
-is decidable.  Furthermore, upon extraction Coq will need to insert unsafe casts
+is decidable.  Furthermore, upon extraction Rocq will need to insert unsafe casts
 for
@@
 \end{coqcode}
-It is easy to prove that catamorphisms respect setoid equalities.  In fact, we
-define it as a setoid morphism between the setoid of \coq{F}-algebras and the
+It is easy to prove that catamorphisms respect setoid equalities.  We
+define \coq{cata} as a setoid morphism between the setoid of \coq{F}-algebras and the
 setoid of functions from \coq{LFix F} to \coq{A}:
@@
 \[
-  f = a \circ F\; f \circ c
+  f = a \circ F\; f \circ c.
 \]
@@
 arbitrary  algebra/coalgebra pair and, in fact, a recursive function definition
-like this cannot be directly accepted by Coq.
+like this cannot be directly accepted by Rocq.
 
@@
 % standard practice when building general recursive functions in
-% Coq\footnote{\dcas{This technique had a name in the community, but I cannot
+% Rocq\footnote{\dcas{This technique had a name in the community, but I cannot
 % remember. Something predicates}}. The core definitions are the following:
@@
 % Here, \coq{RecF_inv} provides the structurally smaller element for
-% the Coq recursive definition. We then couple coalgebras with a proof that they
+% the Rocq recursive definition. We then couple coalgebras with a proof that they
 % are recursive, and add a coercion from recursive coalgebras to coalgebras.
@@
 % times is straightforward (e.g.\ they are recursive on some inductive type in
-% Coq). However, in many cases, proving that a coalgebra is recursive directly
+% Rocq). However, in many cases, proving that a coalgebra is recursive directly
 % can be tedious and cumbersome. To ease development, we prove that if a
@@
 %  \item Write a specification as a morphism \coq{f : A ~> B}.
-%  \item Use Coq's generalised rewriting and the laws of morphisms to 
+%  \item Use Rocq's generalised rewriting and the laws of morphisms to
 %    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
@@
 %Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
-%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
+%\emph{extensionally equal}. Then, Rocq's extraction mechanism will erase any
 %proofs from the code, and inline our definitions. To ease the application of
@@
 comparable complexity%
-%% Commenting out footnote to save space, since we mention the Coq QSort in related work already...
+%% Commenting out footnote to save space, since we mention the Rocq QSort in related work already...
 \footnote{The QSort definition and correctness proof in the
@@
 package is under 200LOC --
-\href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
+\href{https://github.com/mattam82/Rocq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Rocq-Equations/blob/8.20/examples/quicksort.v},
 and other online proofs of correctness are of similar complexity
@@
 % algebra, since this is where most of the comparisons happen. The recursive
-% coalgebra for Mergesort is defined as the following Coq functions:
+% coalgebra for Mergesort is defined as the following Rocq functions:
 % \begin{coqcode}
@@
 where \coq{Lmap times_two} is a list map function defined as a hylomorphism, and
-\coq{times_two} multiplies every element of the list by two. We can use Coq's
+\coq{times_two} multiplies every element of the list by two. We can use Rocq's
 generalised rewriting~\cite{Sozeau09} and \coq{hylo_fusion_l} to fuse \coq{times_two} into
@@
 a hand-written implementation, and that it has been derived by fusing two
-different recursive programs using regular Coq rewrite tactics.
+different recursive programs using regular Rocq rewrite tactics.
 
@@
 \end{coqcode}
-Unfortunately, this is not provable in Coq without adding the parametricity
+Unfortunately, this is not provable in Rocq without adding the parametricity
 axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}, but we can prove it for specific
@@
 \end{coqcode}
-
 Now we use \coq{tau} to define the \coq{append} function as follows:
@@
 \end{coqcode}
-Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
+Here, \coq{ilist_coalg} is a recursive coalgebra from Rocq lists to the
 \coq{ListF} container. The acid rain theorem allows us to fuse this definition
@@
 %  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
-%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
+%        \item Extraction of Distributed Mobile code in Rocq~\cite{MiculanP12}.
 %\end{itemize}
-% and recursion schemes in Coq
+% and recursion schemes in Rocq
 % (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}). Additionally, we compare our
@@
 There is existing prior work on formalising recursion schemes in a proof
-assistant.  Tesson et al.\ demonstrated the efficacy of leveraging Coq to
+assistant.  Tesson et al.\ demonstrated the efficacy of leveraging Rocq to
 establish an approach for implementing a robust system dedicated to verifying
@@
 lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
-formalised recursion schemes in Coq~\cite{MurataE19}. Their development does not
+formalised recursion schemes in Rocq~\cite{MurataE19}. Their development does not
 include hylomorphisms and dynamorphisms, and relies on the functional
@@
 
-Larchey-Wendling and Monin encode recursion schemes in Coq,
+Larchey-Wendling and Monin encode recursion schemes in Rocq,
 by formalising computational graphs of algorithms~\cite{larchey2022braga}.
@@
 % \label{sec:reccoq}
-Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
+Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Rocq,
 using a recursion scheme in which termination is entirely enforced by its
@@
 a nonstandard approach to achieve structural recursion.  For example, the
-mergesort in Coq's standard library uses an ``explicit stack of pending merges''
+mergesort in Rocq's standard library uses an ``explicit stack of pending merges''
 in order to avoid issues with nonstructural definitions. There is a major
@@
 as a form of mergesort''~\cite{AbreuDHJMS23}.  There are common approaches in
-Coq to deal with termination~\cite{DBLP:journals/pacmpl/SozeauM19,
+Rocq to deal with termination~\cite{DBLP:journals/pacmpl/SozeauM19,
   DBLP:conf/icfp/Sozeau07}, but none of these approaches address program
@@
 compiler~\cite{10.1145/3133900}. The correctness of tree traversal fusion has
-been verified mechanically in Coq~\cite{10.1145/3377555.3377884}, but the
+been verified mechanically in Rocq~\cite{10.1145/3377555.3377884}, but the
 authors do not formalise unfolds and hylomorphisms.
@@
 % \emph{structural decrease} of arguments in the recursive calls of a function.
-% Coq uses such a check, and it works for many classic functional programming
+% Rocq uses such a check, and it works for many classic functional programming
 % patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
@@
 
-% In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
+% In the context of Rocq, Danvy explains the use of fold/unfold lemmas to mechanize
 % equational reasoning about recursive programs~\cite{DANVY_2022}. Fold-unfold
@@
 later extracted into idiomatic and runnable code. Our mechanisation is
-\emph{fully axiom-free}. This formalisation allows us to use Coq as a framework
+\emph{fully axiom-free}. This formalisation allows us to use Rocq as a framework
 for program calculation in a way that is close to common practice in pen and
@@
 formulations to try to improve the performance of the rewriting tactics
-(e.g.\ canonical structures).  Currently, Coq is unable to inline a number of
+(e.g.\ canonical structures).  Currently, Rocq is unable to inline a number of
 trivially inlineable definitions.  We will study alternative definitions, or
-extensions to Coq's code extraction mechanisms to force the full inlining of all
+extensions to Rocq's code extraction mechanisms to force the full inlining of all
 container code that is used in hylomorphisms.  Finally, proving termination
@@
 type $A$, maps everything to it, $X \mapsto A$, only needs one shape ($A$, L9
-of Fig~\ref{fig:poly}), and zero positions (In Coq, \coq{False},
+of Fig~\ref{fig:poly}), and zero positions (In Rocq, \coq{False},
 L10 of Fig~\ref{fig:poly}). The reason why there are zero valid positions in
@@
 \coq{nat + unit * unit}. The positions of this functor
-are \mintinline[breaklines]{Coq}{False + (unit + unit)}, which is isomorphic
+are \mintinline[breaklines]{Rocq}{False + (unit + unit)}, which is isomorphic
 to \coq{unit + unit}, which will only be valid in the RHS.
 
-The Coq value \coq{inl n} of type
+The Rocq value \coq{inl n} of type
 \coq{nat + bool * bool} would look as follows as a container
@@
 
-Similarly, the Coq value \coq{inr (true, false)} as a container
+Similarly, the Rocq value \coq{inr (true, false)} as a container
 extension is as follows:
@@
 \begin{minipage}{.52\textwidth}
-  \begin{minted}[linenos, fontsize=\small]{Coq}
+  \begin{minted}[linenos, fontsize=\small]{Rocq}
 Context `(C1 : Cont S1 P1)
@@
 \begin{minipage}{.47\textwidth}
-  \begin{minted}[linenos, firstnumber=23, fontsize=\small]{Coq}
+  \begin{minted}[linenos, firstnumber=23, fontsize=\small]{Rocq}
 Definition Ss : Type := S1 + S2.
@@
 is to define such equivalence simultaneously to the definition for  \coq{LFix}.
-Such definition would be rejected by Coq. There are other approaches that can
+Such definition would be rejected by Rocq. There are other approaches that can
 circumvent this problem, by separating the definition of least fixed points from
@@
 Furthermore, this extra layer interferes with code extraction, in that it
-requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
+requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Rocq
 may not always able to inline them.

### 2025-07-07 paper/paper.tex
Fixed awkward line break

@@
 optimisation called \emph{fusion}.  A lot of program calculations have this
-pattern: start from a simple specification, e.g.\
-$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
+pattern: start from a simple specification, e.g.
+\\$\mathsf{map}\;(\lambda x.\;x \times 2) \circ \mathsf{sort}$, and use program
 equivalences and algebraic laws to rewrite it to an optimised version. We will

### 2025-07-07 paper/paper.tex
Preparing cameraready sources
identity and associativity

@@
 each object $A$ and for two arrows $A \xrightarrow{f} B$ and $B \xrightarrow{g}
-C$ there always exists an arrow $A \xrightarrow{g\circ f} C$ obeying the
+C$ there always exists an arrow $A \xrightarrow{g\circ f} C$ obeying the identity and
 associativity law. We denote $\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to

### 2025-07-07 zipfile.sh
git archive for paper

@@ -1,4 +1,4 @@
 #!/bin/zsh
 
 rm -rf ./archive && mkdir archive
-git archive --format zip --prefix coq-hylos/ --worktree-attributes HEAD -o archive/coq-hylos.zip
+git archive --format zip --worktree-attributes HEAD -o archive/coq-hylos.zip

### 2025-07-07 paper/paper.tex
minor

@@
 %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{}
+\EventEditors{Yannick Forster and Chantal Keller}
 \EventNoEds{2}
-\EventLongTitle{ITP 2025)}
+\EventLongTitle{16th International Conference on Interactive Theorem Proving (ITP 2025)}
 \EventShortTitle{ITP 2025}
@@
 \EventYear{2025}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
+\EventDate{September 29--October 2, 2025}
+\EventLocation{Reykjavik, Iceland}
 \EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
+\SeriesVolume{352}
+\ArticleNo{32}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 singleton set. The reader who is not accustomed with category theory can assume
-types are sets, giving the intuition that the category $\Set$ can also be viewed
-as the category of (simple) types and programs between them.
+types are sets and programs are functions, giving the intuition that the category $\Set$ can also be viewed as the category of (simple) types and programs between them.
 

### 2025-07-07 paper/paper.tex
Link to code!

@@
 program calculation to apply a fusion optimisation. In the supplemental material
-accompanying this paper, we formalise both mergesort and quicksort, but in this
+accompanying this paper (\url{https://github.com/dcastrop/coq-hylomorphisms}), we formalise both mergesort and quicksort, but in this
 section, we focus solely on quicksort. The recursive structure of quicksort is

### 2025-07-07 paper/paper.tex
Camera ready
Font
minors

--- 2025-07-07 paper/paper.tex
@@
 functor without assuming what is the type of data contained in them.  For
-example, a program $f : \texttt{List }X \to \texttt{Maybe} X$ which returns
-nothing if the list is empty and the head of the list otherwise can work for all
-types $X$ uniformly because it needs not to inspect the data inside the list.
+example, a program $\phi_{X} : \texttt{List }X \to \texttt{Maybe} X$, which returns
+nothing if the list is empty and the head of the list otherwise, does not need to know the content of the list to perform its task, in other words, it can work for all
+types $X$ \emph{uniformly}.
 
@@
 by $\cata{\alpha_{X}} : (\mu F, \InIso) \to (X, a_{X})$ and pronounced
-``\emph{catamorphism}'' or ``\emph{fold}'' satisfying:
+``\emph{catamorphism}'' or ``\emph{fold}''. This is formally stated as follows:
 \begin{equation}

### 2025-07-07 paper/paper.tex

@@
 
-\usepackage{inconsolata}
+% \usepackage{inconsolata}
 % \usepackage{draftwatermark}
@@
 
-%\nolinenumbers %uncomment to disable line numbering
+\nolinenumbers %uncomment to disable line numbering
 
@@
   := fix f x H
-     := match c x as cx return (forall e : Pos (shape cx), RecF c (cont cx e)) -> B with
-        | cx => fun H => a { shape := shape cx ; cont := fun e => f (cont cx e) (H e) }
+     := match c x as cx 
+            return (forall e : Pos (shape cx), RecF c (cont cx e)) -> B with
+        | cx => fun H => 
+            a { shape := shape cx ; cont := fun e => f (cont cx e) (H e) }
         end (RecF_inv H).

### 2025-07-07 paper/paper.tex
Update
ref MiculanP12

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\supplement{}
+
+\supplementdetails[subcategory={Source Code}, cite={}, swhid={}]{Software}{https://github.com/dcastrop/coq-hylomorphisms}
 
@@
 extensions to Rocq's code extraction mechanisms to force the full inlining of all
-container code that is used in hylomorphisms.  Finally, proving termination
-still remains a hurdle. In our framework this reduces to proving that the
-anamorphism terminates in all inputs, and we provide a convenient connection to
-well-founded recursion. Furthermore, recursive coalgebras compose with natural
-transformations, which allows the reuse of a number of core recursive
-coalgebras. A possible interesting future line of work is the use of the
-approach by Abreu et al.~\cite{AbreuDHJMS23} in combination with ours to improve
-code extraction from divide-and-conquer computations whose termination does not
-require an external proof.
-
-\newpage
+container code that is used in hylomorphisms.  Following work by Miculan and
+Paviotti~\cite{MiculanP12} an interesting next step would be to embed the notion
+of hylomorphism into a type theory with a notion of distributed computations
+(e.g. ML5), the resulting extracted code would be then implemented in a
+distributed programming language (e.g. Erlang).
+Finally, proving termination still remains a hurdle. In our framework this
+reduces to proving that the anamorphism terminates in all inputs, and we provide
+a convenient connection to well-founded recursion. Furthermore, recursive
+coalgebras compose with natural transformations, which allows the reuse of a
+number of core recursive coalgebras. A possible interesting future line of work
+is the use of the approach by Abreu et al.~\cite{AbreuDHJMS23} in combination
+with ours to improve code extraction from divide-and-conquer computations whose
+termination does not require an external proof.
 

### 2025-07-07 paper/paper.tex
Hanging words

@@
 singleton set. The reader who is not accustomed with category theory can assume
-types are sets and programs are functions, giving the intuition that the category $\Set$ can also be viewed as the category of (simple) types and programs between them.
+for simplicity that  types may be viewed as sets and programs as functions
+between sets, giving the intuition that the category $\Set$ can also be viewed
+as the category of (simple) types and programs between them.
 
@@
 ``\emph{anamorphism}'' or ``\emph{unfold}''. We spell out the uniqueness
-property:
+property of unfolds:
 \begin{equation}
@@
 setoid equality, i.e. if \coq{LFixR F} represents the setoid equality of
-\coq{LFix F}, and we have two \coq{x y : LFix F}, then \coq{LFixR F x y} iff
+\coq{LFix F}. Let \coq{x y : LFix F}, then \coq{LFixR F x y} iff
 \begin{coqcode}
